##  Attitudes towards abortion


 <a name="q:abortion">*</a> Abortion is a divisive issue in the United States,
particularly among Christians, some of whom believe that abortion is
absolutely forbidden by the Bible. Do attitudes towards abortion
differ among Christian denominations? The data in
[link](http://www.utsc.utoronto.ca/~butler/d29/abortion.txt) come from
the American Social Survey for the years  1972--1974. The variables
are:



* `year`: year of survey, 1972, 1973 or 1974

* `religion`: Christian denomination: Southern Protestant,
other Protestant, Catholic.

* `education`: years of education: low (0--8), medium
(8--12), high (more than 12).

* `attitude` towards abortion (response). There were three abortion
questions in the survey, asking whether the respondent thought that
abortion should  be legal in these three circumstances:


* when there is a strong possibility of a birth defect

* when the mother's health is threatened

* when the pregnancy is the result of rape.

A respondent who responded "yes" to all three questions was
recorded as having a Positive attitude towards abortion; someone who
responded "no" to all three was recorded as Negative, and anyone
who gave a mixture of Yes and No responses was recorded as Mixed.

* `frequency` The number of respondents falling into the
given category combination.




(a) Read in and display the data. Do you seem to have the right
things?


Solution


The columns are aligned with each other but not with the headings,
so `read_table2` it has to be:
```{r }
my_url="http://www.utsc.utoronto.ca/~butler/d29/abortion.txt"
abortion=read_table2(my_url)
abortion
```

     

This looks good. `religion, education, attitude` are correctly
categorical with apparently the right levels, and `frequency`
is a (whole) number.

The only weirdness is that year got read in as a number, because it
*was* a number. `read_table2` had no way of knowing that
we might have wanted to treat this as categorical. But you don't need to
observe that.



(b) We have three choices for fitting logistic regressions:


* `glm` with `family="binomial"`

* `polr` (from package `MASS`)

* `multinom` (from package `nnet`)

Why should we be using `polr` for this analysis?
Explain briefly.


Solution


It all comes down to what kind of response variable we have. Our
response variable `attitude` has *three* categories that can
be put in order: negative, mixed, positive (or the other way
around). So we have an ordinal response, which calls for
`polr`. You can eliminate the other possibilities, because
`glm` requires a *two-*category response and
`multinom` requires a nominal response whose categories do
*not* have a natural order.
In short, "because we have an ordered (ordinal) response". 



(c) `attitude` is text, and needs to be an ordered factor,
with the values in the right order.  Create a vector, using
`c()` if you wish, that contains the factor levels in the
right order, and then create a new `attitude` factor using
the `ordered` function, feeding it first the variable as read
from the file, and second the new ordered levels that you want it to
have.


Solution


This is actually harder to describe than it is to do.
First thing is to decide on the ordering you want. You can go
low to high or high to low (it doesn't matter).
`r tufte::margin_note("The        results might look different which way you go, but it won't        make any *material* difference.")` I'm going negative to
positive. Either way, you want `Mix` in the middle:
```{r }
lev=c("Neg","Mix","Pos")
```

       

Then use the original `attitude` plus this `lev` as
input to `ordered`:

```{r }
abortion = abortion %>%
mutate(attitude.ord=ordered(attitude,lev))
abortion
```

 

I printed out the result to convince myself that I had the right
thing, but you don't have to. Note that `attitude.ord` has the
same values as `attitude`, but is `ord` (that is, an
ordered factor) rather than text.

This is the easy way. To save yourself some typing, you can get the
`attitude` levels from the data:

```{r }
lev = abortion %>% distinct(attitude) %>% pull(attitude)
lev
```

 

or 

```{r }
lev2 = abortion %>% count(attitude) %>% pull(attitude)
lev2
```

 

`distinct` arranges the values in the order it found them in
the data, which is a sensible low-to-high here; `count`
arranges the levels in alphabetical order, which is *not*
sensible, so you'd need to rearrange them, like this:

```{r }
lev3=lev2[c(2,1,3)]
lev3
```

 

or, if you like working with data frames better:

```{r }
lev4 = abortion %>% count(attitude) %>%
slice(c(2,1,3)) %>% pull(attitude)
lev4
```

 

where you'd have to go as far as `count` the first time to find
what to slice, and then run the whole thing the second time.



(d) Fit a model that will predict the correctly-ordered
attitude towards abortion 
from religious denomination and education. Don't forget that each
row of the data file encodes a lot of people. You don't need to
display the results.


Solution


This is a veiled hint to remember the `weights=`
thing. Don't forget to use your `attitude` in the proper
order that you went to such great pains to make in the last part!
```{r }
library(MASS)
abortion.1=polr(attitude.ord~religion+education,
data=abortion,weights=frequency)
```

       



(e) Fit two more models, one containing `religion` only,
and the other containing `education` only. Again, no need to
look at the results.


Solution


This is a standard application of `update`. Note that
this carries along the `data=`, and *also* the
`weights=`, so you don't need either of them again; just
make sure you say what *changes*, and everything else
(including the form of the model) will stay the same:

```{r }
abortion.2=update(abortion.1,.~.-education)
abortion.3=update(abortion.1,.~.-religion)
```

 

The syntax means "take everything in the model abortion.1, and then take out education" for the first one.
If you don't like `update`, you can also copy and paste and edit:

```{r }
abortion.2a=polr(attitude.ord~religion,
data=abortion,weights=frequency)
abortion.3a=polr(attitude.ord~education,
data=abortion,weights=frequency)
```

       




(f) Investigate whether each of `religion` and `education`
should stay in the model, or whether they can be removed. Do this
two ways, using `drop1` and using `anova`. What do you
conclude? 


Solution


I think `drop1` is easier, so let's do that first:
```{r dropping}
drop1(abortion.1,test="Chisq")
```

       

Dropping nothing is rather clearly the best thing to do. Note how easy
that is, because you test both variables at once. If one of them had
been removable, you would have removed it, and then done
`drop1` again to see whether the other one could come out as well.

The `anova` way is more complicated, but sometimes that's the
only way that works. The idea is that you compare the fit of the model
containing both variables with the model obtained by dropping one of
the variables, and see whether you really need the bigger model. This
has to be done twice, once for each explanatory variable you're testing.

First, we test `religion`, by comparing the model with it (and
`education`) and the model without it (with just
`education`, but the other way around (since the smaller model
goes first):
```{r }
anova(abortion.3,abortion.1)
```

       

This P-value is small, so the bigger model is better:
`religion` should stay. (To be sure that you did the right
thing, look in the `anova` output at the two `Model`
lines: the extra thing in the bigger model needs to be the thing
you're testing for.)

Now, we test `education` by comparing the model with both and
the model without `education` (but with `religion` since
that's important):

```{r }
anova(abortion.2,abortion.1)
```

       

Same conclusion: `education` has to stay too.
Note that the P-values in `drop1` and in `anova`, and
also the test statistics (`LRT` and `LR stat.`) are the
same, so it literally doesn't matter which way you do it.



(g) Which of your fitted models is the best? Explain briefly.



Solution


We couldn't take either of the explanatory variables out, so the
model with both variables is best, the one I called
`abortion.1`. 



(h) <a name="part:preds">*</a> 
Obtain predicted probabilities for each attitude towards abortion
for each combination of education level and religious
denomination. To do that, (i) obtain those levels of education and
religion, (ii) make a data frame out of all combinations of them,
(iii) feed your best fitted model and that data frame into
`predict` with any necessary options.


Solution


Use either `distinct` or `count` to get the levels of
`education` and `religion`:
```{r }
educations = abortion %>% distinct(education) %>% pull(education)
religions = abortion %>% distinct(religion) %>% pull(religion)
new=crossing(education=educations,religion=religions)
new
pp=predict(abortion.1,new,type="p")
cbind(new,pp)
```

       



(i) Using your predictions from (<a href="#part:preds">here</a>), describe
the effect of education. (This should be the same for all values of `religion`).


Solution


Pick a religious denomination, say `Cath` since that's
the first one. Focus on the three rows with that
`religion`, here rows 1, 4 and 7.  I'd say the striking
thing is that people with `Low` education are most likely
to have a negative attitude towards abortion. Or you could say
that they are *least* likely to have a positive attitude
towards abortion. That's equally good. Or you can turn it around
and say that the `High` education people are *more*
likely to be in favour of abortion (or less likely to be opposed
to it).
This holds true for all religious denominations, as you can check.
This is one of those cases where I don't mind much what you say,
as long as the data (here the predictions) support it.



(j) Using your predictions from (<a href="#part:preds">here</a>), describe
the effect of `religion`. (This should be the same for all
levels of `education`.)


Solution


As in the previous part: pick a level of `education`, and
then judge the effect of `religion`. 
The `Low` level of education appeared to be the most
opposed to abortion before, so the effects of `religion`
might be the most noticeable there. For those, the Catholics are
most likely to be negatively inclined towards abortion, and the
least likely to be postively inclined. There is not much
difference between the two flavours of Protestant.
Some further thoughts:
What actually surprises me is that you hear (in the media) about
Christians being steadfastly opposed to abortion. This is
something that these data do not support at all: an appreciable
fraction of people from each of the denominations, especially
with medium or high levels of education, are actually in favour
of abortion in all three of the circumstances described in the question.
The models we fitted assume an effect of `religion`
regardless of education, and an effect of `education`
regardless of religion. But it might be that the effect of
education depends on which religious denomination you're looking
at. The way to assess whether that is true is to add an
*interaction*, as you would in analysis of variance. (We
haven't talked about that in this course yet, which is why I
didn't ask you to do it.) In R, `a:b` means 
"the interaction between factors `a` and `b`" and
`a*b` means 
"the main effects of `a` and `b` and their interaction". In our case, therefore, we
should add `education:religion` to our model and test it
for significance:
```{r }
abortion.4=update(abortion.1,.~.+education:religion)
anova(abortion.1,abortion.4)
```

       

The interaction *is* significant, so it should stay. What effect
does it have on the predictions? We've done all the setup earlier, so
we just run `predict` again on the new model:

```{r }
pp=predict(abortion.4,new,type="p")
cbind(new,pp)
```

 

For Southern Protestants and Other Protestants, the effect of
education is as we described it before: the lower the amount of
education, the less favourably inclined towards abortion (that is,
`Neg` is more likely and `Pos` is less likely). This is
still true for Catholics, but the effect of education is less
noticeable: the probabilities of each response, over the levels of
education, vary much less. This, I 
think, is where the interaction comes from: for Catholics, education
has a much smaller effect on attitudes towards abortion than it does
for either of the Protestant denominations. 

I would have liked to have you explore this, but the question was
already too long, so I called a halt where I did.

Again, we should tidy up after ourselves:

```{r }
detach("package:MASS", unload=T)
```

 




