##  Acid rain


 One of the first noticeable effects of climate change was
"acid rain". This is formed by the water vapour in clouds combining
with nitrous oxide and sulfur dioxide, which come from coal and oil
production. How does the acidity of rain affect the acidity of the
soil on which it falls? (This will have an effect on the kind of
plants that can be grown in that soil.) Acidity is measured using the
pH scale, where a pH of 7 is chemically neutral, a number less than 7
is acidic, and a number greater than 7 is alkaline.

An experiment was conducted at the Florida Institute of Food and
Agricultural Sciences, to determine how acidity of rain affects soil
acidity. Experimental plots were irrigated with rainwater that was
prepared to have one of two different pH levels, 3.7 and 4.5. The
acidity of the soil was then measured at three different depths,
0--15, 15--30, and 30--46 centimetres. This was done on three
different dates, in April and June 1991. The data are in
[link](http://ritsokiguess.site/datafiles/acidrain.csv).



(a) Read in and display the data.

Solution


This time, it's a `.csv`:

```{r acidrain-1 }
my_url <- "http://ritsokiguess.site/datafiles/acidrain.csv"
acidrain <- read_csv(my_url)
acidrain
```

 

There are 4 columns, soil depth, date, pH of the rain that was
applied (all explanatory) and soil acidity, the response. You'll note
that we have one soil acidity value per combination of the other
things (it was a designed experiment).

We're going to ignore the date for now, treating the observations on
different dates as replicates.

$\blacksquare$

(b) Draw a grouped boxplot to show how soil acidity depends on
the pH of the rain and the soil depth. (The first time you do this
you might not get the boxplot you want. How can you fix that?)

Solution


The problem is likely to be that either your `x` or your
`fill` for your boxplot is numerical (`rain_pH` is
`dbl`) rather than the categorical variable you need.
Try to use one of the explanatory variables as `x` and the other
one as `fill` (or `colour`):
```{r acidrain-2 }
ggplot(acidrain, aes(x = rain_pH, y = soil_acidity, fill = soil_depth)) +
  geom_boxplot()
```

     

That looks as if it worked, but it didn't. See the boxplot below for
how it's supposed to be.  I need `x` for the boxplot needs to
be categorical. The easiest way to make it such is to wrap it in
`factor`:

```{r acidrain-3 }
ggplot(acidrain, aes(x = factor(rain_pH), y = soil_acidity, fill = soil_depth)) +
  geom_boxplot()
```

     

Even though `soil_depth` looks like numbers, the printout of
the data frame reveals that it is text, so that is OK.

If you prefer, exchange `x` and `fill`:

```{r acidrain-4 }
ggplot(acidrain, aes(fill = factor(rain_pH), y = soil_acidity, x = soil_depth)) +
  geom_boxplot()
```

     
$\blacksquare$

(c) What does your grouped boxplot say about likely
interactions? Explain briefly.

Solution


Think about the effect of one of your explanatory variables, for
each level of the other. For example, in the last plot, the effect
of the rain pH pm on the soil acidity is very small at the largest
and smallest depths, but at the middle soil depth 15--30, the
average (median) soil acidity is a lot less when the rain pH is
*higher* (which seems odd to me). The effect of rain pH being
different according to soil pH is what suggests an interaction
effect. 
Another way of thinking about this is imagining what an
interaction plot would look like. This would be a trace going
through the middle (strictly, mean rather than median) of each set
of boxplots of one colour. In the last boxplot, the red trace
would go close to straight across, while the blue one would dip in
the middle. Not parallel, so suggesting an interaction.
Either approach is good.

$\blacksquare$

(d) Fit an ANOVA with interaction. What do you conclude from
it? You may want to create a column that is the factor version of
`rain_pH` first.

Solution


Following my own hint:
```{r acidrain-5 }
acidrain <- acidrain %>% mutate(frph = factor(rain_pH))
soil.1 <- aov(soil_acidity ~ frph * soil_depth, data = acidrain)
summary(soil.1)
```

 

The interaction is nowhere near significant, perhaps surprisingly. But
bear in mind that there are only 18 observations in total, so each box
on the boxplot is based on *three* observations only. So the
interaction would have to be a lot bigger to be significant.

The usual procedure after finding a non-significant
interaction is to take it out:

```{r acidrain-6 }
soil.2 <- update(soil.1, . ~ . - frph:soil_depth)
summary(soil.2)
```

 

The P-values have come down a bit (the result of gaining two df for
error while the error SS only got a tiny bit bigger), but not nearly
enough to be significant.


$\blacksquare$


