##  Air pollution


 The data in
[link](http://www.utsc.utoronto.ca/~butler/d29/airpollution.csv) are
measurements of air-pollution variables recorded at 12 noon on 42
different days at a location in Los Angeles. The file is in
`.csv` format, since it came from a spreadsheet.  Specifically,
the variables (in suitable units), in the same order as in the data
file, are:



* wind speed

* solar radiation

* carbon monoxide

* Nitric oxide (also known as nitrogen monoxide)

* Nitrogen dioxide

* Ozone

* Hydrocarbons


The aim is to describe pollution using fewer than these seven variables.



(a) Read in the data and demonstrate that you have the right
number of rows and columns in your data frame.


Solution


This is a `.csv` file, so:
```{r }
my_url <- "http://www.utsc.utoronto.ca/~butler/d29/airpollution.csv"
air <- read_csv(my_url)
air
```

     

There should  be 42 rows (for the 42 days), and 7 columns (for the 7
variables), and there are. 
    


(b) <a name="part:fivenum">*</a> 
Obtain a five-number summary for each variable. You can do this in
one go for all seven variables.


Solution


Like this (the cleanest):
```{r }
air %>% map_df(~ quantile(.))
```

     

using `map` from `purrr`: "for each (column of) `air`, calculate the quantiles of it"..
I have to figure out how to
identify which number from the five number summary each of these is,
but in this case you can easily figure it out since the min is the
smallest and the max has to be the biggest in each column.

Or, with some more work, this:

```{r }
air %>%
  pivot_longer(everything(), names_to="xname", values_to="x") %>% 
  nest(-xname) %>%
  mutate(q = map(data, ~ enframe(quantile(.$x)))) %>%
  unnest(q) %>%
  pivot_wider(names_from=name, values_from=value) %>% 
  select(-data)
```

This time, I put the percentiles in columns and variable names in rows (because it seemed to be easier that way).
 

just like the weather problem (in fact, *exactly* like the weather
problem).

    


(c) Obtain a principal components analysis. Do it on the
correlation matrix, since the variables are measured on different
scales. You don't need to look at the results yet.


Solution


This too is all rather like the previous question:
```{r }
air.1 <- princomp(air, cor = T)
```

     
    


(d) Obtain a scree plot. How many principal components might be
worth looking at? Explain briefly. (There might be more than one
possibility. If so, discuss them all.)


Solution


`ggscreeplot` the thing you just obtained, having loaded
package `ggbiplot`: 
```{r }
ggscreeplot(air.1)
```

     

There is a technicality here, which is
that `ggbiplot`, the package, loads `plyr`, which
contains a lot of the same things as `dplyr` (the latter is a
cut-down version of the former). If you load `dplyr` and
*then* `plyr` (that is to say, if you load the
`tidyverse` first and then `ggbiplot`), you will end up
with trouble, and probably the wrong version of a lot of functions. To
avoid this, load `ggbiplot` *first*, and then you'll be
OK. 

Now, finally, we might diverge from the previous question. There are
actually *two* elbows on this plot, at 2 and at 4, which means
that we should entertain the idea of either 1 or 3 components. I would
be inclined to say that the elbow at 2 is still "too high up" the
mountain --- there is still some more mountain below it.

The points at 3 and 6 components look like elbows too, but they are
*pointing the wrong way*. What you are looking for when you
search for elbows are points that are the end of the mountain and the
start of the scree. The elbows at 2 (maybe) and 4 (definitely) are
this kind of thing, but the elbows at 3 and at 6 are not.
    


(e) Look at the `summary` of the principal components
object. What light does this shed on the choice of number of
components? Explain briefly.


Solution


```{r }
summary(air.1)
```

     

The first component only explains 33\% of the variability, not very
much, but the first *three* components together explain 70\%,
which is much more satisfactory. So I would go with 3 components.

There are two things here: finding an elbow, *and* explaining a
sensible fraction of the variability. You could explain more of the
variability by taking more components, but if you are not careful you
end up explaining seven variables with, um, seven variables.

If you go back and look at the scree plot, you'll see that the first
elbow is really rather high up the mountain, and it's really the
*second* elbow that is the start of the scree.

If this part doesn't persuade you that three components is better than
one, you need to pick a number of components to use for the rest of
the question, and stick to it all the way through.
    


(f) <a name="part:preferred">*</a> How do each of your preferred number of components depend
on the variables that were measured? Explain briefly.


Solution


When this was a hand-in question, there were three marks for it,
which was  a bit of a giveaway!
Off we go:
```{r }
air.1$loadings
```

     

You'll have to decide where to draw the line between "zero" and
"nonzero". It doesn't matter so much where you put the line, so your
answers can differ from mine and still be correct.

We need to pick the loadings that are "nonzero", however we define
that, for example:



* component 1 depends (negatively) on carbon monoxide and nitrogen dioxide.

* component 2 depends (negatively) on solar radiation and ozone
and possibly positively on nitric oxide.

* component 3 depends (positively) on wind and hydrocarbons.


It is a good idea to translate the variable names (which are
abbreviated) back into the long forms.
    


(g) Make a data frame that contains (i) the original data, (ii) a column of row numbers, (iii) the principal component scores. Display some of it.

Solution


All the columns contain numbers, so `cbind` will do
it. (The component scores are seven columns, so
`bind_cols` won't do it unless you are careful.):
```{r }
cbind(air, air.1$scores) %>%  
  mutate(row = row_number()) -> d
head(d)
```

     

This is probably the easiest way, but you see that there is a mixture
of base R and Tidyverse. The result is actually a base R `data.frame`, so displaying it will display *all* of it, hence my use of `head`.
If you want to do it the all-Tidyverse
way
`r tufte::margin_note("There really ought to be a radio station *CTDY: All Tidyverse, All The Time*.")`
then you need to bear in mind that `bind_cols` *only* 
accepts vectors or data frames, not matrices, so a bit of care is needed first:

```{r }
air.1$scores %>%
  as_tibble() %>%
  bind_cols(air) %>%
  mutate(row = row_number()) -> dd
dd
```

 

I think the best way to think about this is to start with what is
farthest from being a data frame or a vector (the matrix of principal
component scores, here), bash that into shape first, and then glue the
rest of the things to it.

Note that we used all Tidyverse stuff here, so the result is a
`tibble`, and displaying it for me displays the first ten rows as
you'd expect. (This may be different in an R Notebook, since I think
there you get the first ten rows anyway.)


(h) Display the row of your new data frame for the observation
with the smallest (most negative) score on component 1. Which row is
this? What makes this observation have the most negative score on
component 1?

Solution


I think the best strategy is to sort by  component 1 score (in the default ascending order), and then display the first row:
```{r }
d %>% arrange(Comp.1) %>% slice(1)
```

     

It's row 8.

We said earlier that component 1 depends negatively on carbon monoxide
and nitrogen dioxide, so that an observation that is *low* on
component 1 should be *high* on these things.
`r tufte::margin_note("You might  have said that component 1 depended on other things as well, in  which case you ought to consider whether observation 8 is, as  appropriate, high or low on these as well.")`

So are these values high or low? That was the reason for having you
make the five-number summary <a href="#part:fivenum">here</a>. For
observation 8, `CO` is 6 and `NO2` is 21; looking back
at the five-number summary, the value of `CO` is above Q3, and
the value of `NO2` is the highest of all. So this is entirely
what we'd expect.


(i) Which observation has the lowest (most negative) value on
component 2? Which variables ought to be high or low for this
observation? Are they? Explain briefly.

Solution


This is a repeat of the ideas we just saw:
```{r }
d %>% arrange(Comp.2) %>% slice(1)
```

     

and for convenience, we'll grab the quantiles again:

```{r }
air %>% map_df(~ quantile(.))
```

 

Day 34. We said that component 2 depends (negatively) on solar
radiation and ozone and possibly positively on nitric oxide. This
means that day 34 ought to be *high* on the first two and low on
the last one (since it's at the low end of component 2). Solar
radiation is, surprisingly, close to the median (75), but ozone, 24,
is very near the highest, and nitric oxide, 1, is one of a large
number of values equal to the lowest. So day 34 is pointing the right
way, even if its variable values are not quite what you'd expect.
This business about figuring out whether values on variables are high
or low is kind of fiddly, since you have to refer back to the
five-number summary to see where the values for a particular
observation come. Another way to approach this is to calculate
*percentile ranks* for everything. Let's go back to our original data frame and replace everything with its percent rank: 

```{r }
air %>% mutate_all(~ percent_rank(.)) -> pct_rank
pct_rank
```

 

Observation 34 is row 34 of this:

```{r }
pct_rank %>% slice(34)
```

 

Very high on ozone, (joint) lowest on nitric oxide, but middling on
solar radiation.
The one we looked at before, observation 8, is this:

```{r }
pct_rank %>% slice(8)
```

 

High on carbon monoxide, the highest on nitrogen dioxide.


(j) Obtain a biplot, with the row numbers labelled, and explain briefly how your conclusions from the previous two parts are consistent with it.

Solution



```{r }
ggbiplot(air.1, labels = d$row)
```

 

Day 8 is way over on the left. The things that point in the direction
of observation 8 (`NO2, CO` and to a lesser extent `NO`
and `HC`) are the things that observation 8 is high on. On the
other hand, observation 8 is around the middle of the arrows for
`wind`, `solar.radiation` and `O3`, so that day
is not especially remarkable for those. 

Observation 34 is
nearest the bottom, so we'd expect it to be high on ozone (yes), high
on solar radiation (no), low on nitric oxide (since that points the
most upward, yes) and also maybe low on wind, since observation 34 is
at the "back end" of that arrow. Wind is 6, which is at the first
quartile, low indeed.

The other thing that you see from the  biplot is that there are four
variables pointing more or less up and to the left, and at right
angles to them, three other variables pointing up-and-right or
down-and-left. You could imagine rotating those arrows so that the
group of 4 point upwards, and the other three point left and
right. This is what factor analysis does, so you might imagine that
this technique might give a clearer picture of which variables belong
in which factor than principal components does. Hence what follows.


(k) Run a factor analysis on the same data, obtaining two factors. Look at the factor loadings. Is it clearer which variables belong to which factor, compared to the principal components analysis? Explain briefly.

Solution


```{r }
air.2 <- factanal(air, 2, scores = "r")
air.2$loadings
```

 

I got the factor scores since I'm going to look at a biplot
shortly. If you aren't, you don't need them.

Factor 1 is rather more clearly carbon monoxide, nitric oxide and
nitrogen dioxide. Factor 2 is mostly ozone, with a bit of solar
radiation and carbon monoxide. I'd say this is clearer than before.

A biplot would tell us whether the variables are better aligned with
the axes now:

```{r }
biplot(air.2$scores, air.2$loadings)
```

 

At least somewhat. Ozone points straight up, since it is the dominant part of factor 2 and not part of factor 1 at all. Carbon monoxide and the two oxides of nitrogen point to the right. 

Extra: 
`wind`, `solar.radiation` and `HC` don't appear
in either of our factors, which also shows up here:

```{r }
air.2$uniquenesses
```

 

Those variables all have *high* uniquenesses.

What with the high uniquenesses, and the fact that two factors explain
only 42\% of the variability, we
really ought to look at 3 factors, the same way that we said we should look at
3 components:

```{r }
air.3 <- factanal(air, 3)
air.3$loadings
```

 

In case you are wondering, `factanal` automatically uses the
correlation matrix, and so takes care of variables measured on
different scales without our having to worry about that.

The rotation has only helped somewhat here. Factor 1 is mainly
`NO2` with some influence of `CO` and `HC`;
factor 2 is mainly ozone (with a bit of solar radiation and carbon monoxide),
and factor 3 is mainly `NO` with a bit of `CO`.

I think I mentioned most of the variables in there, so the uniquenesses
should not be too bad:

```{r }
air.3$uniquenesses
```

 

Well, not great: `wind` and `solar.radiation` still have
high uniquenesses because they are not *strongly* part of any
factors.

If you wanted to, you could obtain the factor scores for the 3-factor
solution, and plot them on a three-dimensional plot using
`rgl`, rotating them to see the structure. A three dimensional
"biplot"
`r tufte::margin_note("A three-dimensional biplot ought to be called a *triplot*.")` 
would also be a cool thing to look at.




