##  Alligator food


 What do alligators most like to eat? 219 alligators were captured
in four Florida lakes. Each alligator's stomach contents were
observed, and the food that the alligator had eaten  was classified
into one of five categories: fish, invertebrates (such as snails or
insects), reptiles (such as turtles), birds, and "other" (such as
amphibians, plants or rocks). The researcher noted for each alligator
what that alligator had most of in its stomach, as well as the gender
of each alligator and whether it was "large" or "small" (greater
or less than 2.3 metres in length). The data can be found in
[link](http://ritsokiguess.site/datafiles/alligator.txt). The
numbers in the data set (apart from the first column) are all
frequencies. (You can ignore that first column "profile".)

Our aim is to predict food type from the other variables.



(a) Read in the data and display the first few lines. Describe
how the data are not "tidy".


Solution


Separated by exactly one space:
```{r alligator-1 }
my_url <- "http://ritsokiguess.site/datafiles/alligator.txt"
gators.orig <- read_delim(my_url, " ")
gators.orig
```

       

The last five columns are all frequencies. Or, one of the variables
(food type) is spread over five columns instead of being contained in
one. Either is good.

My choice of "temporary" name reflects that I'm going to obtain a
"tidy" data frame called `gators` in a moment.

$\blacksquare$

(b) Use `pivot_longer` to arrange the data
suitably for analysis (which will be using
`multinom`). Demonstrate (by looking at the first few rows
of your new data frame) that you now have something tidy.


Solution


I'm creating my "official" data frame here:

```{r alligator-2}
gators.orig %>% 
  pivot_longer(Fish:Other, names_to = "Food.type", values_to = "Frequency") -> gators
gators
```


       
I gave my
column names Capital Letters to make them consistent with the others
(and in an attempt to stop infesting my brain with annoying
variable-name errors when I fit models later).

Looking at the first few lines reveals that I now have a column of
food types and one column of frequencies, both of which are what I
wanted. I can check that I have all the different food types by
finding the distinct ones:

```{r alligator-3 }
gators %>% distinct(Food.type)
```

 

(Think about why `count` would be confusing here.)

Note that `Food.type` is text (`chr`) rather than being a
factor. I'll hold my breath and see what happens when I fit a model
where it is supposed to be a factor.

$\blacksquare$

(c) What is different about this problem, compared to
Question <a href="#q:abortion">here</a>, that would make 
`multinom` the right tool to use?


Solution


Look at the response variable `Food.type` (or whatever
you called it): this has multiple categories, but they are
*not ordered* in any logical way. Thus, in short, a nominal
response. 

$\blacksquare$

(d) Fit a suitable multinomial model predicting food type from
gender, size and lake. Does each row represent one alligator or more
than one? If more than one, account for this in your modelling.


Solution


Each row of the tidy `gators` represents as many
alligators as are in the `Frequency` column. That is, if
you look at female small alligators in Lake George that ate
mainly fish, there are three of those.^[When you have variables that are categories, you might have more than one individual with exactly the same categories; on the other hand, if they had measured *Size* as, say, length in centimetres, it would have been very unlikely to get two alligators of exactly the same size.]
This to remind you to include the `weights` piece,
otherwise `multinom` will assume that you have *one*
observation per line and not as many as the number in
`Frequency`.

*That* is the
reason that `count` earlier would have been confusing:
it would have told you how many *rows* contained each
food type, rather than how many *alligators*, and these
would have been different:
```{r alligator-4 }
gators %>% count(Food.type)
gators %>% count(Food.type, wt = Frequency)
```

         

Each food type appears on 16 rows, but is the favoured diet of very
different numbers of *alligators*. Note the use of `wt=`
to specify a frequency variable.^[Discovered by me two minutes  ago.]

You ought to understand *why* those are different.

All right, back to modelling:

```{r alligator-5 }
library(nnet)
gators.1 <- multinom(Food.type ~ Gender + Size + Lake,
  weights = Frequency, data = gators
)
```

       

This worked, even though `Food.type` was actually text. I guess
it got converted to a factor. The ordering of the levels doesn't
matter here anyway, since this is not an ordinal model.

No need to look at it, since the output is kind of confusing anyway: 
```{r alligator-6 }
summary(gators.1)
```

 

You get one coefficient for each variable (along the top) and for each
response group (down the side), using the first group as a baseline
everywhere. These numbers are hard to interpret; doing predictions is
much easier.

$\blacksquare$

(e) Do a test to see whether `Gender` should stay in
the model. (This will entail fitting another model.) What do you conclude?


Solution


The other model to fit is the one *without* the variable
you're testing:
```{r alligator-7 }
gators.2 <- update(gators.1, . ~ . - Gender)
```

       

I did `update` here to show you that it works, but of course
there's no problem in just writing out the whole model again and
taking out `Gender`, preferably by copying and pasting:

```{r alligator-8 }
gators.2x <- multinom(Food.type ~ Size + Lake,
  weights = Frequency, data = gators
)
```

 

and then you compare the models with and without `Gender` using `anova`:

```{r alligator-9 }
anova(gators.2, gators.1)
```

 

The P-value is not small, so the two models fit equally well, and
therefore we should go with the smaller, simpler one: that is, the one
without `Gender`.

Sometimes `drop1` works here too (and sometimes it doesn't, for
reasons I haven't figured out):

```{r alligator-10, error=T}
drop1(gators.1, test = "Chisq")
```

 

I don't even know what this error message means, never mind what to do
about it.

$\blacksquare$

(f) Predict the probability that an alligator
prefers each food type, given its size, gender (if necessary) and
the lake it was found 
in, using the more appropriate of the two models that you have
fitted so far.  This means (i) 
making a data frame for prediction, and (ii) obtaining and
displaying the predicted probabilities in a way that is easy to read.


Solution

Our best model `gators.2` contains size and lake, so we need to predict for all combinations of those. 

You might think of this first, which is easiest:

```{r}
predictions(gators.2, variables = c("Size", "Lake"), type = "probs")
```

except that this uses the original dataframe, in which the sizes and lakes are repeated (once for each food type, of which there are five):

```{r}
gators
```

You can persist with the above, if you are careful:

```{r}
predictions(gators.2, variables = c("Size", "Lake"), type = "probs") %>% 
  select(group, predicted, Size, Lake) %>% 
  pivot_wider(names_from = group, values_from = predicted)
```

and now you have four numbers in each cell (that are actually all the same), corresponding to the four repeats of size and lake in the original dataset.

A little reading of the help for `pivot_wider` reveals that there is an option `values_fn`. This is for exactly this case, where pivoting wider has resulted in multiple values per cell. The input to `values_fn` is the name of a function that will be used to summarize the four values in each cell. Since those four values are all the same in each case (they are predictions for the same values of size and lake), you could summarize them by any of mean, median, max, min, etc:^[Just don't use a measure of *spread*, which will give you zero.]

```{r}
predictions(gators.2, variables = c("Size", "Lake"), type = "probs") %>% 
  select(group, predicted, Size, Lake) %>% 
  pivot_wider(names_from = group, values_from = predicted, values_fn = mean) -> preds1
preds1
```

So that works, but you might not have thought of all that.

The other way is to get hold of the different names of lakes and sizes, for which this is one way, and then use `datagrid`:

```{r alligator-11 }
Lakes <- gators %>% distinct(Lake) %>% pull(Lake)
Lakes
Sizes <- gators %>% distinct(Size) %>% pull(Size)
Sizes
```

I didn't need to think about `Genders` because that's not in
the better model. See below for what happens if you include it
anyway. 

I have persisted with the Capital Letters, for consistency.

Then:

```{r}
new <- datagrid(model = gators.2, Size = Sizes, Lake = Lakes)
new
```

The `Frequency` column is the mean `Frequency` observed, which was (kinda) part of the model, but won't affect the predictions that we get.

Next:

```{r}
predictions(gators.2, newdata = new, type = "probs")
```

This is, you'll remember, long, with one row for each diet (in `group`), so we need to `select` the columns of interest to us, and then pivot wider:

```{r}
predictions(gators.2, newdata = new, type = "probs") %>% 
  select(rowid, group, predicted, Size, Lake) %>% 
  pivot_wider(names_from = group, values_from = predicted)
```

Either of these two ways is good to get to this point, or some other way that also gets you here.

If you thought that the better model was the one with `Gender`
in it, or you otherwise forgot that you didn't need `Gender`
then you needed to do something like this as well:

```{r alligator-14 }
Genders <- gators %>% distinct(Gender) %>% pull(Gender)
new <- datagrid(model = gators.2, Lake = Lakes, Size = Sizes, Gender = Genders)
new
```


If you predict this in the model *without* `Gender`, you'll get
the following:

```{r alligator-15, size="footnotesize"}
predictions(gators.2, newdata = new, type = "probs") %>% 
  select(rowid, group, predicted, Lake, Size, Gender) %>% 
  pivot_wider(names_from = group, values_from = predicted)
```

 

Here, the predictions for each gender are *exactly the same*,
because not having `Gender` in the model means that we take its
effect to be *exactly zero*.

Alternatively, if you really thought the model with `Gender` was the
better one, then you'd do this:

```{r}
predictions(gators.1, newdata = new, type = "probs") %>% 
  select(rowid, group, predicted, Lake, Size, Gender) %>% 
  pivot_wider(names_from = group, values_from = predicted)
```


and this time there *is* an effect of gender, but it is
smallish, as befits an effect that is not significant.^[There were only 216 alligators total, which is a small sample size for this kind of thing, especially with all those parameters to estimate.]
    
$\blacksquare$

(g) What do you think is the most important way in which the
lakes differ? (Hint: look at where the biggest predicted
probabilities are.)


Solution


Here are the predictions again:
```{r alligator-17 }
preds1
```

       
Following my own hint: the preferred diet in George and Hancock lakes
is fish, but the preferred diet in Oklawaha and Trafford lakes is (at
least sometimes) invertebrates. That is to say, the preferred diet in
those last two lakes is less likely to be invertebrates than it is in
the first two (comparing for alligators of the same size).  This is
true for both large and small alligators, as it should be, since there
is no interaction in the model.
That will do, though you can also note that reptiles are more
commonly found in the last two lakes, and birds sometimes appear
in the diet in Hancock and Trafford but rarely in the other two
lakes. 

Another way to think about this is to hold size constant and
compare lakes (and then check that it applies to the other size
too). In this case, you'd find the biggest predictions among the
first four rows, and then check that the pattern persists in the
second four rows. (It does.)

I think looking at predicted probabilities like this is the
easiest way to see what the model is telling you. 

$\blacksquare$

(h) How would you describe the major difference between the
diets of the small and large alligators?


Solution


Same idea: hold lake constant, and compare small and large, then
check that your conclusion holds for the other lakes as it should.
For example, in George Lake, the large alligators are more
likely to eat fish, and less likely to eat invertebrates,
compared to the small ones. The other food types are not that
much different, though you might also note that birds appear
more in the diets of large alligators than small ones. 
Does that hold in the other lakes? I think so, though there is
less difference for fish in Hancock lake than the others (where
invertebrates are rare for both sizes). Birds don't commonly
appear in any alligator's diets, but where they do, they are
commoner for large alligators than small ones.
    
$\blacksquare$



