

 The "APACHE II" is a scale for assessing patients who
arrive in the intensive care unit (ICU) of a hospital. These are seriously
ill patients who may die despite the ICU's best attempts. APACHE
stands for "Acute Physiology And Chronic Health Evaluation".
`r tufte::margin_note("As with many of these acronyms, you get the idea that the acronym came first and they devised some words to fit it.")`
The scale score is calculated from several physiological measurements
such as body temperature, heart rate and the Glasgow coma scale, as
well as the patient's age. The final result is a score between 0 and
71, with a higher score indicating more severe health issues. Is it
true that a patient with a higher APACHE II score has a higher
probability of dying?

Data from one hospital are in
[link](http://www.utsc.utoronto.ca/~butler/d29/apache.txt). The columns
are: the APACHE II score, the total number of patients who had that
score, and the number of patients with that score who died.



(a)[2] Read in and display the data (however much of it
displays). Why are you convinced that have the right thing?

Solution


Data values separated by one space, so:
```{r }
my_url="http://www.utsc.utoronto.ca/~butler/d29/apache.txt"
icu=read_delim(my_url," ")
icu
```

     

I had to stop and think about what to call the data frame, since one
of the columns is called `apache`.

Anyway, I appear to have an `apache` score between 0 and something, a
number of patients and a number of deaths (that is no bigger than the
number of patients). If you check the original data, the `apache`
scores go up to 41 and are all the values except for a few near the
end, so it makes perfect sense that there would be 38 rows.

Basically, any comment here is good, as long as you make one and it
has something to do with the data.

`apache` scores could be as high as 71, but I imagine a patient would
have to be *very* ill to get a score anywhere near that high.


(b)[2] Does each row of the data frame relate to one patient or
sometimes to more than one? Explain briefly.

Solution


Sometimes to more than one. The number in the `patients`
column says how many patients that line refers to: that is to say
(for example) the row where `apache` equals 6 represents
*all* the patients whose `apache` score was 6, however many
of them there were (14 in this case).
I had to be careful with the wording because the first two rows of
the data frame actually *do* refer to only one patient each
(who survived in both cases), but the later rows do refer to more
than one patient.


(c)[3] Explain why this is the kind of situation where you need a
two-column response, and create this response variable, bearing in
mind that I will (later) want you to estimate the probability of
dying, given the `apache` score.

Solution


This needs a two-column response precisely *because* each row
represents (or could represent) more than one observation.
The two columns are the number of observations referring to the
event of interest (dying), and the number of observations where
that didn't happen (survived). We don't actually have the numbers
of survivals, but we can calculate these by subtracting from the
numbers of patients (since a patient must have either lived or
died): 
```{r }
response = icu %>% mutate(survivals=patients-deaths) %>%
select(deaths, survivals) %>% as.matrix()
response
```

     

noting that the deaths column has to come *first* since that's
what we want the probability of. It has to be a `matrix`, so
`as.matrix` is the final step. You can quickly check that the
two numbers in each row add up to the number of `patients` for
that row.

Or do everything outside of the data
frame: 

```{r }
survivals=with(icu,patients-deaths)
resp=with(icu, cbind(deaths, survivals))
resp
class(resp)
```

 

Or use the dollar sign instead of the `with`s. Any of those is
good. 

I have no objection to your displaying the response matrix.


(d)[3] Fit a logistic regression to estimate the probability of
death from the `apache` score, and display the results.

Solution


```{r }
apache.1=glm(response~apache,family="binomial",data=icu)
summary(apache.1)
```



My naming convention has gotten messed up again. This should really be
called `deaths.1` or something like that, but that would be a
really depressing name.


(e)[2] Is there a significant effect of `apache` score on the
probability of survival? Explain briefly.

Solution


A gimme two points. The P-value for `apache` is $4.94
\times 10^{-13}$, very small, so `apache` score definitely
has an effect on the probability of survival.


(f)[2] Is the effect of a larger `apache` score to increase or to
decrease the probability of death? Explain briefly.

Solution


The slope coefficient for `apache` is 0.1156, positive, and
since we are modelling the probability of death (the first column
of the response matrix), this says that as `apache` goes
up, the probability of death goes up as well.
If you made your response matrix with the columns the wrong way
around, the slope coefficient for `apache` should be
$-0.1156$, but the explanation should come to the same place,
because this says that the probability of survival goes down as
`apache` goes up.


(g)[3] Obtain the predicted probability of death for each of the
`apache` scores that were in the data set. Display these predicted
probabilities next to the `apache` values that they came
from. (You can display all of them.)

Solution


This is the easier version of `predict` since we don't have
to make a new data frame of values to predict from:
```{r }
p=predict(apache.1,type="response")
cbind(icu,p)
```

     

The `type="response"` is needed to make sure the predictions
come out as probabilities. If you omit it, you get log-odds.

The predicted probability (of dying) does indeed go up as
`apache` goes up.


(h)[4] Make a plot of predicted death probability against
`apache` score (joined by lines) with, also on the plot, the
observed proportion of deaths within each `apache` score,
plotted against `apache` score. Does there seem to be good
agreement between observation and prediction?

Solution


This means calculating the observed proportions first, adding the
predicted probabilities, and then making the plot, like this:
```{r }
icu %>% mutate(obs_prop=deaths/patients) %>%
mutate(pred=p) %>%
ggplot(aes(x=apache,y=pred))+geom_line()+
geom_point(aes(y=obs_prop))
```

     

You don't need to make a column in the pipeline with the predictions
in it; you can just use what I called `p` directly in the
`aes`. 

Note that you *do* need to have a new `aes` inside the
`geom_point`, however, because the $y$ of the plot has
changed: it needs to be the observed proportion now.

What you actually have to do in this situation depends on what you
have. In this case, we have the total number of patients at each
`apache` score, but you might have the number of patients
surviving in one column and dying in another, in which case you'd need
to calculate the total first.

I'd say the agreement is pretty good, except for the one patient with
`apache` of 41 but who somehow survived.

That's what I asked for, and is full marks if you got it. However, the
points are not all based on the same number of observations. One way
to show that on your plot is to vary the size of the plotted
point
`r tufte::margin_note("By size is meant the *area* of the circle,  which is what our brains perceive as the size of   two-dimensional, like the area of a slice in a pie chart. On the  plot, the radius of the circle for 20 is less than twice that of the circle for 10, because the area depends on the radius *squared*.")`
according to the number of patients it was based on. This is not hard
to do, since we have exactly that in `patients`:

```{r }
icu %>% mutate(obs_prop=deaths/patients) %>%
mutate(pred=p) %>%
ggplot(aes(x=apache,y=pred))+geom_line()+
geom_point(aes(y=obs_prop,size=patients))
```

     
The points that are far from the prediction are mostly based on a
small number of patients, and the observed proportions for
`apache` scores with a lot of patients are mostly close to the
prediction. Note that the `size`, because it is based on one of
the variables, goes *inside* the `aes`. If you wanted the
points all to be of size 5, say, you'd do it this way:

```{r eval=F}
geom_point(aes(y=obs_prop),size=5)
```

 



