##  Killing aphids


 An experiment was designed to examine how well the insecticide rotenone kills
aphids that feed on the chrysanthemum plant called *Macrosiphoniella sanborni*. 
The explanatory variable is the log concentration (in milligrams per litre) of the
insecticide. At each of the five different concentrations,
approximately 50 insects were exposed. The number of insects exposed
at each concentration, and the number killed, are shown below.


```

Log-Concentration   Insects exposed    Number killed   
0.96                       50              6               
1.33                       48              16              
1.63                       46              24              
2.04                       49              42              
2.32                       50              44              

```




(a) Get these data into R. You can do this by copying the data
into a file and reading that into R (creating a data frame), or you
can enter the data manually into R using `c`, since there are
not many values. In the latter case, you can create a data frame or
not, as you wish. Demonstrate that you have the right data in R.
 
Solution

 
There are a couple of ways. My
current favourite is the `tidyverse`-approved
`tribble` method. A `tribble` is a 
"transposed `tibble`", in which you copy and paste the data,
inserting column headings and commas in the right places. The
columns don't have to line up, since it's the commas that
determine where one value ends and the next one begins:
```{r }
dead_bugs <- tribble(
  ~log_conc, ~exposed, ~killed,
  0.96, 50, 6,
  1.33, 48, 16,
  1.63, 46, 24,
  2.04, 49, 42,
  2.32, 50, 44
)
dead_bugs
```

    

Note that the last data value has no comma after it, but instead has
the closing bracket of `tribble`.  

You can have extra spaces if you wish. They will just be ignored.
If you are clever in R
Studio, you can insert a column of commas all at once (using
"multiple cursors").
I used to do it like this. I make vectors of each column using `c` and then glue the columns together into a data frame:
```{r }
log_conc <- c(0.96, 1.33, 1.63, 2.04, 2.32)
exposed <- c(50, 48, 46, 49, 50)
killed <- c(6, 16, 24, 42, 44)
dead_bugs2 <- tibble(log_conc, exposed, killed)
dead_bugs2
```

     

The values are correct --- I checked them.

Now you see why `tribble` stands for "transposed tibble": if you want to construct a data frame by hand, you have to work with columns and then glue them together, but `tribble` allows you to work "row-wise" with the data as you would lay it out on the page.

The other obvious way to read the data values without typing them is to copy
them into a file and read *that*. The values as laid out are
aligned in columns. They might be separated by tabs, but they are
not. (It's hard to tell without investigating, though a tab is by
default eight spaces and these values look farther apart than that.)
I copied them into a file `exposed.txt` in my current folder
(or use `file.choose`):

```{r }
bugs2 <- read_table("exposed.txt")
bugs2
```

 

This didn't quite work: the last column `Number killed` got
split into two, with the actual number killed landing up in
`Number` and the column `killed` being empty. If you
look at the data file, the data values for `Number killed` are
actually aligned with the word `Number`, which is why it came
out this way. Also, you'll note, the column names have those
"backticks" around them, because they contain illegal characters
like a minus sign and spaces. Perhaps a good way to
pre-empt
`r tufte::margin_note("My daughter learned the word pre-empt because we  like to play a bridge app on my phone; in the game of bridge, you  make a pre-emptive bid when you have no great strength but a lot of  cards of one suit, say seven, and it won't be too bad if that suit  is trumps, no matter what your partner has. If you have a weakish hand with a lot of cards in one suit, your opponents are probably going  to be able to bid and make something, so you pre-emptively bid first  to try and make it difficult for them.")`  all these problems is to
make a copy of the data file with the illegal characters replaced by
underscores, which is my file `exposed2.txt`:

```{r }
bugs2 <- read_table("exposed2.txt")
bugs2
```

 

This is definitely good. We'd have to be careful with Capital Letters
this way,  but it's definitely good.

You may have thought that this was a lot of fuss to make about reading
in data, but the point is that data can come your way in lots of
different forms, and you need to be able to handle whatever you
receive so that you can do some analysis with it.
 

(b) <a name="part:expo">*</a> Looking at the data, would you expect there to be a
significant effect of log-concentration? Explain briefly.
 
Solution


The numbers of insects killed goes up *sharply* as the
concentration increases, while the numbers of insects exposed
don't change much. So I would expect to see a strong, positive
effect of concentration, and I would expect it to be strongly
significant, especially since we have almost 250 insects altogether.
 


(c) We are going to do a logistic regression to predict how
likely an insect is to be killed, as it depends on the
log-concentration. Create a suitable response variable, bearing in
mind (i) that we have lots of insects exposed to each different
concentration, and (ii) what needs to go into each column of the response.

 
Solution


There needs to be a two-column response variable. The first column
needs to be the number of "successes" (insects killed, here) and
the second needs to be the number of "failures" (insects that
survived). We don't actually have the latter, but we know how many
insects were exposed in total to each dose, so we can work it
out. Like this:
```{r }
dead_bugs %>%
  mutate(survived = exposed - killed) %>%
  select(killed, survived) %>%
  as.matrix() -> response
response
```

   

`glm` requires an R `matrix` rather than a data
frame, so the last stage of our pipeline is to create one (using the
same numbers that are in the data frame: all the `as.`
functions do is to change what type of thing it is, without changing
its contents). 

It's also equally good to create the response *outside* of the
data frame and use `cbind` to glue its columns together:

```{r }
resp2 <- with(
  dead_bugs,
  cbind(killed, survived = exposed - killed)
)
resp2
```

 
 


(d) Run a suitable logistic regression, and obtain a summary of
the results.

 
Solution


I think you know how to do this by now:
```{r }
bugs.1 <- glm(response ~ log_conc, family = "binomial", data = dead_bugs)
summary(bugs.1)
```

   
 


(e) Does your analysis support your answer to (<a href="#part:expo">here</a>)?
Explain briefly.

 
Solution


That's a *very* small P-value, $1.1\times 10^{-15}$, on
`log_conc`, so there is no doubt that concentration has an
effect on an insect's chances of being killed. This is exactly what
I guessed in (<a href="#part:expo">here</a>), which I did before looking at the
results, honest!
 


(f) Obtain predicted probabilities of an insect's being killed at
each of the log-concentrations in the data set. (This is easier than
it sometimes is, because here you don't create a new data frame for
`predict`.)

 
Solution


Just this (notice there are only *two* things going into
`predict`): 
```{r }
prob <- predict(bugs.1, type = "response")
cbind(dead_bugs$log_conc, prob)
```

   
or, if you frame everything in terms of the `tidyverse`, turn
the predictions from a `matrix` into a `tibble` first,
and then use `bind_cols` to glue them together:

```{r }
as_tibble(prob) %>% bind_cols(dead_bugs)
```

 
The advantage of showing the whole input data frame is that you can
compare the observed with the predicted. For example, 44 out of 50
insects were killed at log-dose 2.32, which is a proportion of 0.88,
pretty close to the prediction of 0.91.
 

(g) People in this kind of work are often interested in the
"median lethal dose". In this case, that would be the
log-concentration of the insecticide that kills half the
insects. Based on your predictions, roughly what do you think the
median lethal dose is?

 
Solution


The log-concentration of 1.63 is predicted to kill just over half
the insects, so the median lethal dose should be a bit less than
1.63. It should not be as small as 1.33, though, since that
log-concentration only kills less than a third of the insects. So I
would guess somewhere a bit bigger than 1.5. Any guess somewhere in
this ballpark is fine: you really cannot be very accurate.

Extra: this is kind of a strange prediction problem, because we know what
the *response* variable should be, and we want to know what the
explanatory variable's value is. Normally we do predictions the
other way around.
`r tufte::margin_note("This kind of thing is sometimes called an inverse prediction.")`
So the only way to get a more accurate figure is
to try some different log-concentrations, and see which one gets
closest to a probability 0.5 of killing the insect.

Something like this would work:
```{r }
lc.new <- tibble(log_conc = seq(1.5, 1.63, 0.01))
prob <- predict(bugs.1, lc.new, type = "response")
cbind(lc.new, prob)
```

   

The closest one of these to a probability of 0.5 is 0.4971, which goes
with a log-concentration of 1.57: indeed, a bit bigger than 1.5 and a
bit less than 1.63. The `seq` in the construction of the new
data frame is "fill sequence": go from 1.5 to 1.63 in steps of
0.01. The rest of it is the same as before.

Now, of course this is only our "best guess", like a single-number
prediction in regression. There is uncertainty attached to it (because
the actual logistic regression might be different from the one we
estimated), so we ought to provide a confidence interval for it. But
I'm riding the bus as I type this, so I can't look it up right now.

Later: there's a function called `dose.p` 
in `MASS` that appears to do this:

```{r }
library(MASS)
lethal <- dose.p(bugs.1)
lethal
```

 

We have a sensible point estimate (the same 1.57 that we got by hand),
and we have a standard error, so we can make a confidence interval by
going up and down twice it (or 1.96 times it) from the estimate. The
structure of the result is a bit arcane, though:

```{r }
str(lethal)
```

 

It is what R calls a "vector with attributes". To get at the pieces and calculate the interval, we have to do something like this:

```{r }
(lethal_est <- as.numeric(lethal))
(lethal_SE <- as.vector(attr(lethal, "SE")))
```

 

and then make the interval:

```{r }
lethal_est + c(-2, 2) * lethal_SE
```

 

1.47 to 1.68.

I got this idea from page 4.14 of
[link](http://www.chrisbilder.com/stat875old/schedule_new/chapter4.doc). I
think I got a little further than he did. An idea that works more
generally is to get several intervals all at once, say for the
"quartile lethal doses" as well:

```{r }
lethal <- dose.p(bugs.1, p = c(0.25, 0.5, 0.75))
lethal
```

 

This looks like a data frame or matrix, but is actually a 
"named vector", so `enframe` will get at least some of this and turn
it into a genuine data frame:

```{r }
enframe(lethal)
```

 

That doesn't get the SEs, so we'll make a new column by grabbing the "attribute" as above:

```{r }
enframe(lethal) %>% mutate(SE = attr(lethal, "SE"))
```

 

and now we make the intervals by making new columns containing the lower and upper limits:

```{r }
enframe(lethal) %>%
  mutate(SE = attr(lethal, "SE")) %>%
  mutate(LCL = value - 2 * SE, UCL = value + 2 * SE)
```

 

Now we have intervals for the median lethal dose, as well as for the doses that kill a quarter and three quarters of the aphids.



 



