##  Understanding athletes' height, weight, sport and gender


 On a previous assignment, we used MANOVA on the athletes
data to demonstrate that there was a significant relationship between
the combination of the athletes' height and weight, with the sport they
play and the athlete's gender. The problem with MANOVA is that it
doesn't give any information about the *kind* of relationship. To
understand that, we need to do discriminant analysis, which is the
purpose of this question.

The data can be found at
[link](http://www.utsc.utoronto.ca/~butler/c32/ais.txt). 



(a) Once again, read in and display (some of) the data, bearing
in mind that the data values are separated by *tabs*. (This
ought to be a free two marks.)

Solution


Nothing new here:

```{r }
my_url <- "http://www.utsc.utoronto.ca/~butler/c32/ais.txt"
athletes <- read_tsv(my_url)
athletes
```

 


(b) Use `unite` to make a new column in your data frame
which contains the sport-gender *combination*. Display it. (You
might like to display only a few columns so that it is clear that
you did the right thing.) Hint: you've seen `unite` in the
peanuts example in class.

Solution


The columns to combine are called `Sport` and `Sex`,
with Capital Letters. The syntax for `unite` is that you
give the name of the new combo column first, and then the names of
the columns you want to combine, either by listing them or by
using a select-helper. They will be separated by an underscore by
default, which is usually easiest to handle.
`r tufte::margin_note("The opposite      of *unite* is *separate*, which splits a combined      column like my *combo* into separate columns; it too uses    underscore as the default separator.")` 
In `unite`, you can
group the columns to "unite" with `c()`, as in class, or
not, as here. Either way is good.
`r tufte::margin_note("You used to have to group    them, but you don't any more. Hence my old code has them grouped,    but my new code does not.")`
We'll be using height and weight in the
analysis to come, so I decided to display just those:
```{r }
athletes %>%
  unite(combo, Sport, Sex) -> athletesc
athletesc %>% select(combo, Ht, Wt)
```

     

I gave the data frame a new name, since I might want to come back to
the original later. Also, displaying only those columns gives more
width for the display of my `combo`, so that I can be sure I
got it right.

Extra: there is another column, `SSF`, that begins with S, so the
select-helper thing is not so obviously helpful here. But the two
columns we want start with S followed by either e or p, so we could do this:

```{r }
athletes %>%
  unite(combo, matches("^S(e|p)")) %>%
  select(combo, Ht, Wt)
```

 

The `matches` takes a so-called regular expression. This one
says ``starting at the beginning of the column name, find an uppercase
S followed by either a lowercase e or a lowercase p''. This picks out
the columns and only the columns we want. In the opposite order,
though (either order is fine).

I have a feeling we can also take advantage of the fact that the two
columns we want to `unite` are the only two text ones:

```{r error=T}
athletes %>%
  unite(combo, select_if(is.character)) %>%
  select(combo, Ht, Wt)
```

 

Not this, but there must be something like this.

(later)

I posted a question on the R Studio community site:
[link](https://community.rstudio.com/t/combining-unite-and-select-if/6033/2)
and got an answer! This is how it goes:

```{r }
athletes %>%
  unite(combo, names(select_if(., is.character))) %>%
  select(combo, Ht, Wt)
```

 

The key was the `names`, and the extra dot 
("output from previous step") in `select_if`.


(c) Run a discriminant analysis "predicting" sport-gender
combo from height and weight. Display the results. (No comment
needed yet.)

Solution


That would be this. I'm having my familiar trouble with names:
```{r }
combo.1 <- lda(combo ~ Ht + Wt, data = athletesc)
```

     

If you used a new name for the data frame with the sport-gender
combinations in it, use that new name here.

The output:

```{r }
combo.1
```

 

I comment here that there are two linear discriminants because there
are two variables (height and weight) and actually 17 groups (not
quite $2\times 10$ because some sports are played by athletes of only
one gender). The smaller of 2 and $17-1$ is 2. (I often ask about
this, but am choosing not to here.)


(d) What kind of height and weight would make an athlete have a
large (positive) score on `LD1`? Explain briefly.

Solution


The Coefficients of Linear Discriminants for `LD1` are both
positive, so an athlete with a large positive score on
`LD1` has a large height and weight: that is to say, they
are tall and heavy.


(e) Make a guess at the sport-gender combination that has the
*highest* score on LD1. Why did you choose the combination you did?

Solution


I could have made you guess the *smallest* score on LD1, but
that would have been too easy (female gymnasts). 
For this one, you want a sport-gender combination that is typically
tall and heavy, and you can look in the table of Group Means to
help you find a candidate group.
I think the two best guesses are male basketball players (tallest
and nearly the heaviest) and male field athletes (heaviest and
among the group of athletes that are second-tallest behind the
male basketball players). I don't so much mind what you guess, as
long as you make a sensible call about a group that is reasonably
tall and reasonably heavy (or, I suppose, that matches with what
you said in the previous part, whatever that was).


(f) <a name="part:ld2">*</a> What combination of height and weight would make an athlete have a
*small* (that is, very negative) score on LD2? Explain briefly.

Solution


The italics in the question are something to do with questions
that have a link to them in Bookdown. I don't know how to fix
that.
Going back to the Coefficients of Linear Discriminants, the
coefficient for Height is negative, and the one for Weight is
positive. What will make an athlete come out small (very
negative) on this is if they have a *large* height and a
*small* weight. 
To clarify your thinking on this, think of
the heights and weights as being standardized, so that a big one
will be positive and a small one will be negative. To make
`LD2` very negative, you want a "plus" height to multiply
the minus sign, and a "minus" weight multiplying the plus sign.
Extra: what is happening here is that `LD1` gives the most
important way in which the groups differ, and `LD2` the
next-most important. There is generally a positive correlation
between height and weight (taller athletes are generally heavier),
so the most important "dimension" is the big-small one with tall
heavy athletes at one end and short light athletes at the other.
The `Proportion of trace` in the output says that
`LD1` is definitely more important, in terms of separating
the groups, than `LD2` is, but the latter still has
*some* value.


(g) Obtain predictions for the discriminant analysis, and use
these to make a plot of `LD1` score against `LD2`
score, with the individual athletes distinguished by what sport they play
and gender they are. (You can use colour to distinguish them, or you
can use shapes. If you want to go the latter way, there are clues in
my solutions to the MANOVA question about these athletes.)

Solution


The prediction part is only one step:
```{r }
p <- predict(combo.1)
```

     

One point for this.

This, in case you are wondering, is obtaining predicted group
membership and LD scores for the original data, that is, for our 202
athletes. 

I prefer (no obligation) to take a look at what I have. My `p`
is actually a `list`:
```{r }
class(p)
glimpse(p)
```

 

Our standard procedure is to `cbind` the predictions together with the original data (including the combo), and get a huge data frame (in this case):

```{r }
d <- cbind(athletesc, p)
head(d)
```

 
And so, to the graph:

```{r error=T}
ggplot(d, aes(x = x.LD1, y = x.LD2, colour = combo)) + geom_point()
```

 
If you can distinguish seventeen different colours, your eyes are
better than mine! You might prefer to use seventeen different shapes,
although I wonder how much better that will be:

```{r error=T}
ggplot(d, aes(x = x.LD1, y = x.LD2, shape = combo)) + geom_point() +
  scale_shape_manual(values = 1:17)
```

     

You have to do something special to get as many as seventeen
shapes. This idea came from the MANOVA question in the last
assignment. 

Or even this:

```{r error=T}
ggplot(d, aes(x = x.LD1, y = x.LD2, shape = combo, colour = combo)) + geom_point() +
  scale_shape_manual(values = 1:17)
```

 
Perhaps having colours *and* shapes makes the combos easier to
distinguish. We're beginning to stray onto the boundary between
statistics and aesthetics here!

Extra: earlier, I asked you to guess which group(s) of athletes had a
high (positive) score on LD1. These are the ones on the right side of
this plot: male basketball players bottom right and male field
athletes top right. Was that what you guessed? What about the other
guesses you might have made?


(h) Look on your graph for the four athletes with the smallest
(most negative) scores on `LD2`. What do they have in common?
Does this make sense, given your answer to part (<a href="#part:ld2">here</a>)?
Explain briefly.

Solution


These are the four athletes at the bottom of the plot. If you can
distinguish the colours, two of these are red and two of them are
orange, so they are all basketball players (two male and two
female). If you plotted the shapes, and you used the same shapes I
did, two of them are circles and the other two are upward-facing
triangles, leading you to the same conclusion. (You could also
denote each combo by a letter and plot with those letters, as per
the solutions to the last assignment.)
Back in part (<a href="#part:ld2">here</a>), I said that what would make an
athlete come out very negative on `LD2` is if they were
*tall* and *not heavy*. This is the stereotypical
description of a basketball player, so it makes perfect sense to
me. 
Extra: some basketball players are tall and *heavier*; these
are the ones on the right of the plot, with a larger `LD1`
score, to reflect that they are both tall and heavy, but with an
`LD2` score closer to zero, reflecting that, given how tall
they are, their weight is about what you'd expect. LD2 is really
saying something like "weight relative to height", with someone
at the top of the picture being unusually heavy and someone at the
bottom unusually light.


(i) Obtain a (very large) square table, or a (very long) table
with frequencies, of actual and predicted sport-gender
combinations. You will probably have to make the square table very
small to fit it on the page. For that, displaying the columns in two
or more sets is OK (for example, six columns and all the rows, six
more columns and all the rows, then the last five columns for all
the rows).  Are there any sport-gender combinations that
seem relatively easy to classify correctly?  Explain briefly.

Solution


Let's see what happens:
```{r error=T}
tab <- with(d, table(combo, class))
tab
```

     

That's kind of long.

For combos that are easy to classify, you're looking for a largish
number on the diagonal of the table (classified correctly), bearing in
mind that you only see about four columns of the table at once, and
(much) smaller numbers in the rest of the row and column. I don't mind
which ones you pick out, but see if you can find a few:



* Male basketball players (9 out of 12 classified correctly)

* Male field athletes (7 out of 10 classified correctly)

* Female netball players (13 out of about 23)

* Female rowers (10 out of about 22)

Or you can turn it into a tibble:

```{r }
tab %>% as_tibble()
```

 
This makes the `tidyverse` output, with frequencies. You
probably want to omit the zero ones:

```{r size="tiny"}
tab %>% as_tibble() %>% filter(n > 0)
```

 
This is the same output as below. See there for comments.

The other, perhaps easier, way to tackle this one is the
`tidyverse` way, making a "long" table of frequencies. Here is some of it. You'll be able to click to see more:

```{r }
d %>% count(combo, class)
```

 

The zeroes never show up here.
The `combo` column is the truth, and the `class` column
is the prediction. Again, you can see where the big frequencies are; a
lot of the female netball players were gotten right, but there were a
lot of them to begin with.

Extra: let's see if we can work out *proportions* correct. I've
changed my mind from how I originally wrote this. I still use
`count`, but I start with the overall misclassification. Let's
take it in steps:

```{r }
d %>%
  count(combo, class) %>%
  mutate(stat = ifelse(combo == class, "correct", "wrong"))
```

 

That makes a new column `stat` that contains whether the
predicted sport-gender combination was correct or wrong. For an
overall misclassification rate we have to count these, but *not*
simply counting the number of rows; rather, we need to total up the
things in the `n` column:

```{r error=T}
d %>%
  count(combo, class) %>%
  mutate(stat = ifelse(combo == class, "correct", "wrong")) %>%
  count(stat, wt = n)
```

 

This tells us how many predictions overall were right and how many
wrong. 

To make those into proportions, another `mutate`, dividing by
the total of `n`:

```{r error=T}
d %>%
  count(combo, class) %>%
  mutate(stat = ifelse(combo == class, "correct", "wrong")) %>%
  count(stat, wt = n) %>%
  mutate(proportion = n / sum(n))
```

 

65\% of the sport-gender combinations were misclassified. This is
awful, but is a lot better than guessing (we'd then get about 5\% of
them right and about 95\% wrong).

There's a subtlety here that will make sense when we do the
corresponding calculation by sport-gender combination. To do
*that*, we put a `group_by(combo)` either before or after
we define `stat` (it doesn't matter which way):

```{r error=T}
d %>%
  count(combo, class) %>%
  group_by(combo) %>%
  mutate(stat = ifelse(combo == class, "correct", "wrong")) %>%
  count(stat, wt = n) %>%
  mutate(proportion = n / sum(n))
```

 

That last `sum(n)`: what is it summing over? The answer is
"within `combo`", since that is the `group_by`. You
see that the two `proportion` values within, say,
`BBall_female`, add up to 1.

We don't actually see all the answers, because there are too many of
them. Let's try to get the proportion correct and wrong in their own
columns. This almost works:

```{r error=T}
d %>%
  count(combo, class) %>%
  group_by(combo) %>%
  mutate(stat = ifelse(combo == class, "correct", "wrong")) %>%
  count(stat, wt = n) %>%
  mutate(proportion = n / sum(n)) %>%
  pivot_wider(names_from=stat, values_from=proportion)
```

 

This doesn't work because everything outside of the `pivot_wider` is
tested for uniqueness; if it's unique, it gets its own row. Thus,
`BBall_male` and 3 is different from `BBall_male` and
9. But we only want one row of `BBall_male`. I think the
easiest way around this is to get rid of `n`, since it has
served its purpose:

```{r error=T}
d %>%
  count(combo, class) %>%
  group_by(combo) %>%
  mutate(stat = ifelse(combo == class, "correct", "wrong")) %>%
  count(stat, wt = n) %>%
  mutate(proportion = n / sum(n)) %>%
  select(-n) %>%
  pivot_wider(names_from=stat, values_from=proportion, values_fill = list(proportion=0))
```

One extra thing: some of the `proportion` values were missing, because there weren't any misclassified (or maybe correctly-classified!) athletes. The `values_fill` sets any missings in `proportion` to zero.
 

While we're about it, let's arrange in
order of misclassification probability:

```{r error=T}
d %>%
  count(combo, class) %>%
  group_by(combo) %>%
  mutate(stat = ifelse(combo == class, "correct", "wrong")) %>%
  count(stat, wt = n) %>%
  mutate(proportion = n / sum(n)) %>%
  select(-n) %>%
  pivot_wider(names_from=stat, values_from=proportion, values_fill = list(proportion=0)) %>% 
  replace_na(list(correct = 0, wrong = 0)) %>%
  arrange(wrong)
```

 

The most distinctive athletes were the female gymnasts (tiny!),
followed by the male basketball players (tall) and the female field
athletes (heavy). These were easiest to predict from their height and
weight. The ones at the bottom of the list were very confusible since
the discriminant analysis guessed them all wrong!
So what were the most common *misclassifications*? Let's go back
to this:
```{r error=T}
head(d)
d %>%
  count(combo, class) %>%
  mutate(stat = ifelse(combo == class, "correct", "wrong"))
```

 
We want to express those `n` values as proportions out of their
actual sport-gender combo, so we group by `combo` before
defining the proportions:

```{r error=T}
d %>%
  count(combo, class) %>%
  mutate(stat = ifelse(combo == class, "correct", "wrong")) %>%
  group_by(combo) %>%
  mutate(proportion = n / sum(n))
```

 

Only pick out the ones that were gotten wrong, and arrange the remaining
proportions in descending order:

```{r error=T}
d %>%
  count(combo, class) %>%
  mutate(stat = ifelse(combo == class, "correct", "wrong")) %>%
  group_by(combo) %>%
  mutate(proportion = n / sum(n)) %>%
  filter(stat == "wrong") %>%
  arrange(desc(proportion))
```

 

The embarrassment champion is the three male tennis players that were
taken to be --- female rowers! Most of the other mistakes are more
forgivable: the male rowers being taken for male water polo players,
for example. 




