##  How do height and weight depend on sport played by elite athletes?


 This question uses the data on Australian elite athletes,
which can be found at
[link](http://www.utsc.utoronto.ca/~butler/c32/ais.txt). 202 elite
athletes had various physical and physiological measurements taken,
and for each athlete, their gender (in the column `Sex`) and
the sport they play (`Sport`) was also noted. The data values
are separated by *tabs*, meaning that `read_tsv` will
read them in.



(a) Read in the data and verify that you have 202 rows and 13
columns. 

Solution


As in the hint:
```{r }
my_url <- "http://www.utsc.utoronto.ca/~butler/c32/ais.txt"
athletes <- read_tsv(my_url)
athletes
```

     

202 rows and 13 columns indeed.


(b) We are going to see whether the height-weight combination
of an athlete depends significantly on which sport they play and which gender
they are. Explain (very) briefly why a multivariate analysis of
variance will be necessary to do this.

Solution


We now have *two* response variables, height and weight,
rather than just one. In order to do a test with more than one
response variable, we need to use multivariate ANOVA rather than
regular ANOVA. (The first sentence is enough.)

Extra: the explanatory variables, sport and gender, are both
categorical here, which makes a MANOVA work. If either of them had
been quantitative, we would have needed to do a multivariate
regression, which is beyond the scope of what we do in this course.


(c) Create a response variable for a MANOVA. (You don't need to
show what you got, since it is rather big.) The columns you want are
called `Ht` and `Wt`.

Solution


Use `cbind` to glue the two columns together:
```{r }
response <- with(athletes, cbind(Ht, Wt))
```

     

Either way around, height or weight first, is good. Also good is the
dollar sign, which you'll need to do twice (and thus the `with`
way is more elegant):

```{r }
response_a <- cbind(athletes$Ht, athletes$Wt)
```

 


(d) Run a multivariate analysis of variance to see whether the
height-weight combination depends significantly on gender, sport or the
combination of both. Display the results. Use the small-m
`manova` way to do this one.

Solution


The straightforward way is small-m `manova`. I don't know
what I should be calling my fitted model object, so I'm making
something up. Don't forget to use the right names for the
variables, and to include an interaction:
```{r }
htwt.1 <- manova(response ~ Sex * Sport, data = athletes)
summary(htwt.1)
```

     

The other way is apparently to use large-M `Manova` from package
`car`. This goes as a two-stage process: fit an `lm`
first with the same model formula as above, and then pass *that*
into `Manova`. This way, you just display the final result,
rather than passing it into `summary`:

```{r error=T}
htwt.2 <- lm(response ~ Sex * Sport, data = athletes)
htwt.3 <- Manova(htwt.2)
```

 

Except that this doesn't work, for reasons that I don't
understand. `htwt.2` is fine; it's the `Manova` that is
not working.


(e) Briefly justify removing the interaction term from your
previous model, and fit a model without it. You can use either
`manova` or `Manova` for this; they should both
work. (You only need to use one of them.)  Display your results.

Solution


The interaction should come out, since it is not
significant. (This is what I meant by "briefly.")
To do that, in each case, replace the `*` with a `+`, or use
`update`. Most of those work:
```{r }
htwt.4 <- manova(response ~ Sex + Sport, data = athletes)
summary(htwt.4)
```

   

or

```{r error=T}
htwt.5 <- update(htwt.1, . ~ . - Sex:Sport)
```

 

It seems that `manova` things don't like `update`. If
that happened to you, try it another way.

or, with `Manova`, the same two ways:

```{r }
htwt.6 <- lm(response ~ Sex + Sport, data = athletes)
htwt.7 <- Manova(htwt.6)
htwt.7
```

 

or

```{r }
htwt.8 <- update(htwt.2, . ~ . - Sex:Sport)
htwt.9 <- Manova(htwt.8)
htwt.9
```

 

These both work, and give the same answer. I think this is because the
thing being `update`d is a regular `lm`; the
`Manova` bit comes after, and is calculated from the
`lm` fit.

Any of these is good, and gets you the same answer.

Extra: it you run `summary` on the output from `Manova`,
you get more detail:

```{r }
summary(htwt.9)
```

 

There are actually *four* competing tests that generalize the
one-response-variable $F$ statistic in different ways. The one we've
been using is called Pillai's Trace, but if you go looking in a text
on multivariate analysis, you'll also see the others, especially
Wilks's $\Lambda$ (that's a capital lambda). They will generally agree
in terms of conclusion, so you can look at them all if you like, but
it doesn't hurt much to look at only one.

I am in danger of running out of numbers for my fitted models (and am
definitely in danger of forgetting which one is which)!


(f) See if you can make a graph that shows what's going on
here. Bear in mind that you have two quantitative variables and
*two* categorical ones, so you will need to find a way to
display everything. Hint: what would you do with two quantitative
variables and *one* categorical one? Can you find a way to
generalize that idea by displaying the other categorical variable
differently? I think we have seen a couple of ideas that will work,
somewhere in the lecture notes (but in a different context). Or you
can search for ideas, of course.
For full marks, obtain a plot which does not give you any
warnings. (The grader will know which of the possible plots give
warnings, and will deduct extra marks if you assert that a plot
gives no warnings when in fact it does. That is to say, if you get
any warnings, *you need to include those in what you hand in*).

Solution


This will probably involve a bit of digging.
The starting point would be a scatterplot (the two quantitative
variables) with the points distinguished by colour (one of the
categorical variables). What other options do we have? 
If you look eg. at the navigation on the left of
[link](http://sape.inf.usi.ch/quick-reference/ggplot2/size), you'll
see that there are two likely possibilities: `shape` and
`size`. `shape` uses a different plotting symbol for
each value of a categorical variable; `size` draws the
plotting symbol bigger or smaller according to the value of a
categorical variable. 
The remaining question is the one of which aesthetic to "map" to
which variable? To guide you in this, note that we have ten
different sports and two different genders. In my opinion, colours
are the easiest to distinguish a lot of, though even then 10 is
pushing it, then plotting symbols. Symbol sizes are hard to
distinguish a lot of, so it's best to limit this (if you use it at
all) to gender.
With that in mind, I would go for this one:
```{r }
ggplot(athletes, aes(x = Ht, y = Wt, colour = Sport, shape = Sex)) + geom_point()
```

     

If you investigate, you'll find that `ggplot` can handle about
six different shapes (and we have ten sports), and it doesn't care for
using `size` to distinguish values of a categorical
variable. It will try to draw the graph, but will give you a warning,
such as these:

```{r }
ggplot(athletes, aes(x = Ht, y = Wt, colour = Sport, size = Sex)) + geom_point()
```

     

```{r }
ggplot(athletes, aes(x = Ht, y = Wt, shape = Sport, size = Sex)) + geom_point()
```

     

```{r }
ggplot(athletes, aes(x = Ht, y = Wt, shape = Sport, colour = Sex)) + geom_point()
```

     

Whatever plot you come up with, you can put height and weight on the
other axes, since neither one of them is uniquely a response variable.

Any of these graphs with warnings do the job, kind of, so they get 2 points.

One of the many well-thought-out things about `ggplot2` is that
making this graph was actually not difficult once you had figured out
that you also had `shape` and `size` at your
disposal. (In addition, `ggplot2` will help you make some
aesthetic decisions about which variable is distinguished by what kind
of thing. This will help you work towards a plot that doesn't have any
warnings. These warnings are pretty clear about what not to do, and by
implication what you need to do instead. Colour, for example, is
always safe warning-wise.)

The model for this kind of graph that we saw is the plot of
probability (of preferring different brands of a product) as it depended
on age (two quantitative variables) for each brand and gender (two
categorical variables). I had three brands and two genders there, so I
used colour for brand and shape for gender, but it would have worked
about as well if I'd used shape for brand and colour for gender.

The warning message for `shape` contains the words ``specify
shapes manually if you must have them''. I was wondering how you can
avoid that. Here's how:

```{r }
ggplot(athletes, aes(x = Ht, y = Wt, shape = Sport, colour = Sex)) + geom_point() +
  scale_shape_manual(values = 1:10)
```

     

I agree with `ggplot2` that this many shapes are hard to tell
apart,
`r tufte::margin_note("Having said that, the shapes are less anbiguous than  the colours, because if you're willing to study the legend, you can  work out exactly which sport a shape belongs to, whereas the colours  might be hard to tell apart at all.")` but if you can figure this out,
you achieve the goal of producing a plot with no warnings, so you get
full marks. (We need 10 shapes because there are 10 different sports,
so we have to specify 10 different values in `values=`: any 10
different values will do. A list of the possible shapes is in
[link](https://stackoverflow.com/questions/16813278/cycling-through-point-shapes-when-more-than-6-factor-levels),
at the bottom. You can even use regular letters, but you have to refer
to them by numeric code:

```{r }
ggplot(athletes, aes(x = Ht, y = Wt, shape = Sport, colour = Sex)) + geom_point() +
  scale_shape_manual(values = c(66, 70, 71, 78, 82, 83, 52, 84, 3, 87))
```

     

I had to find something for `TSprnt`, since all the letters
appeared to be taken.

On consideration, I like this last graph the best, because the letters
are more or less reminders of which sport each point is (they are
"mnemonic", so you don't have to look them up all the time, as with
colours or shapes: you can eventually remember which is which). 

Whatever you come up with, expect: 3 marks if you get a plot with no
warnings, 2 marks if you get a plot with warnings, 1 mark if you get a
plot that would produce warnings but you don't indicate that there are
warnings, either by saying so or by including the warnings in what you
hand in.

I didn't ask you to interpret your plot (just getting it was enough
work), but there seems to be a common theme that, for sports played by
both genders, that the athletes are in about the same place relative
to the other ones in their gender (eg., the basketball players are
tall and heavy, the field athletes are really heavy but not very
tall), but male athletes are both taller and heavier than female
athletes playing the same sport. This consistency of pattern was
probably why the interaction was nowhere near significant: there are
additive separate effects of sport and of gender on the height-weight
combination. 



