## Ferritin and normality



 In the lecture notes, we looked at some
data on different athletes from the Australian Institute of
Sport. This data set can be found at
[http://ritsokiguess.site/datafiles/ais.txt](http://ritsokiguess.site/datafiles/ais.txt). Recall that the
values are separated by *tabs*.  In this question, we will assess
one of the variables in the data set for normality.



(a) Read the data set into R. 
  

Solution


`read_tsv` is the right thing:
```{r }
my_url="http://ritsokiguess.site/datafiles/ais.txt"
athletes=read_tsv(my_url)
athletes
```       

I listed the data to check that I had it right, but I didn't ask you
to. (If you didn't have it right, that will show up soon enough.)
  
$\blacksquare$

(b) One of the variables, `Ferr`, is a measurement of
Ferritin for each athlete. Obtain a normal quantile plot of the
Ferritin values, for all the athletes together. What do you
conclude about the shape of the distribution? Explain briefly.
  

Solution


As you would expect:

```{r }
ggplot(athletes, aes(sample=Ferr))+
stat_qq()+stat_qq_line()
``` 

This is almost a classic right skew: the values are too bunched up at
the bottom and too spread out at the top. The curved shape should be
making you think "skewed" and then you can work out which way it's
skewed. 
  
$\blacksquare$

(c) It is possible that the shape you found in the previous
part is because the athletes from all the different sports were
mixed together. Use `ggplot` to obtain one normal quantile
plot for each sport, collected together on one plot. 
  

Solution


Your previous plot had all the sports mixed together. To that
you add something that will put each sport in its own facet:

```{r }
ggplot(athletes,aes(sample=Ferr))+stat_qq()+stat_qq_line()+
facet_wrap(~Sport)
``` 
  
$\blacksquare$

(d) Looking at the plots in the previous part, would you say
that the Ferritin values for each sport individually have a more
normal shape than they do for all the sports together? Explain
briefly. 
  

Solution


There are a couple of ways you can go, and as ever I'm looking
mostly for consistency of argument. The two major directions you
can go are (i) most of these plots are still curved the same way
as the previous one, and (ii) they are mostly straighter than
they were before.
Possible lines of argument include that pretty much all of these
plots are right-skewed still, with the same upward-opening
curve. Pretty much the only one that doesn't is Gymnastics, for
which there are only four observations, so you can't really
tell. So, by this argument, Ferritin just *does* have a
right-skewed distribution, and breaking things out by sport
doesn't make much difference to that.
Or, you could go another way and say that the plot of all the
data together was *very* curved, and these plots are much
less curved, that is to say, much less skewed. Some of them,
such as basketball and netball, are almost straight, and they
are almost normally distributed. Some of the distributions, such
as track sprinting (`TSprnt`), are definitely still
right-skewed, but not as seriously so as before.
Decide what you think and then discuss how you see it.
  
$\blacksquare$



