

 Does attendance at baseball games depend on how warm it is,
and if so, how? The data in
[http://ritsokiguess.site/datafiles/baseball.txt](http://ritsokiguess.site/datafiles/baseball.txt) were
collected by a minor league baseball team. The first column is the
temperature at 6:00pm on the night of a game, and the second column is
the attendance at  that night's game.



(a) Read in the data and verify that you have seven
temperatures and attendances.

Solution


```{r }
my_url="http://ritsokiguess.site/datafiles/baseball.txt"
night.games=read.table(my_url,header=T)
night.games
```     

Seven, by counting them. Or,

```{r }
dim(night.games)
``` 

Seven rows and two columns.

$\blacksquare$

(b)\label{part:scplot} Make a scatterplot for predicting
attendance from temperature. Is there an apparent relationship? Is
it approximately linear?

Solution


```{r penzance}
attach(night.games)
plot(attendance~temperature)
```     
This is a very weak relationship. If there is anything happening at
all, the attendance appears to go up with temperature and then down
again. So whatever trend there is, it does not appear to be linear.

 

$\blacksquare$

(c) Fit a linear relationship for predicting attendance from
temperature. Take a look at the summary. Is there something in here
that supports the conclusion you drew in (\ref{part:scplot})?


Solution


```{r }
night.games.1=lm(attendance~temperature)
summary(night.games.1)
``` 

The $F$ test for the regression (or the $t$-test for the slope of
`temperature`) is not significant, meaning that there is no
evidence of a linear relationship beyond chance. That supports the
"no relationship" conclusion from (\ref{part:scplot}), but it also
kind of supports the "up and down" trend idea, because that is not a
linear trend. Either way, the message is "a linear trend does not describe the data". 

xxx Some thoughts for me:

- fit quadratic term?

- put residual plot in with multiple regression?

 


