

 Can students throw a baseball farther than a softball? A
statistics class, containing 24 students, went out to a football field
to try to answer this question. Each student warmed up and then threw
each type of ball as far as they could. The order of ball types was
randomized: some students threw the baseball first, and some threw the
softball first. (A softball is bigger than a baseball, so we might
expect that a softball would be harder to throw a long way than a
baseball.) The data are in
[http://www.utsc.utoronto.ca/~butler/c32/throw.txt](http://www.utsc.utoronto.ca/~butler/c32/throw.txt) in three
columns: the first is a number identifying the student, the second is
the distance thrown with the baseball (in yards) and the third is the
distance thrown with the softball (also in yards).



(a) Read the data into R. You'll need to supply some names
to the columns.


Solution


This kind of thing:
```{r }
myurl="http://www.utsc.utoronto.ca/~butler/c32/throw.txt"
throws=read_delim(myurl," ",col_names=c("student","baseball","softball"))
throws
```

This is one of those times where we have to tell R what names to give
the columns. Or you can put `col_names=F` and leave the
columns called `X1, X2, X3` or whatever they end up as.



(b) Calculate a column of differences, baseball minus softball,
*in* the data frame.


Solution


Add it to the data frame using `mutate`. Use the
right-arrow assignment to create what I called
`throws2` below, or put something like
`throws2=` on the beginning of the line. Your choice.

```{r }
throws %>% mutate(diff=baseball-softball) ->
throws2
``` 



(c) Carry out a sign test in R, testing the null hypothesis
that the median difference is zero, against the alternative that
it is greater than zero. Obtain a P-value. Your option whether you use
`smmr` or not.


Solution


I think using `smmr` is way easier, so I'll do that
first. There is even a shortcut in that the null median defaults to
zero, which is exactly what we want here:

```{r }
library(smmr)
sign_test(throws2,diff)
``` 

We want, this time, the upper-tailed one-sided test, since we want to
prove that students can throw a baseball a *longer* distance
than a softball. Thus the P-value we want is 0.000033.

To build it yourself, you know the steps by now.
First step is to count how many differences are greater and less than zero:
```{r }
table(throws2$diff>0)
``` 

or

```{r }
table(throws2$diff<0)
``` 

or, since we have things in a data frame,

```{r }
throws2 %>% count(diff>0)
``` 

or count those less than zero. I'd take any of those.

Note that these are *not all the same*. One of the differences is
in fact exactly zero. The technically right thing to do with the zero
difference is to throw it away (leaving 23 differences with 2 negative
and 21 positive). I would take that, or 2 or 3 negative differences
out of 24 (depending on whether you count "greater than zero" or
"less than zero"). We hope that this won't make a material
difference to the P-value; it'll make some difference, but won't (we
hope) change the conclusion about whether to reject.

Second step is to get a P-value for whichever one of those you got,
from the appropriate binomial distribution. 

The P-value is the probability of getting 21 (or 22) positive
differences out of 24 (or 23) or more, since this is the end of the
distribution we should be at if the alternative hypothesis is correct.
Thus any of these will get you a defensible P-value:

```{r }
sum(dbinom(21:23,23,0.5))
sum(dbinom(22:24,24,0.5))
sum(dbinom(21:24,24,0.5))
sum(dbinom(0:2,23,0.5))
sum(dbinom(0:2,24,0.5))
sum(dbinom(0:3,24,0.5))
``` 

The first and fourth of those are the same as `smmr` (throwing
away the exactly-median value). 

As we hoped, there is no *material* difference here: there is no
doubt with any of these possibilities that we will reject a median
difference of zero in favour of a median difference greater than
zero. 


```{r }

```     


