##  More beer please


 Previously, you did a cluster analysis of
ten brands of beer, as rated by 32 students. This time, we will do a
non-metric multidimensional scaling of those same brands of beer. The
data are in [link](http://ritsokiguess.site/datafiles/beer.txt).



(a) Noting that we want to assess distances between brands of beer,
read in the data and do whatever you need to do to work out
distances between the beers. Show your result.
 
Solution



       
This is really a copy of last time. We need to transpose the data
frame to get the beers in *rows* (`dist` works on
distances between rows), then feed everything but the student IDs
into `dist`:
```{r }
my_url <- "http://ritsokiguess.site/datafiles/beer.txt"
beer <- read_table2(my_url)
beer
d <- beer %>%
  dplyr::select(-student) %>%
  t() %>%
  dist()
```



I did it the funny-looking way. The cluster analysis question offers
an alternative.
 
$\blacksquare$

(b) Obtain a non-metric multidimensional scaling of the
beers. (Comment coming up in a moment.)
 
Solution


```{r }
library(MASS)
beer.1 <- isoMDS(d)
```

     
 
$\blacksquare$

(c) Obtain the `stress` value of the map, and comment on it.
 
Solution


```{r }
beer.1$stress
```

     

The stress is around 10\%, on the boundary between "good" and
"fair". It seems as if the map should be more or less worth
using. (Insert your own hand-waving language here.)
 
$\blacksquare$

(d) Obtain a map of the beers, labelled with the names of the
beers. 
 
Solution


This is slightly different from class, where I plotted the
languages actually *at* their locations. But here, the beer
names are longer, so we should plot the points and label them. I'd
make a data frame first, and probably pipe it into the plot,
thus. Don't forget we have to get the names of the 10 beers, and
not the 32 students! The names of the columns of the data frame
include an identifier column for the students, so skip the first one:

```{r }
beer_names <- beer %>% dplyr::select(-student) %>% names()
beer_names
data.frame(beer.1$points, beer = beer_names) %>%
  ggplot(aes(x = X1, y = X2, label = beer)) +
  geom_point() + geom_text_repel()
```

     
 
$\blacksquare$

(e) Find a pair of beers close together on your map. Are they
similar in terms of student ratings? Explain briefly.
 
Solution


I think Sam Adams and Gordon Biersch, right in the middle of the
map. We can pull them out by name:
```{r }
beer %>% dplyr::select(SamAdams, GordonB)
```

     

These are, with a few exceptions (the most glaring being the 18th
student), within a couple of points of each other. So I would say they
are similar. Another way to show this is to make a scatterplot of
them, and draw on it the line where the ratings are the same. Since
the ratings are whole numbers, they are likely to be duplicated, so I
"jitter" them as I plot them, to prevent overplotting:

```{r adkjalkjdsg}
ggplot(beer, aes(x = SamAdams, y = GordonB)) + geom_jitter() +
  geom_abline(intercept = 0, slope = 1)
```

 

Are they close to the line, or not? Dunno. Maybe I should plot the
ratings of some far-apart beers, and see whether it looks any
different, for example `Becks` and `SierraN`:

```{r adkjalkjdsh}
ggplot(beer, aes(x = Becks, y = SierraN)) + geom_jitter() +
  geom_abline(intercept = 0, slope = 1)
```

 

Not much. I was expecting something a lot more impressive. Another way
is to summarize the rating *differences* for the pairs of beers:

```{r }
diff <- with(beer, SamAdams - GordonB)
summary(diff)
```

 

and

```{r }
diff <- with(beer, Becks - SierraN)
summary(diff)
```

 

Now we see something: two beers that are similar ought to have a
median rating difference close to 0 most of the time. That is the
case for the first pair; the median *is* zero, and the IQR is
only 2. But for the second pair, Becks is typically rated lower than
Sierra Nevada (negative median) and the IQR is larger (4 rather than
2). 
 
$\blacksquare$

(f) In our cluster analysis, we found that Anchor Steam, Pete's
Wicked Ale, Guinness and Sierra Nevada were all in the same
cluster. Would you expect them to be close together on your map? Are
they? Explain briefly.
 
Solution


If they are in the same cluster, we would expect them to
"cluster" together on the map. Except that they don't, really.
These are the four beers over on the right of our map. They are kind
of in the same general neighbourhood, but not really what you would
call close together. (This is a judgement call, again.) In fact, none
of the beers, with the exception of Sam Adams and Gordon Biersch in
the centre, are really very close to any of the others.

That may be the story, in the end. With the cluster analysis, we were
forcing the clustering to happen, whether it was really there or
not. We haven't seen a test for the "significance" of clusters, so
the clusters we found may not mean very much.
 
$\blacksquare$



