##  Rating beer


 Thirty-two students each rated 10 brands of beer:


* Anchor Steam

* Bass

* Beck's

* Corona

* Gordon Biersch

* Guinness

* Heineken

* Pete's Wicked Ale

* Sam Adams

* Sierra Nevada

The ratings are on a scale of 1 to 9, with a higher
rating being better.
The data are in
[link](http://www.utsc.utoronto.ca/~butler/d29/beer.txt).  I
abbreviated the beer names for the data file. I hope you can figure
out which is which.


(a) Read in the data, and look at the first few rows.
 
Solution


Data values are aligned in columns, but the column headers are
not aligned with them, so `read_table2`:
```{r }
my_url <- "http://www.utsc.utoronto.ca/~butler/d29/beer.txt"
beer <- read_table2(my_url)
beer
```

       
32 rows (students), 11 columns (10 beers, plus a column of student
IDs).  All seems to be kosher. If beer can be kosher.
`r tufte::margin_note("I  investigated. It can; in fact, I found a long list of kosher beers  that included Anchor Steam.")`
 

(b) The researcher who collected the data wants to see which
beers are rated similarly to which other beers. Try to create a
distance matrix from these data and explain why it didn't do what
you wanted. (Remember to get rid of the `student` column
first.) 
 
Solution


The obvious thing is to feed these ratings into `dist`
(we are *creating* distances rather than re-formatting
things that are already distances). We need to skip the first
column, since those are student identifiers:
```{r }
beer %>%
  select(-student) %>%
  dist() -> d
glimpse(d)
```

   

The 496 distances are:

```{r }
32 * 31 / 2
```

 

the number of ways of choosing 2 objects out of 32, when order does
not matter.
Feel free to be offended by my choice of the letter `d` to
denote both data frames (that I didn't want to give a better name to)
and dissimilarities in `dist` objects.

You can look at the whole thing if you like, though it is rather
large. A `dist` object is stored internally as a long vector
(here of 496 values); it's displayed as a nice triangle. The clue here
is the thing called `Size`, which indicates that we have a
$32\times 32$ matrix of distances *between the 32 students*, so
that if we were to go on and do a cluster analysis based on this
`d`, we'd get a clustering of the *students* rather than
of the *beers*, as we want. (If you just print out `d`,
you'll see that is of distances between 32 (unlabelled) objects, which
by inference must be the 32 students.)

It might be interesting to do a cluster analysis of the 32 students
(it would tell you which of the students have similar taste in beer),
but that's not what we have in mind here.
 

(c) The R function `t()` *transposes* a matrix: that
is, it interchanges rows and columns. Feed the transpose of your
read-in beer ratings into `dist`. Does this now give
distances between beers?
 
Solution


Again, omit the first column. The pipeline code looks a bit weird:
```{r }
beer %>%
  select(-student) %>%
  t() %>%
  dist() -> d
```

   

so you should feel free to do it in a couple of steps. This way shows
that you can also refer to columns by number:

```{r }
beer %>% select(-1) -> beer2
d <- dist(t(beer2))
```

 

Either way gets you to the same place:

```{r }
d
```

 

There are 10 beers with these names, so this is good.
 

(d) Try to explain briefly why I used `as.dist` in the
class example (the languages one) but `dist` here. (Think
about the form of the input to each function.)
 
Solution


`as.dist` is used if you *already* have
dissimilarities (and you just want to format them right), but
`dist` is used if you have 
*data on variables* and you want to *calculate*
dissimilarities. 
 

(e) <a name="part:beer-dendro">*</a> Obtain a clustering of the beers, using Ward's method. Show
the dendrogram.
 
Solution


This:
```{r khas}
beer.1 <- hclust(d, method = "ward.D")
plot(beer.1)
```

       
 

(f) What seems to be a sensible number of clusters? Which
beers are in which cluster?
 
Solution


This is a judgement call. Almost anything sensible is
reasonable. I personally think that two clusters is good, beers
Anchor Steam, Pete's Wicked Ale, Guinness and Sierra Nevada in
the first, and Bass, Gordon Biersch, Sam Adams, Corona, Beck's,
and Heineken in the second.
You could make a case for three clusters, splitting off
Corona, Beck's and Heineken  into their own cluster, or even
about 5 clusters as 
Anchor Steam, Pete's Wicked Ale; Guinness, Sierra Nevada; Bass,
Gordon Biersch, Sam Adams; Corona; Beck's, Heineken.

The idea is to have a number of clusters sensibly smaller than
the 10 observations, so that you are getting some actual
insight. Having 8 clusters for 10 beers wouldn't be very
informative! (This is where you use your own knowledge about
beer to help you rationalize your choice of number of clusters.) 

Extra: as to why the clusters split up like this, I think the four
beers on the left of my dendrogram are "dark" and the six on
the right are "light" (in colour), and I would expect the
students to tend to like all the beers of one type and not so
much all the beers of the other type.

You knew I would have to investigate this, didn't you? Let's aim
for a scatterplot of all the ratings for the dark  beers,
against the ones for the light beers. 

Start with the data frame read in from the file:

```{r }
beer
```

       

The aim is to find the average rating for a dark beer and a light beer
for each student, and then plot them against each other. Does a
student who likes dark beer tend not to like light beer, and vice versa?

Let's think about what to do first.

We need to: `pivot_longer` all the rating columns into one, labelled
by `name` of beer. Then create a variable that is `dark`
if we're looking at one of the dark beers and `light`
otherwise. `ifelse` works like "if" in a spreadsheet: a
logical thing that is either true or false, followed by a value if
true and a value if false. There is a nice R command `%in%`
which is `TRUE` if the thing in the first variable is to be
found somewhere in the list of things given next (here, one of the
apparently dark beers). (Another way to do this, which will appeal to
you more if you like databases, is to create a second data frame with
two columns, the first being the beer names, and the second being
`dark` or `light` as appropriate for that beer. Then you
use a "left join" to look up beer type from beer name.)

Next, group by beer type within student. Giving two things to
`group_by` does it this way: the second thing within 
(or "for each of") the first. 

Then calculate the mean
rating within each group. This gives one column of students, one
column of beer types, 
and one column of rating means. 

Then we need to `pivot_wider` beer type
into two columns so that we can make a scatterplot of the mean ratings
for light and dark against
each other. 

Finally, we make a scatterplot. 

You'll see the final version of this that worked, but rest assured
that there were many intervening versions of this that didn't!

I urge you to examine the chain one line at a time and see what each
line does. That was how I debugged it.

Off we go:


```{r iyrpoydf,message=F}
beer %>%
  pivot_longer(-student, names_to="name", values_to="rating") %>%
  mutate(beer.type = ifelse(name %in%
    c("AnchorS", "PetesW", "Guinness", "SierraN"), "dark", "light")) %>%
  group_by(student, beer.type) %>%
  summarize(mean.rat = mean(rating)) %>%
  pivot_wider(names_from=beer.type, values_from=mean.rat) %>%
  ggplot(aes(x = dark, y = light)) + geom_point()
```

 

After all that work, not really. There are some students who like
light beer but not dark beer (top left), there is a sort of vague
straggle down to the bottom right, where some students like dark beer
but not light beer, but there are definitely students at the top
right, who just like beer! 

The only really empty part of this plot is
the bottom left, which says that these students don't hate both kinds
of beer; they like either dark beer, or light beer, or both.

The reason a `ggplot` fits into this "workflow" is that the
first thing you feed into `ggplot` is a data frame, the one
created by the chain here. Because it's in a pipeline, 
you don't have the
first thing on `ggplot`, so you can concentrate on the
`aes` ("what to plot") and then the "how to plot it". 
Now back to your regularly-scheduled programming.
 

(g) Re-draw your dendrogram with your clusters indicated.
 
Solution


`rect.hclust`, with your chosen number  of clusters:
```{r sdkjdh}
plot(beer.1)
rect.hclust(beer.1, 2)
```

       

Or if you prefer 5 clusters, like this:

```{r ljashkjsdah}
plot(beer.1)
rect.hclust(beer.1, 5)
```

 

Same idea with any other number of clusters. If you follow through
with your preferred number of clusters from the previous part, I'm good.
 

(h) Obtain a K-means
clustering with 2 clusters.
`r tufte::margin_note("If you haven't gotten to K-means clustering yet, leave this and save it for later.")`
Note that you will need to use the (transposed) 
*original  data*, not the distances. Use a suitably large value of
`nstart`. (The data are ratings all on the same scale, so
there is no need for `scale` here. In case you were
wondering.) 
 
Solution


```{r echo=FALSE}
set.seed(457299)
```

       
I used 20 for `nstart`. This is the pipe way:
```{r }
beer.2 <- beer %>%
  select(-1) %>%
  t() %>%
  kmeans(2, nstart = 20)
```

       

Not everyone (probably) will get the same answer, because of the
random nature of the procedure, but the above code should be good
whatever output it produces.
 

(i) How many beers are in each cluster?
 
Solution


On mine:
```{r }
beer.2$size
```

       

You might get the same numbers the other way around.
 

(j) *Which* beers are in each cluster? You can do this
simply by obtaining the cluster memberships and using
`sort` as in the last question, or you can do it as I did
in class by obtaining the 
names of the things to be clustered and picking out the ones of
them that are in cluster 1, 2, 3, \ldots .)
 
Solution


The cluster numbers of each beer are these:

```{r }
beer.2$cluster
```

  

This is what is known in the business as a "named vector": it has values (the cluster numbers) and each value has a name attached to it (the name of a beer).

Named vectors are handily turned into a data frame with `enframe`:

```{r }
x <- enframe(beer.2$cluster)
x
```

Or, to go back the other way, `deframe`:

```{r}
deframe(x)
```

 

or, give the columns better names and arrange them by cluster:

```{r }
enframe(beer.2$cluster, name = "beer", value = "cluster") %>%
  arrange(cluster)
```

 

These happen to be the same clusters as in my 2-cluster solution using
Ward's method.
 


