##  Bell peppers and too much water


 A pathogen called *Phytophthora capsici* causes bell
peppers to wilt and die. It is thought that too much water aids in the
spread of the pathogen. Two fields are under study, labelled
`a` and `b`. The first step in the research project is
to compare the mean soil  water content of the two fields.  There is
a suspicion that field `a` will have a higher water content
than field `b`. The data
are in the
file
[link](http://www.utsc.utoronto.ca/~butler/d29/bellpepper.csv).


(a) Read the file in using `read_csv`, and
list the resulting data frame.
 
Solution


Reading directly from the URL is easiest:
```{r }
my_url <- "http://www.utsc.utoronto.ca/~butler/d29/bellpepper.csv"
pepper <- read_csv(my_url)
pepper
```

     

If you like, find out how many observations you have from each field, thus:

```{r }
pepper %>% count(field)
```

 

Fourteen and sixteen.
 

(b) Make side-by-side boxplots of the water content values for
the two fields. How do the fields seem to compare?
 
Solution


This kind of thing:
```{r dartmouth}
ggplot(pepper, aes(x = field, y = water)) + geom_boxplot()
```

     

This one is rather interesting: the distribution of water contents for
field `a` is generally higher than that for field `b`,
but the median for `a` is actually *lower*.

The other reasonable plot is a facetted histogram, something like this:

```{r }
ggplot(pepper, aes(x = water)) + geom_histogram(bins = 6) +
  facet_grid(field ~ .)
```

 

The distribution of water content in field `b` is actually
bimodal, which is probably the explanation of the funny thing with the
median. What actually seems to be happening (at least for these data)
is that the water content in field B is either about the same as field
A, or a lot less (nothing in between).  I can borrow an idea from
earlier to find the five-number summaries for each field:

```{r }
pepper %>%
  nest(-field) %>%
  mutate(qq = map(data, ~ enframe(quantile(.$water)))) %>%
  unnest(qq) %>%
  mutate(pctile = parse_number(name)) %>%
  select(-name) %>%
  spread(pctile, value)
```

 

This is a weird one: all the quantiles are greater for field A
*except* for the median.
 

(c) Do a two-sample $t$-test to test whether there is evidence
that the mean water
content in field `a` is higher than that of field
`b`. What do you conclude? Explain briefly. (You'll need to
figure out a way of doing a one-sided test, or how to adapt the
results from a two-sided test.)
 
Solution


```{r }
t.test(water ~ field, alternative = "greater", data = pepper)
```

     

Note the use of `alternative` to specify that the first group
mean (that of field `a`) is bigger than the second, field
`b`, under the alternative hypothesis.

The P-value, 0.0274, is less than 0.05, so we reject the null (equal
means) in favour of the `a` mean being bigger than the
`b` mean: field `a` really does have a higher mean water
content. 

Another way to tackle this is to do a two-sided test and adapt the P-value:

```{r }
t.test(water ~ field, data = pepper)
```

 

This time we do *not* go straight to the P-value. First we check
that we are on the correct side, which we are since the sample mean
for field `a` *is* bigger than for field `b`. Then we are
entitled to take the two-sided P-value 0.0548 and *halve* it to
get the same 0.0274 that we did before.
 

(d) Is the result of your test consistent with the boxplot, or
not? Explain briefly.
 
Solution


The test said that field `a` had a greater mean water
content. Looking at the boxplot, this is consistent with where the
boxes sit (`a`'s box is higher up than
`b`'s). However, it is not consistent with the medians,
where `b`'s median is actually *bigger*.
You have two possible right answers here: comparing the boxes with
the test result (they agree) or comparing the medians with the
test result (they disagree). Either is good. If you like, you
could also take the angle that the two boxes overlap a fair bit,
so it is surprising that the test came out significant. (The
resolution of this one is that we have 30 measurements altogether,
14 and 16 in the two groups, so the sample size is not tiny. With
smaller samples, having overlapping boxes would probably lead to a
non-significant difference.) 
 


