## Number 1 songs



 The data file
[link](http://stat405.had.co.nz/data/billboard.csv) contains a lot of
information about songs popular in 2000. This dataset is untidy.  Our
ultimate aim is to answer "which song occupied the \#1 position for the largest number of weeks?". To do that, we will build a pipe that
starts from the data frame read in from the URL above, and finishes
with an answer to the question. I will take you through this step by
step. Each part will involve adding something to the pipe you built
previously (possibly after removing a line or two that you used to
display the previous result).



(a) Read the data and display what you have.


Solution


```{r }
billboard=read_csv("http://stat405.had.co.nz/data/billboard.csv")
```

       

There are a *lot* of columns. What does this look like?

```{r }
billboard
```

 

On yours, you will definitely see a little arrow top right saying
"there are more columns", and you will have to click on it several
times to see them all.

 

(b) The columns `x1st.week` through
`x76th.week` contain the rank of each song in the Billboard
chart in that week, with week 1 being the first week that the song
appeared in the chart.  Convert all these columns into two: an
indication of week, called `week`, and of rank, called
`rank`. Most songs appeared in the Billboard chart for a
lot less than 76 weeks, so there are missing values, which you
want to remove.  (I say "indication of week" since this will
probably be text at the moment). Display your new data frame. Do
you have fewer columns?  Why do you have a lot more rows? Explain
briefly.


Solution


This is `gather`ing up all those columns, with
`na.rm=T` to get rid of the missings:
```{r }
billboard %>% gather(week,rank,x1st.week:x76th.week,na.rm=T) 
```

         

Another way to do this is with a select-helper: all those column names
end with `week`, so we can select them all thus:

```{r }
billboard %>% gather(week,rank,ends_with("week"),na.rm=T)
```

 

There are now only 9 columns, a lot fewer than we started with. This
is (I didn't need you to say) because we have collected together all
those `week` columns into one (a column called `rank`
with an indication of which `week` it came from). The logic of
the `gather` is that all those columns contain ranks (which is
what make them the same), but they are ranks from different weeks
(which is what makes them different).

What has actually happened is that we have turned "wide" format into
"long" format. This is not very insightful, so I would like you to
go a bit further in your explanation. The original data frame encodes
the rank of each song in each week, and what the `gather` has
done is to make that explicit: in the new data frame, each song's rank
in each week appears in *one* row, so that there are as many rows
as there are song-week combinations. The original data frame had 317
songs over 76 weeks, so this many:

```{r }
317*76
```

 

song-week combinations.

Not every song appeared in the Billboard chart for 76 weeks, so our tidy
data frame has a lot fewer rows than this.

You need to say that the original data frame had each song appearing
once (on one line), but now each song appears on multiple rows, one
for each week that the song was in the chart. Or something equivalent
to that.

 

(c) Display just your two new columns (for the first few
rows). Add something appropriate onto the end of your pipe to do this.


Solution


A `select` is the thing:
```{r }
billboard %>% gather(week,rank,x1st.week:x76th.week,na.rm=T) %>%
select(week, rank)
```

         



(d) Both your `week` and `rank` columns are
(probably) text. Create new columns that contain just the numeric
values, and display just your new columns, again adding onto the
end of your pipe. If it so happens that `rank` is already a number, leave it as it is.
(In the previous part, you probably had some
code that picked out a few columns to display them. Get rid of
that.) 
 

Solution


My `rank` is already a number, so I leave that. 
I'm defining a new column `rank_number` that is just a copy 
of `rank`, so that I can check the two new things 
that should now both be numbers.
`parse_number` is the easiest. Create new column(s),
with `mutate`, that are the `parse_number`-ed
versions of the old one(s).
```{r }
billboard %>% gather(week,rank,x1st.week:x76th.week,na.rm=T) %>%
mutate(week_number=parse_number(week),
rank_number=rank) %>%
select(ends_with("number"))
```

         
You see that these are indeed numbers. (I gave my new columns names
that ended with `number`, which meant that I could select them
with the select-helper `ends_with`. I'm not insisting that you
do this, but it's a handy trick.)

If your `rank` already looks like a number but happens to be
text, you can also convert it with `as.numeric` (or
`as.integer`, since it is actually text that looks like a whole
number. For converting `rank`, these are also good, but for
converting `week`, you need something that will pull out the
number. (`str_extract` from `stringr` will also do it,
but that's beyond our scope now. It's on page 212 of the R book if
you wish to investigate it. But `parse_number` is a lot
easier.) 

 

(e) The meaning of your week-number column is that it refers
to the number of weeks *after* the song first appeared in the
Billboard chart. That is, if a song's first appearance (in
`date.entered)` is July 24, then week 1 is July 24, week 2
is July 31, week 3 is August 7, and so on. Create a column
`current` by adding the appropriate number of *days*,
based on your week number, to `date.entered`. Display
`date.entered`, your week number, and `current` to
show that you have calculated the right thing. Note that you can
add a number of days onto a date and you will get another date.


Solution


There is a (small) gotcha here: if you read carefully, you'll
see that "week 1" is actually "week 0"  in terms of the
number of days to add on to `date.entered`. So you have
to subtract one from the number of weeks before you multiply
it by seven to get a number of days.
After that thinking, this:
```{r }
billboard %>% gather(week,rank,x1st.week:x76th.week,na.rm=T) %>%
mutate(week_number=parse_number(week),
rank_number=rank) %>%
mutate(current=date.entered+(week_number-1)*7) %>%
select(date.entered,week_number,current)
```

         

Don't forget to use your `week`-turned-into-number, or else it
won't work! (This bit me too, so you don't need to feel bad.)

You can also combine the three column-definition statements into one
mutate. It doesn't matter; as soon as you have defined a column, you
can use it in defining another column, even within the same
`mutate`. 

Anyway, the rows displayed are all `week_number` 1, so the
`current` date should be the same as `date.entered`, and
is. (These are all the first week that a song is in the Billboard
chart). 

You might be thinking that this is not much of a check, and you would
be right. A handy trick is to display a random sample of 10 (say) out
of the 5,000-odd rows of the data frame. To do that, add the line
`sample_n(10)` on the end, like this:

```{r echo=F}
set.seed(457299)
```

 

```{r }
billboard %>% gather(week,rank,x1st.week:x76th.week,na.rm=T) %>%
mutate(week_number=parse_number(week),
rank_number=rank) %>%
mutate(current=date.entered+(week_number-1)*7) %>%
select(date.entered,week_number,current) %>%
sample_n(10)  
```

 

This gives a variety of rows to check. The first `current`
should be $27-1=26$ weeks, or about 6 months, after the date the song
entered the chart, and so it is; the second one should be $3-1=2$
weeks after entry, and it is. The third one should be 3 weeks after
September 23; this is (as I figure it) September $23+21=44$, and
September has 30 days, so this is really October 14. Check.

Your random selection of rows is likely to be different from mine, but
the same kind of thinking will enable you to check whether it makes
sense. 

 

(f) Reaching the \#1 rank on the Billboard chart is one of
the highest accolades in the popular music world. List all the
songs that reached `rank` 1. For these songs, list the
artist (as given in the data set), the song title, and the date(s)
for which the song was ranked number 1. Arrange the songs in date
order of being ranked \#1. Display all the songs (I found 55 of them).


Solution


To the previous pipe, add the last lines below. You can use
either `rank` (text) or what I called
`rank_number` (a number). It doesn't matter here,
since we are only checking for equal-to, not something like
"less than":
```{r size="scriptsize"}
billboard %>% gather(week,rank,x1st.week:x76th.week,na.rm=T) %>%
mutate(week_number=parse_number(week),
rank_number=rank) %>%
mutate(current=date.entered+(week_number-1)*7) %>%
filter(rank==1) %>%
arrange(current) %>%
select(artist.inverted,track,current) 
```

         

You'll see the first ten rows, as here, but with clickable buttons to
see the next 10 (and the previous 10 if you have moved beyond 1--10). 
The "artist" column is called `artist.inverted` because, if
the artist is a single person rather than a group, their last name is
listed first. The song title appears in the column `track`. 

The song by Destiny's Child spills into 2001 because it entered the
chart in 2000, and the data set keeps a record of all such songs until
they drop out of the chart. I'm not sure what happened to the song
that was \#1 on January 8, 2000; maybe it entered the chart in
1999
`r tufte::margin_note("Which was the title of a song by Prince.")` and so is not
listed here.

 

(g) Use R to find out which song held the \#1 rank for the
largest number of weeks. For this, you can assume that the song
titles are all unique (if it's the same song title, it's the same
song), but the artists might not be (for example, Madonna might
have had two different songs reach the \#1 rank). The information
you need is in the output you obtained for the previous part, so
it's a matter of adding some code to the end of that.
The last mark was for displaying *only* the song that was
ranked \#1 for the largest number of weeks, or for otherwise
making it easy to see which song it was.


Solution


This is a question of using `count`, but on the
`track` title:
```{r }
billboard %>% gather(week,rank,x1st.week:x76th.week,na.rm=T) %>%
mutate(week_number=parse_number(week),
rank_number=rank) %>%
mutate(current=date.entered+(week_number-1)*7) %>%
filter(rank==1) %>%
arrange(current) %>%
select(artist.inverted,track,current) %>%
count(track)
```

         

Then you can scan down the `n` column, find that the
biggest number is 11, and say: it's the song "Independent Women Part I" by Destiny's Child. This is 3 points (out of 4, when the question
was to be handed in).

But, this is a data frame, so anything we can do to a data frame we
can do to this, like listing out only the row(s) where `n` is
equal to its maximum value:

```{r }
billboard %>% gather(week,rank,x1st.week:x76th.week,na.rm=T) %>%
mutate(week_number=parse_number(week),
rank_number=rank) %>%
mutate(current=date.entered+(week_number-1)*7) %>%
filter(rank==1) %>%
arrange(current) %>%
select(artist.inverted,track,current) %>%
count(track) %>%
filter(n==max(n))
```

 

or arranging them in (most logically, descending) order by `n`
to make it easier to pick out the top one:

```{r }
billboard %>% gather(week,rank,x1st.week:x76th.week,na.rm=T) %>%
mutate(week_number=parse_number(week),
rank_number=rank) %>%
mutate(current=date.entered+(week_number-1)*7) %>%
filter(rank==1) %>%
arrange(current) %>%
select(artist.inverted,track,current) %>%
count(track) %>%
arrange(desc(n))
```

 

Either of those will net you the 4th point.

If you want to be a little bit more careful, you can make an
artist-track combination as below. This would catch occasions where
the same song by two different artists made it to \#1, or two
different songs that happened to have the same title did. It's not
very likely that the same artist would record two *different*
songs with the same title, though it is possible that the same song by
the same artist could appear in the Billboard chart on two different
occasions.
`r tufte::margin_note("As, for example, when Prince died.")`

I think I want to create an artist-song combo fairly early in my pipe,
and then display *that* later, something like this. This means
replacing `track` by my `combo` later in the pipe,
wherever it appears:

```{r }
billboard %>% gather(week,rank,x1st.week:x76th.week,na.rm=T) %>%
mutate(week_number=parse_number(week),
rank_number=rank) %>%
mutate(combo=paste(track,artist.inverted,sep=" by ")) %>%
mutate(current=date.entered+(week_number-1)*7) %>%
filter(rank==1) %>%
arrange(current) %>%
select(combo,current) %>%
count(combo) %>%
arrange(desc(n))
```

 

I don't think it makes any difference here, but it might in other
years, or if you look over several years where you might get cover
versions of the same song performed by different artists.

Zero-point bonus: how many of these artists have you heard of? How
many have your parents heard of? (I followed popular music quite
closely much earlier than this, in the early 1980s in the UK. I
remember both Madonna and U2 when they *first* became
famous. U2's first single was called "Fire" and it just scraped into
the UK top 40. Things changed after that.)

 


