##  Types of childbirth


 Childbirths can be of two types: a
"vaginal" birth in which the child is born through the mother's
 vagina in the normal fashion, and a "cesarean section" where a
 surgeon cuts through the wall of the mother's abdomen, and the baby
 is delivered through the incision. Cesarean births are used when
 there are difficulties in pregnancy or during childbirth that would
 make a vaginal birth too risky.
A hospital kept track of the number of vaginal and Cesarean births
 for the twelve months of 2012. Of interest is whether the Cesarean
 rate (the ratio of Cesarean births to all births) was increasing,
 decreasing or remaining stable over that time.
The data may be found at
 [link](http://ritsokiguess.site/datafiles/birthtypes.txt). The
 columns are the names of the months (in 2012), the number of cesarean
 births and the number of vaginal births. (The data are not real, but
 are typical of the kind of thing you would observe.)


(a) Read the data into R and display your data frame.

Solution


This is a space-delimited text file, which means:

```{r birthtypes-1 }
my_url <- "http://ritsokiguess.site/datafiles/birthtypes.txt"
births <- read_delim(my_url, " ")
births
```

 

Some text and two numbers for each month. Check.

$\blacksquare$

(b) Create a column of actual dates and also a column of
cesarean rates, as defined above. Store your new data frame in a
variable and display it. For the dates, assume that each date is of
the 1st of the month that it belongs to.

Solution


The easiest way is to use `str_c` or `paste` to
create a text date with year, month and day in some order, and
then to use the appropriate function from `lubridate` to
turn that into an actual `date`. If you use
`str_c`, you (probably) need the `sep` thing to
make sure the values get a space between them; `paste`
does this automatically. (The next question is whether
`ymd` or whatever can cope without spaces, but I'm not
exploring that.)
The cesarean rate is `cesarean` divided by
`cesarean` plus `vaginal`:
```{r birthtypes-2 }
library(lubridate)
b2 <- births %>%
  mutate(datestr = str_c("2012", month, "1", sep = " ")) %>%
  mutate(thedate = ymd(datestr)) %>%
  mutate(cesarean_rate = cesarean / (cesarean + vaginal))
b2
```

        

If you don't like that, create columns that contain 2012 and 1 all
the way down. If you set a column name equal to a single value, that
single value gets repeated the right number of times:^[This is  an example of R's so-called "recycling rules".]

```{r birthtypes-3 }
births %>% mutate(year = 2012, day = 1)
```

 

and then use `unite` as in class. The distinction is that
`unite` *only* works on columns. It also "swallows up"
the columns that it is made out of; in this case, the original year,
month and day disappear:

```{r birthtypes-4 }
b3 <- births %>%
  mutate(year = 2012, day = 1) %>%
  unite(datestr, year, month, day) %>%
  mutate(thedate = ymd(datestr)) %>%
  mutate(cesarean_rate = cesarean / (cesarean + vaginal))
b3
```

 

I don't mind which order you glue your year, month and day together,
as long as you construct the dates with the consistent
`lubridate` function. 

$\blacksquare$

(c) Plot the cesarean rate against time, with a smooth
trend. Do you see an upward trend, a downward trend, no trend, or
something else?

Solution


This is a scatterplot with time on the $x$ axis:
```{r tann}
ggplot(b2, aes(x = thedate, y = cesarean_rate)) + geom_point() + geom_smooth()
```

        

I like this better than joining the points by lines, since we already
have a trend on the plot, but you can do that in some contrasting way:

```{r fletzlinger}
ggplot(b2, aes(x = thedate, y = cesarean_rate)) + geom_point() +
  geom_line(linetype = "dashed") + geom_smooth()
```

 
I see a downward trend. ("A downward trend with a wiggle" if you
like.) There is a certain unevenness
in the trend of the actual data, but the overall picture appears to be
downhill. 

$\blacksquare$

(d) Try to summarize the trend you just found with a
correlation. What goes wrong? How can you fix it?

Solution


Something like this is the obvious guess:

```{r birthtypes-5, error=T}
with(b2, cor(thedate, cesarean_rate))
```

 

This fails because `thedate` is not of itself a number. But
lurking in the background is how the date is actually
*represented*: as a number of days since Jan 1, 1970. Thus,
passing it through `as.numeric` might turn it into that:

```{r birthtypes-6 }
b2 %>% mutate(numeric_date = as.numeric(thedate)) -> b5
b5
```

 

A little mental calculation suggests that these dates in 2012 are a
bit over 40 years, that is $40 \times 365 \simeq 14000$ days, since
the "zero" date of Jan 1, 1970, and so it turns out. This suggests
that we can calculate a correlation with the numeric dates:

```{r birthtypes-7 }
with(b5, cor(numeric_date, cesarean_rate))
```

 

and we can make a test of the null hypothesis that the correlation is
zero (against a two-sided alternative) thus:

```{r birthtypes-8 }
with(b5, cor.test(numeric_date, cesarean_rate))
```

 

That downward trend is more than just chance, with a P-value just
under 0.01. Having said that, though, if you look at the confidence
interval for the correlation, it includes almost all the negative
values it could be, so that with only 12 observations we really know
very little about the correlation other than that it appears to be
negative. 

Extra: 
 In practice, you would typically have a much longer time series
 of measurements than this, such as monthly measurements for
 several years. In looking at only one year, like we did here,
 we could get trapped by seasonal effects: for example, cesarean
 rates might always go down through the year and then jump up
 again in January. Looking at several years would enable us to
 disentangle seasonal effects that happen every year from
 long-term trends. (As an example of this, think of Toronto
 snowfall: there is almost always snow in the winter and there
 is never snow in the summer, a seasonal effect, but in
 assessing climate change, you want to think about long-term
 trends in snowfall, after allowing for which month you're
 looking at.)
 
$\blacksquare$


