##  Body fat as repeated measures


 This one is also stolen from STAC32. Athletes are concerned
with measuring their body fat percentage. Two different methods are
available: one using ultrasound, and the other using X-ray
technology. We are interested in whether there is a difference in the
mean body fat percentage as measured by these two methods, and if so,
how big that difference is. Data on 16 athletes are at
[link](http://ritsokiguess.site/datafiles/bodyfat.txt).



(a) Read in the data and check that you have a sensible number of
rows and columns.



Solution


This kind of thing:
```{r bodyfat2-1 }
my_url <- "http://ritsokiguess.site/datafiles/bodyfat.txt"
bodyfat <- read_delim(my_url, " ")
bodyfat
```

 

16 rows (athletes) and 3 columns, one for each measurement
method and one labelling the athletes. All good.



$\blacksquare$

(b) Carry out a suitable (matched-pairs) $t$-test to determine
whether the means are the same or different. 



Solution


Feed the two columns into `t.test` along with
`paired=T`. Remember that this is (in effect) a one-sample
$t$-test, so that you can't use a `data=`. You therefore have
to wrap everything in a `with`:
```{r bodyfat2-2 }
with(bodyfat, t.test(xray, ultrasound, paired = TRUE))
```

 

The test we want is two-sided, so we didn't have to take any special
steps to get that.

$\blacksquare$



(c) What do you conclude from the test?



Solution


The P-value of 0.7623 is not at all small, so there is no way we can
reject the null hypothesis.^[My hat stays on my head.] There
is no evidence of a difference in means; we can act as if the two
methods produce the same mean body fat percentage. 
That is to say, on this evidence we can use either method, whichever
one is cheaper or more convenient.

 
$\blacksquare$

(d) Run a repeated-measures analysis of variance, treating the two
methods of measuring body fat as the repeated measures (ie., playing
the role of "time" that you have seen in most of the other repeated
measures analyses). There is no "treatment" here, so there is
nothing to go on the right side of the squiggle. Insert a `1`
there to mean "just an intercept". Display the results.



Solution


Construct the response variable, run `lm`, construct the
within-subjects part of the design, run `Manova`:

```{r bodyfat2-3 }
bodyfat %>%
  select(xray:ultrasound) %>%
  as.matrix() -> response
bodyfat.1 <- lm(response ~ 1)
methods <- colnames(response)
methods.df <- data.frame(methods=factor(methods))
bodyfat.2 <- Manova(bodyfat.1, idata = methods.df, idesign = ~methods)
bodyfat.2
```

 

$\blacksquare$


(e) Compare your repeated-measures analysis to your matched-pairs
one. Do you draw the same conclusions?



Solution


The P-value for `methods`, which is testing the same thing
as the matched pairs, is 0.7623, which is actually
*identical* to the matched pairs $t$-test, and so the
conclusion is identical also. That is, there is no difference
between the two methods for measuring body fat. 
This goes to show that repeated measures gives the same answer as
a matched-pairs $t$-test in the situation where they both
apply. But repeated measures is, as we have seen, a lot more general.

Since this really is repeated measures, we ought to be able to use
a mixed model here too. We need "long" or "tidy" format, which
we don't have yet. One pipe to save them all, to paraphrase Lord
of the Rings^[The movies of Lord of the Rings were filmed      in New Zealand, which is also the country in which R was first      designed.] --- put all the fat measurements in one column with a
label saying which `method` they were obtained with; create
a column which is the athlete number as a factor; fit the linear
mixed model; see what we can drop from it:
```{r bodyfat2-4 }
bodyfat
bodyfat %>%
  pivot_longer(-athlete, names_to="method", values_to="fat") %>%
  mutate(fathlete = factor(athlete)) %>%
  lmer(fat ~ method + (1 | fathlete), data = .) %>%
  drop1(test = "Chisq")
```

     

Once again, there is no difference between methods, and though the
P-value is different from the matched pairs or profile analysis, it is
very close to those.

If you're not clear about the tidy data frame used for input to
`lmer`, pull the top two lines off the pipeline and see what they produce:

```{r bodyfat2-5 }
bodyfat %>%
  pivot_longer(-athlete, names_to="method", values_to="fat") %>%
  mutate(fathlete = factor(athlete))
```

 

Each athlete now appears twice: once with their `fat` measured
by `xray`, and again with it measured by
`ultrasound`. The column `fathlete` is a factor.

The mixed model took me two goes to get right: I forgot that I needed the
`data=.` in `lmer`, because it works like `lm`
with the model formula first, not the input data. If the pipeline is
going too fast for you, create the tidy data frame and save it, then
use the saved data frame as input to `lmer`.

$\blacksquare$





