##  Body fat revisited


 Athletes are concerned with measuring their body fat
percentage. Two different methods are available: one using ultrasound,
and the other using X-ray technology. We are interested in whether
there is a difference in the mean body fat percentage as measured by
these two methods, and if so, how big that difference is. Data on 16
athletes are at
[link](http://www.utsc.utoronto.ca/~butler/c32/bodyfat.txt). 

We saw this data set before.




(a) Read in the data again.



Solution


This kind of thing. Since you looked at the data (didn't you?),
you'll know that the values are separated by single spaces:

```{r }
myurl="http://www.utsc.utoronto.ca/~butler/c32/bodyfat.txt"
bodyfat=read_delim(myurl," ")
bodyfat  
```

 




(b) Calculate the differences, and make a normal quantile plot of
them. Is there any evidence that normality of differences fails?
Explain briefly. 



Solution


This is a good place to look ahead. We'll need the differences in two
places, most likely: first for the normal quantile plot, and second
for the matched-pairs sign test. So we should calculate and save them first:

```{r }
bodyfat %>% mutate(diff=xray-ultrasound) -> bodyfat2
```

 

I seem to be using a 2 on the end to name my
dataframe-with-differences, but you can use whatever name you like.

Then, not forgetting to use the data frame that we just made:

```{r }
ggplot(bodyfat2,aes(sample=diff))+stat_qq()+stat_qq_line()
```

   
This is showing a little evidence of skewness or outliers  (depending
on your point of view: either is good). The lowest and highest values
are both too high, and the pattern of points on the plot is kind of
curved (which would be evidence of skewness). Or you could say that
the two highest values are too high, with the other values being more
or less in line (that would be evidence of outliers at the upper
end). I like outliers better than skewness, since those bottom-end
points are not far off the line. I would also accept ``no substantial
problems'', if you can make the case that those two highest points are
not too far off the line. With only 16 observations as we have here,
even truly normal data would stray off the line a bit.

As ever, your explanation is more important than your conclusion. Can
you justify what you think?

If you took your differences the other way around, as
`ultrasound` minus `xray`, your plot will also be the
other way around, with the "outliers" at the bottom. That's good
too.





(c) Previously, we did a matched-pairs $t$-test for these data. In
the light of your normal quantile plot, do you think that was a good
idea? Explain briefly.



Solution


We are looking for the differences to be approximately normal,
bearing in mind that we have a sample of size 16, which is not
that large. Say what you think here; the points, if I were giving
any here, would be for the way in which you support it.
The comment I made before when we did a matched-pairs $t$-test
was that the P-value was so large and non-significant that it was
hard to imagine any other test giving a significant
result. Another way of saying that is that I considered these
differences to be "normal enough", given the circumstances.
You might very well take a different view. You could say that
these differences are clearly not normal, and that the sample size
of 16 is not large enough to get any substantial help from the
Central Limit Theorem. From that point of view, running the
$t$-test is clearly not advisable.




(d) Use the sign test appropriately to compare the two methods for
measuring body fat. (Use `smmr` if you wish.) What do you
conclude, as ever in the context of the data?



Solution


That means using a sign test to test the null hypothesis that the
median difference is zero, against the alternative that it is not
zero. (I don't see anything here to indicate that we are looking only
for positive or only for negative differences, so I think two-sided is
right. You need some reason to do a one-sided test, and there isn't
one here.)

Remembering again to use the data frame that has the differences in it:

```{r }
sign_test(bodyfat2, diff, 0)
```

 

The two-sided P-value is 0.4545, so we are nowhere near rejecting the
null hypothesis that the median difference is zero. There is no
evidence that the two methods for measuring body fat show any
difference on average.

The table of aboves and belows says that there were 6 positive
differences and 10 negative ones. This is not far from an even split,
so the lack of significance is entirely what we would expect. 

Extra: this is the same conclusion that we drew the last time we
looked at these data (with a matched-pairs $t$-test). That supports
what I said then, which is that the $t$-test was so far from being
significant, that it could be very wrong without changing the
conclusion. That is what seems to have happened.




