##  Buses to Boulder


 A bus line operates a route from Denver to Boulder (these
places are in Colorado). The
schedule says that the journey time should be 60 minutes. 11
randomly chosen journey times were recorded, and these are in the
file [link](http://ritsokiguess.site/datafiles/buses.txt), with
journey times shown in minutes.


(a) Read the data into R, and display the data frame that you
read in.
 
Solution


Since you can read the data directly from the URL, do that (if
you are online) rather than having to copy and paste and save,
and then find the file you saved.
Also, there is only one column, so you can pretend that there
were multiple columns, separated by whatever you like. It's least
typing to pretend that they were separated by commas like a
`.csv` file:
```{r boulder-1 }
my_url <- "http://ritsokiguess.site/datafiles/buses.txt"
journey.times <- read_csv(my_url)
journey.times
```

       

Using `read_delim` with any delimiter (such as `" "`)
will also work, and is thus also good.

Variable names in R can have a dot (or an underscore, but not a space)
in them. I have grown accustomed to using dots to separate words. This
works in R but not other languages, but is seen by some as
old-fashioned, with underscores being the modern way.^[In some  languages, a dot is used to concatenate bits of text, or as a way of  calling a method on an object. But in R, a dot has no special  meaning, and is used in function names like `t.test`. Or  `p.value`.] 
You can also use what is called "camel case" 
by starting each "word" after the first with an uppercase
letter like this:

```{r boulder-2, eval=F}
journeyTimes <- read_csv(my_url)
```

 

You have to get the capitalization and punctuation right when you use your variables,
no matter what they're called. In any of the cases above, there is no
variable called `journeytimes`.  As Jenny Bryan (in
[link](http://www.stat.ubc.ca/~jenny/STAT545A/block01_basicsWorkspaceWorkingDirProject.html)) 
puts it, boldface in original:
Implicit contract with the computer / scripting language: Computer
will do tedious computation for you. In return, you will be
completely precise in your instructions. Typos matter. Case
matters. **Get better at typing.** 
 
$\blacksquare$

(b) Run a suitable test to see whether there is evidence that
the mean journey time differs from 60 minutes. What do you
conclude? (I want a conclusion that says something about journey
times of buses.)
 
Solution


`t.test` doesn't take a `data=` to say which
data frame to use. Wrap it in a `with`:
```{r boulder-3 }
with(journey.times, t.test(minutes, mu = 60))
```

       

We are testing that the mean journey time is 60 minutes, against the
two-sided alternative (default) that the mean is not equal to 60 minutes. The
P-value, 0.1971, is a lot bigger than the usual $\alpha$ of 0.05, so we cannot
reject the null hypothesis. That is, there is no evidence that the
mean journey time differs from 60 minutes.

As you remember, we have not proved that the mean journey time
*is* 60 minutes, which is what "accepting the null hypothesis"
would be. We have only failed to reject it, in a shoulder-shrugging
kind of way: "the mean journey time *could* be 60 minutes". The
other acceptable word is "retain"; when you say "we retain the null hypothesis", you imply something 
like "we act as if the mean is 60 minutes, at least until we find something better."
 
$\blacksquare$

(c) Give a 95\% confidence interval for the mean journey
time. (No R code is needed here.)
 
Solution


Just read it off from the output: 57.72 to 69.74 minutes.
 
$\blacksquare$

(d) Do you draw consistent conclusions from your test and
confidence interval? Explain briefly.
 
Solution


The test said that we should not reject a mean of 60
minutes. The confidence interval says that 60 minutes is inside
the interval of plausible values for the population mean, which
is another way of saying the same thing. (If we had rejected 60
as a mean, 60 would have been *outside* the confidence interval.)
 
$\blacksquare$

(e) Draw a boxplot of the journey times. Do you see a reason
to doubt the test that you did above?
 
Solution

 The grouping variable is a
"nothing" as in the Ken and Thomas question (part (d)):
```{r brixham}
ggplot(journey.times, aes(x = 1, y = minutes)) + geom_boxplot()
```

       

The assumption behind the $t$-test is that the population from which
the data come has a normal distribution: ie.\ symmetric with no
outliers. A small sample (here we have 11 values) even from a normal
distribution might look quite non-normal (as in Assignment 0 from last
week), so I am not hugely concerned by this boxplot. However, it's
perfectly all right to say that this distribution is skewed, and
therefore we should doubt the $t$-test, because the upper whisker is
longer than the lower one. In fact, the topmost value is very nearly
an outlier:^[Whether you think it is or not may depend on how  many bins you have on your histogram. With 5 bins it looks like an  outlier, but with 6 it does not. Try it and see.]

```{r babbacombe}
ggplot(journey.times, aes(x = minutes)) + geom_histogram(bins = 5)
```

 

and there might be skewness as well, so maybe I should have been concerned.

I would be looking for some intelligent comment on the boxplot: what it
looks like vs.\ what it ought to look like. I don't so much mind what
that comment is, as long as it's intelligent enough.

Perhaps I should draw a normal quantile plot:

```{r boulder-4 }
ggplot(journey.times, aes(sample = minutes)) + stat_qq() + stat_qq_line()
```

 

The normal quantile plot is saying that the problem is actually at the
*bottom* of the distribution: the lowest value is not low enough,
but the highest value is actually *not* too high. So this one
seems to be on the edge between OK and being right-skewed (too bunched
up at the bottom). My take is that with this small sample this is not
too bad. But you are free to disagree.

If you don't like the normality, you'd use a *sign test* and test
that the *median* is not 60 minutes, which you would (at my
guess) utterly fail to reject:

```{r boulder-5 }
library(smmr)
sign_test(journey.times, minutes, 60)
ci_median(journey.times, minutes)
```

 

and so we do. The median could easily be 60 minutes.
 
$\blacksquare$



