
## Making it stop

 If you are driving, and you hit the brakes, how far do you travel before coming to a complete stop? 
Presumably this depends on how fast you are going. 
Knowing this relationship is important in setting speed limits on roads. For example, on a very bendy road, the speed limit needs to be low, because you cannot see very far ahead, and there could be something just
out of sight that you need to stop for. 

Data were collected for a typical car and driver, as shown in \url{http://ritsokiguess.site/datafiles/stopping.csv}. These are American data, so the speeds are miles per hour and the stopping distances are in feet.



(a) Read in and display (probably all of) the data.

Solution


The usual:

```{r}
my_url <- "http://ritsokiguess.site/datafiles/stopping.csv"
stopping <- read_csv(my_url)
stopping
```
There are only eight observations.


$\blacksquare$


(b) Make a suitable plot of the data.

Solution


Two quantitative variables means a scatterplot. Stopping distance is the outcome, so that goes on the $y$-axis:


```{r}
ggplot(stopping, aes(x=speed, y=distance)) + geom_point()
```


$\blacksquare$


(c) Describe any trend you see in your graph.

Solution


It's an upward trend, but not linear: the stopping distance seems to increase faster at higher speeds.



$\blacksquare$


(d) Fit a linear regression predicting stopping distance from speed. (You might have some misgivings about doing this, but do it anyway.)

Solution


Having observed  a curved relationship, it seems odd to fit a straight line. But we are going to do it anyway and then critique what we have:

```{r}
stopping.1 <- lm(distance~speed, data=stopping)
summary(stopping.1)
```

Extra: note that R-squared is actually really high. We come back to that later.


$\blacksquare$


(e) Plot the residuals against the fitted values for this regression.

Solution


```{r}
ggplot(stopping.1, aes(x=.fitted, y=.resid)) + geom_point()
```


$\blacksquare$


(f) What do you learn from the residual plot? Does that surprise you? Explain briefly. 

Solution


The points on the residual plot form a (perfect) curve, so the original relationship was a curve. This is exactly what we saw on the scatterplot, so to me at least, this is no surprise.

(Make sure you say *how you know* that the original relationship was a curve from looking at the residual plot. Joined-up thinking. 
There are *two* ways we know that the relationship is a curve. Get them both.)


$\blacksquare$


(g) What is the actual relationship between stopping distance and speed, according to the physics? See if you can find out. Cite any books or websites that you use: that is, include a link to a website, or give enough information about a book that the grader could find it.

Solution


I searched for "braking distance and speed" and found the first two things below, that seemed to be useful. Later, I was thinking about the fourth point (which came out of my head) and while searching for other things about that, I found the third thing:

- a [British road safety website](https://www.brake.org.uk/info-and-resources/facts-advice-research/road-safety-facts/15-facts-a-resources/facts/1255-speed), that says "The braking distance depends on how fast the vehicle was travelling before the brakes were applied, and is proportional to the square of the initial speed."
- the [Wikipedia article on braking distance](https://en.wikipedia.org/wiki/Braking_distance), which gives the actual formula. This is the velocity squared, divided by a constant that depends on the coefficient of friction. (That is why your driving instructor tells you to leave a bigger gap behind the vehicle in front if it is raining, and an even bigger gap if it is icy.)
- an [Australian math booklet](http://www.amsi.org.au/teacher_modules/pdfs/Maths_delivers/Braking5.pdf) that talks specifically about braking distance and derives the formula (and the other equations of motion).
- also, if you have done physics, you might remember the equation of motion $v^2 = u^2 + 2as$, where $u$ is the initial velocity, $v$ is the final velocity, $a$ is the acceleration and $s$ is the distance covered. In this case, $v=0$ (the car is stationary at the end), and so $-u^2/2a = s$. The acceleration is negative (the car is slowing down), so the left side is, despite appearances, positive. There seems to be a standard assumption that deceleration due to braking is constant (the same for all speeds), at least if you are trying to stop a car in a hurry.

These are all saying that we should add a speed-squared term to our regression, and then we will have the relationship exactly right, according to the physics.

Extra: Another way  to measure how far you are behind the vehicle in front is time.
Many of the British "motorways" (think 400-series highways) were built when I was young, and I remember [a TV commercial](https://www.youtube.com/watch?v=mf5d2DP4Pp0) that said "Only a Fool Breaks the Two Second Rule". `r tufte::margin_note("This is perhaps not a commercial so much as a public safety message.")` In those days (the linked one is from the 1970s), `r tufte::margin_note("There are some typical British cars of the era in the commercial.")` a lot of British drivers were not used to going that fast, or on roads that straight, so this was a way to know how big a gap to leave, so that you had time to take evasive action if needed. The value of the two-second rule is that it works for any speed, and you don't have to remember a bunch of stopping distances. (When I did my (Canadian) driving theory test, I think I worked out and learned a formula for the stopping distances that I could calculate in my head. I didn't have to get very close since the test was multiple-choice.)



$\blacksquare$


(h) Fit the relationship that your research indicated (in the previous part) and display the results. Comment briefly on the R-squared value.

Solution


Add a squared term in `speed`:

```{r}
stopping.2 <- lm(distance~speed+I(speed^2), data=stopping)
summary(stopping.2)
```

The R-squared now is basically 1, so that the model fits very close to perfectly.

Extra: you probably found in your research that the distance should be just something times speed squared, with no constant or linear term. Here, though, we have a significant linear term as well.
That is probably just chance, since the distances in the data look as if they have been rounded off. With more accurate values, I think the linear term would have been closer to zero.

If you want to go literally for the something-times-speed-squared, you can do that. This doesn't quite work:

```{r}
stopping.3x <- lm(distance~I(speed^2), data=stopping)
summary(stopping.3x)
```
because it still has an intercept in it. In R, the intercept is denoted by `1`. It is always included, unless you explicitly remove it. Some odd things start to happen if you remove the intercept, so it is not a good thing to do unless you know what you are doing. The answers [here](https://stats.stackexchange.com/questions/7948/when-is-it-ok-to-remove-the-intercept-in-a-linear-regression-model) have some good discussion. Having decided that you *are* going to remove the intercept,
you can remove it the same way as anything else (see `update` in the multiple regression lecture) with "minus". I haven't shown you this, so if you do it, you will need to cite your source: that is, say where you learned what to do:

```{r}
stopping.3 <- lm(distance~I(speed^2)-1, data=stopping)
summary(stopping.3)
```

The R-squared is still extremely high, much higher than for the straight line. The coefficient value, as I said earlier (citing Wikipedia), depends on the coefficient of friction; the stopping distances you see typically are based on a dry road, so you have to allow extra distance (or time: see above) if the road is not dry.

$\blacksquare$




(i) Somebody says to you "if you have a regression with a high R-squared, like 95%, there is no need to look for a better model." How do you respond to this? Explain briefly.

Solution


An example of a regression with an R-squared of 95% is the straight-line fit from earlier in this question. This is an example of a regression that fits well but is not appropriate because it doesn't capture the form of the relationship. 

In general, we are saying that no matter how high R-squared is, we might still be able to improve on the model we have. The flip side is that we might not be able to do any better (with another data set) than an R-squared of, say, 30%, because there is a lot of variability that is, as best as we can assess it, random and not explainable by anything.

Using R-squared as a measure of absolute model quality is, thus, a mistake. Or, to say it perhaps more clearly, asking "how high does R-squared have to be to indicate a good fit?" is asking the wrong question.
The right thing to do is to concentrate on getting the form of the model right, and thereby get the R-squared as high as we can for that data set (which might be very high, as here, or not high at all).


$\blacksquare$






