##  Can caffeine improve your performance on a test?


 
Does caffeine help students do better on a certain test? To
find out, 36 students were randomly allocated to three groups (12 in
each group).  Each student received a fixed number of cups of coffee
while they were studying, but the students didn't know whether they
were receiving all full-strength coffee ("high"), all decaf coffee
("low") or a 50-50 mixture of the two ("moderate"). For each
subject, their group was recorded as well as their score on the
test. The data are in
[link](http://ritsokiguess.site/datafiles/caffeine.csv), as a
`.csv` file.



(a) Read in and examine the data. How are the values laid out?


Solution


`read_csv` because it's a `.csv` file:
```{r }
my_url <- "http://ritsokiguess.site/datafiles/caffeine.csv"
caffeine.untidy <- read_csv(my_url)
caffeine.untidy
```

     

The first column is the number of the subject (actually within each
group, since each student only tried one amount of caffeine). Then
follow the test scores for the students in each group, one group per column.

I gave the data frame a kind of dumb name, since (looking ahead) I
could see that I would need a less-dumb name for the tidied-up data,
and it seemed sensible to keep `caffeine` for that.



(b) Explain briefly how the data are not "tidy".


Solution


The last three columns are all scores on the test: that is, they
all measure the same thing, so they should all be in the same column.
Or, there should be a column of scores, and a separate column
naming the groups. Or, there were 36 observations in the data, so
there should be 36 rows. You always have a variety of ways to
answer these, any of which will do.



(c) Use a suitable tool from the `tidyverse` to create one
column of test scores and and one column of group labels. Call your
column of group labels `amount`. Is it a `factor`?


Solution


We are combining several columns into one, so this is `pivot_longer`:

```{r}
caffeine.untidy %>% 
  pivot_longer(-Sub, names_to = "amount", values_to = "score") -> caffeine
```



I didn't ask you to list the resulting data frame, but it is smart to
at least look for yourself, to make sure `pivot_longer` has done
what you expected.

```{r }
caffeine
```

 

A column of amounts of caffeine, and a column of test scores. This is
what we expected. There should be 12 each of the `amount`s,
which you can check if you like:

```{r }
caffeine %>% count(amount)
```

 

Indeed.

Note that `amount` is text, not a factor. Does this matter? We'll see.

This is entirely the kind of situation where you need `pivot_longer`,
so get used to seeing where it will be useful.



(d) Obtain side-by-side boxplots of test scores by amount of caffeine.


Solution



```{r }
ggplot(caffeine, aes(x = amount, y = score)) + geom_boxplot()
```

Note that this is *much more difficult* if you don't have a tidy data frame. (Try it and see.) 



(e) Does caffeine amount seem to have an effect? If so, what
kind of effect?


Solution


On average, exam scores seem to be higher when the amount of
caffeine is higher (with
the effect being particularly pronounced for High caffeine). 
If you want to, you can also say the the effect of caffeine seems
to be small, relative to the amount of variability there is (there
is a lot). The point is that you say *something* supported by
the boxplot.



(f) Run a suitable analysis of variance to determine whether
the mean test score is equal or unequal for the three groups. What
do you conclude?


Solution


Something like this:
```{r }
caff.1 <- aov(score ~ amount, data = caffeine)
summary(caff.1)
```

     

The P-value on the $F$-test is less than 0.05, so we reject the null
hypothesis (which says that all the groups have equal means) in favour
of the alternative: the group means are not all the same (one or more
of them is different from the others).

Notice that the boxplot and the `aov` are quite happy for
`amount` to be text rather than a factor (they actually do want
a factor, but if the input is text, they'll create one).



(g) Why is it a good idea to run Tukey's method here?
 

Solution


The analysis of variance $F$-test is significant, so that the
groups are not all the same. Tukey's method will tell us which
group(s) differ(s) from the others. There are three groups, so
there are differences to find that we don't know about yet.



(h) Run Tukey's method. What do you conclude?


Solution


This kind of thing:

```{r }
caff.3 <- TukeyHSD(caff.1)
caff.3
```

 

The high-caffeine group definitely has a higher mean test score than
the no-caffeine group. (The Moderate group is not significantly
different from either of the other groups.)
Both the
comparisons involving Moderate could go either way (the interval for
the difference in means includes zero). The None-High comparison, 
however, is away from zero, so this is the significant one. As is
usual, we are pretty sure that the difference in means (this way
around) is negative, but we are not at all clear about how big it is,
because the confidence interval is rather long.
`r tufte::margin_note("We'd need a  lot more students to make it narrower, but this is not surprising  since students vary in a lot of other ways that were not measured here.")`


Extra: the normality and equal spreads assumptions look perfectly good, given the boxplots, and I don't think there's any reason to consider any other test. You might like to assess that with normal quantile plots:

```{r}
ggplot(caffeine, aes(sample=score)) + stat_qq() +
  stat_qq_line() + facet_wrap(~amount, ncol=2)
```


There's nothing to worry about there normality-wise. If anything, there's a little evidence of *short* tails (in the None group especially), but you'll recall that short tails don't affect the mean and thus pose no problems for the ANOVA. Those three lines also have pretty much the same slope, indicating very similar spreads. Regular ANOVA is the best test here. (Running eg. Mood's median test would be a mistake here, because it doesn't use the data as efficiently (counting only aboves and belows) as the ANOVA does, and so the ANOVA will give a better picture of what differs from what.)