##  Rainfall in California


 The data in
[link](http://ritsokiguess.site/datafiles/calirain.txt) are
rainfall and other measurements for 30 weather stations in
California. Our aim is to understand how well the annual rainfall at
these stations (measured in inches) can be predicted from the other
measurements, which are the altitude (in feet above sea level), the
latitude (degrees north of the equator) and the distance from the
coast (in miles).



(a) Read the data into R. You'll have to be careful here, since the
values are space-delimited, but sometimes by more than one space, to
make the columns line up.  `read_table2`, with filename or
url, will read it in.
One of the variables
is called `rainfall`, so as long as you *do not* call
the data frame that, you should be safe.


Solution


I used `rains` as the name of my data frame: 
```{r calirain-1 }
my_url="http://ritsokiguess.site/datafiles/calirain.txt"
rains=read_table2(my_url)
```

 

I have the right number of rows and columns.

There is also `read_table`, but that requires *all* the
columns, including the header row, to be lined up. You can try that
here and see how it fails.

I don't need you to investigate the data yet (that happens in the next
part), but this is interesting (to me):

```{r calirain-2 }
rains
```

 

Some of the station names are two words, but they have been smooshed
into one word, so that `read_table2` will recognize them as a
single thing. Someone had already done that for us, so I didn't even
have to do it myself. 

If the station names had been two genuine words, a `.csv` would
probably have been the best choice (the actual data values being
separated by commas then, and not spaces).

$\blacksquare$

(b) Make a boxplot of the rainfall figures, and explain why the
values are reasonable. (A rainfall cannot be negative, and it is
unusual for a annual rainfall to exceed 60 inches.) A
`ggplot` boxplot needs *something* on the $x$-axis: the
number 1 will do.


Solution



```{r calirain-3 }
ggplot(rains,aes(y=rainfall,x=1))+geom_boxplot()
```

     
There is only one rainfall over 60 inches, and the smallest one is
close to zero but positive, so that is good.

Another possible plot here is a histogram, since there is only one quantitative variable:

```{r calirain-4 }
ggplot(rains, aes(x=rainfall))+geom_histogram(bins=7)
```



This clearly shows the rainfall value above 60 inches, but some other things are less clear: are those two rainfall values around 50 inches above or below 50, and are those six rainfall values near zero actually above zero? 
Extra: What stations have those extreme values? Should you wish to find out:
```{r calirain-5 }
rains %>% filter(rainfall>60)
```

 
This is a place right on the Pacific coast, almost up into Oregon (it's almost
the northernmost of all the stations). So it makes sense that it would
have a high rainfall, if anywhere does. (If you know anything about
rainy places, you'll probably think of Vancouver and Seattle, in the
Pacific Northwest.) Here it is:
[link](https://www.google.ca/maps/place/Crescent+City,+CA,+USA/@41.7552589,-123.9652917,8.42z/data=!4m5!3m4!1s0x54d066375c6288db:0x76e89ab07375e62e!8m2!3d41.7557501!4d-124.2025913). 
Which station has less than 2 inches of annual rainfall?
```{r calirain-6 }
rains %>% filter(rainfall<2)  
```

 
The name of the station is a clue: this one is in the desert. So you'd
expect very little rain. Its altitude is *negative*, so it's
actually *below* sea level. This is correct. Here is where it is:
[link](https://www.google.ca/maps/place/Death+Valley,+CA,+USA/@36.6341288,-118.2252974,7.75z/data=!4m5!3m4!1s0x80c739a21e8fffb1:0x1c897383d723dd25!8m2!3d36.5322649!4d-116.9325408).

$\blacksquare$

(c) Plot `rainfall` against each of the other quantitative
variables (that is, not `station`). 


Solution


That is, `altitude`, `latitude` and
`fromcoast`. The obvious way to do this (perfectly
acceptable) is one plot at a time:
```{r calirain-7 }
ggplot(rains,aes(y=rainfall,x=altitude))+geom_point()
```

     

```{r calirain-8 }
ggplot(rains,aes(y=rainfall,x=latitude))+geom_point()
```

 

and finally 

```{r calirain-9 }
ggplot(rains,aes(y=rainfall,x=fromcoast))+geom_point()
```

 

You can add a smooth trend to these if you want. Up to you. Just the
points is fine with me.

Here is a funky way to get all three plots in one shot:

```{r calirain-10 }
rains %>% 
  pivot_longer(altitude:fromcoast, names_to="xname",values_to="x") %>%
  ggplot(aes(x=x,y=rainfall))+geom_point()+
  facet_wrap(~xname,scales="free")
```

 

This always seems extraordinarily strange if you haven't run into it
before. The strategy is to put *all* the $x$-variables you want
to plot into *one* column and then plot your $y$ against the
$x$-column.  Thus: make a column of all the $x$'s glued
together, labelled by which $x$ they are, then plot $y$ against $x$
but make a different sub-plot or "facet" for each different
$x$-name. The last thing is that each $x$ is measured on a different
scale, and unless we take steps, all the sub-plots will have the
*same* scale on each axis, which we don't want.

I'm not sure I like how it came out, with three very tall
plots. `facet_wrap` can also take an `nrow` or an
`ncol`, which tells it how many rows or columns to use for the
display. Here, for example, two columns because I thought three was
too many:

```{r calirain-11 }
rains %>% 
  pivot_longer(altitude:fromcoast, names_to="xname",values_to="x") %>%
  ggplot(aes(x=x,y=rainfall))+geom_point()+
  facet_wrap(~xname,scales="free",ncol=2)
```

 

Now, the three plots have come out about square, or at least "landscape", which I like a lot better.

$\blacksquare$

(d) Look at the relationship of each other variable with
`rainfall`. 
Justify the assertion that `latitude` seems most strongly
related with `rainfall`. Is that relationship positive or negative? linear? Explain
briefly. 


Solution


Let's look at the three variables in turn:


* `altitude`: not much of anything. The stations near
sea level have rainfall all over the place, though the three
highest-altitude stations have the three highest rainfalls apart
from Crescent City.

* `latitude`: there is a definite upward trend here, in
that stations further north (higher latitude) are likely to have
a higher rainfall. I'd call this trend linear (or, not obviously
curved), though the two most northerly stations have one higher
and one much lower rainfall than you'd expect.

* `fromcoast`: this is a weak downward trend, though
the trend is spoiled by those three stations about 150 miles
from the coast that have more than 40 inches of rainfall.

Out of those, only `latitude` seems to have any meaningful
relationship with `rainfall`.

$\blacksquare$

(e) Fit a regression with `rainfall` as the response
variable, and `latitude` as your explanatory variable. What
are the intercept, slope and R-squared values? Is there a 
*significant* relationship between `rainfall` and your
explanatory variable? What does that mean?


Solution


Save your `lm` into a
variable, since it will get used again later:
```{r calirain-12 }
rainfall.1=lm(rainfall~latitude,data=rains)
summary(rainfall.1)
```

 

My intercept is $-113.3$, slope is $3.6$ and R-squared is $0.33$ or
33\%. (I want you to pull these numbers out of the output and round
them off to something sensible.) The slope is significantly nonzero,
its P-value being 0.00085: rainfall really does depend on latitude, although
not strongly so.

Extra: Of course, I can easily do the others as well, though you don't have to:

```{r calirain-13 }
rainfall.2=lm(rainfall~fromcoast,data=rains)
summary(rainfall.2)
```

 

Here, the intercept is 23.8, the slope is $-0.05$ and R-squared is a
dismal 0.04 (4\%).  This is a way of seeing that this relationship is
really weak, and it doesn't even have a curve to the trend or anything
that would compensate for this. I looked at the scatterplot again and
saw that if it were not for the point bottom right which is furthest
from the coast and has almost no rainfall, there would be almost no
trend at all. The slope here is not significantly different from zero,
with a P-value of 0.265.

Finally:

```{r calirain-14 }
rainfall.3=lm(rainfall~altitude,data=rains)
summary(rainfall.3)
```

 

The intercept is 16.5, the slope is 0.002 and the R-squared is 0.09 or
9\%, also terrible. The P-value is 0.105, which is not small enough to
be significant. 

So it looks as if it's only `latitude` that
has any impact at all. This is the only explanatory variable with a
significantly nonzero slope. On its own, at least.

$\blacksquare$

(f) Fit a multiple regression predicting `rainfall` from
all three of the other (quantitative) variables. Display the
results. Comment is coming up later.


Solution


This, then:
```{r calirain-15 }
rainfall.4=lm(rainfall~latitude+altitude+fromcoast,data=rains)
summary(rainfall.4)
```

 

$\blacksquare$

(g) What is the R-squared for the regression of the last part?
How does that compare with the R-squared of your regression in part (e)?


Solution


The R-squared is 0.60 (60\%), which is quite a bit bigger than the
R-squared of 0.33 (33\%) we got back in (e). 

$\blacksquare$

(h) What do you conclude about the importance of the variables
that you did *not* include in your model in
(e)? Explain briefly.


Solution


Both variables `altitude` and `fromcoast` are
significant in this regression, so they have *something to add* over and above `latitude` when it comes to
predicting rainfall, even though (and this seems odd) they have no
apparent relationship with `rainfall` on their own. 
Another way to say this is that the three variables work together
as a team to predict rainfall, and together they do much better
than any one of them can do by themselves.  
This also goes to show that the scatterplots we began
with don't get to the heart of multi-variable relationships,
because they are only looking at the variables two at a time.

$\blacksquare$

(i) Make a suitable hypothesis test that the variables
`altitude` and `fromcoast` significantly improve the
prediction of `rainfall` over the use of `latitude`
alone. What do you conclude?


Solution


This calls for `anova`. Feed this two fitted models,
smaller (fewer explanatory variables) first. The null hypothesis
is that the two models are equally good (so we should go with the
smaller); the alternative is that the larger model is better, so
that the extra complication is worth it:
```{r calirain-16 }
anova(rainfall.1,rainfall.4)  
```

 

The P-value is small, so we reject the null in favour of the
alternative: the regression with all three explanatory variables fits
better than the one with just `latitude`, so the bigger model
is the one we should go with.

If you have studied these things: this one is a 
"multiple-partial $F$-test", for testing the combined significance of more than one $x$
but less than all the $x$'s.^[If you had just one $x$, you'd  use a $t$-test for its slope, and if you were testing all the $x$'s, you'd use the global $F$-test that appears in the regression output.]

$\blacksquare$




