## Endogenous nitrogen excretion in carp



 A paper in Fisheries Science reported on variables that
affect "endogenous nitrogen excretion" or ENE in carp raised in
Japan. A number of carp were divided into groups based on body weight,
and each group was placed in a different tank. The mean body weight of
the carp placed in each tank was recorded. The carp were then fed a
protein-free diet three times daily for a period of 20 days. At the
end of the experiment, the amount of ENE in each tank was measured, in
milligrams of total fish body weight per day. (Thus it should not
matter that some of the tanks had more fish than others, because the
scaling is done properly.)

For this question, write a report in R Markdown that answers the
questions below and contains some narrative that describes your
analysis. Create an HTML document from your R Markdown.



(a) Read the data in from
[link](http://www.utsc.utoronto.ca/~butler/c32/carp.txt). There are 10
tanks. 


Solution


Just this. Listing the data is up to you, but doing so and
commenting that the values appear to be correct will improve your report.
```{r }
my_url="http://www.utsc.utoronto.ca/~butler/c32/carp.txt"
carp=read_delim(my_url," ")  
carp
```

 



(b)??scatter-one?? Create a scatterplot of ENE (response) against bodyweight
(explanatory). Add a smooth trend to your plot.


Solution


```{r }
ggplot(carp,aes(x=bodyweight,y=ENE))+geom_point()+
geom_smooth()
```

 

This part is just about getting the plot. Comments are coming in a
minute. Note that `ENE` is capital letters, so that
`ene` will not work.



(c) Is there an upward or downward trend (or neither)? Is the
relationship a line or a curve? Explain briefly.


Solution


The trend is downward: as bodyweight increases, ENE
decreases. However, the decrease is rapid at first and then levels
off, so the relationship is nonlinear. I want some kind of
support for an assertion of non-linearity: anything that says that
the slope or rate of decrease is not constant is good.



(d)??linear?? Fit a straight line to the data, and obtain the R-squared
for the regression.


Solution


`lm`. The first stage is to fit the straight line, saving
the result in a  variable, and the second stage is to look at the
"fitted model object", here via `summary`:
```{r }
carp.1=lm(ENE~bodyweight,data=carp)
summary(carp.1)
```

 
Finally, you need to give me a (suitably rounded) value for
R-squared: 46.6\% or 47\% or the equivalents as a decimal. I just
need the value at this point.
This kind of R-squared is actually pretty good for natural data, but
the issue is whether we can improve it by fitting a non-linear
model.\endnote{The suspicion being that we can, since the
scatterplot suggested serious non-linearity.}



(e) Obtain a residual plot (residuals against fitted values)
for this regression. Do you see any problems? If so, what does that
tell you about the relationship in the data?


Solution



This is the easiest way: feed the output of the regression
straight into `ggplot`:

```{r }
ggplot(carp.1,aes(x=.fitted,y=.resid))+geom_point()
```

 



(f) Fit a parabola to the data (that is, including an
$x$-squared term). Compare the R-squared values for the models in
this part and part (??linear??). Does that suggest that the parabola
model is an improvement here over the linear model?


Solution


Add bodyweight-squared to
the regression. Don't forget the `I()`:
```{r }
carp.2=lm(ENE~bodyweight+I(bodyweight^2),data=carp)
summary(carp.2)
```

 

R-squared has gone up from 47\% to 74\%, a substantial
improvement. This suggests to me that the parabola model is a
substantial improvement.\endnote{Again, not a surprise, given our initial
scatterplot.} 

I try to avoid using the word "significant" in this context, since
we haven't actually done a test of significance.

The reason for the `I()` is that the up-arrow has a special
meaning in `lm`, relating to interactions between factors (as
in ANOVA), that we don't want here. Putting `I()` around it
means "use as is", that is, raise bodyweight to power 2, rather than
using the special meaning of the up-arrow in `lm`.

Because it's the up-arrow that is the problem, this applies whenever
you're raising an explanatory variable to a power (or taking a
reciprocal or a square root, say).



(g) Is the test for the slope coefficient for the squared term
significant? What does this mean?


Solution


Look along the `bodyweight`-squared line to get a P-value
of 0.031. This is less than the default 0.05, so it *is*
significant.
This means, in short, that the quadratic model is a significant
*improvement* over the linear one.\endnote{Now we can use that word
"significant".} 
Said longer: the null hypothesis being tested is that the slope
coefficient of the squared term is zero (that is, that the squared
term has nothing to add over the linear model). This is rejected,
so the squared term has *something* to add in terms of
quality of prediction.



(h) Make the scatterplot of part (??scatter-one??), but add
the fitted curve. Describe any way in which the curve fails to fit well.


Solution


This is a bit slippery, because the points to plot and the
fitted curve are from different data frames. What you do in this
case is to put a `data=` in one of the `geom`s,
which says "don't use the data frame that was in the  `ggplot`, but use this one instead". 
I would think about
starting with the regression object `carp.2` as my base
data frame, since we want (or I want) to do two things with
that: plot the fitted values and join them with lines. Then I
want to add the original data, just the points:
```{r }
ggplot(carp.2,aes(x=carp$bodyweight,y=.fitted),colour="blue")+
geom_line(colour="blue")+
geom_point(data=carp,aes(x=bodyweight,y=ENE))
```

       

This works, but is not very aesthetic, because the bodyweight that is
plotted against the fitted values is in the wrong data frame, and so
we have to use the dollar-sign thing to get it from the right one.

A better way around this is "augment" the data with output from the regression object. 
This is done using `augment` from 
package `broom`:
```{r augment1, eval=F, error=TRUE}
library(broom)
carp.2a=augment(carp, carp.2)
carp.2a
```

 

so now you see what `carp.2a` has in it, and then:

```{r error=TRUE}
g=ggplot(carp.2a,aes(x=bodyweight,y=.fitted))+
geom_line(colour="blue")+
geom_point(aes(y=ENE))
```

 

This is easier coding: there are only two non-standard things. The
first is that the fitted-value lines should be a distinct colour like
blue so that you can tell them from the data points. The second thing
is that for the second `geom_point`, the one that plots the data,
the $x$ coordinate `bodyweight` is correct so that we don't
have to change that; we only have to change the $y$-coordinate, which
is `ENE`. The plot is this:

```{r error=T}
g
```

 

Concerning interpretation, you have a number of possibilities
here. The simplest is that the points in the middle are above the
curve, and the points at the ends are below. (That is, negative
residuals at the ends, and positive ones in the middle, which gives
you a hint for the next part.) Another is that the parabola curve
fails to capture the *shape* of the relationship; for example, I
see nothing much in the data suggesting that the relationship should go
back up, and even given that, the fitted curve doesn't go especially
near any of the points.

I was thinking that the data should be fit better by something like
the left half of an upward-opening parabola, but I guess the curvature
on the left half of the plot suggests that it needs most of the left
half of the parabola just to cover the left half of the plot.

The moral of the story, as we see in the next part, is that the
parabola is the wrong curve for the job.



(i) Obtain a residual plot for the parabola model. Do you see
any problems with it? (If you do, I'm not asking you to do anything
about them in this question, but I will.)


Solution


The same idea as before for the other residual plot.  Use the
fitted model object `carp.2` as your data frame for the
`ggplot`:
```{r }
ggplot(carp.2,aes(x=.fitted,y=.resid))+geom_point()
```

 

I think this is *still* a curve (or, it goes down and then
sharply up at the end). Either way, there is still a pattern. 

That was all I needed, but as to what this means: our parabola was a
curve all right, but it appears not to be the right *kind* of
curve. I think the original data looks more like a hyperbola (a curve
like $y=1/x$) than a parabola, in that it seems to decrease fast and
then gradually to a limit, and *that* suggests, as in the class
example, that we should try an asymptote model. Note how I specify it,
with the `I()` thing again, since `/` has a special meaning 
to `lm` in the same way that 
`^` does:
```{r }
carp.3=lm(ENE~I(1/bodyweight),data=carp)
summary(carp.3)
```

 

That fits *extraordinarily* well, with an R-squared up near
98\%. The intercept is the asymptote, which suggests a (lower) limit
of about 5.2 for `ENE` (in the limit for large bodyweight). We
would have to ask the fisheries scientist whether this kind of thing
is a reasonable biological mechanism. It says that a carp always has
some ENE, no matter how big it gets, but a smaller carp will have a
lot more.

Does the fitted value plot look reasonable now? This is `augment` again since the fitted values and observed data come from different data frames:
```{r augment2, error=T}
library(broom)
augment(carp, carp.3) %>% 
ggplot(aes(x=bodyweight,y=.fitted))+
geom_line(colour="blue")+
geom_point(aes(y=ENE))
```

 

I'd say that does a really nice job of fitting the data. But it would
be nice to have a few more tanks with large-bodyweight fish, to
convince us that we have the shape of the trend right.

And, as ever, the residual plot. That's a lot easier than the plot we
just did:

```{r error=T}
ggplot(carp.3,aes(x=.fitted,y=.resid))+geom_point()
```

 

All in all, that looks pretty good (and certainly a vast improvement
over the ones you got before).

When you write up your report, you can make it flow better by writing
it in a way that suggests that each thing was the obvious thing to do
next: that is, that *you* would have thought to do it next,
rather than me telling you what to do.

My report (as an R Markdown file) is at
[link](http://www.utsc.utoronto.ca/~butler/c32/carp.Rmd). Download it,
knit it, play with it.




