

 You remember the study of attitudes towards research on
cat's brains? We are going to use the same data again, but for a
different purpose this time. The data are at
[link](http://ritsokiguess.site/datafiles/decision.txt).



(a) Read in the data again.


Solution


Nothing very surprising here:
```{r catbrain-b-1 }
my_url <- "http://ritsokiguess.site/datafiles/decision.txt"
decide <- read_delim(my_url)
str(decide)
```

       

I wanted some data with variables of different types, and this one was handy.

$\blacksquare$

(b) A normal quantile
plot, `ggplot`-style. There are two parts to the process: in
the `aes` goes `sample=` (on the right side of the
equals goes the data), and instead of a `geom` goes
`stat_qq` and `stat_qq_line` Make a normal quantile plot of the
`relativism` scores, with a line as well. Comment on what you see.


Solution


According to the hints, it looks like this:
```{r maeyers}
ggplot(decide, aes(sample = relativism)) + stat_qq() + stat_qq_line()
```

       

So what is this telling us? If you ignore the very lowest values of
`relativism`, the points form a curve, which indicates
skewness. The highest values are (slightly) too bunched up and the
lowest values, aside from the very low ones, are too spread out, which
means skewed to the left. But the very lowest ones are all bunched up,
as if 3 is a lower limit, which is exactly *not* what you'd
expect with left skewness.

Now try it with `idealism`, and comment on its normality or lack thereof. 

$\blacksquare$




