##  The brain of a cat, revisited

Earlier, we looked at an ethics study that had to do with a fictional brain of a fictional cat. I said
there was actually a *town* called Catbrain. It's in England,
near Bristol, and seems to be home to a street of car dealerships.


(a) Find the latitude and longitude of "Catbrain UK" (though I don't think there are any others).

Solution

Make sure you have these two packages loaded:

```{r }
library(leaflet)
library(tmaptools)
```

To find the latitude and longitude of Catbrain:

```{r}
catbrain <- tibble(place = "Catbrain UK")
catbrain %>% mutate(ll = list(geocode_OSM(place))) %>% 
  unnest_wider(ll) %>% 
  unnest_wider(coords) -> catbrain
catbrain
```

Remember that the output from `geocode_OSM` is a list, and it has in it a thing called `coords` that is the longitude and latitude together, and another thing called `bbox` that we don't use. So we have to `unnest` *twice* to get the longitude (as `x`) and the latitude (as `y`) out for drawing in a moment.

(b) Draw a map of Catbrain using Leaflet.


Solution

That goes this way:

```{r}
leaflet(data = catbrain) %>% 
  addTiles() %>% 
  addCircleMarkers(lng = ~x, lat = ~y) -> catbrain_map
catbrain_map

```


 
 
There are car dealerships are along Lysander Road. Zoom in
to see them. Or zoom out to see where this is. You can keep zooming out until you know where you are, using the plus and minus buttons, or your mouse wheel.


The name Catbrain, according to
[link](http://www.bristolpost.co.uk/news/history/name-catbrain-hill-came-825247),
means "rough stony soil", from Middle English, and has nothing to do
with cats or their brains at all.


Extra: I was actually surprised that this worked at all, because with only one point, how does it know what scale to draw the map? Also, unless your UK geography is really good, you won't have any clue about exactly where this is. That's the reason for the next part.

(c) Make a dataframe containing some other British cities as well as Catbrain, and find their latitudes and longitudes.


Solution

I chose the cities below, mostly somewhere near Catbrain. You could fire up a Google map, zoom it out until it contains something you know, and pick some places you've heard of. (I happen to know British geography pretty well, so I just picked some mostly nearby places out of my head. I didn't really want to pick London, but I figured this was the one *you* might know.)

```{r cache=T}
catbrain2 <- tribble(
  ~where,
  "Catbrain UK",
  "Bristol UK",
  "Taunton UK",
  "Newport UK",
  "Gloucester UK",
  "Cardiff UK",
  "Birmingham UK",
  "London UK",
  "Caldicot UK"
)
catbrain2 %>%
  rowwise() %>% 
  mutate(ll = list(geocode_OSM(where))) %>% 
  unnest_wider(ll) %>% 
  unnest_wider(coords) -> catbrain2
catbrain2
```

The first time I did this, I forgot the `rowwise`, which we didn't need the first time (there was only one place), but here, it causes odd problems if you omit it. 

(d) Draw a map containing the places you picked.

Solution

The map-drawing is almost the same, just changing the dataframe:

```{r}
leaflet(data = catbrain2) %>% 
  addTiles() %>% 
  addCircleMarkers(lng = ~x, lat = ~y)
```

 

Now, if you have any sense of the geography of the UK, you know where
you are.  The big river (the Severn) is
the border between England and Wales, so
the places above and to the left of it are in Wales, including
Caldicot (see question about Roman pottery).

You can zoom this map *in* (once you have figured out which of the circles is Catbrain) and find Lysander Road again, and also the M5 (see below).

More irrelevant extra: the M5 is one of the English "motorways" (like 400-series highways
or US Interstates). The M5 goes from Birmingham to Exeter. You can
tell that this is England because of the huge number of traffic
circles, known there as "roundabouts". One of the first things they
teach you in British driving schools is how to handle roundabouts:
which lane to approach them in, which (stick-shift) gear to be in, and
when you're supposed to signal where you're going. I hope I still
remember all that for when I next drive in England!
    



