##  Sugar in breakfast cereals


 The data for this question are in
[http://ritsokiguess.site/datafiles/cereal-sugar.txt](http://ritsokiguess.site/datafiles/cereal-sugar.txt). The
story here is whether breakfast cereals marketed to children have a
lot of sugar in them; in particular, whether they have more sugar on
average than cereals marketed to adults.


(a) Read in the data (to R) and display the data set. Do you have
a variable that distinguishes the children's cereals from the
adults' cereals, and another that contains the amount of sugar?



Solution


```{r }
my_url="http://ritsokiguess.site/datafiles/cereal-sugar.txt"
cereals=read_delim(my_url," ")
cereals
``` 

The variable `who` is a categorical variable saying who the
cereal is intended for, and the variable `sugar` says how much
sugar each cereal has.





(b) Calculate the mean sugar content for each group of cereals (the
adults' ones and the children's ones). Do they look similar or different?



Solution


`group_by` and `summarize`:
```{r }
cereals %>% group_by(who) %>%
summarize(sugar_mean=mean(sugar))
``` 

These means look very different, though it would be better to look at
a boxplot (coming up in a moment).



(c) Make side-by-side boxplots of the sugar contents of the two
types of cereal. What do you see that is out of the ordinary? 




Solution


The usual:

```{r }
ggplot(cereals,aes(x=who,y=sugar))+geom_boxplot()
``` 

I see outliers: two high ones on the adults' cereals, and one high and
one low on the children's cereals.

My thought above about the means being very different is definitely supported by the medians being very different on the boxplots. We should have no trouble declaring that the "typical" amounts of sugar in the adults' and children's cereals are different. 





(d) Explain briefly why you would not trust a two-sample $t$-test
with these data. (That is, say what the problem  is, and why it's a
problem.)



Solution


The problem is the outliers (which is rather a giveaway), but the
reason it's a problem is that the two-sample $t$-test assumes
(approximately) normal data, and a normal distribution doesn't
have outliers.
Not only do you need to note the outliers, but you also need to
say why the outliers cause a problem *in this case*. Anything
less than that is not a complete answer. 




(e) Run a suitable test to see whether the "typical" amount of
sugar differs between adult's and children's cereals. Justify the test
that you run. (You can use the version of your test that lives in a
package, if that is easier for you.) What do you conclude, in the
context of the data?



Solution


Having ruled out the two-sample $t$-test, we are left with Mood's
median test. I didn't need you to build it yourself, so you can
use package `smmr` to run it with:
```{r }
library(smmr)
median_test(cereals,sugar,who)
```     

We conclude that there *is* a difference between the median
amounts of sugar between the two groups of cereals, the P-value of 0.00000007
being extremely small.

Why did it come out so small? Because the amount of sugar was smaller
than the overall median for almost all the adult cereals, and larger
than the overall median for almost all the children's ones. That is,
the children's cereals really do have more sugar.

Mood's median test doesn't come with a confidence interval (for the
difference in population medians), because whether or not a certain
difference in medians is rejected depends on what those medians
actually are, and the idea of the duality of the test and CI doesn't
carry over as we would like.

My daughter likes chocolate Cheerios, but she also likes Shredded
Wheat and *Bran Flakes*. Go figure. (Her current favourite is
Raisin Bran, even though she doesn't like raisins by themselves.)

Mood's median test is the test we should trust, but you might be
curious about how the $t$-test stacks up here:

```{r }
t.test(sugar~who,data=cereals)
``` 

The P-value is *even smaller*, and we have the advantage of
getting a confidence interval for the difference in means: from about
30 to about 40 units less sugar in the adult cereals. Whatever the
units were.





