##  Roller coasters


 A poll on the Discovery Channel asked people to nominate the
best roller-coasters in the United States. We will examine the 10
roller-coasters that received the most votes. Two features of a
roller-coaster that are of interest are the distance it drops from
start to finish, measured here in feet\endnote{Roller-coasters work by
gravity, so there must be some drop.} and the duration of the ride,
measured in seconds. Is it true that roller-coasters with a bigger
drop also tend to have a longer ride? The data are at
[link](http://www.utsc.utoronto.ca/~butler/c32/coasters.csv).\endnote{These
are not to be confused with what your mom insists that you place
between your coffee mug and the table.}



(a) Read the data into R and verify that you have a sensible
number of rows and columns.


Solution


A `.csv`, so the usual for that:

```{r }
my_url="http://www.utsc.utoronto.ca/~butler/c32/coasters.csv"
coasters=read_csv(my_url)
coasters
```

 

The number of marks for this kind of thing has been decreasing through
the course, since by now you ought to have figured out how to do it
without looking it up.

There are 10 rows for the promised 10 roller-coasters, and there are
several columns: the drop for each roller-coaster and the duration of
its ride, as promised, as well as the name of each roller-coaster and the state
that it is in. (A lot of them seem to be in Ohio, for some reason that
I don't know.) So this all looks good.



(b) Make a scatterplot of duration (response) against drop
(explanatory), labelling each roller-coaster with its name in such a
way that the labels do not overlap. Add a regression line to your plot.


Solution


The last part, about the labels not overlapping, is an
invitation to use `ggrepel`, which is the way I'd
recommend doing this. (If not, you have to do potentially lots
of work organizing where the labels sit relative to the points,
which is time you probably don't want to spend.) Thus:
```{r }
library(ggrepel)
ggplot(coasters,aes(x=drop,y=duration,label=coaster_name))+
geom_point()+geom_text_repel()+geom_smooth(method="lm",se=F)
```

       

The `se=F` at the end is optional; if you omit it, you get that
"envelope" around the line, which is fine here.

Note that with the labelling done this way, you can easily identify
which roller-coaster is which.



(c) Would you say that roller-coasters with a larger drop tend
to have a longer ride? Explain briefly.


Solution


I think there are two good answers here: "yes" and "kind of".
Supporting "yes" is the fact that the regression line does go
uphill, so that overall, or on average, roller-coasters with a
larger drop do tend to have a longer duration of ride as well.
Supporting "kind of" is the fact that, though the regression
line goes uphill, there are a lot of roller-coasters that are
some way off the trend, far from the regression line.
I am happy to go with either of those. I could also go with
"not really" and the same discussion that I attached to ``kind
of''. 



(d) Find a roller-coaster that is unusual compared to the
others. What about its combination of `drop` and
`duration` is unusual?


Solution


This is an invitation to find a point that is a long way off the
line. I think the obvious choice is my first one below, but I
would take either of the others as well:


* "Nitro" is a long way above the line. That means it has
a long duration, relative to its drop. There are two other
roller-coasters that have a larger drop but not as long a
duration. In other words, this roller-coaster drops slowly,
presumably by doing a lot of twisting, loop-the-loop and so on.

* "The Beast" is a long way below the line, so it has a
short duration relative to its drop. It is actually the
shortest ride of all, but is only a bit below average in terms
of drop. This suggests that The Beast is one of those rides
that drops a long way quickly.

* "Millennium Force" has the biggest drop of all, but a
shorter-than-average duration. This looks like another ride
with a big drop in it.

A roller-coaster that is "unusual" will have a residual that
is large in size (either positive, like Nitro, or negative, like
the other two). I didn't ask you to find the residuals, but if
you want to, `augment` from `broom` is the
smoothest way to go:
```{r }
library(broom)
duration.1=lm(duration~drop,data=coasters) 
augment(duration.1,coasters) %>%
select(coaster_name,duration,drop,.resid) %>%
arrange(desc(abs(.resid)))
```

       

`augment` produces a data frame (of the original data frame
with some new columns that come from the regression), so I can feed it
into a pipe to do things with it, like only displaying the columns I
want, and arranging them in order by absolute value of residual, so
that the roller-coasters further from the line come out first. This
identifies the three that we found above. The fourth one, ``Ghost
Rider'', is like Nitro in that it takes a (relatively) long time to
fall not very far.
You can also put `augment` in the *middle* of a pipe, but then
you seem to lose things that were not in the regression:

```{r }
coasters %>%
lm(duration~drop, data=.) %>%
augment() %>%
arrange(desc(abs(.resid)))
```

 

I wanted to hang on to the roller-coaster names, so I went the other
way. The advantage of this way is less typing. You have to specify the
`data=.` thing in the regression because the data frame is
*not* the first input to `lm` (the model is), but you have
a much simpler `augment` because its first input is the model
that came out of the previous step. If the second input to
`augment` is missing, as it is here, it ``attempts to
reconstruct the data from the model''\endnote{A quote from the package
vignette.} which I think means that things like the coaster names that
were not part of the regression won't be part of `augment`'s
output either. That's my understanding. My first way explicitly
supplies the original data frame to `augment` so there is no
question about what it is using.




