##  College plans


 5199 male high school seniors in Wisconsin
`r tufte::margin_note("I don't  know why Wisconsin again, but that's what it is.")` were classified by
socio-economic status (low, lower-middle, upper-middle, high), by
the degree that their parents encouraged them in their education (low
or high),
and whether or not they had plans to go to college (yes or no). How,
if at all, are these categorical variables related? The data can be
found at
[link](http://www.utsc.utoronto.ca/~butler/d29/college-plans.txt). 



(a) Read in the data and check that you have a column for each
variable and a column of frequencies.

Solution


Delimited by one space:
```{r }
my_url <- "http://www.utsc.utoronto.ca/~butler/d29/college-plans.txt"
wisc <- read_delim(my_url, " ")
wisc
```

     

As promised. We only have 16 observations, because we have all
possible combinations of categorical variable combinations, 4 social
strata, times 2 levels of encouragement, times 2 levels of college
plans. 

Each line of the data file summarizes a number of students, not just
one.  For example, the first line says that 749 students were in the
lower social stratum, received low encouragement and have no college
plans. If we sum up the frequencies, we should get 5199 because there
were that many students altogether:

```{r }
wisc %>% summarize(tot = sum(frequency))
```

 


(b) Fit a log-linear model containing all possible
interactions. You don't need to examine it yet.

Solution


```{r }
wisc.1 <- glm(frequency ~ social.stratum * encouragement * college.plans,
  data = wisc, family = "poisson"
)
```

   



(c) Find out which terms (interactions) could be removed. Do you
think removing any of them is a good idea?


Solution


This is `drop1`. If you forget the `test=`, you won't
get any P-values:
```{r }
drop1(wisc.1, test = "Chisq")
```

   

This P-value is not small, so the three-way interaction can be removed.



(d) Remove anything you can, and fit again. Hint: `update`.


Solution


In this kind of modelling, it's easier to describe what changes
should be  made to get from one model to another, rather than
writing out the whole thing from scratch again.
Anyway, the three-way interaction can come out:
```{r }
wisc.2 <- update(wisc.1, . ~ . - social.stratum:encouragement:college.plans)
```

   



(e) Continue to examine what can be removed, and if reasonable,
remove it, until you need to stop. Which terms are left in your final model?


Solution


Start with `drop1`:
```{r }
drop1(wisc.2, test = "Chisq")
```

   

These are all strongly significant, so they have to stay. There is
nothing else we can remove. All the two-way interactions have to stay
in the model.



(f) Make two-way tables of any remaining two-way interactions, and
describe any relationships that you see.


Solution


We have three two-way tables to make.

My first one is social stratum by parental encouragement. Neither of
these is really a response, but I thought that social stratum would
influence parental encouragement rather than the other way around, hence:

```{r }
xtabs(frequency ~ social.stratum + encouragement, data = wisc) %>%
  prop.table(margin = 1)
```

 

This says that there tends to be more parental encouragement, the
higher the social stratum. 
Next, this:

```{r }
xtabs(frequency ~ social.stratum + college.plans, data = wisc) %>%
  prop.table(margin = 1)
```

 

In this one (and the next), `college.plans` is the response, in
columns, so we want to have the *rows* adding up to 1. 

The higher the social stratum, the more likely is a male
high school senior to have plans to go to college. (The social stratum
is not in order, so you'll have to jump from the second row to the
third to the fourth to the first to assess this. Lower and lower
middle are not in order, but the others are.)

Finally, this:

```{r }
xtabs(frequency ~ encouragement + college.plans, data = wisc) %>%
  prop.table(margin = 1)
```

 

And here you see an *enormous* effect of parental encouragement
on college plans: if it is low, the high-school senior is very
unlikely to be considering college.

Nothing, in all honesty, that is very surprising here. But the two-way
interactions are easier to interpret than a three-way one would have
been.

Here, we think of college plans as being a response, and this
analysis has shown that whether or not a student has plans to go to
college depends separately on the socio-economic status and the level
of parental encouragement (rather than on the combination of both, as
would have been the case had the three-way interaction been
significant). 





