##  Computerized accounting


 Beginning accounting students need to
learn to learn to audit in a computerized environment. A sample of
beginning accounting students took each of two tests: the Computer
Attitude Scale (CAS) and the Computer Anxiety Rating Scale (CARS). A
higher score in each indicates greater anxiety around computers. The test scores are scaled to be between 0 and 5. Also
noted was each student's gender. The data are in
[http://ritsokiguess.site/datafiles/compatt.txt](http://ritsokiguess.site/datafiles/compatt.txt). The data values are separated by spaces.



(a) Read the data into R. Do you have what you expected? Explain briefly. 


Solution


Read in and display the data. This, I think, is the easiest way.
```{r compatt-1 }
my_url <- "https://raw.githubusercontent.com/nxskok/datafiles/master/compatt.txt"
anxiety=read_delim(my_url," ")
anxiety
```     

There is a total of 35 students with a CAS score, a CARS score and a
gender recorded for each. This is in line with what I was
expecting. (You can also note that the genders appear to be a mixture
of males and females.)
    
$\blacksquare$

(b) How many males and females were there in the sample?


Solution


Most easily `count`:
```{r compatt-2 }
anxiety %>% count(gender)
```     

This also works (and is therefore good):

```{r compatt-3 }
anxiety %>% group_by(gender) %>% summarize(count=n())
``` 

I want you to use R to do the counting (that is, don't just go through
the whole data set  and count the males and
females yourself). This is because you might have thousands of data
values and you need to learn how to get R to count them for
you.

15 females and 20 males, *which you should say*. I made a point
of *not* saying that it is enough to get the output with the
answers on it, so you need to tell me what the answer is.
    
$\blacksquare$

(c) Do the CAS scores tend to be higher for females or for males? Draw a suitable graph to help you decide, and come to a conclusion.


Solution


Gender is categorical and CAS score is quantitative, so a boxplot would appear to be the thing:
```{r compatt-4 }
ggplot(anxiety,aes(x=gender,y=CAS))+geom_boxplot()
```     

The median for males is slightly higher, so male accountants are more anxious around computers than female accountants are.

If you wish, side-by-side (or, better, above-and-below) histograms
would also work:

```{r compatt-5 }
ggplot(anxiety,aes(x=CAS))+geom_histogram(bins=6)+
facet_wrap(~gender,ncol=1)
``` 

If you go this way, you have to make a call about where the centres of the histograms are. I guess the male one is slightly further to the right, but it's not so easy to tell. (Make a call.)
    
$\blacksquare$

(d) Find the median CAS scores for each gender. Does this support what you saw on your plot? Explain briefly.


Solution


Group-by and summarize:
```{r compatt-6 }
anxiety %>% group_by(gender) %>% summarize(med=median(CAS))
```     

The median is a bit higher for males, which is what I got on my
boxplot (and is apparently the same thing as is on the histograms, but
it's harder to be sure there).
    
$\blacksquare$

(e) Find the mean and standard deviation of both CAS and CARS scores (for all the students combined, ie.\ not separated by gender) *without* naming those columns explicitly.


Solution


Without naming them explicitly means using some other way to pick them out of the data frame, such as `summarize` with `across`. 

The basic `across` comes from asking yourself what the *names* of those columns have in common: they start with C and the gender column doesn't:

```{r compatt-7 }
anxiety %>% summarize(across(starts_with("C"), 
                             list(mean = \(cc) mean(cc), sd = \(cc) sd(cc))))
``` 


Inside the calculation of the mean and SD, the columns you have chosen can have any name. You can use `x`, or, I used `cc` as a reminder to me that these columns have names that begin with C.^[The name `c` has a special meaning in R, so I avoided that.]

Another way is to ask what *property* these two columns have in common: they are the only two numeric (quantitative) columns. This means using an `across` with a `where` inside it, thus:

```{r compatt-8 }
anxiety %>% summarize(across(where(is.numeric), 
                             list(mean = \(num) mean(num), sd = \(num) sd(num))))
```     

Again, you can call the selected columns `x`, or you can have a name like `num` that reminds you that these are the numeric ones.

Read the first one as "across all the columnns whose names start with C, find the mean and SD of them." The second one is a little clunkier: "acrosss all the columns for which `is.numeric` is true, find the mean and SD of them". A shorter way for the second one is "across all the numeric (quantitative) columns, find their mean and SD", but then you have to remember exactly how to code that. The reason for the `list` is that we are calculating two statistics for each column that we find. I am using a "named list" so that the mean gets labelled with a `mean` on the end of the column name, and the SD gets an `sd` on the end. For this kind of `summarize`, the names of the summaries can be short (and don't need to contain a variable name), because they will be glued onto the end of the names of the variables for which they are the mean and SD.

Either of these is good, or anything equivalent (like noting that the two anxiety scales both `ends\_with` S):

```{r compatt-9 }
anxiety %>% summarize(across(ends_with("S"), 
                             list(mean = \(s) mean(s), sd = \(s) sd(s))))
``` 

Because I didn't say otherwise, you should say what the means and SDs are, rounding off suitably: the CAS scores have mean 2.82 and SD 0.48, and the CARS scores have mean 2.77 and SD 0.67.

Yet another way to do it is to select the columns you want first (which
you can also do by number so as not to name them, or use the select-helper that you put inside the `across`), and then find the mean
and SD of all of them:

```{r compatt-10 }
anxiety %>% select(2:3) %>% 
    summarize(across(everything(), 
                     list(mean = \(x) mean(x), sd = \(x) sd(x))))
``` 

This doesn't work:

```{r compatt-11 }
summary(anxiety)
``` 

because, although it gets the means, it does not get the standard
deviations. (I added the SD to the original question to make you find
a way other than this.)

In summary, find a way to get those answers without naming those
columns in your code, and I'm good.

In case you were wondering about how to do this separately by gender, well, put the `group\_by` in like you did before:

```{r compatt-12 }
anxiety %>% group_by(gender) %>%
  summarize(across(where(is.numeric), 
                   list(mean = \(num) mean(num), sd = \(num) sd(num))))
``` 

or

```{r compatt-13 }
anxiety %>% group_by(gender) %>%
    summarize(across(starts_with("C"), 
                    list(mean = \(cc) mean(cc), sd = \(cc) sd(cc))))
``` 

The male means are slightly higher on both tests, but the male
standard deviations are a little smaller.
You might be wondering whether the test scores are related. They are both quantitative, so the obvious way to find out is a scatterplot:

```{r compatt-14 }
ggplot(anxiety,aes(x=CAS, y=CARS))+geom_point()
``` 

The two variables can be on either axis, since there is no obvious
response or explanatory variable. A higher score on one scale goes
with a higher score on the other, suggesting that the two scales are
measuring the same thing.

This plot mixes up the males and females, so you might like to distinguish them, which goes like this:

```{r compatt-15 }
ggplot(anxiety,aes(x=CAS,y=CARS,colour=gender))+geom_point()
``` 

There is a slight (but only slight) tendency for the males to be up
and to the right, and for the females to be down and to the left. This
is about what you would expect, given that the male means are slightly
bigger on both scores, but the difference in means is not that big
compared to the SD.
    
$\blacksquare$




