## Chocolate chips



 A famous cookie manufacturer claims that
their bags of chocolate chip cookies contain "more than 1100 chocolate chips on average". A diligent group of students buys 16
bags of these cookies and counts the number of chocolate chips in each
bag. The results are in [http://www.utsc.utoronto.ca/~butler/c32/chips.txt](http://www.utsc.utoronto.ca/~butler/c32/chips.txt).



(a) Read in and display (some of) the data.


Solution


I'll pretend it's a
`.csv` this time, just for fun.  Give the data frame a
name different from `chips`, so that you don't get
confused:
```{r }
bags=read_csv("chips.txt")
bags
``` 

That looks sensible.       



(b) Build your own sign test in R for testing that the median is
1100 chocolate chips, against the alternative that it is greater.
(Do this as in class: count the appropriate thing,
compare it with an appropriate binomial distribution, and obtain a
P-value.


Solution


The null median is 1100, so we count the number
of values above and below:

```{r }
bags %>% count(chips<1100)
``` 

The un-standard thing there is that we can put a logical condition
directly into the `count`. If you don't think of that, you can
also do this, which creates a new variable `less` that is
`TRUE` or `FALSE` for each bag appropriately:

```{r }
bags %>% mutate(less=(chips<1100)) %>% count(less)
``` 

or the more verbose

```{r }
bags %>% mutate(less=(chips<1100)) %>%
group_by(less) %>% summarize(howmany=n())
``` 

Just one value below, with all the rest above. 
Getting the right P-value, properly, requires some careful thought
(but you will probably get the right answer anyway). If the
alternative hypothesis is true, and the median is actually bigger than
1100 (say, 1200), you would expect half the data values to be bigger
than 1200 and half smaller. So *more* than half the data values
would be bigger than *1100*, and fewer than half of them would be
less than 1100. So, if we are going to reject the null (as it looks as
if we will), that small number of values below 1100 is what we want.

The P-value is the probability of a value 1 or less in a binomial
distribution with $n=16, p=0.5$:

```{r }
sum(dbinom(0:1,16,0.5))
``` 

Or, equivalently, count *up* from 15:

```{r }
sum(dbinom(15:16,16,0.5))
``` 

This is correctly one-sided, so we don't have to do anything with
it. 



(c) Use my R package `smmr` to reproduce your sign test
above, and verify that you get consistent results. (See the
maze-design question for instructions on installing this, if you
haven't yet.)


Solution


This will mean reading the output carefully:
```{r }
library(smmr)
sign_test(bags,chips,1100)
```       

This time, we're doing a one-sided test, specifically an
*upper-tail* test, since we are looking for evidence that the
median is *greater than* 1100. The results are exactly what we
got "by hand": 15 values above and one below, and a P-value (look
along the `upper` line) of 0.00026.
 The two-sided P-value of 0.00052 rounds to the same 0.0005 as SAS got.

Alternatively, you can do this:

```{r }
sign_test0(bags$chips,1100)
``` 

with the same result (but only go this way if you need to).



(d) Use `smmr` to obtain a 95\% confidence interval for the
median number of chocolate chips per bag of cookies.


Solution


Once everything is in place, this is simplicity itself:
```{r }
ci_median(bags,chips)
```       

1135 to 1325. I would round these off to whole numbers, since the data
values are all whole numbers. These values are all above 1100, which
supports the conclusion we got above that the median is above
1100. This is as it should be, because the CI is "all those medians that would *not* be rejected by the sign test". 

Or, 

```{r }
ci_median0(bags$chips)
``` 




