##  Growing corn


 A new type of corn seed has been developed.
The people developing it want to know if the type of soil the seed
is planted in has an impact on how well the seed performs, and if so,
what kind of impact. Three
outcome measures were used: the yield of corn produced (from a fixed
amount of seed), the amount of water needed, and the amount of
herbicide needed. The data are in
[link](http://ritsokiguess.site/datafiles/cornseed.csv). 32 fields
were planted with the seed, 8 fields with each soil type.



(a) Read in the data and verify that you have 32 observations
with the correct variables.


Solution


The usual:
```{r cornseed-1 }
my_url <- "https://raw.githubusercontent.com/nxskok/datafiles/master/cornseed.csv"
cornseed <- read_csv(my_url)
cornseed
```

     

We have 32 rows; we have a categorical soil type, three
numerical columns containing the yield, water and herbicide values,
and we also have a label for each of the 32 fields (which is actually
a number, but we don't have to worry about that, since we won't be
using `field` for anything).
    
$\blacksquare$

(b) Run a multivariate analysis of variance to see whether
the type of soil has any effect on any of the variables. What do you
conclude from it?


Solution


The usual thing: create the response, use `manova` (or
`Manova` from `car` if you like, but it's not necessary):
```{r cornseed-2 }
response <- with(cornseed, cbind(yield, water, herbicide))
cornseed.1 <- manova(response ~ soil, data = cornseed)
summary(cornseed.1)
```

     

With a P-value (just) less than 0.05, soil type has some effect on the
response variables: that is, it affects one or more of the three
responses, or some combination of them. ANOVA conclusions are usually
vague, and MANOVA conclusions are vaguer than most. We will try to
improve on this. But with an only-just-significant P-value, we should
not be expecting miracles.

We ought to check Box's M test:

```{r}
summary(BoxM(response, cornseed$soil))
```

No problems with unequal variances or correlations, at least.

Here and below, `field` is neither a response variable nor an
explanatory variable; it is an experimental unit, so `field`
acts as an ID rather than anything else. So `field` should not
be part of any of the analyses; if it did appear, the only way it
could is as a factor, for example if this was somehow a repeated
measures analysis over the three response variables. In that case,
`lmer`, if you were going that way, would use `field` as
a random effect.

The variables to include are the
yield, water and herbicide as measured response variables, and soil
type, as the categorical explanatory variable. (For the discriminant
analysis, these get turned around: the grouping variable `soil`
acts like a response and the others act as explanatory.)
    
$\blacksquare$

(c) Run a discriminant analysis on these data, "predicting"
soil type from the three response variables. Display the results.


Solution


```{r cornseed-3 }
cornseed.2 <- lda(soil ~ yield + water + herbicide, data = cornseed)
cornseed.2
```

     

No `field` in here, for reasons discussed above. (I'm not even
sure how you *can* run a discriminant analysis with a factor
variable on the right of the squiggle.) The fields were numbered by
soil type:

```{r cornseed-4 }
cornseed %>% select(field, soil)
```

 

so evidently if you know the field number you can guess the field
type, but we didn't care about that: we cared about whether you can
distinguish the fields by yield, water, herbicide or combination
thereof. 
    
$\blacksquare$

(d) <a name="part:corn-svd">*</a> 
Which linear discriminants seem to be worth paying attention to?
Why did you get three linear discriminants? Explain briefly.


Solution


Look for  "proportion of trace" in the output.

The first one is *way* bigger than the others, which says that
the first linear discriminant is way more important (at separating the
groups) than either of the other two.

As to why we got three: there are 3 variables and 4 groups (soil
types), and the smaller of 3 and $4-1$ is 3.
    
$\blacksquare$

(e) Which response variables do the important linear
discriminants depend on? Answer this by extracting something from
your discriminant analysis output.


Solution


The table "coefficients of linear discriminants".
We said earlier that the only important discriminant is
`LD1`. On that, the only notably non-zero coefficient is for
`herbicide`; the ones for `yield` and `water` are
close to zero. That is to say, the effects of the soil types play out
through herbicide and not either of the other two variables.

I didn't ask you to, but you could check this by seeing how
`herbicide` differs according to soil type:

```{r cornseed-5 }
ggplot(cornseed, aes(x = soil, y = herbicide)) + geom_boxplot()
```



The fields in `loam` soil needed more herbicide than the others.

Or by `water`:

```{r cornseed-6 }
ggplot(cornseed, aes(x = soil, y = water)) + geom_boxplot()
```

 

There isn't much difference in the amount of water needed between any
of the fields, no matter what soil type. 

This confirms that `water` is not distinguished by soil type,
while `herbicide` is (at least to some extent).
    
$\blacksquare$

(f) Obtain predictions for the discriminant analysis. (You
don't need to do anything with them yet.)


Solution


Just this, therefore:
```{r cornseed-7 }
cornseed.pred <- predict(cornseed.2)
```

     
    
$\blacksquare$

(g) Plot the first two discriminant scores against each other,
coloured by soil type. You'll have to start by making a data frame
containing what you need.


Solution


I changed my mind from the past about how to do this. I make a big data frame out of the data and predictions (with `cbind`) and go from there:
```{r cornseed-8 }
d <- cbind(cornseed, cornseed.pred)
head(d)
```

     
Then we use this as input to `ggplot`:

```{r cornseed-9 }
ggplot(d, aes(x = x.LD1, y = x.LD2, colour = soil)) + geom_point()
```

 

    
$\blacksquare$

(h) On your plot that you just made, explain briefly how `LD1`
distinguishes at least one of the soil types.



Solution


Find a soil type that is typically high (or low or average) on
LD1. Any one or more of these will do: loam soils are typically high on LD1,
clay soils or salty soils are typically low on LD1; sandy soils are
typically average on LD1. (There are exceptions, but I'm looking for
"typically".)
  
$\blacksquare$

(i) On your plot, does `LD2` appear to do anything to
separate the groups? Is this surprising given your earlier findings?
Explain briefly.



Solution


All the soil groups appear go to about the full height of the plot:
that is to say, none of the groups appear to be especially at the
top or the bottom. That means that `LD2` does not separate
the groups at all. Back in part (<a href="#part:corn-svd">here</a>), we said that
the first linear discriminant is way more important than either of
the other two, and here we see what that means: `LD2` does nothing to
separate the groups. So it's not a surprising finding at all.
I thought earlier about asking you to plot only the first linear
discriminant, and now we see  why: only the first one separates the
groups. If you wanted to do that, you could  make a boxplot of the
discriminant scores by `soil` group, thus:
```{r cornseed-10 }
ggplot(d, aes(x = soil, y = x.LD1)) + geom_boxplot()
```

   
This says more or less the same thing as your plot of `LD1` and
`LD2`: `loam` has the highest `LD1` score,
`sandy` is about in the middle, and `clay` and
`salty` have typically negative `LD1` scores, similar to
each other, though there is one outlying `salty` that looks a
lot more like a `loam`.
  

$\blacksquare$

(j) Make a table of actual and predicted `soil`
group. Which soil type was classified correctly the most often? 



Solution



```{r cornseed-11, error=T}
with(d, table(obs = soil, pred = class))
```

   

Or, the `tidyverse` way, which is below.

There were 8 fields of each soil type. The soil type that has the most
of its fields classified correctly (based on the values of the
response variables) has the biggest number down the diagonal of the
table: looking at 3, 6, 5 and 4, we see that the `loam` soil
type had the most of its fields classified correctly, so this was the
most distinct from the others. (We also saw this on the plot of
`LD1` vs. `LD2`: the `loam` fields were all over
on the right.)

This was easier because we had the same number of fields of each
type. If we didn't have that, the right way to go then would be to work out
*row* percentages: 
"out of the fields that were actually sandy, what percent of them got classified as sandy", 
and so on.

This is not a perfect classification, though, which is about what you
would expect from the soil types being intermingled on the plot of
`LD1` vs. `LD2`. If you look at the table,
`salty` and `sandy` are fairly distinct also, but
`clay` is often confused with both of them. On the plot of
`LD1` and `LD2`, `salty` is generally to the left
of `sandy`, but `clay` is mixed up with them both.
The tidyverse way of doing this is equally good. This is the tidied-up way:

```{r cornseed-12 }
d %>% count(soil, class) %>% 
  pivot_wider(names_from = class, values_from = n, values_fill = 0)
```

 
Six out of eight `loam`s were correctly classified, which is
better than anything else.

Extra: we can calculate misclassification rates, first overall, which is easier:

```{r cornseed-13, error=T}
d %>%
  count(soil, class) %>%
  mutate(soil_stat = ifelse(soil == class, "correct", "wrong")) %>%
  count(soil_stat, wt = n)
```

 

```{r cornseed-14, error=T}
d %>%
  count(soil, class) %>%
  mutate(soil_stat = ifelse(soil == class, "correct", "wrong")) %>%
  count(soil_stat, wt = n) %>%
  mutate(prop = nn / sum(nn))
```



Note the use of `wt` on the second `count` to count the
number of *observations* from the first `count`, not the
number of *rows*.

This shows that 44\% of the soil types were misclassified, which
sounds awful, but is actually not so bad, considering. Bear in mind
that if you were just guessing, you'd get 75\% of them wrong, so
getting 44\% wrong is quite a bit better than that. The variables
(especially `herbicide`) are at least somewhat informative
about soil type; it's better to know them than not to.

Or do it by actual soil type:
```{r cornseed-15 }
d %>%
  count(soil, class) %>%
  group_by(soil) %>%
  mutate(soil_stat = ifelse(soil == class, "correct", "wrong")) %>%
  count(soil_stat, wt = n)
```

 

```{r cornseed-16, error=T}
d %>%
  count(soil, class) %>%
  group_by(soil) %>%
  mutate(soil_stat = ifelse(soil == class, "correct", "wrong")) %>%
  count(soil_stat, wt = n) %>%
  mutate(prop = n / sum(n)) %>%
  select(-n) %>%
  pivot_wider(names_from=soil_stat, values_from=prop)
```

 

Loam soil was the easiest to get right, and clay was easiest to get
wrong. However, these proportions were each based on only eight
observations, so it's probably wise *not* to say that loam is
*always* easiest to get right.

I didn't have you look at posterior probabilities here.^[Rest assured that I did on the final exam!] With 32 fields, this is rather a lot
to list them all, but what we can do is to look at the ones that were
misclassified (the true soil type differs from the predicted soil
type). Before that, though, we need to make a data frame with the stuff in
it that we want to look at. And before *that*, I want to round
the posterior probabilities to a small number of decimals.

Then, we can fire away with this:
```{r cornseed-17, error=T}
d %>%
  mutate(across(starts_with("posterior"), ~ round(., 3))) %>%
  mutate(row = row_number()) -> dd
dd %>% filter(soil != class)
```

 
Most of the posterior probabilities are neither especially small nor
especially large, which adds to the impression that things are really
rather uncertain. For example, field 8 could have been either loam
(0.352) or sandy (0.373). There was one field that was actually salty
but looked like a loam one (with `LD1` score around 2); this is
field 20, that needed a lot of herbicide; it was rated to have an 82\%
chance of being loam and only 1\% chance of salty.

Let's remind ourselves of why we were doing this: the MANOVA was
significant, so at least some of the fields were different on some of
the variables from some of the others. What we found by doing the
discriminant analysis was that only the first discriminant was of any
value in distinguishing the soil types by the variables we measured,
and *that* was mostly `herbicide`. So the major effect
that soil type had was on the amount of herbicide needed, with the
loam soils needing most.

I wanted to finish with one more thing, which was to look again at the
soils that were actually loam:

```{r cornseed-18 }
dd %>%
  filter(soil == "loam") %>%
  select(soil, yield, water, herbicide, class, starts_with("posterior"))
```

 
Fields 7 and 8 could have been pretty much any type of soil;
`sandy` came out with the highest posterior probability, so
that's what they were predicted (wrongly) to be. Some of the fields,
1, 3 and 5, were clearly (and correctly) loam. For 1 and 5, you can clearly
see that this is because `herbicide` was high, but field 3 is
more of a mystery. For this field, `herbicide` is *not*
high, so one or more of the other variables must be pointing towards
`loam`. 

We can obtain predicted
`LD1` scores for various combinations of "typical" values of
the response variables and see what has what effect on `LD1`:

```{r cornseed-19 }
summary(cornseed)
```

 

The problem is that the variables have different spreads. Let's do
some predictions (ie.\ calculations) of LD1 score for combinations of
quartiles of our response variables. I like quartiles because these
are "representative" values of the variables, typical of how far up
and down they go. This process is one you've seen before:

```{r cornseed-20 }
yields <- c(51, 67)
waters <- c(26, 37)
herbicides <- c(3, 6)
new <- crossing(yield = yields, water = waters, herbicide = herbicides)
pred <- predict(cornseed.2, new)
cbind(new, pred$x) %>% arrange(desc(LD1))
```

 

I arranged the predicted LD1 scores in descending order, so the most
loam-like combinations are at the top. The top two combinations look
like loam; they both have high `herbicide`, as we figured
before. But they also have high `yield`. That might go some way
towards explaining why field 3, with its non-high `herbicide`,
was confidently predicted to be `loam`:

```{r cornseed-21 }
cornseed %>% filter(field == 3)
```

 

This has a very high `yield`, and *that* is what is making
us (correctly) think it is `loam`.

I suddenly remembered that I hadn't done a biplot of this one, which I
could, since it's a discriminant analysis:

```{r cornseed-22 }
ggbiplot(cornseed.2, groups = cornseed$soil)
```

 

This shows the dominant influence of `herbicide` on LD1 score
(more herbicide is more positive), and that `water` has nothing
to say (in terms of distinguishing soil types) and `yield` has
not much to say, their arrows being short. That observation with a
non-high `herbicide` that was predicted to be \textrm{loam} had
the highest `yield` of all, so even the small influence of
`yield` on `LD1` made a big difference here.
  

$\blacksquare$



