##  A correlation matrix


 Here is a correlation matrix between five variables. This
correlation matrix was based on $n=50$ observations. Save the data
into a file.


```

1.00 0.90 -0.40 0.28 -0.05
0.90 1.00 -0.60 0.43 -0.20
-0.40 -0.60 1.00 -0.80 0.40
0.28 0.43 -0.80 1.00 -0.70
-0.05 -0.20 0.40 -0.70 1.00

```




(a) Read in the data, using `col_names=F` (why?). Check that you
have five variables with names invented by R.

Solution


I saved my data into `cov5.txt`,
`r tufte::margin_note("Not to be confused    with *covfefe*.")` delimited by single spaces, so:
```{r }
corr <- read_delim("cov5.txt", " ", col_names = F)
corr
```

   

I needed to say that I have no variable names and I want R
to provide some. As you see, it did: `X1` through
`X5`. You can also supply your own names in this fashion:

```{r }
my_names <- c("first", "second", "third", "fourth", "fifth")
corr2 <- read_delim("cov5.txt", " ", col_names = my_names)
corr2
```

 


(b) Run a principal components analysis from this correlation matrix.

Solution


Two lines, these:
```{r }
corr.mat <- as.matrix(corr)
corr.1 <- princomp(covmat = corr.mat)
```

     

Or do it in one step as

```{r }
corr.1a <- princomp(as.matrix(corr))
```

 

if you like, but I think it's less clear what's going on.


(c) <a name="part:howmany">*</a> Obtain a scree plot. Can you justify
the use of two components (later, factors), bearing in mind that we
have only five variables?

Solution


```{r sdljhljsahja}
ggscreeplot(corr.1)
```

     

There is kind of an elbow at 3, which would suggest two components/factors.
`r tufte::margin_note("There is also kind of an elbow at 4, which would suggest three factors, but that's really too many with only 5 variables. That wouldn't be much of a *reduction* in the number of variables, which is what principal components is trying to achieve.")`

You can also use the eigenvalue-bigger-than-1 thing: 

```{r }
summary(corr.1)
```

 

Only the first
*two* eigenvalues are bigger than 1, and the third is quite a bit
smaller. So this would suggest two factors also. The third eigenvalue
is in that kind of nebulous zone between between being big and being
small, and the percent of variance explained is also ambiguous: is
86\% good enough or should I go for 96\%? These questions rarely have
good answers. But an issue is that you want to summarize your
variables with a (much) smaller number of factors; with 5 variables,
having two factors rather than more than two seems to be a way of
gaining some insight.


(d) Take a look at the first two component loadings. Which variables appear
to feature in which component? Do they have a positive or negative effect?

Solution


```{r }
corr.1$loadings
```

     
This is messy. 

In the first component, the loadings are all about the
same in size, but the ones for `X3` and `X5` are
negative and the rest are positive. Thus, component 1 is contrasting
`X3` and `X5` with the others.

In the second component,  the emphasis is on `X1`, `X2`
and `X5`, all with negative loadings, and possibly `X4`,
with a positive loading.

Note that the first component is basically "size", since the
component loadings are all almost equal in absolute value. This often
happens in principal components; for example, it also happened with
the running records in class.

I hope the factor analysis, with its rotation, will straighten things
out some. 


(e) Create a "covariance list" (for the purposes of
performing a factor analysis on the correlation matrix).

Solution


This is about the most direct way:

```{r }
corr.list <- list(cov = corr.mat, n.obs = 50)
```

     
recalling that there were 50 observations. The idea is that we feed
this into `factanal` instead of the correlation matrix, so that
the factor analysis knows how many individuals there were (for testing
and such).

Note that you need the correlation matrix *as a `matrix`*,
not as a data frame. If you ran the `princomp` all in one step,
you'll have to create the correlation matrix again, for example like this:

```{r }
corr.list2 <- list(cov = as.matrix(corr), n.obs = 50)
```

 

The actual list looks like this:

```{r }
corr.list
```

 

An R list is a collection of things *not all of the same type*,
here a matrix and a number, and is a handy way of keeping a bunch of
connected things together. You use the same dollar-sign notation as
for a data frame to
access the things in a list:

```{r }
corr.list$n.obs
```

 

and logically this is because, to R, a data frame *is* a special
kind of a list, so anything that works for a list also works for a
data frame, plus some extra things besides.
`r tufte::margin_note("In computer  science terms, a data frame is said to **inherit** from a list: it is a list plus extra stuff.")` 

The same idea applies to extracting things from the output of a
regression (with `lm`) or something like a `t.test`: the
output from those is also a list. But for those, I like `tidy`
from `broom` better.


(f) Carry out a factor analysis with two factors. We'll
investigate the bits of it in a 
moment.  

Solution


```{r }
corr.2 <- factanal(factors = 2, covmat = corr.list)
```

     


(g) <a name="part:load">*</a> Look at the factor loadings. Describe how the factors are
related to the original variables. Is the interpretation clearer
than for the principal components analysis?

Solution


```{r }
corr.2$loadings
```

     

Oh yes, this is a lot clearer. Factor 1 is variables 3 and 5
contrasted with variable 4; factor 2 is variables 1 and 2. No
hand-waving required.

Perhaps now is a good time to look back at the correlation matrix and
see why the factor analysis came out this way:

```{r }
corr
```

 

Variable `X1` is highly correlated with `X2` but not
really any of the others. Likewise variables `X3, X4, X5` are
more or less highly correlated among themselves but not with the
others (`X2` and `X3` being an exception, but the big
picture is as I described). So variables that appear in the same
factor should be highly correlated with each other and not with
variables that are in different factors. But it took the factor
analysis to really show this up.


(h) Look at the uniquenesses. Are there any that are unusually
high? Does that surprise you, given your answer to
(<a href="#part:load">here</a>)? (You will probably have to make a judgement call here.)

Solution


```{r }
corr.2$uniquenesses
```

     

The ones for `X3` and `X5` are higher than the rest;
this is because their loadings on factor 1 are lower than the
rest. Since those loadings are still high, I wouldn't be worried about
the uniquenesses.

The point here is that an (excessively) high uniqueness indicates a
variable that doesn't appear in *any* factor. The easy link to
make is "all the variables appear in a factor, so there shouldn't be any very high uniquenesses". If, say, `X3` doesn't have a high
loading on any factor, `X3` would have a high uniqueness (like
0.9, and none of these values approach that).



