##  Mating songs in crickets


 Male tree crickets produce "mating songs" by rubbing their
wings together to produce a chirping sound. It is hypothesized that
female tree crickets identify males of the correct species by how fast
(in chirps per second) the male's mating song is. This is called the
"pulse rate".  Some data for two species of crickets are in
[link](http://ritsokiguess.site/datafiles/crickets.txt). The
columns, which are unlabelled, are temperature and pulse rate
(respectively) for *Oecanthus exclamationis* (first two
columns) and *Oecanthus niveus* (third and fourth columns). The
columns are separated by tabs. There are some missing values in the
first two columns because fewer *exclamationis* crickets than
*niveus* crickets were measured.
The research question is whether males
of the different species have different average pulse rates. It is
also of interest to see whether temperature has an effect, and if
so, what.
Before we get to that, however, we have some data organization to do.


(a) Read in the data, allowing for the fact that you have no
column names. You'll see that the
columns have names `X1` through `X4`. This is
OK.


Solution


Tab-separated, so `read_tsv`; no column names, so `col_names=F`:
```{r }
my_url <- "http://ritsokiguess.site/datafiles/crickets.txt"
crickets <- read_tsv(my_url, col_names = F)
crickets
```

 

As promised.

If you didn't catch the tab-separated part, this probably happened to you:

```{r }
d <- read_delim(my_url, " ", col_names = F)
```

 

This doesn't look good:

```{r }
problems(d)
```

 

The "expected columns" being 1 should bother you, since we know
there are supposed to be 4 columns. At this point, we take a look at
what got read in:

```{r }
d
```



and there you see the `t` or "tab" characters separating the
values, instead of spaces. (This is what I tried first, and once I
looked at this, I realized that `read_tsv` was what I needed.)

(b) Tidy these untidy data, going as directly as you can to something tidy. (Some later parts show you how it used to be done.) Begin by: (i) adding a column of row numbers, (ii) `rename`-ing the columns to species name, an underscore, and the variable contents (keeping `pulserate` as one word), and then use `pivot_longer`. Note that the column names encode *two* things.



Solution



Take this one piece of the pipeline at a time: that is, first check that you got the renaming right and looking at what you have, before proceeding to the `pivot_longer`. The syntax of `rename` is new name equals old name, and I like to split this over several lines to make it easier to read:

```{r}
crickets %>% 
  mutate(row=row_number()) %>% 
  rename(
    exclamationis_temperature = X1,
    exclamationis_pulserate = X2,
    niveus_temperature = X3,
    niveus_pulserate = X4
  ) 
```

The first part of each column name is the species and the second part is what was measured each time, separated by an underscore. To handle that in `pivot_longer`, you give *two* names of new columns to create (in `names_to`), and say what they're separated by:

```{r}
crickets %>% 
  mutate(row=row_number()) %>% 
  rename(
    exclamationis_temperature = X1,
    exclamationis_pulserate = X2,
    niveus_temperature = X3,
    niveus_pulserate = X4
  ) %>% 
  pivot_longer(-row, names_to=c("species", "measurement"), names_sep="_", values_to = "obs")
```

This is tidy now, but we went a step too far: that column `measurement` should be *two* columns, called `temperature` and `pulserate`, which means it should be made wider:

```{r}
crickets %>% 
  mutate(row=row_number()) %>% 
  rename(
    exclamationis_temperature = X1,
    exclamationis_pulserate = X2,
    niveus_temperature = X3,
    niveus_pulserate = X4
  ) %>% 
  pivot_longer(-row, names_to=c("species", "measurement"), names_sep="_", values_to = "obs") %>% 
  pivot_wider(names_from=measurement, values_from=obs)

```


The row numbers are cricket-within-species, which isn't very meaningful, but we needed something for the `pivot_wider` to key on, to recognize what needed to go in which row. The way it works is it uses anything not mentioned in `names_from` or `values_from` as a "key": each unique combination belongs in a row. Here that would be the combination of row and species, which is a good key because each species appears once with each row number. 



(c) If you found (b) a bit much to take in, the rest of the way we take a rather more leisurely approach towards the tidying. 

These data are rather far from being tidy. There need to be
three variables, temperature, pulse rate and species, and there
are $14+17=31$ observations altogether. This one is tricky in that
there are temperature and pulse rate for each of two levels of a
factor, so I'll suggest combining the temperature and chirp rate
together into one thing for each species, then pivoting them longer ("combining"),
then pivoting them wider again ("splitting"). Create new columns, named for  each species,
that contain the temperature and pulse rate for that species in
that order, `unite`d together.
For the rest of this question, start from the data frame you read
in, and build a pipe, one or two steps at a time, to save creating
a lot of temporary data frames.


Solution


Breathe, and then begin. `unite` creates new columns by
joining together old ones:
`r tufte::margin_note("As *str-c* or *paste* do, actually, but the advantage of *unite* is that it gets rid of the other columns, which you probably no longer need.")`
```{r }
crickets %>%
  unite(exclamationis, X1:X2) %>%
  unite(niveus, X3:X4)
```

 

Note that the original columns `X1:X4` are *gone*, which
is fine, because the information we needed from them is contained in
the two new columns. `unite` by default uses an underscore to
separate the joined-together values, which is generally safe since you
won't often find those in data.

Digression: `unite`-ing with a space could cause problems if
the data values have spaces in them already. Consider this list of names:

```{r }
names <- c("Cameron McDonald", "Durwin Yang", "Ole Gunnar Solskjaer", "Mahmudullah")
```

 

Two very former students of mine, a Norwegian soccer player, and a
Bangladeshi cricketer. Only one of these has played for Manchester United:

```{r }
manu <- c(F, F, T, F)
```

 

and let's make a data frame:

```{r }
d <- tibble(name = names, manu = manu)
d
```



Now, what happens if we `unite` those columns, separating them
by a space?

```{r }
d %>% unite(joined, name:manu, sep = " ")
```

 

If we then try to separate them again, what happens?

```{r }
d %>%
  unite(joined, name:manu, sep = " ") %>%
  separate(joined, c("one", "two"), " ")
```

 

Things have gotten lost: most of the original values of `manu`
and some of the names. If we use a different separator character,
either choosing one deliberately or going with the default underscore,
everything works swimmingly:

```{r }
d %>%
  unite(joined, name:manu, sep = ":") %>%
  separate(joined, c("one", "two"), ":")
```

 

and we are back to where we started.

If you run just the `unite` line (move the pipe symbol to the
next line so that the `unite` line is complete as it stands),
you'll see what happened.
      


(d) The two columns `exclamationis` and `niveus`
that you just created are both temperature-pulse rate combos, but
for different species. Collect them together into one
column, labelled by species. (This is a straight `tidyr` `pivot_longer`, even though the columns contain something odd-looking.)


Solution


Thus, this, naming the new column `temp_pulse` since it
contains both of those things. Add to the end of the pipe you
started building in the previous part:

```{r }
crickets %>%
  unite(exclamationis, X1:X2) %>%
  unite(niveus, X3:X4) %>%
  pivot_longer(exclamationis:niveus, names_to = "species", values_to = "temp_pulse")
```

 

Yep. You'll see both species of
crickets, and you'll see some missing values at the bottom, labelled,
at the moment, `NA_NA`. 

This is going to get rather long, but don't fret: we debugged the two
`unite` lines before, so if you get any errors, they must
have come from the `pivot_longer`. So that would be the place to check.



(e) Now split up the temperature-pulse combos at the underscore, into
two separate columns. This is `separate`. When specifying
what to separate by, you can use a number ("split after this many characters") or a piece of text, in quotes ("when you see this text, split at it"). 


Solution


The text to split by is an underscore (in quotes), since
`unite` by default puts an underscore in between the
values it pastes together. Glue the `separate` onto the
end. We are creating two new variables `temperature` and
`pulse_rate`:
```{r }
crickets %>%
  unite(exclamationis, X1:X2) %>%
  unite(niveus, X3:X4) %>%
  pivot_longer(exclamationis:niveus, names_to = "species", values_to = "temp_pulse") %>% 
  separate(temp_pulse, c("temperature", "pulse_rate"), "_")
```

 

You'll note that `unite` and `separate` are opposites ("inverses") of each other, but we haven't just done something and then undone it, because we have a `pivot_longer` in between; in fact, arranging it this way has done precisely the tidying we wanted.
      


(f) Almost there.  Temperature and pulse rate are still text
(because `unite` turned them into text), but they should be
numbers. Create new variables that are numerical versions of
temperature and pulse rate (using `as.numeric`). Check that
you have no extraneous variables (and, if necessary, get rid of
the ones you don't want). (Species is also text and really ought
to be a factor, but having it as text doesn't seem to cause any
problems.)
You can, if you like, use `parse_number` instead of
`as.numeric`. They should both work. The distinction I
prefer to make is that `parse_number` is good for text
with a number in it (that we want to pull the number out of),
while `as.numeric` is for turning something that looks like
a number but isn't one into a genuine number.
`r tufte::margin_note("You could      just as well make the point that the text 20.8 contains the      number 20.8 and nothing else, so that parsing it as text in search of a number      will pull out 20.8 as a number. If that logic works for you, go      with it.")`


Solution


`mutate`-ing into a column that already exists overwrites
the variable that's already there (which saves us some effort
here). 
```{r }
crickets.1 <- crickets %>%
  unite(exclamationis, X1:X2) %>%
  unite(niveus, X3:X4) %>%
  pivot_longer(exclamationis:niveus, names_to = "species", values_to = "temp_pulse") %>% 
  separate(temp_pulse, c("temperature", "pulse_rate"), "_") %>%
  mutate(temperature = as.numeric(temperature)) %>%
  mutate(pulse_rate = as.numeric(pulse_rate))
crickets.1
```

 

I saved the data frame this time, since this is the one we will use
for our analysis.

The warning message tells us that we got genuine missing-value NAs
back, which is probably what we want. Specifically, they got turned
from missing *text* to missing *numbers*!
`r tufte::margin_note("You might think that  missing is just missing, but R distinguishes between types of missing.")`
The R word
"coercion" means values being changed from one type of thing to
another type of thing.  (We'll ignore the missings and see if they
cause us any trouble. The same warning messages will show up on graphs
later.)  So I have 34 rows (including three rows of missings) instead
of the 31 rows I would have liked. Otherwise, success.

There is (inevitably) another way to do this. We are doing the
`as.numeric` twice, exactly the same on two different columns,
and when you are doing the same thing on a number of columns, here a
`mutate` with the same function, you have the option of using
`across`. This is the same idea 
that we used way
back to compute numerical summaries of a bunch of columns:

```{r }
crickets %>%
  unite(exclamationis, X1:X2) %>%
  unite(niveus, X3:X4) %>%
  pivot_longer(exclamationis:niveus, names_to = "species", values_to = "temp_pulse") %>% 
  separate(temp_pulse, c("temperature", "pulse_rate"), "_") %>%
  mutate(across(c(temperature, pulse_rate), ~as.numeric(.)))
```

 

Can't I just say that these are columns 2 and 3?

```{r }
crickets %>%
  unite(exclamationis, X1:X2) %>%
  unite(niveus, X3:X4) %>%
  pivot_longer(exclamationis:niveus, names_to = "species", values_to = "temp_pulse") %>% 
  separate(temp_pulse, c("temperature", "pulse_rate"), "_") %>%
  mutate(across(2:3, ~as.numeric(.)))
```

 

Yes. Equally good. What goes into the `across`
is the same as can go into a `select`: column numbers, names,
or any of those "select helpers" like `starts_with`.

You might think of using `across` here on the quantitative columns, but remember the reason for doing this: all the columns are text, before you convert temperature and pulse rate to numbers, and so there's no way to pick out just the two columns you want that way. 

Check that the temperature and pulse rate columns are now labelled
`dbl`, which means they actually *are* decimal numbers
(and don't just look like decimal numbers).

Either way, using `unite` and then `separate` means that
all the columns we created we want to keep (or, all the ones we would
have wanted to get rid of have already been gotten rid of).

Now we could actually do some statistics. That we do elsewhere.
      



