##  Crickets revisited


 This is a reorganization of the crickets problem that you
may have seen before (minus the data tidying).

Male tree crickets produce "mating songs" by rubbing their wings
together to produce a chirping sound. It is hypothesized that female
tree crickets identify males of the correct species by how fast (in
chirps per second) the male's mating song is. This is called the
"pulse rate".  Some data for two species of crickets are in
[link](http://www.utsc.utoronto.ca/~butler/c32/crickets2.csv) as a CSV
file. The columns are species (text), temperature, and pulse
rate (numbers). This is the tidied version of the data set that the
previous version of this question had you create.
The research question is whether males
of the different species have different average pulse rates. It is
also of interest to see whether temperature has an effect, and if
so, what.


(a) Read the data into R and display what you have.


Solution


Nothing terribly surprising here:

```{r }
my_url="http://www.utsc.utoronto.ca/~butler/c32/crickets2.csv"
crickets=read_csv(my_url)
crickets
```

 

31 crickets, which is what I remember. What species are there?

```{r }
crickets %>% count(species)
```

 

That looks good. We proceed.



(b) Do a two-sample $t$-test to see whether the mean pulse rates
differ between species. What do you conclude?


Solution


Drag your mind way back to this:
```{r }
t.test(pulse_rate~species,data=crickets) 
```

 

There is strong evidence of a difference in means (a P-value around
0.00001), and the confidence interval says that the mean chirp rate is
higher for *exclamationis*. That is, not just for the crickets
that were observed here, but for *all* crickets of these two
species. 
      


(c) Can you do that two-sample $t$-test as a regression?


Solution


Hang onto the "pulse rate depends on species" idea and try
that in `lm`:
```{r }
pulse.0=lm(pulse_rate~species,data=crickets)
summary(pulse.0)
```

         

I had to use "model 0" for this since I already have a
`pulse.1` below and I didn't want to go down and renumber
everything. 

Look along the `speciesniveus` line. Ignoring the fact that it
is negative, the $t$-statistic is almost the same as before (5.17 vs.\
5.22) and so is the P-value ($1.4 \times 10^{-5}$ vs.\ $1.6 \times
10^{-5}$). 

Why aren't they exactly the same? Regression is assuming equal
variances everywhere (that is, within the two `species`), and
before, we did the Welch-Satterthwaite test that does not assume equal
variances. What if we do the pooled $t$-test instead?

```{r }
t.test(pulse_rate~species,data=crickets,var.equal=T) 
```

 

Now the regression and the $t$-test *do* give exactly the same
answers. We'll think about that equal-spreads assumption again later.



(d) The analysis in the last part did not use temperature,
however. Is it possible that temperature also has an effect? To
assess this, draw a scatterplot of pulse rate against temperature,
with the points distinguished, somehow, by the species they are
from.
`r tufte::margin_note("This was the actual reason I thought of this      question originally:    I wanted you to do this.")`


Solution


One of the wonderful things about `ggplot` is that doing
the obvious thing works:
```{r }
ggplot(crickets,aes(x=temperature,y=pulse_rate,colour=species))+
geom_point()
```

       
    


(e) What does the plot tell you that the $t$-test doesn't? How
would you describe differences in pulse rates between species now?


Solution


The plot tells you that (for both species) as temperature goes
up, pulse rate goes up as well. *Allowing for that*, the
difference in pulse rates between the two species is even
clearer than it was before. To see an example, pick a
temperature, and note that the mean pulse rate at that
temperature seems to be at least 10 higher for
*exclamationis*, with a high degree of consistency.
The $t$-test mixed up all the pulse rates at all the different
temperatures. Even though the conclusion was clear enough, it
could be clearer if we incorporated temperature into the analysis.
There was also a potential source of unfairness in that the
*exclamationis* crickets tended to be observed at higher
temperatures than *niveus* crickets; since pulse rates
increase with temperature, the apparent difference in pulse
rates between the species might have been explainable by one
species being observed mainly in higher temperatures. This was
*utterly invisible* to us when we did the $t$-test, but it
shows the importance of accounting for all the relevant
variables when you do your analysis.
`r tufte::margin_note("And it shows the        value of looking at relevant plots.")` If the species had been
observed at opposite temperatures, we might have
concluded
`r tufte::margin_note("Mistakenly.")` 
that *niveus* have the
higher pulse rates on average. I come back to this later when I
discuss the confidence interval for species difference that
comes out of the regression model with temperature.
      


(f) Fit a regression predicting pulse rate from species and
temperature. Compare the P-value for species in this regression to
the one from the $t$-test. What does that tell you?


Solution


This is actually a so-called "analysis of covariance model",
which properly belongs in D29, but it's really just a regression:
```{r }
pulse.1=lm(pulse_rate~species+temperature,data=crickets)
summary(pulse.1)
```

 

The P-value for species is now $6.27\times 10^{-14}$ or
0.00000000000006, which is even less than the P-value of 0.00001 that
came out of the $t$-test. That is to say, when you know temperature,
you can be even more sure of your conclusion that there is a
difference between the species.

The R-squared for this regression is almost 99\%, which says that if
you know both temperature and species, you can predict the pulse rate
almost exactly.

In the regression output, the slope for species is about $-10$. It is
labelled `speciesniveus`. Since species is categorical,
`lm` uses the first category, *exclamationis*, as the
baseline and expresses each other species relative to that. Since the
slope is about $-10$, it says that at any given temperature, the mean
pulse rate for *niveus* is about 10 less than for
*exclamationis*. This is pretty much what the scatterplot told
us.

We can go a little further here:

```{r }
confint(pulse.1)
```

 

The second line says that the pulse rate for *niveus* is
between about 8.5 and 11.5 less than for *exclamationis*, at
any given temperature (comparing the two species at the same
temperature as each other, but that temperature could be
anything). This is a lot shorter than the CI that came out of the
$t$-test, that went from 14 to 32. This is because we are now
accounting for temperature, which also makes a difference. (In the
$t$-test, the temperatures were all mixed up). What we also see is
that the $t$-interval is shifted up compared to the one from the
regression. This is because the $t$-interval conflates
`r tufte::margin_note("Mixes up.")` 
two things: the *exclamationis* crickets do have a
higher pulse rate, but they were also observed at higher temperatures,
which makes it look as if their pulse rates are more
higher
`r tufte::margin_note("This is actually grammatically correct.")` than they
really are, when you account for temperature.

This particular model constrains the slope with temperature to be the
same for both species (just the intercepts differ). If you want to
allow the slopes to differ between species, you add an interaction
between temperature and species:

```{r }
pulse.2=lm(pulse_rate~species*temperature,data=crickets)
summary(pulse.2)
```

 

To see whether adding the interaction term added anything to the
prediction,
`r tufte::margin_note("Though it's hard to imagine being able to improve on an R-squared of 99%.")` 
compare the model with and without using `anova`:

```{r }
anova(pulse.1,pulse.2)  
```

 

There's no significant improvement by adding the interaction, so
there's no evidence that having different slopes for each species is
necessary. This is the same interpretation as any `anova` for
comparing two regressions: the two models are not significantly
different in fit, so go with the simpler one, that is, the one without
the interaction.

Note that `anova` gave the same P-value as did the
$t$-test for the slope coefficient for the interaction in
`summary`, 0.254 in both cases. This is because there were only
two species and therefore only one slope coefficient was required to
distinguish them. If there had been three species, we would have had
to look at the `anova` output to hunt for a difference among
species, since there would have been two slope coefficients, each with
its own P-value.
`r tufte::margin_note("This wouldn't have told us about the overall  effect of species.")` 

If you haven't seen interactions before, don't worry about this. The
idea behind it is that we are testing whether we needed lines with
different slopes and we concluded that we don't. Don't worry so much
about the mechanism behind `pulse.2`; just worry about how it
somehow provides a way of modelling two different slopes, one for each
species, which we can then test to see whether it helps.

The upshot is that we do not need different slopes; the model
`pulse.1` with the same slope for each species describes what
is going on.

`ggplot` makes it almost laughably easy to add regression lines
for each species to our plot, thus:

```{r }
ggplot(crickets,aes(x=temperature,y=pulse_rate,colour=species))+
geom_point()+geom_smooth(method="lm",se=F)
```

 

The lines are almost exactly parallel, so having the same slope for
each species makes perfect sense.
      


(g) Make suitable residual plots for the regression
`pulse.1`. 


Solution


First, the plot of residuals against fitted values (after all,
it *is* a regression):
```{r }
ggplot(pulse.1,aes(x=.fitted,y=.resid))+geom_point()
```

         

This looks nice and random.

Now, we plot the residuals against the explanatory variables. There
are two, temperature and species, but the latter is categorical. We'll
have some extra issues around species, but before we get to that, we
have to remember that the data and the output from the regression are
in different places when we plot them. There are different ways to get
around that. Perhaps the simplest is to use `pulse.1` as our
"default" data frame and then get `temperature` from the
right place:

```{r }
ggplot(pulse.1,aes(x=crickets$temperature,y=.resid))+geom_point()
```

 

I don't see anything untoward there.

Species. We want to compare the residuals for the two species, which
is categorical. Since the residuals are quantitative, this suggests a
boxplot. Remembering to get species from the right place again, that
goes like this:

```{r }
ggplot(pulse.1,aes(x=crickets$species,y=.resid))+geom_boxplot()
```

 

For the residuals, the median should be zero within each group, and
the two groups should be approximately normal with mean 0 and about
the same spread. Same spread looks OK, since the boxes are almost
exactly the same height, but the normality is not quite there, since
both distributions are a little bit skewed to the right. That would
also explain why the median residual in each group is a little bit
less than zero, because the mathematics requires the overall
*mean* residual to be zero, and the right-skewness would make the
mean higher than the median.

Is that non-normality really problematic? Well, I could look at the
normal quantile plot of all the residuals together:

```{r }
ggplot(pulse.1,aes(sample=.resid))+stat_qq()+stat_qq_line()
```

 

There's a little weirdness at the top, and a tiny indication of a
curve (that would suggest a little right-skewedness), but not really
much to worry about. If that third-highest residual were a bit
lower (say, 3 rather than 3.5) and maybe if the lowest residual was a
bit lower, I don't think we'd have anything to complain about at all.

So, I'm not worried.




