## Coefficient of VariationF

 The coefficient of variation of a vector `x` is defined as the standard deviation of `x` divided by the mean of `x`. 



(a) Write a function called `cv` that calculates the coefficient of variation of its input and returns the result. You should use base R's functions that reliably compute the pieces that you need.

Solution


I like to make a function skeleton first to be sure I remember all the pieces. You can do this by making a code chunk, typing `fun`, waiting a moment, and selecting the "snippet" that is offered to you. That gives you this, with your cursor on `name`:

```{r}
name <- function(variables) {

}
```

Give your function a name (I asked you to use `cv`), hit tab, enter a name like x for the input, hit tab again, and then fill in the body of the function, to end up with something like this:

```{r}
cv <- function(x) {
mean_x <- mean(x)
sd_x <- sd(x)
sd_x/mean_x
}
```

I think this is best, for reasons discussed below.

Some observations:

- in the last line of the function, you calculate something, and that something is the thing that gets returned. You *do not* need to save it in a variable, because then you have to return that variable by having its name alone on the last line.
- R has a `return()` function, but it is *bad style* to use it to return a value at the end of a function. The time to use it is  when you test something early in the function and want to return early with a value like zero or missing without doing any actual calculation. Looking ahead a bit, you might want to return "missing" if the mean is zero *before* dividing by it, which you could do like this, this being *good* style:

```{r}
cv2 <- function(x) {
mean_x <- mean(x)
if (mean_x == 0) return(NA)
sd_x <- sd(x)
sd_x/mean_x
}
```

- use the built-in `mean` and `sd` rather than trying to roll your own, because they have been tested by many users over a long time and they work. That's what I meant by "pieces" in the question.

- you might be tempted to do something like this:

```{r}
cv3 <- function(x) {
sd(x)/mean(x)
}
```

This will work, but it is not a good habit to get into, and thus not best as an answer to this question. This one line of code says three things: "work out the SD of `x`", "work out the mean of `x`", and "divide them by each other". It is much clearer, for anyone else reading the code (including yourself in six months when you have forgotten what you were thinking) to have the three things one per line, so that anyone reading the code sees that each line does one thing and what that one thing is. There is a reason why production code and [code golf](https://en.wikipedia.org/wiki/Code_golf) are two very different things. Code that is easy to read is also easy to maintain, either by you or others.

- using something other than `mean` and `sd` as the names of your intermediate results is a good idea because these names are already used by R (as the names of the functions that compute the mean and SD). Redefining names that R uses can make other code behave unpredictably and cause hard-to-find bugs.^[This is something I've done in the past and been bitten by, so I am trying to get myself not to do it any more.]


$\blacksquare$


(b) Use your function to find the coefficient of variation of the set of integers 1 through 5.

Solution


This can be as simple as

```{r}
cv(1:5)
```

or, equally good, define this into a vector first:

```{r}
y <- 1:5
cv(y)
```

or, displaying a little less understanding, type the five numbers into the vector first (or directly as input to the function):

```{r}
y <- c(1, 2, 3, 4, 5)
cv(y)
cv(c(1, 2, 3, 4, 5))
```



$\blacksquare$



(c) Define a vector as follows:

```{r}
v <- c(-2.8, -1.8, -0.8, 1.2, 4.2)
```

What is its coefficient of variation, according to your function? Does this make sense? Why did this happen? Explain briefly.

Solution


Try it and see:

```{r}
cv(v)
```

A very large number, much bigger than any of the data values; with these human-sized numbers, we'd expect a human-sized coefficient of variation as well. 

What actually happened was that the mean of `v` is this:

```{r}
mean(v)
```

Zero, or close enough, so that in calculating the coefficient of variation, we divided by something that was (almost) zero and got a result that was (almost) infinite.^[In computing with decimal numbers, things are almost never exactly zero or exactly infinite; they are very small or very  big. The mean here, which you would calculate to be zero, is less than the so-called machine epsilon, which is about 10 to the minus 16 in R (R works in double precision). A mean that small is, to the computer, indistinguishable from zero. It came out that way because the last value is added to a total that is by that point negative, and so you have a loss of accuracy because of subtracting nearly equal quantities. I learned all about this stuff in my first real computer science course, which I think was a numerical math course, some absurd number of years ago. It looks as if this gets taught in CSCC37 these days.]

The possibility of getting a zero mean is why most people only calculate a coefficient of variation if all of the numbers are positive, which brings us to the next part:


$\blacksquare$


(d) Most people only calculate a coefficient of variation if there are no negative numbers. Rewrite your function so that it gives an error if there are any negative numbers in the input, and test it with the vector `v` above. Hint: you might need to add `error=TRUE` to your chunk header to allow your document to preview/knit (inside the curly brackets at the top of the chunk, after a comma).

Solution


This is a case for `stopifnot`, or of `if` coupled with `stop`. Check this up front, as the first thing you do before you calculate anything else. As to what to check, there are several possibilities:

- stop if any of the numbers are negative
- continue if all of the numbers are positive
- stop if the *smallest* number is negative (the smallest number is negative if and only if not all the numbers are positive)

R has functions `any` and `all` that do what you'd expect:

```{r}
w <- 1:5
w
any(w>3.5)
all(w<4.5)
```


Are there any numbers greater than 3.5 (yes, 4 and 5); are all the numbers less than 4.5 (no, 5 isn't).

Cite your sources for these if you use either of them, since this is the first place in the course that I'm mentioning either of them.

Remember that if you use `stopifnot`, the condition that goes in there is what has to be true if the function is to run; if you use `if` and `stop`, the condition is what will *stop* the function running. With that in mind, I would code my three possibilities above this way. First off, here's the original:

```{r}
cv <- function(x) {
mean_x <- mean(x)
sd_x <- sd(x)
sd_x/mean_x
}
```

then, stop if any of the numbers are negative:

```{r, error=TRUE}
cv <- function(x) {
if (any(x<0)) stop("A value is negative")
mean_x <- mean(x)
sd_x <- sd(x)
sd_x/mean_x
}
cv(v)
```

continue if all the numbers are positive

```{r, error=TRUE}
cv <- function(x) {
stopifnot(all(x>0))
mean_x <- mean(x)
sd_x <- sd(x)
sd_x/mean_x
}
cv(v)
```

stop if the smallest value is negative

```{r, error=TRUE}
cv <- function(x) {
if (min(x)<0) stop("Smallest value is negative")
mean_x <- mean(x)
sd_x <- sd(x)
sd_x/mean_x
}
cv(v)
```

There are (at least) three other possibilities: you can negate the logical condition and interchange `if/stop` and `stopifnot`, thus (at the expense of some clarity of reading):

continue if it is not true that any of the numbers are negative

```{r, error=TRUE}
cv <- function(x) {
stopifnot(!any(x<0))
mean_x <- mean(x)
sd_x <- sd(x)
sd_x/mean_x
}
cv(v)
```

(you might be thinking of De Morgan's laws here)

stop if it is not true that all the numbers are positive

```{r, error=TRUE}
cv <- function(x) {
if (!all(x>0)) stop("Not all values are positive")
mean_x <- mean(x)
sd_x <- sd(x)
sd_x/mean_x
}
cv(v)
```

continue if the smallest value is not negative

```{r, error=TRUE}
cv <- function(x) {
stopifnot(min(x)>=0)
mean_x <- mean(x)
sd_x <- sd(x)
sd_x/mean_x
}
cv(v)
```
or another way to do the last one, a more direct negation of the condition, which at my guess needs some extra brackets:

```{r, error=TRUE}
cv <- function(x) {
stopifnot(!(min(x)<0))
mean_x <- mean(x)
sd_x <- sd(x)
sd_x/mean_x
}
cv(v)
```

This one is hard to parse: what does that last message mean? I would take a negative off each side and read it as "min of x is negative is TRUE", but that takes extra effort. 

I said that last one needed some extra brackets. This is, I thought, to get the order of operations right (operator precedence); it turns out not to matter because "not" has lower precedence than most other things, so that these do actually work (the "not" is evaluated *after* the less-than and the other things, so last of all here, even though it appears to be "glued" to the `min`): 

```{r}
!min(v)<0
!min(1:5)<0
```

See [this](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Syntax.html) for details. See especially the second set of examples, the ones beginning with "Special operators", and see especially-especially the comment at the bottom of these examples! That is to say, *you* should put in the extra brackets unless you also make the case that they are not needed, because anyone reading your code is guaranteed to be confused by it when they read it (including you in six months, because you *will not* remember the operator priority of "not").

My take is that one of the first three of the seven possibilities for coding `stopifnot` or `if` with `stop` is the best, since these more obviously encode the condition for continuing or stopping as appropriate. There are two things here: one is that you have to get the code right, but the second is that you have to get the code *clear*, so that it is obvious to anyone reading it that it does the right thing (once again, this includes you in six months). On that score, the first three alternatives are a direct expression of what you want to achieve, and the last four make it look as if you found a way of coding it that worked and stopped there, without thinking about whether there were any other, clearer or more expressive, possibilities.


$\blacksquare$




