##  Wolves and caribou


 In Denali National Park, Alaska, the size of the wolf
population depends on the size of the caribou population (since
wolves hunt and kill caribou). This is a large national park, so
caribou are found in very large herds, so big, in fact, that the
well-being of the entire herd is not threatened by wolf
attacks.
`r tufte::margin_note("In fact, it is believed that wolves help keep caribou    herds strong by preventing over-population: that is, the weakest    caribou are the ones taken by wolves.")` 
Can the size of the caribou population
be used to predict the size of the wolf population?
The data can be found at
[link](http://www.utsc.utoronto.ca/~butler/c32/caribou.txt). The
columns 
are: the date of the survey,
`r tufte::margin_note("The survey is always taken in    the fall, but the date varies.")` 
the name of the park employee in charge  of the survey, the caribou population (in hundreds) and the wolf  population (actual count).
`r tufte::margin_note("Counting animals in a region,    especially rare, hard-to-find animals, is a whole science in    itself. These numbers are probably estimates (with some    uncertainty).")` 


(a) Take a look at the data file. How would you describe its
format? Read it into R, and check that you got something sensible.


Solution


This looks at first sight as if it's separated by spaces, but
most of the data values are separated by *more than one*
space. If you look further, you'll see that the values are
*lined up in columns*, with the variable names aligned at
the top. This is exactly the kind of thing that
`read_table` will read. We start with the usual
`library(tidyverse)`:
```{r }
library(tidyverse)
my_url <- "http://www.utsc.utoronto.ca/~butler/c32/caribou.txt"
denali <- read_table(my_url)
denali
```

     

That worked: four columns with the right names, and the counts of
caribou and wolf are numbers. There are only seven years of surveys;
in real-life data you would have more. But the point here is working
with dates.

The only (small) weirdness is that the
dates are text rather than having been converted into dates. This is
because they are not year-month-day, which is the only format that
gets automatically converted into dates when read in. (You could use
`mdy` from `lubridate` to make them dates.)

Extra: you might have wondered how the names survived, even though
they have spaces in them, sometimes more than one. Here's how the file looks:


```

date       name             caribou wolf
09/01/1995 David S.         30       66
09/24/1996 Youngjin K.      34       79
10/03/1997 Srinivasan M.    27       70
09/15/1998 Lee Anne J.      25       60
09/08/1999 Stephanie T.     17       48
09/03/2000 Angus Mc D.      23       55
10/06/2001 David S.         20       60

```


What `read_table` looks for is columns that contain spaces
*all the way down*, and separates the values there. For example,
between the year of`date` and the first name in `name`
there is a space all the way down. After the names and before the
caribou counts there are several spaces, and there is one space
between the words `caribou` and `wolf` in the header
line that goes all the way down. Thus four columns, `date`,
`name`, `caribou` and `wolf`. This means that the
spaces within the names don't cause any problems at all, since the
spaces aren't in the same place in *every* line.
`r tufte::margin_note("The only way this would fail is if *every* first name had the same number of letters in it; then the space between first name and initial of last name *would* be in the same place in every line.")`



(b) Create a new data frame where the column labelled
`date` is now a genuine R `Date`, using something
from `lubridate`.

Solution


What you do is to look at the format of the dates as they are
now. They appear to be month-day-year, American
style.
`r tufte::margin_note("Not a surprise since Denali National Park is in  Alaska.")`  
Thus the function needed is `mdy`. It doesn't matter
whether the months are names or numbers:
```{r }
denali %>% mutate(date = mdy(date)) -> denali
denali
```

     

I lived on the edge and overwrote both my column and the whole data
frame.
`r tufte::margin_note("It's actually not *really* living on the edge,  because if it doesn't work, you go back and read the data in from  the file again.")`

The dates are displayed in ISO format, year-month-day. You see at the
top of the column that they now really *are* dates, not just
pieces of text that look like dates. 


(c) Create new columns containing the days of
the week and the month names for the dates.


Solution


This involves digging in the `lubridate` help to find out how to extract things from a date. It turns out that `wday` extracts the day of the week from a date, by default as a number, and `month` gets the month, also by default as a number:
```{r }
denali %>% mutate(mon = month(date), wd = wday(date))
```

     

This is not what we wanted, though; we wanted the names of the months
and of the days. To fix that, add `label=T` to both functions:

```{r }
denali %>% mutate(mon = month(date, label = T), wd = wday(date, label = T))
```

 
and that cracks it.

No need to save this data frame anywhere, since we're not using any of
this later.

Extra: the `ord` means "ordered factor", which makes sense
since these are categorical variables with a natural order. This means
that you could do something like counting the number of surveys in
each month like this:

```{r }
denali %>%
  mutate(mon = month(date, label = T), wd = wday(date, label = T)) %>%
  count(mon)
```

 
      


(d) Enough playing around with dates. Make a scatterplot of
caribou population (explanatory) against wolf population
(response). Do you see any relationship?


Solution


Nothing terribly surprising here:
```{r }
ggplot(denali, aes(x = caribou, y = wolf)) + geom_point()
```

       
If you like, add a smooth trend to it:
`r tufte::margin_note("This wiggles more than  I would like, with such a small number of observations. Try putting  someting like *span=2* in the smooth to make it less wiggly.")`

```{r }
ggplot(denali, aes(x = caribou, y = wolf)) + geom_point() + geom_smooth(se = F)
```

       

This is an upward trend: when one population is large, the other one
is large too. This is typical for predator-prey relationships: when
there is more to eat (more caribou) the wolf population goes up, and
when less, it goes down. 
      


(e) On your plot from the previous part, label each point with
the year it belongs to. You can do this in two steps: first make a
new column containing just the years, and then use it as labels
for the points on the plot.

Solution


I'm going to use `geom_text_repel` for the labels from package
`ggrepel`. The year values are gotten using the
`lubridate` function `year`:

```{r }
denali %>%
  mutate(year = year(date)) %>%
  ggplot(aes(x = caribou, y = wolf, label = year)) + geom_point() + geom_text_repel()
```

 

I thought about joining up the points in year order. This is actually
*not* `geom_line` as you would have guessed, since what that
does is to join points in the order of the variable on the
$x$-axis.
`r tufte::margin_note("I have to say that I didn't know that until just now.")`
To join points in the order that they are in the data (what we want
here, because the points are in time order in the data), use instead 
`geom_path`:

```{r }
denali %>%
  mutate(year = year(date)) %>%
  ggplot(aes(x = caribou, y = wolf, label = year)) + geom_point() +
  geom_text_repel() + geom_path()
```

 

In 1996, both populations were large, and both showed a steady decline
until 1999. In 2000 and 2001, both populations seemed to be on the way
up again, and you can imagine that in a couple of years, things would
go back to about where they were in 1995.



(f) Make a plot of caribou population against time (this is done
the obvious way). What seems to be happening to the caribou
population over time?


Solution


Make a scatterplot, with the survey date as explanatory
variable, and caribou population as response (since time always
goes on the $x$-axis):
```{r }
ggplot(denali, aes(x = date, y = caribou)) + geom_point() + geom_line()
```

 

I used an ordinary `geom_line` this time, to connect neighbouring
years, as is often done with a time series. The overall trend is
downward, though the 1999 value might be a low from which the
population is recovering.
      


(g) The caribou and wolf populations over time are 
really "time series". 
See if you can make a plot of *both* the
caribou and wolf populations against time. You can make two
$y$-axes, one for caribou and one for wolf; this will probably
require some research on your part to figure out.


Solution


The obvious starting point is to note that both the
`caribou` and `wolf` columns are animal
populations, just of different animals. One way of plotting both
populations is to `pivot_longer` them up into one longer column, and
then plot them against time, with the two animals distinguished
by colour:
```{r }
denali %>%
  pivot_longer(caribou:wolf, names_to="animal", values_to="population") %>%
  ggplot(aes(x = date, y = population, colour = animal)) +
  geom_point() + geom_line()
```

       
This is not quite the story, though, because the caribou and wolf
populations are on different scales. The caribou population is
numbered in hundreds, while the wolf population is an actual count. 

The surveys are late in the year, so the one that is nearly in 1996 is
actually the 1995 survey.

What would be nice would be to have a secondary $y$-axis, so that
there were two $y$-scales, one for each animal. This is very easy to
manipulate, though (you can change either scale and get a very
different-looking graph), so we ought to be careful.

All right, so let's put the caribou on the left:

```{r }
ggplot(denali, aes(x = date, y = caribou)) + geom_line()
```

 

Or we can add a `colour` aesthetic to distinguish the caribou
from the wolf populations, that we're going to add in a moment. This looks rather odd at first:

```{r }
ggplot(denali, aes(x = date, y = caribou, colour = "caribou")) + geom_line()
```

 

Now we think about adding the wolf numbers. This is done by adding a
second `geom_line`, overriding the `y` and the
`colour` to specify that this is wolf now:

```{r }
ggplot(denali, aes(x = date, y = caribou, colour = "caribou")) +
  geom_line() +
  geom_line(aes(y = wolf, colour = "wolf"))
```

 

What has happened is that we get lines of different colour for each
animal, with a legend. So far so good. The problem is that the wolf
numbers are about 2.5 times bigger than the caribou
numbers,
`r tufte::margin_note("Which means, if you stop to think about it, that  there are *actually* about 40 times more caribou than wolves.")`
so that
we don't get a good sense of how they go up and down together. If we
divided the wolf numbers by 2.5, we would see this better:

```{r }
ggplot(denali, aes(x = date, y = caribou, colour = "caribou")) +
  geom_line() +
  geom_line(aes(y = wolf / 2.5, colour = "wolf"))
```

 

Now we get to the secondary $y$-axis. We want to label this
`wolf` and have it reflect that we actually made the graph by
dividing the wolf values by 2.5:

```{r }
ggplot(denali, aes(x = date, y = caribou, colour = "caribou")) +
  geom_line() +
  geom_line(aes(y = wolf / 2.5, colour = "wolf")) +
  scale_y_continuous(sec.axis = sec_axis(~ . * 2.5, name = "wolf"))
```

 

Woo, and, very possibly, hoo. I got most of these ideas from
[link](https://rpubs.com/MarkusLoew/226759). 

Now we see how the populations vary over time, and also that they vary
together. 

This is about the only double-$y$-axis setup that I like, with scales
chosen so that both the series vary about the same amount. By
"discreetly" changing the wolf scale, you could make it look as if
one population was much bigger than the other, or varied much more
than the other. Lies and statistics.

In my opinion, too many people just plot series against time, possibly
with a second $y$-axis.
`r tufte::margin_note("And all too often with Excel (spit).")`
Variables that vary together, like the wolf
and caribou populations here, ought to be plotted \emph{against each
other} on a scatterplot, possibly with the time points labelled.

The ambitious among you may like to compare the graphs here with
other predator-prey relationships. If you are of a mathematical bent,
you might look into the Lotka-Volterra equations, which is a system of
two differential equations describing how changes in one population
cause changes in the other population.

      



