##  Running and blood sugar


 A diabetic wants to know how aerobic exercise affects his
blood sugar. When his blood sugar reaches 170 (mg/dl), he goes out for
a run at a pace of 10 minutes per mile. He runs different distances on
different days. Each time he runs, he measures his blood sugar after
the run. (The preferred blood sugar level is between 80 and 120 on
this scale.) The data are in the file
[link](http://ritsokiguess.site/datafiles/runner.txt).  Our aim is
to predict blood sugar from distance.



(a) Read in the data and display the data frame that you read
in.
 
Solution


From the URL is easiest. These are delimited by one space, as
you can tell by looking at the file:
```{r }
my_url <- "http://ritsokiguess.site/datafiles/runner.txt"
runs <- read_delim(my_url, " ")
runs
```

     

That looks like my data file.
 

(b) Make a scatterplot and add a smooth trend to it.
 
Solution


```{r plymouth}
ggplot(runs, aes(x = distance, y = blood_sugar)) + geom_point() +
  geom_smooth()
```

     

`blood_sugar` should be on the vertical axis, since this is
what we are trying to predict. Getting the `x` and the
`y` right is easy on these, because they are the $x$ and $y$
for your plot.
 

(c) Would you say that the relationship between blood sugar and
running distance is approximately linear, or not? It is therefore
reasonable to use a regression of blood sugar on distance? Explain briefly.
 
Solution


I'd say that this is about as linear as you could ever wish
for. Neither the pattern of points nor the smooth trend have any
kind of noticeable bend in them. (Observing a lack of curvature in
either the points or the smooth trend is enough.) The trend
is a linear one, so using a regression will be just fine. (If it
weren't, the rest of the question would be kind of dumb.)
 

(d) Fit a suitable regression, and obtain the regression output.
 
Solution


Two steps: `lm` and then `summary`:
```{r }
runs.1 <- lm(blood_sugar ~ distance, data = runs)
summary(runs.1)
```

     
 

(e) How would you *interpret* the slope? That is, what is
the slope, and what does that mean about blood sugar and running distance?
 
Solution


The slope is $-25.37$. This means that for each additional mile run,
the runner's blood sugar will decrease on average by about 25 units.

You can check this from the scatterplot. For example, from 2 to 3
miles, average blood sugar decreases from about 140 to about 115, a
drop of 25.
 

(f) Is there a (statistically) significant relationship between
running distance and blood sugar? How do you know? Do you find this
surprising, given what you have seen so far? Explain briefly.
 
Solution


Look at the P-value either on the `distance` line (for its
$t$-test) or for the $F$-statistic on the bottom line. These are
the same: 0.000000023. (They will be the same any time there is
one $x$-variable.) This P-value is *way* smaller than 0.05,
so there *is* a significant relationship between running distance
and blood sugar. This does not surprise me in the slightest,
because the trend on the scatterplot is *so* clear, there's
no way it could have happened by chance if in fact there were no
relationship between running distance and blood sugar.
 

(g) This diabetic is planning to go for a 3-mile run tomorrow
and a 5-mile run the day after. Obtain suitable 95\% intervals that
say what his blood sugar might be after each of these runs. 
 
Solution


This is a prediction interval, in each case, since we are talking about
*individual* runs of 3 miles and 5 miles (not the mean blood
sugar after *all* runs of 3 miles, which is what a confidence
interval for the mean response would be).
The procedure is to set up a data frame with the two
`distance` values in it, and then feed that and the
regression object into `predict`, coming up in a moment.
```{r }
dists <- c(3, 5)
dist.new <- tibble(distance = dists)
dist.new
```

     
The important thing is that the name of the column of the new data
frame must be *exactly* the same as the name of the explanatory
variable in the regression. If they don't match, `predict`
won't work. At least, it won't work properly.
`r tufte::margin_note("It won't give you an error, but it will go back to the *original* data frame to get distances to predict from, and you will get very confused.")`

Then, `predict`:

```{r }
pp <- predict(runs.1, dist.new, interval = "p")
pp
```

 

and display this with the distances by the side:

```{r }
cbind(dist.new, pp)
```

 

or

```{r }
data.frame(dist.new, pp)
```

 
Blood sugar after a 3-mile run is predicted to be between 104 and 127;
after a 5-mile run it is predicted to be between 52 and 77.5.

Extra: both `cbind` and `data.frame` are "base R" ways of
combining a data frame with something else to make a new data
frame. They are not from the `tidyverse`. The
`tidyverse` way is via `tibble` or `bind_cols`,
but they are a bit more particular about what they will take:
`tibble` takes vectors (single variables) and
`bind_cols` takes vectors or data frames. The problem here is
that `pp` is not either of those:

```{r }
class(pp)
```

 

so that we have to use `as_tibble` first to turn it into a
data frame, and thus:

```{r }
pp %>% as_tibble() %>% bind_cols(dist.new)
```

 

which puts things backwards, unless you do it like this:

```{r }
dist.new %>% bind_cols(as_tibble(pp))
```

 

which is a pretty result from very ugly code. 

I also remembered that if you finish with a `select`, you get the columns in the order they were in the `select`:

```{r }
pp %>%
  as_tibble() %>%
  bind_cols(dist.new) %>%
  select(c(distance, everything()))
```

 

`everything` is a so-called "select helper". It means 
"everything except any columns you already named", so this whole thing has the effect of listing the columns 
with `distance` first and all the other columns afterwards, in the order that they were in before.
 

(h) Which of your two intervals is longer? Does this make
sense? Explain briefly.
 
Solution


The intervals are about 22.25 and 25.5 units long. The one for a
5-mile run is a bit longer. I think this makes sense because 3
miles is close to the average run distance, so there is a lot of
"nearby" data. 5 miles is actually longer than any of the runs
that were actually done (and therefore we are actually
extrapolating), but the important point for the prediction
interval is that there is less nearby data: those 2-mile runs
don't help so much in predicting blood sugar after a 5-mile
run. (They help *some*, because the trend is so linear. This
is why the 5-mile interval is not *so* much longer. If the
trend were less clear, the 5-mile interval would be more
noticeably worse.)
 


