##  Making it over the mountains


 In 1846, the Donner party (Donner and Reed
families) left Springfield, Illinois for California in covered
wagons. After reaching Fort Bridger, Wyoming, the leaders decided to
find a new route to Sacramento. They became stranded in the eastern
Sierra Nevada mountains at a place now called Donner Pass, when the
region was hit by heavy snows in late October. By the time the
survivors were rescued on April 21, 1847, 40 out of 87 had died.

After the rescue, the age and gender of each person in the party was
recorded, along with whether they survived or not. The data are in
[link](http://ritsokiguess.site/datafiles/donner.txt). 



(a) Read in the data and display its structure. Does that agree
with the description in the previous paragraph?


Solution


Nothing very new here:
```{r }
my_url <- "http://ritsokiguess.site/datafiles/donner.txt"
donner <- read_delim(my_url, " ")
donner
```

     

The ages look like ages, and the other variables are categorical as
promised, with the right levels. So this looks sensible.

I usually write (or rewrite) the data description myself, but the one
I found for these data at
[link](https://onlinecourses.science.psu.edu/stat504/node/159) was so
nice that I used it as is. You can see that the data format used on
that website is not as nice as ours. (I did some work to make it look
nicer for you, with proper named categories rather than 0 and 1.)
    


(b) Make graphical or numerical summaries for each pair of
variables. That is, you should make a graph or numerical summary for
each of 
`age` vs. `gender`, `age` vs.\
`survived` and `gender` vs. `survived`. In choosing the
kind of graph or summary that you will use, bear in mind that
`survived` and `gender` are factors with two
levels. 


Solution


Thinking about graphs first: we have two numeric-vs-factor graphs
(the ones involving `age`), which I think should be
boxplots, though they could be something like side-by-side
histograms (if you are willing to grapple with
`facet_grid`). The other two variables are both
factors, so a good graph for them would be something
like a grouped bar plot (as in the question about parasites earlier).
If you prefer numerical summaries: you could do mean age (or some
other summary of age) for each group defined by gender or
survivorship, and you could do a cross-tabulation of frequencies
for gender and survival. I'll take any mixture of graphs and
numerical summaries that addresses each pair of variables somehow
and summarizes them in a sensible way.
Starting with `age` vs. `gender`:
```{r }
ggplot(donner, aes(x = gender, y = age)) + geom_boxplot()
```

     

or:

```{r }
ggplot(donner, aes(x = age)) + geom_histogram(bins = 10) + facet_grid(gender ~ .)
```


or:

```{r }
donner %>% group_by(gender) %>% summarize(m = mean(age))
```

 

Age vs. `survived` is the same idea:

```{r }
ggplot(donner, aes(x = survived, y = age)) + geom_boxplot()
```

     

or:

```{r }
ggplot(donner, aes(x = age)) + geom_histogram(bins = 10) + facet_grid(survived ~ .)
```


or:

```{r }
donner %>% group_by(survived) %>% summarize(m = mean(age))
```

 

For `survived` against `gender`, the obvious thing is a
cross-tabulation, gotten like this:

```{r }
with(donner, table(gender, survived))
```

 

or like this:

```{r }
donner %>% group_by(gender, survived) %>% summarize(n = n())
```

 

For a graph, borrow the grouped bar-plot idea from the parasites question:

```{r }
ggplot(donner, aes(x = gender, fill = survived)) + geom_bar(position = "dodge")
```

 

I think this way around for `x` and `fill` is better,
since we want to ask something like "out of the females, how many survived" 
(that is, gender is explanatory and survival response). 
    


(c) For each of the three graphs or summaries in the previous
question, what do they tell you about the relationship between the
pair of variables concerned? Explain briefly.


Solution


I think the graphs are the easiest thing to interpret, but use
whatever you got:


* The females on average were younger than the males. (This
was still true with the medians, even though those very old
males might have pulled the mean up.)

* The people who survived were on average younger than those
who didn't (or, the older people tended not to survive).

* A greater proportion of females survived than males.

A relevant point about each relationship is good.
    


(d) Fit a logistic regression predicting survival from age and
gender. Display the summary.


Solution


Each row of the data frame is one person, so we can use the
`survived` column without going through that two-column
response business.
However, the response variable `survived` is a categorical
variable expressed as text, so we need to make it into a
`factor` first. Either create a new variable that is the
factor version of `survived`, or do it right in the `glm`:
```{r }
donner.1 <- glm(factor(survived) ~ age + gender, family = "binomial", data = donner)
summary(donner.1)
```

   
We ought to take a moment to think about what is being predicted here:

```{r }
levels(factor(donner$survived))
```

 

The baseline is the first of these, `no`, and the thing that is
predicted is the probability of the second one, `yes` (that is,
the probability of surviving).
    


(e) Do both explanatory variables have an impact on survival?
Does that seem to be consistent with your numerical or graphical
summaries? Explain briefly.



Solution


Both explanatory variables have a P-value (just) less than 0.05, so
they both have an impact on survival: taking either one of them out
of the logistic regression would be a mistake.
To see if this makes sense, go back to your plots or summaries, the
ones involving survival. For age, the mean or median age of the
survivors was less than for the people who died, by five year
(median) or eight years (mean), so it makes sense that there would
be an age effect. For gender, two-thirds of the women survived and
two-thirds of the men died, so there ought to be a gender effect and is.
  


(f) Are the men typically older, younger or about the same age as
the women? Considering this, explain carefully what the negative
`gendermale` slope in your logistic regression means.



Solution


The men are typically older than the women. 
The negative (and significant) `gendermale` slope means that
the probability of a male surviving is less than that of a woman
*of the same age*. Or, though the males are typically
older, *even after you allow for that*, the males have worse
survival. (`genderfemale` is the baseline.)
You need to get at the idea of "all else equal" when you're
assessing regression slopes of any kind: regular regression,
logistic regression or survival analysis (coming up later). That's
why I said "carefully" in the question. If I say "carefully" or
"precisely", a complete answer is looking for a specific thing to
show that you understand the issue completely.
  


(g) Obtain the median and quartiles of age, and construct a data
frame that contains all combinations of those ages and the two
genders, with columns called exactly `age` and
`gender`. Use this data frame to obtain predicted probabilities
of survival for each of those combinations of age and gender.



Solution


I have a standard way of doing this:


* Make a list of all the variable values I'm going to predict
for, using plural names

* Make a data frame containing all combinations, using
`crossing`, giving the columns the right names

* Feed this data frame into `predict`, along with a
fitted model, and get a column of predictions.

* Glue those predictions onto the data frame of values to
predict for, and display them.

That way I can do it without thinking too hard.
There is a lazy way to do step 1 (which I'll show first), and a way
that is more work but more reliable (because you have to concentrate
less).
Step 1, the lazy way:
```{r }
donner %>% select(age) %>% summary()
donner %>% count(gender)
```

   

Then construct vectors called `ages` and `genders` by
literally copying the values output from above:

```{r }
ages <- c(24, 28, 40)
genders <- c("female", "male")
```

 

There is a risk here, because it is very easy to make typos while
copying, so it would be more reliable to compute these (and, as long
as you trust your computation, reproducible as well). Thinking of
reproducibility, if the data set changes, your computation of
quartiles etc.\ will still work, whereas your copying above won't.

So, let's get the median and quartiles of age:

```{r }
donner %>%
  summarize(
    q1 = quantile(age, 0.25),
    med = median(age),
    q3 = quantile(age, 0.75)
  )
```

 

only these came out sideways, so we can "transpose" them:

```{r }
donner %>%
  summarize(
    q1 = quantile(age, 0.25),
    med = median(age),
    q3 = quantile(age, 0.75)
  ) %>%
  pivot_longer(everything(), names_to="stat", values_to="value")
```

 

and now if we "pull out" the `value` column, we are good:

```{r }
ages <- donner %>%
  summarize(
    q1 = quantile(age, 0.25),
    med = median(age),
    q3 = quantile(age, 0.75)
  ) %>%
  pivot_longer(everything(), names_to="stat", values_to="value") %>% 
  pull(value)
ages
```

 

The same kind of idea will get us both genders without typing any genders:

```{r }
genders <- donner %>%
  count(gender) %>%
  pull(gender)
genders
```

 

We don't care how many there are of each gender; it's just a device to
get the different genders.
`r tufte::margin_note("The counts are in a column called  *n* which we calculate and then ignore.")` This is another way:

```{r }
genders <- donner %>%
  select(gender) %>%
  distinct() %>%
  pull(gender)
genders
```

 

The rationale behind this one is that `distinct()` removes
duplicates (or duplicate combinations if you have more than one
variable), leaving you with only the distinct ones (the ones that are
different from each other). 

I don't think the genders are ever going to change, but it's the
principle of the thing.

Step 2: 

```{r }
donner.new <- crossing(age = ages, gender = genders)
donner.new
```

 

All combinations of the three ages and two genders.

Step 3:

```{r }
p <- predict(donner.1, donner.new, type = "response")
p
```



Step 4:

```{r }
cbind(donner.new, p)
```

 

These, remember, are predicted probabilities of *surviving*. 

  


(h) Do your predictions support your conclusions from earlier
about the effects of age and gender? Explain briefly.



Solution


We said before that the probability of survival was lower if the age
was higher. This is confirmed here: for example, look at rows 1, 3
and 5, which are all females of increasing ages; the probability of
survival decreases. (Or look at males, in rows 2, 4 and 6; the
effect is the same.)

To see the effect of gender, look at two predictions of different
genders but the same age (eg. rows 1 and 2). The female is always
predicted to have the higher survival probability. This is also what
we saw before. The effect of gender is substantial, but not strongly
significant, because we only have 45 observations, not so many when
all we know about each person is whether they survived or not.
I wanted you to think about these different ways to understand the
model, and to understand that they all say the same thing, in
different ways (and thus you can look at whichever of them is most
convenient or comprehensible). For the logistic and survival models,
I find looking at predictions to be the easiest way to understand
what the model is saying.
  



