##  The effects of Substance A


 In a dose-response experiment, animals (or
cell cultures or human subjects) are exposed to some toxic substance,
and we observe how many of them show some sort of response. In this
experiment, a mysterious Substance A is exposed at various doses to
100 cells at each dose, and the number of cells at each dose that
suffer damage is recorded. The doses were 10, 20, ... 70 (mg), and
the number of damaged cells out of 100 were respectively 10, 28, 53,
77, 91, 98, 99.



(a) Find a way to get these data into R, and show that you have
managed to do so successfully.


Solution


There's not much data here, so we don't need to create a file,
although you can do so if you like (in the obvious way: type the
doses and damaged cell numbers into a `.txt` file or
spreadsheet and read in with the appropriate `read_`
function). 
Or, use a `tribble`:
```{r doseresponse-1 }
dr <- tribble(
  ~dose, ~damaged,
  10, 10,
  20, 28,
  30, 53,
  40, 77,
  50, 91,
  60, 98,
  70, 99
)
dr
```

       
Or, make a data frame with the values typed in:

```{r doseresponse-2 }
dr2 <- tibble(
  dose = seq(10, 70, 10),
  damaged = c(10, 28, 53, 77, 91, 98, 99)
)
dr2
```

 

`seq` fills a sequence "10 to 70 in steps of 10", or you can
just list the doses.

I like this better than making two columns *not* attached to a
data frame, but that will work as well.

For these, find a way you like, and stick with it.

    

$\blacksquare$

(b) Would you expect to see a significant effect of dose for
these data?
Explain briefly.


Solution


The number of damaged cells goes up sharply as the dose goes up
(from a very small number to almost all of them). So I'd expect
to see a strongly significant effect of dose. This is far from
something that would happen by chance.
 

$\blacksquare$

(c) Fit a logistic regression modelling the probability of a
cell being damaged as it depends on dose. Display the
results. (Comment on them is coming later.)


Solution


This has a bunch of observations at each dose (100 cells, in
fact), so we need to do the two-column response thing: the
successes in the first column and the failures in the
second. This means that we first need to calculate the number of
cells at each dose that were not damaged, by subtracting the
number that *were* damaged from 100. R makes this easier
than you'd think, as
you see. A `mutate` is the way to
go: create a new column in `dr` and save back in
`dr` (because I like living on the edge):
```{r doseresponse-3 }
dr <- dr %>% mutate(undamaged = 100 - damaged)
dr
```

       

The programmer in you is probably complaining "but, 100 is a number and `damaged` is a vector of 7 numbers. How does R know to subtract *each one* from 100?" Well, R has what's known as
"recycling rules": if you try to add or subtract (or elementwise
multiply or divide) two vectors of different lengths, it recycles the
smaller one by repeating it until it's as long as the longer one. So
rather than `100-damaged` giving an error, it does what you
want.^[The usual application of this is to combine a number  with a vector. If you try to subtract a length-2 vector from a  length-6 vector, R will repeat the shorter one three times and do  the subtraction, but if you try to subtract a length-2 vector from a length-*7* vector, where you'd have to repeat the shorter one a fractional number of times, R will do it, but you'll get a warning, because this is probably *not* what you wanted. Try it and see.]

I took the risk of saving the new data frame over the old one. If it
had failed for some reason, I could have started again.

Now we have to make our response "matrix" with two columns, using `cbind`:

```{r doseresponse-4 }
response <- with(dr, cbind(damaged, undamaged))
response
```

 

Note that each row adds up to 100, since there were 100 cells at each
dose. This is actually trickier than it looks: the two things in
`cbind` are columns (vectors), and `cbind` glues them
together to make an R `matrix`:

```{r doseresponse-5 }
class(response)
```

 

which is what `glm` needs here, even though it looks a lot like
a data frame (which wouldn't work here). This *would* be a data
frame:

```{r doseresponse-6 }
dr %>% select(damaged, undamaged) %>% class()
```

 

and would therefore be the wrong thing to give `glm`. So I had
to do it with `cbind`, or use some other trickery, like this:

```{r doseresponse-7 }
dr %>%
  select(damaged, undamaged) %>%
  as.matrix() -> resp
class(resp)
```

 

Now we fit our model:

```{r doseresponse-8 }
cells.1 <- glm(response ~ dose, family = "binomial", data = dr)
summary(cells.1)
```

 
    

$\blacksquare$

(d) Does your output indicate that the probability of damage
really does *increase* with dose? (There are two things here:
is there really an effect, and which way does it go?)


Solution


The slope of `dose` is significantly nonzero (P-value
less than $2.2 \times 10^{-16}$, which is as small as it can
be). The slope itself is *positive*, which means that as
dose goes up, the probability of damage goes up. 
That's all I needed, but if you want to press on: the slope is
0.113, so an increase of 1 in dose goes with an increase of
0.113 in the *log-odds* of damage. Or it multiplies the
odds of damage by $\exp(0.113)$. Since 0.113 is small, this is
about $1.113$ (mathematically, $e^x\simeq 1+x$ if $x$ is small),
so that the increase is about 11\%. 
If you like, you can get a rough CI for the slope by going up
and down twice its standard error (this is the usual
approximately-normal thing). Here that would be
```{r doseresponse-9 }
0.113323 + c(-2, 2) * 0.008315
```

       

I thought about doing that in my head, but thought better of it, since
I have R just sitting here. The interval is short (we have lots of
data) and definitely does not contain zero.
    

$\blacksquare$

(e) Obtain predicted damage probabilities for some representative doses. 


Solution

This is `predictions` with `variables`, which will do the predictions for the five-number summary of the doses (or something very like that):

```{r}
predictions(cells.1, variables = "dose")
```

The advantage to this is that you don't need to do the `datagrid` thing to specify which doses you want.

$\blacksquare$

(f) Draw a graph of the predicted probabilities, and to that add the observed proportions of damage at each dose. Hints: you will have to calculate the observed proportions first. See [here](https://vincentarelbundock.github.io/marginaleffects/articles/mfx01_predictions.html), near the bottom, to find out how to add data to one of these graphs. The `geom_point` line is the one you need.

Looking at  the predicted probabilities, would you say that
the model fits well? Explain briefly.


Solution

Let's take a look at our original dataframe:

```{r}
dr
```

To that we need to add a column of proportion damaged, which is `damaged` divided by `damaged` plus `undamaged`. This last ought to be 100, but data can go missing for any number of reasons, so it pays not to assume that they add up to 100 every time:

```{r}
dr %>% mutate(prop = damaged / (damaged + undamaged)) -> dr2
dr2
```

Check. I saved this to add to the graph later.

Here's the graph of predicted values:

```{r}
plot_cap(cells.1, condition = "dose")
```

The document linked in the hint says that this is an ordinary `ggplot` graph, so you add things to it with a plus as for any `ggplot` graph. That means that you can add a `geom_point` with a `data =` and an `aes`, except that the obvious doesn't quite work:

```{r}
plot_cap(cells.1, condition = "dose") +
  geom_point(data = dr2, aes(x =  dose, y = prop), inherit.aes = FALSE)
```

The message is not helpful, but I can tell you where it comes from. When you add something to a plot like this, all the things in the original `ggplot` (which you don't see because it is hidden inside `plot_cap`) are "inherited" by anything else that you add to the plot, so that you either have to overwrite them with something new (as I did with `x` and `y`) or you get the previous values, one of which was evidently `conf.low`. To override this behaviour, which we don't want because *we* have nothing called `conf.low`, add `inherit.aes = FALSE` to the `geom_point`:^[This is done in the `geom_point` on the website, which is where I learned about it.]

```{r}
plot_cap(cells.1, condition = "dose") +
  geom_point(data = dr2, aes(x =  dose, y = prop), colour = "red", inherit.aes = FALSE)
```

I also made the data points red (you don't need to, but if you want to, put the colour *outside* the `aes`, to make it clear that the colour is not determined by any of the variables in your dataframe).

The predicted
probabilities ought to be close to the observed proportions. They are in fact *very* close to them, so the model
fits very well indeed.

Your actual words are a judgement call, so precisely what you
say doesn't matter so much, but *I* think this model fit is
actually closer than you could even hope to expect, it's that
good. But, your call. I think your answer ought to contain
"close" or "fits well" at the very least. 
     
$\blacksquare$




