##  Drug treatment programs


 One of the goals of drug treatment programs is to lengthen
the time until the patient returns to using drugs. (It is not
generally possible to prevent patients from *ever* using drugs
again.) In one study, over 600 former drug users took part. Two
different programs, a short program and a long program, were offered
at two different sites, labelled A and B. The data can be found in
[link](http://ritsokiguess.site/datafiles/drugusers.txt). The
variables are these:



* `ID`: patient ID number

* `age`: patient age at enrollment into the study

* `ndrugtx`: number of previous drug treatments

* `treat`: 0 for short treatment program, 1 for long program

* `site`: 0 for site A, 1 for site B

* `time`: time until return to drug use

* `censor`: whether the subject returned to drug use (1) or
not (0) during the follow-up period

* `herco`: whether subject used heroine or cocaine in the
last 3 months: 1 is both, 2 is one (either heroine or cocaine), 3 is neither.



(a) Read in the data and check in one way or another that you
have what was promised above.


Solution


This:
```{r }
my_url <- "http://ritsokiguess.site/datafiles/drugusers.txt"
drugusers <- read_delim(my_url, " ")
drugusers
```

 

This shows that you have over 600 rows and the variables described.
    

$\blacksquare$

(b) There are some missing values in the dataframe. Demonstrate
this using `summary`. Pipe the dataframe into
`drop_na` and show that they have gone. (`drop_na`
removes all rows that have missing values in them.)


Solution


First off, `summary` is a quick way to show how many missing
values there are:
`r tufte::margin_note("It doesn't work with text columns, but it  *does* work if you temporarily turn the text columns into  factors, eg. by using *mutate-if*. However, we don't have any text  columns here, so what we do here is good for this data set.")`
```{r }
summary(drugusers)
```

 

Age has five missing values and "number of previous drug treatments"
has seventeen.

Following the instructions, and saving back into the original dataframe:

```{r }
drugusers %>% drop_na() -> drugusers
```

 
and then

```{r }
summary(drugusers)
```

 

No NA left. Gosh, as they say, that was easy.
Extra: how many rows did we lose?

```{r }
nrow(drugusers)
```

 

There were 628 rows before, so we lost 18. (There were 22 missing
values, but some of them were two on one row, so we only lost 18
rows.)

This is a very unsophisticated way of dealing with missing
values. Another way is to "impute" them, that is, to guess what they
would have been, and then fill in the guessed values and use them as
if they were the truth, for example by regressing the columns with missing values on all the others, and using the regression predictions in place of the missing values.


$\blacksquare$

(c) Some of these variables are recorded as numbers but are
actually categorical. Which ones? Re-define these variables in your
data frame so that they have sensible (text) values.


Solution


These variables are actually categorical rather than quantitative:


* `treat`

* `site`

* `censor`

* `herco`

Most of them have only two levels, so it doesn't matter whether
we make them categorical or leave them as numbers, but for
`herco` it matters. Let's give them all sensible values,
mostly with `ifelse`,
`r tufte::margin_note("Case-when is much clearer than using nested if-elses when you have three or more categories, as for *herco*.")` thus:
```{r }
drugusers %>% mutate(
  treat = ifelse(treat == 0, "short", "long"),
  site = ifelse(site == 0, "A", "B"),
  censor = ifelse(censor == 1, "returned", "no-return"),
  herco = case_when(
    herco == 1 ~ "both",
    herco == 2 ~ "one",
    herco == 3 ~ "neither"
  )
) -> drugusers
```

       

I'm living on the edge and overwriting everything:

```{r }
drugusers
```

 


$\blacksquare$

(d) Create a suitable reponse variable for a Cox proportional
hazards regression that predicts time until return to drug use from
the other variables.
This requires some care, because you need to be sure about what the
censoring variable actually represents and what you need it to represent.


Solution


This is `Surv` in package `survival`. The response
variable needs to encode two things: the time until the event of
interest (return to drug use) and whether or not that event
happened for each patient.
`r tufte::margin_note("Some people define the response variable right inside the *coxph*, in the same way as putting something like *log(y)* as a response in an *lm*, but I think, especially while you're getting used to the process, it's better to create the response variable first and look at it to make sure it's the right thing.")`
In this case, that is
`censor="returned"`. 
```{r }
y <- with(drugusers, Surv(time, censor == "returned"))
```

       

Use whatever name you gave to the level of `censor` that means "returned to drug use". 
    

$\blacksquare$

(e) Look at the first few values of your response variable. Why
is the fifth one marked with a `+`? Explain briefly.


Solution


`head` works as well with a vector (displaying the first
six values) as it does with a data frame:
```{r }
head(y)
```

      

The fifth value is marked with a `+` because it is a censored
value: this is a patient who was never observed to go back to drug
use. You can tell this by looking at the `head` of the entire
data frame:

```{r }
head(drugusers)
```

 

since this patient has `censor="no-return"`. The other ones have
`censor="returned"`; these are all "uncensored" in the jargon.

Typically, censored values will be bigger than uncensored ones,
because (in general) the individual will be observed until the study
ends, and studies of this kind carry on for years:

```{r }
ggplot(drugusers, aes(x = censor, y = time)) + geom_boxplot()
```

 
Yep. The smallest time for a censored observation would be an upper outlier
if it were observed for an uncensored observation.

One nice side-effect of turning `censor` into a categorical variable is that it can now
distinguish groups as a boxplot requires.

I discovered something rather amusing when I originally wrote this (a
year ago). Suppose you want to compare times for the two treatment
groups, and you *also* want to distinguish censored from
non-censored observations. Then, this works:

```{r }
ggplot(drugusers, aes(x = treat, y = time, colour = censor)) +
  geom_boxplot()
```

 

For each treatment, you get side-by-side boxplots of the times for
censored (red) and uncensored (blue) observations, and so you see for
both treatments (short and long) the censored times are typically
longer than the uncensored ones.

(This you may recognize as a "grouped boxplot", for when we have two
categorical variables and one quantitative one.)

I borrow this idea for two-way ANOVA (coming up later).
    

$\blacksquare$

(f) Fit a Cox proportional hazards model, predicting from all
the other variables (except for `row` and `ID`) that you haven't used
yet. Display the results.


Solution


```{r }
drugusers.1 <- coxph(y ~ age + ndrugtx + treat + site + herco, data = drugusers)
summary(drugusers.1)
```

   

Another way to handle 
"all the other $x$'s except `row`, `ID`, `time` and `censor`" is this:

```{r }
drugusers.1a <- coxph(y ~ . - row - ID - time - censor, data = drugusers)
broom::tidy(drugusers.1a)
```

 

Same. I used `tidy` from `broom` to shorten the output a bit.
    

$\blacksquare$

(g) Find which explanatory variables can be removed at $\alpha=0.05$
(there should be two of them). Bear in mind that we have categorical variables, so that looking at the output from `summary` is not enough.


Solution


The hint is meant to suggest to you that looking at `drop1` is the right way to go:
```{r error=T}
drop1(drugusers.1, test = "Chisq")
```

       

Note that `herco`, a categorical variable with three levels,
has 2 degrees of freedom here, since a test of 
"no effect of `herco`" is testing that survival is the same at 
*all three* levels of `herco`.


$\blacksquare$

(h) Remove *all* the non-significant explanatory variables
and re-fit your model. By carrying out a suitable test demonstrate
that your smaller model is the better one.


Solution


`site` and `herco` are the two variables to come
out.
`r tufte::margin_note("The researchers were probably relieved that there        was not quite a significant effect of *site*.")` I like
`update`, but there is no
problem about copying-pasting your `coxph` and taking out
what you no longer need.
```{r }
drugusers.2 <- update(drugusers.1, . ~ . - site - herco)
```

 

Having fit two models, we can use `anova` to compare them. The
right test gets done, so no need for `test=`:

```{r }
anova(drugusers.2, drugusers.1)
```

 

There is no significant difference between these two 
models,
`r tufte::margin_note("Not at the 0.05 level, anyway.")` so we can go with the 
smaller, simpler one
(with just `age`, `ndrugtx` and `treat`).
    

$\blacksquare$

(i) <a name="part:posneg">*</a> Display your better model. Are all of the
explanatory variables significant? Do their slope coefficients have
sensible signs (plus or minus), based on what you know or can guess
about drug treatments? Explain briefly.


Solution


```{r }
summary(drugusers.2)
```

       

The three remaining explanatory variables are all clearly significant:
the patient's age, the number of previous drug treatments, and whether
the treatment was short or long. This is legit (we don't need to run
`drop1` again) because the remaining explanatory variables are
all quantitative or have only two levels, so that the single-df tests
in `summary` are what we need.

Do their slope coefficients have sensible signs? Well, this requires
careful thought. A positive coefficient means that increasing that
variable *increases the hazard of the event*: ie., it makes the
event likelier to happen sooner. Here, the "event" is 
"return to drug use":



* Age has a negative coefficient, so an older person is likely to take
*longer* to return to drug use, other things being equal. This
makes some kind of sense, if you imagine drug use as being related to
maturity, or an older drug user as being more strongly committed to
"turning their life around", so that a drug treatment of any kind is
going to be more effective on an older patient.

* The number of previous treatments has a positive coefficient, so that a
patient who has had a lot of previous treatments is likely to go back
to drugs sooner. Such a person might be an "addict" for whom
treatments really do not work, or might not be committed to giving up drugs.


* `treatshort` has a positive coefficient.  This says that
if you give a patient a short treatment, they are more likely (other
things being equal) to go back to drugs sooner, as compared to the
baseline long treatment.  That is, a longer treatment is more
effective than a shorter one. Given a significant effect of
treatment length, this is the way around you would expect it to be.


    

$\blacksquare$

(j) We have three variables left in our model, `age`,
`ndrugtx` and `treat`. The quartiles of age are 27 and
37, the quartiles of `ndrugtx` are 1 and 6, and the two
possible values of `treat` are `short` and `long`. Create a data frame
with variables of these names and all possible combinations of their
values (so there should be 8 rows in the resulting data
frame). Display the resulting data frame.


Solution


This data frame is going to be used for prediction, so I will
call it `new` and construct it in pieces as I did before
(thus meaning that I don't have to think too hard about what I'm doing):
```{r }
ages <- c(27, 37)
ndrugtxs <- c(1, 6)
treats <- c("short", "long")
new <- crossing(age = ages, ndrugtx = ndrugtxs, treat = treats)
new
```

       

8 rows as promised.
    

$\blacksquare$

(k) Obtain predicted survival probabilities for each of the
values of `age`, `ndrugtx` and `treat` used in
the previous part. You don't need to display it (we are going to
plot it shortly). 

Solution


`survfit` is the survival analysis version of
`predict` and works the same way, so this is all you need:
```{r }
pp <- survfit(drugusers.2, new, data = drugusers)
```

       

Make sure that you use your best model, ie.\ the second one. The
`data=` is needed for the plot below, not in itself for the
prediction. 

$\blacksquare$

(l) Plot your predicted survival curves. 


Solution


This:
```{r }
ggsurvplot(pp, conf.int = F)
```

   

The only thing to remember is that you plot your *predictions*,
not the model from which they came.

If your plot didn't come out, you may have to go back and re-do the
`survfit` with the `data=` at the end.

For reference in a minute:

```{r }
new
```

 

$\blacksquare$

(m) Which of your combinations of values is predicted to take
the longest to return to drug use? Which is predicted to take the
shortest time? Explain briefly.


Solution


Remember that "up and to the right" is the best survival
curve: that is, the people on this survival curve are predicted
to take the longest to return to drug use. On my plot, this is
the pale blue survival curve, stratum 5. Going back to my
combinations data frame, this is 37-year-olds with only one
previous drug treatment and the longer drug treatment this time.

The worst is my green survival curve, stratum 4, which is
the exact opposite of this: 27-year-olds, 6 previous drug
treatments, shorter treatment this time.

"Returning to drug use" is like "death" in that you want it
to be a long time before it happens, so "best" is top right on
the plot of survival curves. In other curcumstances, you might
want the event to happen *sooner*, in which case the
lower-left survival curve would be the "best" one.
    

$\blacksquare$

(n) Are your survival curve plot and your conclusions from part
(<a href="#part:posneg">here</a>) consistent, or not? Explain briefly.


Solution


The survival curves say that being older, having fewer previous
treatments and being on the long treatment are better in terms
of taking longer to return to drug use. Our analysis of whether
the slope coefficients in `drugusers.2` were positive or
negative came to exactly the same conclusion. So the survival
curves and part (<a href="#part:posneg">here</a>) are entirely consistent.

On my plot with the legend, you can assess the effects of the
individual variables: for example, to assess the effect of age,
find two combos that differ only in age, say strata 1 and 5, the
red and light blue ones. Of these, the light blue survival curve is
higher, so age 37 is better in terms of survival than age
27. This will work whichever such pair you pick: for example,
strata 3 and 7, the olive green and purple curves, compare
the same way. 

Extra: more comparisons for you to do: to assess the effect of number of
previous treatments, compare eg. strata 1 and 3, red and olive
green, and to assess the effect of treatment length, compare
eg. strata 5 and 6, light blue and darker blue.

All this struggling to identify colours makes me think of
[link](https://blog.xkcd.com/2010/05/03/color-survey-results/), in
which the guy behind the webcomic XKCD did a survey where he
showed people a whole bunch of different colours and asked the
people to name the colours.
    
$\blacksquare$





