## Students and exercise

 Some students in a Statistics class were asked how many
minutes they typically exercised in a week. The data are shown in
\url{http://ritsokiguess.site/datafiles/exercise.txt}. 

Some of the
students identified as male and some as female. Our concern is how the males
and females compare in terms of the amount of exercise they do.


(a) Take a look at the data file. (Click on the link, or paste the copied link into your web browser.) How is it laid out?

Solution

Aligned in columns. Or, separated by spaces, but a variable number
of them. (The latter is a hint that \verb=read_delim= will not
work, and the former is a hint about what \emph{will} work.)

$\blacksquare$


(b) Read in and display (some of) the data. (This means to
display enough of what you read in to convince others that you read
in the right kind of thing.)

Solution

"Aligned in columns" is the best way of saying it so that you
know to use \verb=read_table=. Hence:

```{r}
my_url <- "http://ritsokiguess.site/datafiles/exercise.txt"
exercise <- read_table(my_url)
exercise
```

Just entering the data frame name displays the first ten rows, which
is usually enough to convince anyone that you have the right thing. 

Call the data frame what you like, as long as it describes what's in
the data frame in some way.


$\blacksquare$


(c) Make a suitable plot of this data frame. 

Solution

Two variables, one quantitative and one categorical, means that a
boxplot is the thing:

```{r}
ggplot(exercise, aes(x=gender, y=minutes)) + geom_boxplot()
```


$\blacksquare$


(d) Does there appear to be any substantial difference in the
average amount of time that males and females spend exercising?
Explain briefly. ("average" could be mean or median. Which is it
here?)

Solution

A boxplot shows the median. So we learn here that the median time
spent exercising per week is very slightly higher for
females. However, there is a lot of variability (the height of the
boxes), and so the difference between the medians is very
small. Hence, there is definitely \emph{no} substantial difference
between males and females. 

Less insightfully, the difference in median between males and females is
very small (but that doesn't rule out the possibility that the
spread is very small also).

$\blacksquare$


(e) How do you know that both distributions, for males as well
as females, are skewed to the right? Explain (very) briefly.

Solution

The upper whiskers, at the top of the box, are longer than the
lower ones (at the bottom).

$\blacksquare$


(f) For data like this, why does it make practical sense that
the distributions are skewed to the right?

Solution

Nobody can exercise less than 0 minutes per week, but there is no
upper limit: a student in the class can exercise as much as they
want. This means that there could be unusually high values, but
not unusually low values. 

Extra: Distributions that have a limit on one side are often skewed to the other side. 
This one has a limit on the left, so it is skewed to the *right*. This is most likely to be true if 
there are observations close to the limit, such as the people in this data set that don't exercise at all (and there are some of those).



$\blacksquare$





