## Facebook friends and grey matter



 Is there a relationship between the number
of Facebook friends a person has, and the density of grey matter in
the areas of the brain associated with social perception and
associative memory? To find out, a 2012 study measured both of these
variables for a sample of 40 students at City University in London
(England). The data are at
[http://www.utsc.utoronto.ca/~butler/c32/facebook.txt](http://www.utsc.utoronto.ca/~butler/c32/facebook.txt). The grey
matter density is on a $z$-score standardized scale. The values are
separated by *tabs*.

The aim of this question is to produce an R Markdown report that
contains your answers to the questions below. 

You should aim to make your report flow smoothly, so that it would be
pleasant for a grader to read, and can stand on its own as an analysis
(rather than just being the answer to a question that I set you).
Some suggestions: give your report a title and arrange it into
sections with an Introduction; add a small amount of additional text
here and there explaining what you are doing and why. I don't expect
you to spend a large amount of time on this, but I *do* hope
you will make some effort. (My report came out to 4 Word pages.)


(a) Read in the data and make a scatterplot for predicting the
number of Facebook friends from the grey matter density. On your
scatterplot, add a smooth trend.


Solution


Begin your document with a code chunk containing
`library(tidyverse)`. The data values are
separated by tabs, which you will need to take into account:
```{r }
my_url="http://www.utsc.utoronto.ca/~butler/c32/facebook.txt"
fb=read_tsv(my_url)
fb
ggplot(fb,aes(x=GMdensity,y=FBfriends))+geom_point()+geom_smooth()
```       



(b) Describe what you see on your scatterplot: is there a
trend, and if so, what kind of trend is it? (Don't get too taken in
by the exact shape of your smooth trend.) Think ``form, direction,
strength''. 


Solution


I'd say there seems to be a weak, upward, apparently linear
trend. The points are not especially close to the trend, so I
don't think there's any justification for calling this other
than "weak". (If you think the trend is, let's say,
"moderate", you ought to say what makes you think that: for
example, that the people with a lot of Facebook friends also
tend to have a higher grey matter density. I can live with a
reasonably-justified "moderate".)
The reason I said not to get taken in by the shape of the smooth
trend is that this has a "wiggle" in it: it goes down again
briefly in the middle. But this is likely a quirk of the data,
and the trend, if there is any, seems to be an upward one.



(c) Fit a regression predicting the number of Facebook friends
from the grey matter density, and display the output.


Solution


That looks like this. You can call the "fitted model object"
whatever you like, but you'll need to get the capitalization of
the  variable names correct:
```{r }
fb.1=lm(FBfriends~GMdensity,data=fb)
summary(fb.1)
```       

I observe, though I didn't ask you to, that the R-squared is pretty
awful, going with a correlation of 
```{r }
sqrt(0.1904)
``` 

which *would* look
like as weak of a trend as we saw.
`r tufte::margin_note("Correlations have to go up beyond 0.50 before they start looking at all interesting.")`



(d) Is the slope of your regression line significantly
different from zero? What does that mean, in the context of the
data?


Solution


The P-value of the slope is 0.005, which is less than
0.05. Therefore the slope *is* significantly different from
zero. That means that the number of Facebook friends really does
depend on the grey matter density, for the whole population of
interest and not just the 40 students observed here (that were a
sample from that population). I don't mind so much what you
think the population is, but it needs to be clear that the
relationship applies to a population. 
Another way to approach this is to say that you would expect
this relationship to show up again in another similar
experiment. That also works, because it gets at the idea of
reproducibility. 



(e) Are you surprised by the results of
parts (b) and (d)? Explain briefly.


Solution


I *am* surprised, because I thought the trend on the
scatterplot was so weak that there would not be a significant
slope. I guess there was enough of an upward trend to be
significant, and with $n=40$ observations we were able to get a
significant slope out of that scatterplot. With this many
observations, even a weak correlation can be significantly
nonzero. 
You can be surprised or not, but you need to have some kind of
consideration of the strength of the trend on the scatterplot as
against the significance of the slope. For example, if you
decided that the trend was "moderate" in strength, you would
be justified in being less surprised than I was. 
Here, there is the usual issue that we have proved that the
slope is not zero (that the relationship is not flat), but we
may not have a very clear idea of what the slope actually
*is*. There are a couple of ways to get a confidence
interval. The obvious one is to use R as a calculator and go up
and down twice its standard error (to get a rough idea):
```{r }
82.45+2*27.58*c(-1,1)
```       
The `c()` thing is to get both confidence limits at once. The
smoother way is this:

```{r }
confint(fb.1)
``` 

Feed `confint` a "fitted model object" and it'll give you
confidence intervals (by default 95\%) for all the parameters in it. 

The confidence interval for the slope goes from about 27 to about
138. That is to say, a one-unit increase in grey matter density goes
with an increase in Facebook friends of this much. This is not
especially insightful: it's bigger than zero (the test was
significant), but other than that, it could be almost
anything. *This* is where the weakness of the trend comes back to
bite us. With this much scatter in our data, we need a *much*
larger sample size to estimate accurately how big an effect grey
matter density has.



(f) Obtain a scatterplot with the regression line on it.


Solution


Just a modification
of (a):
```{r }
ggplot(fb,aes(x=GMdensity,y=FBfriends))+geom_point()+
geom_smooth(method="lm")
```       



(g) Obtain a plot of the residuals from the regression against
the fitted values, and comment briefly on it.


Solution


This is, to my mind, the easiest way:
```{r }
ggplot(fb.1,aes(x=.fitted,y=.resid))+geom_point()
```       

There is some "magic" here, since the fitted model object is not
actually a data frame, but it works this way.
That looks to me like a completely random scatter of
points. Thus, I am completely happy with the straight-line regression
that we fitted, and I see no need to improve it.

(You should make two points here: one, describe what you see, and two,
what it implies about whether or not your regression is satisfactory.)

Compare that residual plot with this one:

```{r }
ggplot(fb.1,aes(x=.fitted,y=.resid))+
geom_point()+geom_smooth()
```       
Now, why did I try adding a smooth trend, and why is it not
necessarily a good idea? The idea of a residual plot is that there
should be no trend, and so the smooth trend curve ought to go straight
across. The problem is that it will tend to wiggle, just by chance, as
here: it looks as if it goes up and down before flattening out. But if
you look at the points, *they* are all over the place, not close
to the smooth trend at all. So the smooth trend is rather
deceiving. Or, to put it another way, to indicate a real problem, the
smooth trend would have to be a *lot* farther from flat than this
one is. I'd call this one basically flat.




