##  Where should the fire stations be?


 In city planning, one major issue is where to locate fire
stations. If a city has too many fire stations, it will spend too much
on running them, but if it has too few, there may be unnecessary fire
damage because the fire trucks take too long to get to the fire.

The first part of a study of this kind of issue is to understand the
relationship between the distance from the fire station (measured in
miles in our data set) and the amount of fire damage caused (measured
in thousands of dollars). A city recorded the fire damage and distance
from fire station for 15 residential fires (which you can take as a
sample of "all possible residential fires in that city"). The data
are in [link](http://www.utsc.utoronto.ca/~butler/d29/fire_damage.txt). 



(a) Read in and display the data, verifying that you have the
right number of rows and the right columns.


Solution


A quick check of the data reveals that the data values are
separated by exactly  one space, so:
```{r }
my_url="http://www.utsc.utoronto.ca/~butler/d29/fire_damage.txt"
fire=read_delim(my_url," ")
fire
```

   

15 observations (rows), and promised, and a column each of distances
and amounts of fire damage, also as promised.

One mark for reading in the data, and one for saying something
convincing about how you have the right thing.
    


(b) <a name="part:ttest">*</a> Obtain a 95\% confidence interval for the
mean fire damage. (There is nothing here from STAD29, and your
answer should have nothing to do with distance.)


Solution


I wanted to dissuade you  from thinking too hard here. It's just
an ordinary one-sample $t$-test, extracting the interval from it:
```{r }
t.test(fire$damage)
```

     

Or

```{r }
with(fire,t.test(damage))
```

 

Ignore the P-value (it's testing that the mean is the default
*zero*, which makes no sense). The confidence interval either way
goes from 21.9 to 30.9 (thousand dollars).
    


(c) Draw a scatterplot for predicting the amount of fire damage
from the distance from the fire station. Add a smooth trend to your
plot. 


Solution


We are predicting fire damage, so that goes on the $y$-axis:
```{r }
ggplot(fire,aes(x=distance,y=damage))+geom_point()+geom_smooth()
```

     



(d) <a name="part:howgood">*</a> Is there a relationship between distance from fire station
and fire damage? Is it linear or definitely curved? How strong is
it? Explain briefly.


Solution


When the distance is larger, the fire damage is definitely larger,
so there is clearly a relationship. I would call this one
approximately linear: it wiggles a bit, but it is not to my mind
obviously curved. I would also call it a strong relationship,
since the points are close to the smooth trend.
    


(e) Fit a regression predicting fire damage from distance. How
is the R-squared consistent (or inconsistent) with your answer from
part~(<a href="#part:howgood">here</a>)?


Solution


The regression is an ordinary `lm`:
```{r }
damage.1=lm(damage~distance,data=fire)
summary(damage.1)
```

     

We need to display the results, since we need to see the R-squared in
order to say something about it.

R-squared is about 92\%, high, indicating a strong and linear
relationship. Back in part~(<a href="#part:howgood">here</a>), I said that the
relationship is linear and strong, which is entirely consistent with
such an R-squared. (If you said something different previously, say
how it does or doesn't square with this kind of R-squared value.)

Points: one for fitting the regression, one for displaying it, and two
(at the grader's discretion) for saying what the R-squared is and how
it's consistent (or not) with part~(<a href="#part:howgood">here</a>).

Extra: if you thought the trend was "definitely curved", you would
find that a parabola (or some other kind of curve) was definitely
better than a straight line. Here's the parabola:

```{r }
damage.2=lm(damage~distance+I(distance^2),data=fire)
summary(damage.2)
```

 

There's no evidence here that a quadratic is better.

Or you might even have thought from the wiggles that it was more like cubic:

```{r }
damage.3=update(damage.2,.~.+I(distance^3))
summary(damage.3)
```



No evidence that a cubic is better; that increase in R-squared up to
about 94\% is just chance (bearing in mind that adding *any* $x$,
even a useless one, will increase R-squared).

How bendy is the cubic?

```{r }
ggplot(fire,aes(x=distance,y=damage))+geom_point()+
geom_smooth(method="lm")+
geom_line(data=damage.3,aes(y=.fitted), colour="red")
```

 

The cubic, in red, does bend a little, but it doesn't do an obvious
job of going through the points better than the straight line does. It
seems to be mostly swayed by that one observation with damage over 40,
and choosing a relationship by how well it fits one point is flimsy at
the best of times.  So, by Occam's Razor, we go with the line rather
than the cubic because it (i) fits equally well, (ii) is simpler.
    


(f) <a name="part:cim">*</a> Obtain a 95\% confidence interval for the mean fire damage
\emph{for a residence that is 4 miles from the nearest fire
station}. (Note the contrast with part~(<a href="#part:ttest">here</a>).)


Solution


This is a confidence interval for a mean response at a given value
of the explanatory variable. This is as opposed to
part~(<a href="#part:ttest">here</a>), which is averaged over *all* distances.
So, follow the steps. Make a tiny data frame with this one value
of `distance`:
```{r }
new=tibble(distance=4)
```

     

and then feed it into `predict`:

```{r }
pp=predict(damage.1,new,interval="c")
```

 

and then put it side-by-side with the value it's a prediction for:

```{r }
cbind(new,pp)
```

 

28.5 to 31.4 (thousand dollars).



(g) Compare the confidence intervals of parts
(<a href="#part:ttest">here</a>) and (<a href="#part:cim">here</a>). Specifically, compare their
centres and their lengths, and explain briefly why the results
make sense.


Solution


Let me just put them side by side for ease of comparison:
part~(<a href="#part:ttest">here</a>) is:
```{r }
t.test(fire$damage)
```

       

and part~(<a href="#part:cim">here</a>)'s is

```{r }
pp
```

 

I printed them out like this since these give the centres of the
intervals as well as the lower and upper limits.

The centre of the interval is higher for the mean damage when the
distance is 4. This is because the mean distance is a bit less than 4:

```{r }
fire %>% summarize(m=mean(distance))
```

 

We know it's an upward trend, so our best guess at the mean damage is
higher if the mean distance is higher (in (<a href="#part:cim">here</a>), the
distance is *always* 4: we're looking at the mean fire damage for
*all* residences that are 4 miles from a fire station.)

What about the lengths of the intervals? The one in (<a href="#part:ttest">here</a>)
is about $30.9-21.9=9$ (thousand dollars) long, but the one in
(<a href="#part:cim">here</a>) is only $31.4-28.5=2.9$ long, much shorter. This
makes sense because the relationship is a strong one: knowing the
distance from the fire station is very useful, because the bigger it
is, the bigger the damage going to be, with near certainty. Said
differently, if you know the distance, you can estimate the damage
accurately.  If you don't know the distance (as is the case in
(<a href="#part:ttest">here</a>)), you're averaging over a lot of different
distances and thus there is a lot of uncertainty in the amount of fire
damage also.

If you have some reasonable discussion of the reason why the centres
and lengths of the intervals differ, I'm happy. It doesn't have to be
the same as mine.
      





