##  Dissimilarities between fruits


 Consider the fruits apple, orange, banana, pear,
strawberry, blueberry. We are going to work with these four
properties of fruits:


* has a round shape

* Is sweet

* Is crunchy

* Is a berry



(a)  Make a table with fruits as columns, and with
rows "round shape", "sweet", "crunchy", "berry". In each cell
of the table, put a 1 if the fruit has the property named in the
row, and a 0 if it does not. (This is your opinion, and may not
agree with mine. That doesn't matter, as long as you follow through
with whatever your choices were.)
 
Solution


Something akin to this:


```

Fruit        Apple Orange Banana Pear Strawberry Blueberry
Round shape    1      1      0     0       0         1
Sweet          1      1      0     0       1         0
Crunchy        1      0      0     1       0         0
Berry          0      0      0     0       1         1

```

You'll have to make a choice about "crunchy". I usually eat
pears before they're fully ripe, so to me, they're crunchy.
 

(b) We'll define the dissimilarity between two fruits to be the
number of qualities they *disagree* on. Thus, for example, the
dissimilarity between Apple and Orange is 1 (an apple is crunchy and
an orange is not, but they agree on everything else). Calculate the
dissimilarity between each pair of fruits, and make a square table
that summarizes the results. (To save yourself some work, note that
the dissimilarity between a fruit and itself must be zero, and the
dissimilarity between fruits A and B is the same as that between B
and A.) Save your table of dissimilarities into a file for the next part.
 
Solution


I got this, by counting them:

```
 
Fruit         Apple  Orange   Banana   Pear  Strawberry  Blueberry
Apple           0       1       3       2        3          3
Orange          1       0       2       3        2          2
Banana          3       2       0       1        2          2
Pear            2       3       1       0        3          3
Strawberry      3       2       2       3        0          2
Blueberry       3       2       2       3        2          0

```

I copied this into a file `fruits.txt`. Note that (i) I
have aligned my columns, so that I will be able to use
`read_table` later, and (ii) I have given the first column
a name, since `read_table` wants the same number of column
names as columns.

Extra: yes, you can do this in R too. We've seen some of the tricks
before. 

Let's start by reading in my table of fruits and properties, which
I saved in
[link](http://www.utsc.utoronto.ca/~butler/d29/fruit1.txt):
```{r }
my_url="http://www.utsc.utoronto.ca/~butler/d29/fruit1.txt"
fruit1=read_table(my_url)
fruit1 
```

     

We don't need the first column, so we'll get rid of it:

```{r error=T}
fruit2 = fruit1 %>% select(-Property)
fruit2
```

 

The loop way is the most direct. We're going to be looking at
combinations of fruits and other fruits, so we'll need two loops one
inside the other. It's easier for this to work with column numbers,
which here are 1 through 6, and we'll make a matrix `m` with
the dissimilarities in it, which we have to initialize first. I'll
initialize it to a $6\times 6$ matrix of `-1`, since the final
dissimilarities are 0 or bigger, and this way I'll know if I forgot
anything.

Here's where we are at so far:

```{r eval=F}
fruit_m=matrix(-1,6,6)
for (i in 1:6) {
for (j in 1:6) {
fruit_m[i,j]=3 # dissim between fruit i and fruit j
}
}
```

 

This, of course, doesn't run yet. The sticking point is how to
calculate the dissimilarity between two columns. I think that is a
separate thought process that should be in a function of its own. The
inputs are the two column numbers, and a data frame to get those
columns from:

```{r }
dissim=function(i,j,d) {
x = d %>% select(i)
y = d %>% select(j)
sum(x!=y)
}
dissim(1,2,fruit2)
```

 

Apple and orange differ by one (not being crunchy). The process is:
grab the $i$-th column and call it `x`, grab the $j$-th column
and call it `y`. These are two one-column data frames with four
rows each (the four properties). `x!=y` goes down the rows, and
for each one gives a `TRUE` if they're different and a
`FALSE` if they're the same. So `x!=y` is a collection
of four T-or-F values. This seems backwards, but I was thinking of
what we want to do: we want to count the number of different
ones. Numerically, `TRUE` counts as 1 and `FALSE` as 0,
so we should make the thing we're counting (the different ones) come
out as `TRUE`. To count the number of `TRUE`s (1s), add
them up. 

That was a complicated thought process, so it was probably wise to
write a function to do it. Now, in our loop, we only have to call the
function (having put some thought into getting it right):

```{r }
fruit_m=matrix(-1,6,6)
for (i in 1:6) {
for (j in 1:6) {
fruit_m[i,j]=dissim(i,j,fruit2)
}
}
fruit_m
```

 

The last step is re-associate the fruit names with this matrix. This
is a `matrix` so it has a `rownames` and a
`colnames`. We set both of those, but first we have to get the
fruit names from `fruit2`:

```{r }
fruit_names=names(fruit2)
rownames(fruit_m)=fruit_names
colnames(fruit_m)=fruit_names
fruit_m
```

 

This is good to go into the cluster analysis (happening later).

There is a `tidyverse` way to do this also. It's actually a lot
like the loop way in its conception, but the coding looks
different. We start by making all combinations of the fruit names with
each other, which is `crossing`:

```{r }
combos=crossing(fruit=fruit_names,other=fruit_names)
combos
```

 

Now, we want a function that, given any two fruit *names*, works
out the dissimilarity between them. A happy coincidence is that we can
use the function we had before, *unmodified*! How? Take a look: 

```{r }
dissim=function(i,j,d) {
x = d %>% select(i)
y = d %>% select(j)
sum(x!=y)
}
dissim("Apple","Orange",fruit2)
```

 

`select` can take a column number *or a column name*, so
that running it with column names gives the right answer.

Now, we want to run this function for each of the pairs in
`combos`. The "for each" is `fruit` and `other`
in parallel, so it's `map2` rather than `map`. Also, the
dissimilarity is a whole number each time, so we need
`map2_int`. So we can do this:

```{r }
combos %>% mutate(dissim=map2_int(fruit,other,dissim,fruit2))
```

 

This would work just as well using `fruit1` rather than
`fruit`, since we are picking out the columns by name rather
than number.

To make this into something we can turn into a `dist` object
later, we need to `spread` the column `other` to make a
square array:

```{r }
fruit_spread = combos %>% 
mutate(dissim=map2_int(fruit,other,dissim,fruit2)) %>%
spread(other,dissim)
fruit_spread
```

 

Done!
 

(c) Do a hierarchical cluster analysis using complete
linkage. Display your dendrogram.
 
Solution


First, we need to take one of our matrices of dissimilarities
and turn it into a `dist` object. Since I asked you to
save yours into a file, let's start from there. Mine is aligned
columns: 
```{r }
dissims=read_table("fruits.txt")
dissims
```

      

Then turn it into a `dist` object. The first step is to take
off the first column, since `as.dist` can get the names from
the columns:

```{r }
d = dissims %>% select(-fruit) %>%
as.dist()
d
```



If you forget to take off the first column, this happens:

```{r error=T}
as.dist(dissims)
```

 

The key thing here is "non-square matrix": you have one more column
than you have rows, since you have a column of fruit names.

This one is `as.dist` since you already have dissimilarities
and you want to arrange them into the right type of
thing. `dist` is for *calculating* dissimilarities, which
we did before, so we don't want to do that now.

Now, after all that work, the actual cluster analysis and dendrogram:
```{r hkadh}
fruits.1=hclust(d,method="complete")
plot(fruits.1)
```

     
 

(d) How many clusters, of what fruits, do you seem to have?
Explain briefly.
 
Solution


I reckon I have three clusters: strawberry and blueberry in one,
apple and orange in the second, and banana and pear in the
third. (If your dissimilarities were different from mine, your
dendrogram will be different also.)
 

(e) Pick a pair of clusters (with at least 2 fruits in each)
from your dendrogram. Verify that 
the complete-linkage distance on your dendrogram is correct.
 
Solution


I'll pick strawberry-blueberry and and apple-orange. I'll arrange
the dissimilarities like this:

```

apple   orange
strawberry    3       2
blueberry     3       2

```

The largest of those is 3, so that's the complete-linkage
distance. That's also what the dendrogram says.
(Likewise, the smallest of those is 2, so 2 is the
single-linkage distance.) That is to say, the largest distance or
dissimilarity 
from anything in one cluster to anything in the other is 3, and
the smallest is 2.
I don't mind which pair of clusters you take, as long as you spell
out the dissimilarity (distance) between each fruit in each
cluster, and take the maximum of those. Besides, if your
dissimilarities are different from mine, your complete-linkage
distance could be different from mine also. The grader will have
to use her judgement!
`r tufte::margin_note("That's two cups of coffee I owe the      grader now.")`  
The important point is that you assess the dissimilarities between
fruits in one cluster and fruits in the other. The dissimilarities
between fruits in the same cluster don't enter into it xxx.
`r tufte::margin_note("I      now have a mental image of John Cleese saying *it don't enter      into it* in the infamous Dead Parrot sketch, https://www.youtube.com/watch?v=vnciwwsvNcc}. Not to      mention       *How to defend yourself against an assailant armed with fresh      fruit*, https://www.youtube.com/watch?v=4JgbOkLdRaE.    ")`
As it happens, all my complete-linkage distances between clusters
(of at least 2 fruits) are 3. The single-linkage ones are
different, though:
```{r lhsdjhad}
fruits.2=hclust(d,method="single")
plot(fruits.2)
```

     

All the single-linkage cluster distances are 2. (OK, so this wasn't a
very interesting example, but I wanted to give you one where you could
calculate what was going on.)
 


