##  Making some R functions


 Let's write some simple R functions to convert temperatures,
and later to play with text.



(a) A temperature in Celsius is converted to one in Kelvin by
adding 273.15. (A temperature of $-273.15$ Celsius, 0 Kelvin, is the "absolute zero" 
temperature that nothing can be colder than.) Write a
function called `c_to_k` that converts an input Celsius
temperature to one in Kelvin, and test that it works.

Solution


This is mostly an exercise in structuring your function
correctly. Let's call the input `C` (uppercase C, since
lowercase c has a special meaning to R):
```{r }
c_to_k <- function(C) {
  C + 273.15
}
c_to_k(0)
c_to_k(20)
```

     

This is the simplest way to do it: the last line of the function, if
calculated but not saved, is the value that gets returned to the
outside world. The checks suggest that it worked.

If you're used to Python or similar, you might prefer to calculate the
value to be returned and then return it. You can do that in R too:

```{r }
c_to_k <- function(C) {
  K <- C + 273.15
  return(K)
}
c_to_k(0)
c_to_k(20)
```

 

That works just as well, and for the rest of this question, you can go
either way.^[R style is to use the last line of the function  for the return value, unless you are jumping out of the function  before the end, in which case use `return`.]

$\blacksquare$

(b) Write a function to convert a Fahrenheit temperature to
Celsius. The way you do that is to subtract 32 and then multiply by
$5/9$. 

Solution


On the model of the previous one, we should call this
`f_to_c`. I'm going to return the last line, but you can
save the calculated value and return that instead:
```{r }
f_to_c <- function(F) {
  (F - 32) * 5 / 9
}
f_to_c(32)
f_to_c(50)
f_to_c(68)
```

     

Americans are very good at saying things like "temperatures in the 50s", 
which don't mean much to me, so I like to have benchmarks to
work with: these are the Fahrenheit versions of 0, 10, and 20 Celsius.

Thus "in the 50s" means "between about 10 and 15 Celsius".

$\blacksquare$

(c) *Using the functions you already wrote*, write a function
to convert an input Fahrenheit temperature to Kelvin.

Solution


This implies that you can piggy-back on the functions you just
wrote, which goes as below. First you convert the Fahrenheit to
Celsius, and then you convert *that* to Kelvin. (This is less
error-prone than trying to use algebra to get a formula for this
conversion and then implementing that):
```{r }
f_to_k <- function(F) {
  C <- f_to_c(F)
  K <- c_to_k(C)
  return(K)
}
f_to_k(32)
f_to_k(68)
```

     

These check because in Celsius they are 0 and 20 and we found the
Kelvin equivalents of those to be these values earlier.

I wrote this one with a `return` because I thought it made the
structure clearer: run one function, save the result, run another
function, save the result, then return what you've got.

$\blacksquare$

(d) Rewrite your Fahrenheit-to-Celsius convertor to take a
suitable default value and check that it works as a default.

Solution


You can choose any default you like. I'll take a default of 68
(what I would call "a nice day"):
```{r }
f_to_c <- function(F = 68) {
  (F - 32) * 5 / 9
}
f_to_c(68)
f_to_c()
```

     

The change is in the top line of the function. You see the result: if
we run it without an input, we get the same answer as if the input had
been 68.

$\blacksquare$

(e) What happens if you feed your Fahrenheit-to-Celsius convertor
a *vector* of Fahrenheit temperatures? What if you use it in a
`mutate`? 

Solution


Try it and see:
```{r }
temps <- seq(30, 80, 10)
temps
f_to_c(temps)
```

     

Each of the Fahrenheit temperatures gets converted into a Celsius
one. This is perhaps more useful in a data frame, thus:

```{r }
tibble(temps = seq(30, 80, 10)) %>%
  mutate(celsius = f_to_c(temps))
```

 

All the temperatures are side-by-side with their equivalents.

Here's another way to do the above:

```{r }
temps <- seq(30, 80, 10)
temps %>%
  enframe(value = "fahrenheit") %>%
  mutate(celsius = f_to_c(temps))
```

 

`enframe` creates a two-column data frame out of a vector (like
`temps`). A vector can have "names", in which case they'll be
used as the `name` column; the values will go in a column
called `value` unless you rename it, as I did.

$\blacksquare$

(f) Write another function called `wrap` that takes two
arguments: a piece of text called `text`, which defaults to
`hello`, and another piece of text called `outside`,
which defaults to `*`. The function returns `text`
with the text `outside` placed before and after, so that
calling the function with the defaults should return
`*hello*`. To do this, you can use `str_c` from
`stringr` (loaded with the `tidyverse`) which places
its text arguments side by side and glues them together into one
piece of text.  Test your function briefly.

Solution


This:
```{r }
wrap <- function(text = "hello", outside = "*") {
  str_c(outside, text, outside)
}
```

     

I can run this with the defaults:

```{r }
wrap()
```

 

or with text of my choosing:

```{r }
wrap("goodbye", "_")
```

 

I think that's what I meant by "test briefly".

$\blacksquare$

(g) What happens if you want to change the default
`outside` but use the default for `text`? How do you
make sure that happens? Explore.

Solution


The obvious thing is this, which doesn't work:
```{r }
wrap("!")
```

     

This takes *`text`* to be `!`, and `outside`
to be the default. How do we get `outside` to be `!`
instead? The key is to specify the input by name:

```{r }
wrap(outside = "!")
```

 

This correctly uses the default for `text`.

If you specify inputs without names, they are
taken to be in the order that they appear in the function
definition. As soon as they get out of order, which typically happens
by using the default for something early in the list, as we did here
for `text`, you have to specify names for anything that comes
after that. These are the names you put on the function's top line.

You can always use names:

```{r }
wrap(text = "thing", outside = "**")
```

 

and if you use names, they don't even have to be in order:

```{r }
wrap(outside = "!?", text = "fred")
```

 
$\blacksquare$

(h) What happens if you feed your function `wrap` a vector
for either of its arguments? What about if you use it in a
`mutate`? 

Solution


Let's try:
```{r }
mytext <- c("a", "b", "c")
wrap(text = mytext)
```

     

```{r }
myout <- c("*", "!")
wrap(outside = myout)
```

 

If one of the inputs is a vector, the other one gets "recycled" as
many times as the vector is long. What if they're both vectors?

```{r }
mytext2 <- c("a", "b", "c", "d")
wrap(mytext2, myout)
```

 

This uses the two inputs in parallel, repeating the short one as
needed. But this, though it works, gives a warning:

```{r }
wrap(mytext, myout)
```

 

This is because the shorter vector (of length 2 here) doesn't go
evenly into the longer one (length 3). It gives a warning because this
is probably not what you wanted.

The `mutate` thing is easier, because all the columns in a data
frame have to be the same length. `LETTERS` is a vector with
the uppercase letters in it:

```{r }
tibble(mytext = LETTERS[1:6], myout = c("*", "**", "!", "!!", "_", "__")) %>%
  mutate(newthing = wrap(mytext, myout))
```

$\blacksquare$





