##  Social interaction among old people


 A graduate student wrote a thesis comparing different
treatments for increasing social interaction among geriatric
patients. He recruited 21 patients at a state mental hospital and
randomly assigned them to treatments: Reality Orientation
(`ro`), Behavior Therapy (`bt`) or no treatment
(`ctrl`). Each subject was observed at four times, labelled
`t1` through `t4` in the data file
[link](http://www.utsc.utoronto.ca/~butler/d29/geriatrics.txt). The
response variable was the percentage of time that the subject was
"engaging in the relevant social interaction", so that a higher
value is better.

The principal aim of the study was to see whether there were
differences among the treatments (one would hope that the real
treatments were better than the control one), and whether there were
any patterns over time.



(a) Read in the data and display at least some of it.


Solution


The usual, separated by a single space:

```{r }
my_url <- "http://www.utsc.utoronto.ca/~butler/d29/geriatrics.txt"
geriatrics <- read_delim(my_url, " ")
geriatrics
```

 

Correctly 21 observations measured at 4 different times. We also have
subject numbers, which might be useful later.



(b) Create a response variable and fit a suitable `lm`
as the first stage of the repeated-measures analysis.


Solution


This:
```{r }
response <- with(geriatrics, cbind(t1, t2, t3, t4))
geriatrics.1 <- lm(response ~ treatment, data = geriatrics)
```

     

There is no need to look at this, since we are going to feed it into
`Manova` in a moment, but in case you're curious, you see (in `summary`) a
regression of each of the four columns in `response` on
`treatment`, one by one. 
    


(c) Run a suitable `Manova`. There is some setup
first. Make sure you do that.


Solution


Make sure `car` is loaded, and do the `idata` and
`idesign` thing:
```{r }
times <- colnames(response)
times.df <- data.frame(times=factor(times))
geriatrics.2 <- Manova(geriatrics.1, idata = times.df, idesign = ~times)
```

     

In case you're curious, `response` is an R `matrix`:

```{r }
class(response)
```

 

and
not a data frame (because it was created by `cbind` which makes
a matrix out of vectors). So, to pull the names off the top, we really
do need `colnames` (applied to a matrix) rather than just
`names` (which applies to a data frame). 
    


(d) Display the results of your repeated-measures
analysis. What do you conclude? Explain briefly.


Solution


Just "printing" it will get you what you want:
```{r }
geriatrics.2
```

     

The interaction is significant, so the effects of the treatments are
different at different times. (It makes most sense to say it this way
around, since treatment is something that was controlled and time was not.)

You, I hope, know better than to look at the main effects when there
is a significant interaction!
    


(e) To understand the results that you got from the repeated
measures analysis, you are going to draw a picture (or two). To do
*that*, we are going to need the data in "long" format with
one response value per line (instead of four). Use `pivot_longer`
suitably to get the data in that format, and demonstrate that you
have done so.


Solution


The usual layout:

```{r}
geriatrics %>% 
  pivot_longer(t1:t4, names_to="time", values_to = "intpct") -> geriatrics.long
geriatrics.long
```


I have *one* column of interaction percents, and
*one* column of times. If you check the whole thing, you'll see
that `pivot_longer` gives all the measurements for subject 1, then subject 2, and so on.

The long data frame is, well, long.

It's not necessary to pull out the numeric time values, though you
could if you wanted to, by using
`parse_number`.
    


(f) Calculate and save the mean interaction percents for each
time-treatment combination.


Solution


`group_by` followed by `summarize`, as ever:
```{r }
geriatrics.long %>%
  group_by(treatment, time) %>%
  summarize(mean = mean(intpct)) -> means
means
```

       



(g) Make an interaction plot. Arrange things so that time goes
across the page. Use your data frame of means that you just calculated.


Solution


Once you have the means, this is not too bad:

```{r ger_int}
ggplot(means, aes(x = time, y = mean, group = treatment, colour = treatment)) +
  geom_point() + geom_line()
```

 

The "second factor" `treatment` appears as both
`group` and `colour`.
    


(h) Describe what you see on your interaction plot, and what it
says about why your repeated-measures analysis came out as it did.


Solution


The two "real" treatments `bt` and `ro` both go up
sharply between time 1 and time 2, and then come back down so that
by time 4 they are about where they started. The control group
basically didn't change at all, and if anything went *down*
between times 1 and 2, a completely different pattern to the others.
The two treatments `bt` and `ro` are not exactly
parallel, but they do at least  have qualitatively the same
pattern.
`r tufte::margin_note("That is to say, it's the same kind of shape.")` It
is, I think, the fact that the control group has a
*completely* different pattern over time that makes the
interaction come out significant.
`r tufte::margin_note("I am kind of wrong about    that, as we see.")`
I'm going to explore that some more later, but first I want to get
you to draw a spaghetti plot.
    


(i) Draw a spaghetti plot of these data. That is, use
`ggplot` to plot the interaction percent against time for
each subject,
joining the points for the *same subject* by lines whose colour
shows what treatment they were on. Use the "long" data frame for
this (not the data frame of means).


Solution


This is almost easier to do than it is to ask you to do:

```{r ger_spag}
ggplot(geriatrics.long, aes(x = time, y = intpct, colour = treatment, group = subject)) +
  geom_line()
```

   

The basic difficulty here is to get all the parts. We need both a
`colour` and a `group`; the latter controls the joining
of points by lines (if you have both). Fortunately we already had
subject numbers in the original data; if we had not had them, we would
have had to create them. `dplyr` has a function
`row_number` that we could have used for that; we'd apply the row
numbers to the original wide data frame, before we made it long, so
that the correct subject numbers would get carried along.

Whether you add a `geom_point()` to plot the data points, or not,
is up to you. Logically, it makes sense to include the actual data,
but aesthetically, it looks more like spaghetti if you leave the
points out. Either way is good, as far as I'm concerned.

I didn't ask you to comment on the spaghetti plot, because the story
is much  the same as the interaction plot. There is a lot of
variability, but the story within each group is basically what we
already said: the red lines go sharply up and almost as sharply back
down again, the blue lines do something similar, only not as sharply
up and down, and the green lines do basically nothing.

I said that the control subjects' time pattern was noticeably
different from the others. Which made me think: what if we remove the
control subjects? Would there still be an interaction?
`r tufte::margin_note("This is rather like removing time zero in the example in class.")`

All right, we need to start with the original wide data frame, and
from *that* select everything but `ctrl`:

```{r }
gg <- geriatrics %>% filter(treatment != "ctrl")
gg
```

 

So now there are two treatments left, seven people on
each:
`r tufte::margin_note("The factor *treatment* still has three levels,  but only two of them have any remaining data.")` 

```{r }
gg %>% count(treatment)
```

 

Then we do
the same stuff over again: construct the response, run the
`lm`, create the stuff for `idata` and `idesign`,
and run the `Manova`. There's really nothing new here:

```{r }
attach(gg)
response <- cbind(t1, t2, t3, t4)
gg.1 <- lm(response ~ treatment)
times <- colnames(response)
times.df <- data.frame(times=factor(times))
gg.2 <- Manova(gg.1, idata = times.df, idesign = ~times)
gg.2
detach(gg)
```

 

There is still an interaction, but it's not as significant as it was
before. I think it is still significant because the shape of the two
time trends is not the same: the red `bt` group goes up further
and down further. I was musing that the higher values are also more
variable, which would suggest a transformation, but I haven't explored that.
If the interaction had turned out to be nonsignificant this way? You
might think about trying to remove it from the model, except that in
this kind of model, `treatment` is a "between-subjects factor" 
and `times` is a "within-subjects factor", so they
are different kinds of things. What you do in that case is to ignore
the non-significant interaction and interpret the main effects: there
is no way to "gain df for error" like we did in two-way
ANOVA. Supposing, in this case, that we were using $\alpha=0.01$, we
would say that the interaction is not significant. Then we look at the
main effects: there is no effect of treatment, but there is an effect
of time. Or, to put it another way, once you allow for an effect of
time, there is no difference between the two remaining
treatments.
`r tufte::margin_note("There is often an effect of time, which is why you would be taking multiple time measurements, but the issue is when you take that into account, you are giving yourself an improved chance, in general, to find a treatment effect. This is exactly the same idea as using a matched pairs design to give yourself an improved chance of finding a treatment effect, even though the subjects might be quite different from each other. In fact, repeated measures *is* matched pairs with more than two measurements on each person. Which makes me think, I should have you do a matched pairs as repeated measures.")`

Thinking back to our spaghetti plot, we are now comparing the red and
blue treatments. They both go up at time 2 and down afterwards, which
is the time effect, but even once you allow for this time trend, there is
too much scatter to be able to infer a difference between the treatments.

Extra (maybe branch off into another question?) I was thinking that this is not terribly clear, so I thought I would
fake up some data where there is a treatment effect and a time effect
(but no interaction), and draw a spaghetti plot, so you can see the
difference, idealized somewhat of course. Let's try to come up with
something with the same kind of time effect, up at time 2 and then
down afterwards, that is the same for two drugs A and B. Here's what I
came up with:

```{r }
fake <- read.csv("fake.csv", header = T)
fake
```

 

You can kind of get the idea that the pattern over time is up and then
down, so that it finishes about where it starts, but the numbers for
drug A are usually bigger than the  ones for drug B, consistently over
time. So there ought not to be an interaction, but there ought to be
both a time effect and a drug effect.

Let's see whether we can demonstrate that. First, a spaghetti plot,
which involves getting the data in long format first. I'm saving the
long format to use again later.

```{r }
fake %>% 
  pivot_longer(t1:t4, names_to="times", values_to="score") -> fake.long
fake.long %>%
  ggplot(aes(x = times, y = score, colour = drug, group = subject)) +
  geom_point() + geom_line()
```

 

The reds are consistently higher than the blues (drug effect), the
pattern over time goes up and then down (time effect), but the time
effect is basically the same for both drugs (no interaction).

I got the plot wrong the first time, because I forgot whether I was
doing an interaction plot (where `group=` and `colour=`
are the same) or a spaghetti plot (where `group` has to be
`subject` and the colour represents the treatment, usually). 

Let's do the repeated-measures ANOVA and see whether my guess above is right:

```{r }
response <- with(fake, cbind(t1, t2, t3, t4))
fake.1 <- lm(response ~ drug, data = fake)
times <- colnames(response)
times.df <- data.frame(times=factor(times))
fake.2 <- Manova(fake.1, idata = times.df, idesign = ~times)
```

 

After typing this kind of stuff out a few too many times, I hope
you're getting the idea "function". Also, the construction of the
response is kind of annoying, where you have to list all the time
columns. The trouble is, `response` has to be a `matrix`,
which it is:

```{r }
class(response)
```

 

but if you do the obvious thing of selecting the columns of the data
frame that you want:

```{r }
fake %>% select(t1:t4) -> r
class(r)
```

 

you get a data frame instead. I think this would work:

```{r }
r <- fake %>% select(t1:t4) %>% as.matrix()
class(r)
```

 

The idea is that you select the columns you want as a data frame first
(with `select`), and then turn it into a `matrix` at the
end. 

This is the kind of thing you'd have to do in a function, I think,
since you'd have to have some way of telling the function which are
the "time" columns. Anyway, hope you haven't forgotten what we were
doing:
`r tufte::margin_note("I got sidetracked, surprise surprise.")`

```{r }
fake.2
```

 

The drug-by-time interaction is not (quite) significant, so working at
$\alpha=0.05$ (since that will help me make my point), we go ahead and
interpret the main effects: there is a time effect (the increase at
time 2 that I put in on purpose), and, allowing for the time effect,
there is also a difference between the drugs (because the drug A
scores are a bit higher than the drug B scores).

What if we ignored the time effect? You'd think we could do something
like this, treating the measurements at different times as replicates:

```{r }
head(fake.long)
fake.3 <- aov(score ~ drug, data = fake.long)
summary(fake.3)
```

 

but this would be *wrong*, because we are acting as if we have 40
independent observations, which we don't (this is the point of doing
repeated measures in the first place). It looks as if we have achieved
something by getting a lower P-value for `drug`, but we
haven't really, because we have done so by cheating.

What we could do instead is to average the scores for each subject
over all the times,
`r tufte::margin_note("This would be allowable, since we are  averaging *over* the time-dependence; we are creating 10  independent averages, from the 10 subjects. People do this kind of  thing, instead of having to deal with the repeated measures.")` for
which we go back to the original data frame:

```{r }
fake
fake %>%
  mutate(avg.score = (t1 + t2 + t3 + t4) / 4) %>%
  aov(avg.score ~ drug, data = .) %>%
  summary()
```

 

Ah, now, this is very interesting. I was hoping that by throwing away
the time information (which is useful), we would have diminished
the significance of the drug effect. By failing to include the
time-dependence in our model, we ought to have introduced some extra
variability, which ought to weaken our test. But this test gives
*exactly the same* P-value as the one in the MANOVA, and it looks
like exactly the same test (the $F$-value is the same too). So it
looks as if this is what the MANOVA is doing, to assess the
`drug` effect: it's averaging over the times. Since the same
four (here) time points are being used to compute the average for each
subject, we are comparing like with like at least, and even if there
is a large time effect, I suppose it's going to have the same effect
on each average. For example, if as here the scores at time 2 are
typically highest, all the averages are going to be composed of one
high score and three lower ones. So maybe I have to go back and dilute
my conclusions about the significance of treatments earlier: it's actually
saying that there is a difference between the two remaining treatments
*averaged over time* rather than *allowing for time* as I
said earlier.
    



