##  Handling shipments


 A company called Global Electronics from
time to time imports shipments of a certain large part used as a
component in several of its products. The size of the shipment varies
each time. Each shipment is sent to one of two warehouses (labelled A
and B) for handling. The data in
[link](http://ritsokiguess.site/datafiles/global.csv) show the
`size` of each shipment (in thousands of parts) and the direct
`cost` of handling it, in thousands of dollars. Also shown is
the `warehouse` (A or B) that handled each shipment.



(a) Read the data into R and display your data frame. 


Solution


If you open the data file in your web browser, it will probably
open as a spreadsheet, which is not really very helpful, since
then it is not clear what to do with it. You could, I suppose,
save it and upload it to R Studio Cloud, but it requires much less
brainpower to open it directly from the URL:

```{r }
url <- "http://ritsokiguess.site/datafiles/global.csv"
shipments <- read_csv(url)
```

     

If you display your data frame and it looks like this, you are good
(you can give the data frame any name):

```{r }
shipments
```

 
(b) Describe how many rows and columns your data frame has, and what they contain. 

It has 10 rows and 3 columns. *You need to say this.*

That is, there were 10 shipments recorded, and for each of them, 3
variables were noted: the size and cost of the shipment, and the
warehouse it was handled at.
    
We will also be making some graphs of these data later.