##  Predicting GPA of computer science students


 The file
[link](http://ritsokiguess.site/datafiles/gpa.txt) contains some
measurements of academic achievement for a number of university
students studying computer science:



* High school grade point average

* Math SAT score

* Verbal SAT score

* Computer Science grade point average

* Overall university grade point average.




(a) Read in the data and display it (or at least the first ten lines).


Solution


The usual:

```{r }
my_url <- "http://ritsokiguess.site/datafiles/gpa.txt"
gpa <- read_delim(my_url, " ")
gpa
```

 

Two SAT scores and three GPAs, as promised.



(b) <a name="part:hsu-scatter">*</a> Make a scatterplot of high school GPA against university
GPA. Which variable should be the response and which
explanatory? Explain briefly. Add a smooth trend to your plot.


Solution


High school comes before university, so high school should be
explanatory and university should be the response. (High school
grades are used as an admission criterion to university, so we
would hope they would have some predictive value.)
```{r }
ggplot(gpa, aes(x = high_GPA, y = univ_GPA)) + geom_point() +
  geom_smooth()
```

   
    


(c) Describe any relationship on your scatterplot: its direction, its
strength and its shape. Justify your description briefly.


Solution


Taking these points one at a time:


* direction: upward (a higher high-school GPA generally goes
with a higher university GPA as well. Or you can say that the
lowest high-school GPAs go with the lowest university GPAs,
and high with high, at least most of the time).

* strength: something like moderately strong, since while
the trend is upward, there is quite a lot of scatter. (This is
a judgement call: something that indicates that you are basing
your description on something reasonable is fine.)

* shape: I'd call this "approximately linear" since there
is no clear curve here. The smooth trend wiggles a bit, but
not enough to make me doubt a straight line.

Looking ahead, I also notice that when high-school GPA is high,
university GPA is also consistently high, but when high-school
GPA is low, the university GPA is sometimes low and sometimes
high, a lot more variable. (This suggests problems with fan-in
later.) In a practical sense, what this seems to show is that
people who do well at university usually did well in high-school
as well, but sometimes their high-school grades were not
that good. This is especially true for people with university
GPAs around 3.25.
      


(d) <a name="part:highonly">*</a> Fit a linear regression for predicting university GPA
from high-school GPA and display the results.


Solution


Just this, therefore:
```{r }
gpa.1 <- lm(univ_GPA ~ high_GPA, data = gpa)
summary(gpa.1)
```

    

Extra: this question goes on too long, so I didn't ask you to look at the
residuals from this model, but my comments earlier suggested that we
might have had some problems with fanning-in (the variability of
predictions getting less as the high-school GPA increases). In case
you are interested, I'll look at this here. First, residuals against
fitted values:

```{r }
ggplot(gpa.1, aes(x = .fitted, y = .resid)) + geom_point() + geom_smooth()
```

 

Is that evidence of a trend in the residuals? Dunno. I'm inclined to
call this an "inconsequential wiggle" and say it's OK. Note that the
grey envelope includes zero all the way across.

Normal quantile plot of residuals:

```{r }
ggplot(gpa.1, aes(sample = .resid)) + stat_qq() + stat_qq_line()
```

 

A somewhat long-tailed distribution: compared to a normal
distribution, the residuals are a bit too big in size, both on the
positive and negative end.

The problem I was really worried about was the potential of
fanning-in, which we can assess by looking at the absolute residuals:

```{r }
ggplot(gpa.1, aes(x = .fitted, y = abs(.resid))) + geom_point() + geom_smooth()
```

 

That is definitely a downward trend in the size of the residuals, and
I think I was right to be worried before. The residuals should be of
similar size all the way across.

The usual problem of this kind is fanning-*out*, where the
residuals get *bigger* in size as the fitted values increase. The
bigger values equals more spread is the kind of thing that a
transformation like taking logs will handle: the bigger values are all
brought downwards, so they will be both smaller and less variable than
they were. This one, though, goes the other way, so the only kind of
transformation that might help is one at the other end of the scale
(think of the Box-Cox lambda scale), something like maybe reciprocal,
$\lambda=-1$ maybe.

The other thought I had was that there is this kind of break around a
high-school GPA of 3 (go back to the scatterplot of (<a href="#part:hsu-scatter">here</a>)): when the
high-school GPA is higher than 3, the university GPA is very
consistent (and shows a clear upward trend), but when the high-school
GPA is less than 3, the university GPA is very variable and there
doesn't seem to be any trend at all. So maybe two separate analyses
would be the way to go.

All right, how does Box-Cox work out here, if at all?

```{r }
library(MASS)
boxcox(univ_GPA ~ high_GPA, data = gpa)
```

 

It doesn't. All right, that answers *that* question. 

When I loaded `MASS`, I also loaded its `select`
function,
`r tufte::margin_note("That is what that masked message above was  about.")` and I might want to use the 
`tidyverse` `select`
function later, and things could get confused. So let's "unload"
`MASS` now:

```{r }
detach("package:MASS", unload = T)
```

 
        


(e) Two students have been admitted to university. One has
a high school GPA of 3.0 and the other a high school GPA of   
3.5. Obtain suitable intervals that summarize the GPAs that each of these
two students might obtain in university.


Solution


Since we are talking about individual students, rather than
the  mean of *all* students with these GPAs, prediction
intervals are called for. The first step is to make a data
frame to predict from. This has to contain columns for all the
explanatory variables, just `high_GPA` in this case:
```{r }
gpa.new <- tibble(high_GPA = c(3, 3.5))
gpa.new
```

      

and then feed that into `predict`:

```{r }
preds <- predict(gpa.1, gpa.new, interval = "p")
preds
```

 

and display that side by side with the values it was predicted from:

```{r }
cbind(gpa.new, preds)
```

 

or this way, if you like it better:

```{r }
as_tibble(preds) %>% bind_cols(gpa.new) %>% select(high_GPA, everything())
```

 

Thus the predicted university GPA for the student with high school GPA
3.0 is between 2.6 and 3.7, and for the student with high school GPA
3.5 is between 2.9 and 4.0. (I think this is a good number of decimals
to give, but in any case, you should actually *say* what the
intervals are.)

I observe that these intervals are almost exactly the same
length. This surprises me a bit, since I would have said that 3.0 is
close to the average high-school GPA and 3.5 is noticeably higher. If
that's the case, the prediction interval for 3.5 should be longer
(since there is less "nearby data"). Was I right about that?

```{r }
gpa %>% summarize(
  mean = mean(high_GPA),
  med = median(high_GPA),
  q1 = quantile(high_GPA, 0.25),
  q3 = quantile(high_GPA, 0.75)
)
```

 

More or less: the mean is close to 3, and 3.5 is close to the third
quartile. So the thing about the length of the prediction interval is
a bit of a mystery. Maybe it works better for the confidence interval
for the mean:

```{r }
preds <- predict(gpa.1, gpa.new, interval = "c")
cbind(gpa.new, preds)
```

 

These intervals are a lot shorter, since we are talking about
*all* students with the high-school GPAs in question, and we
therefore no longer have to worry about variation from student to
student (which is considerable). But my supposition about length is
now correct: the interval for 3.5, which is further from the mean, is
a little longer than the interval for 3.0. Thinking about it, it seems
that the individual-to-individual variation, which is large, is
dominating things for our prediction interval above.
        


(f) <a name="part:all">*</a> Now obtain a regression predicting university GPA from
high-school GPA as well as the two SAT scores. Display your results.


Solution


Create a new regression with all the explanatory variables you want in it:

```{r }
gpa.2 <- lm(univ_GPA ~ high_GPA + math_SAT + verb_SAT, data = gpa)
summary(gpa.2)
```

 
        


(g) Test whether adding the two SAT scores has improved the
prediction of university GPA. What do you conclude?



Solution


Since we added *two* explanatory variables, the $t$-tests in
`gpa.2` don't apply (they tell us whether we can take out
*one* $x$-variable). We might have some suspicions, but that's
all they are.  So we have to do `anova`:
```{r }
anova(gpa.1, gpa.2)
```

   

If you put the models the other way around, you'll get a negative
$F$-statistic and degrees of freedom, which doesn't make
much sense (although the test will still work).

The null hypothesis here is that the two models fit equally
well. Since the P-value is not small, we do not reject that null
hypothesis, and therefore we conclude that the two models *do*
fit equally well, and therefore we prefer the smaller one, the one
that predicts university GPA from just high-school GPA. (Or,
equivalently, we conclude that those two SAT scores don't add anything
to the prediction of how well a student will do at university, once
you know their high-school GPA.)

This might surprise you, given what the SATs are supposed to be
*for*. But that's what the data say.
  


(h) Carry out a backward elimination starting out from your
model in part (<a href="#part:all">here</a>). Which model do you end up with?
Is it the same model as you fit in (<a href="#part:highonly">here</a>)?


Solution


In the model of (<a href="#part:all">here</a>), `math_SAT` was the
least significant, so that comes out first. (I use
`update` but I'm not insisting that you do:)
```{r }
gpa.3 <- update(gpa.2, . ~ . - math_SAT)
summary(gpa.3)
```

 
Here is where we have to stop, since both high-school GPA and
verbal SAT score are significant, and so taking either of them
out would be a bad idea. 
This is a *different* model than the one of
(<a href="#part:highonly">here</a>). This is the case, even though the model
with high-school GPA only was not significantly worse than the
model containing everything. (This goes to show that
model-building doesn't always have nice clear answers.)
In the model I called `gpa.2`, neither of the SAT
scores were anywhere near significant (considered singly), but
as soon as we took out one of the SAT scores (my model
`gpa.3`), the other one became significant. This goes
to show that you shouldn't take out more than one explanatory
variable based on the results of the $t$-tests, and even if
you test to see whether you should have taken out both of the SAT,
you won't necessarily get consistent
results. Admittedly, it's a close decision whether to
keep or remove `verb_SAT`, since its P-value is
close to 0.05.
The other way of tackling this one is via `step`, which
does the backward elimination for you (not that it was much
work here):
```{r }
step(gpa.2, direction = "backward", test = "F")
```

         

This gives the same result as we did from our backward
elimination. The tables with AIC in them are each step of
the elimination, and the variable at the top is the one that comes out
next. (When `<none>` gets to the top, that's when you stop.)
What happened is that the two SAT scores started off highest, but once
we removed `math_SAT`, `verb_SAT` jumped below
`<none>` and so the verbal SAT score had to stay.

Both the P-value and the AIC say that the decision about keeping or
removing `verb_SAT` is very close.
        


(i) These students were studying computer science at
university. Do you find your backward-elimination result
sensible or surprising, given this? Explain briefly.


Solution


I would expect computer science students to be strong students
generally, good at math and possibly not so good with
words. So it is not surprising that high-school GPA figures
into the prediction, but I would expect math SAT score to have
an impact also, and it does not. It is also rather surprising
that verbal SAT score predicts success at university for these
computer science students; you wouldn't think that having
better skills with words would be helpful. So I'm surprised.
Here, I'm looking for some kind of discussion about what's in
the final backward-elimination model, and what you would
expect to be true of computer science students. 
Let's look at the final model from the backward elimination again:
```{r }
summary(gpa.3)
```

         

The two slope estimates are both positive, meaning that, all else
equal, a higher value for each explanatory variable goes with a higher
university GPA. This indicates that a higher verbal SAT score goes
with a higher university GPA: this is across all the university
courses that a student takes, which you would expect to be math and
computer science courses for a Comp Sci student, but might include
a few electives or writing courses. Maybe what is happening is that
the math SAT score is telling the same story as the high-school GPA
for these students, and the verbal SAT score is saying something
different. (For example, prospective computer science students are
mostly likely to have high math SAT scores, so there's not much
information there.)

I think I need to look at the correlations:

```{r }
cor(gpa)
```

 

We'll ignore `comp_GPA` (i) because we haven't been thinking
about it and (ii) because it's highly correlated with the university
GPA anyway. (There isn't a sense that one of the two university GPAs
is explanatory and the other is a response, since students are taking
the courses that contribute to them at the same time.)

The highest correlation with university GPA of what remains is
high-school GPA, so it's not at all surprising that this features in
all our regressions. The correlations between university GPA and the
two SAT scores are about equal, so there appears to be no reason to
favour one of the SAT scores over the other. But, the two SAT scores
are highly correlated with *each other* (0.835), which suggests
that if you have one, you don't need the other, because they are
telling more or less the same story. 

That makes me wonder how a regression with the SAT math score and
*not* the SAT verbal score would look:

```{r }
gpa.4 <- lm(univ_GPA ~ high_GPA + math_SAT, data = gpa)
summary(gpa.4)
```

 

Math SAT does not quite significantly add anything to the prediction,
which confirms that we do better to use the verbal SAT score
(surprising though it seems). Though, the two regressions with the
single SAT scores, `gpa.3` and `gpa.4`, have almost the
same R-squared values. It's not clear-cut at all. In the end, you have
to make a call and stand by it.
        




