
## Watching TV and education

 The General Social Survey is a large survey of a large number of people. One of the questions on the survey is "how many hours of TV do you watch in a typical day?" Another is "what is your highest level of education attained", on this scale:

- **HSorLess**: completed no more than high h school
- **College**: completed some form of college, either a community college (like Centennial) or a four-year university (like UTSC)
- **Graduate**: completed a graduate degree such as an MSc.

Do people with more education tend to watch more TV? We will be exploring this. The data are in \url{http://ritsokiguess.site/datafiles/gss_tv.csv}.



(a) Read in and display (some of) the data.

Solution


Exactly the usual:

```{r}
my_url <- "http://ritsokiguess.site/datafiles/gss_tv.csv"
gss <- read_csv(my_url)
gss
```


$\blacksquare$


(b) For each level of education, obtain the number of observations, the mean and the median of the number of hours of TV watched.

Solution


`group_by` and `summarize`, using `n()` to get the number of observations (rather than `count` because you want some numerical summaries as well):

```{r}
gss %>% group_by(degree) %>% 
summarise(n=n(), mean=mean(tvhours), med=median(tvhours))
```



$\blacksquare$


(c) What does your answer to the previous part tell you about the shapes of the distributions of the numbers of hours of TV watched? Explain briefly. 

Solution


In each of the three groups, the mean is greater than the median, so I think the distributions are skewed to the right. Alternatively, you could say that you expect to see some outliers at the upper end.


$\blacksquare$


(d) Obtain a suitable graph of your data frame.

Solution


One quantitative variable and one categorical one, so a boxplot. (I hope you are getting the hang of this by now.)

```{r}
ggplot(gss, aes(x=degree, y=tvhours)) + geom_boxplot()
```


$\blacksquare$


(e) Does your plot indicate that your guess about the distribution shape was correct? Explain briefly.

Solution


I guessed before that the distributions would be right-skewed, and they indeed are, with the long upper tails. Or, if you suspected upper outliers, they are here as well.

Say what you guessed before, and how your graph confirms it (or doesn't, if it doesn't.)


$\blacksquare$


(f) Run a suitable test to compare the average number of hours of TV watched for people with each amount of education. ("Average" could be mean or median, whichever you think is appropriate.)

Solution


From the boxplot, the distributions are definitely not all normal; in fact, none of them are. So we should use Mood's median test, thus:

```{r}
median_test(gss, tvhours, degree)
```


$\blacksquare$


(g) What do you conclude from your test, in the context of the data?

Solution


The P-value of $6.6\times 10^{-13}$ is extremely small, so we conclude that not all of the education groups watch the same median amount of TV. 
Or, there are differences in the median amount of TV watched among the three groups.

An answer of "the education groups are different" is *wrong*, because you don't know that they are *all* different. It might be that some of them are different and some of them are the same. The next part gets into that.


$\blacksquare$


(h) Why might you now want to run some kind of follow-up test? Run the appropriate thing and explain briefly what you conclude from it, in the context of the data.

Solution


The overall Mood test is significant, so there are some differences between the education groups, but we don't know where they are.  Pairwise median tests will reveal where any differences are:

```{r}
pairwise_median_test(gss, tvhours, degree)
```

The people whose education is high school or less are significantly different from the other two education levels. The boxplot reveals that this is because they watch *more* TV on average. The college and graduate groups are not significantly different (in median TV watching).

Extra 1:

You might have been surprised that the College and Graduate medians were not significantly different. After all, they look quite different on the boxplot. Indeed, the P-value for comparing just those two groups is 0.0512, only just over 0.05. But remember that we are doing three tests at once, so the Bonferroni adjustment is to multiply the P-values by 3, so this P-value is "really" some way from being significant. 
I thought I would investigate this in more detail:

```{r}
gss %>% filter(degree != "HSorLess") %>% 
median_test(tvhours, degree)
```

The College group are about 50-50 above and below the overall median, but the Graduate group are two-thirds below. This suggests that the Graduate group watches less TV, and with these sample sizes I would have expected a smaller P-value. But it didn't come out that way. 

You might also be concerned that there are in total more values below the grand median (106) than above (only 85). This must mean that there are a lot of data values *equal* to the grand median:

```{r}
gss %>% filter(degree != "HSorLess") -> gss1
gss1 %>% summarize(med=median(tvhours))
```

and

```{r}
gss1 %>% count(tvhours)
```

Everybody gave a whole number of hours, and there are not too many different ones; in addition, a lot of them are equal to the grand median of 2.

Extra 2:

Regular ANOVA and Welch ANOVA should be non-starters here because of the non-normality, but you might be curious about how they would perform:

```{r}
gss.1 <- aov(tvhours~degree, data=gss)
summary(gss.1)
TukeyHSD(gss.1)
```

and

```{r}
oneway.test(tvhours~degree, data=gss)
gamesHowellTest(tvhours~factor(degree), data=gss)
```

The conclusions are actually identical to our Mood test, and the P-values are actually not all that much different. Which makes me wonder just how bad the sampling distributions of the sample means are. Bootstrap to the rescue:

```{r}
gss %>% nest(d=c(tvhours)) %>% 
mutate(samp=map(d, ~rerun(1000, sample(.$tvhours, replace = T)))) %>% 
mutate(means=map(samp, ~map_dbl(., ~mean(.x)))) %>% 
unnest(means) %>% 
ggplot(aes(x=means)) + geom_histogram(bins=30) + 
facet_wrap(~degree)
```

Coding this made my head hurt, but building it one line at a time, I pretty much got it right first time. In words:

- get a separate data frame `d` for each education level (with just the data for that one)
- create a column called `samp` with 1000 bootstrap samples of `tvhours` (for each education level)
- calculate the mean of each bootstrap sample (thus 1000 of those for each education level)
- display each bootstrapped mean along with its education level
- make a histogram of each.

I had to use a lot of bins because `bins` is a total and each sampling distribution has a different mean. So each histogram has something like 10 bins.

Anyway, even though the data was very much not normal, these sampling distributions are very normal-looking, suggesting that something like Welch ANOVA would have been not nearly as bad as you would have guessed. This is evidently because of the big sample sizes. (This also explains why the two other flavours of ANOVA gave results very similar to Mood's median test.)


$\blacksquare$







