##  Go away and don't come back!


 When a person has a heart attack and survives it, the major
concern of health professionals is to prevent the person having a
second heart attack. Two factors that are believed to be important are
anger and anxiety; if a heart attack survivor tends to be angry or
anxious, they are believed to put themselves at increased risk of a
second heart attack.

Twenty heart attack survivors took part in a study. Each one was given
a test designed to assess their anxiety (a higher score on the test
indicates higher anxiety), and some of the survivors took an anger
management course. 
The data are in
[link](http://ritsokiguess.site/datafiles/ha2.txt); `y` and
`n` denote "yes" and "no" respectively. The columns denote
(i) whether or not the person had a second heart attack, (ii) whether
or not the person took the anger management class, (iii) the anxiety
score. 



(a) Read in and display the data.

Solution


```{r }
my_url <- "http://ritsokiguess.site/datafiles/ha2.txt"
ha <- read_delim(my_url, " ")
ha
```

     

The anxiety scores are numbers; the other two variables are "yes"
and "no", which makes perfect sense.

$\blacksquare$

(b) <a name="part:fit">*</a> 
Fit a logistic regression predicting whether or not a heart attack
survivor has a second heart attack, as it depends on anxiety score
and whether or not the person took the anger management
class. Display the results.

Solution


This:

```{r }
ha.1 <- glm(factor(second) ~ anger + anxiety, family = "binomial", data = ha)
summary(ha.1)
```

 

There's no need to worry about two-column responses here, because each
row of the data frame refers to only one person, and the response
variable `second` is already a categorical variable with two
categories (that needs to be turned into a `factor` for
`glm`). 

$\blacksquare$

(c) In the previous part, how can you tell that you were
predicting the probability of having a second heart attack (as
opposed to the probability of not having one)?

Solution


The levels of a factor are taken in alphabetical order, with
`n` as the baseline, so we are predicting the probability
of the second one `y`.

$\blacksquare$

(d) <a name="part:preds">*</a>
For the two possible values `y` and `n` of
`anger` and the anxiety scores 40, 50 and 60, make a data
frame containing all six combinations, and use it to obtain
predicted probabilities of a second heart attack. Display your
predicted probabilities side by side with what they are predictions
for.

Solution


This time, I give you the values I want the predictions for (as
opposed to calculating something like quartiles from the data), so
we might as well just type them in.
Step 1 is to save them (my preference is under plural names):
```{r }
anxieties <- c(40, 50, 60)
angers <- c("y", "n")
```

     

Step 2 is to make a data frame of combinations using
`crossing`. I'll call mine `new`:

```{r }
new <- crossing(anxiety = anxieties, anger = angers)
new
```

 

Step 3 is to do the predictions. Into `predict` go (in order)
the fitted model, the new data frame that we just made, and something
to make the predictions be probabilities:

```{r }
p <- predict(ha.1, new, type = "response")
```

 

Step 4 is to put these next to the data frame we
created. `cbind` is easiest, since we don't know what kind of
thing `p` might be, and `bind_cols` is pickier:

```{r }
cbind(new, p)
```

 
$\blacksquare$

(e) Use your predictions from the previous part to describe the
effect of changes in `anxiety` and `anger` on the
probability of a second heart attack.

Solution


The idea is to change one variable \emph{while leaving the other
fixed}. (This is the common refrain of "all else equal".)

Pick a level of `anger`, say `n` (it doesn't matter
which) and look at the effect of `anxiety`. The probability
of a second heart attack increases sharply from 0.17 to 0.40 to
0.69. So an increased anxiety score is associated with an
increased probability of second heart attack (all else equal).

To assess the effect of taking the anger management course, pick
an `anxiety` value, say 40, and compare the probabilities
for `anger` `n` and `y`. For someone who has
not taken the anger management course, the probability is 0.17,
but for someone who has, it drops all the way to 0.07. (The
pattern is the same at the other anxiety scores.)

Extra: the reason it doesn't matter what value of the other
variable you look at (as long as you keep it fixed) is that the
model is "additive" with no interaction, so that the effect of
one variable does not depend on the value of the other one. If we
wanted to see whether the effect of anxiety was different
according to whether or not the person had taken the anger
management course, we would add an interaction between
`anxiety` and `anger`. But we won't be doing this
kind of thing until we get to analysis of variance, so you don't
need to worry about it yet.

$\blacksquare$

(f) Are the effects you described in the previous part
consistent with the `summary` output from `glm` that
you obtained in (<a href="#part:fit">here</a>)? Explain briefly how they are, or
are not. (You need an explanation for each of `anxiety` and
`anger`, and you will probably get confused if you look at
the P-values, so don't.)

Solution


In the previous part, we found that increased anxiety went with an
increased probability of second heart attack. Back in
(<a href="#part:fit">here</a>), we found a positive slope of 0.11904 for anxiety,
which also means that a higher anxiety score goes with a higher
probability of second heart attack.
That was not too hard. The other one is a little more work.
`anger` is categorical with two categories `n` and
`y`. The first one, `n`, is the baseline, so
`angery` shows how `y` compares to `n`. The
slope $-1.04211$ is negative, which means that someone who has
taken the anger management course has a *lower* probability
of a second heart attack than someone who hasn't (all else
equal). This is the same story that we got from the predictions. 
That's it for that, but I suppose I should talk about those
P-values. 
`anxiety` is significant, so it definitely has an effect on
the probability of a second heart attack. The pattern is clear
enough even with this small data set. Here's a visual:
```{r }
ggplot(ha, aes(x = second, y = anxiety)) + geom_boxplot()
```

     

This is the wrong way around in terms of cause and effect, but it
shows pretty clearly that people who suffer a second heart attack have
a higher level of anxiety than those who don't.

That's clear enough, but what about `anger`? That is *not*
significant, but there seems to be a visible effect of `anger`
on the predicted probabilities of (<a href="#part:preds">here</a>): there we saw
that if you had done the anger management course, your probability of
a second heart attack was lower. But that's only the predicted
probability, and there is also uncertainty about that, probably quite
a lot because we don't have much data. So if we were to think about
confidence intervals for the predicted probabilities, they would be
wide, and for the two levels of `anger` at a fixed
`anxiety` they would almost certainly overlap.

Another way of seeing that is a visual, which would be a side-by-side
bar chart:

```{r }
ggplot(ha, aes(x = anger, fill = second)) + geom_bar(position = "dodge")
```

 

A small majority of people who took the anger management did not have
a second heart attack, while a small minority of those who did not,
did.
`r tufte::margin_note("Read that carefully.")` But with these small numbers, the
difference, even though it points the way we would have guessed, is
not large enough to be significant:

```{r }
with(ha, table(anger, second))
```

 

This is not nearly far enough from an equal split to be
significant. (It's the same kind of idea as for Mood's median test in
C32/C33.)

$\blacksquare$



