## Handspans of males and females

 Take your right hand, and stretch the fingers out as far as
you can. The distance between the tip of your thumb and the tip of
your little (pinky) finger is your handspan. The students in a
Statistics class at Penn State measured their handspans and also
whether they identified as male or female. The data are at [http://ritsokiguess.site/datafiles/handspan.txt](http://ritsokiguess.site/datafiles/handspan.txt),
with handspans
measured in inches.  Thinking of these as a random sample of all
possible students, is it true that males have a larger mean handspan
than females? This is what we will explore.


(a) Read in and display (some of) the data.

Solution

This is a delimited (by spaces) file, so:

```{r}
library(tidyverse)
my_url <- "http://ritsokiguess.site/datafiles/handspan.txt"
span <- read_delim(my_url, " ")
span
```     


$\blacksquare$


(b) Make a suitable graph of the two columns.

Solution

One quantitative variable and one categorical one, so a boxplot:

```{r}
ggplot(span, aes(x=sex, y=handspan)) + geom_boxplot()
```     

$\blacksquare$


(c) Run a suitable two-sample $t$-test to address the question
of interest. What do you conclude, in the context of the data?

Solution

We are trying to show that males have a *larger* mean
handspan, so we need an `alternative`. To see which: there
are two sexes, F and M in that order, and we are trying to show
that F is less than M:

```{r}
t.test(handspan~sex, data=span, alternative="less")
```     

The P-value is very small, so there is no doubt that males have larger
average handspans than females.

$\blacksquare$


(d) Obtain a 90\% confidence interval for the difference
in mean handspan between males and females. Do you need to run
any more code? Explain briefly.

Solution

A confidence interval is two-sided, so we have to re-run the test
without the \texttt{alternative} to make it two-sided. Note also that
we need a 90\% interval, which is different from the default 95\%, so
we have to ask for that too:

```{r}
t.test(handspan~sex, data=span, conf.level=0.90)
``` 

The interval is $-2.93$ to $-2.15$, *which you should say*. It
would be even better to say that males have a mean handspan between
2.15 and 2.93 inches larger than that of females. You also need to round off your answer: the data are 
given to 0 or 1 decimals, so your interval should be given to 1 or 2 decimals (since the confidence interval is for a mean). 

On a question like this, the grader is looking for three things: 

- getting the output
- saying what the interval is
- rounding it to a suitable number of decimals. 

Thus, getting the output alone is only one out of three things.

$\blacksquare$


(e) Explain briefly why you might have some concerns about the
validity of the $t$-tests you ran in this question. Or, if you don't have any concerns, explain briefly why that is.

Solution

The major assumption here is that the male and female handspans
have (approximate) normal distributions. The boxplots we drew
earlier both had low-end outliers, so the normality is
questionable. 

Also, say something about the sample sizes and whether or not you think they are large enough to be helpful.

How big are our sample sizes?

```{r}
span %>% count(sex)
```


My suspicion is that we are saved by two things: the sample sizes
are large enough for the central limit theorem to help us, and in
any case, the conclusion is so clear that the assumptions can
afford to be off by a bit. 


Extra: one way to think about whether we should be concerned about
the lack of normality is to use the `bootstrap` to see what
the sampling distribution of the sample mean might look like for males and for females. (This is the stuff in Lecture 5a.) The
way this works is to sample from each distribution *with
replacement*, work out the mean of each sample, then repeat many
times, once for the females and once for the males.

To start with the females, the first thing to do is to grab *only* the rows containing the females. This, using an idea from Lecture 5a that we see again properly later, is `filter`:

```{r}
span %>% filter(sex=="F") -> females
females
```

There are 103 females. From these we need to take a "large" number of bootstrap samples to get a sense of how the mean handspan of the females varies:

```{r}
set.seed(457299)
tibble(sim = 1:1000) %>% 
  rowwise() %>% 
  mutate(the_sample = list(sample(females$handspan, replace = TRUE))) %>% 
  mutate(the_mean = mean(the_sample)) -> d
```



Then we make a histogram of the bootstrap sampling distribution of the sample mean for the females:

```{r}
ggplot(d, aes(x = the_mean)) + geom_histogram(bins = 10)
```

I don't know what you think of this. There are a few more extreme values than I would like, and
it looks otherwise a bit left-skewed to me. But maybe I  am worrying too much.

The males one works exactly the same way:

```{r}
span %>% filter(sex=="M") -> males
tibble(sim = 1:1000) %>% 
  rowwise() %>% 
  mutate(the_sample = list(sample(males$handspan, replace = TRUE))) %>% 
  mutate(the_mean = mean(the_sample)) -> d
```

```{r}
ggplot(d, aes(x = the_mean)) + geom_histogram(bins = 10)
```

There is a similar story here. I think these are good enough overall, and so I am happy with the two-sample $t$-test, but it is not as clear-cut as I was expecting.



$\blacksquare$




