

 A research lab is developing a new compound to relieve
severe cases of hay fever. There were two active ingredients, labelled
A and B, with each one having three levels, low, medium and
high. There were thus $3\times 3=9$ treatment combinations. 36
subjects were available, and these were randomly assigned to
combinations of levels of factor A and factor B, so that each
combination was tested on 4 people. The response variable was the
number of hours of pain relief (so a higher value is better). The data
are in ****.



(a)[2] Read the data and display its structure. Verify that you
have what you were expecting.

Solution


```{r }
hayfever=read.table("hayfever.txt",header=T)
hayfever
```

     


(b)[2] interaction plot

Solution


```{r }
hayfever %>% group_by(a,b) %>%
 summarize(relief.mean=mean(relief)) %>%
 ggplot(aes(x=a,y=relief.mean,colour=b,group=b))+
geom_point()+geom_line()
```

     


(c)[2] Run an analysis of variance with interaction.

Solution


```{r }
hayfever.1=aov(relief~a*b,data=hayfever)
summary(hayfever.1)
```

     


(d)[2] simple effects

