##  Treating hay fever


 Hay fever is an allergic reaction to things like grasses or
pollen which can make it unpleasant to go outside when the weather is
at its best.  A research lab is developing a new compound to relieve
severe cases of hay fever. There were two active ingredients, labelled
A and B, with each one having three levels, low, medium and
high. There were thus $3\times 3=9$ treatment combinations. 36
subjects were available, and these were randomly assigned to
combinations of levels of factor A and factor B, so that each
combination was tested on 4 people. The response variable was the
number of hours of relief from symptoms (so a higher value is
better). The data are in
[link](http://ritsokiguess.site/datafiles/hayfever.txt).



(a) Read the data and display its structure. Verify that you
have what you were expecting.

Solution


Aligned columns separated by spaces, so `read_table`:
```{r }
my_url <- "http://ritsokiguess.site/datafiles/hayfever.txt"
hayfever <- read_table(my_url)
hayfever
```

     

I have 36 observations (patients). There are two categorical columns
`a` and `b` corresponding to the two active ingredients,
and they each seem to have levels low, medium and high.
`r tufte::margin_note("It's important to be clear about the distinction between a categorical variable, that lives in a data frame column, and its levels, the values that appear in the column. This is especially important if you're trying to decide whether a data frame is tidy, since typically an untidy data frame will have factor levels as column names rather than the factor itself, and you need to be able to tell the difference.")`

The `replicate` column labels each observation *within*
its A-B combination, so that each treatment combination was indeed
replicated four times. We won't be using this column in our analysis;
I think it's a refugee from the original untidy format the data was in
before it came to you.


(b) Calculate the mean hours of relief for each combination of
levels of the two active ingredients. Save your results in a data
frame and display that.

Solution


This is a group-by and summarize, but there are two active ingredients and they *both* have to go in the group-by:
```{r }
hayfever %>%
  group_by(a, b) %>%
  summarize(m = mean(relief)) -> d
d
```



I'm using my usual name `d` for a temporary data frame. I could
have put brackets around my whole pipeline to display its result, but
I still need to save the data frame `d` to use in a moment.

These levels are in the wrong logical order, but they are in the right order in the original data frame, so we can use `fct_inorder` first, thus:

```{r }
hayfever %>%
  mutate(a = fct_inorder(a), b = fct_inorder(b)) %>%
  group_by(a, b) %>%
  summarize(m = mean(relief)) -> d2
d2
```

 


(c) Make an interaction plot, showing how the mean pain relief depends on the combination of levels of A and B. It is probably easiest to use the data frame you obtained in the previous part.

Solution


My column of mean relief values is called `m`. Use whatever
name you gave it.  I'm going to use my proper-order data frame for
this:
```{r }
ggplot(d2, aes(x = a, y = m, colour = b, group = b)) +
  geom_point() + geom_line()
```

     
Or, you probably had this:
```{r }
ggplot(d, aes(x = a, y = m, colour = b, group = b)) +
  geom_point() + geom_line()
```

     

Since `a` and `b` both have three levels, you could just as well use them the other way around:

```{r }
ggplot(d2, aes(x = b, y = m, colour = a, group = a)) +
  geom_point() + geom_line()
```

 

The plot looks different depending on how you draw it, but the
conclusion from it (below) will be the same.


(d) What do you conclude from your interaction plot? Explain briefly.

Solution


The three lines are not particularly close to being parallel, so I
would expect to see an interaction between the two active
ingredients (that is, the number of hours of pain relief depends
on the combination of the two of them).

Extra: It is always a judgment call to decide whether the lines
are "approximately parallel" or not. It depends on how much
variability there is in the data, which the interaction plot
doesn't show. Another idea is to add the data to the interaction
plot, colour-coded in the same way. I would do this by taking out
the `geom_point` for the means and add one instead for the
data, taken from the original data frame:
```{r }
ggplot(d2, aes(x = a, y = m, colour = b, group = b)) +
  geom_line() +
  geom_point(data = hayfever, aes(y = relief))
```

     

Technique: for that last `geom_point`, put in anything that
changes: a new data frame, and a new `y` for the plot, but the
`x` and `colour` and `group` are the same as they
were before, so I don't need to specify them.

The points are *very* close to the lines, so there is almost no
residual variability. This makes it more convincing that the
interaction is real and will be significant.


(e) Run an analysis of variance with interaction. What do you conclude?

Solution


```{r }
hayfever.1 <- aov(relief ~ a * b, data = hayfever)
summary(hayfever.1)
```

     

The interaction is (strongly) significant. The hours of relief depend on the combination of levels of the active ingredients A and B.

Don't try to interpret the main effects of A and B from here. That's what simple effects are for, coming up.


(f) Analyze the simple effects of B when A is
`medium`. (This means doing an appropriate `aov` and
an appropriate Tukey, if warranted.)

Solution


First, we pull out only the data where A is `medium`, and then we do a one-way analysis of B on that data. This is the slick way, though you can certainly save the result of `filter` first:
```{r }
hayfever %>%
  filter(a == "medium") %>%
  aov(relief ~ b, data = .) -> simple_medium
summary(simple_medium)
```

     

There is definitely some effect of ingredient B when A is medium. What
is it? Tukey will tell us:

```{r }
TukeyHSD(simple_medium)
```

 

There is no difference between medium and high (levels of B), but both of these are better in terms of relief than low is.


(g) Analyze the simple effects of B when A is `high`.

Solution


Same idea: pull out only the data where A is `high`, do a one-way analysis of B, and do Tukey if needed:
```{r }
hayfever %>%
  filter(a == "high") %>%
  aov(relief ~ b, data = .) -> simple_high
summary(simple_high)
TukeyHSD(simple_high)
```

     

When A is high, there is definitely an effect of B again, but this
time high (for B) is better than medium is better than low. (All the
P-values in Tukey are very small.)

You could guess this from the interaction plot as well (whichever one
you came up with): when A is high, it is better for B to be high as
well, but when A is medium or low, there is not much difference
between B being medium or high.

Extra: here, the effect of B is different, depending on what A is. This is
what a significant interaction *means*. If there were no
significant interaction, the effect of B would always be the same, no
matter what A was.



