## Lengths of heliconia flowers



 The tropical flower *Heliconia* is
fertilized by hummingbirds, a different species for each variety of
Heliconia. Over time, the lengths of the flowers and the form of the
hummingbirds' beaks have evolved to match each other. The length of
the Heliconia flower is therefore an important measurement. Does it
have a normal distribution for each variety?

The data set at
[http://www.utsc.utoronto.ca/~butler/c32/heliconia.csv](http://www.utsc.utoronto.ca/~butler/c32/heliconia.csv) contains
the lengths (in millimetres) of samples of flowers from each of three
varieties of Heliconia: *bihai*, *caribaea* red, and
*caribaea* yellow.



(a) Read the data into R. There are different numbers of length
measurements for each variety. How does this show up in the data
frame? (Look at all the rows, not just the first ten.)


Solution


The usual `read_csv`:
```{r }
heliconia=read_csv("heliconia.csv")
```       

I suggested to look at *all* the rows. Here's why:

```{r }
heliconia %>% print(n=Inf) 
``` 

The varieties with fewer values have missings (NAs) attached to the
end. This is because all the columns in a data frame have to have the
same number of values. (The missings won't impact what we do below ---
we get a warning but not an error, and the plots are the same as they
would be without the missings --- but you might be aesthetically
offended by them, in which case you can read what I do later on.)



(b) Make a normal quantile plot for the variety
*bihai*. 


Solution


There's a certain amount of repetitiveness here (that we work
around later):
```{r }
ggplot(heliconia,aes(sample=bihai))+stat_qq()+stat_qq_line()
```         

I'm saving the comments until we've seen all three.



(c) Make a normal quantile plot for the variety
*Caribaea* red (note that the variable name in the data
frame has an underscore in it).


Solution

          
Same idea again:
```{r }
ggplot(heliconia,aes(sample=caribaea_red))+stat_qq()+stat_qq_line()
```         



(d) Make a normal quantile plot for the variety
*Caribaea* yellow (this also has an underscore in it).


Solution


And, one more time:

```{r }
ggplot(heliconia,aes(sample=caribaea_yellow))+stat_qq()+stat_qq_line()
```         

I did a lot of copying and pasting there.



(e) Which of the three varieties is closest to having a normal
distribution? Explain (very) briefly.


Solution


Look at the three plots, and see which one stays closest to the
line. To my mind, this is clearly the last one, *Caribaea*
yellow. So your answer ought to be ``*Caribaea* yellow,
because the points are closest to the line''. This, I would say,
is acceptably close to normal, so using a $t$-test here would be fine.
The answer "the last one" is not quite complete, because I
asked you which *variety*, so your answer needs to name a variety.



(f) For each of the two other varieties, apart from the one you
mentioned in the last part, describe briefly how their distributions fail
to be normal.


Solution


Let's look at *bihai* first. I see this one as an almost
classic curve: the points are above the line, then below, then
above again. If you look at the data scale ($y$-axis), the points
are too bunched up to be normal at the bottom, and too spread out
at the top: that is, skewed to the *right*.
You might also (reasonably) take the view that the points at the
bottom are close to the line (not sure about the very smallest
one, though), but the points at the top are farther away, so that
what we have here is two outliers at the top. I'm OK with that.
It's often difficult to distinguish between skewness and outliers
(at the end of the long tail). What you conclude can often depend
on how you look.
We also had to look at the second plot, *caribaea*
red. This is a rather strange one: the points veer away from the
line at the ends, but look carefully: it is *not* outliers
at both ends, but rather the points are *too bunched up* to
be normal at both ends: that is, the distribution has \emph{short
tails} compared to the normal. It is something more like a
uniform distribution, which has no tails at all, than a normal
distribution, which won't have outliers but it *does* have
some kind of tails. So, "short tails". 
Extra: that's all you needed, but I mentioned above that you might have
been offended aesthetically by those missing values that were not
really missing. Let's see if we can do this aesthetically. As you
might expect, it uses several of the tools from the "tidyverse".
First, tidy the data. The three columns of the data frame are all
lengths, just lengths of different things, which need to be
labelled. This is `gather` from `tidyr`. 
```{r }
heliconia.long=heliconia %>% 
gather(variety,length,bihai:caribaea_yellow,na.rm=T)
heliconia.long  
``` 

This is now aesthetic as well as tidy: all those `NA` lines
have gone (you can check that 
there are now $16+23+15=54$ rows of actual data, as there should be).

Now, how to get a normal quantile plot for each variety? This is
`facet_wrap` on the end of the `ggplot` again. 

```{r }
ggplot(heliconia.long,aes(sample=length))+
stat_qq()+stat_qq_line()+
facet_wrap(~variety,scale="free")
``` 

These are a bit elongated vertically. The `scale="free"` allows
a different vertical scale for each plot (otherwise there would be one
vertical scale for all three plots); I decided that was best here
since the typical lengths for the three varieties are
different. *Caribaea* yellow is more or less straight,
*bihai*  has outliers (and may also be curved), *caribaea*
red has that peculiar S-bend shape.

I didn't really like the vertical elongation. I'd rather have the
plots be almost square, which they would be if we put them in three
cells of a $2 \times 2$ grid. `facet_wrap` has `nrow`
and `ncol` which you can use one or both of to make this
happen. This creates an array of plots with two columns and as many
rows as needed:

```{r }
ggplot(heliconia.long,aes(sample=length))+
stat_qq()+stat_qq_line()+
facet_wrap(~variety,scale="free",ncol=2)
``` 

I think the square plots make it easier to see the shape of these:
curved, S-bend, straightish.
Almost the same code will get a histogram for each variety, which I'll
also make squarish:

```{r }
ggplot(heliconia.long,aes(x=length))+
geom_histogram(binwidth=1)+facet_wrap(~variety,scale="free",ncol=2)

``` 

*bihai* has those two outliers, *caribaea* red has no tails
to speak of (or you might say "it's bimodal", which would be another
explanation of the pattern on the normal quantile plot\endnote{If you
have studied a thing called "kurtosis", the fourth moment about
the mean, you'll know that this measures *both* tail length
*and* peakedness, so a short-tailed distribution also has a
strong peak. Or, maybe, in this case, two strong peaks.}), and
*caribaea* yellow is shoulder-shruggingly normal (I looked at
that and said, "well, I *guess* it's normal".)  After you've
looked at the normal quantile plots, you see what a crude tool a
histogram is for assessing normality.




