##  Hunter-gatherers in Australia


 A hunter-gatherer society is one where people get their food
by hunting, fishing or foraging rather than by agriculture or by
raising animals. Such societies tend to move from place to place.
Anthropologists have studied hunter-gatherer societies in forest
ecosystems across the world. The average population density of these
societies is 7.38 people per 100 km$^2$. Hunter-gatherer societies on
different continents might have different population densities,
possibly because of large-scale ecological constraints (such as
resource availability), or because of other factors, possibly social
and/or historic, determining population density.

Some hunter-gatherer societies in Australia were studied, and the
population density per 100 km$^2$ recorded for each. The data are in
[http://ritsokiguess.site/datafiles/hg.txt](http://ritsokiguess.site/datafiles/hg.txt). 



(a) Read the data into R. Do you have the correct variables?
How many hunter-gatherer societies in Australia were studied?
Explain briefly.


Solution


The data values are separated by (single) spaces, so `read_delim`
is the thing:
```{r }
url="http://ritsokiguess.site/datafiles/hg.txt"
societies=read_delim(url," ")
```     

I like to put the URL in a variable first, because if I don't, the
`read_delim` line can be rather long. But if you want to do it
in one step, that's fine, as long as it's clear that you are doing the
right thing.

Let's look at the data frame:

```{r }
societies
``` 

I have the name of each society and its population density, as
promised (so that is correct). There were 13 societies that were
studied. For me, they were all displayed. For you, you'll probably see only the first ten, and you'll have to click Next to see the last three.
    
$\blacksquare$

(b) The question of interest is whether these Australian
hunter-gatherer societies are like the rest of the world in terms of mean
population density. State suitable null and alternative
hypotheses. *Define any symbols you use*: that is, if you use a
symbol, you also have to say what it means. 


Solution


The mean for the world as a whole ("average", as stated earlier)
is 7.38. Let $\mu$ denote the population mean for Australia (of
which these societies are a sample). Then our hypotheses are:
$$ H_0: \mu=7.38$$
and
$$ H_a: \mu \ne 7.38.$$
There is no reason for a one-sided alternative here, since all we
are interested in is whether Australia is different from the rest
of the world.
*Expect to lose a point* if you use the symbol $\mu$ without
saying what it means.

$\blacksquare$

(c) Test your hypotheses using a suitable test. What do you
conclude, in the context of the data?


Solution


A $t$-test, since we are testing a mean:
```{r }
t.test(societies$density,mu=7.38)
```     

The P-value is 0.0023, less than the usual $\alpha$ of 0.05, so we
*reject* the null hypothesis and conclude that the mean
population density is not equal to 7.38. That is to say, Australia is
different from the rest of the world in this sense.

As you know, "reject the null hypothesis" is only part of the
answer, so gets only part of the marks.
    
$\blacksquare$

(d) Do you have any doubts about the validity of your test?
Explain briefly, using a suitable graph to support your
explanation. 


Solution


The assumption behind the $t$-test is that the data are
approximately normal. We can assess that in several ways, but the
simplest (which is perfectly acceptable at this point) is a
histogram. You'll need to pick a suitable number of bins. This one
comes from Sturges' rule:
```{r }
ggplot(societies,aes(x=density))+geom_histogram(bins=5)
```     
Your conclusion might depend on how many bins you chose for your
histogram. Here's 8 bins (which is really too many with only 13
observations, but it actually shows the shape well): 

```{r }
ggplot(societies,aes(x=density))+geom_histogram(bins=8)
``` 

or you can get a number of bins from one of the built-in functions,
such as:

```{r }
mybins=nclass.FD(societies$density)
mybins
``` 

This one is small. The interquartile range is large and $n$ is small,
so the binwidth will be large and therefore the number of bins will be
small. 

Other choices: a one-group boxplot:

```{r }
ggplot(societies,aes(x=1,y=density))+geom_boxplot()
``` 

This isn't the best for assessing normality as such, but it will tell
you about lack of symmetry and outliers, which are the most important
threats to the $t$-test, so it's fine here. Or, a normal quantile plot:

```{r }
ggplot(societies,aes(sample=density))+
stat_qq()+stat_qq_line()
``` 

This is actually the best way to assess normality, but I'm not
expecting you to use this plot here, because we may not have gotten to
it in class yet. (If you have read ahead and successfully use the
plot, it's fine.)

After you have drawn your chosen plot (you need *one* plot), you
need to say something about normality and thus whether you have any
doubts about the validity of your $t$-test. This will depend on the
graph you drew: if you think your graph is symmetric and outlier-free,
you should have no doubts about your $t$-test; if you think it has
something wrong with it, you should say what it is and express your
doubts. My guess is that you will think this distribution is skewed to
the right. Most of my plots are saying that.
`r tufte::margin_note("The normal  quantile plot is rather interesting: it says that the uppermost  values are approximately normal, but the *smallest* eight or so  values are too bunched up to be normal. That is, normality fails not  because of the long tail on the right, but the bunching on the  left. Still right-skewed, though.")`

On the website where I got these data, they were using the data as
an example for another test, precisely *because* they thought the
distribution was right-skewed. Later on, we'll learn about the sign
test for the median, which I think is actually a better test here.
    
$\blacksquare$





