##  The Collatz sequence


 The Collatz sequence is a sequence of integers $x_1, x_2,
\ldots$ defined in a
deceptively simple way: if $x_n$ is the current term of the sequence,
then $x_{n+1}$ is defined as $x_n/2$ if $x_n$ is even, and $3x_n+1$ if
$x_n$ is odd.
We are interested in understanding how this sequence
behaves; for example, what happens to it as $n$ gets large, for
different choices of the first term $x_1$? We will explore this
numerically with R; the ambitious among you might like to look into
the mathematics of it.



(a) What happens to the sequence when it reaches 4? What would be
a sensible way of defining where it ends? Explain briefly.

Solution


When the sequence reaches 4 (that is, when its current term is 4),
the next term is 2 and the one after that is 1. Then the following
term is 4 again ($(3 \times 1)+1$) and then it repeats
indefinitely, $4, 2, 1, 4, 2, 1, \ldots$. 
I think a sensible way to define where the sequence ends is to say
"when it reaches 1", since if you start at 2 you'll never reach
4 (so "when it reaches 4" won't work), and it seems at least
plausible that it will hit the cycle 4, 2, 1 sometime.


(b) Write an R function called `is_odd` that returns
`TRUE` if its input is an odd number and `FALSE` if it
is even (you can assume that the input is an integer and not a
decimal number). To do *that*, you can use the function `%%` where 
`a %% b` is the remainder when `a` is divided by
`b`. To think about oddness or evenness, consider the
remainder when you divide by 2.

Solution


Let's try this out. For example, 5 is odd and 6 is even, so
```{r }
5 %% 2
6 %% 2
```

     

When a number is odd, its remainder on dividing by 2 is 1, and when
even, the remainder is 0. There is an additional shortcut here in that
1 is the numeric value of `TRUE` and 0 of `FALSE`, so
all we have to do is calculate the remainder on dividing by 2, turn it
into a `logical`, and return it:

```{r }
is_odd <- function(x) {
  r <- x %% 2
  as.logical(r)
}
```

 

You probably haven't seen `as.logical` before, but it's the
same idea as `as.numeric`: turn something that looks like a
`TRUE` or `FALSE` into something that actually
*is*. 

We should test it:

```{r }
is_odd(19)
is_odd(12)
is_odd(0)
```

 

0 is usually considered an even number, so this is good.


(c) Write an R function called
`hotpo1`
`r tufte::margin_note("*Hotpo* is short for half or triple-plus-one.")` 
that takes an
integer as input and returns the next number in the Collatz
sequence. To do this, use the function you just wrote that
determines whether a number is even or odd.

Solution


The logic is "if the input is odd, return 3 times it plus 1, otherwise return half of it". 
The R structure is an
`if-then-else`: 
```{r }
hotpo1 <- function(x) {
  if (is_odd(x)) 3 * x + 1 else x / 2
}
```

     

In R, the condition that is tested goes in brackets, and then if the
value-if-true and the value-if-false are single statements, you just
type them. (If they are more complicated than that, you put them in
curly brackets.) Now you see the value of writing `is_odd`
earlier; this code almost looks like the English-language description
of the sequence. If we had not written `is_odd` before, the
condition would have looked something like 

```{r eval=F}
if (x %% 2 == 1) 3 * x + 1 else x / 2
```

 

which would have been a lot harder to read.

All right, let's try that out:

```{r }
hotpo1(4)
hotpo1(7)
hotpo1(24)
```

 

That looks all right so far.


(d) Now write a function `hotpo` that will return the whole
Collatz sequence for an input $x_1$. For this, assume that you will
eventually get to 1.

Solution


This is a loop, but not a `for` loop (or something that we
could do `rowwise`), because we don't know how many
times we have to go around. This is the kind of thing that we
should use a `while` loop for: 
"keep going while a condition is true".
In this case, we should keep going if we haven't reached 1 yet. If
we haven't reached 1, we should generate the next value of the
sequence and glue it onto what we have so far. To initialize the
sequence, we start with the input value. There is an R trick to
glue a value onto the sequence, which is to use `c` with a
vector and a value, and save it back into the vector:
```{r }
hotpo <- function(x) {
  sequence <- x
  term <- x
  while (term > 1) {
    term <- hotpo1(term)
    sequence <- c(sequence, term)
  }
  sequence
}
```

     
I use `term` to hold the current term of the sequence, and
overwrite it with the next one (since I don't need the old one any
more). 

Does it work?

```{r }
hotpo(4)
hotpo(12)
hotpo(97)
```

 

97 is a wild ride, but it does eventually get to 1. 

Extra: where I originally saw this, which was "Metamagical Themas"
by Douglas Hofstadter, he was illustrating the programming language
Lisp and the process of recursion, whereby you define a function in
terms of itself. This one is a natural for that, because the Collatz
sequence starting at $x$ is $x$ along with the Collatz sequence
starting at the next term. For example, if you start at 12, the next
term is 6, so that the Collatz sequence starting at 12 is 12 followed by
the Collatz sequence starting at 6. There is no dependence any further
back. You can do recursion in R also; there is no problem with a
function calling itself:

```{r }
hotpo_rec <- function(x) {
  if (x == 1) 1 else c(x, hotpo_rec(hotpo1(x)))
}
```

 

Recursive functions have two parts: a "base case" that says how you
know you are done (the 1 here), and a "recursion" that says how you
move to a simpler case, here working out the next term, getting the
whole sequence for that, and gluing the input onto the front. It seems
paradoxical that you define a function in terms of itself, but what
you are doing is calling a simpler sequence, in this case one that is
length one shorter than the sequence for the original input. Thus, we
hope,
`r tufte::margin_note("Nobody knows whether you *always* get to 1, but also nobody has ever found a case where you don't. Collatz's conjecture, that you will get to 1 eventually, is known to be true for all starting $x_1$ up to some absurdly large number, but not for *all* starting points.")`
we will eventually reach 1.

Does it work?

```{r }
hotpo_rec(12)
hotpo_rec(97)
```

 

It does.

Recursive functions are often simple to understand, but they are not
always very efficient. They can take a lot of memory, because they
have to handle the intermediate calls to the function, which they have
to save to use later (in the case of `hotpo_rec(97)` there are a
lot of those). Recursive functions are often paired with a technique
called "memoization", where each time you calculate the function's
value, you *save* it in another array. The first thing you do in
the recursive function is to check whether you already have the
answer, in which case you just look it up and return it. It was a lot
of work here to calculate the sequence from 97, but if we had saved
the results, we would already have the answers for 292, 146, 73, 220
and so on, and getting those later would be a table lookup rather than
another recursive calculation.


(e) Write two (very small) functions that take an entire sequence
as input and return (i) the length of the sequence and (ii) the
maximum value it attains.

Solution


These are both one-liners. Call the input whatever you like:

```{r }
hotpo_len <- function(sequence) length(sequence)
hotpo_max <- function(sequence) max(sequence)
```

 

Because they are one-liners, you don't even need the curly brackets,
although there's no problem if they are there.

Testing:

```{r }
hotpo_len(hotpo(12))
hotpo_max(hotpo(97))
```

 

This checks with what we had before.


(f) Make a data frame consisting of the values 11 through 20, and,
using `tidyverse` ideas, obtain a data frame containing the
Collatz sequences starting at each of those values, along with their
lengths and their maximum values. Which sequence is longest? Which
one goes up highest?

Solution


This one uses `rowwise` ideas:
`r tufte::margin_note("I should have been more careful in my functions to make sure everything was integers, and, in particular, to do integer division by 2 because I knew that this division was going to come out even.")` 

```{r }
tibble(x = 11:20) %>%
  rowwise %>% 
  mutate(sequence = list(hotpo(x))) %>%
  mutate(length = hotpo_len(sequence)) %>%
  mutate(high = hotpo_max(sequence))
```

     

First, we obtain a list-column containing the sequences (which is why its calculation needs a `list` around it), then two
ordinary columns of their lengths and their maximum values. 

The sequences for 18 and 19 are the longest, but the sequence for 15
goes up the highest.



