##  Cars


 My cars data file can be found at
[link](http://www.utsc.utoronto.ca/~butler/c32/cars.csv). 
The values in
the data file are separated by commas; the car names are up to 29
characters long.  Display your results for each part after (a). In R,
displaying a `tibble` normally shows its first ten lines, which
is all you need here; there's no need to display all the lines.


(a) Read the data into R and list the values.


Solution


`read_csv` will do it:
```{r }
my_url <- "http://www.utsc.utoronto.ca/~butler/c32/cars.csv"
cars <- read_csv(my_url)
cars
```

 
      


(b) Display only the car names and the countries they come from.


Solution



```{r }
cars %>% select(car, country)
```

 

This *almost* works, but not quite:

```{r }
cars %>% select(starts_with("c"))
```

 

It gets *all* the columns that start with `c`, which
includes `cylinders` as well.
      


(c) Display everything *except* horsepower:


Solution


Naming what you *don't* want is sometimes easier:
```{r }
cars %>% select(-hp)
```

 
      


(d) Display only the cars that have 8-cylinder engines (but
display all the variables for those cars).


Solution


This:
```{r }
cars %>% filter(cylinders == 8)
```

 
8 of them, all from the US.
      


(e) Display the cylinders and horsepower for the cars that have
horsepower 70 or less.


Solution


This one is selecting some observations and some variables:
```{r }
cars %>% filter(hp <= 70) %>% select(cylinders:hp)
```

 

Cylinders and horsepower are consecutive columns, so we can select
them either using the colon `:` or by
`c(cylinders,hp)`. 

You can also do the `filter` and the
`select` the other way around.
This one works because the *rows* you want to
choose are determined by a column you're going to keep. If you wanted
to display the cylinders and horsepower of the cars with `mpg`
over 30, you would have to choose the rows first, because after you've
chosen the columns, there is no `mpg` any more.
      


(f) Find the mean and SD of gas mileage of the cars with 4 cylinders.


Solution



```{r }
cars %>% filter(cylinders == 4) %>% summarize(m = mean(MPG), s = sd(MPG))
```

 

Or you can get the mean and SD of gas mileage for all numbers of
cylinders, and pick out the one you want:

```{r }
cars %>% group_by(cylinders) %>% summarize(m = mean(MPG), s = sd(MPG))
```

 

Top row is the same as before. And since the output is a data frame,
you can do any of these things with *that*, for example:

```{r }
cars %>%
  group_by(cylinders) %>%
  summarize(m = mean(MPG), s = sd(MPG)) %>%
  filter(cylinders == 4)
```

 

to pick out just the right row.
This is a very easy kind of question to set on an exam. Just so you know.
      



