##  Diet and growth in boys


 A dietician is studying the effect of
different diets on children's growth. In part of the study, the
dietician is investigating two religious sects, labelled `a`
and `b` in our data set. Both sects are vegetarian; the
difference between them is that people in Sect A only eat vegetables
from below the ground, and Sect B only eats vegetables from above the
ground. The height and weight of the boys^[This was not sexism, but a recognition that boys and girls will be of different heights for reasons unrelated to diet. Doing it this way makes the analysis simpler.] are measured at regular
intervals. The data in
[link](http://ritsokiguess.site/datafiles/kids-diet.txt) are the
heights of the boys at age 12.



(a) Read in the data and find out how many observations you have
and which variables.


Solution


The data values are separated by one space, so:
```{r kids-diet-1 }
my_url <- "http://ritsokiguess.site/datafiles/kids-diet.txt"
diet <- read_delim(my_url, " ")
diet
```

     

21 observations on two variables, `sect` and
`height`. (You should state this; it is not enough to make the
reader figure it out for themselves.)

The heights are evidently in centimetres.

You can call the data frame whatever you like.


$\blacksquare$

(b) Obtain side-by-side boxplots of the heights for boys from
each sect. Does it look as if the heights of the boys in each sect
are different? Comment briefly. 


Solution


The boxplot is the kind of thing we've seen before:

```{r kids-diet-2 }
ggplot(diet, aes(x = sect, y = height)) + geom_boxplot()
```

 

It looks to me as if the boys in Sect B are taller on average.


$\blacksquare$

(c) Looking at your boxplots, do you see any problems with
doing a two-sample $t$-test? Explain briefly.


Solution


The assumption is that the data in each group are 
"approximately normal". Boxplots don't tell you about normality specifically,
but they tell you whether there are any outliers (none here) and
something about the shape (via the lengths of the whiskers). I'd
say the Sect A values are as symmetric as we could hope for. For
Sect B, you can say either that they're skewed to the left (and
that therefore we have a problem), or that the heights are close
enough to symmetric (and that therefore we don't). For me, either
is good. 
As ever, normal quantile plots can offer more insight. With
data in this form, the
two samples are mixed up, but using facets is the way to go. Philosophically, we draw a normal quantile plot of *all* the heights, and then say at the end that we would actually like a separate plot for each sect:
```{r kids-diet-3 }
diet %>%
  ggplot(aes(sample = height)) + stat_qq() + stat_qq_line() +
  facet_wrap(~sect, ncol = 1)
```

 

I decided that I wanted short squat plots rather than tall skinny ones.

With the sizes of the samples, I really don't see any problems
here. Most of the evidence for the left skewness in Sect B is actually coming
from that largest value being too small. Sect A is as good as you
could wish for. Having extreme values being *not extreme enough* is not a problem, since it won't be distorting the mean. 

The other way of doing this is to use `filter` to pull out the
rows you want and then feed that into the plot:

```{r kids-diet-4 }
secta <- filter(diet, sect == "a") %>%
  ggplot(aes(sample = sect)) + stat_qq() + stat_qq_line()
```

 

and the same for sect B. This is the usual `ggplot`-in-pipeline
thing where you don't have a named data frame in the `ggplot`
because it will use whatever came out of the previous step of the
pipeline.
    

$\blacksquare$

(d) Run a $t$-test to determine whether the mean heights differ
significantly. What do you conclude? Explain briefly. (Run the
$t$-test even if your previous work suggests that it is not the
right thing to do.)


Solution


The wording states that a two-sided test is correct, which is the
default, so you don't need anything special:

```{r kids-diet-5 }
t.test(height ~ sect, data = diet)
```

 

This is a two-sample test, so it takes a `data=`. 

Our null hypothesis is that the two sects have equal mean height. The
P-value of 0.103 is larger than 0.05, so we do not reject that null
hypothesis. That is, there is no evidence that the sects differ in
mean height. (That is, our earlier thought that the boys in Sect B were taller
is explainable by chance.)

You *must* end up with a statement about mean heights, and when
you do a test, you must state the conclusion in the context of the
problem, whether I ask you to or not. 
"Don't reject the null hypothesis" 
is a step on the way to an answer, not an answer in
itself. If you think it's an answer in itself, you won't be of much
use to the world as a statistician.

You might have been thinking that Mood's median test was the
thing, if you were worried about that skewness in Sect B. My guess is
that the $t$-test is all right, so it will be the better test
(and give the smaller P-value) here, but if you want to do it, you
could do it this way:

```{r kids-diet-6 }
library(smmr)
median_test(diet, height, sect)
```

 

My suspicion (that I wrote before doing the test) is correct: there is
even less evidence of a difference in median height between the
sects. The table shows that both sects are pretty close to 50--50
above and below the overall median, and with sample sizes this small,
they are certainly not significantly different from an even split. The
small frequencies bring a warning about the chi-squared approximation
possibly not working (that `smmr` suppresses). We had one like
this elsewhere, but there the result was very significant, and this one
is very non-significant. However, the implication is the same: even if
the P-value is not very accurate (because the expected frequencies for
sect B are both 4.5), the conclusion is unlikely to be wrong because
the P-value is so far from 0.05.
    
$\blacksquare$




