##  Investigating motor activity in rats


 A researcher named King was investigating
the effect of the drug midazolam on motor activity in rats. Typically,
the first time the drug is injected, a rat's motor activity decreases
substantially, but rats typically develop a "tolerance", so that
further injections of the drug have less impact on the rat's motor
activity.

The data shown in
[link](http://www.utsc.utoronto.ca/~butler/d29/king.csv) were all taken
in one day, called the "experiment day" below. 24 different rats
were used. Each rat, on the experiment day, was injected with a fixed
amount of midazolam, and at each of six five-minute intervals after
being injected, the rat's motor activity was measured (these are
labelled `i1` through `i6` in the data). The rats
differed in how they had been treated before the experiment day. The
control group of rats had previously been injected repeatedly with a
saline solution (no active ingredient), so the experiment day was the
first time this group of rats had received midazolam. The other two
groups of rats had both received midazolam repeatedly before the
experiment day: the "same" group was injected on experiment day in
the same environment that the previous injections had taken place (this
is known in psychology as a "conditioned tolerance"), but the
"different" group had the previous injections in a different
environment than on experiment day.

The column `id` identifies the rat from which each sequence of
values was obtained.



(a) Explain briefly why we need to use a repeated measures
analysis for these data.


Solution


Each rat is measured at six different times (`i1` through
`i6`): that is to say, each row of the data set consists of
repeated measurements on the *same* rat. (If each row had
used six *different* rats to obtain the six measurements, we
would have been back in familiar territory and could have used a
regular analysis of variance.)
    


(b) Read in the data and note that you have what was promised
in the question.


Solution


The usual:
```{r }
my_url="http://www.utsc.utoronto.ca/~butler/d29/king.csv"
king=read_csv(my_url)
king
```

     

There are 24 rats (rows). The columns label the rat (`id`) and
the times at which motor activity was measured (`i1` through
`i6`). The remaining column, `context`, describes how
the rats were treated before experiment day, with the levels being the
same ones given in the question.
    


(c) We are going to do a repeated-measures analysis using the
"profile" method shown in class. Create a suitable response
variable for this method.


Solution


`cbind` the appropriate columns together, to make a matrix:
```{r }
response=with(king,cbind(i1,i2,i3,i4,i5,i6))
```

     

This is the "simple" but tedious way, and produces a matrix because
the `i1` through `i6` are vectors (single columns):

```{r }
class(response)
```

 

`i1:i6` does not work here, because we are outside of the
`tidyverse`, and in that world, `:` only means
"through" (as in "this through that") when the things on either
side of it are or represent numbers.

The clever way to get the response is to `select` the columns
and then turn them into a matrix. This *does* permit the colon
because we are now in the `tidyverse`:

```{r }
response = king %>% 
select(i1:i6) %>%
as.matrix() 
class(response)  
```

 
It is indeed a matrix.

I tried to be extra-clever and use `starts_with`, but I have
another column `id` that starts with `i` that I do
*not* want to be part of the response. So I had to abandon that
idea, but not before trying this:

```{r }
response = king %>%
select(matches("i[0-9]")) %>%
as.matrix()
head(response)
```

 

`head` displays the first six lines (of anything). We don't
normally need it because we are typically dealing with
`tibble`-like data frames that display only ten rows of
themselves by default. But this worked. The `matches` part
takes a so-called "regular expression" which is a very flexible way
of matching anything: in this case, a column whose name starts with
`i` followed by exactly one digit (something between 0 and 9
inclusive). 
    


(d) Set up the "within-subjects" part of the analysis. That
means getting hold of the names of the columns that hold the
different times, saving them, and also making a data frame out of them:


Solution


```{r }
times=colnames(response)
times.df=data.frame(times)
```

   
    


(e) Fit the repeated-measures ANOVA. This will involve fitting
an `lm` first, if you have not already done so.


Solution


Fit the `lm` first, and then pass that into `Manova`
from `car`:
```{r }
king.1=lm(response~context,data=king)
king.2=Manova(king.1,idata=times.df,idesign=~times)
king.2
```

     
    


(f) What do you conclude from your repeated-measures ANOVA?
Explain briefly, in the context of the data.


Solution


The interaction term is significant, with a P-value less than
0.05. This is where we start and stop looking.
This means that the effect of time on motor activity (that is, the
way the motor activity depends on time) is different for each
`context`. That's all we can say now.
Grading note: as long as the setup and MANOVA are done somewhere,
I don't mind which part they are labelled with. But you need to do
the setup, initial `lm` and `Manova`
*somewhere* so that everything comes out right in the end.
    


(g) To understand the results of the previous part, we are going
to make a spaghetti plot. In preparation for that, we need to save
the data in "long format" with one observation on *one* time
point in each row. Arrange that, and show by displaying (some of)
the data that you have done so.


Solution


This is `gather` yet again: gather up columns `i1`
through `i6` and call them something like `activity`:
```{r }
king.long = king %>% gather(time,activity,i1:i6)
king.long
```

     

Displaying the resulting data frame is a good way to display ``some
of'' it. You can always look at more if you like. There are more rows
and fewer columns than there were before, which is
encouraging. `gather` works columnwise: it gathers up all the
values in the first time column `i1` first, then `i2`,
and so on.
    


(h) Make a spaghetti plot: that is, plot motor activity against
the time points, joining the points for each *rat* by lines,
and colouring the points and lines according to the *context*.


Solution


That means this, using `group` to indicate which points to
join by lines, since it's different from the `colour`: 
```{r }
ggplot(king.long,aes(x=time,y=activity,colour=context,group=id))+
geom_point()+geom_line()
```

     

I'd say the `geom_point` is optional, so that this is also
good, perhaps better even:

```{r }
ggplot(king.long,aes(x=time,y=activity,colour=context,group=id))+
geom_line()
```

 
    


(i) Looking at your spaghetti plot, why do you think your
repeated-measures ANOVA came out as it did? Explain briefly.



Solution


What you're after is an explanation of how the *patterns* over
time are different for the three `context`s. If you can find
something that says that, I'm good. For example, even though all of
the rats experienced a decrease in motor activity between times 1
and 2, the rats in the `same` group didn't decrease as
much. Or, the rats in the `same` group continued to decrease
beyond time 2, whereas the rats in the `control` and
`different` groups tended to level off after time 2, not
decreasing so much after that.
If you like, you can draw an interaction plot by working out the
means for each `context`-`time` group first:
```{r }
king.long %>% group_by(context,time) %>%
summarize(m=mean(activity)) %>%
ggplot(aes(x=time,y=m,colour=context,group=context))+
geom_point()+geom_line()
```

   

This seems to illustrate the same things as I found on the spaghetti
plot. It gains in clarity by only looking at means, but loses by not
considering the variability. Your call.

This kind of thing also runs with `lmer` from package
`lme4`. It uses the long data frame, thus, treating `id`
(identifying the rats) as a random effect:

```{r }
king.3=lmer(activity~context*time+(1|id),data=king.long)
```

 

What can we drop? The only thing under consideration is the interaction:

```{r }
drop1(king.3,test="Chisq")
```

 

and we get the same conclusion as before, but with a much smaller P-value.

With this kind of modelling, there is no distinction between
"within" and "between", so that even though `time` is a
within-subjects factor and `context` is between subjects, if
the interaction had not been significant, we could have dropped it
from the model, and then we would have had an effect of `time`
and an effect of `context`, independent of each other. I was
actually looking for an example with a non-significant interaction,
but I couldn't find one. 
  



