##  Treating leprosy


 Two drugs are being tested in the treatment of
leprosy. These are labelled A and D. There is also a control drug,
labelled F. The response variable is a post-treatment score of leprosy
bacilli (measured at six different sites on each patient). A lower
score is better.

Thus far, we have a standard one-way analysis of variance. But the
researchers wanted greater precision in assessing the effects (if any)
of the drugs, so they also measured a pre-treatment score of leprosy
bacilli. The data are in the file
[link](http://www.utsc.utoronto.ca/~butler/d29/leprosy.txt). The
pre-treatment and post-treatment scores are labelled `pre` and
`post` respectively.



(a) Read in the data and check that you have apparently the
right thing.
```{r }
```

   
 
Solution


Take a look at the data file. The values have multiple spaces
between them, but they are aligned with each other and the
column headings, so `read_table` is the thing:
```{r }
my_url="http://www.utsc.utoronto.ca/~butler/d29/leprosy.txt"
lepro=read_table(my_url)
lepro
```

 

Call it what you like.

That looks good, with variables of the right names. 
 

(b)<a name="part:lepro-scatter">*</a> Make a scatterplot of post-treatment score against
pre-treatment score, with the points for each drug drawn in a
different colour. 
 
Solution


This is the kind of thing that `ggplot` does without
batting an eyelid:
```{r }
ggplot(lepro,aes(x=pre,y=post,colour=drug))+geom_point()
```

       
 

(c) Does it appear that including the pre-treatment score was a
good idea? Explain briefly.
 
Solution


The overall trend on the scatterplot is that a higher `pre`
tends to go with a higher `post`, regardless of drug, so
including this information appears to be informative.
I personally suspect that there's some fan-out happening on the
pre-post relationship, but I'm not planning to make you explore that.
 

(d) What about this dataset suggests that analysis of
covariance is a method worth trying?
 
Solution


The key is a mixture of categorical and quantitative explanatory
variables. Here we have a categorical variable `drug` and a
quantitative one `pre`. 
If we had only one type of explanatory variable, we could do a
regression or an ANOVA as appropriate. But we don't. In some ways,
it's not worth making a fuss about the distinction, because
regressions and ANOVAs are all linear models anyway. But you may
see the term "analysis of covariance", so it's worth your while
to know what it's about. 
 

(e) Fit an analysis of covariance model to predict
post-treatment score. Include an interaction between your
explanatory variables. (You don't need to look at the output from
the model.)
 
Solution


This is what you'd guess. `lm` handles the interaction
properly, even though `pre` is a  quantitative variable.

```{r }
lepro.1=lm(post~pre*drug,data=lepro)
```

     

I wanted to take a look, so I did:

```{r }
summary(lepro.1)
```

 

For testing the interaction, there are *two* slope coefficients
that should be zero if there is no interaction. So we have to test
this with `drop1`, which is next.
 

(f) Pass your fitted model of the last part into
`drop1`. Is
the interaction term significant?


Solution


Just this:
```{r }
drop1(lepro.1, test="F")
```

       

There is only a test for the interaction term because you can't take out the main effects until you've taken out the interaction.

The P-value for the interaction is very large (0.5606) so it is
nowhere near significant. We can drop the interaction.



(g) Fit a model without the interaction. Is this a sensible thing to
do (in addition, that is, to the fact that I just asked you to do it)?


Solution


Change the `*` to a `+`:
```{r }
lepro.2=lm(post~pre+drug,data=lepro)
```

     

Or use `update` (not much in it, here):

```{r }
lepro.2a=update(lepro.1,.~.-pre:drug)
```

 

We just said that the interaction could come out, since it wasn't
significant, so this is exactly the model that we should be fitting.



(h) Take a look at the `summary` of your preferred
model. Is there a significant effect of pre-treatment score?
Describe the effects of the different drugs on the post-treatment
score. (Don't do any tests for `drug`.) Does your comparison
of drugs make sense?
 
Solution


Mine was the no-interaction model `lepro.2`:
```{r }
summary(lepro.2)
```

     

The pre-treatment term is definitely significant, with a P-value of
0.0000025. So pre-treatment score definitely has an impact on
post-treatment score.

I didn't ask you to test for significance of drugs. I just wanted you
to assess their coefficients. Drug A is being used as the baseline, so
its coefficient is zero. Drug D has a slightly positive coefficient
(0.109) so its average bacilli score is slightly higher (for any
pre-treatment score) than for drug A. Drug F, which was the placebo,
has a slope of 3.446, so its average bacilli score is a fair bit
higher than for either of the other drugs. This makes sense because a
higher score is worse, and the two "real" drugs are both better than
the fake one.

Whether there is a real drug difference, I didn't ask you to assess,
but you could do it by `drop1` again, this way:
```{r }
drop1(lepro.2, test="F")
```

 

This is actually not significant.
`r tufte::margin_note("This is why I didn't ask you to test this, since it would have confused the story.")`
This is one of those cases where the non-significant `drug` has
a slightly *bigger* AIC than `<none>`, so `drop1`
considers it best to leave it in the model.
 

(i) Obtain predicted values for `post` for each of the
three drugs at `pre` scores 5, 12 and 20. To do this, obtain
a new data frame that has all 9 combinations of drugs and
`pre` scores, and then feed this into `predict` using
your preferred model.
`r tufte::margin_note("Analysis of covariance is just a linear  model, so *predict* works the same here as in regression.")`
 
Solution


First, make the new data frame for predicting from, using
`crossing`. I'm doing this in small steps for clarity:
first, I define all the drugs and `pre` values, and then I
feed them into `crossing`:
```{r }
drugs=c("A","D","F")
pres=c(5,12,20)
lepro.new=crossing(drug=drugs,pre=pres)
lepro.new
```

     

The names of the columns in the new data frame have to be
*exactly* `drug` and `pre`. Anything else, and
`predict` won't work.

Now I obtain the predictions, from my best model `lepro.2`. I
don't need intervals or anything like that:

```{r }
preds=predict(lepro.2,lepro.new)
preds
```

 

only it would be easier to see the predictions side by side with the
values they are predictions for:

```{r }
allpreds=cbind(lepro.new,preds)
allpreds
```

 

I gave this a name in case I feel like using it again later.
 

(j) Now, plot the data with the fitted lines on. You'll need to do
the trick of using the original data frame (and plot the points),
and the data frame of predictions (and join those points with lines).
 
Solution



```{r }
ggplot(lepro,aes(x=pre,y=post,colour=drug))+geom_point()+
geom_line(data=allpreds,aes(y=preds,linetype=drug))
```

 

Inside the `aes` for the `geom_line`, I only have to
specify the things that have changed, so that the `x`
(`pre`) and the `colour` (`drug`) are
"inherited" unchanged from the first `aes`.

`geom_line` has the `aes` as its default first argument,
so I have to specify `data=` by name.

The `linetype=` is really overkill, but I just wanted to show
you that you can distinguish the drugs by line type as
well.
`r tufte::margin_note("The line types show up in the legend too, though they're not so easy to see.")`
 

(k) Are the lines on your plot parallel, with the same slopes? Is this what you would
expect? Explain briefly.
 
Solution


My lines are parallel. This is exactly what I would expect, since
my best model has no interaction, and the interaction is what
would make the lines *not* be parallel. 
If your best model
*did* have the interaction term still in it, your predictions
would have been these:
```{r }
preds=predict(lepro.1,lepro.new)
cbind(lepro.new,preds)
```

     

The slopes for each drug are no longer the same (because nothing is
forcing them to be the same). For example, if you compare drug A at
pre score 5 and 12, it goes up by 5.2, while drug D at the same pre
scores goes up by 7.5.

The `ggplot` in this case is actually a lot easier than the
other `ggplot`, because the lines are now just the regression
lines for each group.  So `geom_smooth` will get them:

```{r }
ggplot(lepro,aes(x=pre,y=post,colour=drug))+
geom_point()+geom_smooth(method="lm")
```

 

The grey intervals are a bit confusing.
`r tufte::margin_note("You can get rid of them by saying *se=F* inside the *geom_smooth*, as normal.")` They are
confidence intervals for the mean `post` score (as we did for regression
early on in the course). But I left them there to show that they
overlap substantially and thus that those slopes are not 
"really different". That's the lack of interaction. There is, as you see, a
substantial scatter in the points that would make it very difficult to
prove that those three slopes are really different. 
 



