##  Confusing letters


 Two researchers studied how often people mistook one letter
for another one. They did this by showing their subjects each
(uppercase) letter for a few milliseconds only, and asking them which
letter (out of a list) they just saw. The researchers obtain a
"confusion matrix" in which, for example, the W-Q entry is the total
of W's taken as Q and Q's taken as W. This confusion matrix is in
[link](http://ritsokiguess.site/datafiles/letterrec.txt). Note that
the confusions are actually *similarities* rather than
dissimilarities: a large number means that the two letters concerned
are easily confused, that is, similar. Similarities can be converted
to dissimilarities by subtracting them from a larger number than
anything in the table (that is, subtracting from 21 is good here).



(a) Read in the confusion matrix and convert it to
dissimilarities and then to a
`dist` object, displaying the `dist` object that you
obtain. 


Solution


`read_table` to read in the data, having first noted that
we have aligned columns with spaces between:
```{r letterrec-1 }
my_url <- "http://ritsokiguess.site/datafiles/letterrec.txt"
letters <- read_table(my_url)
letters
```

    

The first column didn't have a heading, so `read_table` filled
in `X1` for it (there was data down below, so it needed to have
a name). This is in contrast to something like `read_delim`,
where you have to have as many headers as columns. In
`read_table`, you know where the headers have to be, so if any
are missing they can be filled in, but with `read_delim` you
don't have that knowledge, so you have to have exactly the right
number of headers, one per data column exactly.

These are similarities, so convert to dissimilarities by subtracting
from 21. This is a shortcut way to do that, once you've gotten rid of
everything that is not numbers:

```{r letterrec-2, error=T}
letters %>% dplyr::select(-X1) -> letters2
letters2 <- 21 - letters2
letters2
```

 

This looks weird, or at least the stuff above the diagonal does, but
`dist` works with the stuff *below* the diagonal (unless
you tell it otherwise), so all will be good.

`as.dist` comes next:

```{r letterrec-3, error=T}
d <- as.dist(letters2)
d
```

 

This works because the letters that are confused are actually
column names of the data frame. 

You can check (and probably should, at least for yourself) that the
distances in `d` correspond properly to the ones in
`letters`. For example, the letters C and G were confused 12
times, $21-12=9$, and the entry for C and G in `d` is indeed
9. 

Note that the actual confusion numbers were in the data file as the
bottom half of the matrix, with the top half being
zeroes. `as.dist` handled this with no problem. (You can check
the help for `as.dist` to find out how it deals with this kind
of thing.)


$\blacksquare$

(b) Run a multidimensional scaling, with the default two
dimensions, on your `dist` object, and display the
results. (Making a graph is coming later.)

Solution


This:
```{r letterrec-4 }
d.1 <- cmdscale(d)
d.1
```

     

Or you can do it with `eig=T`, which gets you some more information:

```{r letterrec-5 }
d.1a <- cmdscale(d, eig = T)
d.1a
```

 

The most interesting thing here is the `GOF` at the bottom,
which is not that high (whichever one of the two values you take),
suggesting that the two-dimensional representation is not very
good. Further evidence for that is in `eig`, the
"eigenvalues", which are about 493, 232, 140, 50, 40, 0 and some
negative ones, which suggests that three dimensions might be better
than two for representing the data, because the first *three*
eigenvalues seem noticeably bigger than the others. (This is the same
thinking as the `svd` or percent of trace in discriminant
analysis.)  But that's by the way: we'll stick with two dimensions.

The important thing to remember is that if you go the `eig=T`
way, you have to pull out the points to plot from the thing called
`points`, so that you plot `d.1` itself  but `d.1a$points`.

$\blacksquare$

(c) Obtain a vector of the names of the letters that were
confused in this study. 

Solution


Easiest way is to pull out the first column of the data frame that
you read in from the file (if you can remember what it was called):
```{r letterrec-6 }
letter_names <- letters$X1
```

     

That silly column name `X1` that `read_table`
supplied. 

Extra: You can even get the letter names from the thing I called `d`,
but I can't remember how, so I have to cheat. I said that `d`
has a "print method"^[The way that a multicoloured function  like `print` works is that when you ask to show something,  like `d`, R first looks to see what kind of thing you want to show (by calling `class`), and determines that it is a `dist` object. Then it looks to see if there is a function called `print.dist` (there is), and if there is, it calls that (to produce that nice display). If there isn't, it calls `print.default`, which just displays its input without doing anything special. This is why printing the output from an `lm` looks very different from printing a data frame: the first calls `print.lm` and the second calls `print.data.frame`, or `print.tbl_df` for a tibble.] 
that controls how it looks:

```{r letterrec-7 }
d
```



but its innards are a whole lot more complicated than that:

```{r letterrec-8 }
print.default(d)
```

 

or

```{r letterrec-9 }
unclass(d)
```

 

This one gets rid of any special kind of thing that `d` is, and
displays it like a thing without any special properties.

It's the "attribute" called `Labels` that we need to grab:

```{r letterrec-10 }
attributes(d)$Labels
```

 

$\blacksquare$

(d) Plot your multidimensional scaling map. To do this, first
create a data frame containing the points you want to plot and their
labels, and then plot the points labelled by the right thing.

Solution


The "labels" need to be the letter names, which is why I made
you find them in the previous part. I'm going to do this with a
pipe, or else I'll create *another* thing called `d`
and overwrite the one I wanted to keep, again:
```{r letterrec-11, error=T}
data.frame(d.1, names = letter_names)
```

     

So far so good. The coordinates have gained names `X1` and
`X2` (as they did before, but I wanted to check). This is what
happens when you turn a matrix with nameless columns into a data
frame. So I can proceed:

```{r letterrec-12, error=T}
data.frame(d.1, names = letter_names) %>%
  ggplot(aes(x = X1, y = X2, label = names)) +
  geom_point() + geom_text_repel() +
  coord_fixed()
```

     

I need the  last line for the same reason as before: I want to treat
the two axes equally. If you don't have that, you have a distorted map
where one of the coordinates appears to be more important than the other.

If you can't get `ggrepel` to behave itself, an equally good
alternative is to plot the letter names instead of the labelled
dots. To do that, take out the `geom_point` and add an
ordinary `geom_text`. This will plot, *at* the location
given by the `x` and `y`, the text in `label`,
centred at the location. (If you have `geom_point()` as well,
you'll get a black dot in the middle of each piece of text.)
`geom_text` has options to justify the text relative to the
point, so you can see both, but I've always found these rather
finicky, so I'd rather let `geom_text_repel` do the work of
figuring out where to put the text relative to the point.

```{r letterrec-13, error=T}
data.frame(d.1, names = letter_names) %>%
  ggplot(aes(x = X1, y = X2, label = names)) +
  geom_text() + coord_fixed()
```

 

Then pick (in the next part) a pair of letters that is close together,
and proceed. I like the labelled dots better (as a matter of taste),
but this way is a perfectly good way to answer the question, so is
perfectly acceptable here.

$\blacksquare$

(e) Pick a pair of letters appear close together on your
map. Does it make sense, from looking at these letters, that they
would be easy to confuse? Explain briefly.

Solution


My map has four pairs of close-together letters: C and G, H and M,
N and W, Q and D. Pick one of those pairs. I don't mind which pair
you pick. (If you got something different, pick from what your map
shows.) 
I think it is not at all surprising that these pairs of letters
got confused, because the letters of each pair have similar shapes
(which is all you'd have to react to if you see them for "a few    milliseconds" each): C and G circular with a hole on the right,
H, M, N and W vertical lines on the outside with something across
the middle, Q and D almost circular. (It's up to you whether you
consider H, M, N and W as a group of four or as two pairs.)

$\blacksquare$

(f) Verify that your chosen pair of letters was often confused
in the data.

Solution


The data we read in from the file was this:
```{r letterrec-14 }
letters
```

     

Then look for your pair of letters:


\begin{tabular}{lr}
letters & confusion\\
\hline            
C,G & 12\\
H,M & 10\\
N,W & 13\\
Q,D & 20\\     
\hline
\end{tabular}


These are four of the biggest numbers in the matrix, which is as it
should be. You need to find the number of confusions for your pair of
letters and assert that it is large (compared to the others).

These are actually not *all* the large ones: M and W, H and N, M
and N are also large (which lends some support to these letters being
a foursome^[Set of four. Match-play golf has a game called  *foursomes* where the two players on each team take it in turns to hit the ball, as opposed to the game called *fourballs*, where each of the two players plays their own ball, and the team's score on a hole is the better of the two players' scores.] rather than two pairs).

If you prefer, you can work from your `dist` object, the thing
I called `d`:

```{r letterrec-15 }
d
```



This time, you're looking for a *small* dissimilarity between
your pair of letters:


\begin{tabular}{lr}
letters & dissimilarity\\
\hline            
C,G & 9\\
H,M & 11\\
N,W & 8\\
Q,D & 1\\     
\hline
\end{tabular}


These, again, are smaller than most, though not the smallest
overall. So, if you go this way, you need to assert that the
corresponding number in your `dist` object is *small*.

$\blacksquare$


