##  Fabricated data


 The data in
[link](http://ritsokiguess.site/datafiles/simple-manova.txt)
(probably made up) are measurements of two 
variables `y1` and `y2` on three groups, labelled
`a`, `b` and `c`. We want to see whether
`y1` or `y2` or some combination of them differ among
the groups.



(a) Read in the data and check that you have 12 observations on
2 numeric variables from 3 different groups.


Solution


```{r }
my_url <- "http://ritsokiguess.site/datafiles/simple-manova.txt"
simple <- read_delim(my_url, " ")
simple
```

   

As promised: 12 observations, 2 numeric variables `y1` and
`y2`, and one categorical variable (as text) containing three
groups `a,b,c`.
    

$\blacksquare$

(b) Run a one-way ANOVA predicting `y1` from
`group`. Are there any significant differences in `y1`
among the groups? If necessary, run Tukey. What do you conclude?


Solution


```{r }
simple.1 <- aov(y1 ~ group, data = simple)
summary(simple.1)
TukeyHSD(simple.1)
```

     

The $F$-test said that there were differences among the groups, so I
ran Tukey and found that group `c` is significantly bigger on
`y1` than the other two groups.


$\blacksquare$

(c) Repeat all the previous part for `y2`.


Solution


```{r }
simple.2 <- aov(y2 ~ group, data = simple)
summary(simple.2)
TukeyHSD(simple.2)
```

     

Same idea: there is a difference, and Tukey reveals that this time
group `a` is significantly *smaller* than the other groups
on `y2`. 
    

$\blacksquare$

(d) Make a plot of `y1` against `y2`, with the
points distinguished by which group they are from.


Solution


`y1` and `y2` can be either way around, since
they are both response variables (!):

```{r }
ggplot(simple, aes(x = y1, y = y2, colour = group)) + geom_point()
```


    

$\blacksquare$

(e) How is it that group `b` is not distinguishable on
either `y1` or `y2` individually (from the ANOVAs),
but *is* distinguishable on your graph? Explain briefly.


Solution


Group `b` has the same kind of `y1` values as group
`a` and the same kind of `y2` values as group
`c`. This means looking at either variable singly is not
enough to distinguish group `b`. But group `b`
stands out on the graph as being "top left" of the picture. That
means that `y1` has to be low *and* `y2` has to
be high, both together, to pick out group `b`. 
That seems like a lot of work, but the important point
that I want you to raise is that the points in group `b`
are distinguished by a particular *combination* of values of
`y1` and `y2`.
    

$\blacksquare$

(f) Run a one-way multivariate analysis of variance, obtaining
a P-value. Note the very small P-value.


Solution


This is just like the seed weight and yield example in class
(deliberately so):
```{r }
response <- with(simple, cbind(y1, y2))
simple.3 <- manova(response ~ group, data = simple)
summary(simple.3)
```

     

The P-value here, 0.00027, is smaller than either of the P-values for
the one-variable ANOVAs. This expresses the idea that the group
difference is essentially multivariate (it depends on the
*combination* of `y1` and `y2` values). I only
wanted you to note that it is smaller than either of the other
ANOVAs; you don't need to say anything about why.

We were in the fortunate position of being able to draw a picture,
because we had only two response variables, and we could plot them
against each other with the groups labelled. (If we had three response
variables, like the peanuts example, we wouldn't be able to do this,
barring some kind of 3D-plotting procedure.)
    
$\blacksquare$




