##  Fear of math


 Two new short courses have been proposed for helping
students who suffer from severe math phobia. The courses are labelled
A and B. Ten students were randomly allocated to one of these two
courses, and each student's score on a math phobia test was recorded
after they completed their course. The math phobia test produces
whole-number scores between 0 and 10, with a higher score indicating a
greater fear of mathematics. The data can be found in
[link](http://www.utsc.utoronto.ca/~butler/c32/mathphobia.txt). We start
with R for this question.



(a) Read in the data and check, however you like, that you have 10
observations, 5 from each course.


Solution


This doesn't need much comment:
```{r }
my_url="http://www.utsc.utoronto.ca/~butler/c32/mathphobia.txt"
math=read_delim(my_url," ")
math
```

 

This will do, counting the `a` and `b`. Or, to save
yourself that trouble:

```{r }
math %>% count(course)
```

 

Five each. The story is to get the computer to do the grunt work for
you, if you can make it do so. Other ways:

```{r }
math %>% group_by(course) %>% summarize(count=n())
```

 

and this:

```{r }
with(math,table(course))
```

 

giving the same answer. Lots of ways.

Extra: there is an experimental design issue here. You might have noticed that each student did only *one* of the courses. Couldn't students do both, in a matched-pairs kind of way? Well, it's a bit like the kids learning to read in that if the first of the courses reduces a student's anxiety, the second course won't appear to do much good (even if it actually would have been helpful had the student done that one first). This is the same idea as the kids learning to read: once you've learned to read, you've learned to read, and learning to read a second way won't help much. The place where matched pairs scores is when you can "wipe out" the effect of one treatment before a subject gets the other one. We have an example of kids throwing baseballs and softballs that is like that: if you throw one kind of ball, that won't affect how far you can throw the other kind.



(b) Do a two-sample $t$-test to assess whether there is a
difference in mean phobia scores after the students have taken the two
courses. What do you conclude? (You have no \textsl{a
priori}
`r tufte::margin_note("That is, before looking at the data. This is  Latin. It's also the place that the Bayesian *prior distribution*  comes from. The *posterior distribution* comes from the Latin  *a posteriori*, which means *afterwards*, that is, after  you have looked at the data.")` 
reason to
suppose that a particular one of the tests will produce a higher mean
than the other, so do a two-sided test.)



Solution


A two-sided test is the default, so there is not much to do here:

```{r }
t.test(phobia~course,data=math)
```

 

The P-value of 0.4456 is nowhere near less than 0.05, so there is no
evidence at all that the  
mean math phobia scores are different between the two courses. 




(c) Draw boxplots of the math phobia scores for each group (one
line of code). What is the most striking thing that you notice?



Solution


```{r }
ggplot(math,aes(x=course,y=phobia))+geom_boxplot()
```

 

Boxplot `a` is just weird. The bar across the middle is
actually at the top, and it has no bottom. (Noting something sensible
like this is enough.) Boxplot `b` is hugely spread
out.
`r tufte::margin_note("The two groups have very different spreads, but that is  not a problem as long as we remember to do the Welch-Satterthwaite  test that does not assume equal spreads. This is the default in R,  so we are good, at least with that.")`

By way of explanation: the course `a` scores have a number of
values equal so that the 3rd quartile and the median are the name, and
also that the first quartile and the minimum value are the same:

```{r }
tmp=math %>% filter(course=="a")
tmp %>% count(phobia)
summary(tmp)
```

 

The phobia scores from course A are two 6's, two 7's and an 8. The
median and third quartile are both 7, and the first quartile is the
same as the lowest value, 6.

Technique note: I wanted to do two things with the phobia scores from
course A: count up how many of each score, and show you what the
five-number summary looks like. One pipe won't do this (the pipe
"branches"), so I saved what I needed to use, before it branched,
into a data frame `tmp` and then used `tmp` twice. Pipes
are powerful, but not *all*-powerful.





(d) Explain briefly why a $t$-test would not be good for these
data. (There are two things that you need to say.)



Solution


The easiest way to structure this is to ask yourself first what the
$t$-test needs, and second whether you have it.
The $t$-test assumes (approximately) normal data. The boxplot for
group `a` doesn't even look symmetric, and the one for group
`b` has an oddly asymmetric box. So I think the normality is
in question here, and therefore another test would be
better. (This is perhaps a bit glib of an answer, since there are
only 5 values in each group, and so they can certainly look
non-normal even if they actually are normal, but these values are
all integers, so it is perhaps wise to be cautious.)
We have the machinery to assess the normality for these, in one shot:
```{r }
ggplot(math,aes(sample=phobia))+
stat_qq()+stat_qq_line()+
facet_wrap(~course,ncol=1, scales="free")
```

   

I don't know what *you* make of those, but they both look pretty
straight to me (and there are only five observations, so it's hard to
judge). Course `b` maybe has a "hole" in it (three large
values and two small ones). Maybe. I dunno. What I would *really*
be worried about is outliers, and at least we don't have those.
I mentioned in class that the $t$-tests are robust to non-normality. I
ought to have expanded on that a bit: what really makes the $t$-test
still behave itself with non-normality is when you have *large*
samples, that is, when the Central Limit Theorem has had a chance to
take hold. (That's what drives the normality not really being
necessary in most cases.) But, even with small samples, exact
normality doesn't matter so much.  Here, we have two tiny samples, and
so we have to insist a bit more, but only a bit more, on a
more-or-less normal shape in each group. (It's kind of a double
jeopardy in that the situation where normality matters most, namely
with small samples, is where it's the hardest to judge, because
samples of size 5 even from a normal distribution can look very
non-normal.)
But, the biggest threats to the $t$-test are big-time skewness and
outliers, and we are not suffering too badly from those.
  



(e) Run a suitable test to compare the "typical" scores for the
two courses. (You can use the version from a package rather than
building your own.) What do you conclude?



Solution


This is an invite to use `smmr`:
```{r }
library(smmr)
median_test(math,phobia,course)
```

     

We are nowhere near rejecting equal medians; in fact, both courses are
very close to 50--50 above and below the overall median.

If you look at the frequency table, you might be confused by
something: there were 10 observations, but there are only $1+2+2+2=7$
in the table. This is because three of the observations were equal to
the overall median, and had to be thrown away:

```{r }
math %>% summarize(med=median(phobia))
math %>% count(phobia)
```

 

The overall median was 7. Because the actual data were really discrete
(the phobia scores could only be whole numbers), we risked losing a
lot of our data when we did this test (and we didn't have much to
begin with). The other thing to say is that with small sample sizes,
the frequencies in the table have to be *very* lopsided for you
to have a chance of rejecting the null. Something like this is what
you'd need:

```{r }
x=c(1,1,2,6,6,6,7,8,9,10)
g=c(1,1,1,1,1,2,2,2,2,2)
d=tibble(x,g)
median_test(d,x,g)
```

 

I faked it up so that we had 10 observations, three of which were
equal to the overall median. Of the rest, all the small ones were in
group 1 and all the large ones were in group 2. This is lopsided
enough to reject with, though, because of the small frequencies, there
actually was a warning about 
"chi-squared approximation may be inaccurate".
`r tufte::margin_note("There *was*, in the   *chisq.test* inside *median_test*, but in  *smmr* I didn't pass that warning back to the outside world.")`




