##  Salaries of mathematicians


 A researcher in a scientific
foundation wanted to evaluate the relationship between annual salaries
of mathematicians and three explanatory variables:


* an index of work quality

* number of years of experience

* an index of publication success.


The data can be found at
[link](http://www.utsc.utoronto.ca/~butler/d29/mathsal.txt). Data from
only a relatively small number of mathematicians were available.



(a) Read in the data and check that you have a sensible number
of rows and the right number of columns. (What does "a sensible  number of rows" mean here?)

Solution


This is a tricky one. There are aligned columns, but *the column headers are not aligned with them*. 
Thus `read_table2` is what you need.
```{r }
my_url="http://www.utsc.utoronto.ca/~butler/d29/mathsal.txt"
salaries=read_table2(my_url)
salaries
```

        

24 observations ("only a relatively small number") and 4 columns,
one for the response and one each for the explanatory variables.

Or, if you like,

```{r }
dim(salaries)
```

 

for the second part: 24 rows and 4 columns again.
I note, with only 24 observations, that we don't really have enough
data to investigate the effects of three explanatory variables, but
we'll do the best we can. If the pattern, whatever it is, is clear
enough, we should be OK.
 

(b) Make scatterplots of `salary` against each of the three explanatory variables. If you can, do this with *one* `ggplot`.

Solution


The obvious way to do this is as three separate scatterplots,
and that will definitely work. But you can do it in one go if
you think about facets, and about having all the $x$-values in
one column (and the names of the $x$-variables in another
column):
```{r ivybridge}
salaries %>% gather(xname, x, -salary) %>%
ggplot(aes(x=x, y=salary))+geom_point()+
facet_wrap(~xname, ncol=2, scales="free")
```

       

If you don't see how that works, run it yourself, one line at a time. 

I was thinking ahead a bit while I was coding that: I wanted the three
plots to come out about square, and I wanted the plots to have their
own scales. The last thing in the `facet_wrap` does the latter,
and arranging the plots in two columns (thinking of the plots as a set
of four with one missing) gets them more or less square.

If you don't think of those, try it without, and then fix up what you
don't like.
 

(c) Comment briefly on the direction and strength of each
relationship with `salary`.

Solution


To my mind, all of the three relationships are going uphill (that's the "direction" part). `experience` is the
strongest, and `pubsucc` looks the weakest (that's the
"strength" part). If you want to say there is no relationship
with `pubsucc`, that's fine too. This is a judgement
call. 
Note that all the relationships are more or less linear (no
obvious curves here). We could also investigate the relationships
among the explanatory variables:
```{r }
cor(salaries)
```

       
Mentally cut off the first row and column (`salary` is the
response). None of the remaining correlations are all that high, so we
ought not to have any multicollinearity problems.
 

(d)<a name="regone">*</a> Fit a regression predicting salary from the other three
variables, and obtain a `summary` of the results.

Solution


```{r }
salaries.1=lm(salary~workqual+experience+pubsucc,data=salaries)
summary(salaries.1)
```

   
 

(e) How can we justify the statement 
"one or more of the explanatory variables helps to predict salary"? How is this
consistent with the value of R-squared?

Solution


"One or more of the explanatory variables helps" is an
invitation to consider the (global) $F$-test for the whole
regression. This has the very small P-value of $1.124\times
10^{-10}$ (from the bottom line of the output): very small, so
one or more of the explanatory variables *does* help, and
the statement is correct.
The idea that something helps to predict salary suggests
(especially with such a small number of observations) that we
should have a high R-squared. In this case, R-squared is 0.9109,
which is indeed high.
 

(f) Would you consider removing any of the variables from this
regression? Why, or why not?

Solution


Look at the P-values attached to each variable. These are all
very small: 0.003, 0.00000003 and 0.0003, way smaller than
0.05. So it would be a mistake to 
take any, even one, of the variables out: doing so would make the
regression much worse.
If you need convincing of that, see what happens when we take
the variable with the highest P-value out --- this is `workqual`:
```{r }
salaries.2=lm(salary~experience+pubsucc,data=salaries)
summary(salaries.2)
```

       

R-squared has gone down from 91\% to 86\%: maybe not so much in the
grand scheme of things, but it is noticeably less. Perhaps better,
since we are comparing models with different numbers of explanatory
variables, is to compare the *adjusted* R-squared: this has gone
down from 90\% to 85\%. The fact that this has gone down *at all*
is enough to say that taking out `workqual` was a
mistake.
`r tufte::margin_note("Adjusted R-squareds are easier to compare in this  context, since you don't have to make a judgement about whether it has changed substantially, whatever you think substantially means.")`

Another way of seeing whether a variable has anything to add in a
regression containing the others is a **partial regression  plot**. 
We take the residuals from `salaries.2` above and plot
them against the variable we removed, namely
`workqual`.
`r tufte::margin_note("The residuals have to be the ones from a  regression *not* including the $x$-variable you're testing.")` If
`workqual` has nothing to add, there will be no pattern; if it
*does* have something to add, there will be a trend. Like
this. I use `augment` from `broom`:

```{r dartington}
library(broom)
salaries.2 %>% augment(salaries) %>%
ggplot(aes(x=workqual, y=.resid))+geom_point()
```

 

This is a mostly straight upward trend. So we
need to add a linear term in `workqual` to the
regression.
`r tufte::margin_note("Or not take it out in the first place.")`
 

(g) Do you think it would be a mistake to take *both* of
`workqual` and `pubsucc` out of the regression? Do a
suitable test. Was your guess right?

Solution


I think it would be a big mistake. Taking even one of these
variables out of the regression is a bad idea (from the
$t$-tests), so taking out more than one would be a *really* bad idea.
To perform a test, fit the model without these two explanatory variables:
```{r }
salaries.3=lm(salary~experience,data=salaries)
```

     

and then use `anova` to compare the two regressions, smaller
model first:

```{r }
anova(salaries.3,salaries.1)
```

 

The P-value is extremely small, so the bigger model is definitely
better than the smaller one: we really do need all three
variables. Which is what we guessed.


(h) Back in part (<a href="#regone">here</a>), you fitted a regression with all
three explanatory variables. By making suitable plots, assess
whether there is any evidence that (i) that the linear model should
be a curve, (ii) that the residuals are not normally 
distributed, (iii) that there is "fan-out", where the residuals are getting
bigger *in size* as the fitted values get bigger? Explain
briefly how you came to your conclusions in each case.

Solution


I intended that (i) should just be a matter of looking at residuals
vs.\ fitted values:
```{r }
ggplot(salaries.1, aes(x=.fitted, y=.resid))+geom_point()
```

    

There is no appreciable pattern on here, so no evidence of a curve (or
apparently of any other problems).

Extra: you might read this that we should check residuals against the
$x$-variables as well, which is a similar trick to the above one for
plotting response against each of the explanatories. There is one step
first, though: use `augment` from `broom` first to get a
dataframe with the original $x$-variables *and* the residuals in
it. The following thus looks rather complicated, and if it confuses
you, run the code a piece at a time to see what it's doing:

```{r }
salaries.1 %>% augment(salaries) %>%
gather(xname, x, workqual:pubsucc) %>%
ggplot(aes(x=x, y=.resid))+geom_point()+
facet_wrap(~xname, scales="free", ncol=2)
```

 

These three residual plots are also pretty much textbook random, so no problems here either.

For (ii), look at a normal quantile plot of the residuals, which is not as difficult as the plot I just did:

```{r }
ggplot(salaries.1, aes(sample=.resid))+stat_qq()+stat_qq_line()
```

 

That is really pretty good. Maybe the *second* smallest point is
a bit far off the line, but otherwise there's nothing to worry about. A
quick place to look for problems is the extreme observations, and the
largest and smallest residuals are almost exactly the size we'd expect
them to be.

Our graph for assessing fan-in or fan-out is to plot the *absolute* values of the residuals against the fitted values. The plot from (i) suggests that we won't have any problems here, but to investigate:

```{r }
ggplot(salaries.1, aes(x=.fitted, y=abs(.resid)))+geom_point()+geom_smooth()
```

 

This is pretty nearly straight across. You might think it increases a
bit at the beginning, but most of the evidence for that comes from the
one observation with fitted value near 30 that happens to have a
residual near zero. Conclusions based on one observation are not to be
trusted!
In summary, I'm happy with this linear multiple regression, and I
don't see any need to do anything more with it. I am, however, willing
to have some sympathy with opinions that differ from mine, if they are
supported by those graphs above.
 



