##  Running a maze


 A researcher is trying to design a maze that can be run by
rats in about 60 seconds. One particular maze was run by a sample of
21 rats, with the times shown in
[link](http://www.utsc.utoronto.ca/~butler/c32/maze.txt). 



(a) Read the data into R. What (if anything) are the data values
delimited by?

Solution


Take a look at the data file first. There is only one column of
data, so you can treat it as being delimited by anything you like:
a space, or a comma (the file can also be treated as a
`.csv`), etc.:
```{r }
myurl <- "http://www.utsc.utoronto.ca/~butler/c32/maze.txt"
times <- read_delim(myurl, " ")
times
```

 



(b) Run a sign test, doing it yourself as we did in class:
count the number of values above and below 60, take the *smaller*
of those, and find the probability of a value of that or smaller still
on a binomial distribution with $n=21$ and $p=0.5$ (we have 21 data
points), doubling the answer because the test is two-sided.



Solution



Count how many values are above and below 60:

```{r }
times %>% count(time > 60)
```

 
5 above and 16 below. Then find out how likely it is that a binomial
with $n=21, p=0.5$ would produce 5 or fewer successes:

```{r }
p <- sum(dbinom(0:5, 21, 0.5))
p
```

 

or if you prefer count upwards from 16:

```{r }
sum(dbinom(16:21, 21, 0.5))
```

 

and double it to get a two-sided P-value:

```{r }
2 * p
```

 

We'll compare this with `smmr` in a moment.




(c) Install my package `smmr`, if you haven't already. To do
this, you first need to install the package `devtools` (if you
haven't already),
by going to the console and typing

```{r eval=F}
install.packages("devtools")
```

 

When that's all done, install `smmr` thus:

```{r eval=F}
library(devtools)
install_github("nxskok/smmr")
```

 

That all needs to be done only once. Then, each R Studio session where
you want to use `smmr` needs this:

```{r }
library(smmr)
```

 

As usual, only the `library` thing only needs to be done every
time. 

When you have `smmr` installed, use `sign_test` from
that package to re-run your sign test. Do you get the same P-value?



Solution


The sign test function takes a data frame, an (unquoted) column
name from that data frame of data to test the median of, and a
null median (which defaults to 0 if you omit it):
```{r }
library(smmr)
sign_test(times, time, 60)
```

     

This shows you two things: a count of the values below and above the
null median, and then the P-values according to the various
alternative hypotheses you might have. 

In our case, we see again the 16 maze-running times below 60 seconds
and 5 above (one of which was a long way above, but we don't care
about that here). We were testing whether the median was different
from 60, so we look at the two-sided P-value of 0.0266, which is
exactly what we had before.

If `sign_test` doesn't work for you (perhaps because it needs
a function `enquo` that you don't have), there is an
alternative function `sign_test0` that doesn't use it. It
requires as input a *column* of values (extracted from the data
frame) and a null median, thus:

```{r }
with(times, sign_test0(time, 60))
```

 

The output should be, and here is, identical.




(d) Package `smmr` also has a function
`pval_sign`, which has the same input as
`sign_test`, but with the null median *first*.
Run it on your data and see what it gives.



Solution


Try it and see:
```{r }
pval_sign(60, times, time)
```

     

The two-sided P-value, and that is all. We'll be using this in a minute.

Alternatively, there is also this, which needs a null median and a
*column* as input:

```{r }
with(times, pval_sign0(60, time))
```

 




(e) Obtain a 95\% confidence interval for the median based on these
data. Do this two ways. First, use the trial and error way from class
(either the try-lots-of-values way or the bisection way; either is good).
Second, use `ci_median` from `smmr`. The latter takes
as input a data frame, a column name (unquoted) and optionally a
`conf.level` that defaults to 0.95.



Solution


The reason for showing you `pval_sign` in the previous
part is that this is a building block for the confidence interval.
What we do is to try various null medians
and find out which ones give P-values less than 0.05 (outside the
interval) and which ones bigger (inside). 
We know that the value 60 is
outside the 95\% CI, and the sample median is close to 50 (which we
expect to be inside), so sensible values to try for the upper end of
the interval would be between 50 and 60:

```{r }
pval_sign(58, times, time)
pval_sign(55, times, time)
```

 

So, 55 is inside the interval and 58 is outside. I could investigate
further in similar fashion, but I thought I would try a whole bunch of null
medians all at once. That goes like this, `rowwise` because `pval_sign` expects *one* null-hypothesis median, not several all at once:

```{r }
tibble(meds = seq(55, 58, 0.25)) %>% 
  rowwise() %>% 
  mutate(pvals = pval_sign(meds, times, time))
```

 

So values for the median all the way up to and including 57.5 are in
the confidence interval.


Now for the other end of the interval. I'm going to do this a
different way: more efficient, but less transparent. The first thing I
need is a pair of values for the median: one inside the interval and
one outside. Let's try 40 and 50:

```{r }
pval_sign(40, times, time)
pval_sign(50, times, time)
```

 

OK, so 40 is outside and 50 is inside. So what do I guess for the next
value to try? I could do something clever like assuming that the
relationship between hypothesized median and P-value is *linear*,
and then guessing where that line crosses 0.05. But I'm going to
assume *nothing* about the relationship except that it goes
uphill, and therefore crosses 0.05 somewhere. So my next guess is
halfway between the two values I tried before:

```{r }
pval_sign(45, times, time)
```

 

So, 45 is inside the interval, and my (slightly) improved guess at the
bottom end of the interval is that it's between 40 and 45. So next, I
try halfway between *those*:

```{r }
pval_sign(42.5, times, time)
```

 

42.5 is outside, so the bottom end of the interval is between 42.5 and 45.

What we are doing is narrowing down where the interval's bottom end
is. We started by knowing it to within 10, and now we know it to
within 2.5. So if we keep going, we'll know it as accurately as we wish.

This is called a "bisection" method, because at each step, we're
dividing our interval by 2.

There is one piece of decision-making at each step: if the P-value for
the median you try is greater than 0.05, that becomes the top end of
your interval (as when we tried 45); if it is less, it becomes the
bottom end (when we tried 42.5).

This all begs to be automated into a loop. It's not a
`for`-type loop, because we don't know how many times we'll be
going around. It's a `while` loop: keep going while something
is true. Here's how it goes:

```{r }
lo <- 40
hi <- 50
while (abs(hi - lo) > 0.1) {
  try <- (hi + lo) / 2
  ptry <- pval_sign(try, times, time)
  print(c(try, ptry))
  if (ptry < 0.05) {
    lo <- try
  } else {
    hi <- try
  }
}
lo
pval_sign(lo, times, time)
hi
pval_sign(hi, times, time)
```

 

The loop stopped because 44.297 and 44.375 are less than 0.1
apart. The first of those is outside the interval and the second is
inside. So the bottom end of our interval is 44.375, to this
accuracy. If you want it more accurately, change 0.1 in the
`while` line to something smaller (but then you'll be waiting
longer for the answer). 

I put the `print` statement in the loop so that you could see
what values were being tried, and what P-values they were
producing. What happens with these is that the P-value jumps at each
data value, so you won't get a P-value exactly 0.05; you'll get one
above and one below.

Likewise, you can use the function with a zero on its name and feed it
a column rather than a data frame and a column name:

**********************************

```{r }
tibble(meds =  seq(55, 58, 0.25)) %>% 
  rowwise() %>% 
  mutate(pvals =  with(times, pval_sign0(meds, time)))
```

 

Or adapt the idea I had above for bisection.
All that was a lot of work, but I wanted you to see it all once, so that you
know where the confidence interval is coming from. `smmr` also
has a function `ci_median` that does all of the above without
you having to do it. As I first wrote it, it was using the trial and
error thing with `rowwise`, but I chose to rewrite it with the
bisection idea, because I thought that would be more accurate.

```{r }
ci_median(times, time)
```

 

This is a more accurate interval than we got above. (The
`while` loop for the bisection keeps going until the two
guesses at the appropriate end of the interval are less than 0.01
apart, by default.)
`r tufte::margin_note("You can change this by adding something like *tol=1e-4* to the end of your *ci-median*.")` 

If you want some other confidence level, you add `conf.level`
on the end, as you would for `t.test`:

```{r }
ci_median(times, time, conf.level = 0.75)
```

 

A 75\% CI, just for fun. This is a shorter interval than the 95\% one,
as it should be.

Likewise there is a `ci_median0` that takes a column and an
optional confidence level:

```{r }
with(times, ci_median0(time))
with(times, ci_median0(time, conf.level = 0.75))
```

 

with the same results. Try `ci_median` first, and if it
doesn't work, try `ci_median0`.




