##  Running a maze


 A researcher is trying to design a maze that can be run by
rats in about 60 seconds. One particular maze was run by a sample of
21 rats, with the times shown in
[link](http://ritsokiguess.site/datafiles/maze.txt). 



(a) Read the data into R. What (if anything) are the data values
delimited by?


(b) Run a sign test, doing it yourself as we did in class:
count the number of values above and below 60, take the *smaller*
of those, and find the probability of a value of that or smaller still
on a binomial distribution with $n=21$ and $p=0.5$ (we have 21 data
points), doubling the answer because the test is two-sided.




(c) Install my package `smmr`, if you haven't already. To do
this, you first need to install the package `devtools` (if you
haven't already),
by going to the console and typing

```{r maze-qq-1, eval=F}
install.packages("devtools")
```

 

When that's all done, install `smmr` thus:

```{r maze-qq-2, eval=F}
library(devtools)
install_github("nxskok/smmr")
```

 

That all needs to be done only once. Then, each R Studio session where
you want to use `smmr` needs this:

```{r maze-qq-3 }
library(smmr)
```

 

As usual, only the `library` thing only needs to be done every
time. 

When you have `smmr` installed, use `sign_test` from
that package to re-run your sign test. Do you get the same P-value?




(d) Package `smmr` also has a function
`pval_sign`, which has the same input as
`sign_test`, but with the null median *first*.
Run it on your data and see what it gives.





(e) Obtain a 95\% confidence interval for the median based on these
data. Do this two ways. First, use the trial and error way from class
(either the try-lots-of-values way or the bisection way; either is good).
Second, use `ci_median` from `smmr`. The latter takes
as input a data frame, a column name (unquoted) and optionally a
`conf.level` that defaults to 0.95.





