##  Mental context


 It is believed that being in the same mental context for
learning and for testing leads to better test scores. An experiment
was carried out to test this. During the learning phase, subjects
learned a list of 80 words in a room painted orange and decorated with
posters, paintings and other paraphernalia.
`r tufte::margin_note("This is a fancy  word for *stuff*.")` A memory test was given to all subjects
immediately after they had learned the words, to give the impression
that the experiment was over. (The results of this test were
discarded.) One day later, subjects were unexpectedly re-tested under
different experimental conditions and asked to write down all the
words of the original list that they could remember. The re-test took
place in five different conditions, which subjects were randomly
allocated to one of:



* Same context: in the original orange-painted room.

* Different context: in a very different room, painted grey and
located in a different part of campus.

* Imaginary context: in the grey room, but subjects are asked to
remember the orange room in which they took the original test. To
help them with this, the experimenter begins by asking them several
questions about the orange room and the objects in it.

* Photographed context: in the grey room, but subjects are shown
photographs of the orange room.

* Placebo context: in the grey room, with subjects first being
asked to recall their living room.


In each case, the response variable was the number of words on the
original list successfully recalled by each subject.

50 subjects in total completed the experiment, 10 under each of the 5
conditions. 

The researchers had four research hypotheses to test with the data:



* Groups where the context matches (is the same, or is simulated
by imagining or photograph) will perform better than groups with
different or placebo contexts.

* The group with the same context will differ from the group with
imaginary or photographed contexts.


* The imaginary-context group will differ from the
photographed-context group.

* The different-context group will differ from the placebo-context group.


The research hypotheses are (as is usual) written as
*alternative* hypotheses. We can rewrite them as null
hypotheses, with some extra wording to facilitate converting them
into contrasts, like this:


* The mean of the `same`, `imaginary` and
`photograph` groups (group means) is equal to the mean of the
`different` and `placebo` groups (group means).

* The mean of the `imaginary` and `photograph`
groups is equal to the (mean of the) `same` group.

* The `imaginary` and the `photograph` groups will
have the same mean.

* The `different` and the `placebo` groups will
have the same mean.

The data are in
[link](http://ritsokiguess.site/datafiles/smith.txt) (the
original researcher's name was Smith). 


(a) Read in the data and verify that you have a column called
`context` that is text and a column called
`words` that is a (whole) number.


Solution


The usual thing --- read in the data appropriately and look at the
data frame you got:
```{r }
my_url <- "http://ritsokiguess.site/datafiles/smith.txt"
smith <- read_delim(my_url, " ")
smith
```

       

`dbl` actually means "decimal number", but these look like whole numbers, so I think we are good.
      


(b) Turn `context` into a `factor`, within the
data frame. (We are going to be doing contrasts). Display how the
data frame now looks.


Solution


Use `mutate` and assign everything back to what it was
before:
```{r }
smith <- smith %>% mutate(context = factor(context))
smith
```

    

Note that `context` is now labelled as the factor that we made.



(c) What are the names of the five contexts in the data set
(just display them),
and what order are they in?


Solution


You have lots of choices now, since `context` is a factor
(and can also be treated as text). 
The most obvious factor-based way is this:
```{r }
levels(smith$context)
```

       

or use `summary`:

```{r }
smith %>% select(context) %>% summary()
```

 

Or treat `context` as text, such as this way:
```{r }
smith %>% count(context)
```

       

These all display the contexts in alphabetical order.

Or display the distinct values:

```{r }
smith %>% distinct(context)
```

 

or, without using the `tidyverse`,

```{r }
unique(smith$context)
```

 
These two *don't* display things in alphabetical order, because
they display things as they appear in the file. This is not the order
we want, though the second way *does* display the `Levels`
in the right order.
      


(d) Write each of the four research hypotheses (in the
null-hypothesis versions) as R vectors that can be used to make
contrasts. (This will mean getting the contexts in the right
order. If you get stuck, do the last two first, since they're
easiest. The first one is actually the hardest.)


Solution


All right, let's go from the bottom:


* Different and placebo have same means. These are 1st and 4th.
```{r }
c4 <- c(1, 0, 0, -1, 0)
```

         

* Imaginary and photograph have same means. 2nd and 3rd:
```{r }
c3 <- c(0, 1, -1, 0, 0)
```

   


* Imaginary and photograph together same as `same`. This
time we have two means being compared with one, so we need to give
the two means half weight. 2nd and 3rd against 5th:
```{r }
c2 <- c(0, 1 / 2, 1 / 2, 0, -1)
```

   


* (Same and imaginary and photograph) vs.\ (different or placebo):
2nd, 3rd and 5th with weight $1/3$ (three of them) against 1st and
4th with weight $1/2$ (2 of them):
```{r }
c1 <- c(-1 / 2, 1 / 3, 1 / 3, -1 / 2, 1 / 3)
```

   

Variations: you can switch the sign on everything within a
contrast (since it doesn't matter which is plus and which is
minus). You can also multiply through by anything to get rid of
the fractions, for example these:
```{r eval=F}
c2 <- c(0, 1, 1, 0, -2)
c1 <- c(-3, 2, 2, -3, 2)
```

       

All that matters is that the coefficients add up to zero, and that they are the right size and sign relative to each other.
      


(e) Pick two of your contrasts (doesn't matter which two) and
demonstrate that they are orthogonal.


Solution


Multiply your chosen contrasts together elementwise, and show
that the results add to zero, eg. by showing the whole thing:
```{r }
c1 * c2
```

    
which adds to zero because it is $2/6$ minus $1/3$,
or by explicitly summing the elementwise product:

```{r }
sum(c1 * c3)
```

 

Any pair is good.
      


(f) Collect your contrasts together into a matrix, and
tell `lm` that these are the contrasts for `context`.


Solution


This:
```{r }
m <- cbind(c1, c2, c3, c4)
contrasts(smith$context) <- m
```

       

Slipping off into another aside, you might have been wondering whether
there was a way to test 
*all* the contrasts for orthogonality at once. There is, and it
depends on the rules for matrix multiplication. We want to test all
the *columns* of `m` for orthogonality, but matrix
multiplication works by combining a *row* with a column. No
matter, *transposing* a matrix interchanges rows and columns, so
that in math, we want to look at the matrix $M^T M$. In R,
`%*%` means "matrix multiply".
`r tufte::margin_note("In R, percents around something mean that it is a special version of that something. Hence the notation for matrix-multiply and the pipe symbol. A regular * when used for multiplying matrices in R will multiply them element by element.")` Thus,

```{r }
t(m) %*% m
```

 
This matrix-multiplies the transpose of `m` by `m`
itself. There are numbers down the top-left to bottom-right diagonal,
but don't worry about these, since a contrast doesn't have to be
orthogonal with itself. The thing to note is that 
*all the other  elements* of this matrix are 
*zero*: that means that each of
the four contrasts is orthogonal to each of the other three.
      


(g) Fit a model with `lm`, and display the results.


Solution


We're past the hard part:
```{r }
smith.1 <- lm(words ~ context, data = smith)
summary(smith.1)
```

       
      


(h) For each of the original research hypotheses, what do you
infer about them?


Solution


Even though I generated the contrasts backwards, I maintained
the numbering so that they refer to the original numbered
hypotheses. You might have named them something more mnemonic;
that works too, and is quite possibly better.
Anyway:


* Matching context better than non-matching context:
strongly significant (P-value 0.000085). (Strictly, this is a
two-sided test, but the way I constructed the contrast, this
should be significantly *positive* if the research
hypothesis is true, and it is.)

* Same context *not* different from
imaginary/photographed context (P-value 0.823)

* Imaginary context *not* different from photographed context (P-value 0.938).

* Different context *not* different from placebo
context (P-value 0.699). 


I wanted you to match up the research hypotheses with the
P-values, as above, and state a conclusion about each one. If
you do that, I am happy.

Taking this back to the original research, I think the
first hypothesis was the most immediately important of the four:
we were able to show that having (or faking up) the original
room helped with recall. After that, it didn't matter how it was
done: being in the original room was not different from
imagining the original room (by thinking about it or seeing a
photo of it). Failing to recall the original room was equally
bad, whether the subjects were in a different room and not asked
to think about the original orange room, or whether they were
literally asked to think about another room entirely.

Contrasts enabled us to tackle this problem and gain a very clear
conclusion: recalling the original orange room is helpful, but
it doesn't matter how you do it.

How big of a difference does thinking about the orange room
make? You can use the 2 SE thing to get a 95\% confidence interval:
the difference in the (mean of the) recall-orange-room means and
the (mean of the) don't-recall means is between about 4.6 and
12.6 words (out of 80). I leave it for you to decide whether
that is practically important as well as being statistically significant.

      



