##  Growth of Mizuna lettuce seeds


 In 2010, a group of students planted some Mizuna lettuce
seeds, and recorded how they grew. The data were saved in an Excel
spreadsheet, which is at
[link](http://ritsokiguess.site/datafiles/mizuna.xlsx). The columns
are: the date, the height (in cm) of (I presume) the tallest plant,
the amount of water added since the previous date (ml), the
temperature in the container where the seedlings were growing, and any
additional notes that the students made (edited for length by me). The
top line of the data file is variable names.



(a) Read the
spreadsheet data.


Solution


This is `read_excel` from package `readxl`. I'm
not sure what will happen to the dates yet. Note that this needs
a "local" copy of the spreadsheet (that is, you have to
download it and save it on your computer, then upload it to
`rstudio.cloud`), possibly using `file.choose` to
help R find it. I put my copy in the same project folder as I was
working in, so I just need the file name:
```{r mizuna-r-1 }
library(readxl)
mizuna <- read_excel("mizuna.xlsx")
mizuna
```

     

The dates *did* get read properly. `dttm` is
"date-time", so I guess it's allowing for the possibility that my
dates had times attached as well. Unlike with SAS (as we see later),
the years came out right.

$\blacksquare$

(b) Make a plot of `height` against your
dates, with the points joined by lines.


Solution


```{r mizuna-r-2 }
ggplot(mizuna, aes(x = date, y = height)) + geom_point() + geom_line()
```

       

$\blacksquare$

(c) Label each point on the plot with the
amount of water added up to that point.


Solution


This is `water` again. The way to do this is to load
`ggrepel`, then add `geom_text_repel` to the
plot, by adding `label=water` to the *original*
`aes`:
```{r mizuna-r-3 }
library(ggrepel)
ggplot(mizuna, aes(x = date, y = height, label = water)) +
  geom_point() + geom_line() + geom_text_repel(colour = "red")
```

 

I made the text red, so that you can see it more easily. It "repels"
away from the points, but not from the lines joining them. Which makes
me wonder whether this would work better (I explain `alpha` afterwards):

```{r mizuna-r-4 }
library(ggrepel)
ggplot(mizuna, aes(x = date, y = height, label = water)) +
  geom_point() + geom_line() + geom_label_repel(colour = "red", alpha = 0.7)
```

 

The difference between `text` and `label` is that
`text` just uses the text of the variable to mark the point,
while `label` puts that text in a box.

I think it works better. You can see where the line goes (under the boxes with
the labels in them), but you can see the labels clearly.

What that `alpha` does is to make the thing it's attached to
(the labels) partly *transparent*. If you leave it out (try it),
the black line disappears completely under the label boxes and you
can't see where it goes at all. The value you give for `alpha`
says how transparent the thing is, from 1 (not transparent at all)
down to 0 (invisible). I first tried 0.3, and you could hardly see the
boxes; then I tried 0.7 so that the boxes were a bit more
prominent but the lines underneath were still slightly visible, and I
decided that this is what I liked. I think making the labels a
different colour was a good idea, since that helps to distinguish the
number on the label from the line underneath.

You can apply `alpha` to pretty much any `ggplot` thing
that might be on top of something else, to make it possible to see
what's underneath it. The commonest use for it is if you have a
scatterplot with a lot of points; normally, you only see some of the
points, because the plot is then a sea of black. But if you make the
points partly transparent, you can see more of what's nearby that
would otherwise have been hidden. 

At some point, I also have to show
you folks `jitter`, which plots in slightly different places
points that would otherwise overprint each other exactly, and you
wouldn't know how many of them there were, like the outliers on the
boxplots of German children near the new airport.

$\blacksquare$


