##  Do you like your mobile phone?


 A phone company commissioned a survey of their customers'
satisfaction with their mobile devices. The responses to the survey
were on a so-called Likert scale of "very unsatisfied",
"unsatisfied", "satisfied", "very satisfied". Also recorded were
each customer's gender and age group (under 18, 18--24, 25--30, 31 or
older). (A survey of this kind does not ask its respondents for their
exact age, only which age group they fall in.) The data, as
frequencies of people falling into each category combination, are in [link](http://www.utsc.utoronto.ca/~butler/d29/mobile.txt).



(a) <a name="part:gather">*</a> Read in the data and take a look at the format. Use a tool
that you know about to arrange the frequencies in *one* column,
with other columns labelling the response categories that the
frequencies belong to. Save the new data frame. (Take a look at it
if you like.)
 
Solution


```{r }
my_url="http://www.utsc.utoronto.ca/~butler/d29/mobile.txt"
mobile=read_delim(my_url," ")
mobile
```

 
With multiple columns that are all frequencies, this is a job for
`gather`:
```{r }
mobile.long = mobile %>% gather(satisfied,frequency,very.unsat:very.sat)
mobile.long
```

     

Yep, all good. See how `mobile.long` contains what it should?
(For those keeping track, the original data frame had 8 rows and 4
columns to gather, and the new one has $8\times 4=32$ rows.)
 


(b) We are going to fit ordered logistic models below. To
do that, we need our response variable to be a factor with its levels
in the right order. By looking at the data frame
you just created, determine what kind
of thing your intended response variable currently is.



Solution


I looked at `mobile.long` in the previous part, but if you
didn't, look at it here:

```{r }
mobile.long
```

     

My intended response variable is what I called `satisfied`.
This is `chr` or "text", not the `factor` that I
want. 




(c) If your intended response variable is not a factor, create a factor in your data frame with levels in the right order. Hint: look at the order your levels are *in the data*.



Solution


My intended response `satisfied` is text, not a factor, so
I need to do this part.
The hint is to look at the column `satisfied` in
`mobile.long` and note that the satisfaction categories
appear in the data *in the order that we want*. This is good
news, because we can use `fct_inorder` like this:
```{r }
mobile.long %>%
mutate(satis=fct_inorder(satisfied)) -> mobile.long
```

     

If you check, by looking at the data frame, `satis` is 
a `factor`, and you can also do this to verify that its levels
are in the right order:

```{r }
with(mobile.long, levels(satis))
```

 
Success.

Extra: so now you are asking, what if the levels are in the *wrong* order in the data? Well, below is what you used to have to do, and it will work for this as well.
I'll first find what levels of satisfaction I have. This
can be done by counting them, or by finding the distinct ones:
```{r }
mobile.long %>% count(satisfied)
```

     

or

```{r }
mobile.long %>% distinct(satisfied)
```

 

If you count them, they come out in alphabetical order. If you ask for
the distinct ones, they come out in the order they were in
`mobile.long`, which is the order the *columns* of those
names were in `mobile`, which is the order we want. 

To actually grab those satisfaction levels as a vector (that we will
need in a minute), use `pluck` to pull the column out of the
data frame as a vector:

```{r }
v1 = mobile.long %>% 
distinct(satisfied) %>%
pluck("satisfied")
v1
```

 

which is in the correct order, or

```{r }
v2 = mobile.long %>% 
count(satisfied) %>%
pluck("satisfied")  
v2
```

 

which is in alphabetical order. The problem with the second one is
that we know the correct order, but there isn't a good way to code
that, so we have to rearrange it ourselves. The correct order from
`v2` is 4, 2, 1, 3, so:

```{r }
v3=c(v2[4],v2[2],v2[1],v2[3])
v3
v4=v2[c(4,2,1,3)]
v4
```

 

Either of these will work. The first one is more typing, but is
perhaps more obvious. There is a third way, which is to keep things as
a data frame until the end, and use `slice` to pick out the
rows in the right order:

```{r }
v5 = mobile.long %>% 
count(satisfied) %>%
slice(c(4,2,1,3)) %>%  
pluck("satisfied")  
v5
```

 

If you don't see how that works, run it yourself, one line at a time.

The other way of doing this is to physically type them into a vector,
but this carries the usual warnings of requiring you to be very
careful and that it won't be reproducible (eg. if you do another
survey with different response categories). 

So now create the proper response
variable thus, using your vector of categories:

```{r }
mobile.long %>%
mutate(satis=ordered(satisfied,v1)) -> mobile.long2
mobile.long2
```

 

`satis` has the same values as `satisfied`, but its
label `ord` means that it is an ordered factor, as we want.



(d) <a name="part:thefit">*</a>
Fit ordered logistic models to predict satisfaction from (i) gender
and age group, (ii) gender only, (iii) age group only. (You don't
need to examine the models.) Don't forget a suitable
`weights`!
 
Solution


(i):

```{r }
library(MASS)
mobile.1=polr(satis~gender+age.group,weights=frequency,data=mobile.long)
```

 

For (ii) and (iii), `update` is the thing (it works for any
kind of model):

```{r }
mobile.2=update(mobile.1,.~.-age.group)
mobile.3=update(mobile.1,.~.-gender)
```

 

We're not going to look at these, because the output from
`summary` is not very illuminating. What we do next is to try
to figure out which (if either) of the explanatory variables
`age.group` and `gender` we need.
 


(e) Use `drop1` on your model containing both explanatory
variables to determine whether you can remove either of them. Use
`test="Chisq"` to obtain P-values.



Solution


`drop1` takes a fitted model, and tests each term in
it in turn, and says which (if any) should be removed. Here's how it goes:

```{r }
drop1(mobile.1,test="Chisq")
```

 

The possibilities are to remove `gender`, to remove
`age.group` or to remove nothing. 
The best one is "remove nothing", because it's the one on the output with the smallest
AIC. Both P-values are small, so it would be a mistake to remove
either of the explanatory variables.



(f) Use `anova` to decide whether we are justified in
removing `gender` from a model containing both `gender`
and `age.group`. Compare your P-value with the one from `drop1`.
 
Solution


This is a comparison of the model with both variables
(`mobile.1`) and the model with `gender` removed
(`mobile.3`). Use `anova` for this, smaller
(fewer-$x$) model first:
```{r }
anova(mobile.3,mobile.1)
```

     

The P-value is (just) less than 0.05, so the models are significantly
different. That means that the model with both variables in fits
significantly better than the model with only `age.group`, and
therefore that taking `gender` out is a mistake.

The P-value is identical to the one from `drop1` (because they
are both doing the same test).
 

(g) Use `anova` to see whether we are justified in removing
`age.group` from a model  containing both `gender` and
`age.group`. Compare your P-value with the one from
`drop1` above.
 
Solution


Exactly the same idea as the last part. In my case, I'm comparing
models `mobile.2` and `mobile.1`:
```{r }
anova(mobile.2,mobile.1)
```

    

This one is definitely significant, so I need to keep
`age.group` for sure. Again, the P-value is the same as the one
in `drop1`.
 

(h) Which of the models you have fit so far is the most
appropriate one? Explain briefly.
 
Solution


I can't drop either of my variables, so I have to keep them both:
`mobile.1`, with both `age.group` and `gender`.
  

(i) Obtain predicted probabilities of a
customer falling in the various satisfaction categories, as it
depends on gender and age group. To do that, you need to feed
`predict` three things: the fitted model that contains both
age group and gender, the data frame that you read in from the file
back in part (<a href="#part:gather">here</a>) (which contains all the combinations of age group
and gender), and an appropriate `type`.

 
Solution


My model containing both $x$s was `mobile.1`, the data frame
read in from the file was called `mobile`, and I  need
`type="p"` to get probabilities:
```{r }
probs=predict(mobile.1,mobile,type="p")
mobile %>% select(gender, age.group) %>%
cbind(probs)
```

   

This worked for me, but this might happen to you, with the same commands as above:

```{r error=T, echo=F}
probs=predict(mobile.1,mobile,type="p")
mobile %>% MASS::select(gender, age.group) %>%
cbind(probs)
```

 

Oh, this didn't work. Why not? There don't seem to be any errors.

This is the kind of thing that can bother you for *days*. The
resolution (that it took me a long time to discover) is that you might
have the `tidyverse` *and also* `MASS` loaded, in
the wrong order, and `MASS` also has a `select` (that
takes different inputs and does something different). If you look back
at part (<a href="#part:thefit">here</a>), you might have seen a message there when you
loaded `MASS` that `select` was "masked". When you
have two packages that both contain a function with the same name, the
one that you can see (and that will get used) is the one that was
loaded *last*, which is the `MASS` select (not the one we
actually wanted, which is the `tidyverse` select). There are a
couple of ways around this. One is to un-load the package we no longer
need (when we no longer need it). The mechanism for this is shown at
the end of part (<a href="#part:unload">here</a>). The other is to say explicitly
which package you want your function to come from, so that there is no
doubt. The `tidyverse` is actually a collection of
packages. The best way to find out which one our `select` comes
from is to go to the Console window in R Studio and ask for the help
for `select`. With both `tidyverse` and `MASS`
loaded, the help window offers you a choice of both `select`s;
the one we want is "select/rename variables by name", and the actual
package it comes from is `dplyr`.

With that in mind, we can get the right `select` like this:

```{r }
probs=predict(mobile.1,mobile,type="p")
mobile %>% dplyr::select(gender, age.group) %>%
cbind(probs)
```

   

This gets the right `select` for sure: package name, two
colons, function name means 
"the function by that name that lives in that package".

This is a non-standard use of `cbind` because I wanted to grab
only the gender and age group columns from `mobile` first, and
then `cbind` *that* to the predicted probabilities. The
missing first input to `cbind` is 
"whatever came out of the previous step", 
that is, the first two columns of `mobile`.

I only included the first two columns of `mobile` in the
`cbind`, because the rest of the columns of `mobile`
were frequencies, which I don't need to see. (Having said that, it
would be interesting to make a *plot* using the observed
proportions and predicted probabilities, but I didn't ask you for that.)

This is an unusual `predict`, because we didn't have to make a
data frame (with my usual name `new`) containing all the
combinations of things to predict for. We were lucky enough to have
those already in the original data frame `mobile`. 

The usual way to do this is something like the trick that we did for
getting the different satisfaction levels:

```{r }
genders = mobile.long %>% distinct(gender) %>% pluck("gender")
age.groups = mobile.long %>% distinct(age.group) %>%
pluck("age.group")
```

 

This is getting perilously close to deserving a function written to do
it (strictly speaking, we should, since this is three times we've
used this idea now).

Then `crossing` to get the combinations, and then
`predict`: 

```{r }
new=crossing(gender=genders,age.group=age.groups)
new
pp=predict(mobile.1,new,type="p")
cbind(new,pp)
```

 
This method is fine if you want to do this
question this way; the way I suggested first ought  to be easier, but 
the nice thing about this is its mindlessness: you always do it about
the same way.

 


(j) <a name="part:unload">*</a> 
Describe any patterns you see in the predictions, bearing in mind the
significance or not of the explanatory variables.

 
Solution


I had both explanatory variables being significant, so I would
expect to see both an age-group effect *and* a gender effect.
For both males and females, there seems to be a decrease in
satisfaction as the customers get older, at least until age 30 or
so. I can see this because the predicted prob.\ of "very  satisfied" 
decreases, and the predicted prob.
of "very unsatisfied" increases. The 31+ age group are very
similar to the 25--30 group for both males and females. So that's
the age group effect.
What about a gender effect? Well, for all the age groups, the males
are more likely to be very satisfied than the females of the
corresponding age group, and also less likely to to be very
unsatisfied. So the gender effect is that males are more satisfied
than females overall. (Or, the males are less discerning. Take your pick.)
When we did the tests above, age group was very definitely
significant, and gender less so (P-value around 0.03). This
suggests that the effect of age group ought to be large, and the
effect of gender not so large. This is about what we observed: the
age group effect was pretty clear, and the gender effect was
noticeable but small: the females were less satisfied than the
males, but there wasn't all that much difference.
Now we need to "unload" MASS, because, as we saw, *it* has a
`select` that we don't want to have
interfering with the `dplyr` one:
```{r }
detach("package:MASS",unload=T)
```

    
 



