##  Exercise and anxiety and bullying mice


 Does exercise help to reduce anxiety?
To assess this, some researchers randomly assigned
mice to either 
an enriched environment where there was an exercise wheel available,
or a standard environment with no exercise options. After three weeks
in the specified environment, for five minutes a day for two weeks,
the mice were each exposed to a "mouse bully" --- a mouse who was very
strong, aggressive, and territorial. One measure of mouse anxiety is
amount of time hiding in a dark compartment, with mice who are more
anxious spending more time in darkness. The amount of time spent in
darkness is recorded for each of the mice. 

The data can be found at
[link](http://www.utsc.utoronto.ca/~butler/d29/stressedmice.txt). 



(a)[2] Read the data into R, and display your data frame. Count the number of mice in each group. 

Solution


These are aligned columns with spaces in between, so we need `read_table`:
```{r }
my_url="http://www.utsc.utoronto.ca/~butler/d29/stressedmice.txt"
mice=read_table(my_url)
mice
```

   

You can call the data frame whatever you like.

If you must, you can physically count the number of mice in each group, but you ought to get in the habit of coding this kind of thing:

```{r }
mice %>% count(Environment)
```

 

Seven in each.
 


(b)[2] Draw side-by-side boxplots of time spent in darkness for each
group of mice.

Solution


This:
```{r artichoke}
ggplot(mice, aes(x=Environment, y=Time))+geom_boxplot()
```

     

You did remember to put capital letters on the variable names, didn't you?
 


(c)[2] Do the boxplots support the hypothesis about exercise and
anxiety? Explain briefly.

Solution


The hypothesis about exercise and anxiety is that mice who
exercise more should be less anxious. How does that play out in
this study? Well, mice in the enriched environment at least have
the opportunity to exercise, which the mice in the standard
environment do not, and anxiety is measured by the amount of time
spent in darkness (more equals more anxious).
So we'd expect the mice in the standard environment to spend more
time in darkness, if that hypothesis is correct.
That's exactly what the boxplots show, with very little
doubt.
`r tufte::margin_note("This means that I would expect to reject a null hypothesis of equal means, but I get ahead of myself.")`
Your answer needs to make two points: (i) what you would expect to
see, if the hypothesis about anxiety and exercise is true, and
(ii) whether you actually did see it. You can do this either way
around: for example, you can say what you see in the boxplot, and
then make the case that this *does* support the idea of more
exercise corresponding with less anxiety.
 


(d)[2] Carry out a $t$-test for comparing the mean time spent in
darkness for the mice in the two groups. Think carefully about the
details of the $t$-test (and what you need evidence in favour of).

Solution


We are trying to prove that exercise goes with *less*
anxiety, so a one-sided test is called for. The other thing to
think about is how R organizes the groups for
`Environment`: in alphabetical order. Thus
`Enriched` is first (like on the boxplot). We're trying to
prove that the mean `Time` is *less* for
`Enriched` than for `Standard`, so we need `alternative="less"`:
```{r }
with(mice,t.test(Time~Environment,alternative="less"))
```

     
 


(e)[2] What do you conclude, in terms of anxiety and exercise (at
least for mice)? Explain  briefly.

Solution


The P-value (from the previous part) is 0.000037, which is way less
than 0.05 (or 0.01 or whatever $\alpha$ you chose). So the null
hypothesis (equal means) is resoundingly rejected in favour of the
one-sided alternative that the mean anxiety (as measured by time
spent in darkness) is less for the mice who (can) exercise.
You need to end up by doing a one-sided test. An alternative to
what I did is to do a two-sided test in the previous part. Then
you can fix it up by recognizing that the means are the right way
around for the research hypothesis (the mean time in darkness is
way less for `Enriched`), and then dividing the two-sided
P-value by 2. But you need to do the "correct side" thing: just
halving the two-sided P-value is not enough, because the sample
mean for `Enriched` might have been *more* than for
`Standard`. 
 


(f)[2] Does anything in the previous parts suggest any problems with
the analysis you just did? Explain briefly.

Solution


Look at the side-by-side boxplots. The strict assumptions hiding
behind the $t$-tests are that the data in each group come from
normal distributions (equal standard deviations are not
required). Are the data symmetric? Are there any outliers? Well, I
see a high outlier in the `Enriched` group, so I have some
doubts about the normality. On the other hand, I only have seven
observations in each group, so there is no guarantee even if the
populations from which they come are normal that the samples will
be. So maybe things are not so bad.
This is one of those situations where you make a case and defend it. I
don't mind so much which case you make, as long as you can defend
it. Thus, something like either of these two is good:


* I see an outlier in the `Enriched` group. The data
within each group are supposed to be normally distributed, and
the `Enriched` group is not. So I see a problem.

* I see an outlier in the `Enriched` group. But the
sample sizes are small, and an apparent outlier could arise by
chance. So I do not see a problem.

Extra: another way to think about this is normal quantile plots to assess
normality within each group. This uses the facetting trick to get a separate normal quantile plot for each `Environment`:
```{r }
ggplot(mice, aes(sample=Time))+stat_qq()+stat_qq_line()+
facet_wrap(~Environment, scales="free")
```

     

For the `Enhanced` group, the upper-end outlier shows up. In a
way this plot is no more illuminating than the boxplot, because you
still have to make a call about whether this is "too big". Bear in
mind also that these facetted normal quantile plots, with two groups,
come out tall and skinny, so vertical deviations from the line are
exaggerated. On this plot, the lowest value also looks too low.

For the `Standard` group, there are no problems with normality
at all.

What happens if we change the shape of the plots?

```{r }
ggplot(mice, aes(sample=Time))+stat_qq()+stat_qq_line()+
facet_wrap(~Environment, scales="free", ncol=1)
```

     

This makes the plots come out in one column, that is, short and squat. I'd still call the highest one in `Enhanced` an outlier, but the lowest value now looks pretty close to what you'd expect.
 


