##  Movie ratings again


??q:movies-b?? This question again uses the movie rating data at
[link](http://www.utsc.utoronto.ca/~butler/c32/movie-lengths.csv).



(a) Read the data into R and obtain the number of movies of
each rating and the *median* length of movies of each rating.


Solution


Reading in is as in the other question using these data (just copy
your code, or mine). No credit for that, since you've done it before.
```{r }
my_url="http://www.utsc.utoronto.ca/~butler/c32/movie-lengths.csv"
movies=read_csv(my_url)
movies
```

 

Now, the actual for-credit part, which is a `group_by` and
`summarize`: 

```{r }
movies %>% group_by(rating) %>%
 summarize(count=n(),med=median(length))
```

 

The G movies have a smaller median than the others, but also the PG-13
movies seem to be longer on average (not what we found before). 
    


(b) Obtain a suitable graph that assesses the assumptions for
ANOVA. Why do you think it is not reasonable to run ANOVA here? Explain
briefly. 


Solution


The graph would seem to be a boxplot, side by side for each group:
```{r }
ggplot(movies,aes(x=rating, y=length))+geom_boxplot()
```

     

We are looking for approximate normal distributions with approximately
equal spreads, which I don't think we have: there are outliers, at the
low end for G movies, and at the high end for PG and PG-13
movies. Also, you might observe that the distribution of lengths for R
movies is skewed to the right. (Noting either the outliers or skewness
as a reason for not believing normality is enough, since all we need
is *one* way that normality fails.)

I think the spreads (as measured by the interquartile ranges) are
acceptably similar, but since we have rejected normality, it is a bit
late for that.

So I think it is far from reasonable to run an ANOVA here. In my
opinion 15 observations in each group is not enough to gain much from
the Central Limit Theorem either.

Extra: since part of the assumption for ANOVA is (approximate)
normality, it would also be entirely reasonable to make normal
quantile plots, one for each movie type, facetted. Remember the
process: you pretend that you are making a normal quantile plot for
all the data together, regardless of group, and then at the last
minute, you throw in a `facet_wrap`. I've written the code out
on three lines, so that you can see the pieces: the "what to plot",
then the normal quantile plot part, then the facetting:

```{r }
ggplot(movies,aes(sample=length))+
stat_qq()+stat_qq_line()+
facet_wrap(~rating)
```

 

Since there are four movie ratings, `facet_wrap` has arranged
them into a $2\times 2$ grid, which satisfyingly means that each
normal quantile plot is more or less square and thus easy to
interpret. 

The principal problem unveiled by these plots is outliers. It looks as
if the G movies have one low outlier, the PG movies have two high
outliers, the PG-13 movies have one or maybe three high outliers
(depending on how you count them), and the R movies have none. Another
way to look at the last two is you could call them curved, with too
much bunching up at the bottom and (on PG-13) too much spread-out-ness
at the top, indicating right-skewed distributions. The distribution
of lengths of the R-rated movies is too bunched up at the bottom, but
as you would expect for a normal at the top. The R movies show the
right-skewedness in an odd way: usually this skewness shows up by
having too many high values, but this time it's having too *few*
*low* values.

The assumption for ANOVA is that all four of these are at least
approximately normal (with the same spread). We found problems with
the normality on at least three of them, so we definitely have doubts
about trusting ANOVA here.

I could have used `scales=free` here to get a separate $y$-axis
for each plot, but since the $y$-axis is movie length each time, and
all four groups would be expected to have at least roughly similar
movie lengths, I left it as it was. (The other advantage of leaving
the scales the same is that you can compare spread by comparing the
slopes of the lines on these graphs; since the lines connect the
observed and theoretical quartiles, a steeper slope means a larger
IQR. Here, the R line is steepest and the PG line is flattest. Compare
this with the spreads of the boxplots.)

Extra extra: if you want, you can compare the normal quantile plots
with the boxplots to see whether you get the same conclusion from
both. For the G movies, the low outlier shows up both ways, and the
rest of the distribution is at least more or less normal. For the PG
movies, I'd say the distribution is basically normal except for the
highest two values (on both plots). For the PG-13 movies, only the
highest value shows up as an outlier, but the next two apparent
outliers on the normal quantile plot are at the  upper end of the long
upper whisker, so the boxplot is saying "right-skewed with one upper outlier" rather than "three upper outliers". The distribution of
the R movies is skewed right, with the bunching at the bottom showing
up as the very small lower whisker.

The boxplots and the normal quantile plots are basically telling the
same story in each case, but they are doing it in a slightly different
way. 
    


(c) Run a Mood's median test (use `smmr` if you
like). What do you conclude, in the context of the data?


Solution


The smart way is to use `smmr`, since it is much easier:
```{r }
library(smmr)
median_test(movies,length,rating)
```

     

The movies do not all have the same median length, or at least one of
the rating types has movies of different median length from the
others. Or something equivalent to that. It's the same conclusion as
for ANOVA, only with medians instead of means.

You can speculate about why the test came out significant. 
My guess is that the G movies are shorter than
average, and that the PG-13 movies are longer than average. (We had
the first conclusion before, but not the second. This is where medians
are different from means.)

The easiest way to see which movie types really differ in length from
which is to do all the pairwise median tests, which is in
`smmr` thus:

```{r }
pairwise_median_test(movies, length, rating)
```

 

The inputs for this are the same ones in the same order as for
`median_test`. (A design decision on my part, since otherwise
*I* would never have been able to remember how to run these!)
Only the first two of these are significant (look in the last
column). We can remind ourselves of the sample medians:

```{r }
movies %>% group_by(rating) %>%
summarize(count=n(),med=median(length))
```

 

The G movies are significantly shorter than the PG and PG-13 movies,
but not quite significantly different from the R movies. This is a
little odd, since the difference in sample medians between G and PG,
significant, is *less* than for G and R (not significant).
There are several Extras here, which you can skip if you don't care
about the background. First, we can do the median test by hand:
This has about four steps: (i) find the median of all the data, (ii) make a
table tabulating the number of values above and below the overall
median for each group, (iii) test the table for association, (iv)
draw a conclusion.
Thus (i):
```{r }
median(movies$length)
```

  

or

```{r }
movies %>% summarize(med=median(length))
```

 

or store it in a variable,
and then (ii):

```{r }
tab1=with(movies,table(length<100,rating))
tab1
```

 

or

```{r }
tab2=with(movies,table(length>100,rating))
tab2
```

 

These differ because there are evidently some movies of length exactly
100 minutes, and it matters whether you count $<$ and $\ge$ (as in
`tab1`) or $>$ and $le$ (`tab2`). Either is good. 

Was I right about movies of length exactly 100 minutes?

```{r }
movies %>% filter(length==100)
```

 

One PG and one R. It makes a difference to the R movies, but if you
look carefully, it makes a difference to the PG movies as well,
because the False and True switch roles between `tab1` and
`tab2` (compare the G movies, for instance).
You
need to store your table in a variable because it has to get passed on
to `chisq.test` below, (iii):

```{r }
chisq.test(tab1, correct=F)
```

 

or 

```{r }
chisq.test(tab2, correct=F)
```

 

Either is correct, or, actually, without the `correct=F`.\endnote{see
discussion elsewhere about Yates' Correction and fixed margins.}

The conclusion (iv) is the same either way: the null of no association
is clearly rejected (with a P-value of 0.0028 or 0.0036 as
appropriate), and therefore whether a movie is longer or shorter than
median length depends on what rating it has: that is, the median
lengths do differ among the ratings. The same conclusion, in other
words, as the $F$-test gave, though with not quite such a small
P-value. 

Second, you might be curious about how
we might do something like Tukey having found some significant
differences (that is, what's lurking in the background of
`pairwise_median_test`). 

Let's first suppose we are comparing G and PG movies. We need
to pull out just those, and then compare them using
`smmr`. Because the first input to `median_test` is a
data frame, it fits neatly into a pipe (with the data frame omitted):

```{r }
movies %>% filter(rating=="G" | rating=="PG") %>%
median_test(length,rating)
```

 

We're going to be doing this about six times --- ${4 \choose 2}=6$ choices
of two rating groups to compare out of the four --- so we should have a
function to do it. I think the input to the function should be a data
frame that has a column called `rating`, and two names of
ratings to compare:

```{r }
comp2=function(rat_1,rat_2,d) {
d %>% filter(rating==rat_1 | rating==rat_2) %>%
median_test(length,rating)
}
```

 

The way I wrote this function is that you have to specify the movie
ratings in quotes. It is *possible* to write it in such a way
that you input them without quotes, `tidyverse` style, but that
gets into "non-standard evaluation" and `enquo()` and
`!!`, which (i) I have to look up every time I want to do it,
and (ii) I am feeling that the effort involved in explaining it to you
is going to exceed the benefit you will gain from it. I mastered it enough
to make it work in `smmr` (note that you specify column names
without quotes there). There are tutorials on this kind of thing if
you're interested.

Anyway, testing:

```{r }
comp2("G","PG",movies)
```

 

That works, but I really only want to pick out the P-value, which is
in the list item `test` in the column `value`, the third
entry. So let's rewrite the function to return just that:

```{r }
comp2=function(rat_1,rat_2,d) {
d %>% filter(rating==rat_1 | rating==rat_2) %>%
median_test(length,rating) %>% pluck("test","value",3)
}
comp2("G","PG",movies)
```

 

Gosh.

What `median_test` returns is an R `list` that has two
things in it, one called `table` and one called
`test`. The thing  called `test` is a data frame with a
column called `value` that contains the P-values. The third of
these is the two-sided P-value that we want.

You might not have seen `pluck` before. This is a way of
getting things out of complicated data structures. This one takes the
output from `median_test` and from it grabs the piece called
`test`. This is a data frame. Next, we want the column called
`value`, and from that we want the third row. These are
specified one after the other to `pluck` and it pulls out the
right thing.

So now our function returns just the P-value.

I have to say that it took me several goes and some playing around in
R Studio to sort this one out. Once I thought I understood
`pluck`, I wondered why my function was not returning a
value. And then I realized that I was saving the value inside the
function and not returning it. Ooops. The nice thing about
`pluck` is that I can put it on the end of the pipeline and and
it will pull out (and return) whatever I want it to.

Let's grab a hold of the different rating groups we have:
```{r }
the_ratings=unique(movies$rating)
the_ratings
```

 

The Pythonisti among you will know how to finish this off: do a
loop-inside-a-loop over the rating groups, and get the P-value for
each pair. You can do that in R, if you must. It's not pretty at all,
but it works:

```{r }
ii=character(0)
jj=character(0)
pp=numeric(0)
for (i in the_ratings) {
for (j in the_ratings) {
pval=comp2(i,j,movies)
ii=c(ii,i)
jj=c(jj,j)
pp=c(pp,pval)
}
}  
tibble(ii,jj,pp)
```

 

This is a lot of fiddling about, since you have to initialize three
vectors, and then update them every time through the loop. It's hard
to read, because the actual business part of the loop is the
calculation of the P-value, and that's almost hidden by all the
book-keeping.  (It's also slow and inefficient, though the slowness
doesn't matter too much here since it's not a very big problem.)

Let's try another way:

```{r }
crossing(first=the_ratings,second=the_ratings)
```



This does "all possible combinations" of one rating with another. We
don't actually need all of that; we just need the ones where the first
one is (alphabetically) strictly less than the second one. This is
because we're never comparing a rating with itself, and each pair of
ratings appears twice, once in alphabetical order, and once the other
way around. The ones we need are these:

```{r }
crossing(first=the_ratings,second=the_ratings) %>%
filter(first<second)
```

 

A technique thing to note: instead of asking 
"how do I pick out the distinct pairs of ratings?", 
I use two simpler tools: first I make
all the combinations of pairs of ratings, and then out of those, pick
the ones that are alphabetically in ascending order, which we know how
to do.

Now we want to call our function `comp2` for each of the things
in `first` *and* each of the things in `second`,
and make a new column called `pval` that contains exactly
that. This (coming fresh from page 332 of the R book, this being the
first time I've ever used it\endnote{This was a year ago when I first
wrote this}) is exactly what the `map2` family of functions
does. In our case, `comp2` returns a decimal number, a
`dbl`, so `map2_dbl` does it.  Thus:
```{r }
crossing(first=the_ratings,second=the_ratings) %>%
filter(first<second) %>% 
mutate(pval=map2_dbl(first,second,~comp2(.x,.y,movies)))
```

 

The logic of `map2_dbl` is 
"for each of the things in `first`, and each of the things in `second`, taken in parallel, call the function `comp2` with those two inputs in that order, always with data frame `movies`". 
The `.x`
and `.y` play the role of the `.` that we usually have
inside a map, but now we're "mapping" over two things rather than
just one, so that they cannot both be called `.`.

One more thing: we're doing 6 tests at once here, so we're giving
ourselves 6 chances to reject a null (all medians equal) that might
have been true. So the true probability of a type I error is no longer
0.05 but something bigger. 

The easiest way around that is to do a so-called Bonferroni
adjustment: instead of rejecting if the P-value is less than 0.05, we
only reject if it is less than $0.05/6$, since we are doing 6
tests. This is a fiddly calculation to do by hand, but it's easy to
build in another `mutate`, thus:
`r tufte::margin_note("In the pairwise median  test in *smmr*, I did this backwards: rather than changing the alpha that you compare each P-value with from 0.05 to 0.05/6, I  flip it around so that you adjust the P-values by *multiplying*  them by 6, and then comparing the adjusted P-values with the usual  0.05. It comes to the same place in the end, except that this way  you can get adjusted P-values that are greater than 1, which makes no sense. You read those as being definitely not significant.")`
```{r }
crossing(first=the_ratings,second=the_ratings) %>%
filter(first<second) %>% 
mutate(pval=map2_dbl(first,second,~comp2(.x,.y,movies))) %>%
mutate(reject=pval<0.05/6)
```

 

And not a loop in sight.

This is how I coded it in `pairwise_median_test`. If you want to
check it, it's on Github:
[link](https://raw.githubusercontent.com/nxskok/smmr/master/R/pairwise_median_test.R). 
The function `median_test_pair` is the same as `comp2`
above. 

So the only significant differences are now G compared to PG and
PG-13. There is not a significant difference in median movie length
between G and R, though it is a close call. We thought the PG-13
movies might have a significantly different median from other rating
groups beyond G, but they turn out not to have. (The third and fourth
comparisons would have been significant had we not made the Bonferroni
adjustment to compensate for doing six tests at once; with that
adjustment, we only reject if the P-value is less than
$0.05/6=0.0083$, and so 0.0106 is not quite small enough to reject
with.) 

Listing the rating groups sorted by median would give you an idea of
how far different the medians have to be to be significantly different:

```{r }
medians=movies %>% group_by(rating) %>%
summarize(med=median(length)) %>%
arrange(desc(med))
medians
```

 

Something rather interesting has happened: even though the comparison of
G and PG (18 apart) is significant, the comparison of G and R (21
apart) is not significant. This seems very odd, but it happens because
the Mood median test is not actually literally comparing the sample
medians, but only assessing the splits of values above and below the
median of the combined sample. A subtlety, rather than an error, I'd say.

Here's something extremely flashy to finish with:

```{r }
crossing(first=the_ratings,second=the_ratings) %>%
filter(first<second) %>% 
mutate(pval=map2_dbl(first,second,~comp2(.x,.y,movies))) %>%
mutate(reject=pval<0.05/6) %>%
left_join(medians,by=c("first"="rating")) %>%
left_join(medians,by=c("second"="rating"))
```

 

The additional two lines look up the medians of the rating groups in
`first`, then `second`, so that you can see the actual
medians of the groups being compared each time. You see that medians
different by 30 are definitely different, ones differing by 15 or less
are definitely not different, and ones differing by about 20 could go
either way.

I think that's *quite* enough of that.
    



