

## Question title

 The data in file
[http://www.utsc.utoronto.ca/~butler/c32/ncbirths.csv](http://www.utsc.utoronto.ca/~butler/c32/ncbirths.csv) are about
500 randomly chosen births of babies in North Carolina. There is a lot
of information: not just the weight at birth of the baby, but whether
the baby was born prematurely, the ages of the parents, whether the
parents are married, how long (in weeks) the pregnancy lasted (this is
called the "gestation") and so on. We have seen these data before.



(a) Read in the data from the file into R, bearing in mind what
type of file it is. 


Solution


This is a `.csv` file (it came from a spreadsheet), so it
needs reading in accordingly. Work directly from the URL (rather
than downloading the file, unless you are working offline):
```{r }
myurl="http://www.utsc.utoronto.ca/~butler/c32/ncbirths.csv"
bw=read_csv(myurl)
``` 



(b) Find a 95% confidence interval for the mean birth weight of
all babies born in North Carolina (of which these babies are a
sample). At the end, you should state what the confidence interval is.
Giving some output is necessary, but *not* enough by itself.

If your variable name has a space or other special character (like a
question mark) in it, remember that you have to surround its name with
backticks, as discussed the first time we looked at these data.



Solution


This:

```{r }
t.test(bw$`Weight (pounds)`)  
``` 

or (the same, but remember to match your brackets):

```{r }
with(bw,t.test(`Weight (pounds)`))
``` 

The confidence interval goes from 6.94 to 7.20 pounds.

There is an annoyance about `t.test`. Sometimes you can use
`data=` with it, and sometimes not. When we do a two-sample
$t$-test later, there is a "model formula" with a squiggle in it,
and there we can use `data=`, but here not, so you have to use
the dollar sign or the `with` to say which data frame to get
things from. The distinction seems to be that emph{if you are using a
model formula}, you can use `data=`, and if not, not.

This is one of those things that is a consequence of R's history. The
original `t.test` was without the model formula and thus
without the `data=`, but the model formula got "retro-fitted"
to it later. Since the model formula comes from things like
regression, where `data=` is legit, that had to be retro-fitted
as well. Or, at least, that's my understanding.





(c) Birth weights of babies born in the United States have a mean
of 7.3 pounds. Is there any evidence that babies born in North
Carolina are less heavy on average? State appropriate hypotheses, do your
test, obtain a P-value and state your conclusion, in terms of the
original data.



Solution


Null is that the population mean (the mean weight of all babies born
in North Carolina) is $H_0: mu=7.3$ pounds, and the alternative  is
that the mean is less: $H_a: mu<7.3$ pounds. This is a one-sided
alternative, which we need to feed into `t.test`:

```{r }
t.test(bw$`Weight (pounds)`,mu=7.3,alternative="less")
``` 
$ %$ 

Or with `with`. If you see what I mean.

The P-value is 0.0003, which is *less* than any $alpha$ we might
have chosen: we *reject* the null hypothesis in favour of the
alternative, and thus we conclude that the mean birth weight of babies
in North Carolina
is indeed less than 7.3 pounds.

"Reject the null hypothesis" is *not* a complete answer. You
need to say something about what rejecting the null hypothesis means
*in this case*: that is, you must make a statement about birth
weights of babies.





(d) The theory behind the $t$-test says that the distribution of
birth weights should be (approximately) normally distributed. Obtain a
histogram of the birth weights. Does it look approximately normal?
Comment briefly. (You'll have to pick a number of bins for your
histogram first. I don't mind very much what you pick, as long as it's
not obviously too many or too few bins.)



Solution


We did this before (and discussed the number of bins before), so
I'll just reproduce my 10-bin histogram (which is what I preferred,
but this is a matter of taste):
```{r }
ggplot(bw,aes(x=`Weight (pounds)`))+geom_histogram(bins=10)
``` 

So, we were assessing normality. What about that?

It is mostly normal-looking, but I am suspicious about those
*very* low birth weights, the ones below about 4 pounds. There
are too many of those, as I see it.

If you think this is approximately normal, you need to make some
comment along the lines of ``the shape is approximately symmetric with
no outliers''. I think my first answer is better, but this answer is
worth something, since it is a not completely unreasonable
interpretation of the histogram.

A normal quantile plot is better for assessing normality
than a histogram is, but I won't make you do one until we have seen
the idea in class. Here's the normal quantile plot for these data:

```{r }
ggplot(bw,aes(sample=`Weight (pounds)`))+stat_qq()+stat_qq_line()
``` 

This is rather striking: the lowest birthweights (the ones below 5
pounds or so) are *way* too low for a normal distribution to
apply. The top end is fine (except perhaps for that one very heavy
baby), but there are too many low birthweights for a normal
distribution to be believable. Note how much clearer this story is
than on the histogram.

Having said that, the $t$-test, especially with a sample size as big
as this (500), behaves *very* well when the data are somewhat
non-normal (because it takes advantage of the Central Limit Theorem:
that is, it's the *sampling distribution of the sample mean*
whose shape matters). So, even though the data are definitely not
normal, I wouldn't be too worried about our test.

This perhaps gives some insight as to why Freedman-Diaconis said we
should use so many bins for our histogram. We have a lot of low-end
outliers, so that the IQR is actually *small* compared to the
overall spread of the data (as measured, say, by the SD or the range)
and so FD thinks we need a lot of bins to describe the shape. Sturges
is based on data being approximately normal, so it will tend to
produce a small number of bins for data that have outliers.





