##  North Carolina births


 The data in file
[http://www.utsc.utoronto.ca/~butler/c32/ncbirths.csv](http://www.utsc.utoronto.ca/~butler/c32/ncbirths.csv) are about
500 randomly chosen births of babies in North Carolina. There is a lot
of information: not just the weight at birth of the baby, but whether
the baby was born prematurely, the ages of the parents, whether the
parents are married, how long (in weeks) the pregnancy lasted (this is
called the "gestation") and so on. 



(a) Read in the data from the file into R, bearing in mind what
type of file it is. 


Solution


This is a `.csv` file (it came from a spreadsheet), so it
needs reading in accordingly. Work directly from the URL (rather
than downloading the file, unless you are working offline):
```{r }
myurl="http://www.utsc.utoronto.ca/~butler/c32/ncbirths.csv"
bw=read_csv(myurl)
``` 

This shows you which variables the data set has (some of the names got
a bit mangled), and it shows you that they are all integers except for
the birth weight (a decimal number).

The easiest way to find out how many rows and columns there are is
simply to list the data frame:

```{r }
bw
``` 

or you can take a "glimpse" of it:

```{r }
glimpse(bw)
``` 
Either of these displays show that there are 500 rows  (observations,
here births) and 10 columns (variables), and they both show what the
variables are called. So they're both good as an answer to the
question. 

What you'll notice is that the variables have *spaces* in their
names, which will require special handling later. These outputs show
you what to do about those spaces in variable names: surround the
variable name with "backticks". (On my keyboard, that's on the key
to the left of number 1, where the squiggle is, that looks like a
backwards apostrophe. Probably next to `Esc`, depending on the
layout of your keyboard.)

Although almost all of the variables are stored as integers, the ones
that have a question mark in their name are actually "logical", true
or false, with 1 denoting true and 0 false. We could convert them
later if we want to. (A question mark is not a traditional character
to put in a variable name, so we have to surround these variables with
backticks too.)



(b) From your output, verify that you have the
right number of observations and that you have several
variables. Which of your variables correspond to birthweight,
prematureness and length of pregnancy? (You might have to make guesses
based on the names of the variables.) 



Solution


I do indeed have 500 observations on 10 variables ("several"). (If
you don't have several variables, check to see that you didn't use
`read_delim` or something by mistake.) After the 
"500 observations of 10 variables" line(s) in each case, you see all the
variables by name, with what type of values they have\endnote{these
are mostly `int` or "integer".}, and the first few of the
values.\endnote{Other possible variable types are `num` for
(real, decimal) numbers such as birth weight, `chr` for
text, and `Factor` (with the number of levels) for
factors/categorical variables. We don't have any of the last two
here. There is also `lgl` for "logical", things that were
actually recorded as TRUE or FALSE. We have some variables that
are actually logical ones, but they are recorded as integer
values.}

The variable `Weight (pounds)` is the birthweight (in pounds),
`Premie?` is 1 for a premature baby and 0 for a full-term baby,
and `Weeks Gestation` is the number of weeks the pregnancy
lasted. Don't forget to put backticks around each of those when
you use them
later.\endnote{The backticks look different from each other for
annoying technical reasons, but they're all backticks.}
 


(c) The theory behind the $t$-test (which we do later) says that the
distribution of birth weights should be (approximately) normally
distributed. Obtain a histogram of the birth weights. Does it look
approximately normal?  Comment briefly. (You'll have to pick a number
of bins for your histogram first. I don't mind very much what you
pick, as long as it's not obviously too many or too few bins.)



Solution


You'll have seen that I often start with 10 bins, or maybe not quite
that many if I don't have much data, and this is a decent general
principle. That would give
```{r }
ggplot(bw,aes(x=`Weight (pounds)`))+geom_histogram(bins=10)
``` 

which is perfectly acceptable. You can try something a bit more or a
bit less, and see how you like it in comparison. What you are looking
for is a nice clear picture of *shape*. If you have too few bins,
you'll lose the shape:

```{r }
ggplot(bw,aes(x=`Weight (pounds)`))+geom_histogram(bins=4)
``` 

(is that leftmost bin an indication of skewness or some observations
that happen to be smallish?)

And if you have too many, the shape will be there, but
it will be hard to make out in all the noise, with frequencies going
up and down:

```{r }
ggplot(bw,aes(x=`Weight (pounds)`))+geom_histogram(bins=30)
``` 

I generally am fairly relaxed about the number of bins you use, as
long as it's not clearly too few or too many. You might have done
exercises in the past that illustrate that the choice of number of
bins (or the class intervals where you move from one bin to the next,
which is another issue that I won't explore here) can make an
appreciable difference to how a histogram looks.
Extra: I had some thoughts about this issue that I put in a blog
post, that you might like to read:
[http://ritsokiguess.site/docs/2017/06/08/histograms-and-bins/](http://ritsokiguess.site/docs/2017/06/08/histograms-and-bins/). The
nice thing about Sturges' rule, mentioned there, is that you can
almost get a number of bins for your histogram in your head (as long
as you know the powers of 2, that is). What you do is to start with
your sample size, here $n=500$. You find the next power of 2 above
that, which is here $512=2^9$. You then take that power and add 1,
to get 10 bins. If you don't like that, you can get R to calculate
it for you:

```{r }
nclass.Sturges(bw$`Weight (pounds)`)
``` 

The place where Sturges' rule comes from is an assumption of normal
data (actually a binomial approximation to the normal, backwards
though that sounds). If you have less than 30 observations, you'll get
fewer than 6 bins, which won't do much of a job of showing the
shape. Rob Hyndman wrote a
\href{https://robjhyndman.com/papers/sturges.pdf}{critical note} about
Sturges' rule in which he asserts that it is just plain wrong (if you
have taken B57, this note is very readable).

So what to use instead? Well, judgment is still better than something
automatic, but if you want a place to start from, something with a
better foundation than Sturges is the Freedman-Diaconis rule. This, in
its original formulation, gives a bin width rather than a number of bins:

$$ 
w=2(IQR)n^{-1/3}
$$

The nice thing about this is that it uses the interquartile range, so
it won't be distorted by outliers. `geom_histogram` can take a
bin width, so we can use it as follows:

```{r }
w=2*IQR(bw$`Weight (pounds)`)*500^(-1/3)
w
ggplot(bw,aes(x=`Weight (pounds)`))+geom_histogram(binwidth=w)
``` 

R also has

```{r }
nc=nclass.FD(bw$`Weight (pounds)`)
nc
``` 

which turns the Freedman-Diaconis rule into a number of bins rather
than a binwidth; using that gives the same histogram as we got with `binwidth`.

In my opinion, Freedman-Diaconis tends to give too many bins (here
there are 26 rather than the 10 of Sturges). But I put it out there
for you to make your own call.

Another way to go is a "density plot". This is a smoothed-out version of a histogram that is not obviously frequencies in bins, but which does have a theoretical basis. It goes something like this:

```{r }
ggplot(bw, aes(x=`Weight (pounds)`))+geom_density()
``` 

`geom_density` has an optional parameter that controls how smooth
or wiggly the picture is, but the default is usually good.

Alright, before we got distracted, we were assessing normality. What
about that?

It is mostly normal-looking, but I am suspicious about those
*very* low birth weights, the ones below about 4 pounds. There
are a few too many of those, as I see it.

If you think this is approximately normal, you need to make some
comment along the lines of ``the shape is approximately symmetric with
no outliers''. I think my first answer is better, but this answer is
worth something, since it is a not completely unreasonable
interpretation of the histogram.

I have been making the distinction between a histogram (for one
quantitative variable) and side-by-side boxplots (for one quantitative
variable divided into groups by one categorical variable). When you
learned the boxplot, you probably learned it in the context of one
quantitative variable. You can draw a boxplot for that, too, but the
`ggplot` boxplot has an `x` as well as a
`y`. What you do to make a single boxplot is to set the
`x` equal 1, which produces a weird $x$-axis (that you ignore):

```{r }
ggplot(bw,aes(x=1,y=`Weight (pounds)`))+geom_boxplot()
``` 

The high weight is actually an outlier, but look at all those outliers
at the bottom!\endnote{When Tukey, a name we will see again, invented
the boxplot in the 1950s, 500 observations would have been
considered a big data set. He designed the boxplot to produce a
sensible number of outliers for the typical size of data set of his
day, but a boxplot of a large data set tends to have a lot of
outliers that are probably not really outliers at all.}

*I* think the reason for those extra very low values is that they
are the premature births (that can result in *very* small
babies). Which leads to the additional question coming up.





