##  Inferring ice break-up in Nenana


 Nenana, Alaska, is about 50 miles west of Fairbanks.
Every spring, there is a contest in Nenana. A wooden tripod is
placed on the frozen river, and people try to guess the exact minute
when the ice melts enough for the tripod to fall through the ice. The
contest started in 1917 as an amusement for railway workers, and has
taken place every year since. Now, hundreds of thousands of people
enter their guesses on the Internet and the prize for the winner can
be as much as \$300,000. 

Because so much money is at stake, and because the exact same tripod
is placed at the exact same spot on the ice every year, the data are
consistent and accurate. The data are in
[link](http://ritsokiguess.site/datafiles/nenana.txt). 

Yes, we saw these data before.



(a) Read the data into R, as before, or use the data frame that
you read in before.  Note that the values are separated by
*tabs* rather than spaces, so you'll need an appropriate
`read_` to read it in.


Solution


These are "tab-separated values", so `read_tsv` is the
thing, as for the Australian athletes:
```{r nenana-inf-1 }
myurl <- "http://ritsokiguess.site/datafiles/nenana.txt"
nenana <- read_tsv(myurl)
```

       

Use whatever name you like for the data frame. One that is different
from any of the column headers is smart; then it is clear whether you
mean the whole data frame or one of its columns. `ice` or
`melt` or anything like that would also be good.
    
$\blacksquare$

(b) Obtain a 90\% confidence interval for the mean
`JulianDate`. What interval do you get? Looking back at your
histogram, do you have any doubts about the validity of what you
have just done?


Solution


This is a matter of using `t.test` and pulling out the
interval. Since we are looking for a non-standard interval, we
have to remember `conf.level` as the way to get the
confidence level that we want. I'm going with `with` this
time, though the dollar-sign thing is equally as good:
```{r nenana-inf-2 }
with(nenana, t.test(JulianDate, conf.level = 0.90))
```

       

Between 124.5 and 126.6 days into the year. Converting that into
something we can understand (because I want to), there are
$31+28+31+30=120$ days in 
January through April (in a non-leap year), so this says that the mean
breakup date is between about May 4 and May 6. 

The $t$-test is based on an assumption of data coming from a normal
distribution. The histogram we made earlier looks pretty much normal,
so there are no doubts about normality and thus no doubts about the
validity of what we have done, on the evidence we have seen so far. (I
have some doubts on different grounds, based on another of the plots
we did earlier, which I'll explain later, but all I'm expecting you to
do is to look at the histogram and say "Yep, that's normal enough". 
Bear in mind that the sample size is 87, which is large
enough for the Central Limit Theorem to be pretty helpful, so that we don't need the data to be more than "approximately normal" for the sampling distribution of the sample mean to be very close to $t$ with the right df.)
    
$\blacksquare$

(c) An old-timer in Nenana strokes his grey beard and says
"When I were young, I remember the tripod used to fall into the water around May 10". 
In a non-leap year, May 10 is Julian day 130. Test the null hypothesis that the 
mean `JulianDay` is 130, against the alternative that it is less.  What do you conclude?
What practical implication does that have 
(assuming that the old-timer has a good memory)?


Solution


The test is `t.test` again, but this time we have to
specify a null mean and a direction of alternative:
```{r nenana-inf-3 }
with(nenana, t.test(JulianDate, mu = 130, alternative = "less"))
```

       

For a test, look first at the P-value, which is 0.0000000002575: that
is to say, the P-value is very small, definitely smaller than 0.05 (or
any other $\alpha$ you might have chosen). So we *reject* the
null hypothesis, and conclude that the mean `JulianDate` is actually
*less* than 130.

Now, this is the date on which the ice breaks up on average, and we
have concluded that it is *earlier* than it used to be, since we
are assuming the old-timer's memory is correct. 

This is evidence in
favour of global warming; a small piece of evidence, to be sure, but
the ice is melting earlier than it used to all over the Arctic, so
it's not just in Nenana that it is happening. You don't need to get to
the "global warming" part, but I *do* want you to observe that
the ice is breaking up earlier than it used to.
    
$\blacksquare$

(d) Plot `JulianDate` against `Year` on a
scatterplot. What recent trends, if any, do you see? Comment
briefly. (You did this before, but I have some extra comments on
the graph this time, so feel free to just read this part.)


Solution


I liked  the `ggplot` with a smooth trend on it:

```{r nenana-inf-4 }
ggplot(nenana, aes(x = Year, y = JulianDate)) + geom_point() + geom_smooth()
```

 

There was something obvious to see: after about 1960, there is a clear
downward trend: the ice is breaking up earlier on average every
year. Even though there is a lot of variability, the overall trend,
viewed this way, is clear (and consistent with the test we did
earlier). Note that the old-timer's value of 130 is the kind of
`JulianDate` we would typically observe around 1920, which
would make the old-timer over 90 years old.

All right, why did I say I had some doubts earlier? Well, because of
this downward trend, the mean is not actually the same all the way
through, so it doesn't make all that much sense to estimate it, which
is what we were doing earlier by doing a confidence interval or a
hypothesis test. What would actually make more sense is to estimate
the mean `JulianDate` *for a particular year*. This could
be done by a regression: predict `JulianDate` from
`Year`, and then get a 
"confidence interval for the mean response" 
(as you would have seen in B27 or will see in C67). The
trend isn't really linear, but is not that far off. I can modify the
previous picture to give you an idea. Putting in `method="lm"`
fits a line; as we see later, `lm` does regressions in R:

```{r nenana-inf-5 }
ggplot(nenana, aes(x = Year, y = JulianDate)) + geom_point() +
  geom_smooth(method = "lm")
```

 

Compare the confidence interval for the mean `JulianDate` in
1920: 126 to 131 (the shaded area on the graph), with 2000: 121 to
125. A change of about 5 days over 80 years. And with the recent trend
that we saw above, it's probably changing faster than that
now. Sobering indeed.
      
$\blacksquare$




