##  Finding non-missing values


<a name="part:prepare-next">*</a> This is to prepare you for something in the next
question. It's meant to be easy.

In R, the code `NA` stands for "missing value" or
"value not known". In R, `NA` should not have quotes around
it. (It is a special code, not a piece of text.)


(a) Create a vector `v` that contains some numbers and some
missing values, using `c()`. Put those values into a
one-column data frame.
 
Solution


Like this. The arrangement of numbers and missing values doesn't
matter, as long as you have some of each:
```{r }
v=c(1,2,NA,4,5,6,9,NA,11)
mydata=tibble(v)
mydata
```

     

This has one column called `v`.
 

(b) Obtain a new column containing `is.na(v)`. When is this true and when is this false?
 
Solution


```{r }
mydata = mydata %>% mutate(isna=is.na(v))
mydata
```

     

This is `TRUE` if the corresponding element of `v` is
missing (in my case, the third value and the second-last one), and
`FALSE` otherwise (when there is an actual value there).
 

(c) The symbol `!` means "not" in R (and other
programming languages). What does `!is.na(v)` do? Create a
new column containing that.
 
Solution


Try it and see. Give it whatever name you like. My name reflects
that I know what it's going to do:
```{r }
mydata = mydata %>% mutate(notisna=!is.na(v))
mydata
```

     

This is the logical opposite of `is.na`: it's true if there is
a value, and false if it's missing.
 


(d) Use `filter` to display just the
rows of your data frame that have a non-missing value of `v`.

 
Solution


`filter` takes a column to say which rows to pick, in
which case the column should contain something that either *is*
`TRUE` or `FALSE`, or something that can be
interpreted that way:
```{r }
mydata %>% filter(notisna)
```

   

or you can provide `filter` something that can be calculated
from what's in the data frame, and also returns something that is
either true or false:

```{r }
mydata %>% filter(!is.na(v))
```

 

In either case, I only have non-missing values of `v`.
 



