##  Orange juice


 The quality of orange juice produced by a manufacturer
(identity unknown) is constantly being monitored. The manufacturer has
developed a "sweetness index" for its orange juice, for which a
higher value means sweeter juice. Is the sweetness index related to a
chemical measure such as the amount of water-soluble pectin (parts per
million) in the
orange juice? Data were obtained from 24 production runs, and the
sweetness and pectin content were measured for each run. The data are
in [link](http://www.utsc.utoronto.ca/~butler/c32/ojuice.txt). Open that
link up now. You can click on that link just above to open the file.



(a) The data values are separated by a space. Use the appropriate
Tidyverse function to read the data directly from the course website
into a "tibble".


Solution


Start with this (almost always):

```{r }
library(tidyverse)
```

 

The appropriate function, the data values being separated by a space,
will be `read_delim`. Put the URL as the first thing in
`read_delim`, or (better) define it into a variable
first:
`r tufte::margin_note("I say *better* because otherwise the read line gets rather long. This way you read it as *the URL is some long thing that I don't care about especially, and I what I need to do is to read the data from that URL, separated by spaces.*")`

```{r }
url <- "http://www.utsc.utoronto.ca/~butler/c32/ojuice.txt"
juice <- read_delim(url, " ")
```

 

`read_delim` (or `read_csv` or any of the others) tell
you what variables were read in, and also tell you about any "parsing errors" 
where it couldn't work out what was what. Here, we have three
variables, which is entirely consistent with the three columns of data
values in the file.

`read_delim` can handle data values separated by *any*
character, not just spaces, but the separating character, known as a
"delimiter", does *not* have a default, so you have to say what
it is, every time.




(b) Take a look at what got read in. Do you
have data for all 24 runs? 


Solution


Type the name of the data frame in a code chunk (a new one, or
add it to the end of the previous one). Because this is actually
a "tibble", which is what `read_delim` reads in,
you'll only actually see the first 10 lines, but it will tell
you how many lines there are altogether, and you can click on
the appropriate thing to see the rest of it.
```{r }
juice
```

 

I appear to have all the data. If you want further convincing, click
Next a couple of times (on yours) to be sure that the runs go down to
number 24.



(c) In your data frame, where did the column (variable) names
come from? How did R know where to get them from?


Solution


They came from the top line of the data file, so we didn't 
have to specify them. This is the default behaviour of all the
`read_` functions, so we don't have to ask for it
specially. 
In fact, if the top line of your data file is
*not* variable names, *that's* when you have to say
something special. The `read_` functions have an
option `col_names` which can either be `TRUE`
(the default), which means "read them in from the top line",
`FALSE` ("they are not there, so make some up") or a
list of column names to use. You might use the last
alternative when the column names that are in the file are
*not* the ones you want to use; in that case, you would
also say `skip=1` to skip the first line. For example,
with file `a.txt` thus:
`timinput{`.txt}        
you could read the same data but call the columns `x` and
`y` thus:

```{r }
read_delim("a.txt", " ", col_names = c("x", "y"), skip = 1)
```

         



(d) The juice manufacturer was interested in whether there was a
relationship between sweetness and pectin. To assess this, draw a
scatterplot.  Does it look as if there is any kind of a
relationship? (I think `sweetness` is the outcome variable
and `pectin` is explanatory, so draw your scatterplot
appropriately.) 


Solution


This requires a `ggplot` plot. You can go back and look
at the lecture notes to figure out how to make a scatterplot:
the "what to plot" is the $x$-axis and $y$-axis variables,
with the response on the $y$-axis (starting with a data frame
to get the variables from),  and the "how to plot" is
`geom_point` to plot the points:
```{r }
ggplot(juice, aes(x = pectin, y = sweetness)) + geom_point()
```

         

It looks to me as if there is a negative relationship: as pectin goes
up, sweetness tends to go *down*. The trend appears to go top
left to bottom right. 

Having said that, I'm wondering how much of the
apparent trend is caused by those two observations bottom right with
pectin over 350. If you take those away, the trend seems to me to be a
lot less convincing. As an extra, you could add a smooth trend to the plot:

```{r }
ggplot(juice, aes(x = pectin, y = sweetness)) + geom_point() + geom_smooth()
```

         
The smooth trend is kind of downhill, but not very convincing.




