##  Orange juice


We will read in the orange juice data again,
in [link](http://ritsokiguess.site/datafiles/ojuice.txt), and make a suitable graph.


(a) The data values are separated by a space. Use the appropriate
Tidyverse function to read the data directly from the course website
into a "tibble".


Solution

As before,

```{r oj2-1 }
url <- "http://ritsokiguess.site/datafiles/ojuice.txt"
juice <- read_delim(url, " ")
juice
```

24 runs, as we had before.

$\blacksquare$

(b) The juice manufacturer was interested in whether there was a
relationship between sweetness and pectin. To assess this, draw a
scatterplot.  Does it look as if there is any kind of a
relationship? (I think `sweetness` is the outcome variable
and `pectin` is explanatory, so draw your scatterplot
appropriately.) 


Solution


This requires a `ggplot` plot. You can go back and look
at the lecture notes to figure out how to make a scatterplot:
the "what to plot" is the $x$-axis and $y$-axis variables,
with the response on the $y$-axis (starting with a data frame
to get the variables from),  and the "how to plot" is
`geom_point` to plot the points:
```{r oj2-2 }
ggplot(juice, aes(x = pectin, y = sweetness)) + geom_point()
```

         

It looks to me as if there is a negative relationship: as pectin goes
up, sweetness tends to go *down*. The trend appears to go top
left to bottom right. 

Extra: having said that, I'm wondering how much of the
apparent trend is caused by those two observations bottom right with
pectin over 350. If you take those away, the trend seems to me to be a
lot less convincing. You could add a smooth trend to the plot, thus:

```{r oj2-3 }
ggplot(juice, aes(x = pectin, y = sweetness)) + geom_point() + geom_smooth()
```

         
The smooth trend is kind of downhill, but not very convincing.

$\blacksquare$




