##  Ovarian cancer


 R's `survival` package contains several data
sets. One of these is called `ovarian`; it comes from a study
of 26 ovarian cancer patients. The major purpose of this study was to
compare the effects of two treatments on survival time.



(a) Obtain and display (all of) the data set. This is as simple as
loading the package and typing the data set's name.

Solution


Thus. You may need to start with `library(survival)`:

```{r }
ovarian
```

 

There are indeed 26 rows. This is a `data.frame` rather than a
`tibble`, so you see the whole thing, in case you were
expecting something like this:

```{r }
ovarian %>% as.tibble()
```

 

which doesn't change anything in `ovarian`, but changes what
kind of thing it is (and thus how it displays). Usually when you read
something in from a file, you use something like `read_delim`
that makes a `tibble`, but this one wasn't read in from a
file. It was stored in the package as an old-fashioned
`data.frame`, and so that's how it stays.


(b) The columns of interest to us are:


* `futime`: the time for which a patient was followed-up:
the number of days until either they died or the study ended (or
they withdrew from the study for some other reason).

* `fustat`: follow-up status: 1 if the patient died of
ovarian cancer, 0 if they were still alive when the study ended.

* `age`: of patient, at diagnosis, in years

* `rx`: treatment, numbered 1 or 2, but really labels for
the two treatments.

Create and display a suitable response variable `y` for a Cox
proportional-hazards model.

Solution


The idea is to use the appropriate one(s) of these columns in
`Surv`. Remember that the response variable in a survival
model encodes two things: the survival time, and whether or not
the event (here death) actually happened to that patient or not.
I always forget whether the second thing in `Surv` has to
be 1 or 0 if the event happened. The help says that it needs to be
1 or `TRUE` if the event (death) happened, which is what
`fustat` is, so we can use it as it is:
```{r }
y=with(ovarian,Surv(futime,fustat))
y
```

     

This creates a separate variable `y` outside of any data
frame. This is how it has to be done, because although `y`
looks as if it is 26 long (one per patient), it's actually more
complicated than that:

```{r error=T}
ov2 = ovarian %>% mutate(y=Surv(futime,fustat))
```

 

This doesn't work, because internally `y` is a `matrix` with two columns:

```{r }
print.default(y)
```

 

The `attr(,"class")` thing at the bottom of this says that `y` is actually  a `Surv` object, which is how it knows to display as it does, as 26 values some of which have plusses.
`r tufte::margin_note("Internally, printing a *Surv* thing calls the method *print.Surv*, which is defined in the *survival* package.")`

The way R works out the length of something is its total number of
entries, which for this matrix is $26 \times 2=52$. The only things
you can add via `mutate` to a data frame with 26 rows have
length 26 (since everything in a data frame has to be the same length)
or length 1 (in which case it is repeated 26 times).

With some cajoling, you can get this into a data frame, but with more
cajoling than seems necessary. It's much easier to leave it as a
separate thing outside a data frame.
`r tufte::margin_note("In some respects, it's  like that two-column response in logistic regression, where each row of the data frame represents several observations pooled together.")`

I gave `ov2` a "temporary" name because I am not going to use
it again.


(c) In the display of your response variable, some values are
marked with a `+`. Why is that? Explain briefly. (If you use
a technical term, you should explain what it means.)

Solution



These are the censored observations. You can say this, but you
also need to say what that means (this is the "technical term"
referred to in the question). The observations with a `+`
are individuals who were never observed to die, or who were still
alive at the end of the study.

I want you to demonstrate that you know what censored
*means*, not just that you know when you have a censored
observation.

Extra: in a study like this, patients are typically "recruited"
into the study at various different times. Patients who happened
to be in the study near the beginning and who survived can have a
large (censored) value of `y` (like those values over 1000
days). But a patient might join the study later on; if they
survive, they might produce a censored observation with a small
survival time, like the last value 377. I'm sure the doctor would
have liked to follow them for longer, but the funding ran out, and
the doctor had a paper to write. (There is *some* information
in these small censored values, but not much, because most of the
patients, even the ones who eventually died, survived for longer
than 377 days.)

The other thing that might have happened is that a patient with
the 377-censored value died *from something else* unrelated
to ovarian cancer. The study is only concerned with deaths from
ovarian cancer, so such a patient is treated as censored at their
death time. After this point we cannot assess how long
this patient survived *ovarian cancer*.


(d) Fit a Cox proportional-hazards model for predicting
survival time from age and treatment. Note that the numeric values
for treatment make sense only as labels for the two treatments, so
in your model formula make treatment into a factor. Display the
results. 

Solution


The hint suggests something like this:
```{r }
time.1=coxph(y~age+factor(rx),data=ovarian)
summary(time.1)
```

     

Alternatively, define the factor version of `rx` in the data
frame first. This is the slick way to do that:

```{r }
time.1a = ovarian %>% mutate(rxf=factor(rx)) %>%
coxph(y~age+rxf,data=.)
summary(time.1a)
```

 

The answer is the same either way. 
The `data=.` means "use the data frame that came out of the previous step, the one with `rxf` in it."


(e) Is there a significant difference between the treatments in
terms of their effects on survival (from ovarian cancer)?

Solution



Look at the P-value for my `factor(rx)2`, 0.203. This is
not small, so there is no evidence of a difference between
treatments. 

Extra: the reason for the odd label is that we have turned
treatment into a categorical variable; treatment 1 is used as the
baseline, and the negative slope says that the "hazard of death"
is lower for treatment 2 than for treatment 1: that is, people
survive longer on treatment 2, but the difference is not big
enough to be significant (we also have a smallish sample size).

Since there are only two treatments, it would in fact have been OK
to leave them as numbers (with two numbers one unit apart, the
slope would have been the same size as here), but I think it's a
good idea to treat categorical variables as categorical. My own habit
is to use letters or something non-numerical to distinguish
categories. I might have used `t1` and `t2` in this
case, or the names of the different treatments.



(f) Is there a significant effect of age? If there is, describe
the effect that age has on survival.

Solution



The P-value for age is 0.0014, small, so age definitely has a
significant effect on survival. As to what kind of effect, look at
the slope coefficient, 0.15, positive, which means that increased
age-at-diagnosis goes with an *increased* hazard of death, or
that older patients do not survive as long.

I would like you to get to the plain-English words at the
end. Part of your job as a statistician is explaining what you got
to people who are doctors, managers, etc., who won't
understand the terminology. 

Thus, one mark for assessing significance
via P-value, one for looking at the slope coefficient and noting
that it is positive, and one for getting to "older patients do     not survive as long", or 
"older patients have a larger chance of dying sooner". 
(Strictly, this is also "all else equal" as usual,
since survival time might also have depended on treatment, but the
point of this question is for you to get to "older patients do not survive as long".) 

(The interpretation of the slope may seem backwards: a positive
slope means a *shorter* survival time for a larger age. This
is why I talk about "hazard of death", since that guides us to
the correct interpretation.)

Extra: I was curious about what would happen if I just included
`rx` in the model:

```{r }
time.2=update(time.1,.~.-age)
summary(time.2)
```

     

Still not significant, but this model is a *lot worse* because I
took out the significant `age`. What this is doing is mixing up
all the people of different ages (and we know that age has an effect
on survival) and trying (and failing) to discern an effect of
treatment. 

We could have been badly misled by this model if one of the treatments
had predominantly older patients. We know that older patients have
worse survival, so the treatment with older patients would have looked
worse, even if it actually wasn't. The model `time.1` which
contained `age` properly adjusted for the effect of age, so
that was the best way to see whether there was a difference between
treatments. 

What you often see early on in a paper on this kind of stuff is a
graph showing that the treatment groups are similar in terms of
important things like `age`. Here, that could be a boxplot:

```{r }
ggplot(ovarian, aes(x=factor(rx),y=age))+geom_boxplot()
```

 

I needed to do `factor(rx)` because `geom_boxplot`
needs a genuine categorical variable, not just a numerical variable
masquerading as one. If you just leave it as `rx`, as I
discovered, you get *one* boxplot of all the ages together
regardless of treatment. The key, for you as user of software, is not
(necessarily) to get it right the first time, but to know what to do
to fix up the errors you will inevitably get. If you have worked
through the boxplot examples in C32 and D29, you will have enough
experience to remember that a boxplot has to have a categorical `x`
(text will do, but definitely not numbers). This is why I give you so
many things to work through: so that you gain the experience to know
how to fix up problems.

Treatment 1 has a larger spread of ages and treatment 2 has a low
outlier age, but the median ages are very similar.


(g) Make a martingale residual plot for this model. Do you see
any problems? Explain briefly.

Solution


The plot is just the same idea as the one in the notes. Make sure
you have `survminer` installed and loaded:
```{r }
ggcoxdiagnostics(time.1)+geom_smooth()
```

     

Make a call about whether you think the smooth trend deviates too much
from the red dotted line going across at zero. Martingale residuals
can get very negative (and that is OK), so that residual of $-2$ is
not a problem, and this is pulling the smooth trend down a bit (it's
the reason for the dip on the right side of the smooth trend). So I'd
call this OK, but you can make whatever call you like as long as it's
supported by what you see here.

I observe that the obvious fixable thing is where you have a curve
here, one that looks like a parabola (at which point you add squared
terms to your explanatory variables and see if that helps, as for
`bmi` in one of the other problems). This one
is too wiggly to be a parabola (it bends *twice*), and so is more
like a cubic than anything.

The other thing you can note is that the grey envelope is 
"not significantly different from zero", since 0 is clearly within the grey
envelope all the way across.




(h) Find the quartiles of `age`, and make a data frame
containing all combinations of those two ages and the two
treatments. Display what you have. (Feel free to copy the values by
hand, rather than trying to save them and use them.)


Solution


I imagine you can guess what we are going to be doing with these:
predictions, so we'll call the data frame `new` when we get
there. 
Quartiles first:
```{r }
quantile(ovarian$age)
```

   

or, if you prefer,

```{r }
ovarian %>% 
summarize(q1=quantile(age,0.25),
q3=quantile(age,0.75))
```

 

The quartiles are 50.17 and 62.38 (rounding slightly).

Either way is good. 

Then follow my standard procedure (or one of your own devising),
remembering that "treatment" is called `rx` here:

```{r }
ages=c(50.17, 62.38)
rxs=c(1, 2)
new=crossing(age=ages,rx=rxs)
new
```

 



(i) Obtain predicted survival probabilities for each of your
age-treatment combinations, for each of a variety of survival
times. (This is only one thing, despite it sounding like a lot.)


Solution


The magic word here is `survfit`. The structure is the same
as for `predict`:
```{r }
s=survfit(time.1,new,data=ovarian)
summary(s)
```

   

I didn't ask you to display it, so doing so is optional. Also, you
don't need (here) that `data=ovarian` (the predictions will
work just fine without it), but the plot coming up *will not*. So
my recommendation is to put it in.

Extra: The four columns `survival1` through `survival4` are the
survival probabilities for the times shown in the left-hand column
(these are numbers of days), for the four *rows* of my
`new`, in the same order.

There are only a few different times, because these are the numbers of
days at which somebody in the data set died, and the estimated
survival probability does not change at the times in between
these. (You'll see this on a plot in a minute.)

These survival probabilities are pretty easy to eyeball: the best
survival is in stratum 2, which is the younger patients in treatment
2. This we'll come back to.



(j) Draw a plot that compares the survival probabilities at the
different times.


Solution


Thus. The `conf.int=F` means to skip the confidence interval
"envelopes" that I find make the plot rather messy:
```{r }
ggsurvplot(s,conf.int=F)
```

   


(k) According to your plot, how would you describe the effects of
treatment and of age?


Solution


The best survival curve, in terms of surviving longest, is up and to
the right, so the green one is best and the blue one is
worst.
`r tufte::margin_note("In other cases, having the event happen sooner is    better. For example, you might be poisoning rats, in which case    you want them to die quicker. Or the event might be something    desirable like becoming qualified to fly an airplane. In those    cases, down and to the left is better.")` To figure out which those
are, we have to go back to the data frame we created:
```{r }
new
```

  

To see the effect of age, compare strata 1 and 3 (or 2 and 4). This
means comparing the red and blue curves; the red one is clearly better
(in the sense of longer survival time), which means that age has a big
effect on survival, with younger people living longer, other things
being equal (that we saw earlier was significant). You could
equally well compare the green and purple survival curves and come to
the same conclusion.

To assess the effect of treatment, compare strata 1 and 2 (red and
green), or strata 3 and 4 (blue and purple). In both cases, the
stratum corresponding to treatment 2 has slightly better survival (has
a higher chance of living for longer), but there is not as big an
effect as for age. (You'll recall that the treatment difference was
not significant).

One point for stating or implying that you know which stratum is
which, and one each for saying something about the effects of age and
of treatment, including which one is better.

Extra: recall the output from the Cox model:

```{r }
summary(time.1)
```

 

The slope coefficient for treatment 2 (as compared to the baseline
treatment 1) was $-0.83097$, *negative*, which meant that patients on
treatment 2 had a *lower* hazard of death than patients on
treatment 1: that is, that treatment 2 was better for survival than
treatment 1. That is what the plot said also (and the relatively small
difference is consistent with that difference not being significant). 



