##  What sports do these athletes play?


 The data at
[link](http://ritsokiguess.site/datafiles/ais.txt) are physical
and physiological measurements of 202 male and female Australian elite
athletes. The data values are separated by *tabs*. We are going
to see whether we can predict the sport an athlete plays from their
height and weight.

The sports, if you care, are respectively basketball, 
"field athletics" (eg. shot put, javelin throw, long jump etc.),
gymnastics, netball, rowing, swimming, 400m running, tennis, sprinting
(100m or 200m running), water polo.



(a) Read in the data and display the first few rows.

Solution


The data values are separated by tabs, so `read_tsv` is
the thing:
```{r oz-multi-1 }
my_url <- "http://ritsokiguess.site/datafiles/ais.txt"
athletes <- read_tsv(my_url)
athletes
```

     

If you didn't remember that, this also works:

```{r oz-multi-2 }
athletes <- read_delim(my_url, "\t")
```

 

(this is the R way of expressing "tab".)

$\blacksquare$

(b) Make a scatterplot of height vs.\ weight, with the points
coloured by what sport the athlete plays. Put height on the $x$-axis
and weight on the $y$-axis.

Solution


I'm doing this to give you a little intuition for later:
```{r oz-multi-3 }
ggplot(athletes, aes(x = Ht, y = Wt, colour = Sport)) + geom_point()
```

     

The reason for giving you the axes to use is (i) neither variable is
really a response, so it doesn't actually matter which one goes on
which axis, and (ii) I wanted to give the grader something consistent
to look at.


$\blacksquare$


(c) Explain briefly why a multinomial model (`multinom`
from `nnet`) would be the best thing to use to predict sport
played from the other variables.

Solution


The categories of `Sport` are not in any kind of order, and
there are more than two of them.
That's really all you needed, for which two marks was kind of
generous. 

$\blacksquare$

(d) Fit a suitable model for predicting sport played from
height and weight. (You don't need to look at the results.) 100
steps isn't quite enough, so set `maxit` equal to a larger
number to allow the estimation to finish.

Solution


120 steps is actually enough, but any number larger than 110 is
fine. It doesn't matter if your guess is way too high. Like this:

```{r oz-multi-4 }
library(nnet)
sport.1 <- multinom(Sport ~ Ht + Wt, data = athletes, maxit = 200)
```

 

As long as you see the word `converged` at the end, you're
good. 

$\blacksquare$

(e) Demonstrate using `anova` that `Wt` should
not be removed from this model.

Solution


The idea is to fit a model without `Wt`, and then show that
it fits significantly worse. This converges in less than 100 iterations, so you can have `maxit` or not as you prefer:

```{r oz-multi-5 }
sport.2 <- update(sport.1, . ~ . - Wt)
anova(sport.2, sport.1, test = "Chisq")
```

     

The P-value is very small indeed, so the bigger model `sport.1`
is definitely better (or the smaller model `sport.2` is
significantly worse, however you want to say it). So taking
`Wt` out is definitely a mistake. 

This is what I would have guessed (I actually wrote the question in
anticipation of this being the answer) because weight certainly seems
to help in distinguishing the sports. For example, the field athletes
seem to be heavy for their height compared to the other athletes (look
back at the graph you made). 

`drop1`, the obvious thing, doesn't work here:

```{r oz-multi-6, error=T}
drop1(sport.1, test = "Chisq", trace = T)
```

 

I gotta figure out what that error is.
Does `step`?

```{r oz-multi-7 }
step(sport.1, direction = "backward", test = "Chisq")
```

 

Curiously enough, it does. The final model is the same as the initial
one, telling us that neither variable should be removed.

$\blacksquare$

(f) Make a data frame consisting of all combinations of
`Ht` 160, 180 and 200 (cm), and `Wt` 50, 75, and 100
(kg), and use it to obtain predicted probabilities of athletes of
those heights and weights playing each of the sports. Display the
results. You might have to display them smaller, or reduce the
number of decimal places^[For this, use `round`.] 
to fit them on the page.

Solution

To get all combinations of those heights and weights, use `datagrid`:

```{r}
new <- datagrid(model = sport.1, Ht = c(160, 180, 200), Wt = c(50, 75, 100))
new
```

(check: $3 \times 3 = 9$ rows.)

Then predict:

```{r}
p <- predictions(sport.1, newdata = new, type = "probs")
p
```

I saved these, because I want to talk about what to do next. The normal procedure is to say that there is a prediction for each `group` (sport), so we want to grab only the columns we need and pivot wider. But, this time there are 10 sports (see how there are $9 \times 10 = 90$ rows, with a predicted probability for each height-weight combo *for each sport*). I suggested reducing the number of decimal places in the predictions; the time to do that is *now*, while they are all in one column (rather than trying to round a bunch of columns all at once, which is doable but more work). Hence:

```{r}
p %>% 
  mutate(predicted = round(predicted, 2)) %>% 
  select(group, predicted, Ht, Wt) %>% 
  pivot_wider(names_from = group, values_from = predicted)
```

This works (although even then it might not all fit onto the page and you'll have to scroll left and right to see everything).

If you forget to round until the end, you'll have to do something like this:

```{r}
p %>% 
  select(group, predicted, Ht, Wt) %>% 
  pivot_wider(names_from = group, values_from = predicted) %>% 
  mutate(across(BBall:WPolo, \(x) round(x, 2)))
```

In words, "for each column from `BBall` through `WPolo`, replace it by itself rounded to 2 decimals".

There's nothing magic about two decimals; three or maybe even four would be fine. A small enough number that you can see most or all of the columns at once, and easily compare the numbers for size (which is hard when some of them are in scientific notation).

Extra: you can even abbreviate the sport names, like this:

```{r}
p %>% 
  mutate(predicted = round(predicted, 2),
         group = abbreviate(group, 3)) %>% 
  select(group, predicted, Ht, Wt) %>% 
  pivot_wider(names_from = group, values_from = predicted) 
```

Once again, the time to do this is while everything is long, so that you only have one column to work with.^[`abbreviate` comes from base R; it returns an abbreviated version of its (text) input, with a minimum length as its second input. If two abbreviations come out the same, it makes them longer until they are different. It has some heuristics to get things short enough: remove vowels, then remove letters on the end.] Again, you can do this at the end, but it's more work, because you are now renaming columns:

```{r}
p %>% 
  mutate(predicted = round(predicted, 2)) %>% 
  select(group, predicted, Ht, Wt) %>% 
  pivot_wider(names_from = group, values_from = predicted) %>% 
  rename_with(\(x) abbreviate(x, 3), everything())
```

`rename` doesn't work with `across` (you will get an error if you try it), so you need to use `rename_with`^[Which I learned about today.]. This has syntax "how to rename" first (here, with an abbreviated version of itself), and "what to rename" second (here, everything, or `BBall` through `WPolo` if you prefer). 

    
$\blacksquare$

(g) For an athlete who is 180 cm tall and weighs 100 kg, what
 sport would you guess they play? How sure are you that you are
 right? Explain briefly.

Solution


Find this height and weight in your predictions (it's row 6). Look
along the line for the highest probability, which is 0.85 for
`Field` (that is, field athletics). All the other
probabilities are much smaller (the biggest of the others is
0.06). So this means we would guess the athlete to be a field
athlete, and because the predicted probability is so big, we are
very likely to be right.
This kind of thought process is characteristic of discriminant
analysis, which we'll see more of later in the course.
Compare that with the scatterplot you drew earlier: the field
athletes do seem to be distinct from the rest in terms of
height-weight combination. 
Some of the other height-weight combinations are almost equally
obvious: for example, very tall people who are not very heavy are
likely to play basketball. 400m runners are likely to be of
moderate height but light weight. Some of the other sports, or
height-weight combinations, are difficult to judge. Consider also
that we have mixed up males and females in this data set. We might
gain some clarity by including `Sex` in the model, and also
in the predictions. But I wanted to keep this question relatively
simple for you, and I wanted to stop from getting unreasonably
long. (You can decide whether you think it's already too long.)
 
$\blacksquare$



