##  What makes an animal get infected?


 Some animals got infected with a parasite. We are interested
in whether the likelihood of infection depends on any of the age,
weight and sex of the animals. The data are at
[link](http://ritsokiguess.site/datafiles/infection.txt). The
values are separated by tabs.



(a) Read in the data and take a look at the first few lines. Is
this one animal per line, or are several animals with the same age,
weight and sex (and infection status) combined onto one line? How
can you tell?
 
Solution


The usual beginnings, bearing in mind the data layout:
```{r parasite-1 }
my_url <- "http://ritsokiguess.site/datafiles/infection.txt"
infect <- read_tsv(my_url)
infect
```

 

Success.
This appears to be one animal per line, since there is no indication
of frequency (of "how many"). If you were working as a consultant
with somebody's data, this would be a good thing to confirm with them
before you went any further.

You can check a few more lines to
convince yourself
and the story is much the same. The other hint is that you have actual
categories of response, which usually indicates one individual per
row, but not always. If it doesn't, you have some extra work to do to
bash it into the right format.

Extra: let's see whether we can come up with an example of that. I'll make a
smaller example, and perhaps the place to start is 
"all possible combinations" of a few things. 
If you haven't seen `crossing`
before, skip ahead to part (<a href="#part:crossing">here</a>):

```{r parasite-2 }
d <- crossing(age = c(10, 12), gender = c("f", "m"), infected = c("y", "n"))
d
```



These might be one individual per row, or they might be more than one,
as they would be if we have a column of frequencies:

```{r parasite-3 }
d <- d %>% mutate(freq = c(12, 19, 17, 11, 18, 26, 16, 8))
d
```

 

Now, these are multiple observations per row (the presence of
frequencies means there's no doubt about that), but the format is
wrong: `infected` is my response variable, and we want the
frequencies of `infected` being `y` or `n` in
*separate* columns --- that is, we have to *untidy* the data
a bit to make it suitable for modelling. This is `pivot_wider`, the
opposite of `pivot_longer`:

```{r parasite-4 }
d %>% pivot_wider(names_from=infected, values_from=freq)
```

 

Now you can pull out the columns `y` and `n` and make
them into your response, and predict that from `age` and
`gender`. 

The moral of this story is that if you are going to have multiple
observations per row, you probably want the combinations of
*explanatory* variables one per row, but you want the categories
of the *response* variable in separate columns.

Back to where we were the rest of the way.
 
$\blacksquare$

(b) <a name="part:plot">*</a> Make suitable plots or summaries of `infected` against
each of the other variables. (You'll have to think about
`sex`, um, you'll have to think about the `sex`
variable, because it too is categorical.) Anything sensible is OK
here. You might like to think back to what we did in
Question <a href="#q:wolfspider">here</a> for inspiration. (You can also
investigate `table`, which does cross-tabulations.)
 
Solution


What comes to my mind for the numerical variables `age` and
`weight` is boxplots:
```{r fodudu}
ggplot(infect, aes(x = infected, y = age)) + geom_boxplot()
ggplot(infect, aes(x = infected, y = weight)) + geom_boxplot()
```

     

The variables `sex` and `infected` are both
categorical. I guess a good plot for those would be some kind of
grouped bar plot, which I have to think about. 
So let's first try a numerical summary, a
cross-tabulation, which is gotten via `table`:

```{r parasite-5 }
with(infect, table(sex, infected))
```

 

Or, if you like the `tidyverse`:

```{r parasite-6 }
infect %>% count(sex, infected)
```

 

Now, bar plots. Let's start with one variable. The basic bar plot has
categories of a categorical variable along the $x$-axis and each bar
is a count of how many observations were in that category. What is
nice about `geom_bar` is that it will do the counting for you,
so that the plot is just this:

```{r parasite-7 }
ggplot(infect, aes(x = sex)) + geom_bar()
```

 

There are about twice as many males as females.

You may think that this looks like a histogram, which it almost does,
but there is an important difference: the kind of variable on the
$x$-axis. Here, it is a categorical variable, and you count how many
observations fall in each category (at least, `ggplot`
does). On a histogram, the $x$-axis variable is a continuous
*numerical* one, like height or weight, and you have to
*chop it up* into intervals (and then you count how many
observations are in each chopped-up interval).

Technically, on a bar plot, the bars have a little gap between them
(as here), whereas the histogram bars are right next to each other,
because the right side of one histogram bar is the left side of the next.

All right, two categorical variables. The idea is that you have each
bar divided into sub-bars based on the frequencies of a second
variable, which is specified by `fill`. Here's the basic idea:

```{r parasite-8 }
ggplot(infect, aes(x = sex, fill = infected)) + geom_bar()
```

 

This is known in the business as a "stacked bar chart". The issue is
how much of each bar is blue, which is unnecessarily hard to judge
because the male bar is taller. (Here, it is not so bad, because the
amount of blue in the male bar is smaller and the bar is also
taller. But we got lucky here.)

There are two ways to improve this. One is known as a "grouped bar chart", which goes like this:

```{r parasite-9 }
ggplot(infect, aes(x = sex, fill = infected)) +
  geom_bar(position = "dodge")
```

 

The absent and present frequencies for females are next to each other,
and the same for males, and you can read off how big they are from the
$y$-scale. This is my preferred graph for two (or more than two)
categorical variables.

You could switch the roles of `sex` and `infected` and
get a different chart, but one that conveys the same information. Try
it. (The reason for doing it the way around I did is that being
infected or not is the response and `sex` is explanatory, so
that on my plot you can ask "out of the males, how many were infected?", 
which is the way around that makes sense.)

The second way is to go back to stacked bars, but make them the same
height, so you can compare the fractions of the bars that are each
colour. This is `position="fill"`:

```{r parasite-10 }
ggplot(infect, aes(x = sex, fill = infected)) +
  geom_bar(position = "fill")
```

 

This also shows that more of the females were infected than the males,
but without getting sidetracked into the issue that there were more
males to begin with.

I wrote this question in early 2017. At that time, I wrote  (quote):

I learned about this one approximately two hours ago. I just ordered
Hadley Wickham's new book "R for Data Science" from Amazon, and it
arrived today. It's in there. (A good read, by the way. I'm thinking
of using it as a recommended text next year.)
As is so often the way with `ggplot`, the final answer looks
very simple, but there is a lot of thinking required to get there, and
you end up having even more respect for Hadley Wickham for the clarity
of thinking that enabled this to be specified in a simple way.  

(end quote)
 
$\blacksquare$

(c) Which, if any, of your explanatory variables appear to be
related to `infected`? Explain briefly.
 
Solution


Let's go through our output from (<a href="#part:plot">here</a>). In terms of
`age`, when infection is present, animals are (slightly)
older. So there might be a small age effect. Next, when infection
is present, weight is typically a *lot* less. So there ought
to be a big weight effect. 
Finally, from the table, females are
somewhere around 50-50 infected or not, but very few males are
infected. So there ought to be a big `sex` effect as well.
This also appears in the grouped bar plot, where the red
("absent") bar for males is much taller than the blue
("present") bar, but for females the two bars are almost the
same height.
So the story is that we would expect a significant effect of
`sex` and `weight`, and maybe of `age` as well.
 
$\blacksquare$

(d) Fit a logistic regression predicting `infected` from
the other three variables. Display the `summary`.
 
Solution


Thus:
```{r parasite-11, error=T}
infect.1 <- glm(infected ~ age + weight + sex, family = "binomial", data = infect)
```

     

Oh, I forgot to turn `infected` into a factor. This is the
shortcut way to do that:

```{r parasite-12 }
infect.1 <- glm(factor(infected) ~ age + weight + sex, family = "binomial", data = infect)
summary(infect.1)
```

     

The "proper" way to do it is to create a new or redefined column in the data
frame containing the factor version of `infected`. Pipeline
fans among you can do it this way:

```{r parasite-13 }
infect %>%
  mutate(infected = factor(infected)) %>%
  glm(infected ~ age + weight + sex, family = "binomial", data = .) -> infect.1a
summary(infect.1a)
```

 

Either way is good, and gives the same answer. The second way uses the
`data=.` trick to ensure that the input data frame to
`glm` is the output from the previous step, the one with the
factor version of `infected` in it. The `data=.` is
needed because `glm` requires a model formula first rather than
a data frame (if the data were first, you could just omit it). 
 
$\blacksquare$

(e) <a name="part:remove">*</a> Which variables, if any, would you consider removing from
the model? Explain briefly.
 
Solution


This is the same idea as in multiple regression: look at the end
of the line for each variable to get its individual P-value, and
if that's not small, you can take that variable out. `age`
has a P-value of 0.062, which is (just) larger than 0.05, so we
can consider removing this variable. The other two P-values,
0.00089 and 0.024, are definitely less than 0.05, so those
variables should stay.

Alternatively, you can say that the P-value for `age` is
small enough to be interesting, and therefore that `age`
should stay. That's fine, but then you need to be consistent in
the next part.

You probably noted that `sex` is categorical. However, it
has only the obvious two levels, and such a categorical variable
can be assessed for significance this way. If you were worried
about this, the right way to go is `drop1`:

```{r parasite-14 }
drop1(infect.1, test = "Chisq")
```

 

The P-values are similar, but not identical.^[The *test* is this way because it's a generalized linear model rather than a regular regression.]

I have to stop and think about this. There is a lot of theory that
says there are several ways to do stuff in regression, but they are
all identical. The theory doesn't quite apply the same in generalized
linear models (of which logistic regression is one): if you had an
infinite sample size, the ways would all be identical, but in practice
you'll have a very finite amount of data, so they won't agree.

I'm thinking about my aims here: I want to decide whether each
$x$-variable should stay in the model, and for that I want a test that
expresses whether the model fits significantly worse if I take it
out. The result I get ought to be the same as physically removing it
and comparing the models with `anova`, 
eg. for `age`:

```{r parasite-15 }
infect.1b <- update(infect.1, . ~ . - age)
anova(infect.1b, infect.1, test = "Chisq")
```

 

This is the same thing as `drop1` gives. 

So, I think: use `drop1` to assess whether anything should come
out of a model like this, and use `summary` to obtain the
slopes to interpret (in this kind of model, whether they're positive
or negative, and thus what kind of effect each explanatory variable
has on the probability of whatever-it-is.)
 
$\blacksquare$ 

(f) Are the conclusions you drew in (<a href="#part:plot">here</a>) and
(<a href="#part:remove">here</a>) consistent, or not? Explain briefly.
 
Solution


I think they are extremely consistent. When we looked at the
plots, we said that `weight` and `sex` had large
effects, and they came out definitely significant. There was a
small difference in age between the infected and non-infected
groups, and `age` came out borderline significant (with a
P-value definitely larger than for the other variables, so that
the evidence of its usefulness was weaker).

$\blacksquare$ 

(g) <a name="part:crossing">*</a>
The first and third quartiles of `age` are 26 and 130;
the first and third quartiles of `weight` are 9 and 16. Obtain predicted probabilities for all combinations of these and
`sex`. (You'll need to start by making a new data frame, using
`datagrid` to get all the combinations.)
 
Solution

 Here's how `datagrid`
goes. Note my use of plural names to denote
the things I want all combinations of:
```{r parasite-16 }
ages <- c(26, 130)
weights <- c(9, 16)
sexes <- c("female", "male")
new <- datagrid(model = infect.1, age = ages, weight = weights, sex = sexes)
new
```

This is about on the upper end of what you would want to do just using the one line with `datagrid` in it and putting the actual values in instead of `ages`, `weights` etc. To my mind, once it gets longer than about this long, doing on one line starts to get unwieldy. But your taste might be different than mine.
     
Aside:

I could have asked you to include some more values of `age` and
`weight`, for example the median as well, to get a clearer
picture. But that would have made `infect.new` bigger, so I
stopped here. (If we had been happy with *five* representative values of `age` and `weight`, we could have done `predictions` (below) with `variables` and not had to make `new` at all.)

`datagrid` *makes* a data frame from input vectors, so it
doesn't matter if those are different lengths.  In fact, it's also
possible to make this data frame from things like quartiles stored in
a data frame. To do that, you wrap the whole `datagrid` in a `with`:

```{r}
d <- tibble(age = ages, weight = weights, sex = sexes)
d
```

```{r}
new <- with(d, datagrid(model = infect.1, age=age, weight=weight, sex=sex))
new
```

This one is a little confusing because in eg. `age = age`, the first `age` refers to the column that will be in `new`, and the second one refers to the values that are going in there, namely the column called `age` *in the dataframe d*.^[See, I told you it was a little confusing.] 

End of aside.

Next, the predictions:

```{r parasite-17 }
predictions(infect.1, new)
```

 

Extra: I didn't ask you to comment on these, since the question is long
enough already. But that's not going to stop me!

These, in `predicted`, are predicted probabilities of infection.^[When you have one observation per line, the predictions are of the *second* of the two levels of the response variable. When you make that two-column response, the predictions are of the probability of being in the *first* column. That's what it is. As the young people say, don't @ me.]

The way I remember the one-column-response thing is that the first
level is the baseline (as it is in a regression with a categorical
explanatory variable), and the second level is the one whose
probability is modelled (in the same way that the second, third etc.
levels of a categorical explanatory variable are the ones that appear
in the `summary` table).

Let's start with `sex`. The probabilities of a female being
infected are all much higher than of a corresponding male (with the
same age and weight) being infected. Compare, for example, lines 1 and
2. Or 3 and 4. Etc. So `sex` has a big effect.

What about `weight`? As weight goes from 9 to 16, with
everything else the same, the predicted probability of infection goes
sharply *down*. This is what we saw before: precisely, the
boxplot showed us that infected animals were likely to be less heavy.

Last, `age`. As age goes up, the probabilities go (somewhat) up
as well. Compare, for example, lines 1 and 5 or lines 4 and 8. I think
this is a less dramatic change than for the other variables, but
that's a judgement call.

I got this example from (horrible URL warning) here: [link](https://www.amazon.ca/Statistics-Introduction-Michael-J-Crawley/dp/0470022973/ref=pd_sbs_14_3?_encoding=UTF8&pd_rd_i=0470022973&pd_rd_r=97f19951-0ca9-11e9-9275-b3e00a75f9be&pd_rd_w=EfpAv&pd_rd_wg=2NMH8&pf_rd_p=d4c8ffae-b082-4374-b96d-0608daba52bb&pf_rd_r=HAKACF78DSNTGFPDW3YV&psc=1&refRID=HAKACF78DSNTGFPDW3YV)
It starts on page 275 in my edition. He goes at the analysis
a different way, but he finishes with 
another issue that I want to show you. 

Let's work out the residuals and plot them against our quantitative
explanatory variables. I think the best way to do this is
`augment` from `broom`, to create a data frame
containing the residuals alongside the original data:
```{r parasite-18 }
library(broom)
infect.1a <- infect.1 %>% augment(infect)
infect.1a %>% as_tibble()
ggplot(infect.1a, aes(x = weight, y = .resid)) + geom_point() +
  geom_smooth()
```

 

`infect.1a` is, I think, a genuine `data.frame` rather
than a `tibble`.

I don't quite know what to make of that plot. It doesn't look quite
random, and yet there are just some groups of points rather than any
real kind of trend. 

The corresponding plot with age goes the same way:

```{r parasite-19 }
ggplot(infect.1a, aes(x = age, y = .resid)) + geom_point() +
  geom_smooth()
```

 
Crawley found the slightest suggestion of an up-and-down curve in
there. I'm not sure I agree, but that's what he saw.  As with a
regular regression, the residuals against anything should look random,
with no trends. (Though the residuals from a logistic regression can
be kind of odd, because the response variable can only be 1 or 0.)
Crawley tries adding squared terms to the logistic regression, which
goes like this. The `glm` statement is long, as they usually
are, so it's much easier to use `update`:

```{r parasite-20 }
infect.2 <- update(infect.1, . ~ . + I(age^2) + I(weight^2))
```

 

As we saw before, when thinking about what to keep, we want to look at `drop1`:

```{r parasite-21 }
drop1(infect.2, test = "Chisq")
```

 

The squared terms are both significant. The linear terms,
`age` and `weight`, have to stay, regardless of their
significance.^[When you have higher-order terms, you have to keep the lower-order ones as well: higher powers, or interactions (as we see in ANOVA later).]
What do the squared terms do to the predictions? Before, there was a
clear one-directional trend in the relationships with `age` and
`weight`. Has that changed? Let's see. We'll need a few more
ages and weights to investigate with. I think the set of five representative values that comes out of `predictions` with `variables` would be ideal (and saves us having to make another `new`). 

Let's start by assessing the effect of age:

```{r}
predictions(infect.2, variables = "age") %>% 
  select(predicted, age, weight, sex)
```

The actual values of age chosen by `predictions` are as shown. The other columns are constant; the values are the mean weight and the more common sex. We really can see the effect of age with all else constant.

Here, the predicted infection probabilities go up with age and then come down again, as a squared term in age will allow them to do, compared to what we had before:

```{r}
predictions(infect.1, variables = "age") %>% 
  select(predicted, age, weight, sex)
```

where the probabilities keep on going up.

All right, what about the effect of weight? Here's the first model:

```{r}
predictions(infect.1, variables = "weight") %>% 
  select(predicted, age, weight, sex)
```

and here's the second one with squared terms:

```{r}
predictions(infect.2, variables = "weight") %>% 
  select(predicted, age, weight, sex)
```

This one is not so dissimilar: in the linear model, the predicted probabilities of infection start high and go down, but in the model with squared terms they go slightly up before going down.

We couldn't have a squared term in `sex`, since there are only two sexes (in this data set), so the predictions ought to be pretty similar for the two models:

```{r}
predictions(infect.1, variables = "sex") %>% 
  select(predicted, age, weight, sex)
```

and

```{r}
predictions(infect.2, variables = "sex") %>% 
  select(predicted, age, weight, sex)
```

They are actually quite different.
For the model with squared terms in age and weight, the predicted probabilities of infection are a lot higher for both males and females, at least at these (mean) ages and weights.

$\blacksquare$ 
 


