---
title: "Problems and Solutions in Applied Statistics"
author: "Ken Butler"
date: "2021-04-26"
site: bookdown::bookdown_site
documentclass: book
bibliography: []
biblio-style: apalike
link-citations: yes
github-repo: nxskok/pasias
url: 'http://ritsokiguess.site/pasias'
description: "A set of problems and solutions, in R, on various parts of applied statistics"
output:
  pdf_document:
    toc: true
    toc_depth: 1
---

# Introduction

Placeholder


## Packages used somewhere in this book

<!--chapter:end:index.Rmd-->


# Getting used to R and R Studio

Placeholder



<!--chapter:end:01-getting-used.Rmd-->


# Reading in data

Placeholder



<!--chapter:end:02-reading-in.Rmd-->


# Drawing graphs

Placeholder



<!--chapter:end:32-making-graphs.Rmd-->


# Data exploration

Placeholder



<!--chapter:end:03-data-summaries.Rmd-->

# Working with dataframes

```{r, eval=F}
library(tidyverse)
```

```{r, child="packages.Rmd"}
```

```{r jays-dplyr, child="jays-dplyr_qq.Rmd"}
```

```{r kencars,child="kencars_qq.Rmd"}
```

My solutions follow:

```{r jays-dplyr, child="jays-dplyr.Rmd"}
```

```{r kencars,child="kencars.Rmd"}
```

<!--chapter:end:31-working-with-dataframes.Rmd-->


# One-sample inference

Placeholder



<!--chapter:end:04-one-sample-inference.Rmd-->


# Two-sample inference

Placeholder



<!--chapter:end:05-two-sample-inference.Rmd-->


# Power and sample size

Placeholder



<!--chapter:end:06-power.Rmd-->


# The sign test 

Placeholder



<!--chapter:end:07-sign.Rmd-->


# Mood's median test

Placeholder



<!--chapter:end:33-mood-median.Rmd-->


# Matched pairs t and sign test

Placeholder



<!--chapter:end:08-matched-pairs-sign.Rmd-->

# Normal quantile plots

```{r, eval=F}
library(tidyverse)
```

```{r, child="packages.Rmd"}
```


```{r heliconia, child="heliconia_qq.Rmd"}
```

```{r aus-qq, child="aus-qq_qq.Rmd"}
```

My solutions follow:

```{r heliconia, child="heliconia.Rmd"}
```

```{r aus-qq, child="aus-qq.Rmd"}
```

<!--chapter:end:09-normal-quantile.Rmd-->


# Analysis of variance

Placeholder



<!--chapter:end:10-analysis-of-variance.Rmd-->

# Writing reports

```{r, eval=F}
library(tidyverse)
```

```{r, child="packages.Rmd"}
```


```{r carbon, child="carbon_qq.Rmd"}
```

```{r sparrowhawk, child="sparrowhawk_qq.Rmd"}
```

My solutions follow:

```{r carbon, child="carbon.Rmd"}
```

```{r sparrowhawk, child="sparrowhawk.Rmd"}
```

<!--chapter:end:34-reports.Rmd-->


# Tidying data

Placeholder



<!--chapter:end:11-tidying-data.Rmd-->


# Simple regression

Placeholder



<!--chapter:end:12-simple-regression.Rmd-->


# Multiple regression

Placeholder



<!--chapter:end:35-multiple-regression.Rmd-->

# Regression with categorical variables

```{r, eval=F}
library(tidyverse)
```

```{r, child="packages.Rmd"}
```

```{r crickets-b, child="crickets-b_qq.Rmd"}
```

My solutions follow:

```{r crickets-b, child="crickets-b.Rmd"}
```

<!--chapter:end:36-regression-categorical.Rmd-->


# Dates and times

Placeholder



<!--chapter:end:13-dates-and-times.Rmd-->

# Functions

```{r, eval=F}
library(tidyverse)
```

```{r, child="packages.Rmd"}
```


```{r func, child="func_qq.Rmd"}
```

```{r hotpo, child="hotpo_qq.Rmd"}
```

My solutions follow:

```{r func, child="func.Rmd"}
```

```{r hotpo, child="hotpo.Rmd"}
```

<!--chapter:end:14-functions.Rmd-->

# Vector and matrix algebra

```{r, eval=F}
library(tidyverse)
```

```{r, child="packages.Rmd"}
```

```{r, child =  "reg_qq.Rmd"}
```


My solutions follow:

```{r, child =  "reg.Rmd"}
```


<!--chapter:end:51-vector-matrix.Rmd-->


# The Bootstrap

Placeholder



<!--chapter:end:26-bootstrap.Rmd-->

# Bayesian Statistics with Stan

Packages for this chapter:

```{r eval = FALSE}
library(tidyverse)
library(cmdstanr)
library(posterior)
library(bayesplot)
```

```{r, child="packages.Rmd", warning=F}
```


```{r, child="binomial_stan.Rmd"}
```

```{r, child="regression_stan.Rmd"}
```

My solutions follow:

```{r, child="binomial_stan.Rmd"}
```

```{r, child="regression_stan.Rmd"}
```


<!--chapter:end:27-stan.Rmd-->


# Logistic regression

Placeholder



<!--chapter:end:15-logistic-regression.Rmd-->


# Logistic regression with ordinal response

Placeholder



<!--chapter:end:16-ordinal-response.Rmd-->


# Logistic regression with nominal response

Placeholder



<!--chapter:end:37-nominal-response.Rmd-->


# Survival analysis

Placeholder



<!--chapter:end:17-survival-analysis.Rmd-->


# Analysis of variance revisited

Placeholder



<!--chapter:end:18-anova-revisited.Rmd-->


# Analysis of covariance

Placeholder



<!--chapter:end:38-ancova.Rmd-->


# Multivariate analysis of variance

Placeholder



<!--chapter:end:19-manova.Rmd-->


# Repeated measures

Placeholder



<!--chapter:end:20-repeated-measures.Rmd-->


# Discriminant analysis

Placeholder



<!--chapter:end:21-discriminant-analysis.Rmd-->


# Hierarchical cluster analysis

Placeholder



<!--chapter:end:22-cluster.Rmd-->


# K-means cluster analysis

Placeholder



<!--chapter:end:39-kmeans-cluster.Rmd-->


# Drawing maps with Leaflet

Placeholder



<!--chapter:end:41-maps.Rmd-->


# Multidimensional Scaling

Placeholder



<!--chapter:end:23-mds.Rmd-->

# Principal Components

Packages for this chapter:

```{r, eval=F}
library(ggbiplot)
library(tidyverse)
```

```{r, child="packages.Rmd"}
```

```{r weather, child="weather_2014_qq.Rmd"}
``` 


My solutions follow:

```{r weather, child="weather_2014.Rmd"}
``` 


	


<!--chapter:end:24-pc.Rmd-->


# Factor Analysis

Placeholder



<!--chapter:end:42-fa.Rmd-->

# Frequency table analysis

Packages for this chapter:

```{r, eval=F}
library(tidyverse)
```
```{r, child="packages.Rmd"}
```

```{r college, child="college-plans_qq.Rmd"}
```

```{r vote, child="vote_qq.Rmd"}
```

My solutions follow:

```{r college, child="college-plans.Rmd"}
```

```{r vote, child="vote.Rmd"}
```

<!--chapter:end:25-loglin.Rmd-->

