---
title: "Problems and Solutions in Applied Statistics"
author: "Ken Butler"
date: "2021-04-26"
site: bookdown::bookdown_site
documentclass: book
bibliography: []
biblio-style: apalike
link-citations: yes
github-repo: nxskok/pasias
url: 'http://ritsokiguess.site/pasias'
description: "A set of problems and solutions, in R, on various parts of applied statistics"
output:
  pdf_document:
    toc: true
    toc_depth: 2
---

# Introduction

Placeholder


## Packages used somewhere in this book

<!--chapter:end:index.Rmd-->


# Getting used to R and R Studio

We begin with this:

```{r, eval=F}
library(tidyverse)
```

```{r, child="packages.Rmd"}
```

and so to the problems:

```{r, child="r-signup.Rmd"}
```

```{r, child="r0.Rmd"}
```

```{r, child="r0a.Rmd"}
```

```{r different-ways, child="different-ways.Rmd"}
```

<!--chapter:end:01-getting-used.Rmd-->


# Reading in data

```{r, eval=F}
library(tidyverse)
```

```{r, child="packages.Rmd"}
```


```{r, child="oj.Rmd"}
```

```{r, child="soap.Rmd"}
```

```{r, child="global.Rmd"}
```



<!--chapter:end:02-reading-in.Rmd-->


# Drawing graphs

```{r, eval=F}
library(tidyverse)
```

```{r, child="packages.Rmd"}
```


```{r, child="oj2.Rmd"}
```

```{r, child="soap2.Rmd"}
```

```{r, child="global2.Rmd"}
```



<!--chapter:end:32-making-graphs.Rmd-->

# Data exploration

```{r, eval=F}
library(tidyverse)
```

```{r, child="packages.Rmd"}
```


```{r, child="ncbirths.Rmd"}
```

```{r, child="ncbirths-extra.Rmd"}
```

```{r, child="nenana.Rmd"}
```

```{r, child="compatt.Rmd"}
```

```{r, child="thomas.Rmd"}
```

<!--chapter:end:03-data-summaries.Rmd-->

# Working with dataframes

```{r, eval=F}
library(tidyverse)
```

```{r, child="packages.Rmd"}
```

```{r jays-dplyr, child="jays-dplyr.Rmd"}
```

```{r kencars,child="kencars.Rmd"}
```

<!--chapter:end:31-working-with-dataframes.Rmd-->

# One-sample inference

```{r, eval=F}
library(tidyverse)
```

```{r, child="packages.Rmd"}
```

```{r, child="hg.Rmd"}
```

```{r, child="boulder.Rmd"}
```

```{r, child="ncbirths-inf.Rmd"}
```

```{r, child="nenana-inf.Rmd"}
```


<!--chapter:end:04-one-sample-inference.Rmd-->

# Two-sample inference

```{r, eval=F}
library(tidyverse)
```

```{r, child="packages.Rmd"}
```


```{r devices-r, child="devices-r.Rmd"}
```

```{r parking, child="parking.Rmd"}
```

```{r bellpepper, child="bellpepper.Rmd"}
```

```{r mousebully, child="mousebully.Rmd"}
```

```{r kids-diet, child="kids-diet.Rmd"}
```

<!--chapter:end:05-two-sample-inference.Rmd-->

# Power and sample size

```{r, eval=F}
library(tidyverse)
```

```{r, child="packages.Rmd"}
```


```{r power-sim, child="power-sim.Rmd"}
```

```{r power20, child="power20.Rmd"}
```

```{r, proportion-power-sim, child="proportion-power-sim.Rmd"}
```


<!--chapter:end:06-power.Rmd-->

# The sign test 

```{r}
library(tidyverse)
library(smmr)
```

```{r, child="packages.Rmd"}
```


```{r maze, child="maze.Rmd"}
```

```{r cookies-r, child="cookies-r.Rmd"}
```

```{r signpower, child="signpower.Rmd"}
```


<!--chapter:end:07-sign.Rmd-->

# Mood's median test

```{r}
library(tidyverse)
library(smmr)
```

```{r, child="packages.Rmd"}
```


```{r cerealsugar, child="cerealsugar.Rmd"}
```

```{r mathphobia, child="mathphobia.Rmd"}
```

```{r forehead, child="forehead.Rmd"}
```

<!--chapter:end:33-mood-median.Rmd-->

# Matched pairs t and sign test

```{r, eval=F}
library(tidyverse)
library(smmr)
```

```{r, child="packages.Rmd"}
```


```{r bodyfat, child="bodyfat.Rmd"}
```

```{r baseball-softball-r, child="baseball-softball-r.Rmd"}
```

```{r throw, child="throw.Rmd"}
```


```{r salaryinc, child="salaryinc.Rmd"}
```

```{r bodyfat-sign, child="bodyfat-sign.Rmd"}
```


<!--chapter:end:08-matched-pairs-sign.Rmd-->

# Normal quantile plots

```{r, eval=F}
library(tidyverse)
```

```{r, child="packages.Rmd"}
```


```{r heliconia, child="heliconia.Rmd"}
```

```{r aus-qq, child="aus-qq.Rmd"}
```

<!--chapter:end:09-normal-quantile.Rmd-->

# Analysis of variance

```{r, eval=F}
library(tidyverse)
```

```{r, child="packages.Rmd"}
```


```{r movies-r, child="movies-r.Rmd"}
```

```{r deer, child="deer.Rmd"}
```

```{r movies2, child="movies2.Rmd"}
```

```{r carbon, child="carbon.Rmd"}
```

```{r caffeine, child="caffeine.Rmd"}
```


<!--chapter:end:10-analysis-of-variance.Rmd-->

# Writing reports

```{r, eval=F}
library(tidyverse)
```

```{r, child="packages.Rmd"}
```


```{r carbon, child="carbon.Rmd"}
```

```{r sparrowhawk, child="sparrowhawk.Rmd"}
```

<!--chapter:end:34-reports.Rmd-->


# Tidying data

Placeholder



<!--chapter:end:11-tidying-data.Rmd-->


# Simple regression

Placeholder



<!--chapter:end:12-simple-regression.Rmd-->

# Multiple regression

```{r, eval=F}
library(tidyverse)
```

```{r, child="packages.Rmd"}
```


```{r hospital, child="hospital.Rmd"}
```

```{r chemical, child="chemical.Rmd"}
```

```{r mathsal, child="mathsal.Rmd"}
```

```{r gpa, child="gpa.Rmd"}
```

<!--chapter:end:35-multiple-regression.Rmd-->

# Regression with categorical variables

```{r, eval=F}
library(tidyverse)
```

```{r, child="packages.Rmd"}
```

```{r crickets-b, child="crickets-b.Rmd"}
```

<!--chapter:end:36-regression-categorical.Rmd-->

# Dates and times

The usual to begin with:

```{r, eval=F}
library(tidyverse)
```

```{r, child="packages.Rmd"}
```


```{r mizuna-r, child="mizuna-r.Rmd"}
```

```{r birthtypes, child="birthtypes.Rmd"}
```

```{r denali, child="denali-r.Rmd"}
```

```{r whas, child="whas.Rmd"}
```

<!--chapter:end:13-dates-and-times.Rmd-->

# Functions

```{r, eval=F}
library(tidyverse)
```

```{r, child="packages.Rmd"}
```


```{r func, child="func.Rmd"}
```

```{r hotpo, child="hotpo.Rmd"}
```

<!--chapter:end:14-functions.Rmd-->

# The Bootstrap

Packages for this chapter:

```{r, eval=F}
library(tidyverse)
library(bootstrap)
```

```{r, child="packages.Rmd", warning=F}
```

```{r boot_one, child="bootstrap.Rmd"}
```

```{r boot_two, child="bootstrap_median.Rmd"}
```

<!--chapter:end:26-bootstrap.Rmd-->

# Bayesian Statistics with Stan

Packages for this chapter:

```{r eval = FALSE}
library(tidyverse)
library(cmdstanr)
library(posterior)
library(bayesplot)
```

```{r, child="packages.Rmd", warning=F}
```


```{r, child="binomial_stan.Rmd"}
```

```{r, child="regression_stan.Rmd"}
```

<!--chapter:end:27-stan.Rmd-->


# Logistic regression

Placeholder



<!--chapter:end:15-logistic-regression.Rmd-->

# Logistic regression with ordinal or nominal response

```{r, eval=F}
library(MASS)
library(tidyverse)
```
```{r, child="packages.Rmd"}
```

```{r mobile, child="mobile.Rmd"}
```


```{r nonmissing, child="nonmissing.Rmd"}
```



```{r hsb, child="hsb.Rmd"}
```


```{r steak, child="steak.Rmd"}
```

```{r steak-data, child="steak-data.Rmd"}
```

<!--chapter:end:16-ordinal-response.Rmd-->


# Logistic regression with ordinal response

Placeholder



<!--chapter:end:37-nominal-response.Rmd-->

# Survival analysis

```{r, eval=F}
library(survival)
library(tidyverse)
library(survminer)
```

```{r, child="packages.Rmd"}
```

```{r worcester, child="worcester.Rmd"}
```

```{r drug-treatment, child="drug-treatment.Rmd"}
```

```{r myeloma, child="myeloma.Rmd"}
```

```{r ovarian, child="ovarian.Rmd"}
```


<!--chapter:end:17-survival-analysis.Rmd-->


# Analysis of variance revisited

Placeholder



<!--chapter:end:18-anova-revisited.Rmd-->

# Analysis of covariance

Packages for this chapter:

```{r, eval=F}
library(MASS)
library(car)
library(tidyverse)
```

```{r, child="packages.Rmd"}
```



```{r productivity, child="productivity.Rmd"}
```

```{r lepro, child="leprosy.Rmd"} 
```





<!--chapter:end:38-ancova.Rmd-->

# Multivariate analysis of variance

Packages for this chapter:

```{r, eval=F}
library(car)
library(tidyverse)
```

```{r, child="packages.Rmd"}
```


```{r q7,child="manova1.Rmd"}
``` 

```{r urine,child="urine.Rmd"}
```   

```{r athletes-manova, child="athletes-manova.Rmd"}
```


<!--chapter:end:19-manova.Rmd-->


# Repeated measures

Placeholder



<!--chapter:end:20-repeated-measures.Rmd-->


# Discriminant analysis

Placeholder



<!--chapter:end:21-discriminant-analysis.Rmd-->

# Hierarchical cluster analysis

Packages for this chapter:

```{r, eval=F}
library(MASS)
library(ggbiplot)
library(tidyverse)
```

```{r, child="packages.Rmd"}
```


```{r seabed, child="seabed.Rmd"}
```


```{r fruit,child="fruits.Rmd"}
``` 

```{r species,child="species.Rmd"}
```


```{r pittsburgh, child="pittsburgh.Rmd"}
``` 


<!--chapter:end:22-cluster.Rmd-->

# K-means cluster analysis

Packages for this chapter:

```{r, eval=F}
library(MASS)
library(ggbiplot)
library(tidyverse)
```

```{r, child="packages.Rmd"}
```


```{r athletes-km, child="ais-km.Rmd"}
```

```{r decathlon, child="decathlon.Rmd"}
``` 

```{r swiss-cluster, child="swiss-cluster.Rmd"}
```


```{r car-cluster, child="carc.Rmd"}
```

```{r beer,child="beer.Rmd"}
``` 

<!--chapter:end:39-kmeans-cluster.Rmd-->

# Drawing maps with Leaflet

Packages for this chapter:

```{r, eval=F}
library(ggbiplot)
library(tidyverse)
library(ggrepel)
```

```{r, child="packages.Rmd"}
```

```{r, child="catbrain.Rmd"}
```


```{r wisc, child="wisconsin2.Rmd"}
``` 


<!--chapter:end:41-maps.Rmd-->


# Multidimensional Scaling

Placeholder



<!--chapter:end:23-mds.Rmd-->

# Principal Components

Packages for this chapter:

```{r, eval=F}
library(ggbiplot)
library(tidyverse)
```

```{r, child="packages.Rmd"}
```

```{r weather, child="weather_2014.Rmd"}
``` 



	


<!--chapter:end:24-pc.Rmd-->

# Factor Analysis

Packages for this chapter:

```{r, eval=F}
library(ggbiplot)
library(tidyverse)
```

```{r, child="packages.Rmd"}
```

```{r ipip, child="ipip.Rmd"}
``` 

```{r corrmat, child="corrmat.Rmd"}
```   

```{r pollution, child="airpollution.Rmd"}
```

<!--chapter:end:42-fa.Rmd-->

# Frequency table analysis

Packages for this chapter:

```{r, eval=F}
library(tidyverse)
```
```{r, child="packages.Rmd"}
```


```{r college, child="college-plans.Rmd"}
```

```{r vote, child="vote.Rmd"}
```

<!--chapter:end:25-loglin.Rmd-->

