---
title: "Problems and Solutions in Applied Statistics"
author: "Ken Butler"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: []
biblio-style: apalike
link-citations: yes
github-repo: nxskok/pasias
url: 'http://ritsokiguess.site/pasias'
description: "A set of problems and solutions, in R, on various parts of applied statistics"
output:
  pdf_document:
    toc: true
    toc_depth: 2
---

# Introduction {-}

Placeholder


## Packages used somewhere in this book {-}

<!--chapter:end:index.Rmd-->


# Getting used to R and R Studio

In this chapter, my solutions are in with the problems.

```{r 01-getting-used-1, child="packages.Rmd"}
```

```{r 01-getting-used-2, child="jupyter.Rmd"}
```

```{r 01-getting-used-3, child="r-signup.Rmd"}
```

```{r 01-getting-used-4, child="r0.Rmd"}
```

```{r 01-getting-used-5, child="r0a.Rmd"}
```

```{r 01-getting-used-6, child="different-ways.Rmd"}
```



<!--chapter:end:01-getting-used.Rmd-->


# Reading in data

Placeholder



<!--chapter:end:02-reading-in.Rmd-->


# Drawing graphs

Placeholder



<!--chapter:end:32-making-graphs.Rmd-->


# Data exploration

Placeholder



<!--chapter:end:03-data-summaries.Rmd-->


# Working with dataframes

Placeholder



<!--chapter:end:31-working-with-dataframes.Rmd-->


# One-sample inference

Placeholder



<!--chapter:end:04-one-sample-inference.Rmd-->


# Two-sample inference

Placeholder



<!--chapter:end:05-two-sample-inference.Rmd-->


# Power and sample size

Placeholder



<!--chapter:end:06-power.Rmd-->


# The sign test

Placeholder



<!--chapter:end:07-sign.Rmd-->


# Mood median test

Placeholder



<!--chapter:end:33-mood-median.Rmd-->


# Matched pairs t and sign test

Placeholder



<!--chapter:end:08-matched-pairs-sign.Rmd-->

# Normal quantile plots

```{r 09-normal-quantile-1, eval=F}
library(tidyverse)
```

```{r 09-normal-quantile-2, child="packages.Rmd"}
```


```{r 09-normal-quantile-3, child="heliconia_qq.Rmd"}
```

```{r 09-normal-quantile-4, child="aus-qq_qq.Rmd"}
```

My solutions follow:

```{r 09-normal-quantile-5, child="heliconia.Rmd"}
```

```{r 09-normal-quantile-6, child="aus-qq.Rmd"}
```

<!--chapter:end:09-normal-quantile.Rmd-->


# Analysis of variance

Placeholder



<!--chapter:end:10-analysis-of-variance.Rmd-->


# Writing reports

Placeholder



<!--chapter:end:34-reports.Rmd-->

# Learning to code

```{r 61-coding-1, eval=F}
library(tidyverse)
```

```{r 61-coding-2, child="packages.Rmd"}
```

```{r 61-coding-3, child="proggo.Rmd"}
```


<!--chapter:end:61-coding.Rmd-->

# Treating dandruff

```{r 62-dandruff-1, child="packages.Rmd"}
```

```{r 62-dandruff-2, child="dandruff.Rmd"}
```

<!--chapter:end:62-dandruff.Rmd-->


# Tidying data

Placeholder



<!--chapter:end:11-tidying-data.Rmd-->


# Simple regression

Placeholder



<!--chapter:end:12-simple-regression.Rmd-->


# Multiple regression

Placeholder



<!--chapter:end:35-multiple-regression.Rmd-->

# Regression with categorical variables

```{r 36-regression-categorical-1, eval=F}
library(tidyverse)
```

```{r 36-regression-categorical-2, child="packages.Rmd"}
```

```{r 36-regression-categorical-3, child="crickets-b_qq.Rmd"}
```

```{r 36-regression-categorical-4, child="pulsemarch_qq.Rmd"}
```

My solutions follow:

```{r 36-regression-categorical-5, child="crickets-b.Rmd"}
```

```{r 36-regression-categorical-6, child="pulsemarch.Rmd"}
```


<!--chapter:end:36-regression-categorical.Rmd-->


# Dates and times

Placeholder



<!--chapter:end:13-dates-and-times.Rmd-->


# Functions

Placeholder



<!--chapter:end:14-functions.Rmd-->

# Vector and matrix algebra

```{r 51-vector-matrix-1, eval=F}
library(tidyverse)
```

```{r 51-vector-matrix-2, child="packages.Rmd"}
```

```{r 51-vector-matrix-3, child="reg_qq.Rmd"}
```


My solutions follow:

```{r 51-vector-matrix-4, child="reg.Rmd"}
```


<!--chapter:end:51-vector-matrix.Rmd-->


# The Bootstrap

Placeholder



<!--chapter:end:26-bootstrap.Rmd-->


# Bayesian Statistics with Stan

Placeholder



<!--chapter:end:27-stan.Rmd-->


# Logistic regression

Placeholder



<!--chapter:end:15-logistic-regression.Rmd-->


# Logistic regression with ordinal response

Placeholder



<!--chapter:end:16-ordinal-response.Rmd-->


# Logistic regression with nominal response

Placeholder



<!--chapter:end:37-nominal-response.Rmd-->


# Survival analysis

Placeholder



<!--chapter:end:17-survival-analysis.Rmd-->


# Analysis of variance revisited

Placeholder



<!--chapter:end:18-anova-revisited.Rmd-->


# Analysis of covariance

Placeholder



<!--chapter:end:38-ancova.Rmd-->


# Multivariate analysis of variance

Placeholder



<!--chapter:end:19-manova.Rmd-->


# Repeated measures

Placeholder



<!--chapter:end:20-repeated-measures.Rmd-->


# Discriminant analysis

Placeholder



<!--chapter:end:21-discriminant-analysis.Rmd-->

# Hierarchical cluster analysis

Packages for this chapter:

```{r 22-cluster-1, eval=F}
library(MASS)
library(ggbiplot)
library(tidyverse)
```

```{r 22-cluster-2, child="packages.Rmd"}
```


```{r 22-cluster-3, child="seabed_qq.Rmd"}
```


```{r 22-cluster-4,child="fruits_qq.Rmd"}
``` 

```{r 22-cluster-5,child="species_qq.Rmd"}
```


```{r 22-cluster-6, child="pittsburgh_qq.Rmd"}
``` 

My solutions follow:

```{r 22-cluster-7, child="seabed.Rmd"}
```


```{r 22-cluster-8,child="fruits.Rmd"}
``` 

```{r 22-cluster-9,child="species.Rmd"}
```


```{r 22-cluster-10, child="pittsburgh.Rmd"}
``` 


<!--chapter:end:22-cluster.Rmd-->


# K-means cluster analysis

Placeholder



<!--chapter:end:39-kmeans-cluster.Rmd-->


# Drawing maps with Leaflet

Placeholder



<!--chapter:end:41-maps.Rmd-->


# Multidimensional Scaling

Placeholder



<!--chapter:end:23-mds.Rmd-->

# Principal Components

Packages for this chapter:

```{r 24-pc-1, eval=F}
library(ggbiplot)
library(tidyverse)
```

```{r 24-pc-2, child="packages.Rmd"}
```

```{r 24-pc-3, child="weather_2014_qq.Rmd"}
``` 


My solutions follow:

```{r 24-pc-4, child="weather_2014.Rmd"}
``` 


	


<!--chapter:end:24-pc.Rmd-->


# Factor Analysis

Placeholder



<!--chapter:end:42-fa.Rmd-->


# Frequency table analysis

Placeholder



<!--chapter:end:25-loglin.Rmd-->

