##  Bridges in Pittsburgh


 The city of Pittsburgh, Pennsylvania, lies where three
rivers, the Allegheny, Monongahela, and Ohio, 
meet.^[For a long  time, the Pittsburgh Steelers football team played at the Three  Rivers Stadium.] 
It has long been important to build bridges there,
to enable its residents to cross the rivers safely. See
[link](https://en.wikipedia.org/wiki/List_of_bridges_of_Pittsburgh) for
a listing (with pictures) of the bridges. The data at
[link](https://raw.githubusercontent.com/nxskok/datafiles/master/bridges.csv) contains
detail for a large number of past and present bridges in
Pittsburgh. All the variables we will use are categorical. Here they
are:



* `id` identifying the bridge (we ignore)

* `river`: initial letter of river that the bridge crosses

* `location`: a numerical code indicating the location
within Pittsburgh (we ignore)

* `erected`: time period in which the bridge was built (a
name, from `CRAFTS`, earliest, to `MODERN`, most
recent.

* `purpose`: what the bridge carries: foot traffic
("walk"), water (aqueduct), road or railroad.

* `length` categorized as long, medium or short.

* `lanes` of traffic (or number of railroad tracks): a
number, 1, 2, 4 or 6, that we will count as categorical.

* `clear_g`: whether a vertical navigation requirement was
included in the bridge design (that is, ships of a certain height
had to be able to get under the bridge). I think `G` means
"yes". 

* `t_d`: method of construction. `DECK` means the
bridge deck is on top of the construction, `THROUGH` means
that when you cross the bridge, some of the bridge supports are next
to you or above you. 

* `material` the bridge is made of: iron, steel or wood.

* `span`: whether the bridge covers a short, medium or long
distance. 

* `rel_l`: Relative length of the main span of the
bridge (between the two central piers) to the total crossing length.
The categories are `S`, `S-F` and `F`. I don't
know what these mean.

* `type` of bridge: wood, suspension, arch and three types
of truss bridge: cantilever, continuous and simple.


The website [link](https://www.steelconstruction.info/Bridges) is an
excellent source of information about bridges. (That's where I learned
the difference between `THROUGH` and `DECK`.) Wikipedia
also has a good article at
[link](https://en.wikipedia.org/wiki/Bridge). I also found
[link](http://www.metaeventos.net/inscricoes/formularios_off/resumo_preenchido/DINCON/Pina-A_Colimodio-V_Silva-A_PinaFilho-A.pdf)
which is the best description I've seen of the variables.



(a) The bridges are stored in CSV format. Some of the
information is not known and was recorded in the spreadsheet as
`?`. Turn these into genuine missing values by adding
`na="?"` to your file-reading command. Display some of your
data, enough to see that you have some missing data.

Solution


This sort of thing:

```{r pittsburgh-1 }
my_url <- "https://raw.githubusercontent.com/nxskok/datafiles/master/bridges.csv"
bridges0 <- read_csv(my_url, na = "?")
bridges0
```

 

I have some missing values in the `length` column. (You
sometimes see `<NA>` instead of `NA`, as you do here;
this means the missing value is a missing piece of text rather than a
missing number.)^[Sometimes it's necessary to distinguish  between the different types of missing value; if that's the case,  you can use eg. `NA_real_` and `NA_character_-` to distinguish missing decimal numbers from missing text.]

There are 108 bridges in the data set.

I'm saving the name `bridges` for my final data set, after I'm
finished organizing it.

$\blacksquare$

(b) Verify that there are missing values in this dataset. To see
them, convert the text columns temporarily to `factor`s using
`mutate`, and pass the resulting data frame into
`summary`.

Solution


I called my data frame `bridges0`, so this:
```{r pittsburgh-2 }
bridges0 %>%
  mutate(across(where(is.character), \(x) factor(x))) %>%
  summary()
```

   

There are missing values all over the place. `length` has the
most, but `lanes` and `span` also have a fair few.

`mutate` requires `across` and a logical condition, something that is true or false, about each column, and then something to do with it.
In words, "for each column that is text, replace it (temporarily) with the factor version of itself."

Extra: I think the reason `summary` doesn't handle text stuff very
well is that, originally, text columns that were read in from files
*got turned into* factors, and if you didn't want that to happen,
you had to explicitly stop it yourself. Try mentioning
`stringsAsFactors=F` to a veteran R user, and watch their
reaction, or try it yourself by reading in a data file with text
columns using `read.table` instead of
`read_delim`. (This will read in an old-fashioned data frame,
so pipe it through `as_tibble` to see what the columns are.)

When Hadley Wickham designed `readr`, the corner of the
`tidyverse` where the `read_` functions live, he
deliberately chose to keep text as text (on the basis of being honest
about what kind of thing we have), with the result that we sometimes
have to create factors when what we are using requires them rather
than text.

$\blacksquare$

(c) Use `drop_na` to remove any rows of the data frame with missing values in them. How many rows do you have left?

Solution


This is as simple as:
```{r pittsburgh-3 }
bridges0 %>% drop_na() -> bridges
bridges
```

     
I have 70 rows left (out of the original 108).

$\blacksquare$

(d) We are going to assess the dissimilarity between two bridges
by the number of the categorical variables they *disagree*
on. This is called a "simple matching coefficient", and is the
same thing we did in the question about clustering fruits based on
their properties. This time, though, we want to count matches in
things that are *rows* of our data frame (properties of two
different bridges), so we will need to use a strategy like the one I
used in calculating the Bray-Curtis distances.
First, write a function that takes as input two vectors `v`
and `w` and counts the number of their entries that differ
(comparing the first with the first, the second with the second,
\ldots, the last with the last. I can think of a quick way and a
slow way, but either way is good.) To test your function, create two
vectors (using `c`) of the same length, and see whether it
correctly counts the number of corresponding values that are
different. 

Solution


The slow way is to loop through the elements of each vector, using
square brackets to pull out the ones you want, checking them for
differentness, then updating a counter which gets returned at the
end. If you've done Python, this is exactly the strategy you'd
use there:
```{r pittsburgh-4 }
count_diff <- function(v, w) {
  n <- length(v)
  stopifnot(length(v) == length(w)) # I explain this below
  count <- 0
  for (i in 1:n) {
    if (v[i] != w[i]) count <- count + 1
  }
  count
}
```

     

This function makes no sense if `v` and `w` are of
different lengths, since we're comparing *corresponding* elements
of them. The `stopifnot` line checks to see whether `v`
and `w` have the same number of things in them, and stops with
an informative error if they are of different lengths. (The thing
inside the `stopifnot` is what has to be *true*.)

Does it work?

```{r pittsburgh-5 }
v <- c(1, 1, 0, 0, 0)
w <- c(1, 2, 0, 2, 0)
count_diff(v, w)
```

 

Three of the values are the same and two are different, so this is right.

What happens if my two vectors are of different lengths?

```{r pittsburgh-6, error=T}
v1 <- c(1, 1, 0, 0)
w <- c(1, 2, 0, 2, 0)
count_diff(v1, w)
```

 

Error, as produced by `stopifnot`. See how it's perfectly clear
what went wrong?

R, though, is a "vectorized" language: it's possible to work with
whole vectors at once, rather than pulling things out of them one at a
time. Check out this (which is like what I did with the fruits):

```{r pittsburgh-7 }
v != w
```

 

The second and fourth values are different, and the others are the
same. But we can go one step further:

```{r pittsburgh-8 }
sum(v != w)
```

 

The true values count as 1 and the false ones as zero, so the sum is
counting up how many values are different, exactly what we want. So
the function can be as simple as:

```{r pittsburgh-9 }
count_diff <- function(v, w) {
  sum(v != w)
}
```

 

I still think it's worth writing a function do this, though, since
`count_diff` tells you what it does and `sum(v!=w)`
doesn't, unless you happen to know.

$\blacksquare$

(e) Write a function that has as input two row numbers and a data
frame to take those rows from. The function needs to select all the
columns except for `id` and `location`, select the
rows required one at a time, and turn them into vectors. (There may
be some repetitiousness here. That's OK.)  Then those two vectors
are passed into the function you wrote in the previous part, and the
count of the number of differences is returned. This is like the
code in the Bray-Curtis problem. Test your function on rows 3 and 4
of your bridges data set (with the missings removed). There should
be six variables that are different.

Solution


This is just like my function `braycurtis.spec`, except that
instead of calling `braycurtis` at the end, I call
`count_diff`: 

```{r pittsburgh-10 }
row_diff <- function(i, j, d) {
  d1 <- d %>% select(-id, -location)
  x <- d1 %>% slice(i) %>% unlist()
  y <- d1 %>% slice(j) %>% unlist()
  count_diff(x, y)
}
row_diff(3, 4, bridges)
```

 

That's what I said.

Extra: is that right, though? Let's print out those rows and count:

```{r pittsburgh-11 }
bridges %>% slice(c(3, 4)) 
```

 

Out of the ones we're counting, I see differences in purpose, length,
lanes, material, span and type. Six.

I actually think the `unlist` is not needed:

```{r pittsburgh-12 }
row_diff2 <- function(i, j, d) {
  d1 <- d %>% select(-id, -location)
  x <- d1 %>% slice(i)
  y <- d1 %>% slice(j)
  count_diff(x, y)
}
row_diff2(3, 4, bridges)
```

 

Here, `x` and `y` are one-row data frames, but R (via
`count_diff`) is sufficiently flexible to be able to cope with
these rather than vectors (it checks "corresponding elements" of
`x` and `y` for differentness). To my mind, though,
having the `unlist` in is clearer, since it makes it
unambiguous that `x` and `y` are vectors, and we know
that `count_diff` works for vectors since that's what we
tested it with.

$\blacksquare$

(f) Create a matrix or data frame of pairwise dissimilarities
between each pair of bridges (using only the ones with no missing
values). Use loops, or `crossing` and `rowwise`, as
you prefer. Display the first six rows of
your matrix (using `head`) or the first few rows of your data
frame. (The whole thing is big, so don't display it all.)

Solution


First thing, either way, is to find out how many bridges we have left:
```{r pittsburgh-13 }
bridges
```

     

70. So the loops (and the `crossing`) will go up to 70. Loops first:

```{r pittsburgh-14, cache=T}
m <- matrix(-1, 70, 70)
for (i in 1:70) {
  for (j in 1:70) {
    m[i, j] <- row_diff(i, j, bridges)
  }
}
```

 

Even just the top six rows (of all 70 columns) takes up a lot of
space. The grader would only check that it looks about right:

```{r pittsburgh-15, size="scriptsize"}
head(m)
```

 

A cursory glance at this shows that the bridges in the small-numbered
rows of the data frame are similar to each other and different from
the others. This suggests that these small-numbered bridges will end
up in the same cluster (later).

The `tidyverse` way is really similar in conception. First use
`crossing` to create all combinations of `i` and `j`:

```{r pittsburgh-16 }
mm <- crossing(i = 1:70, j = 1:70)
```

 

and then use `rowwise`:

```{r pittsburgh-17, cache=T}
mm %>% 
  rowwise() %>% 
  mutate(diff = row_diff(i, j, bridges)) -> mm
mm
```

If you forget the `rowwise`, the answers (in `diff`) will all be the same, which should make you suspicious.

This is long format, though, so we need to `pivot_wider` the
`j` column to get a square array of dissimilarities:

```{r pittsburgh-18 }
mm %>% pivot_wider(names_from=j, values_from=diff) -> mm
mm
```

 

This shows what we found before, that bridges 3 and 4 differ on 6
variables. 

$\blacksquare$

(g) Turn your matrix or data frame into a `dist`
object. (If you couldn't create a matrix or data frame of
dissimilarities, read them in from
[link](https://raw.githubusercontent.com/nxskok/datafiles/master/mm.csv).) Do not
display your distance object.

Solution


The only tricky thing is that the data frame that I called
`mm` has an extra column called `i` that needs to be
removed first. This also applies if you need to read it in from
the file. Thus:
```{r pittsburgh-19 }
d1 <- as.dist(m)
```

     

or 

```{r pittsburgh-20 }
mm %>% select(-i) %>% as.dist() -> d2
```

 

or, if you got stuck,

```{r pittsburgh-21 }
my_url <- "http://ritsokiguess.site/datafiles/mm.csv"
mmm <- read_csv(my_url)
mmm
mmm %>% select(-i) %>% as.dist() -> d3
d3
```

 
$\blacksquare$

(h) Run a cluster analysis using Ward's method, and display a
dendrogram. The labels for the bridges (rows of the data frame) may
come out too big; experiment with a `cex` less than 1 on the
plot so that you can see them.

Solution


Home stretch now. I found that `cex=0.3` was good for me, though I had to enlarge the graph to see the bridge numbers:
```{r pittsburgh-22 }
bridges.1 <- hclust(d1, method = "ward.D")
plot(bridges.1, cex = 0.3)
```

     

`cex` stands for "character expansion". This is one of the
myriad of things you could adjust on the old base graphics (of which
this is an example). If you wanted to make text *bigger*, you'd
set `cex` to a value bigger than 1.

$\blacksquare$

(i) How many clusters do you think is reasonable for these
data? Draw them on your plot.

Solution


I think you could go with any number from about 3 to something
like 15, but my choice is 5. What you want is for the bridges
within a cluster to be similar and bridges in different clusters
to be different, however you judge that.
```{r pittsburgh-23 }
plot(bridges.1, cex = 0.3)
rect.hclust(bridges.1, 5)
```

     

Whatever number of clusters you go with, draw a corresponding number
of red rectangles.

Note that the low-numbered bridges are all in my first cluster, as I
suspected they would be.

$\blacksquare$

(j) Pick three bridges in the same one of your clusters (it
doesn't matter which three bridges or which cluster). Display the
data for these bridges. Does it make sense that these three bridges
ended up in the same cluster? Explain briefly.

Solution


What I want you to do is to display the data for your chosen three
bridges and make the case that they are "similar". I'm picking
41, 42 and 48 from my third cluster. On yours, scroll right to see the other variables.

```{r pittsburgh-24 }
bridges %>% slice(c(41, 42, 48))
```

     

These bridges are identical on everything except location (which we
weren't considering anyway). So it makes perfect sense that they would
be in the same cluster.

You might not have been so lucky, for example, 10, 12 and 19, which
got joined together further up:

```{r pittsburgh-25 }
bridges %>% slice(c(10, 12, 19))
```

     

These differ on 3 or 4 variables and are the same on all the others,
so you can certainly say that they are more alike than different.

Extra: to get a feel for how different they might be, let's compare
three bridges in different clusters:

```{r pittsburgh-26 }
bridges %>% slice(c(8, 24, 52))
```

 

These are not as different as I was expecting, but they are indeed
different on more of the variables.

It would be interesting to plot these bridges on a map of Pittsburgh,
colour-coded by which cluster they are in. This might give us some
insight about how bridges are alike or different. 

I also remark that the discriminant analysis idea, using the clusters
as known groups, would not work here because we don't have any
quantitative variables to use for the discriminant analysis. 

The most interesting way I can think of is to cross-classify the
bridges by cluster and values of the other variables. These would be
complicated multi-way tables, though, which makes me wonder whether a
"classification tree" like `rpart` would be worth thinking
about. 

I am curious enough to have a crack at this. First we need a data set
with the clusters in it. I'm going with my 5 clusters:

```{r pittsburgh-27 }
bridges.rpart <- bridges %>% mutate(cluster = cutree(bridges.1, 5))
```

 

and then

```{r pittsburgh-28 }
library(rpart)
bridges.tree <- rpart(factor(cluster) ~ river + erected + purpose + length + lanes + clear_g +
  t_d + material + span + rel_l + type, data = bridges.rpart, method = "class")
print(bridges.tree)
```

 

This takes some making sense of, so let's grab a couple of bridges to
predict the cluster of:

```{r pittsburgh-29 }
bridges.rpart %>% slice(c(20, 29)) 
```

 

Let's start with bridge 20 (with ID E42). The line in the printed
output marked 1 and "root" says that there are 70 bridges
altogether, and the best guess (if you know nothing) is to guess
cluster 3, which would guess 47 of the bridges wrong. The five numbers
in the brackets are the proportions of bridges in each cluster at this
point. But we *do* know more about bridge E42. We go to number 2,
which says to look at the `span`. For this bridge it is
`LONG`, so we go down to number 3. Under 3 are 6 and 7, which
say to look at `type`. Bridge E42 is of type `SIMPLE-T`,
so we go to 6. There is nothing under this (lower down in the tree),
and the line ends with a `*`, so we are ready to guess the
cluster. There are 18 bridges with this `span` and one of these
`type`s, and they are *all* in cluster 3, no
errors. Cluster 3 contains long-span bridges of one of those types.

Bridge 29, with ID E51, now. The first thing to look at is
`span` again; this one is `medium`, so we are at
2. Under 2 is 4 and 5; we are invited to look at `material`,
which is `STEEL`, number 5. We have another thing to look at,
10 and 11, which is `river`; in this case, `river` is
`M`, number 10. We are now at the end, guessing cluster 2
(which is also correct); there are 13 bridges of this `span`,
`material` and `river` and only 3 of them were in some
cluster other than 2.

By looking at the tree output, we can describe what makes a bridge be
predicted to land up in each cluster:



* Span is medium or short, material is iron or wood. (This
suggests old bridges.)

* Span is medium or short, material is steel, river is M.

* Span is long, type is arch, cantilever or simple-truss

* Span is medium or short, material is steel, river is A.

* Span is long, type is continuous-truss or suspension.


This story is telling us that the clusters are determined by only a
few of our variables. `span` seems to be the most important,
then either type (if the span is long) or material and maybe river
(otherwise). 

This is an example of a "classification tree" which is a nice
easy-to-follow version of logistic regression.

$\blacksquare$




