##  Location, species and disease in plants


 The table below is a "contingency table", showing
frequencies of diseased and undiseased plants of two different species
in two different locations:


```

Species     Disease present         Disease absent
Location X Location Y  Location X Location Y
A            44         12          38        10
B            28         22          20        18

```


The data were saved as
[link](http://www.utsc.utoronto.ca/~butler/c32/disease.txt). In that
file, the columns are coded by two letters: a `p` or an
`a` to denote presence or absence of disease, and an `x`
or a `y` to denote location X or Y. The data are separated by
multiple spaces and aligned with the variable names. 



(a) Read in and display the data.


Solution


`read_table` again. You know this because, when you looked
at the data file, which of course you did (didn't you?), you saw
that the data values were aligned by columns with multiple spaces
between them:
```{r }
my_url="http://www.utsc.utoronto.ca/~butler/c32/disease.txt"
tbl=read_table(my_url)
tbl
```

 

I was thinking ahead, since I'll be wanting to have one of my columns
called `disease`, so I'm *not* calling the data frame
`disease`. 

You'll also have noticed that I simplified the data frame that I had
you read in, because the original contingency table I showed you has
*two* header rows, and we have to have *one* header row. So
I mixed up the information in the two header rows into one.



(b)??part:nottidy?? Explain briefly how these data are not "tidy".


Solution


The simple answer is that there are 8 frequencies, that each ought
to be in a row by themselves. Or, if you like, there are three
variables, Species, Disease status and Location, and each of
*those* should be in a *column* of its own. 
Either one
of these ideas, or something like it, is good. I need you to
demonstrate that you know something about "tidy data" in this context.



(c) Use a suitable `tidyr` tool to get all the things
that are the same into a single column. (You'll need to make up a
temporary name for the other new column that you create.) Show your
result. 


Solution


`gather` is the tool. All the columns apart from
`Species` contain frequencies, so that's what's "the same". 
They are frequencies in disease-location combinations, so
I'll call the column of "what's different" `disloc`. Feel
free to call it `temp` for now if you prefer:
```{r }
(tbl %>% gather(disloc,frequency,px:ay) -> tbl.2)
```

 

This also works ("gather together everything but `Species`"):

```{r }
(tbl %>% gather(disloc,frequency,-Species) -> tbl.2)
```

 



(d) Explain briefly how the data frame you just created is
still not "tidy" yet.


Solution


The column I called `disloc` actually contains *two*
variables, disease and location, which need to be split up. A
check on this is that we 
have two columns (not including the frequencies), but back in
(??part:nottidy??) we found *three* variables, so there
ought to be three non-frequency columns.



(e) Use one more `tidyr` tool to make these data tidy,
and show your result.


Solution


This means splitting up `disloc` into two separate columns,
splitting after the first character, thus:
```{r }
(tbl.2 %>% separate(disloc,c("disease","location"),1) -> tbl.3)
```

 

This is now tidy: eight frequencies in rows, and three non-frequency
columns. (Go back and look at your answer to part (??part:nottidy??)
and note that the issues you found there have all been resolved now.)



(f) Let's see if we can re-construct the original contingency
table (or something equivalent to it). Use the function
`xtabs`. This requires first a model formula with the frequency
variable on the left of the squiggle, and the other variables
separated by plus signs on the right. Second it requires a data
frame, with `data=`. Feed
your data frame from the previous part into `xtabs`. Save the
result in a variable and display the result.


Solution


```{r }
tbl.4=xtabs(frequency~Species+disease+location,data=tbl.3)
tbl.4
```

 

This shows a pair of contingency tables, one each for each of the two
locations (in general, the variable you put last on the right side of
the model formula). You can check that everything corresponds with the
original data layout at the beginning of the question, possibly with
some things rearranged (but with the same frequencies in the same
places). 



(g) Take the output from the last part and feed it into the
function `ftable`. How has the output been changed? Which do
you like better? Explain briefly.


Solution


This:
```{r }
ftable(tbl.4)  
```

 

This is the same output, but shown more compactly. (Rather like a
vertical version of the original data, in fact.) I like
`ftable` better because it displays the data in the smallest
amount of space, though I'm fine if you prefer the `xtabs`
output because it spreads things out more. This is a matter of
taste. Pick one and tell me why you prefer it, and I'm good.

That's the end of what you had to do, but I thought I would do some
modelling and try to find out what's associated with disease. The
appropriate modelling with frequencies is called "log-linear modelling", 
and it assumes that the log of the frequencies has a
linear relationship with the effects of the other variables. This is
not quite as simple as the log transformations we had before, because
bigger frequencies are going to be more variable, so we fit a
generalized linear model with a Poisson-distributed response and log
link. (It's better if you know what that means, but you ought to be
able to follow the logic if you don't.)

First, fit a model predicting frequency from everything, including all
the interactions. (The reason for doing it this way will become clear later):

```{r }
model.1=glm(frequency~Species*location*disease,data=tbl.3,family="poisson")
drop1(model.1,test="Chisq")
```

 

The residuals are all zero because this model fits perfectly. The
problem is that it is very complicated, so it offers no insight. So
what we do is to look at the highest-order interaction
`Species:location:disease` and see whether it is
significant. It is not, so we can remove it. This is reminiscent of
variable selection in regression, where we pull the least significant
thing out of the model in turn until we can go no further. But here,
we have additional things to think about: we have to get rid of all
the three-way interactions before we can tackle the two-way ones, and
all the two-way ones before we can tackle the main effects. There is a
so-called "nested" structure happening here that says you don't look
at, say, `Species`, until you have removed *all* the
higher-order interactions involving `Species`. Not clear yet?
Don't fret. `drop1` allows you to assess what is currently up
for grabs (here, only the three-way interaction, which is not
significant, so out it comes).

Let's get rid of that three-way interaction. This is another use for
`update` that we've seen in connection with multiple regression
(to make small changes to a big model):

```{r }
model.2=update(model.1,.~.-Species:location:disease)
drop1(model.2,test="Chisq")
```

 

Notice how `update` saved us having to write the whole model
out again.

Now the three two-way interactions are up for grabs:
`Species:location`, `Species:disease` and
`location:disease`. The last of these is the least significant,
so out it comes. I did some copying and pasting, but I had to remember
which model I was working with and what I was removing:

```{r }
model.3=update(model.2,.~.-location:disease)
drop1(model.3,test="Chisq")  
```

 

`Species:disease` comes out, but it looks as if
`Species:location` will have to stay:

```{r }
model.4=update(model.3,.~.-Species:disease)
drop1(model.4,test="Chisq")  
```

 

`Species:location` indeed stays. That means that anything
"contained in" it also has to stay, regardless of its main
effect. So the only candidate for removal now is `disease`: not
significant, out it comes:

```{r }
model.5=update(model.4,.~.-disease)
drop1(model.5,test="Chisq")  
```

 

And now we have to stop.

What does this final model mean? Well, frequency depends significantly
on the `Species:location` combination, but not on anything
else. To see how, we make a contingency table of species by location
(totalling up over disease status, since that is not significant):

```{r }
xtabs(frequency~Species+location,data=tbl.3)  
```

 

Most of the species A's are at location X, but the species B's are
about evenly divided between the two locations. Or, if you prefer
(equally good): location X has mostly species A, while location Y has
mostly species B. You can condition on either variable and compare the
conditional distribution of the other one.

Now, this is rather interesting, because this began as a study of
disease, but disease has completely disappeared from our final model!
That means that nothing in our final model has any relationship with
disease. Indeed, if you check the original table, you'll find that
disease is present slightly more than it's absent, for all
combinations of species and location. That is, neither species nor
location has any particular association with (effect on) disease,
since disease prevalence doesn't change appreciably if you change
location, species or the combination of them.

The way an association with disease would show up is if a
`disease:`something interaction had been significant and had
stayed in the model, that something would have been associated with
disease. For example, if the `disease:Species` table had looked
like this:

```{r }
disease=c("a","a","p","p")
Species=c("A","B","A","B")
frequency=c(10,50,30,30)
xx=data.frame(disease,Species,frequency)
xtabs(frequency~disease+Species)
```

 

For species A, disease is present 75\% of the time, but for species B
it's present less than 40\% of the time. So in this one there ought to be a
significant association between disease and species:

```{r }
xx.1=glm(frequency~disease*Species,data=xx,family="poisson")
drop1(xx.1,test="Chisq")
```

 

And so there is. Nothing can come out of the model. (This is the same
kind of test as a chi-squared test for association, if you know about
that.  The log-linear model is a multi-variable generalization of that.)




