---
title: "play"
output: html_notebook
---

playing with `pivot_longer`:

```{r play-1}
library(tidyverse)
```

a fake family or two, copied from vignette:

```{r play-2}
family <- tribble(
  ~family,  ~dob_child1,  ~dob_child2, ~gender_child1, ~gender_child2,
       1L, "1998-11-26", "2000-01-29",             1L,             2L,
       2L, "1996-06-22",           NA,             2L,             NA,
       3L, "2002-07-11", "2004-04-05",             2L,             2L,
       4L, "2004-10-10", "2009-08-27",             1L,             1L,
       5L, "2000-12-05", "2005-02-28",             2L,             1L,
)
family
```

what I think of is `pivot_longer` with two `names_to`:

```{r play-3}
family %>% 
  mutate(across(starts_with("dob"), \(dob) as.Date(dob))) %>% 
  pivot_longer(-family, names_to=c("variable", "child"), names_sep="_", values_to="value")
```

only it isn't. It would work if all the variables were of the same type.

Here's what Hadley does:

```{r play-4}
family %>% 
  pivot_longer(
    -family, 
    names_to = c(".value", "child"), 
    names_sep = "_", 
    values_drop_na = TRUE
  )
```

haven't grokked ".value" yet. Ah: when I am combining variables of different types, this is how it knows to get the column type right.

Is this how I do my "making it too long" thing, and I have to make it wider? Let's go back to the Toronto weather data:

```{r play-5}
my_url <- 
  "http://ritsokiguess.site/STAC32/toronto_weather.csv"
weather <- read_csv(my_url)
```

```{r play-6}
weather
```

let's do the pivot_wider first this time:

```{r play-7}
weather %>% 
  pivot_wider(names_from=element, values_from = d01:d31)
```

then we make it longer with two names_to, one of which is .value?

```{r play-8}
weather %>% 
  pivot_wider(names_from=element, values_from = d01:d31) %>% 
  pivot_longer(starts_with("d"), names_to=c("day", ".value"), names_sep="_", values_to="temperature")

```

Ooh!



Here's a simpler one (maybe):

```{r play-9}
who
```

```{r play-10}
who %>% pivot_longer(
  cols = new_sp_m014:newrel_f65,
  names_to = c("diagnosis", "gender", "age"), 
  names_pattern = "new_?(.*)_(.)(.*)",
  values_to = "count",
  values_drop_na = T
)
```

