##  Productivity and research-and-development


 An economist compiled data on productivity improvements for
a sample of companies that produce electronic equipment. The companies
were classified according to their expenditures on research and
development in the last three years (classified as low, moderate,
high). Productivity improvement is measured on a 0--100 scale, with
most values being between 0 and 10. The economist also thought that
last year's productivity improvement would be relevant, and so
recorded this as well.

The data set can be found at
[link](http://www.utsc.utoronto.ca/~butler/d29/productivity.txt). 

Feel free to use "R\&D" as an abbreviation for 
"research and development" in your answers below.



(a) Read in and display at least some of the data.

Solution


The data values are separated by one space, so let's use
`read_delim` and display whatever displays (probably the
first ten lines):
```{r }
my_url="http://www.utsc.utoronto.ca/~butler/d29/productivity.txt"
productivity=read_delim(my_url," ")
productivity
```

   

There is a column classifying expenditure along with numerical values
for productivity improvement for this year and last year (labelled
`improvement` and `last`), so it looks as if we are
good.

You are free to give the data frame a shorter name to make it easier
to keep typing it!

Extra: those expenditure levels (that are currently text) will get
sorted into alphabetical order unless we stop them. They are actually
in the right order in the data file, so maybe we should turn them into
a factor (with the levels in the right order) now:

```{r }
productivity %>%
mutate(expenditure=fct_inorder(expenditure)) -> productivity
```

 

If you don't do this, your answers below will be different from
mine. Probably not in any kind of consequential way, but different
nonetheless.


(b) Plot productivity improvement for this year against last
year, identifying the points by the level of
research-and-development expenditure. On your plot, add regression
lines for each expenditure group.

Solution


Two quantitative and one categorical variable, so plot the
categorical variable using colour (or `shape` etc., if you
know about that, but colour is the most obvious thing):
```{r }
ggplot(productivity,aes(x=last,y=improvement,colour=expenditure))+
geom_point()+geom_smooth(method="lm",se=F)
```

     

The `se=F` is optional; no problem if you don't include it. (If
you miss it out, you'll get those CI-of-mean-response grey envelopes
around each line.)


(c) <a name="part:graph-comment">*</a> 
Look at your graph of the previous part. Do you think that (i)
knowing last year's productivity improvement tells you something
about this year's, (ii) the level of expenditure on
research-and-development has an impact on this year's productivity
improvement, (iii) there is an interaction effect between R\&D
expenditure and last year's productivity increase on this year's
productivity increase? In each case, explain briefly.

Solution


Taking the three things in turn:
(i) knowing about last year's productivity increase definitely
helps, because the three trends definitely go up (extra: and are
pretty much linear).
(ii) knowing about the level of expenditure helps because the
coloured trends are in different places (low on the
left and high on the right, whatever colours they are for you).
(iii) I am guessing there is no interaction because the three
lines look more or less parallel (a judgement call: if you think
that there *will* be an interaction because you think the
lines are definitely not "approximately parallel" and therefore
there is an interaction, that's OK too).


(d) Fit an analysis of covariance model without interaction. Run
the results through `drop1` with `test="F"` (the
`F` must be in quotes). What do you conclude?

Solution


This looks exactly like a regression with a categorical variable,
and is. Just the two main effects, thus:
```{r }
improvement.1=lm(improvement~last+expenditure,data=productivity)
drop1(improvement.1,test="F")
```

     

Both the P-values are small, so there is strong evidence of an effect
on this year's productivity increase of last year's productivity
increase and the level of R\&D expenditure. 


(e) Now look at the `summary` of your analysis of
covariance model. Explain briefly how each of the *last three*
numbers in the `Estimate` column are consistent with your graph.

Solution


To begin:
```{r }
summary(improvement.1)
```

     

The last three numbers in the Estimate column are slopes for the
quantitative variable `last` and changes in intercept for the
levels of the categorical variable `expenditure`. Specifically:



* the slope for `last` is positive, so if last year's
productivity increase is higher, so will this year's. This matches
the upward trends within each `expenditure` group on the
graph.

* `expenditure` is categorical, so everything is measured
relative to the baseline. Mine is `low`, but yours is probably `high`. 

* My two estimates for `expenditure` `high` and
`moderate` are both negative, so when expenditure on R\&D is
high or moderate, the productivity increase will be lower this year
(for the same productivity increase last year). Going back to the
graph, if you extend the lines for `moderate` and `high` to
a middling `last` value of something like 10, the
productivity increase this year will be higher when
`expenditure` is low or moderate then when it is high, which
is consistent with those slopes both being positive. (You don't need
to go to this detail, but you should say something convincing from
the graph about how this year's productivity increase is
*higher* if R\&D expenditure is low or moderate compared to
high.)


Extra: the graph also indicates that the higher the
`expenditure` on R\&D is, the more likely last year's
productivity improvement will be higher also. So in practice we are
not likely to be comparing actual companies with different
`expenditure` but the same `last`. Nonetheless, what I
said is what those positive coefficients for `expenditure`
actually mean. Be careful not to say that `low` is lowest
because it's on the left; it's whether it's higher or lower than the
others *at the same value of `last`* that matters.

Extra extra: I asked you to do the test from `drop1` because
there are three levels of `expenditure` (rather than just two),
and the `summary` output only gives you a comparison with the
baseline level, rather than comparing all three
levels. I prefer to do the right test (using `drop1`) first,
and then use `summary` to interpret what I have, or at least
the parts of it that are significant.

Extra extra extra: you would (or at least I would) expect a larger
productivity increase to go with a *larger* expenditure on R\&D,
but that's not how it worked out. This is one of those cases where all
else isn't really equal.


(f) Add an interaction between `last` and
`expenditure` to your analysis of covariance model. Test it
for significance using `drop1`. In the light of what you
have seen, does this surprise you? Explain briefly why you are or
are not surprised.

Solution


I like `update` for this (writing out the whole model is an
alternative):
```{r }
improvement.2=update(improvement.1,.~.+last:expenditure)
drop1(improvement.2,test="F")
```

     

Yes, I too forgot the `test="F"` the first time.

The interaction term is actually significant, at least at
$\alpha=0.05$. This is a surprise to me, because I thought those lines
on the graph were pretty close to parallel, so I wasn't expecting to
see a significant interaction. (What I want from you here is to look
back at your answer to (iii) in part (<a href="#part:graph-comment">here</a>), and
to say how this small P-value is consistent or inconsistent with
whatever you said there. If you said that the lines weren't parallel
enough, then this significant interaction should *not* be a
surprise to you.)

That's all I'm after. I don't need you to speculate on *why* the
test came out significant. I will, in a moment, but you don't need to.

Extra: I didn't expect the interaction to come out significant, but
the P-value is not nearly so small as the ones we had before. I guess
the explanation for this is that the data cluster pretty tightly about
the lines in the graph, so that even a small difference in slopes can
be signficant. You might argue that my red line is not *quite*
as steep as the others, but the difference does look very small.

Let's take a look at the `summary` for my interaction model:

```{r }
summary(improvement.2)
```

 

The last two Estimates, the ones beginning with `last:`, came
from the interaction. Once again `low` is my baseline. These
say how the *slopes* for `last` for the other groups
differ from the slopes for `last` for `low`
expenditure. (Yes, this is confusing. Read it again until you get it.)

The biggest-size difference in slopes, about $0.40$, is between low
expenditure and high expenditure. This says that the line for high
expenditure is this much more steep than the line for low
expenditure. That's where the non-parallelism is on the graph, such as
it is. A small but significant difference in slopes. You can tell that
the effect is small by looking in the Estimate column; those changes
in slopes, $0.40$ and $0.32$, are the smallest things in size out of
everything in that column.

Having seen this, you now realize that I did this question wrong (or,
at least, I led you through it wrong). The right way to do this would
have been to fit the interaction model *first*, see that it is
significant, and then done some predictions to assess the effects of
things: 

```{r }
lasts=c(7,9,11)
expenditures=c("low","moderate","high")
new=crossing(last=lasts,expenditure=expenditures)
new
```

 

(oh, how I wish I'd given those variables shorter names),
`r tufte::margin_note("The expenditure levels have come out in alphabetical order again.")` and then:

```{r }
p=predict(improvement.2,new)
cbind(new,p)
```

 

The effect is more or less as we described it before: as `last`
goes up (for fixed `expenditure`), the predicted productivity
improvements for this year go up, and as `expenditure` level
goes up, the predictions go *down*. But the rates at which they
go up or down are different, which is the significant interaction
coming into play. Having said that, they are not very different, so I
cheated and pretended the interaction was not significant (or that I
was using $\alpha=0.01$), so that you would have something easier to
interpret. Qualitatively, the story is the same either way, because
the sizes of the interaction terms are small compared to the
others. So, even though I cheated, you ended up with more or less the
same conclusions doing it the way I asked you to do it, or doing it
the way I just did it.



