
## Data file operations

 In this question, we read a file from the web and do some
descriptive statistics and a graph. This is very like what you will be
doing on future assignments, so it's good to practice it now.

Take a look at the data file at
[https://www.utsc.utoronto.ca/~butler/c32/jumping.txt](https://www.utsc.utoronto.ca/~butler/c32/jumping.txt). These are
measurements on 30 rats that were randomly made to do different
amounts of jumping by group (we'll see the details later in the
course). The control group did no jumping, and the other groups did
"low jumping" and "high jumping". The first column says which
jumping group each rat was in, and the second is the rat's bone
density (the experimenters' supposition was that more jumping should
go with higher bone density).



(a) What are the two columns of data separated by? (The fancy word
is "delimited"). 

Solution


Exactly one space. This is true all the way down, as you can check.


(b) Make a new R Notebook. Leave the first four lines, but get rid
of the rest of the template document. Start with a code chunk
containing `library(tidyverse)`. Run it.

Solution


You will get either the same message as before or nothing. (I got
nothing because I had already loaded the `tidyverse` in
this session.)


(c) Put the URL of the data file in a variable called
`my_url`. Then use `read_delim` to read in the
file. (See solutions for how.) `read_delim` reads data files
where the data values are always separated by the same single
character, here a space. Save the data frame in a variable `rats`.

Solution


Like this:
```{r }
library(tidyverse)
my_url="https://www.utsc.utoronto.ca/~butler/c32/jumping.txt"
rats=read_delim(my_url," ")
```     

The second thing in `read_delim` is the thing that separates
the data values. Often when you use `read_delim` it'll be a space.


(d) Take a look at your data frame, by making a new code chunk and
putting the data frame's name in it (as we did with
`mtcars`). 

Solution


```{r }
rats
```   

There are 30 rows and two columns, as there should be.


(e) Find the mean bone density for rats that did each amount of jumping.

Solution


This is something you'll see a lot: `group_by` followed by
`summarize`. Reminder: to get that funny thing with the
percent signs (called the "pipe symbol"), type control-shift-M (or
equivalent on a Mac):
```{r }
rats %>% group_by(group) %>%
summarize(m=mean(density))
```     

The mean bone density is clearly highest for the high jumping group,
and not much different between the low-jumping and control groups.


(f) Make a boxplot of bone density for each jumping group.

Solution


On a boxplot, the groups go across and the values go up and down,
so the right syntax is this:
```{r }
ggplot(rats,aes(x=group, y=density))+geom_boxplot()
```   

Given the amount of variability, the control and low-jump groups are
very similar (with the control group having a couple of outliers), but
the high-jump group seems to have a consistently higher bone density
than the others.

This is more or less in line with what the experimenters were
guessing, but it seems that it has to be high jumping to make a
difference. 

You might recognize that this is the kind of data where we would use
analysis of variance, which we will do later on in the course: we are
comparing several (here three) groups.



