##  Effect of drug on rat weight


 Box (1950) gives data on the weights of three groups of
rats.^[This is the same Box as Box-Cox and the M test.] One group was given thyroxin in their drinking water, one group
thiouracil, and the third group was a control. (This description comes
from Christensen (2001).)^[References: Box, George EP, Problems in the analysis of growth and wear curves, Biometrics vol. 6, 362-369, 1950; Christensen R, Advanced Linear Modeling, 2nd edition, Springer, 2001. Yes, that's the Box-Cox Box.]
Weights are measured in
grams at weekly intervals (over a 4-week period, so that each rat is
measured 5 times). The data are in
[link](http://ritsokiguess.site/datafiles/ratweight.csv) as a
`.csv` file.



(a) Read in the data and check that you have a column of
`drug` and five columns of rat weights at different times.
 
Solution


A `.csv` file, so `read_csv`. (I typed the data from
Christensen (2001) into a spreadsheet.)
```{r ratweight-1 }
my_url <- "http://ritsokiguess.site/datafiles/ratweight.csv"
weights <- read_csv(my_url)
weights
```

   

There are 27 rats altogether, each measured five times (labelled time
0 through 4). The rats are also labelled `rat` (the first column), which will be useful later.
 
$\blacksquare$ 

(b) Why would it be *wrong* to use something like
`pivot_longer` to create one column of weights, and separate
columns of drug and time, and then to run a two-way ANOVA? Explain briefly.
 
Solution


Such a solution would assume that we have measurements on
*different* rats, one for each drug-time combination. But we
have sets of five measurements all on the *same* rat: that is
to say, we have repeated measures, and the proper analysis will
take that into account.
 
$\blacksquare$

(c) Create a suitable response variable and fit a suitable
`lm` as the first step of the repeated-measures analysis.
 
Solution


The response variable is the same idea as for any MANOVA: just
glue the columns together:
```{r ratweight-2 }
response <- with(weights, cbind(Time0, Time1, Time2, Time3, Time4))
weights.1 <- lm(response ~ drug, data = weights)
```

     

Now, we *don't* look at `weights.1`, but we *do* use
it as input to `Manova` in a moment.
 
$\blacksquare$

(d) Load the package `car` and run a suitable
`Manova`. To do this, you will need to set up the right thing
for `idata` and `idesign`.
 
Solution


Something like this:
```{r ratweight-3 }
times <- colnames(response)
times.df <- data.frame(times=factor(times))
weights.2 <- Manova(weights.1, idata = times.df, idesign = ~times)
```

     

The thought process is that the columns of the response
(`Time.0` through `Time.4`) are all times. This is the
"within-subject design" part of it: within a rat, the different
response values are at different times. That's the only part of it
that is within subjects. The different drugs are a
"between-subjects" factor: each rat only gets one of the
drugs.^[Things would be a lot more complicated if each rat got a different drug at a different time! But the rats each got one drug *once*, at the beginning, and the issue was the effect of that drug on all the growth that followed.]
 
$\blacksquare$

(e) Take a look at all the output from the MANOVA. Is there a
significant interaction? What does its significance (or lack
thereof) mean?
 
Solution

Look at the `summary`, which is rather long:

```{r ratweight-4 }
summary(weights.2)
```

Start near the bottom with Mauchly's test. This is strongly significant (for the interaction, which is our focus here) and means that sphericity fails and the P-value for the interaction in the univariate test is not to be trusted (it is much too small). Look instead at the Huynh-Feldt adjusted P-value at the very bottom, $1.76 \times 10^{-6}$. This is strongly significant still, but it is a *billion* times bigger than the one in the univariate table! For comparison, the test for interaction in the multivariate analysis has a P-value of 0.0007 or less, depending on which of the four tests you look at (this time, they are not all the same). As usual, the multivariate tests have bigger P-values than the appropriately adjusted univariate tests, but the P-values are all pointing in the same direction.

The significant interaction means that the effect of time on growth is different for
the different drugs: that is, the effect of drug is over the whole
time profile, not just something like 
"a rat on Thyroxin is on average 10 grams heavier than a control rat, over all times".

Since the interaction is significant, that's where we stop, as far as
interpretation is concerned.
 
$\blacksquare$

(f) We are going to draw an interaction plot in a moment. To
set that up, use `pivot_longer` as in the lecture notes to create
one column of weights and a second column of times. (You don't
need to do the `separate` thing that I did in class, though
if you want to try it, go ahead.)
 
Solution

Like this:

```{r ratweight-5}
weights %>% 
  pivot_longer(starts_with("Time"), names_to="time", values_to="weight") -> weights.long
weights.long
```



     

My data frame was called `weights`, so I was OK with having a
variable called `weight`. Watch out for that if you call the
data frame `weight`, though.

Since the piece of the time we want is the number,
`parse_number` (from `readr`, part of the
`tidyverse`) should also work:

```{r ratweight-6 }
weights %>% 
  pivot_longer(starts_with("Time"), names_to="timex", values_to="weight") %>% 
  mutate(time = parse_number(timex)) -> weights2.long
weights2.long %>% sample_n(20)
```

 

I decided to show you a random collection of rows, so that you can see
that `parse_number` worked for various different times. 
 
$\blacksquare$

(g) Obtain an interaction plot. Putting `time` as the
`x` will put time along the horizontal axis, which is the
way we're used to seeing such things. Begin by calculating the mean
`weight` for each `time`-`drug` combination.
 
Solution


`group_by`, `summarize` and `ggplot`, the
latter using the data frame that came out of the
`summarize`. The second factor `drug` goes as the
`colour` and `group` both, since `time` has
grabbed the `x` spot:
```{r ratweight-7 }
weights.long %>%
  group_by(time, drug) %>%
  summarize(mean.weight = mean(weight)) %>%
  ggplot(aes(x = time, y = mean.weight, colour = drug, group = drug)) +
  geom_point() + geom_line()
```

       
 
$\blacksquare$

(h) How does this plot show why the interaction was
significant? Explain briefly.
 
Solution


At the beginning, all the rats have the same average growth, but
from time 2 (or maybe even 1) or so, the rats on thiouracil
grew more slowly. The idea is not just that thiouracil has a
*constant* effect over all times, but that the *pattern*
of growth is different for the different drugs: whether or not
thiouracil inhibits growth, and, if so, by how much, depends on
what time point you are looking at.

Rats on thyroxin or the control drug grew at pretty much the same
rate over all times, so I wouldn't concern myself with any
differences there.

What I thought would be interesting is to plot the growth curves for
*all* the rats individually, colour-coded by which drug the rat
was on. This is the repeated-measures version of the ANOVA interaction
plot with the data on it, a so-called spaghetti plot. (We don't use the lines for the means, here,
instead using them for joining the measurements belonging to the same
subject.)

When I first used this data set, it didn't have a column identifying which rat was which, which made this plot awkward, but now it does (the column `rat`). So we can start directly from the dataframe I created above called `weights.long`:

```{r ratweight-8 }
weights.long
```

  
Each rat is identified by `rat``, which repeats 5 times,
once for each value of `time`:

```{r ratweight-9 }
weights.long %>% count(rat)
```

In the data frame `weights.long`, we plot
`time` ($x$) against `weight` ($y$), grouping the points
according to `rat` and colouring them according to
`drug`. 

```{r mantequilla,fig.width=6,fig.height=6}
library(ggplot2)
ggplot(weights.long, aes(time, weight, group = rat, colour = drug)) + geom_line()
```

As you see, "spaghetti plot" is a rather apt name for this kind of thing.

I like this plot because, unlike the interaction plot, which shows
only means, this gives a sense of variability as well. The blue and
red lines (thyroxin and control) are all intermingled and they go
straight up. So there is nothing to choose between these. The green
lines, though, start off mixed up with the red and blue ones but
finish up at the bottom: the *pattern* of growth of the
thiouracil rats is different from the others, which is why we had a
significant interaction between drug and time.

`drug` is categorical, so `ggplot`
uses a set of distinguishable colours to mark the levels. If our
colour had been a numerical variable, `ggplot` would have used
a range of colours like light blue to dark blue, with lighter being
higher, for example.

What, you want to see that? All right. This one is kind of silly, but
you see the point:

```{r alphington,fig.width=6,fig.height=6}
ggplot(wt, aes(time, weight, group = subject, colour = weight)) + geom_line()
```

 

The line segments get lighter as you go up the page.

Since we went to the trouble of making the "long" data frame, we can also run a repeated measures analysis using the
mixed-model idea (described more fully in the problem of the children
near the new airport):

```{r ratweight-10 }
wt.1 <- lmer(weight ~ drug * time + (1 | rat), data = weights.long)
drop1(wt.1, test = "Chisq")
```

 

The drug-by-time interaction is even more strongly significant than in
the profile analysis. (The output from `drop1` reminds us that
the only thing we should be thinking about now is that interaction.)
 
$\blacksquare$


