## Rescaling

 Suppose we have this vector of values:

```{r}
z <- c(10, 14, 11)
z
```

We want to scale these so that the smallest value is 0 and the largest is 1. We are going to be doing this a lot, so we are going to write a function that will work for any input.



(a) Using a copy of my `z`, work out `min(z)` and `max(z)`. What do they do? Explain (very) briefly.




(b) What do these lines of code do, using the same `z` that I had? Run them and see, and describe briefly what `s` contains.

```{r, eval=FALSE}
lo <- min(z)
hi <- max(z)
s <- (z - lo) / (hi - lo)
s
```






(c) Write a function called `rescale` that implements the calculation above, for any input vector called `x`. (Note that I changed the name.)



(d) Test your function on my `z`, and on another vector of your choosing. Explain briefly why the answer you get from your vector makes sense.



(e) What happens if your input to `rescale` is a vector of numbers all the same? Give an example. Rewrite your function to intercept this case and give a helpful error message.



(f) Make a dataframe (containing any numeric values), and in it create a new column containing the rescaled version of one of its columns, using your function. Show your result.



(g) We might want to rescale the input not to be between 0 and 1, but between two values `a` and `b` that we specify as input. If `a` and/or `b` are not given, we want to use the values 0 for `a` and 1 for `b`. Rewrite your function to rescale the input to be between `a` and `b` instead of 0 and 1. Hint: allow your function to produce values between 0 and 1 as before, and then note that if all the values in a vector `s` are between 0 and 1, then all the values in `a+(b-a)*s`
are between $a$ and $b$.



(h) Test your new function two or more times, on input where you know or can guess what the output is going to be. In each case, explain briefly why your output makes sense.





