

 A company is growing and would like to attract more
employees. The company would like to advertise that salaries there are
increasing. To do this, the company randomly samples 20 employees that
have been working there since January 2016, and for each of these
employees, records their salary in January 2016 and January 2017. The
data, with salaries in thousands of dollars, are in
[https://www.utsc.utoronto.ca/~butler/c32/salaryinc.txt](https://www.utsc.utoronto.ca/~butler/c32/salaryinc.txt). 



(a)[2] Read the data into R and demonstrate that you have two
salaries for each of 20 employees.

Solution


Looking at the file, we see that the values are separated by
exactly one space:
```{r }
my_url="https://www.utsc.utoronto.ca/~butler/c32/salaryinc.txt"
salaries=read_delim(my_url," ")
salaries
```     

There are 20 employees (rows), and two columns of salaries: for each
employee in the data set, their salary in January 2016 and in January
2017 (thus, two salaries for each employee).


(b)[2] To compare the salaries, explain briefly why a
matched-pairs test would be better than a two-sample test.

Solution


A matched-pairs test would be better because we have two
observations (salaries) for each subject (employee). A two-sample
test would be appropriate if we had two separate sets of
employees, one set with their salaries recorded in 2016 and the
other with their salaries recorded in 2017. That is not what we
have here.
You can go after this either way: why a matched-pairs approach is
appropriate, or why a two-sample approach is not (or a bit of both).


(c)[4] Make a suitable graph to assess the assumptions for a
matched-pairs $t$-test. What does your graph tell you?

Solution


This requires thought first before you do any coding (and this is
the reason for this one being four points). What has to be at
least approximately normally distributed is the set of
*differences*, salary at one time point minus the salary at
the other, for each employee. The individual salaries don't have
to be normally distributed at all. 
We don't have the differences here, so we have to calculate them
first. The smoothest way is to make a pipeline:
```{r }
salaries %>%
mutate(diff=jan2017-jan2016) %>%
ggplot(aes(sample=diff))+stat_qq()+stat_qq_line()
```     

A couple of coding notes: (i) you can take the differences 2016 minus
2017 if you like (then they will tend to be negative), (ii)
`ggplot` used in a pipeline like this does *not* have a
data frame first (the data frame used is the nameless output from the
`mutate`, with the differences in it). 

Also, there's no problem doing the `mutate`, saving that, and
then feeding the saved data frame into `ggplot`. If you find
that clearer, go for it.

As for what I see: I think those points get a bit far from the
line at the high and low ends: the high values are too high and the
low values are too low, which is to say that we have outliers at both
ends, or the distribution has long tails (either way of saying it is
good). 

The important conclusion here is whether these differences are normal
*enough* to trust a matched pairs $t$-test here. We have a sample
of size 20, so the central limit theorem will help us some, but you
can reasonably say that these tails are too long and that we should
not trust a matched-pairs $t$-test.

I actually wanted you to practice doing a matched-pairs $t$-test
anyway, hence my comment in the next part, but the result is probably
not *so* trustworthy.


(d)[3] Carry out a suitable matched-pairs $t$-test on these data. (If you
thought in the previous part that this was the wrong thing to do, do
it anyway for the purposes of this assignment.) What do you conclude?

Solution


The company is trying to prove that salaries are *increasing*
over time, so we need a one-sided alternative.
Following through the procedure, even though you may not trust it much:
```{r }
with(salaries, t.test(jan2016, jan2017, alternative="less", paired=T))
```     

You could also have the years the other way around, in which case the
alternative has to be the other way around as well:

```{r }
with(salaries, t.test(jan2017, jan2016, alternative="greater", paired=T))
``` 

Or, if you saved the data frame with the differences in it, do a
one-sample test on those, again making sure that you get the
`alternative` right. I didn't save it, so I'm calculating the
differences again:

```{r }
salaries %>%
mutate(diff=jan2017-jan2016) %>%
with(.,t.test(diff, mu=0, alternative="greater"))
``` 

Whichever way you do it, the P-value is the same $2.271 \times
10^{-9}$, which is a whole lot less than 0.05, so there is no doubt at
all that salaries are increasing.

(Your intuition ought to have expected something like this, because
everyone's 2017 salary appears to be greater than their 2016 salary.)

Extra: you might be feeling that we ought to be doing a matched-pairs
sign test, which you could do this way:

```{r }
library(smmr)
salaries %>%
mutate(diff=jan2017-jan2016) %>%
sign_test(diff,0)
``` 

and then take the "upper" P-value, which is in the same ballpark as
the one from the $t$-test. So the salaries really are increasing,
whether you believe the $t$-test or not. And note that \emph{every
single employee's salary increased}.

(Again, the "missing" data frame in `sign_test` is the nameless
one with the differences in it.)


(e)[2] The company would like to estimate  \emph{how
much} salaries are increasing, on average. Obtain some output that
will enable the company to assess this, and tell the CEO which piece
of the output they should look at.

Solution


A confidence interval. 95\% is fine. As before, we have to run
`t.test` again because we ran a one-sided test and a
confidence interval for us is two-sided:
```{r }
with(salaries, t.test(jan2017, jan2016, paired=T))
```     

Between about \$5,000 and about \$7,500. This is what to tell the CEO.

 

