##  Crimes in San Francisco -- the data


 The data in [link](http://ritsokiguess.site/datafiles/sfcrime.csv) is a huge dataset of crimes committed in San
Francisco between 2003 and 2015. The variables are:



* `Dates`: the date and time of the crime

* `Category`: the category of crime, eg. "larceny" or
"vandalism" (response).

* `Descript`: detailed description of crime.

* `DayOfWeek`: the day of the week of the crime.

* `PdDistrict`: the name of the San Francisco Police
Department district in which the crime took place.

* `Resolution`: how the crime was resolved

* `Address`: approximate street address of crime

* `X`: longitude

* `Y`: latitude


Our aim is to see whether the category of crime depends on the day of
the week and the district in which it occurred. However, there are a
lot of crime categories, so we will focus on the top four
"interesting" ones, which we will have to discover.



(a) Read in the data and verify that you have these columns and a
lot of rows. (The data may take a moment to read in. You will see why.)


Solution


```{r cache=T}
my_url="http://ritsokiguess.site/datafiles/sfcrime.csv"
sfcrime=read_csv(my_url)
sfcrime
```

     

Those columns indeed, and pushing a million rows! That's why it took
so long!

There are also 39 categories of crime, so we need to cut that down
some. There are only ten districts, however, so we should be able to
use that as is.
    


(b) How is the response variable here different to the one in
the question about steak preferences (and therefore why would
`multinom` from package `nnet` be the method of choice)?


Solution


Steak preferences have a natural order, while crime categories do
not. Since they are unordered, `multinom` is better than
`polr`. 
    


(c) Find out which crime categories there are, and 
arrange them in order of how many crimes there were in each
category. 


Solution


This can be the one you know, a group-by and summarize, followed
by `arrange` to sort:

```{r }
sfcrime %>% group_by(Category) %>%
summarize(count=n()) %>%
arrange(desc(count))
```

 

or this one does the same thing and saves a step:

```{r }
sfcrime %>% count(Category) %>%
arrange(desc(n))
```

 

For this one, do the `count` step first, to see what you
get. It produces a two-column data frame with the column of counts
called `n`. So now you know that the second line has to be
`arrange(desc(n))`, whereas before you tried `count`,
all you knew is that it was arrange-desc-something.

You need to sort in descending order so that the categories you want
to see actually do appear at the top.
    


(d) Which are the four most frequent "interesting" crime
categories, that is to say, not including "other offenses" and
"non-criminal"? Get them into a vector called
`my.crimes`. See if you can find a way of doing this  that
doesn't involve typing them in (for full marks).


Solution


The most frequent interesting ones are, in order, larceny-theft,
assault, drug-narcotic and vehicle theft. 
The fact that "other offenses" is so big indicates that there
are a lot of possible crimes out there, and even 39 categories of
crime isn't enough. "Non-criminal" means, I think, that the
police were called, but on arriving at the scene, they found that
no law had been broken.
I think the easy way to get the "top four" crimes out is to pull
them out of the data frame that `count` produces. They are
rows 1, 4, 5 and 6, so add a `slice` to your pipeline:
```{r }
my.rows=c(1,4,5,6)
my.crimes = sfcrime %>% count(Category) %>%
arrange(desc(n)) %>%
slice(my.rows) %>% pull(Category)
my.crimes
```

     

I just want the `Category` column (as a vector), and
`pull` is the way to get that. (If I don't do `pull`, I
get a data frame.)

If you can't think of anything, just type them into a vector with
`c`, or better, copy-paste them from the console. But that's a
last resort, and would cost you a point. If you do this, they have to
match exactly, UPPERCASE and all.
    


(e) (Digression, but needed for the next part.) The R vector
`letters` contains the lowercase letters from `a` to
`z`. Consider the vector `('a','m',3,'Q')`. Some of
these are found amongst the lowercase letters, and some not. Type
these into a vector `v` and explain briefly why 
`v %in% letters` produces what it does.



Solution


This is the ultimate "try it and see":
```{r }
v=c('a','m',3,'Q')
v %in% letters
```

   

The first two elements of the answer are `TRUE` because
lowercase-a and lowercase-m can be found in the lowercase letters
somewhere. The other two are false because the number 3 and the
uppercase-Q cannot be found anywhere in the lowercase letters.

The name is `%in%` because it's asking whether each element of
the first vector (one at a time) is *in* the set defined by the
second thing: "is `a` a lowercase letter?" ... 
is "`Q` a lowercase letter?" 
and getting the answers "yes, yes, no, no".
  


(f) We are going to `filter` only the rows of our data frame that
have one of the crimes in `my.crimes` as their
`Category`. Also, `select` only the columns
`Category`, `DayOfWeek` and `PdDistrict`. Save
the resulting data frame and display its structure. (You should have a
lot fewer rows than you did before.)



Solution


The hard part about this is to get the inputs to `%in%` the
right way around. We are testing the things in `Category` one
at a time for membership in the set in `my.crimes`, so this:
```{r error=T}
sfcrimea = sfcrime %>% filter(Category %in% my.crimes) %>%
select(c(Category,DayOfWeek,PdDistrict)) 
sfcrimea
```

   

That didn't work. Remember what this was? This is because `MASS` is still loaded:

```{r }
search()
```

 

so we need to get rid of `MASS`:

```{r error=T}
detach("package:MASS", unload=T)
```

 

and try again:

```{r error=T}
sfcrimea = sfcrime %>% filter(Category %in% my.crimes) %>%
select(c(Category,DayOfWeek,PdDistrict)) 
sfcrimea
```

   
I had trouble thinking of a good name for this one, so I put an "a"
on the end. (I would have used a number, but I prefer those for
models.) 

Down to a "mere" 359,000 rows. Don't be stressed that the
`Category` factor still has 39 levels (the original 39 crime
categories); only four of them have any data in them:

```{r }
sfcrimea %>% count(Category)
```

 

So all of the crimes that are left are one of the four Categories we
want to look at.
  


(g) Save these data in a file `sfcrime1.csv`.



Solution


This is `write_csv` again:
```{r }
write_csv(sfcrimea,"sfcrime1.csv")
```

   
  



