##  Trying on shirts


 Does it make a difference if you see somebody else trying on
a shirt before you do, and if so, does it matter what kind of person
it is? 

An experiment was carried out in a university bookstore, with
students who wanted to try on a shirt serving as (unsuspecting)
experimental subjects. When a student wanted to try on a shirt, the
sales associate told them that there was only one shirt left, and it
was being tried on by an "other customer". This "other customer"
was actually a "confederate" of the experimenter (that means, they
were pretending to be a real customer but were actually part of the
experiment). The "other customer" was always female: either an
attractive well-dressed model, or an average-looking student wearing
jeans. The "other customer" would come out of the
dressing room and hand the shirt to the sales associate, who would
give it to the student who wanted to try it on. When the student had
tried on the shirt, they were asked to rate it on a 7-point scale (on
five different dimensions, and those five scores were then
averaged). Does a student's evaluation depend on whether the student
was male or female, and whether the "other customer" was a model or
a student? There was also a control group, where the student was
handed the shirt directly  by the sales associate, without the 
"other customer" being involved at all. 

There were thus five treatments:
male students who saw a model, male students who saw a student, female
students who saw a model, female students who saw a student, and the
control group. There were 25 students in each treatment group.

The data from this experiment can be found at
[link](http://ritsokiguess.site/datafiles/shirts.csv).



(a) Read in and display the data. How many observations do you
have, and is that what you should have?

Solution


`read_csv`:
```{r }
my_url <- "http://ritsokiguess.site/datafiles/shirts.csv"
shirts <- read_csv(my_url)
shirts
```

     

There were 5 treatment groups with 25 students in each, so there
should be, and are, $5 \times 25=125$ observations.

$\blacksquare$

(b) Turn `treatment` into a `factor` in your data
frame. (You can use the same name as the text `treatment`
that you read in from the file.)

Solution


Thus:
```{r }
shirts <- shirts %>% mutate(treatment = factor(treatment))
```

     

and for checking:

```{r }
shirts
```

 

see that `treatment` is now definitely a factor rather than a
categorical variable represented as a piece of text.

$\blacksquare$

(c) List the treatments in the order that they are in within
your `factor`. (There are lots of ways to do this; any one of
them except for `distinct` is good.)

Solution


This is the most direct:
```{r }
levels(shirts$treatment)
```

     

or you can do one of the `tidyverse`-style ways, such as

```{r }
shirts %>% count(treatment)
```

 

or

```{r }
shirts %>% distinct(treatment)
```

 

except that `distinct` shows you the values in the order that
they appear in the data file, which is *not* the order that they
appear in the factor. So that's why I asked you not to do it that way.

Alphabetical order,  is the moral of the story.

Extra: you can make them appear in the order they appear in the data,
if you want to (and if you know what you are doing):

```{r }
shirts %>% mutate(trt2 = fct_inorder(treatment)) -> d
levels(d$trt2)
```

 

`fct_inorder` *creates* a factor whose levels are in the
order that they appeared in the data. If this is what you want, you
can go ahead and do this, and use this order when you create your
contrasts. It will work, and it will be good even though your
contrasts will look different from mine.

$\blacksquare$

(d) <a name="part:mean-eval">*</a> 
Obtain a table of mean evaluation scores for each treatment group.
  

Solution


We will use this later when assessing the significance of the
contrasts. It's the usual group-by and summarize:
```{r }
shirts %>%
  group_by(treatment) %>%
  summarize(m = mean(score))
```

     

Extra: how does this come out if you have your factor levels in a
non-alphabetical order?

```{r }
d %>% group_by(trt2) %>% summarize(m = mean(score))
```

 

Ah, it respects the order of the levels of the factor: that is to say,
the treatment groups appear here in the order that they were in the
data, because the factor `trt2` was created to make that
happen.

`forcats` has other tools to re-order the levels of a factor,
lump factor levels together to make a category "other", and so
on. These also work if you start with a categorical variable as text:
it *creates* a factor with those properties.
  
$\blacksquare$

(e) The experimenters wanted to compare four specific things in
their analysis:


* evaluation scores between male students who saw a (female) model and
male students who saw a (female) student

* evaluation scores between female students who saw a (female)
model and female students who saw a (female) student

* evaluation scores for male and for female students (overall)

* evaluation scores for the (average of the) four genuine
treatments and for the control group

Create contrasts, with suitable names, using vectors with
appropriate values.
  

Solution


These, in that order, are comparisons of treatments 4 and 5:
```{r }
c_mms <- c(0, 0, 0, 1, -1)
```

   

treatments 2 and 3:

```{r }
c_fms <- c(0, 1, -1, 0, 0)
```

 

(in these two the 1 and $-1$ can also be the other way around)

the average of 2 and 3 vs.\ the average of 4 and 5:

```{r }
c_mf <- c(0, 0.5, 0.5, -0.5, -0.5)
```

 

or you can use 1 and $-1$ instead of the 0.5s, or you can switch the
signs around. I like the 0.5 values to remind me it's an average of
two means.

Finally

```{r }
c_tc <- c(1, -0.25, -0.25, -0.25, -0.25)
```

 

or multiply through by 4 to get rid of the fractions, or have the
negative sign on the first one and positive signs on the others.

I tried to give the contrasts mnemonic but short names, so that I
would remember which was which.
  
$\blacksquare$

(f) Pick two of your contrasts (doesn't matter which two) and
demonstrate that they are orthogonal.
  

Solution


Multiply them together (elementwise, which is what `*`
does) and add them up, showing that you get zero, for example:
```{r }
sum(c_mf * c_tc)
```

     

Check (mops brow).
  
$\blacksquare$

(g) Collect all your contrasts together into a matrix and
declare that they are contrasts for `treatment` within your
data frame (whatever you called it).
  

Solution


I called my data frame `shirts`, so I need to do this:
```{r }
m <- cbind(c_mms, c_fms, c_mf, c_tc)
contrasts(shirts$treatment) <- m
```

       

There's no output here; we'll see in a moment whether it worked.

Extra: once you've created the matrix `m`, it gives you a
second way to test the contrasts for orthogonality, all at once:

```{r }
t(m) %*% m
```

 

This matrix-multiplies the transpose of `m` by `m`, so
it's testing each column of `m` to see whether it's orthogonal
to each other column. The columns don't have to be orthogonal to
themselves, hence the non-zeros down the diagonal, but all the
off-diagonal entries are zero. Hence each contrast is orthogonal to
each other contrast.

Or even:

```{r }
z <- t(m) %*% m
all(z[row(z) != col(z)] == 0)
```

 

That says (a little breathlessly) that it is true that all the
elements of $M^TM$ that are off the diagonal are zero.^[The thing inside the square brackets says only to look at the elements of $M^TM$ whose row number and whose column number are different; it is perhaps easier to reason that elements of a matrix whose row number and column number are the *same* are *on* the diagonal, for example the element in row 2, column 2.]
  
$\blacksquare$

(h) Predict evaluation score from 
treatment *as a  regression*, and display the results.
  

Solution


Once you have everything set up, it's just a matter of going
through the process:
```{r }
score.1 <- lm(score ~ treatment, data = shirts)
summary(score.1)
```

       

Note how the contrasts have appeared as "slopes" in our
regression. (If this didn't happen for you, go back and check what you
did. Probably something gave you an error before, in that case.)

Your slopes can differ in terms of sign and possibly value from mine,
but the $t$-statistics and P-values should be the same as mine.
  
$\blacksquare$

(i) For each of your contrasts, assess whether or not it is
significant, and explain briefly what that means in the context of the
data. If a contrast is significant, use your answer to
part (<a href="#part:mean-eval">here</a>) to help in your interpretation. 
    

Solution


I'll take my four contrasts as they appear in the output from `lm`:


* Comparing the male students who saw the model with those
who saw the student: this is strongly significant, and the
table of means shows that the males who saw the model rated
the shirt significantly *higher* than those who saw the
"average-looking" female student.

* Comparing the female students who saw the model with
those who saw the student: this difference is not quite
significant, so we conclude that it didn't matter who the
female students saw coming out of the dressing room.

* There is a significant difference between the average
evaluation score that males gave to the shirt and that
females did. The table of means reveals that the male
average was higher, mostly because males *really* liked
the shirt when they thought the model had been wearing it!

* There is no significant difference overall between
treatment and control groups in terms of average
score. (This is mostly because the treatment groups differ
among themselves, with the scores for students who saw the
model being higher and those for students who saw the female
student being lower than for the control group.)

It is enough, on the non-significant ones, to say what it is
that the contrast is testing, and then to note that it is not
significant. On the significant ones, I want you to say what
is higher than what.
You can also get these conclusions from the `lm`
output, but it requires a bit more care:


* I set up `c_mms` with the score for the students
who had seen the model as plus and those who saw the student
as minus. My slope here is positive, so the score for
male students who saw the model is higher.

* My slope for `c_fms` is also positive, and I set
this one up the same way with model as positive and student
as negative, so the female students rated the shirt higher
(but not significantly higher) when they saw the model.

* My `c_mf` slope is significantly negative. I set
up this contrast with females as plus and males as minus, so
this means males gave a significantly *higher* score on
average (for reasons discussed above).

* The last slope is very close to 0, consistent with there
being no difference (on average) between treatment and control.

Perhaps the story to get from this analysis is that male
students are *such* suckers!
    
$\blacksquare$




