## The power of the sign test


 I've mentioned several times that the sign test has less
power than the $t$-test. Let's investigate this with a specific example.

Let's suppose we are testing $H_0: \mu=40$ against $H_a: \mu \ne 40$,
where $\mu$ is the population mean (and median, as we shall see). Our
population actually has a normal distribution with mean 50 and SD 15,
so that the null hypothesis is *wrong* and we want to reject it
most of the time. On the other hand, the population actually *is*
normally-distributed and so the $t$-test is the right one to use.

(This is an old question, so I tackle the simulated power differently
than I did it in class this time. But see if you can follow what I do
here.)



(a) Use `power.t.test` to find the probability that a
$t$-test correctly rejects the null hypothesis using a sample size
of $n=10$.


Solution


```{r }
power.t.test(delta=50-40,n=10,sd=15,type="one.sample",alternative="two.sided")
```       

The power is 0.469. Not great, but we'll see how this stacks up
against the sign test.




(b) What code in R would draw a random sample of size 10 from the
*true* population distribution and save the sample in a variable?



Solution


The data actually have a normal distribution with mean 50 and
SD 15, so we use `rnorm` with this mean and SD, obtaining
10 values:
```{r }
x=rnorm(10,50,15)  
x
``` 




(c) What code would count how many of the sampled values are less
than 40 and how many are greater (or equal)? 



Solution


 The way we know this is to put `x` into a data frame first:
```{r }
tibble(x) %>% count(x<40)
``` 

2 values less (and 8 greater-or-equal).




(d) It turns out the sign test would reject $H_0: M=40$ against
$H_a: M \ne 40$ at $\alpha=0.05$ if the smaller of the numbers in the
last part is 1 or less. ($M$ is the population median.) 
Add to your pipeline to obtain `TRUE`
if you should reject the null for your
data and `FALSE` otherwise. 



Solution


This is actually easier than you might think. The output from
`count` is a data frame with a column called `n`,
whose minimum value you want. I add to my pipeline:
```{r }
tibble(x) %>% count(x<40) %>%
summarize(the_min=min(n)) %>%
mutate(is_rejected=(the_min<=1))
``` 

This will fail sometimes. If all 10 of your sample values are greater
than 40, which they might turn out to be, you'll get a table with only
one line, `FALSE` and 10; the minimum of the `n` values
is 10 (since there is only one), and it will falsely say that you
should not reject. The fix is

```{r }
tibble(x) %>% count(x<40) %>%
summarize(the_min=min(n)) %>%
mutate(is_rejected=(the_min<=1 | the_min==10))
``` 

The above is almost the right thing, but not quite: we only want that value
that I called `is_rejected`, rather than the whole data frame,
so a `pull` will grab it:

```{r }
tibble(x) %>% count(x<40) %>%
summarize(the_min=min(n)) %>%
mutate(is_rejected=(the_min<=1 | the_min==10)) %>%
pull(is_rejected)
``` 

You might be wondering where the "1 or less" came from. Getting a
P-value for the sign test involves the binomial distribution: if the
null is correct, each data value is independently either above or
below 40, with probability 0.5 of each, so the number of values below
40 (say) is binomial with $n=10$ and $p=0.5$. The P-value for 1
observed value below 40 and the rest above is

```{r }
2*pbinom(1,10,0.5)  
``` 

which is less than 0.05; the P-value for 2 values below 40 and the
rest above is 

```{r }
2*pbinom(2,10,0.5)    
``` 

which is bigger than 0.05. 

You might have encountered the term "critical region" for a
test. This is the values of your test statistic that you would reject
the null hypothesis for. In this case, the critical region is 1 and 0
observations below 40, along with 1 and 0 observations above 40.

When you're thinking about power, I think it's easiest to think in
terms of the critical region (rather than directly in terms of
P-values) since you have a certain $\alpha$ in mind all the way
through, 0.05 in the power examples that I've done. The steps are then:



* Work out the critical region for your test, the values of the
test statistic (or sample mean or sample count) that would lead to
rejecting the null hypothesis.


* Under your particular alternative hypothesis, find the
probability of falling into your critical region.


When I say "work out", I mean either calculating (along the lines of
STAB57), or simulating, as we have done here.



(e) Use `rerun` to simulate the above process 1000 times:
drawing a random sample from a normal distribution with mean 50 and SD
15, counting the number of values below and above 40, rejecting if the
minimum of those is 1 or less, then counting the number of rejections
out of 1000.



Solution


The way we've used `rerun` is to use it to select the
random samples, and then we use `map` ideas to do what we
want to do with each random sample, along the lines of what we did
with the one random sample above.
This is liable to go wrong the first few times, so make sure that
each line works before you go on to the next. (`rerun` will
produce you a `list` of random samples, with each of which
you want to do something.) While you're debugging, try it with a
small number of random samples like 5.
I start with setting the random number seed, so it comes out the
same each time. I discuss the results below and the code below that.
```{r }
set.seed(457299)
rerun(1000,rnorm(10,50,15)) %>% 
map(~tibble(x=.)) %>% 
map(~count(.,x<40)) %>% 
map(~summarize(.,the_min=min(n))) %>% 
map(~mutate(.,is_rejected=(the_min<=1 | the_min==10))) %>% 
map_lgl(~pull(.,is_rejected)) %>% 
tibble(was_true=.) %>% 
count(was_true)
```     

The estimated power of the sign test is 0.243, since that was the
number of times a simulated sample gave us 0 or 1 values above or
below 40 (and the rest on the other side).

All right, that code is seriously scary. Let me try to take you
through it.



* The `rerun` line is the same kind of thing we had before:
generate 1000 different random samples of size 10 from a normal
distribution with mean 50 and SD 15.

* The output from the previous step is a `list` of
vectors. But we like data frames to count things in, so for each
vector we turn it into a data frame, filling a column called
`x` with whatever was in each vector (that random sample). So
now we have 1000 data frames each containing a column called
`x`.

* Next, in each of those data frames, count up how many of the
`x` values are less than 40. This will produce a data frame
each time containing a column `n` that is the
frequencies. Here and below, the `.`  is used to denote
"it": that is, each of the elements of the list created originally
by `rerun` that we are doing something with. Also, at the
moment, the output for each element of the list is a data frame, so
we stick with `map` for the time being.

* Next, for each of those tables of frequencies, find the smallest
one and call it `the_min`. (As discussed above, all the
values might be bigger than 40, in which case `the_min` will
be 10 rather than 0, which we handle next.)

* Next, we create a new column called `is_rejected` which
says that we should reject a median of 40 if the minimum value we
just calculated is 1 or less, or if it happens to be 10, in which
case that would have been the only entry in the frequency table, so
that the missing one would have been zero.

* Next, we pull out only the true-or-false value in
`is_rejected`. At last, the answer here is not a data frame
but a simple logical value; `map_lgl` is like
`map_dbl` except that the thing we are doing returns a
`TRUE` or a `FALSE` rather than a number.

* At this point we have a vector of 1000 true-or-false. We want to
count them, so we put them into a data frame (with a column called
`was_true`), and in the last line, count them up. There are
243 (correct) rejections and 757 (incorrect) non-rejections.


You may now breathe again. 

I'm now thinking a better way to do this is to write a function that
takes a sample (in a vector) and returns a TRUE or FALSE according to
whether or not a median of 40 would be rejected for that sample:

```{r }
is_reject=function(x) {
tibble(x=x) %>%
count(x<40) %>%
summarize(the_min=min(n)) %>%
mutate(is_rejected=the_min<=1 | the_min==10) %>%
pull(is_rejected)
}
``` 

Now, we have to use that. This function will be "mapped" over for
each of the random samples that come out of `rerun`, but now
there will be only one `map` because the complication of the
multiple `maps` has been subsumed into this one function. I'll
set my random number seed so that I get the same results as before:

```{r }
set.seed(457299)
rerun(1000,rnorm(10,50,15)) %>%
map_lgl(~is_reject(.)) %>%
tibble(rejected=.) %>%
count(rejected)
``` 

Same results, and yeah, I like that a lot better.



(f) Which is more powerful in this case, the sign test or the
$t$-test? How do you know?



Solution


The power of the sign test is estimated as 0.243, which is quite a bit less
than the power of the $t$-test, which we found back in (a) to be
0.469. So the $t$-test, in this situation where it is valid, is
the right test to use: it is (a) valid and (b) more powerful.
So the $t$-test is more powerful. One way to think about how
*much* more powerful is to ask ``how much smaller of a sample
size would be needed for the $t$-test to have the same power as
this sign test?'' The power of my sign test was 0.243, so in
`power.t.test` we set
`power` equal to that and
omit the sample size `n`:

```{r }
power.t.test(delta=50-40,power=0.243,sd=15,type="one.sample",alternative="two.sided")
```       
A sample of size 6 gives the same power for the $t$-test that a
sample of size 10 does for the sign test. The ratio of these two
sample sizes is called the *relative efficiency* of the two
tests: in this case, the $t$-test is $10/6=1.67$ times more
efficient. The data that you have are being used ``more
efficiently'' by the $t$-test.
It is possible to derive\endnote{Meaning, I forget how to do it.
But it has something to do with looking at alternatives that are
very close to the null.}  the limiting relative efficiency of
the $t$ test relative to the sign test when the data are actually
normal, as the sample size gets larger. This turns out not to
depend on how far wrong the null is (as long as it is the same for
both the $t$-test and the sign test). This ``asymptotic relative
efficiency'' is $\pi/2=1.57$. Our relative efficiency for power
0.243, namely 1.67, was pretty close to this, even though our
sample sizes 10 and 6 are not especially close to infinity.
This says that, if your data are actually from a normal
distribution, you do a lot better to use the $t$-test than the
sign test, because the sign test is wasteful of data (it only uses
above/below rather than the actual values). 
If your data are *not* from a normal distribution, then the
story can be very different. 
Of course you knew I would investigate this. There is a
distribution called the "Laplace" or "double exponential"
distribution, that has very long tails.\endnote{If you've ever run
into the exponential distribution, you'll recall that this is
right skewed with a very long tail. The Laplace distribution looks
like two of these glued back to back.} The distribution is not in
base R, but there is a package called `smoothmest` that
contains a function `rdoublex` to generate random values from
this distribution. So we're going to do a simulation investigation
of the power of the sign test for Laplace data, by the same
simulation technique that we did above. Like the normal, the Laplace
distribution is symmetric, so its mean and median are the same
(which makes our life easier).\endnote{This is about the *only*
way in which the normal and Laplace distributions are alike.}

Let's test the hypothesis that the median is zero. We'll suppose that
the true median is 0.5 (this is called `mu` in
`rdoublex`). The first problem we run into is that we can't use
`power.t.test` because they assume normal data, which we are
far from having. So we have to do two simulations: one to simulate the
power of the $t$ test, and one to simulate the power of the sign test.

To simulate the $t$ test, we first have to generate some Laplace data
with the true mean of 0.5. We'll use a sample size of 50 throughout
these simulations.

```{r }
library(smoothmest)
rl=rdoublex(50,mu=0.5)
rl
``` 

This seems to have some unusual values, far away from zero:

```{r }
tibble(rl) %>%
ggplot(aes(sample=rl))+
stat_qq()+stat_qq_line()
``` 

You see the long tails compared to the normal.

Now, we feed these values into `t.test` and see whether we
reject a null median of zero (at $\alpha=0.05$):

```{r }
tt=t.test(rl)  
tt
``` 

Or we can just pull out the P-value and even compare it to 0.05:

```{r }
pval=tt$p.value  
pval
is.reject=(pval<=0.05)
is.reject
``` 
$

This one has a small P-value and so the null median of 0 should be
(correctly) rejected.

We'll use these ideas to simulate the power of the $t$-test for these
data, testing a mean of 0. This uses the same ideas as for any power
simulation; the difference here is the true distribution:

```{r }
rerun(1000,rdoublex(50,mu=0.5)) %>%
map(~t.test(.,mu=0)) %>%
map_dbl("p.value") ->
pvals
``` 

and then count them:

```{r }
tibble(pvals) %>% count(pvals<=0.05)
``` 

And now we simulate the sign test. Since what we want is a P-value
from a vector, the easiest way to do this is to use
`pval_sign0` from `smmr`, which returns exactly the
two-sided P-value that we want, so that the procedure is a step simpler:

```{r }
rerun(1000,rdoublex(50,mu=0.5)) %>%
map_dbl(~pval_sign0(0,.)) ->
pvals_sign
``` 

and then

```{r }
tibble(pvals_sign) %>% count(pvals_sign<=0.05)
``` 

For data from this Laplace
distribution, the power of this $t$-test is 0.696, but the power of
the sign test on the same data is 0.761, *bigger*.  For
Laplace-distributed data, the sign test is *more* powerful than
the $t$-test.

This is not to say that you will ever run into data that comes from
the Laplace distribution. But the moral of the story is that the sign
test *can* be more powerful than the $t$-test, under the right
circumstances (and the above simulation is the "proof" of that
statement). So a blanket statement like ``the sign test is not very
powerful'' needs to be qualified a bit: when your data come from a
sufficiently long-tailed distribution, the sign test can be more
powerful relative to the $t$-test than you would think.

I finish by "unloading" the two packages that got loaded:

```{r }
detach(package:smoothmest, unload=T)
detach(package:MASS, unload=T)
``` 





