##  Making soap


 A company operates two production lines in a factory for
making soap bars. The production lines are
labelled A and B. A production line that moves faster may produce more
soap, but may possibly also produce more "scrap" (that is, bits of
soap that 
can no longer be made into soap bars and will have to be thrown away).  

The data are in [link](http://www.utsc.utoronto.ca/~butler/c32/soap.txt).



(a) Read the data into R.  Display the data. There should be 27
rows. Are there?


Solution


Read directly from the URL, most easily:
```{r }
url <- "http://www.utsc.utoronto.ca/~butler/c32/soap.txt"
soap <- read_delim(url, " ")
soap
```

    

27 rows. `line`, which is either `a` or `b`, was
correctly deduced to be text.



(b) Obtain a histogram of the `scrap` values, using 10
bins for your histogram.
 

Solution



```{r }
ggplot(soap, aes(x = scrap)) + geom_histogram(bins = 10)
```

 



(c) Comment briefly on the shape of the histogram. Is it approximately
symmetric, skewed to the left, skewed to the right or something else? (By "comment  briefly" I mean "say in a few words why you gave the answer you did.")


Solution


I would call this "bimodal". There are two peaks to the
histogram, one around 250 and one around 370, with a very small
frequency in between (the bar around 300). 
Apart from the bimodality, there is no particular evidence for a
long tail on either end, so I don't think you could otherwise
call it anything other than symmetric.
Having said that (this is going beyond the question), the way a
histogram looks can depend on the bins you choose to draw it
with. This is 8 bins rather than 10:
```{r }
ggplot(soap, aes(x = scrap)) + geom_histogram(bins = 8)
```

 

The middle low-frequency bin has gone, and this one just looks
symmetric, with a kind of "flat top".



(d) Make side-by-side boxplots of scrap values for each
production line.


Solution



```{r }
ggplot(soap, aes(x = line, y = scrap)) + geom_boxplot()
```



One categorical, one quantitative variable, so boxplots make sense.



(e) Do you think your boxplot says that there are differences in the
amount of scrap produced by the two production lines, or not?
Explain briefly.


Solution


I would say that there *is* a difference between the two
production lines, with line A producing an average (median) of
about 330 and line B producing a median of about 275. But you
could also make the case that, although the medians are rather
different, there is a lot of variability and hence a lot of
overlap between the two boxplots, and therefore that there is
not a "substantial" difference.
I would say that either of those answers are good \emph{if you
back them up with proper reasons}. This is going to be a
common theme in this course: I am going to ask you to make a
decision and support it, where the reasons you provide are often
more
important than the decision you make.
You might be wondering whether the medians, or means, since
there is no serious skewness here and definitely no outliers,
are "significantly different". This is inference, which we
will come to later, but a preview looks like this:
```{r }
t.test(scrap ~ line, data = soap)
```

       

They are not: the P-value of 0.22 is not anywhere near as small as
0.05, so we can't reject the null hypothesis that the two lines have
equal mean amount of scrap. 

Rusty on this stuff? Don't worry; we're going to come back to it later
in the course.

I was also wondering about something else: that bimodal
histogram. Could that be explained by the scrap values being two
different production lines being mixed together? One way to understand
that is to have two separate histograms, one for each line, side by
side, which is what facetting does. There is an extra wrinkle here
that I explain afterwards:

```{r }
ggplot(soap, aes(x = scrap)) + geom_histogram(bins = 10) + facet_grid(line ~ .)
```

 

I could have used `facet_wrap`, but that would have put the
histograms side by side, and I wanted them one above the other (for
ease of comparison, since they'll be on the same
scale). `facet_grid` is like `facet_wrap`, but offers
you more control over where the facets go: you can arrange them above
and below by a variable, or left and right by a variable. Whatever is
facetting the plots up and down (on the $y$ axis) goes before the
squiggle, and whatever facets them left and right goes after. If there
is nothing separating the facets in one direction, here horizontally,
the variable is replaced by a dot.

In some ways, `facet_grid` is also *less* flexible,
because the facets have to be arranged up/down or left/right by a
variable. That worked here, but if you think back to the Australian
athletes, where there were ten different sports, it was
`facet_wrap` that did the right thing, arranging the sports
along rows *and* columns to produce a pleasing display.

All right, that bimodality. I was expecting that the scrap values from
one line would be centred about one value and the scrap values from
the other line would be centred about a different value, with a gap in
between. But that's not what happened at all: the line B values are
all over the place, while it's the line A values that are actually
bimodal all by themselves. I'm not sure whether that really means
anything, since the data sets are pretty small, but it's kind of
interesting. 



(f) We started out with the suspicion that if the line was run
faster, there would be more scrap. We haven't assessed this
yet. Draw a scatter plot with `scrap` on the $y$ axis and
`speed` on the $x$ axis. 


Solution


Same mechanism as before:
```{r }
ggplot(soap, aes(x = speed, y = scrap)) + geom_point()
```

 



(g) What do you think is the most important conclusion from
your plot of the previous part? Describe your conclusion in the
context of the data.


Solution


There seems to be a pretty evident upward trend, apparently
linear, which means that if the speed of the production line is
higher, the amount of scrap produced is also higher.
My last sentence was meant to remind you that 
"there is an upward trend" is *not a complete answer*: 
we are concerned
with what that upward trend tells us about the data. 
This, in other words, confirms the suspicion expressed in the
question, which was therefore a rather large clue: more speed
tends to go with more scrap.
That was as far as I wanted you to go: there seems to be an
association with speed, and there might be an association with
`line` that turned out not to be statistically significant.
What we haven't done is to assess the relationship between speed
and scrap for *each* production line. To do that, we want
to plot the scrap-speed points distinguished for each production
line. `ggplot` makes that easy: you add a
`colour`
`r tufte::margin_note("If you are concerned about the spelling: the guy who wrote ggplot is from New Zealand, where they spell *colour* the same way we do. However, if you want to use *color*, that works too.")` to say what you want to
distinguish by colour. This is two quantitative variables and one
categorical variable, if you want to think of it that way:
```{r }
ggplot(soap, aes(x = speed, y = scrap, colour = line)) + geom_point()
```

       

Notice that we get a legend, automatically.

What is interesting about this one is the red dots are mostly at the
top (for any given speed), and the blue dots are mostly at the
bottom. That seems to mean that *when we account for speed*,
there is a difference between lines.

I want to show you one more embellishment, which is to put the
regression lines on the plot for each group separately. This is where
`ggplot` is so nice, since I just have to add one thing:

```{r }
ggplot(soap, aes(x = speed, y = scrap, colour = line)) +
  geom_point() + geom_smooth(method = "lm", se = F)
```

       

The points and lines have come out in different colours, without our
having to think too hard.

Both lines show an upward trend, with about the same slope, which
means that regardless of line, increasing the speed goes with
increasing the scrap by the same amount. The fact that the red line is
above the blue one, however, suggests that production line A produces
more scrap at the same speed than production line B.

From a management point of view, there is an interesting dynamic at
work: if you run the production line faster, you'll produce more bars
of soap, but you'll produce more scrap as well. The crucial thing for
the people in the supervisor's office is how much raw material is used
per bar of soap, and if you make the soap bars faster, you might use
more raw material, which will eat into your profits (from one angle),
but you will also have more bars of soap to sell.

Here's another way to see the same thing. I'm *definitely* not
expecting you to follow the code, but you can admire the result!

```{r }
soap2 <- soap %>% select(-line)
ggplot(soap, aes(x = speed, y = scrap)) +
  geom_point(data = soap2, colour = "grey") +
  geom_point(aes(colour = line)) + facet_wrap(~line)
```
$

The idea is that we plot all the points in grey (to 
"put them in the background") 
 and then in each plot we plot the points again,
*coloured, for the group we are looking at*: line A in the left,
line B on the right. This is another way of seeing that line A has
more scrap than line B, given the speed at which the line was being
run. (I discovered this technique only yesterday. I think the code is
remarkably concise for what it does.)

The logic of the code is:


* create a new data frame that contains everything in
`soap` except for `line`

* make a scatter plot of all the points in this new data frame,
coloured grey

* plot the points again (from the original data frame), coloured by
which production line they're from

* produce a separate scatterplot for each production line.


The trick about creating the new data frame was to enable plotting of
all points regardless of group on each subplot ("facet" in
`ggplot` terminology), as well as the ones that come from that
production line.

I don't expect you to be able to follow all the details of the code
below, either,
but I would like you to try and get the logic. What we do is a
regression predicting `scrap` from *two* things:
`speed` and production `line`. The results we get are these:

```{r }
scrap.1 <- lm(scrap ~ speed + line, data = soap)
summary(scrap.1)
```

 

The P-values for `speed` and `line` are the second and
third things in the last column, $7 \times 10^{-16}$  and $1 \times 10^{-6}$
respectively. These are both very strongly significant, in contrast to
the two-sample $t$-test where `line` was not significant.

So does production line make a difference or not?

The plot says that it does, and the meaning of model `scrap.1`
just above is that \emph{`speed` affects scrap when you account
for `line`}, and emph{`line` affects scrap when you
account for speed}. (In the two-sample $t$-test above we didn't
account for speed at all, since the various speeds were all mixed up.)

There is a moral to this story, which I would like you to get even if
you don't get any of the statistics: if a variable makes a
difference, it should be in your model and on your
graph,
`r tufte::margin_note("Meaning that the graph should contain all three variables, *speed*, *scrap* and *line*.")`
because it enables you to get better (more precise) conclusions about your
other variables. Here, there really is a difference between the
production lines, but the $t$-test was too much of a blunt instrument
to unearth it (because `speed` made a difference as well).




