##  Similarity of species 


 Two scientists assessed the dissimilarity
between a number 
of species by recording the number of positions in the protein
molecule cytochrome-$c$ where the two species being compared have
different amino acids. The dissimilarities that they recorded are in
[link](http://ritsokiguess.site/datafiles/species.txt).



(a) Read the data into a data frame and take a look at it.


Solution


Nothing much new here:
```{r }
my_url <- "http://ritsokiguess.site/datafiles/species.txt"
species <- read_delim(my_url, " ")
species
```

     

This is a square array of dissimilarities between the eight species.

The data set came from the 1960s, hence the use of "Man" rather than
"human". It probably also came from the UK, judging by the spelling
of `Mould`.

(I gave the first column the name `what` so that you could
safely use `species` for the whole data frame.)
    


(b) Bearing in mind that the values you read in are
*already* dissimilarities, convert them into a `dist`
object suitable for running a cluster analysis on, and display the
results. (Note that you need to get rid of any columns that don't
contain numbers.)


Solution


The point here is that the values you have are *already*
dissimilarities, so no conversion of the numbers is required. Thus
this is a job for `as.dist`, which merely changes how it
looks. Use a pipeline to get rid of the first column first:
```{r }
species %>%
  select(-what) %>%
  as.dist() -> d
d
```



This doesn't display anything that it doesn't need to: we know that the
dissimilarity between a species and itself is zero (no need to show
that), and that the dissimilarity between B and A is the same as
between A and B, so no need to show everything twice. It might look as
if you are missing a row and a column, but one of the species (Fungus)
appears only in a row and one of them (Man) only in a column.

This also works, to select only the numerical columns:

```{r }
species %>%
  select(where(is.numeric)) %>%
  as.dist()
```

 

Extra: data frames officially have an attribute called "row names",
that is displayed where the row numbers display, but which isn't
actually a column of the data frame. In the past, when we used
`read.table` with a dot, the first column of data read in from
the file could be nameless (that is, you could have one more column of
data than you had column names) and the first column would be treated
as row names. People used row names for things like identifier
variables. But row names have this sort of half-existence, and when
Hadley Wickham designed the `tidyverse`, he decided not to use
row names, taking the attitude that if it's part of the data, it
should be in the data frame as a genuine column. This means that when
you use a `read_` function, you have to have exactly as many
column names as columns. 

For these data, I previously had the column here called
`what` as row names, and `as.dist` automatically got rid
of the row names when formatting the distances. Now, it's a
genuine column, so I have to get rid of it before running
`as.dist`. This is more work, but it's also more honest, and
doesn't involve thinking about row names at all. So, on balance, I
think it's a win.
  


(c) Run a cluster analysis using single-linkage and obtain a dendrogram.


Solution


Something like this:
```{r }
species.1 <- hclust(d, method = "single")
plot(species.1)
```

     
    


(d) Run a cluster analysis using Ward's method and obtain a dendrogram.


Solution


Not much changes  here in the code, but the result is noticeably
different:
```{r }
species.2 <- hclust(d, method = "ward.D")
plot(species.2)
```

  

Don't forget to take care with the `method`: it has to be
`ward` in lowercase (even though it's someone's name) followed
by a D in uppercase.

    


(e) Describe how the two dendrograms from the last two parts
look different.


Solution


This is (as ever with this kind of thing) a judgement call. Your
job is to come up with something reasonable.
For myself, I was thinking about how single-linkage tends to
produce "stringy" clusters that join single objects (species)
onto already-formed clusters. Is that happening here? Apart from
the first two clusters, man and monkey, horse and pig,
*everything* that gets joined on is a single species joined
on to a bigger cluster, including mould and fungus right at the
end. Contrast that with the output from Ward's method, where, for
the most part, groups are formed first and then joined onto other
groups. For example, in Ward's method, mould and fungus are joined
earlier, and also the man-monkey group is joined to the
pigeon-horse-pig group.
`r tufte::margin_note("Tuna is an exception, but usually Ward    tends to join fairly dissimilar things that are nonetheless more    similar to each other than to anything else. This is like    Hungarian and Finnish in the example in class: they are very    dissimilar languages, but they are more similar to each other than    to anything else.")`
You might prefer to look at the specifics of what gets joined. I
think the principal difference from this angle is that mould and
fungus get joined together (much) earlier in Ward. Also, pigeon
gets joined to horse and pig first under Ward, but *after*
those have been joined to man and monkey under
single-linkage. This is also a reasonable kind of observation.
    


(f) Looking at your clustering for Ward's method, what seems to
be a sensible number of clusters? Draw boxes around those clusters.


Solution


Pretty much any number of clusters bigger than 1 and smaller than
8 is ok here, but I would prefer to see something between 2 and
5, because a number of clusters of that sort offers (i) some
insight ("these things are like these other things") and (ii) a
number of clusters of that sort is supported by the data.
To draw those clusters, you need `rect.hclust`, and
before that you'll need to plot the cluster object again. For 2
clusters, that would look like this:
```{r }
plot(species.2)
rect.hclust(species.2, 2)
```

 

This one is "mould and fungus vs. everything else". (My red boxes
seem to have gone off the side, sorry.)

Or we could go to the other end of the scale:

```{r }
plot(species.2)
rect.hclust(species.2, 5)
```

 

Five is not really an insightful number of clusters with 8 species,
but it seems to correspond (for me at least) with a reasonable
division of these species into "kinds of living things". That is, I
am bringing some outside knowledge into my number-of-clusters division.
    


(g) List which cluster each species is in, for your preferred
number of clusters (from Ward's method).


Solution


This is `cutree`. For 2 clusters it would be this:
```{r }
cutree(species.2, 2)
```

     

For 5 it would be this:

```{r }
cutree(species.2, 5)
```

 

and anything in between is in between. 

These ones came out sorted, so there is no need to sort them (so you
don't need the methods of the next question).
    



