

 When you order a steak in a restaurant, the server will ask
you how you would like it cooked, or to be precise, *how much*
you would like it cooked: rare (hardly cooked at all), through medium
rare, medium, medium well to well (which means "well done", so that
the meat has only a little red to it). Could you guess how a person
likes their steak cooked, from some other information about them? The
website [link](fivethirtyeight.com) commissioned a survey where they
asked a number of people how they preferred their steak, along with as
many other things as they could think of to ask. (Many of the
variables below are related to risk-taking, which was something the
people designing the survey thought might have something to do with
liking steak rare.) The variables of interest are all factors or true/false:



* `respondent_ID`: a ten-digit number identifying each
person who responded to the survey.

* `lottery_a`: true if the respondent preferred lottery A
with a small chance to win a lot of money, to lottery B, with a
larger chance to win less money.

* `smoke`: true if the respondent is currently a smoker

* `alcohol`: true if the respondent at least occasionally
drinks alcohol.

* `gamble`: true if the respondent likes to gamble (eg.
betting on horse racing or playing the lottery)

* `skydiving`: true if the respondent has ever been
skydiving.

* `speed`: true if the respondent likes to drive fast

* `cheated`: true if the respondent has ever cheated on a
spouse or girlfriend/boyfriend

* `steak`: true if the respondent likes to eat steak

* `steak_prep` (response): how the respondent likes their
steak cooked (factor, as described above, with 5 levels).

* `female`: true if the respondent is female

* `age`: age group, from 18--29 to 60+.

* `hhold_income`: household income group, from \$0--24,999
to \$150,000+.

* `educ`: highest level of education attained, from 
"less than high school" up to "graduate degree"

* `region`: region (of the US)
that the respondent lives in (five values).


The data are in
[link](http://www.utsc.utoronto.ca/~butler/d29/steak1.csv). This is the
cleaned-up data from a previous question, with the missing values removed.



(a) Read in the data and display the first few lines.


Solution


The usual:

```{r }
steak=read_csv("steak1.csv")
head(steak)
```

 

We should also (re-)load `MASS`, since we'll be needing it:

```{r }
library(MASS)
```

 



(b) We are going to predict `steak_prep` from some of
the other variables. Why is the model-fitting function `polr`
from package `MASS` the best choice for these data
(alternatives being `glm` and `multinom` from package
`nnet`)?


Solution


It all depends on the kind of response variable. We have a
response variable with five ordered levels from Rare to
Well. There are more than two levels (it is more than a
"success" and "failure"), which rules out `glm`, and
the levels are ordered, which rules out `multinom`. As we
know, `polr` handles an ordered response, so it is the
right choice.
    


(c) What are the levels of `steak_prep`, \emph{in the
order that R thinks they are in?} If they are not in a sensible
order, create an ordered factor where the levels are in a sensible order.


Solution


This is the most direct way to find out:
```{r }
preps = steak %>% distinct(steak_prep) %>% pull(steak_prep)
preps
```
$ %$ %$ %$

This is almost the right order (`distinct` uses the order in
the data frame). We just need to switch the first two around, and then
we'll be done:

```{r }
preps1=preps[c(2,1,3,4,5)]
preps1             
```

 

If you used `count`, there's a bit more work to do:

```{r }
preps2 = steak %>% count (steak_prep) %>% pull(steak_prep)
preps2
```

 

because `count` puts them in alphabetical order, so:

```{r }
preps3=preps2[c(4,2,1,3,5)]
preps3
```

 

These use the idea in the
attitudes-to-abortion question: create a vector of the levels in the
*right* order, then create an ordered factor with
`ordered()`. If you like, you can type the levels in the right
order (I won't penalize you for that here), but it's really better to
get the levels without typing or copying-pasting, so that you don't
make any silly errors copying them (which will mess everything up
later).

So now I create my ordered response:

```{r }
steak = steak %>% mutate(steak_prep_ord=ordered(steak_prep,preps1))
```

 
or using one of the other `preps` vectors containing the levels
in the correct order.
$ %$ %$

As far as `polr` is concerned,
it doesn't matter whether I start at `Rare` and go "up", or
start at `Well` and go "down". So if you do it the other way
around, that's fine. As long as you get the levels in a sensible
order, you're good.
    


(d) Fit a model predicting preferred steak preparation in an
ordinal logistic regression from `educ`, `female` and
`lottery_a`. This ought to be easy from your previous work,
but you have to be careful about one thing. No need to print out the
results. 


Solution


The thing you have to be careful about is that you use the
*ordered* factor that you just created as the response:
```{r finetti,cache=T}
steak.1=polr(steak_prep_ord~educ+female+lottery_a,data=steak)
```

     
    


(e) Run `drop1` on your fitted model, with
`test="Chisq"`. Which explanatory variable should be removed
first, if any? Bear in mind that the variable with the
*smallest* AIC should come out first, in case your table
doesn't get printed in order.


Solution


This:

```{r }
drop1(steak.1,test="Chisq")
```

 

My table is indeed out of order (which is why I warned you about it,
in case that happens to you as well). The smallest AIC goes with
`female`, which also has a very non-significant P-value, so
this one should come out first.
    


(f) Remove the variable that should come out first, using
`update`. (If all the variables should stay, you can skip
this part.)


Solution


You could type or copy-paste the whole model again, but
`update` is quicker:
```{r }
steak.2=update(steak.1,.~.-female)
```

     

That's all.

I wanted to get some support for my `drop1` above (since I was
a bit worried about those out-of-order rows). Now that we have fitted
a model with `female` and one without, we can compare them
using `anova`:

```{r }
anova(steak.2,steak.1,test="Chisq")
```

 

Don't get taken in by that "LR stat" on the end of the first row of
the output table; the P-value wrapped onto the second line, and is in
fact exactly the same as in the `drop1` output (it is doing
exactly the same test). As non-significant as you could wish for.

I was curious about whether either of the other $x$'s could come out now:

```{r }
drop1(steak.2,test="Chisq")
```

 

`lottery_a` should come out, but `educ` is edging
towards significance.
    


(g) Using the best model that you have so far, predict the
probabilities of preferring each different steak preparation (method
of cooking) for each combination of the variables that
remain. (Some of the variables are TRUE and FALSE rather than
factors. Bear this in mind.)
Describe the effects of each variable on the predicted
probabilities, if any. Note that there is exactly one person in the
study whose educational level is "less than high school".


Solution


Again, I'm leaving it to you to follow all the steps. My variables
remaining are `educ` and `lottery_a`. The second of
these is a logical, TRUE and FALSE, so I just filled in the two values:
```{r }
educs = steak %>% distinct(educ) %>% pull(educ)
lottery_as=c(FALSE,TRUE)

steak.new=crossing(educ=educs,lottery_a=lottery_as)
steak.new
```

     

My best model so far is the one I called `steak.2`, without
`female`, so this. As before, if you forget what `type`
should be, put in *something*, and it will tell you what your
choices are:

```{r }
p=predict(steak.2,steak.new,type="probs")
cbind(steak.new,p)
```

 

I find this hard to read, so I'm going to round off those
predictions. Three or four decimals seems to be sensible:

```{r }
cbind(steak.new,round(p,3))
```

 

Say something about the effect of changing educational level on the
predictions, and say something about the effect of favouring Lottery A
vs.\ not. I don't much mind what: you can say that there is not much
effect (of either variable), or you can say something like ``people
with a graduate degree are slightly more likely to like their steak
rare and less likely to like it well done'' (for education level) and
``people who preferred Lottery A are slightly less likely to like
their steak rare and slightly more likely to like it well done'' (for
effect of Lottery A). You can see these by comparing the first five
rows to assess the effect of education (or the last five rows, if you
prefer), and you can compare eg. rows 1 and 6 to assess the effect of
Lottery A (compare two lines with the *same* educational level
but *different* preferences re Lottery A). 

I would keep away from saying anything about education level 
"less than high school", since this entire level is represented by exactly
one person.
    


(h) Is it reasonable to remove *all* the remaining
explanatory variables from your best model so far? Fit a model with no explanatory variables,
and do a test. (In R, if the right side of the squiggle is a
`1`, that means "just an intercept". Or you can remove
whatever remains using `update`.) What do you conclude?
Explain briefly.


Solution


The fitting part is the challenge, since the testing part is
`anova` again. The direct fit is this:
```{r fiorentina,cache=T}
steak.3=polr(steak_prep_ord~1,data=steak)
```

     

and the `update` version is this, about equally long, starting
from `steak.2` since that is the best model so far:

```{r }
steak.3a=update(steak.2,.~.-educ-lottery_a)
```

 

You can use whichever you like. Either way, the second part is
`anova`, and the two possible answers should be the same:

```{r }
anova(steak.3,steak.2,test="Chisq")
```

 

or

```{r }
anova(steak.3a,steak.2,test="Chisq")
```

 

At the 0.05 level, removing both of the remaining variables is fine:
that is, nothing (out of these variables) has any impact on the
probability that a diner will prefer their steak cooked a particular
way. 

However, with data like this and a rather exploratory analysis, I
might think about using a larger $\alpha$ like 0.10, and at this
level, taking out both these two variables is a bad idea. You could
say that one or both of them is "potentially useful" or
"provocative" or something like that.

If you think that removing these two variables is questionable, you
might like to go back to that `drop1` output I had above:

```{r }
drop1(steak.2,test="Chisq")
```

 

The smallest AIC goes with `lottery_a`, so that comes out (it
is nowhere near significant):

```{r }
steak.4=update(steak.2,.~.-lottery_a)
drop1(steak.4,test="Chisq")
```

 

and what you see is that educational level is right on the edge of
significance, so that may or may not have any impact. Make a call. But
if anything, it's educational level that makes a difference.
    


(i) In the article for which these data were collected,
[link](https://fivethirtyeight.com/datalab/how-americans-like-their-steak/),
does the author obtain consistent conclusions with yours? Explain
briefly. (It's not a very long article, so it won't take you long to
skim through, and the author's point is pretty clear.)



Solution


The article says that *nothing* has anything to do with steak
preference. Whether you agree or not depends on what you thought
above about dropping those last two variables. So say something
consistent with what you said in the previous part. Two points for
saying that the author said "nothing has any effect", and one
point for how your findings square with that.
Now that you have worked through this great long question, this is
where I tell you that I simplified things a fair bit for you! There
were lots of other variables that might have had an impact on how
people like their steaks, and we didn't even consider those. Why did
I choose what I did here? Well, I wanted to fit a regression predicting
steak preference from everything else, do a big backward
elimination, but:
```{r udinese,cache=T, error=T}
steak.5=polr(steak_prep_ord~.,data=steak)
```

   
The `.` in place of explanatory
variables means "all the other variables", including the nonsensical
personal ID. That saved me having to type them all out. 

Unfortunately, however, it didn't work. The problem is a numerical
one. Regular regression has a well-defined procedure, where the computer follows
through the steps and gets to the answer, every time. Once you go
beyond regression, however, the answer is obtained by a step-by-step
method: the computer makes an initial guess, tries to improve it, then
tries to improve it again, until it can't improve things any more, at
which point it calls it good. The problem here is that `polr`
cannot even get the initial guess! (It apparently is known to suck at
this, in problems as big and complicated as this one.)

I don't normally recommend forward selection, but I wonder whether it
works here:

```{r juventus,cache=T}
steak.5=polr(steak_prep_ord~1,data=steak)
steak.6=step(steak.5,scope=.~lottery_a+smoke+alcohol+gamble+skydiving+
speed+cheated+female+age+hhold_income+educ+region,
direction="forward",test="Chisq",trace=0)
drop1(steak.6,test="Chisq")
```

 

It does, and it says the *only* thing to add out of all the
variables is education level. So, for you, I picked this along with a
couple of other plausible-sounding variables and had you start from there.

Forward selection starts from a model containing nothing and asks
"what can we add?". This is a bit more complicated than backward
elimination, because now you have to say what the candidate things to
add *are*. That's the purpose of that `scope` piece, and
there I had no alternative but to type the names of all the
variables. Backward elimination is easier, because the candidate
variables to remove are the ones in the model, and you don't need a
`scope`. The `trace=0` says "don't give me any output"
(you can change it to a different value if you want to see what that
does), and last, the `drop1` looks at what is actually
*in* the final model (with a view to asking what can be removed,
but we don't care about *that* here). 
  



