

??q:steak-data?? This question takes you through the data preparation for one
of the other questions. You don't have to do *this*
question, but you may find it interesting or useful.

When you order a steak in a restaurant, the server will ask
you how you would like it cooked, or to be precise, *how much*
you would like it cooked: rare (hardly cooked at all), through medium
rare, medium, medium well to well (which means "well done", so that
the meat has only a little red to it). Could you guess how a person
likes their steak cooked, from some other information about them? The
website [link](fivethirtyeight.com) commissioned a survey where they
asked a number of people how they preferred their steak, along with as
many other things as they could think of to ask. (Many of the
variables below are related to risk-taking, which was something the
people designing the survey thought might have something to do with
liking steak rare.) The variables of interest are all factors or true/false:



* `respondent_ID`: a ten-digit number identifying each
person who responded to the survey.

* `lottery_a`: true if the respondent preferred lottery A
with a small chance to win a lot of money, to lottery B, with a
larger chance to win less money.

* `smoke`: true if the respondent is currently a smoker

* `alcohol`: true if the respondent at least occasionally
drinks alcohol.

* `gamble`: true if the respondent likes to gamble (eg.
betting on horse racing or playing the lottery)

* `skydiving`: true if the respondent has ever been
skydiving.

* `speed`: true if the respondent likes to drive fast

* `cheated`: true if the respondent has ever cheated on a
spouse or girlfriend/boyfriend

* `steak`: true if the respondent likes to eat steak

* `steak_prep` (response): how the respondent likes their
steak cooked (factor, as described above, with 5 levels).

* `female`: true if the respondent is female

* `age`: age group, from 18--29 to 60+.

* `hhold_income`: household income group, from \$0--24,999
to \$150,000+.

* `educ`: highest level of education attained, from ``less
than high school'' up to "graduate degree"

* `region`: region (of the US)
that the respondent lives in (five values).


The data are in
[link](http://www.utsc.utoronto.ca/~butler/d29/steak.csv). 



(a) Read in the data and display the first few lines.


Solution


The usual:

```{r }
my_url="http://www.utsc.utoronto.ca/~butler/d29/steak.csv"
steak=read_csv(my_url)
steak
```

 



(b) Run the function `complete.cases` on your data
frame. This will produce a lot of output. Pick
three rows where `complete.cases` is TRUE and three where it
is FALSE, and display those rows of the data frame. Looking at your
results, what do you think `complete.cases` does? Explain briefly.


Solution


If you don't believe the whole output is long, well, here it is:

```{r size="footnotesize"}
complete.cases(steak) 
```

 
It doesn't  matter which three TRUE and which three FALSE values you
pick (as long as they *are* TRUE or FALSE as you claim). 

I'll pick rows 496, 497, and 498 for the TRUE rows, and 540, 541 and
542 for the FALSE ones. Let's assemble these rows into a vector and
use `slice` to display the rows with these numbers:

```{r }
rows=c(496,497,498,540,541,542)
rows
```

 

Like this:

```{r }
steak %>% slice(rows)
```

 



(c) What's the difference between the rows where
`complete.cases` is TRUE and the rows where it is false?


Solution


The rows where `complete.cases` is FALSE have one (or more)
missing values in them; where `complete.cases` is TRUE the
rows have no missing values. (Depending on the rows you choose,
you may not see the missing value(s), as I didn't.)
Extra: this is a bit tricky to investigate more thoroughly, because the
text variables might have missing values in them, and they won't
show up unless we turn them into a factor first:
```{r }
steak %>% mutate_if(is.character,factor) %>% 
summary()
```

     

There are missing values everywhere. What the `_if` functions
do is to do something for each column where the first thing is true:
here, if the column is text, then replace it by the factor version of
itself. This makes for a better summary, one that shows how many
observations are in each category, and, more important for us, how
many are missing (a lot).

All right, so there are 15 columns, so let's investigate missingness
in our rows by looking at the columns 1 through 8 and then 9 through
15, so they all fit on the screen. Recall that you can `select`
columns by number:

```{r }
steak %>% select(1:8) %>% slice(rows)
```

 

and

```{r }
steak %>% select(9:15) %>% slice(rows)
```

 

In this case, the first three rows have no missing values anywhere,
and the last three rows have exactly one missing value. This
corresponds to what we would expect, with `complete.cases`
identifying rows that have any missing values.



(d) Obtain a new data frame that contains only the rows without
missing values (from the data frame you read in from the
file). Hints: save the result of `complete.cases` in a variable
first; `filter` can take anything that produces a true or a
false for each row, and will return the rows for which the thing it
was fed was true.


Solution


The hint, together with the previous part, suggests that the thing
fed into `filter` should be based on
`complete.cases(steak)`. Create a variable containing the
results of `complete.cases` first, then use it in the
`filter`: 
```{r cca}
cc=complete.cases(steak)
steak.complete = steak %>% filter(cc) 
```

     

A quick check that we got rid of the missing values:

```{r ccc}
steak.complete
```

 

There are no missing values *there*. Of course, this is not a
proof, and there might be some missing values further down, but at
least it suggests that we might be good.

For proof, this is the easiest way I know:

```{r ccd}
steak.complete %>% mutate_if(is.character,factor) %>%
summary()
```

 

If there were any missing values, they would be listed on the end of
the counts of observations for each level, or on the bottom of the
five-number sumamries. But there aren't.  So here's your proof.



(e) Write the data into a `.csv` file, with a name like
`steak1.csv`.  Open this file in a spreadsheet and (quickly)
verify that you have the right columns and no missing values.


Solution


This is `write_csv`:
```{r }
write_csv(steak.complete,"steak1.csv")
```

     

Open up Excel, or whatever you have, and take a look. You should have
all the right columns, and, scrolling down, no visible missing values.





