##  Feeling similar, again


 Previously, we looked at an experiment about which objects feel similar to one another and which ones feel different.

30 subjects were each
given 17 different objects to feel, for example 
"inner surface of pine  bark", 
"rubber eraser" and 
"nylon scouring pad". The
subjects had to group the objects into a maximum of 8 groups such that
the objects within a group felt similar, and the ones in different
groups felt different.
A
dissimilarity matrix was formed by counting how many subjects put each
pair of objects into *different* groups, so that the
dissimilarity ranged from 0 (the objects were always put together in
the same group) to 30 (the objects were not put together into the same
group by any of the subjects). 

The data are in
[link](http://ritsokiguess.site/datafiles/stimuli.txt). These have
*no* column names.

This time we are going to try *non-metric* multidimensional
scaling, to see whether that produces a more reasonable map. The
reading in of the data is the same as before (so I have reduced the
marks given for it).



(a) Obtain and display the distance matrix that you used last
time for these data. (I don't care whether you run all the code
again, or whether you just produce the matrix from where you had it
before on R Studio. Correct is all I care about.)


Solution


Copied and pasted from last time:
```{r message=F}
my_url <- "http://ritsokiguess.site/datafiles/stimuli.txt"
stimuli <- read_delim(my_url, " ", col_names = F)
objects <- stimuli$X1
names(stimuli) <- c("object", objects)
stimuli
```

     

That gets a data frame with the right column names. Then:

```{r }
d <- stimuli %>% dplyr::select(-1) %>% as.dist()
```

 

and just to check:

```{r size="tiny"}
w <- getOption("width")
options(width = 132)
d
options(width = w)
```

 
    


(b) Obtain a non-metric multidimensional scaling map of the
objects. (No plotting or comments yet.)


Solution


```{r }
stimuli.1 <- isoMDS(d)
```

     

I'm going to remind myself of what this has in it:

```{r }
names(stimuli.1)
```

 

This is rather like running `cmdscale` with `eig=T`: a
thing called `points` with coordinates to plot, and a measure
of fit, here called `stress`, with, as you'd guess, a smaller
`stress` being better.
    


(c) <a name="part:stress">*</a> Obtain a number that says how well the map reproduces the
distances in the data, and comment on that number.


Solution


What you need here is the "stress":
```{r }
stimuli.1$stress
```

     

This is just over 20\%, which is described in the notes as
"poor". We should thus be skeptical about the map that this produces.
    


(d) Plot the non-metric multidimensional scaling map. Label each
point with its (full) object name, sized suitably.


Solution


Make a data frame first of things to plot, the points and the
names. I'm going to plot the names smaller). I am *not*
naming this data frame `d`, in an attempt to avoid
overwriting things I want to keep:
```{r }
stimuli.1.d <- data.frame(stimuli.1$points, names = objects)
stimuli.1.d
```

     

The repeated stimuli down the left are row names, but since they're
row names, we won't be able to use them in
`ggplot`.
`r tufte::margin_note("Next year, I'll make everyone turn matrices  into data frames first so that we won't deal with old-fashioned data frames, row names and the like.")`

Remember that we are turning a matrix and a column into a data frame,
so we need either the more forgiving `data.frame`, or to turn
`points` into a data frame first, which would go like this:

```{r }
as_tibble(stimuli.1$points) %>%
  mutate(name = objects)
```

 
This time, the columns are called `V1` and `V2`, since
that's what `as_tibble` does. Also note the slightly different
look: fewer decimals, since displaying a tibble rounds numerical
things to three significant digits.
Back to my data frame `stimuli.1.d`.
The points have acquired names `X1` and `X2`, as usual,
so we have all we need:

```{r }
ggplot(stimuli.1.d, aes(x = X1, y = X2, label = names)) +
  geom_point() +
  geom_text_repel(size = 2) +
  coord_fixed()
```

 

I think the full names are good with the text this small.

Actually, this came out a lot like the metric multidimensional scaling
that we did earlier.

I was expecting a greater difference. See whether you can find
out what moved, if anything.
    


(e) <a name="part:incons">*</a> Find all the distances in your distance matrix that are 10
or less (there should be three of them). Find these pairs of objects
on your map. Describe where they are on your map. Do
they appear to be the three closest pairs of objects?


Solution


These are:


* `velvet` and `felt` (distance 4). On the map,
close together at the top.

* `block of styrofoam` and `cork` (distance 10). On the
map, close together at the bottom (though they appear to be
slightly farther apart than `rubbereraser` and
`cardboard`). 

* `unglazed ceramic tile` and `brick` (distance
10). On the map, on the left but not especially close together
(for example, `very fine sandpaper` is in between them
and thus closer to both).

I would say that they are not anything like the three closest
pairs of objects on the map. Consider, for example, \texttt{rigid
plastic sheet} and `wax paper`, or \texttt{very fine
sandpaper} and `unglazed ceramic tile`, or \texttt{rubber
eraser} and `cardboard`, all of which are closer together
on the map than the three pairs of objects with distances 10 or
less.
Three points for saying something about the three pairs of objects
of distance 10 or less, and one point for making some relevant
comment about whether these are the three closest pairs on the
map, eg. by finding one pair (or more than one) of objects on the
map that are closer.
    


(f) Consider again your conclusions in parts (<a href="#part:stress">here</a>)
and (<a href="#part:incons">here</a>). Explain briefly how they are consistent.


Solution


The stress is rather high, which means that the map distances and
actual distances may not correspond very well. This is also
exactly what we found in the last part, by finding one (or two)
pairs of objects that were really close in actuality, but not so
much on the map.
The point here is that a high stress *means* that distances
in real life and on the map won't correspond very well. That's
what I want you to say.
Extra: another way to assess this is with a Shepard
`r tufte::margin_note("This always      makes me think of the number 190 (now 985) bus, but that's the wrong      spelling. Spelled like this, this is also the name of the      illustrator of the original pre-Disney Winnie-the-Pooh      stories. See https://en.wikipedia.org/wiki/E.-H.-Shepard, replacing the dashes by underscores.")`
diagram. That would go like this:
```{r }
stimuli.sh <- Shepard(d, stimuli.1$points)
names(stimuli.sh)
```

  

In here, we plot the actual distances `x` against the map
distances `y`, making a data frame first:

```{r }
with(stimuli.sh, data.frame(x = x, y = y)) %>%
  ggplot(aes(x = x, y = y)) + geom_point()
```

 

The actual distances are on the $x$-axis and the map distances are on
the $y$ axis, as far as I can tell (the help file is not clear on
this). But I think I am right, since a lot of the actual distances
were 28 or 29. Notice how the upward trend, such as it is, is very
fuzzy: when the actual distance is very large, the map distance could
be almost anything.
    


(g) Would a three-dimensional map work better than a
two-dimensional one, bearing in mind that a three-dimensional map
will need something like `rgl` to interpret? Calculate
something that will help you decide, and explain what you think.


Solution


The calculation part is to figure out the stress value for a
three-dimensional map, and compare it with the 20\% that we had
before. `k=3` gets the three dimensions:
```{r }
isoMDS(d, k = 3)$stress
```

     

What this does is to obtain the map and throw all of it away except
for the stress value.
`r tufte::margin_note("Which seems like a very SAS way of doing  things.")` 

I think this stress value, just under 10\%, or on the "fair" end of
"good", is a worthwhile improvement over the just-over-20\% that we
had in two dimensions. So I think this is worth the trouble of having
to use a 3-dimensional plotting tool like `rgl` to interpret it
with. You might agree with me, or you might  not: for example, you
might be looking for a bigger improvement. Either is good, as far as
I'm concerned, as long as your answer does something to balance the
smaller stress value
`r tufte::margin_note("Using more dimensions will  *certainly* decrease the stress, in the same way that adding  an $x$-variable to a regression will increase R-squared; it's the  same issue, of whether the change is big enough to be worth having.")` with the difficulty of interpreting it.

The kind of `rgl` code you'll need is something like this (not tested):

```{r eval=F}
library(rgl)
stimuli.2 <- isoMDS(d, k = 3)$points
plot3d(stimuli.2)
text3d(stimuli.2, text = object.abb)
```

 

What you should find, looking at this (try it!) is that objects close
together in this 3-dimensional place are more nearly close together in
actuality as well, because of the in my opinion notably smaller stress
value. 
    



