##  Things that feel similar to each other


 Which objects feel similar to one another and which ones
feel different? Can we draw them on a "map"? 30
subjects^[Probably students in a psychology course. You know  how it goes.] were each
given 17 different objects to feel, for example "inner surface of pine  bark", 
"rubber eraser" and 
"nylon scouring pad". The
subjects had to group the objects into a maximum of 8 groups such that
the objects within a group felt similar, and the ones in different
groups felt different.^[The maximum number of groups was to  ensure that each subject actually *did* group some objects  together, rather than saying that they all feel different.] A
dissimilarity matrix was formed by counting how many subjects put each
pair of objects into *different* groups, so that the
dissimilarity ranged from 0 (the objects were always put together in
the same group) to 30 (the objects were not put together into the same
group by any of the subjects). 

The data are in
[link](http://ritsokiguess.site/datafiles/stimuli.txt). 



(a) Look at the data, and read in the file appropriately. Do you
have something square in shape, apart from any columns of object
names?  Do your columns have the same names as the objects?


Solution


Looking at the file first, there are *no column names*. So we
have to tell `read_delim` that:
```{r stimuli-1 }
my_url <- "http://ritsokiguess.site/datafiles/stimuli.txt"
stimuli <- read_delim(my_url, " ", col_names = F)
stimuli
```

     

I have 17 rows and 18 columns, but one of the columns is the column of
object names. So I really do have the same number of rows and columns
of dissimilarities: that is, it *is* square in shape apart from
the names.

The columns *do not* have the same names as the
objects; R has used the `X1, X2, ...` names that it uses when
you do not have column headers.

I could have supplied the object names to `col_names`, but
that is a lot more work than what we do below.
    
$\blacksquare$

(b) Obtain the names of the objects. Note that they are rather
long. 


Solution


The object names are in the first column, `X1`, of the data frame:
```{r stimuli-2 }
objects <- stimuli$X1
objects
```

     

I'm saving these for later.
    
$\blacksquare$

(c) Set the column names of your data frame to be your object
names, using `names`. Before you rush to do that, figure out
how many column names you need, and supply values for any extra
ones. Check that your data frame now has the right column names.


Solution


I have 18 columns to name (including the column of object names),
but only 17 names, so I need to supply an extra one:
```{r stimuli-3 }
names(stimuli) <- c("object", objects)
stimuli
```

     
    
$\blacksquare$

(d) <a name="part:distances">*</a>  
Convert your data frame into a distance object. Take a look at the
(rather long) result.


Solution


This is `as.dist`, since we have distances
(dissimilarities) already. Don't forget to take off the first column!
```{r stimuli-4 }
d <- stimuli %>% select(-1) %>% as.dist()
```

     

I can try and show it all here, tiny, but even then it's long because
the column names are long:

```{r stimuli-5, size="tiny"}
w <- getOption("width")
options(width = 132)
d
options(width = w)
```

 

The stuff with `width` was to make it display lots of columns,
and then setting it back afterwards so as not to mess things up later.
If
you try and take `head` of this, you'll lose the structure. I
don't know of a good way to display part of one of these.
    
$\blacksquare$

(e) Obtain and plot a  (metric) multidimensional scaling map of
these data.  Label the points with the name of the
object they represent. (Note that
`geom_text_repel` has an option `size` that controls
the size of the text.)


Solution


This is the procedure. Talking about it is coming in a minute.
```{r stimuli-6 }
d.1 <- cmdscale(d, 2)
data.frame(d.1, stim = objects) %>%
  ggplot(aes(x = X1, y = X2, label = stim)) + geom_point() +
  geom_text_repel(size = 2)
```

     

`cmdscale` gets the coordinates to plot, then we plot them, and
then we place the object names next to the points. I'm not
quite sure what the scale is for `size`, but `size=2`
worked for me here, making the text a bit smaller (so that the labels
don't overlap), but not so small that you can't read it. You'll
probably have to experiment to find a `size` that you like.

If you forget the `2` after the distance matrix in
`cmdscale`, you'll get a two-dimensional solution anyway (since
two dimensions is the default). The output is an array of coordinates
in two dimensions:

```{r stimuli-7 }
d.1
```

 
I note that any map that is the same as this apart from rotation and
reflection is also fine (since the distances will be the same). I saw
a lot of maps that were like mine but upside down (with `cork`
at the top).^[I lie. Last year, *I* got cork at the top,  and a lot of other people got cork at the bottom as you see here.] No problem there.
    
$\blacksquare$

(f) Find a pair of objects that are close together on your
plot. Look back at your answer to part~(<a href="#part:distances">here</a>): is
the distance between those objects small? Explain briefly.


Solution


I don't mind (much) which objects you pick (as long as they are
reasonably close together). Find the actual distance between them
from what I called `d` in part~(<a href="#part:distances">here</a>). Some
possibilities:


* Felt and velvet, Distance 4.

* Block of styrofoam and rubber eraser. Distance 19 (not that
small, but one of the smaller ones).

* Rigid plastic sheet and
wax paper. Distance 12. Smallish.

* unglazed ceramic
tile^[The first association that *unglazed* made for me      was *donut*!] and very fine sandpaper. Distance 12. Smallish.

There won't be a perfect relationship between distance in the
distance matrix and on the map. In this case, there is an upper
limit on distance (30, because 30 people rated the objects for
similarity) and that upper limit is approached by many of the
distances. (This suggests that non-metric multidimensional
scaling, that just reproduces the order of the distances, might be
a better idea for these data.) If objects A and B, and also B and
C, are each close to 30 apart, then objects A and C will also be
close to 30 apart, and that constrains them to be nearly in a
triangle on the map. 
There are some 10s in the distance matrix,
for example between block of styrofoam and cork, and also
between unglazed ceramic tile and brick; these are a bit
further apart on the map, but still close.
    
$\blacksquare$

(g) Obtain a measure of goodness of fit for this
multidimensional scaling map.


Solution


This means fitting again, but this time with `eig=T`, and
pulling out the thing called `GOF`. You can omit the 2,
since that's the default 2 dimensions:
```{r stimuli-8 }
d.2 <- cmdscale(d, 2, eig = T)
d.2$GOF
```

     

I didn't ask you to comment on this, but the adjective that came to my
mind was "disappointing". I think that's to do with the
upper-limit-30 thing again. Also, this time (unlike with Wisconsin)
there was no "right answer", so maybe it just isn't going to be very
good. If you looked at several pairs of points above, you might have
noticed that the correspondence between map distance and actual
distance isn't very good; this is the same issue.

I was curious about whether 3 dimensions would be any better:

```{r stimuli-9 }
d.2a <- cmdscale(d, 3, eig = T)
d.2a$GOF
```

 

That is quite a bit better. The problem with this, though, is that we
need something like `rgl` to explore it with.

$\blacksquare$

(h) Obtain a cluster analysis for the same data, using Ward's
method. Obtain a dendrogram. What seems to be an appropriate number
of clusters?  Mark them on your dendrogram too.


Solution


This seems to be a throwback to last week, but I have my reasons,
which you'll see in a moment:
```{r stimuli-10 }
d.3 <- hclust(d, method = "ward.D")
plot(d.3)
```

     

That looks like 5 clusters to me (chopping the tree at a height of
30). Drawing them:

```{r stimuli-11 }
plot(d.3)
rect.hclust(d.3, 5)
```

 
    
$\blacksquare$

(i) Discuss as briefly as seems reasonable whether your
clusters tell the same story as the map that came from
multidimensional scaling. (There were only three marks for this, so you
don't need to go overboard.) Optionally, think about creating a plot
that will make it easier to see the correspondence between your
clusters and the MDS map.


Solution


I would take my clusters and think about whether the objects in
them are (a) close together on the map and (b) distinct from other
objects on the map. Working from left to right of my dendrogram:


* Together at the bottom of the plot.

* Together at the top left.

* Together at the top left, but rather mixed up with the
previous cluster (in particular, `nylonscouringpad` looks as if it
belongs in the previous cluster).

* Together at the top of the plot.

* Together at the top right.

If you had a different number of clusters, work with what you
have, and if you have done that, you'll be good. My general
conclusion from the above is that my five clusters are mostly
distinct from each other, but the MDS map really only has
four. (If you look at the dendrogram, the two clusters that are
not really distinct from each other were the last ones to be split.)
This is a lot more detail than you need. What I want to see is
some kind of association of clusters with (hopefully) nearby
points on the map, some kind of discussion of whether the clusters
are really distinct groupings on the map, and a sense of whether
you feel the clusters are consistent with what appears on the map.
Extra: this is where I have to say I cheated. I thought this would be
hard to do by trying to match  those names in the clusters
with the ones on the MDS map. So I put myself through some
short-term R coding pain for some substantial long-term gain.
I was thinking, "can I extract the clusters from this    analysis, and plot them on the MDS map in different colours?" That
would go like this:
```{r stimuli-12 }
clusters <- cutree(d.3, 5)
data.frame(d.1, names = stimuli[, 1], cluster = factor(clusters)) %>%
  ggplot(aes(x = X1, y = X2, label = objects, colour = cluster)) +
  geom_point() +
  geom_text_repel(size = 2)
```

     

The first line obtains the clusters as numbers, which I then had to
turn into a factor (to make the plot different colours). 
I copied the rest from
above, and then I added to them, colouring the points according to
what cluster they belonged to. Three of my "rough" objects have
broken off into a cluster of their own; they are, kind of, in their
own little area on the map.

Another, non-graphical, way of doing this is to list the MDS coordinates
along with the clusters, probably sorted by cluster:

```{r stimuli-13 }
data.frame(d.1, object = stimuli[, 1], clusters) %>% arrange(clusters)
```

 

Then you can think about what makes the clusters different in terms of
`X1` and `X2`. For me, clusters 1 and 2 are kind of
mixed up, with `X1` and (usually) `X2` negative; cluster
3 has strongly positive `X2`; cluster 4 has very
*negative* `X2`;^[Oxford semicolon, for the grammar mavens among you.] 
and cluster 5 has strongly positive `X1`.

I wonder whether our three-dimensional solution distinguishes clusters
1 and 2 at all? Same approach again:

```{r stimuli-14 }
save.3d <- data.frame(d.2a$points, objects, clusters) %>%
  arrange(clusters)
save.3d
```

 

It looks as if the combo of negative `X1` and positive
`X3` distinguishes cluster 2 from cluster 1.

As I was writing this, I was thinking that we should throw the
coordinates and the `clusters` into a
*discriminant analysis*, to find out what distinguishes the
clusters. Which is way too weird, and therefore deserves to be
explored (with the three-dimensional solution, for example):

```{r stimuli-15 }
library(MASS)
save.3d.lda <- lda(clusters ~ X1 + X2 + X3, data = save.3d)
ppp <- predict(save.3d.lda)
```

 

I snuck a look at the output and
found that `LD3` is basically worthless, so I can plot
`LD1` against `LD2`, coloured by cluster:

```{r stimuli-16 }
data.frame(ppp$x, cluster = factor(save.3d$clusters)) %>%
  ggplot(aes(x = LD1, y = LD2, colour = cluster)) +
  geom_point()
```

     

This says that the three-dimensional MDS *has* separated clusters
1 and 2, and if we were to plot `d.2a` in `rgl` and
rotate it the right way, we *would* be able to see a difference
between those two clusters as well. (They wouldn't look mixed up as
they do on the two-dimensional map.)

So we can look at the three-dimensional map coordinates and the
discriminant analysis and ask "what distinguishes the clusters?". 
Map coordinates first. I need `points` since I ran
the scaling with `eig=T`:

```{r stimuli-17 }
save.3d
```

 

and the business end of the LDA output:

```{r stimuli-18 }
save.3d.lda$svd
save.3d.lda$scaling
```

 

The plot said that cluster 2 was lowest of all on `LD1`, a  bit
lower than cluster 1. What would make `LD1` small (negative)
would be if `X1` was small and `X2` and `X3` were
large. The cluster 2 observations are the smallest on `X1`
(smaller than cluster 1) and larger on `X3` than cluster 1. So
we *can* enumerate what makes clusters 1 and 2 different. The
plot of the first two `LD`s says, in fact, that under the
3-dimensional multidimensional scaling, all five groups are distinct.

A biplot would be another way to look at that:

```{r stimuli-19 }
ggbiplot(save.3d.lda, groups = factor(save.3d$clusters))
```

 
Points on the right have `X1` and `X2` large, and
`X3` small (cluster 4, and to a lesser extent, cluster
5). Points on the left are the other way around. None of the arrows
point up or down really, but `X1` points up a bit and
`X2` down a bit, so points at the top of the plot are likely to
be high on `X1` and low on `X2`, like cluster 5.

Was that all confusing enough for you?

Anyway, the key to gaining some insight here is to find a way to
*combine* the 
output from the cluster analysis and the multidimensional
scaling. That way you can see both in the same place, and see how they
compare. 
    
$\blacksquare$





