##  Who studies the most outside class?


 A social scientist wanted to see how many hours students
 studied outside of class. She took a random sample of 75 students
 from three different majors: math, social science and English, and
 recorded the number of weekly outside-class study hours for each
 student. The data can be found at
 [link](http://ritsokiguess.site/datafiles/studyhours.txt). The
 social scientist had two research questions: whether math students study
 more outside of class than the other students, and whether English and
 social science students study different amounts of time outside class.


(a) Explain briefly why contrasts would be a better idea here
than ANOVA with Tukey.


Solution


The researcher is not interested in all the comparisons between
the groups, but rather in two specific ones (as detailed in the
question). So we should focus our attention on those
comparisons. This is what contrasts are for.

$\blacksquare$

(b) Read in the data and make side-by-side boxplots of study
hours by major. What do the boxplots suggest about the comparisons
that the researcher wants to make?


Solution


Separated by one space. There appear to be some strange quotes
in there, but we'll ignore those and see whether they cause any
trouble: 
```{r }
my_url <- "http://ritsokiguess.site/datafiles/studyhours.txt"
studyhours <- read_delim(my_url, " ")
studyhours
```

        

So far so good. 75 students, in tidy format.

```{r }
ggplot(studyhours, aes(x = major, y = hours)) + geom_boxplot()
```

 

This suggests that the math students study more than the others, but
the English and social science students study about the same amount.

I see some rather consistent right-skewness here (but no outliers). We
have 25 observations in each group, a decent amount of data, so I'm
not too worried, but a transformation could be a good idea, something
like square root, perhaps. The spreads are not that unequal, but it
*is* true that the Math students have both the largest median
*and* the largest spread, and the social science students are the
lowest on both.

Extra: Box-Cox also works on ANOVA-type data, since that is an `lm`
kind of model:

```{r }
boxcox(hours ~ major, data = studyhours)
```

 

This suggests that some kind of transformation would be a good idea
(no transformation, power 1, is not supported by the data). My guess
of 0.5 is not bad; that, or something a bit less like 0.25 (fourth
root) would be good. Even log is supportable. But we'll proceed
without transformation in this question to give you some practice with
contrasts. Despite appearances, I do (sometimes) like to make things
not so complicated!

$\blacksquare$

(c) Set up contrasts for each of the researcher's research
questions, bearing in mind the levels of `major` and the
order in which they come. (For the `Math` contrast, you want
Math against the average of the other two.)


Solution


We need to know what the categories of `major` are, so that
we can set up the contrasts correctly, so let's start with that:
```{r }
studyhours %>% select(major) %>% summary()
```

 

Oh. Yeah. We never turned `major` into a factor. We'd better do
that, and then try again:

```{r }
studyhours <- studyhours %>% mutate(major = factor(major))
studyhours %>% select(major) %>% summary()
```

 

In alphabetical order.

All right, contrasts. The English vs.\ social science one is easier,
so we'll do that first. Just set it up with a 1 for one of them and a
$-1$ for the other and 0 for math, in the right order according to
what the levels are, viz:

```{r }
c.eng.socsci <- c(1, 0, -1)
```

 

Or `c(-1,0,1)` would be just as good.

The other one is a bit trickier, because you want math against the
average of the others, like
$$\mbox{math}-(\mbox{English}+\mbox{socsci})/2.$$ 
This translates into contrast-ese like this, making sure to get Math
in the middle where it belongs:
`r tufte::margin_note("As I failed to do the first time.")`

```{r }
c.math.others <- c(-0.5, 1, -0.5)
```

 

The coefficients have to add up to zero, so I have to have two halves
to balance my one 1. Or I can use two $-1$s and a 2.

$\blacksquare$

(d) Verify that your two contrasts are orthogonal.


Solution


Multiply them together (elementwise, which is how R does it)
and show that what you get adds up to zero:
```{r }
sum(c.eng.socsci * c.math.others)
```

 

Zero. Orthogonal.

So we are safely in "familiar" territory, not in the 
here-be-dragons
`r tufte::margin_note("On ancient maps, people didn't know what was  in certain parts of the world, because no-one had ever explored  them, so they wrote on the map *here be dragons*.")` land of
non-orthogonal contrasts.

$\blacksquare$

(e) Create a matrix of contrasts (using `cbind`) and let
`lm` know that these are contrasts for `major`, in
data frame `studyhours` (or whatever you called it).


Solution


So, like the example in lecture (where the
`contrasts(model)` was admittedly rather confusing):
```{r }
m <- cbind(c.math.others, c.eng.socsci)
m
contrasts(studyhours$major) <- m
```

        

$\blacksquare$

(f) Fit the model using `lm`. What do you conclude about
your contrasts?


Solution


Pretend the ANOVA is a regression (though you can fit ANOVAs
either way: we just used `aov` before because we were
going to follow up with Tukey):
```{r }
studyhours.1 <- lm(hours ~ major, data = studyhours)
summary(studyhours.1)
```

        

There is a seriously significant difference between the math students
and the others. There is not, however, any kind of difference between
the mean study hours of the English and social science students. 

$\blacksquare$

(g) Are you surprised by the significance or non-significance you
found in the previous part? Explain briefly.


Solution


Go back and compare with the  boxplots. The math students
studied noticeably more than the others, so it is not
surprising that this difference came out significant. The
median study times for the English and social science students
were similar, so it is not at all surprising that this difference
failed to be significant.

When you set up contrasts, *these* are the tests you are
interested in, so there is no reason for following up with
Tukey or anything else. But you have to be able to say ahead of
time which contrasts you want to test. This is in
contrast
`r tufte::margin_note("In contrast. Get it? No? Well, never mind then.")` to Tukey, where you don't have to decide
which comparisons interest you until right at the end.

Another question you might have had is 
"how many contrasts can I do"? To answer this, go back and look at the ANOVA table. If
you run an ANOVA using `lm`, you get the ANOVA table by
passing the fitted model object into `anova` rather than
`summary`:
`r tufte::margin_note("*anova* is one of R's  multi-purpose tools; what it does depends on what you feed it.")`
```{r }
anova(studyhours.1)
```

        

Two degrees of freedom for `major`, so we can do 2
contrasts. (The example in class had four models of chainsaw, so 3 df
for model, so 3 contrasts possible.) What if we don't use all those
contrasts up? Try it and see:

```{r }
m <- cbind(c.math.others)
m
contrasts(studyhours$major) <- m
studyhours.2 <- lm(hours ~ major, data = studyhours)
summary(studyhours.2)
```



The two P-values are identical, but the second one is not now
identified with any contrast (so we just ignore it). If we were only
interested in how math students compared with the rest, we could test
it this way.

Or, we could relabel the students as "math" and "other" and just
do a two-sample $t$-test. To do this, we need to delve into
`forcats` (which is a part of the `tidyverse`). It deals
with categorical variables which are genuine factors, which we haven't
had much cause to do so far. But now we do. What we want to do is to
*recode* the levels of `major`, which is done by
`fct_recode`:

```{r }
studyhoursx <- studyhours %>%
  mutate(mathrest = fct_recode(major,
    "rest" = "english",
    "rest" = "socsci"
  ))
studyhoursx %>% count(mathrest)
```

 

The way `fct_recode` works is that you first give a factor,
here `major`, then you define a list of new levels in terms of
the old ones. We want a student whose major is either English or
Social Science to end up in the new not-math group that we'll call
`rest`, so we need two lines inside `fct_recode`, one
that says that the new `rest` group includes the old
`english` majors, and another that says that `rest` also
includes the old `socsci` majors. Any levels not mentioned
(here `math`) are left unchanged.

I did this wrong the first time: I got the things before and after the
`=` in `fct_recode` the wrong way around.

The last line is to check that the result makes sense. I previously
had 25 students in each major. Now I have 25 students for whom
`mathrest` is `math` (the math majors) and $50=25+25$
students for whom `mathrest` is `rest` (everyone else). 

Now, the $t$-test.

```{r }
t.test(hours ~ mathrest, data = studyhoursx, var.equal = T)
```

 

This is very close to, but not the same as, the test with the contrasts,
because we are not quite testing the same thing. Here, we are
comparing the mean of the math students to the mean of everyone else,
thrown together into one big bag, but with the contrasts, we are
comparing the mean of the math students with the \emph{average of the
means} of the other two groups, treating them as different. (Not
clear? Well, it's not that clear to me either.)

That business with the `var.equal=T`? This is doing a pooled
$t$-test. I did that here because the ANOVA is assuming equal spreads
within the groups (which is what a pooled $t$-test does) and I thought
I should do my best to be consistent. A pooled $t$-test is really a
two-group one-way ANOVA, or at least it is if it is done two-sided.

$\blacksquare$
 

