##  Clustering the Swiss bills


 This question is about the Swiss bank counterfeit bills
again. This time we're going to ignore whether each bill is
counterfeit or not, and see what groups they break into. Then, at
the end, we'll see whether cluster analysis was able to pick out the
counterfeit ones or not.


(a) Read the data in again (just like last time), and look at
the first few rows. This is just the same as before.
 
Solution


The data file was aligned in columns, so:
```{r }
my_url <- "http://www.utsc.utoronto.ca/~butler/d29/swiss1.txt"
swiss <- read_table(my_url)
swiss
```

       
 

(b) The variables in this data frame are on different
scales. Standardize them so that they all have mean 0 and standard
deviation 1. (Don't try to standardize the `status` column!)
 
Solution


```{r }
swiss.s <- swiss %>%
  select(-status) %>%
  scale()
```

    

What kind of thing do we have?

```{r }
class(swiss.s)
```

 

so something like this is needed to display some of it (rather than
all of it):

```{r }
head(swiss.s)
```

 
 

(c) We are going to make a scree plot. First, calculate the total
within-cluster SS for each number of clusters from 2 to 10.
 
Solution

 
When I first made this problem (some years ago),
I thought the obvious answer was a loop, but now that I've been
steeped in the Tidyverse a while, I think `rowwise` is much
clearer, so I'll do that first.
Start by making a `tibble` that has one column called `clusters` containing the numbers 2 through 10:

```{r }
tibble(clusters = 2:10)
```

 
Now, for each of these numbers of clusters, calculate the total within-cluster sum of squares for *it* (that number of clusters). To do that, think about how you'd do it for something like three clusters:

```{r }
kmeans(swiss.s, 3, nstart = 20)$tot.withinss
```

 

and then use that within your `rowwise`:
```{r }
tibble(clusters = 2:10) %>%
  rowwise() %>% 
  mutate(wss = kmeans(swiss.s, clusters, nstart = 20)$tot.withinss) -> wssq
wssq
```

 

Another way is to save *all* the output from the `kmeans`, in a list-column, and then *extract* the thing you want, thus:
```{r }
tibble(clusters = 2:10) %>%
  rowwise() %>% 
  mutate(km = list(kmeans(swiss.s, clusters, nstart = 20))) %>%
  mutate(wss = km$tot.withinss) -> wssq.2
wssq.2
```

The output from `kmeans` is a collection of things, not just a single number, so when you create the column `km`, you need to put `list` around the `kmeans`, and then you'll create a list-column. `wss`, on the other hand, is a single number each time, so no `list` is needed, and `wss` is an ordinary column of numbers, labelled `dbl` at the top. 

The most important thing in both of these is to remember the `rowwise`. Without it, everything will go horribly wrong! This is because `kmeans` expects a *single number* for the number of clusters, and `rowwise` will provide that single number (for the row you are looking at). If you forget the `rowwise`, the whole column `clusters` will get fed into `kmeans` all at once, and `kmeans` will get horribly confused.
 
 
If you insist, do it Python-style as a loop, like this:        
```{r }
clus <- 2:10
wss.1 <- numeric(0)
for (i in clus)
{
  wss.1[i] <- kmeans(swiss.s, i, nstart = 20)$tot.withinss
}
wss.1
```

       

Note that there are 10 `wss` values, but the first one is
missing, since we didn't do one cluster.
`r tufte::margin_note("R vectors start from  1, unlike C arrays or Python lists, which start from 0.")`

The `numeric(0)` says "`wss` has nothing in it, but if it had anything, it would be numbers". Or, you can initialize
`wss` to however long it's going to be (here 10), which is
actually more efficient (R doesn't have to keep making it 
"a bit longer"). If you initialize it to length 10, the 10 values will have
`NA`s in them when you start.
It doesn't matter what `nstart` is: Ideally, big enough to have a decent
chance of finding the best clustering, but small enough that it
doesn't take too long to run.
Whichever way you create your total within-cluster sums of squares, you can use it to make a scree plot (next part). 

 

(d) <a name="part:howmany">*</a> Make a scree plot (creating a data frame
first if you need). How many clusters do you think you
should use?
 
Solution


The easiest is to use the output from the `rowwise`,
which I called `wssq`, this already being a dataframe:

```{r error=T}
ggplot(wssq, aes(x = clusters, y = wss)) + geom_point() + geom_line()
```

       
If you did it the loop way, you'll have to make a data frame
first, which you can then pipe into `ggplot`:

```{r sasjhgajs,fig.height=6,fig.width=6, error=T}
tibble(clusters = 1:10, wss = wss.1) %>%
  ggplot(aes(x = clusters, y = wss)) + geom_point() + geom_line()
```

       
If you started at 2 clusters, your `wss` will start at 2
clusters also, and you'll need to be careful to have something like
`clusters=2:10` (not `1:10`) in the definition of your
data frame. 

Interpretation: I see a small elbow at 4 clusters, so that's how many I think we
should use. Any place you can reasonably see an elbow is good.

The warning is about the missing within-cluster total sum of squares
for one cluster, since the loop way didn't supply a total
within-cluster sum of squares for one cluster.
 

(e) Run K-means with the number of clusters that you found in
(<a href="#part:howmany">here</a>). How many bills are in each cluster?
 
Solution



I'm going to start by setting the random number seed (so that
my results don't change every time I run this). You don't need
to do that, though you might want to in something like R
Markdown code (for example, in an R Notebook):
```{r }
set.seed(457299)
```



Now, down to business:
```{r }
swiss.7 <- kmeans(swiss.s, 4, nstart = 20)
swiss.7$size
```

       
This many. Note that my clusters 1 and 4 (and also 2 and 3) add up to
100 bills. There were 100 genuine and 100 counterfeit bills in the
original data set.
I don't know why "7". I just felt like it.
Extra: you might remember that back before I actually *ran*
K-means on each of the numbers of clusters from 2 to 10. How can we
extract that output? Something like this. Here's where the output was:

```{r }
wssq.2
```

 

Now we need to pull out the 4th row and the `km` column. We need the output as an actual thing, not a data frame, so:

```{r }
wssq.2 %>%
  filter(clusters == 4) %>%
  pull(km) -> swiss.7a
```

 

Is that the right thing?

```{r }
swiss.7a
```

 

Looks like it. But I should check:

```{r }
swiss.7a$centers
```

 

Ah. `swiss.7a` is actually a `list`, as evidenced by the `[[1]]` at the top of the output, so I get things from it thus:

```{r }
swiss.7a[[1]]$centers
```

 

This would be because it came from a list-column; using `pull` removed the data-frameness from `swiss.7a`, but not its listness.
 

(f) Make a table showing cluster membership against actual
status (counterfeit or genuine). Are the counterfeit bills mostly
in certain clusters?
 
Solution


`table`. `swiss.7$cluster` shows the actual
cluster numbers:
```{r }
table(swiss$status, swiss.7$cluster)
```

       

Or, if you prefer,

```{r }
tibble(obs = swiss$status, pred = swiss.7$cluster) %>%
  count(obs, pred)
```

 

or even

```{r }
tibble(obs = swiss$status, pred = swiss.7$cluster) %>%
  count(obs, pred) %>%
  pivot_wider(names_from = obs, values_from = n, values_fill = 0)
```

 

In my case (yours might be different), 99 of the 100 counterfeit bills
are in clusters 1 and 4, and 99 of the 100 genuine bills are in
clusters 2 and 3.
`r tufte::margin_note("This is again where *set.seed* is  valuable: write this text once and it never needs to change.")` So the
clustering has done a very good job of distinguishing the genuine
bills from the counterfeit ones. (You could imagine, if you were an
employee at the bank, saying that a bill in cluster 1 or 4 is
counterfeit, and being right 99\% of the time.) This is kind of a
by-product of the clustering, though: we weren't *trying* to
distinguish counterfeit bills (that would have been the discriminant
analysis that we did before); we were just trying to divide them into
groups of different ones, and part of what made them different was
that some of them were genuine bills and some of them were
counterfeit.
 


