##  Telling whether a banknote is real or counterfeit


 <a name="sec:swiss-money">*</a> A Swiss bank collected  a number of known counterfeit
(fake) 
bills over time, and sampled a number of known genuine bills of the
same denomination.
Is it possible to tell, from measurements taken from a bill, whether
it is genuine or not? We will explore that issue here. The variables
measured were:


* length

* right-hand width

* left-hand width

* top margin

* bottom margin

* diagonal



(a) Read in the data from
[link](http://ritsokiguess.site/datafiles/swiss1.txt), and
check that you have 200 rows and 7 columns altogether.
 
Solution


Check the data file first. It's aligned in columns, thus:
```{r swiss-money-1 }
my_url <- "http://ritsokiguess.site/datafiles/swiss1.txt"
swiss <- read_table(my_url)
swiss
```

       

Yep, 200 rows and 7 columns.
 
$\blacksquare$

(b) Run a multivariate analysis of variance. What do you
conclude? Is it worth running a discriminant analysis? (This is
the same procedure as with basic MANOVAs before.) 
 
Solution


Small-m `manova` will do here:
```{r swiss-money-2 }
response <- with(swiss, cbind(length, left, right, bottom, top, diag))
swiss.1 <- manova(response ~ status, data = swiss)
summary(swiss.1)
```

       

Extra: you might be wondering whether you had to go to all that trouble to
make the response variable. Would this work?

```{r swiss-money-3, error=T}
response2 <- swiss %>% select(length:diag)
swiss.1a <- manova(response2 ~ status, data = swiss)
```

 

No, because `response2` needs to be an R `matrix`, and it isn't:

```{r swiss-money-4 }
class(response2)
```

 

The error message was a bit cryptic (nothing unusual there), but a
data frame (to R) is a special kind of `list`, so that R didn't
like `response2` being a data frame, which it
thought was a list.

This, however, works, since it turns the data frame into a matrix:

```{r swiss-money-5 }
response4 <- swiss %>% select(length:diag) %>% as.matrix()
swiss.2a <- manova(response4 ~ status, data = swiss)
summary(swiss.2a)
```

 
Anyway, the conclusion: the status of a bill (genuine or counterfeit)
definitely has an influence on some or all of those other variables,
since the P-value $2.2 \times 10^{-16}$ (or less) is really small. So
it is worth running a discriminant analysis to figure out where the
differences lie.

As a piece of strategy, for creating the response matrix, you can
always either use `cbind`, which creates a `matrix`
directly, or you can use `select`, which is often easier but
creates a data frame, and then turn *that* into a `matrix`
using `as.matrix`. As long as you end up with a
`matrix`, it's all good.
 
$\blacksquare$

(c) Run a discriminant analysis. Display the output.
 

Solution

 
Now we forget about all that
`response` stuff. For a discriminant analysis, the
grouping variable (or combination of the grouping variables)
is the "response", and the quantitative ones are
"explanatory":
```{r swiss-money-6 }
swiss.3 <- lda(status ~ length + left + right + bottom + top + diag, data = swiss)
swiss.3
```

       
 
$\blacksquare$

(d) How many linear 
discriminants did you get? Is that making sense? Explain briefly.
 
Solution


I got one discriminant, which makes sense because there are two
groups, and the smaller of 6 (variables, not counting the grouping
one) and $2-1$ is 1. 
 
$\blacksquare$

(e) <a name="part:big">*</a> 
Using your output from the discriminant analysis, describe how
each of the linear discriminants that you got is related to your
original variables. (This can, maybe even should, be done crudely:
"does each variable feature in each linear discriminant: yes or no?".)
 
Solution


This is the Coefficients of Linear Discriminants. Make a call about whether each of those coefficients is close to zero (small in size compared to the others), or definitely positive or definitely negative.
These are judgement calls: either you can say that LD1
depends mainly on `diag` (treating the other coefficients
as "small" or close to zero), or you can say that `LD1`
depends on everything except `length`.
 
$\blacksquare$

(f) What values of your variable(s) would make `LD1`
large and positive?
 
Solution


Depending on your answer to the previous part: 
If you said that only `diag` was important, `diag`
being large would make `LD1` large and positive.
If you said that everything but `length` was important,
then it's a bit more complicated: `left` and
`diag` large, `right`, `bottom` and
`top` small (since their coefficients are negative). 
 
$\blacksquare$

(g) <a name="part:means">*</a> Find the means of each variable for each group (genuine
and counterfeit bills). You can get this from your fitted linear
discriminant object.
 
Solution


```{r swiss-money-7 }
swiss.3$means
```

 
 
$\blacksquare$

(h) Plot your linear discriminant(s), however you like. Bear in
mind that there is only one linear discriminant.
 
Solution


With only one linear discriminant, we can plot `LD1` scores on
the $y$-axis and the grouping variable on the $x$-axis. How
you do that is up to you. 

Before we start, though, we need the `LD1` scores. This means
doing predictions. The discriminant scores are in there. We take the
prediction output and make a data frame with all the things in the
original data. My current preference (it changes) is to store the
predictions, and then `cbind` them with the original data,
thus:

```{r swiss-money-8 }
swiss.pred <- predict(swiss.3)
d <- cbind(swiss, swiss.pred)
head(d)
```

      

I needed `head` because `cbind` makes an old-fashioned
`data.frame` rather than a `tibble`, so if you display
it, you get all of it.  

This gives the LD1 scores, predicted groups, and posterior
probabilities as well. That saves us having to pick out the other
things later.
The obvious thing is a boxplot. By examining `d` above (didn't
you?), you saw that the LD scores were in a column called
`LD1`:
```{r antioch, error=T}
ggplot(d, aes(x = status, y = LD1)) + geom_boxplot()
```

   

This shows that positive LD1 scores go (almost without exception) with
genuine bills, and negative ones with counterfeit bills.
It also shows that there are three outlier bills, two counterfeit ones
with unusually high LD1 score, and one genuine one with unusually
*low* LD1 score, at least for a genuine bill.

Or you could do faceted histograms of `LD1` by `status`:

```{r swiss-money-9 }
ggplot(d, aes(x = LD1)) + geom_histogram(bins = 10) + facet_grid(status ~ .)
```

 

This shows much the same thing as `plot(swiss.3)` does (try it).
 
$\blacksquare$

(i) What kind of score on `LD1` do genuine bills
typically have? What kind of score do counterfeit bills typically
have? What characteristics of a bill, therefore, would you look at
to determine if a bill is genuine or counterfeit?
 
Solution


The genuine bills almost all have a *positive* score on
LD1, while the counterfeit ones all have a *negative* one. 
This means that the genuine bills (depending on your answer to
(<a href="#part:big">here</a>)) have a large `diag`, or they have a
large `left` and `diag`, and a small
`right`, `bottom` and `top`.
If you look at your table of means in (<a href="#part:means">here</a>), you'll
see that the genuine bills do indeed have a large `diag`,
or, depending on your earlier answer, a small `right`,
`bottom` and `top`, but not actually a small
`left` (the `left` values are very close for the
genuine and counterfeit coins).

Extra: as to that last point, this is easy enough to think about. A
boxplot seems a nice way to display it:
```{r gtabita}
ggplot(d, aes(y = left, x = status)) + geom_boxplot()
```

      

There is a fair bit of overlap: the median is higher for the
counterfeit bills, but the highest value actually belongs to a genuine one.

Compare that to `diag`:

```{r iggle}
ggplot(d, aes(y = diag, x = status)) + geom_boxplot()
```

 

Here, there is an almost complete separation of the genuine and
counterfeit bills, with just one low outlier amongst the genuine bills
spoiling the pattern.
I didn't look at the predictions (beyond the discriminant scores),
since this question (as set on an assignment a couple of years ago)
was already too long, but there is no difficulty in doing so.
Everything is in the data frame I called `d`:

```{r swiss-money-10 }
with(d, table(obs = status, pred = class))
```

 

(this labels the rows and columns, which is not necessary but is nice.)

The `tidyverse` way is to make a data frame out of the actual
and predicted statuses, and then `count` what's in there:

```{r swiss-money-11 }
d %>% count(status, class)
```

 

This gives a "long" table, with frequencies for each of the
combinations for which anything was observed.

Frequency tables are usually wide, and we can make this one so by pivot-wider-ing `pred`:

```{r swiss-money-12 }
d %>%
  count(status, class) %>%
  pivot_wider(names_from = class, values_from = n)
```

 
One of the genuine bills is incorrectly classified as a counterfeit
one (evidently that low outlier on LD1), but every single one of the
counterfeit bills is classified correctly. That missing value is
actually a frequency that is zero, which you can fix up thus:

```{r swiss-money-13 }
d %>%
  count(status, class) %>%
  pivot_wider(names_from = class, values_from = n, values_fill = 0) 
```

 

which turns any missing values into the zeroes they should be in this
kind of problem.
It would be interesting to see what the posterior probabilities look
like for that misclassified bill:

```{r swiss-money-14 }
d %>% filter(status != class)
```

 

On the basis of the six measured variables, this looks a lot more like
a counterfeit bill than a genuine one.
Are there any other bills where there is any doubt? One way to find out is to find the maximum of the two posterior probabilities. If this is small, 
there is some doubt about whether the bill is real or fake. 0.99 seems like a very stringent cutoff, but let's try it and see:

```{r swiss-money-15, error=T}
d %>%
  mutate(max.post = pmax(posterior.counterfeit, posterior.genuine)) %>%
  filter(max.post < 0.99) %>%
  dplyr::select(-c(length:diag))
```

 
The only one is the bill that was misclassified: it was actually genuine, but
was classified as counterfeit. The posterior probabilities say that it
was pretty unlikely to be genuine, but it was the only bill for which
there was any noticeable doubt at all.

I had to use `pmax` rather than `max` there, because I
wanted `max.post` to contain the larger of the two
corresponding entries: that is, the first entry in `max.post`
is the larger of the first entry of `counterfeit` and the first
entry in `genuine`. If I used `max` instead, I'd get the
largest of *all* the entries in `counterfeit` and
*all* the entries in `genuine`, repeated 200 times. (Try
it and see.) `pmax` stands for "parallel maximum", that is,
for each row separately. This also should work:

```{r swiss-money-16, error=T}
d %>%
  rowwise() %>% 
  mutate(max.post = max(posterior.counterfeit, posterior.genuine)) %>%
  filter(max.post < 0.99) %>%
  select(-c(length:diag))
```

 
Because we're using `rowwise`, `max` is applied to the pairs
of values of `posterior.counterfeit` and `posterior.genuine`, 
*taken one row at a time.*

$\blacksquare$


