##  Test scores in two classes


  Open R Studio. Create a new Text File by selecting File, New File
and Text File. You should see a new empty, untitled window appear at
the top left. In that window, type or copy the data below (which are
scores on a test for students in two different classes):

```

class score
ken 78
ken 62
ken 59
ken 69
ken 81
thomas 83
thomas 77
thomas 63
thomas 61
thomas 79
thomas 72

```

Save the file, using a filename of your choice (with, perhaps, extension
`.txt`). Or, if you prefer, use the one at
[link](http://www.utsc.utoronto.ca/~butler/d29/marks.txt). 


(a) Read the data into a data frame called `marks`, using
`read_delim`, and list the data frame (by typing its name)
to confirm that you read the data values properly. Note that the
top line of the data file contains the names of the variables, as
it ought to.


Solution


I was lazy and used the one on the web, the
values being separated ("delimited") by exactly one space:
```{r }
my_url="http://www.utsc.utoronto.ca/~butler/d29/marks.txt"
marks=read_delim(my_url," ")
marks
```

    

If you copied and pasted, or typed in, the data values yourself, use the local file name (such as `marks.txt`) in place of the URL.

Extra: in the old days, when we used `read.table` (which actually also
works here), we needed to also say `header=T` to note that the
top line of the data file was variable names. With
`read_delim`, that's the default, and if the top line is
*not* variable names, that's when you have to say so. If I cheat,
by skipping the first line and saying that I then have no column names, I get:

```{r }
read_delim(my_url," ",col_names=F,skip=1)
```

 

Column names are supplied (`X1` and `X2`). I could also
supply my own column names, in which case the file needs not to have
any, so I need the `skip` again:

```{r }
read_delim(my_url," ",col_names=c("instructor","mark"),skip=1)
```

 
 


(b) Obtain side-by-side boxplots of the scores for each class.
 
Solution


```{r paignton}
library(tidyverse)
ggplot(marks,aes(x=class,y=score))+geom_boxplot()
```

       

Remember: on a regular boxplot,
`r tufte::margin_note("Boxplots can also go across the page, but for us, they don't.")` 
the groups go across ($x$), the
variable measured goes up ($y$).

Extra: this might work:

```{r }
ggplot(marks,aes(x=class,y=score))+geom_boxplot()+
coord_flip()
```

 

It does. That was a guess. So if you want sideways boxplots, this is
how you can get them. Long group names sometimes fit better on the $y$-axis, in which case flipping the axes will help.
(The `x` and `y` happen
*before* the coordinate-flip, so they are the same as above, not
the same way they come out.)
 

(c) Do the two classes appear to have similar or different
scores, on average? Explain briefly. 
 
Solution


The median for Thomas's class appears to be quite a bit higher
than for Ken's class (the difference is actually about 6
marks). It's up to you whether you think this is a big
difference or not: I want you to have *an* opinion, but I
don't mind so much what that opinion is.
Having said that the medians are quite a bit different, note
that the boxes overlap substantially, so that the
*distributions* of scores are pretty similar (or, the
quartiles of scores are similar, or, the IQR of scores is
similar for the two groups). If you say that, it's good, but I'm
not insisting that you do.
 

(d) Obtain a boxplot of all the scores together, regardless
of which class they came from. 


Solution


Replace your $x$-coordinate by some kind of dummy thing like
`1` (`factor(1)` also works):
```{r torquay}
ggplot(marks,aes(x=1,y=score))+geom_boxplot()
```

       

The $x$-axis is kind of dopey, so you just ignore it. It is possible
to remove it, but that is more work than it's worth, and I didn't get
rid of the ticks below:

```{r }
ggplot(marks,aes(x=1,y=score))+geom_boxplot()+
theme(axis.text.x=element_blank(),
axis.title.x=element_blank())
```

       
 


(e) Compute the median score (of all the scores
together). Does this seem about right, looking at the boxplot?
Explain briefly.


Solution



Three ways to get the median score. I like the first one best:

```{r }
marks %>% summarize(med=median(score))
with(marks,median(score))
median(marks$score)
```

 
`summarize` is the `tidyverse` "verb" that does what
you want here. (The same idea gets the mean score for each class, below.)

The other ways use the basic function `median`. To make that
work, you need to say that the variable `score` whose median
you want lives in the data frame `marks`. These are two ways to
do that.

Extra: if you wanted median by group, this is the approved
(`tidyverse`) way:

```{r }
marks %>% group_by(class) %>%
summarize(med=median(score))
```

 

To get something by group, the extra step is `group_by`, and
then whatever you do after that is done for *each* group.

You can now go back and compare these medians with the ones on the
boxplots in (b). They should be the same. Or you can even do this:

```{r }
marks %>% group_by(class) %>%
summarize(q1=quantile(score,0.25),
med=median(score),
q3=quantile(score,0.75)
)
```

 

You can calculate as many summaries as you like. These ones should
match up with the top and bottom of the boxes on the boxplots. The
only restriction is that the things on the right side of the equals
should return a *single* number. If you have a function like
`quantile` without anything extra that returns more than one number:

```{r }
quantile(marks$score)
```

 

you're in trouble. Only read on if you *really* want to know how
to handle this. Here's step 1:

```{r }
marks %>% 
nest(-class) 
```

 

This is kind of a funky `group_by`. The things in the
`data` column are the *whole* rest of the data frame:
there were 5 students in Ken's class and 6 in Thomas's, and they each
had a `score`, so 5 or 6 rows and 1 column. The column
`data` is known in the trade as a "list-column".

Now, for each of those mini-data-frames, we want to calculate the
quantiles of `score`. 
To do that, use `map`, which is the
`tidyverse` version of "for each": for each of our
mini-data-frames, calculate the five-number summary of the column called `score` in *it*:

```{r }
marks %>% 
nest(-class) %>%
mutate(qq=map(data,~quantile(.$score)))
```

  

I have to be a little bit careful about which data frame I want the
`score` to come from: the ones hidden in `data`, which
are the things we're for-eaching over.

This obtains a new list-column called `qq`, with the
five-number summary for each group.\endnote{It's actually a
coincidence that the five-number summary and Ken's class both have
five values in them.} 

Now we want to display the quantiles. This is the easiest way:

```{r }
marks %>% 
nest(-class) %>%
mutate(qq=map(data,~quantile(.$score))) %>%
unnest(qq)
```

  

The `unnest` turns the list-column back into actual data, so we
get the five quantiles for each class.

The only thing this doesn't do is to show us which quantile is which
(we know, of course, that the first one is the minimum, the last one
is the max and the quartiles and median are in between). It would be
nice to see which is which, though. A trick to do that is to use
`enframe`, thus:

```{r }
quantile(marks$score) %>% enframe()
```

 

or thus:

```{r }
enframe(quantile(marks$score))
```

 

I don't normally like the second way with all the brackets, but we'll
be using it later.

The idea here is that the output from a quantile is a vector, but one
with "names", namely the percentiles themselves. `enframe`
makes a two-column data frame with the names and the values. (You can
change the names of the columns it creates, but here we'll keep track
of which is which.)

So we have a *two*-column data frame with a column saying
which quantile is which. So let's rewrite our code to use this:

```{r }
marks %>% 
nest(-class) %>%
mutate(qq=map(data,~enframe((quantile(.$score)))))
```

 

Note that the `qq` data frames in the list-column now
themselves have two columns.

And finally `unnest` `qq`:

```{r }
marks %>% 
nest(-class) %>%
mutate(qq=map(data,~enframe((quantile(.$score))))) %>%
unnest(qq)  
```

 

Success! Or even:

```{r }
marks %>% 
nest(-class) %>%
mutate(qq=map(data,~enframe((quantile(.$score))))) %>%
unnest(qq) %>%
mutate(qn=parse_number(name)) %>%
select(-name) %>%
spread(qn, value)
```

 

This deliberately untidies the final answer to make it nicer to look
at. (The lines before that create a numeric quantile, so that it sorts
into the right order, and then get rid of the original quantile
percents. Investigate what happens if you do a similar `spread`
without doing that.)




