##  Throwing baseballs and softballs, again


 Previously, you carried out a sign test to determine
whether students could throw a baseball farther than a softball. This
time, we will calculate a confidence interval for the median
difference baseball minus softball, using the results of sign tests.



(a) Read the data into R from
[link](http://ritsokiguess.site/datafiles/throw.txt), giving
appropriate names to the columns, and add a column of
differences. 


Solution


I did it this way, combining the reading
of the data with the calculation of the differences in *one* pipe:

```{r }
myurl <- "http://ritsokiguess.site/datafiles/throw.txt"
throws <- read_delim(myurl, " ", col_names = c("student", "baseball", "softball")) %>%
  mutate(diff = baseball - softball)
throws
```

 

$\blacksquare$

(b) Use `smmr` to find a 95\% confidence interval for the
median difference.


Solution


`ci_median`, with 95\% being the default confidence level:

```{r }
ci_median(throws, diff)
```

 

2 to 9. The ends of a CI for the median will be data values, which are
all whole numbers, so round off that 8.999.

$\blacksquare$

(c) What function in `smmr` will run a two-sided sign test
and return only the P-value? Check that it works by testing whether the
median difference for your data is zero or different from zero.


Solution


The rest of the way, we are trying to reproduce that confidence
interval by finding it ourselves.
The function is called `pval_sign`. If you haven't run into it
before, in R Studio click on Packages, find `smmr`, and
click on its name. This will bring up package help, which
includes a list of all the functions in the package, along with
a brief description of what each one does. (Clicking
on a function name brings up the help for that function.)
Let's check that it works properly by repeating the previous
`sign_test` and verifying that `pval_sign` gives
the same thing:
```{r }
sign_test(throws, diff, 0)
pval_sign(0, throws, diff)
```

       

The P-values are the same (for the two-sided test) and both small, so
the median difference is not zero.

$\blacksquare$

(d) Based on your P-value, do you think 0 is inside the confidence
interval or not? Explain briefly.


Solution


Absolutely not. The median difference is definitely not
zero, so zero cannot be in the confidence interval.
Our suspicion, from the one-sided test from earlier, is that
the differences were mostly positive (people could throw a
baseball farther than a softball, in most cases). So the
confidence interval ought to contain only positive values.
I ask this because it drives what happens below.

$\blacksquare$

(e) Obtain a 95\% confidence interval for the population
median difference, baseball minus softball, using a
trial-and-error procedure that determines whether a number of
possible medians are inside or outside the CI.


Solution


 I've given you a fair bit of freedom to tackle this as you
 wish. Anything that makes sense is good: whatever mixture of
 mindlessness, guesswork and cleverness that you want to employ.
The most mindless way to try some values one at a time and see what
 you get, eg.:
```{r }
pval_sign(1, throws, diff)
pval_sign(5, throws, diff)
```

  

So median 1 is outside and median 5 is inside the 95\% interval. Keep
trying values until you've figured out where the lower and upper ends
of the interval are: where the P-values cross from below 0.05 to
above, or vice versa.

Something more intelligent is to make a long list of potential
medians, and get the P-value for each of them, eg.:

```{r }
d <- tibble(my.med = seq(0, 20, 2))
d %>% rowwise() %>% 
  mutate(pvals = pval_sign(my.med, throws, diff))
```

 

2 is just inside the interval, 8 is also inside, and 10 is
outside. Some closer investigation:

```{r }
d <- tibble(my.med = seq(0, 2, 0.5))
d %>% rowwise() %>% 
  mutate(pvals = pval_sign(my.med, throws, diff))
```

 

The bottom end of the interval actually is 2, since 2 is inside and
1.5 is outside.

```{r }
d <- tibble(my.med = seq(8, 10, 0.5))
d %>% rowwise() %>% 
  mutate(pvals = pval_sign(my.med, throws, diff))
```
 

The top end is 9, 9 being inside and 9.5 outside.

Since the data values are all whole numbers, I think this is accurate enough.
The most sophisticated way is the "bisection" idea we saw before. We
already have a kickoff for this, since we found, mindlessly, that 1 is
outside the interval on the low end and 5 is inside, so the lower
limit has to be between 1 and 5. Let's try halfway between, ie.\ 3:

```{r }
pval_sign(3, throws, diff)
```

 

Inside, so lower limit is between 1 and 3. This can be automated, thus:

```{r }
lo <- 1
hi <- 3
while (abs(hi - lo) > 0.1) {
  try <- (lo + hi) / 2
  ptry <- pval_sign(try, throws, diff)
  if (ptry > 0.05) {
    hi <- try
  } else {
    lo <- try
  }
}
c(lo, hi)
```

 

The difficult bit is to decide whether the value `try` becomes
the new `lo` or the new `hi`. If the P-value for the
median of `try` is greater than 0.05, `try` is inside
the interval, and it becomes the new `hi`; otherwise it's
outside and becomes the new `lo`. Whatever the values are,
`lo` is always outside the interval and `hi` is always
inside, and they move closer and closer to each other.

At the other end of the interval, `lo` is inside and
`hi` is outside, so there is a little switching around within
the loop. For starting values, you can be fairly mindless: for
example, we know that 5 is inside and something big like 20 must be outside:

```{r }
lo <- 5
hi <- 20
while (abs(hi - lo) > 0.1) {
  try <- (lo + hi) / 2
  ptry <- pval_sign(try, throws, diff)
  if (ptry > 0.05) {
    lo <- try
  } else {
    hi <- try
  }
}
c(lo, hi)
```

 

The interval goes from 2 to (as calculated here) about 9. (This is
apparently the same as `ci_median` in `smmr` got.)
`ci_median` uses the bisection method with a smaller "tolerance" than we
did, so its answer is more accurate. It looks as if the interval goes
from 2 to 9: that is, students can throw a baseball on average between
2 and 9 feet further than they can throw a softball.

$\blacksquare$




