##  Baseball and softball spaghetti


 On a previous assignment, we found that students could throw
a baseball further than they could throw a softball. In this question,
we will make a graph called a "spaghetti plot" to illustrate this
graphically. (The issue previously was that the data were matched
pairs: the same students threw both balls.)

This seems to work most naturally by building a pipe, a line or two at
a time. See if you can do it that way. (If you can't, use lots of
temporary data frames, one to hold the result of each part.)



(a) Read in the data again from
[link](http://www.utsc.utoronto.ca/~butler/c32/throw.txt). The
variables had no names, so supply some, as you did before.


Solution


Literal copy and paste: 
```{r }
myurl="http://www.utsc.utoronto.ca/~butler/c32/throw.txt"
throws=read_delim(myurl," ",col_names=c("student","baseball","softball"))
throws
```

       



(b) Create a new column that is the students turned into a
`factor`, adding it to your data frame.


Solution


Feed `student` into `factor`, creating a new
column with `mutate`:
```{r }
throws %>% mutate(fs=factor(student))
```

       

This doesn't look any different from the original student numbers, but
note the variable type at the top of the column.



(c) Gather together all the throwing distances into one column,
making a second column that says which ball was thrown.


Solution


Literally `gather` (from `tidyr`):
```{r }
throws %>% mutate(fs=factor(student)) %>% 
gather(ball,distance,baseball:softball)
```

       

Two columns to include, consecutive ones, or two to omit, the first
and last. So I think it's
easier to name the ones you want to include. 

If you want to show off a little, you can use a select-helper, noting
that the columns you want to gather up all end in "ball":

```{r }
throws %>% mutate(fs=factor(student)) %>% 
gather(ball,distance,ends_with("ball"))
```

       

with the same result. Use whichever you like.



(d) Using your new data frame, make a "scatterplot" of throwing
distance against type of ball.


Solution


The obvious thing:
```{r }
throws %>% mutate(fs=factor(student)) %>% 
gather(ball,distance,baseball:softball) %>%
ggplot(aes(x=ball,y=distance))+geom_point()
```

       

What this plot is missing is an indication of which student threw
which ball. As it stands now, it could be an inferior version of a
boxplot of distances thrown for each ball (which would imply that they
are two independent sets of students, something that is not true).



(e) Add two things to your plot: something that will distinguish
the students by colour (this works best if the thing distinguished
by colour is a factor),
`r tufte::margin_note("You can try it without. See below.")`
and something that will join the two points for the same student by
a line.


Solution


A `colour` and a `group` in the `aes`, and
a `geom_line`:
```{r }
throws %>% mutate(fs=factor(student)) %>% 
gather(ball,distance,baseball:softball) %>%
ggplot(aes(x=ball,y=distance,group=fs,colour=fs))+
geom_point()+geom_line()
```

     

You can see what happens if you use the student as a number:

```{r }
throws %>% mutate(fs=factor(student)) %>% 
gather(ball,distance,baseball:softball) %>%
ggplot(aes(x=ball,y=distance,group=student,colour=student))+
geom_point()+geom_line()
```

     

Now the student numbers are distinguished as a shade of blue (on an
implied continuous scale: even a nonsensical fractional student number
like 17.5 would be a shade of blue). This is not actually so bad here,
because all we are trying to do is to distinguish the students
sufficiently from each other so that we can see where the spaghetti
strands go. But I like the multi-coloured one better.



(f) The legend is not very informative. Remove it from the plot,
using `guides`.


Solution


You won't have seen this before. Here's what to do: Find what's
at the top of the legend that you want to remove. Here that is
`fs`. Find where `fs` appears in your
`aes`. It actually appears in two places: in
`group` and `colour`. I think the legend we want
to get rid of is actually the `colour` one, so we do this:
```{r }
throws %>% mutate(fs=factor(student)) %>% 
gather(ball,distance,baseball:softball) %>%
ggplot(aes(x=ball,y=distance,group=fs,colour=fs))+
geom_point()+geom_line()+
guides(colour=F) 
```

       

That seems to have done it.



(g) What do you see on the final spaghetti plot? What does that tell you
about the relative distances a student can throw a baseball vs.\ a
softball? Explain briefly, blah blah blah.


Solution


Most of the spaghetti strands go downhill from baseball to
softball, or at least very few of them go uphill. That tells us
that most students can throw a baseball further than a softball.
That was the same impression that the matched-pairs $t$-test
gave us. But the spaghetti plot tells us something else. If you
look carefully, you see that most of the big drops are for
students who could throw a baseball a long way. These students
also threw a softball further than the other students, but not
by as much. Most of the spaghetti strands in the bottom half of
the plot go more or less straight across. This indicates that
students who cannot throw a baseball very far will throw a
softball about the same distance as they threw the baseball.
There is an argument you could make here that the difference
between distances thrown is a *proportional* one, something
like "a student typically throws a baseball 20\% further than a softball". 
That could be assessed by comparing not the
distances themselves, but the logs of the distances: in other
words, making a log transformation of all the
distances. (Distances have a lower limit of zero, so you might
expect observed distances to be skewed to the right, which is
another argument for making some kind of transformation.)




