##  Growth of tomatoes


 A biology graduate student exposed each of 32
tomato plants to one of four different colours of light (8 plants to
each colour). The growth rate of each plant, in millimetres per week,
was recorded. The data are in
[link](http://www.utsc.utoronto.ca/~butler/c32/tomatoes.txt). 



(a) Read the data into R and confirm that you have 8 rows and 5
columns of data.


Solution


This kind of thing:
```{r }
my_url="http://www.utsc.utoronto.ca/~butler/c32/tomatoes.txt"
toms1=read_delim(my_url," ")
toms1
```

 

I do indeed have 8 rows and 5 columns.

With only 8 rows, listing the data like this is good. 



(b) Re-arrange the data so that you have *one* column
containing all the growth rates, and another column saying which
colour light each plant was exposed to. (The aim here is to produce
something suitable for feeding into `aov` later.) 


Solution


This is a job for `gather`:
```{r }
toms2 = toms1 %>% gather(colour,growthrate,blue:green)
toms2
```

       

Reminder: data frame to gather, what makes the columns different
(they're different colours), what makes them the same (they're all
growth rates), which columns to gather together (all the colour ones).

Since the column `plant` was never mentioned, this gets
repeated as necessary, so now it denotes "plant within colour group", 
which in this case is not very useful. (Where you have
matched pairs, or repeated measures in general, you *do* want to
keep track of which individual is which. But this is not repeated
measures because plant number 1 in the blue group and plant number 1
in the red group  are *different* plants.)

There were 8 rows originally and 4 different colours, so there should
be, and are, $8 \times 4=32$ rows in the gathered-up data set.



(c) Save the data in the new format to a text file. This is
most easily done using `write_csv`, which is the opposite
of `read_csv`. It requires two things: a data frame, and
the name of a file to save in, which should have a `.csv`
extension.  


Solution


The code is easy enough:
```{r }
write_csv(toms2,"tomatoes2.csv")
```

       

If no error, it worked. That's all you need.

To verify (for my satisfaction) that it was saved correctly:

```{r engine="bash"}
cat tomatoes2.csv 
```

 

On my system, that will list the contents of the file. Or you can just
open it in R Studio (if you saved it the way I did, it'll be in the
same folder, and you can find it in the Files pane.)



(d) Make a suitable boxplot, and use it to assess the assumptions
for ANOVA. What do you conclude? Explain briefly. 


Solution


Nothing terribly surprising here. My data frame is called
`toms2`, for some reason:
```{r }
ggplot(toms2,aes(x=colour, y=growthrate))+geom_boxplot()
```

     

There are no outliers, but there is a little skewness (compare the
*whiskers*, not the placement of the median within the box,
because what matters with skewness is the *tails*, not the middle
of the distribution; it's problems in the tails that make the mean
unsuitable as a measure of centre). The Red group looks the most
skewed. Also, the Yellow group has smaller spread than the others (we
assume that the population variances within each group are equal). The
thing to bear in mind here, though, is that there are only eight
observations per group, so the distributions could appear to have
unequal variances or some non-normality by chance. 

My take is that these data, all things considered, are \emph{just
about} OK for ANOVA. Another option would be to do Welch's ANOVA as
well and compare with the regular ANOVA: if they give more or less the
same P-value, that's a sign that I didn't need to worry.

Extra: some people like to run a formal test on the variances to test
them for equality. My favourite (for reasons explained elsewhere) is
the Levene test, if you insist on going this way. It lives in package
`car`, and *does not* take a `data=`, so you need
to do the `with` thing:

```{r }
library(car)
with(toms2,leveneTest(growthrate,colour))
```

 

The warning is because `colour` was actually text, but the test
did the right thing by turning it into a factor, so that's OK.

There is no way we can reject equal variances in the four groups. The
$F$-statistic is less than 1, in fact, which says that if the four
groups have the same population variances, the sample variances will
be *more* different than the ones we observed on average, and so
there is no way that these sample variances indicate different
population variances. (This is because of 8 observations only per
group; if there had been 80 observations per group, it would have been
a different story.) Decide for yourself whether you're surprised by this.

With that in mind, I think the regular ANOVA will be perfectly good,
and we would expect that and the Welch ANOVA to give very similar results.

I don't need `car` again, so let's get rid of it:

```{r }
detach("package:car",unload=T)
```

 



(e) Run (regular) ANOVA on these data. What do you conclude?
(Optional extra: if you think that some other variant of ANOVA would
be better, run that as well and compare the results.)


Solution


`aov`, bearing in mind that Tukey is likely to follow:

```{r }
toms.1=aov(growthrate~colour,data=toms2)
summary(toms.1)
```

 

This is a tiny P-value, so the mean growth rate for the different
colours is definitely *not* the same for all colours. Or, if you
like, one or more of the colours has a different mean growth rate than
the others.

This, remember, is as far as we go right now.

Extra: if you thought that normality was OK but not equal spreads,
then Welch ANOVA is the way to go:

```{r }
toms.2=oneway.test(growthrate~colour,data=toms2)
toms.2
```

 

The P-value is not *quite* as small as for the regular ANOVA, but
it is still very small, and the conclusion is the same.

If you had doubts about the normality (that were sufficiently great,
even given the small sample sizes), then go with Mood's median test
for multiple groups:

```{r }
library(smmr)
median_test(toms2,growthrate,colour)
```

 

The P-value is again extremely small (though not quite as small as for
the other two tests, for the usual reason that Mood's median test
doesn't use the data very efficiently: it doesn't use how *far*
above or below the overall median the data values are.)

The story here, as ever, is consistency: whatever you thought was
wrong, looking at the boxplots, needs to guide the test you do. This
should probably be a flow chart, but this way works too:



* if you are not happy with normality, go with
`median_test` from `smmr` (Mood's median test). 

* if you are happy with normality and equal variances, go with
`aov`.

* if you are happy with normality but not equal variances, go with
`oneway.test` (Welch ANOVA).


So the first thing to think about is normality, and if you are OK with
normality, then think about equal spreads. Bear in mind that you need
to be willing to tolerate a certain amount of non-normality and
inequality in the spreads, given that your data are only samples from
their populations. (Don't expect perfection, in short.)



(f) If warranted, run a suitable follow-up. (If not warranted, explain briefly
why not.)


Solution


Whichever flavour of ANOVA you ran (regular ANOVA, Welch ANOVA,
Mood's median test), you got the same conclusion for these data:
that the average growth rates were not all the same for the four
colours. That, as you'll remember, is as far as you go. To find
out which colours differ from which in terms of growth rate, you
need to run some kind of multiple-comparisons follow-up, the
right one for the analysis you did. Looking at the boxplots suggests that red is clearly best
and green clearly worst, and it is possible that all the colours
are significantly different from each other.)
If you did regular ANOVA, Tukey is what you need:
```{r }
TukeyHSD(toms.1)
```

       

All of the differences are (strongly) significant, except for yellow
and blue, the two with middling growth rates on the boxplot. Thus we
would have no hesitation in saying that growth rate is biggest in red
light and smallest in green light.

If you did Welch ANOVA, you need Games-Howell, which you have to get
from one of the packages that offers it:

```{r }
library(PMCMRplus)
gamesHowellTest(growthrate~factor(colour),data=toms2)
```

 

The conclusions are the same as for the Tukey: all the means are
significantly different except for yellow and blue.
Finally, if you did Mood's median test, you need this one:

```{r }
pairwise_median_test(toms2, growthrate, colour)
```

 

Same conclusions again. This is what I would have guessed; the
conclusions from Tukey were so clear-cut that it really didn't matter
which way you went; you'd come to the same conclusion.

That said, what I am looking for from you is a sensible choice of
analysis of variance (ANOVA, Welch's ANOVA or Mood's median test) for
a good reason, followed by the *right* follow-up for the test you
did. Even though the conclusions are all the same no matter what you
do here, I want you to get
used to following the right method, so that you will be able to do the
right thing when it *does* matter.




