##  Tortoise shells and eggs


 A biologist measured the length of the carapace (shell) of
female tortoises, and then x-rayed the tortoises to count how many
eggs they were carrying. The length is measured in millimetres. The
data are in
[link](http://ritsokiguess.site/datafiles/tortoise-eggs.txt). The
biologist is wondering what kind of relationship, if any, there is
between the carapace length (as an explanatory variable) and the
number of eggs (as a response variable).



(a) Read in the data, and check that your values look
reasonable. 
 
Solution


Look at the data first. The columns are aligned and separated by
more than one space, so it's `read_table`:
```{r tortoise-1 }
my_url <- "http://ritsokiguess.site/datafiles/tortoise-eggs.txt"
tortoises <- read_table(my_url)
tortoises
```

     

Those look the same as the values in the data file. (*Some*
comment is needed here. I don't much mind what, but something that
suggests that you have eyeballed the data and there are no obvious
problems: that is what I am looking for.)
 

$\blacksquare$

(b) Obtain a scatterplot, with a smooth trend, of the data.

 
Solution


Something like this:
```{r looe}
ggplot(tortoises, aes(x = length, y = eggs)) + geom_point() + geom_smooth()
```

   
 

$\blacksquare$

(c) The biologist expected that a larger tortoise would be able
to carry more eggs. Is that what the scatterplot is suggesting?
Explain briefly why or why not.

 
Solution


The biologist's expectation is of an upward trend. But it looks as
if the trend on the scatterplot is up, then down, ie.\ a curve
rather than a straight line. So this is not what the biologist was
expecting. 
 

$\blacksquare$

(d) Fit a straight-line relationship and display the summary.

 
Solution


```{r tortoise-2 }
tortoises.1 <- lm(eggs ~ length, data = tortoises)
summary(tortoises.1)
```

   

I didn't ask for a comment, but feel free to observe that this
regression is truly awful, with an R-squared of less than 2\% and a
non-significant effect of `length`.
 

$\blacksquare$

(e) Add a squared term to your regression, fit that and display
the  summary.

 
Solution


The `I()` is needed because the raise-to-a-power symbol has
a special meaning in a model formula, and we want to *not*
use that special meaning:
```{r tortoise-3 }
tortoises.2 <- lm(eggs ~ length + I(length^2), data = tortoises)
summary(tortoises.2)
```

 

Another way is to use `update`:

```{r tortoise-4 }
tortoises.2a <- update(tortoises.1, . ~ . + I(length^2))
summary(tortoises.2a)
```

 
 

$\blacksquare$

(f) Is a curve better than a line for these data? Justify your
answer in two ways: by comparing a measure of fit, and  by doing a
suitable test of significance.

 
Solution


An appropriate measure of fit is R-squared. For the straight line,
this is about 0.01, and for the regression with the squared term it
is about 0.43. This tells us that  a straight line fits appallingly
badly, and that a curve fits a *lot* better. 
This doesn't do a test, though. For that, look at the slope of the
length-squared term in the second regression; in particular,
look at its P-value. This is 0.0045, which is small: the squared
term is necessary, and taking it out would be a mistake. The
relationship really is curved, and trying to describe it with a
straight line would be a big mistake.
 

$\blacksquare$

(g) Make a residual plot for the straight line model: that is, plot
the residuals against the fitted values.
 Does this echo
your conclusions of the previous part? In what way? Explain briefly.

 
Solution


Plot the things called `.fitted` and `.resid` from the
regression object, which is not a data frame but you can treat it as
if it is for this:

```{r tortoise-5 }
ggplot(tortoises.1, aes(x = .fitted, y = .resid)) + geom_point()
```

 

Up to you whether you put a smooth trend on it or not:

```{r tortoise-6 }
ggplot(tortoises.1, aes(x = .fitted, y = .resid)) + geom_point() +
  geom_smooth()
```

 
Looking at the plot, you see a curve, up and down. The most
negative residuals go with small or large fitted values; when the
fitted value is in the middle, the residual is usually positive. A
curve on the residual plot indicates a curve in the actual
relationship. We just found above that a curve does fit a lot
better, so this is all consistent.

Aside: the grey "envelope" is wide, so there is a lot of scatter on the
residual plot. The grey envelope almost contains zero all the way
across, so the evidence for a curve (or any other kind of trend) is
not all that strong, based on this plot. This is in great contrast to
the regression with length-squared, where the length-squared term is
*definitely* necessary. 

That was all I wanted, but you can certainly look at other
plots. Normal quantile plot of the residuals:

```{r tortoise-7 }
ggplot(tortoises.1, aes(sample = .resid)) + stat_qq() + stat_qq_line()
```

 

This is not the best: the low values are a bit too low, so that the
whole picture is (a little) skewed to the left.^[The very   negative residuals are at the left and right of the residual plot;  they are there because the relationship is a curve. If you were to  look at the residuals from the model with length-squared, you  probably wouldn't see this.]

Another plot you can make is to assess fan-out: you plot the
*absolute value*^[The value, but throw away the minus sign if it has one.] of the residuals against the fitted values. The idea
is that if there is fan-out, the absolute value of the residuals will
get bigger:

```{r tortoise-8 }
ggplot(tortoises.1, aes(x = .fitted, y = abs(.resid))) + geom_point() +
  geom_smooth()
```

 

I put the smooth curve on as a kind of warning: it looks as if the
size of the residuals goes down and then up again as the fitted values
increase. But the width of the grey "envelope" and the general
scatter of the points suggests that there is really not much happening
here at all. On a plot of residuals, the grey envelope is really more
informative than the blue smooth trend. On this one, there is no
evidence of any fan-out (or fan-in). 

$\blacksquare$





