

 A study was made of the characteristics of urine of young
men. The men were classified into four groups based on their degree of
obesity. (The groups are labelled `a, b, c, d`.) Four variables
were measured, `x` (which you can ignore), pigment creatinine,
chloride and chlorine. The data are in
[link](http://www.utsc.utoronto.ca/~butler/d29/urine.csv) as a
`.csv` file. There are 45 men altogether.



(a) Read in the data and check that you have the right number
of observations and the right variables.
 
Solution


`read_csv`:
```{r }
my_url="http://www.utsc.utoronto.ca/~butler/d29/urine.csv"
urine=read_csv(my_url)
urine
```

     

The variables are as promised, and we do indeed have 45 rows. 
 

(b) Make boxplots of each of the three variables of interest
against obesity group. 
 
Solution


Just churn through it:
```{r peppercorn}
ggplot(urine,aes(x=obesity,y=creatinine))+geom_boxplot()
ggplot(urine,aes(x=obesity,y=chlorine))+geom_boxplot()
ggplot(urine,aes(x=obesity,y=chloride))+geom_boxplot()
```

     

This also works with facets, though it is different from the other
ones we've done, in that we have to collect together the $y$-variables
first rather than the $x$s:

```{r }
urine %>% gather(yname,y,creatinine:chlorine) %>%
ggplot(aes(x=obesity,y=y))+geom_boxplot()+
facet_wrap(~yname,scales="free",ncol=2)
```

 

I decided to throw a couple of things in here: first, the
`scales="free"` thing since the $y$-variables (this time) are
on different scales, and second, the `ncol=2` to arrange the
facets in (3 cells of) a $2\times 2$ grid, rather than having them
come out tall and skinny.\endnote{Like one of those crazy drinks at
Starbucks.} It's unusual to have faceted boxplots, but this is one
of those cases where it makes sense. (The key is different $y$'s but
the same $x$, I think.)

The conclusions about the boxplots are, of course, the same. I think
it makes it easier to have the three boxplots side by side, but it's
up to you whether you think that gain is worth the extra coding.
 

(c) How, if at all, do the groups differ on the variables of interest?
 
Solution


Any sensible comment here is good. You can take the approach that
group D is lowest on creatinine and chloride, but there is not
much to choose (given the amount of variability) on chlorine. Or
you can say that groups B and C are generally highest, except for
chloride where A is higher. Anything of this general kind is fine.
The point here is that some of the groups appear to be different
on some of the variables, which would make a multivariate analysis
of variance (in a moment) come out significant.
 

(d) Run a multivariate analysis of variance, using the three
variables of interest as response variables, and the obesity group
as explanatory. (This is a so-called one-way MANOVA.)
 
Solution


Create the response variable and run `manova`:
```{r }
response=with(urine,cbind(creatinine,chlorine,chloride))
urine.1=manova(response~obesity,data=urine)
summary(urine.1)
```

     
 

(e) What do you conclude? Is this in line with what your
boxplots said? Explain briefly.
 
Solution


The null hypothesis (that each of the variables have the same mean
for each of the groups) is rejected: that is, not all of the
groups have the same mean on all of the variables.
Based on the idea that there seemed to be differences between the
groups in the boxplots, this makes sense. (That's what I saw,
anyway. There was a lot of variability within the groups, which is
why the P-value didn't come out smaller.)
The other way of doing this is the following, using
`Manova` from `car`:
```{r }
response.1=lm(response~obesity,data=urine)
Manova(response.1)
```

     

The result is the same. You don't need do this one here (though you
can as an alternative), but when you come to repeated measures you
*will* need to be able to use `Manova`.

To understand the differences in the variables due to the groups, we
need to run a discriminant analysis (coming up later).

 



