##  Urine and obesity: what makes a difference?


 A study was made of the characteristics of urine of young
men. The men were classified into four groups based on their degree of
obesity. (The groups are labelled `a, b, c, d`.) Four variables
were measured, `x` (which you can ignore), pigment creatinine,
chloride and chlorine. The data are in
[link](http://www.utsc.utoronto.ca/~butler/d29/urine.csv) as a
`.csv` file. There are 45 men altogether.

Yes, you saw this one before. What you found was something like this:

```{r }
my_url="http://www.utsc.utoronto.ca/~butler/d29/urine.csv"
urine=read_csv(my_url)
response=with(urine,cbind(creatinine,chlorine,chloride))
urine.1=manova(response~obesity,data=urine)
summary(urine.1)
```

 

Our aim is to understand why this result was significant.



(a) Read in the data again (copy the code from above) and
obtain a discriminant analysis.
 
Solution


As above, plus:

```{r }
urine.1=lda(obesity~creatinine+chlorine+chloride,data=urine)
urine.1
```

     
 

(b) How many linear discriminants were you expecting? Explain briefly.
 
Solution


There are 3 variables and 4 groups, so the smaller of 3 and
$4-1=3$: that is, 3.
 

(c) Why do you think we should pay attention to the first two
linear discriminants but not the third? Explain briefly.
 
Solution

 The first two ``proportion of
trace'' values are a lot bigger than the third (or, the third
one is close to 0).
 

(d) Plot the first two linear discriminant scores (against each
other), with each obesity group being a different colour.
 
Solution

 First obtain the predictions, and
then make a data frame out of the original data and the
predictions. 

```{r }
urine.pred=predict(urine.1)
d=cbind(urine,urine.pred)
head(d)
```

             

`urine` produced the first five columns and `urine.pred`
produced the rest.

To go a more tidyverse way, we can combine the original data frame and
the predictions using `bind_cols`, but we have to be more
careful that the things we are gluing together are both data frames:

```{r }
class(urine)
class(urine.pred)
```

 

`urine` is a `tibble` all right, but `urine.pred` is a `list`. What does it look like?

```{r }
glimpse(urine.pred)
```

 
A data frame is a list for which all the items are the same length,
but some of the things in here are matrices. You can tell because they
have a number of rows, 45, *and* a number of columns, 3 or
4. They *do* have the right number of rows, though, so something
like `as.data.frame` (a base R function) will smoosh them all
into one data frame, grabbing the columns from the matrices:

```{r }
head(as.data.frame(urine.pred))
```

 

You see that the columns that came from matrices have gained two-part names, the first part from the name of the matrix, the second part from the column name within that matrix. Then we can do this:

```{r }
dd=bind_cols(urine, as.data.frame(urine.pred))
dd
```

 

If you want to avoid base R altogether, though, and go straight to
`bind_cols`, you have to be more careful about the types of
things. `bind_cols` *only* works with vectors and data
frames, not matrices, so that is what it is up to you to make sure you
have. That means pulling out the pieces, turning them from matrices
into data frames, and then gluing everything back together:

```{r error=T}
post=as_tibble(urine.pred$posterior)
ld=as_tibble(urine.pred$x)
ddd=bind_cols(urine, class=urine.pred$class, ld, post)
ddd
```

 
That's a lot of work, but you might say that it's worth it because you
are now absolutely sure what kind of thing everything is. I also had
to be slightly careful with the vector of `class` values; in
`ddd` it has to have a name, so I have to make sure I give it
one.
`r tufte::margin_note("If you run into an error like *Argument 2 must have names* here, that means that the second thing, *class*, needs  to have a name and doesn't have one.")`
Any of these ways (in general) is good. The last way is a more
careful approach, since you are making sure things are of the right
type rather than relying on R to convert them for you, but I don't
mind which way you go.
Now make the plot, making sure that you are using columns with the right names. I'm using my first data frame, with the two-part names:

```{r error=T}
ggplot(d,aes(x=x.LD1,y=x.LD2,colour=obesity))+geom_point()
```

 
 

(e) <a name="part:plot">*</a> Looking at your plot, discuss how (if at all) the
discriminants separate the obesity groups. (Where does each
obesity group fall on the plot?)
 
Solution

 My immediate reaction was
"they don't much". If you look a bit more closely, the
`b` group, in green, is on the right (high
`LD1`) and the `d` group (purple) is on the
left (low `LD1`). The `a` group, red, is
mostly at the top (high `LD2`) but the `c`
group, blue, really is all over the place.

The way to tackle interpreting a plot like this is to look
for each group individually and see if that group is only
or mainly found on a certain part of the plot. 

This can be rationalized by looking at 
the "coefficients of linear discriminants" on the output. `LD1` is
low if creatinine and chloride are low (it has nothing
much to do with `chlorine` since that coefficient
is near zero). Group `d` is lowest on both
creatinine and chloride, so that will be lowest on
`LD1`.  `LD2` is high if `chloride`
is high, or `creatinine` and `chlorine` are
low. Out of the groups `a, b, c`, `a` has
the highest mean on chloride and lowest means on the other
two variables, so this should be highest on `LD2`
and (usually) is.
Looking at the means is only part of the story; if the
individuals within a group are very variable, as they are
here (especially group `c`), then that group will
appear all over the plot. The table of means only says how
the *average* individual within a group stacks up.
```{r }
ggbiplot(urine.1, groups=urine$obesity)
```

 

This shows (in a way that is perhaps easier to see) how the linear
discriminants are related to the original variables, and thus how the
groups differ in terms of the original variables.\endnote{This was why
we were doing discriminant analysis in the first place.} Most of the
B's are high creatinine and high chloride (on the right); most of the
D's are low on both (on the left). LD2 has a bit of `chloride`,
but not much of anything else.
Extra: the way we used to do this was with "base graphics", which involved plotting the `lda` output itself:

```{r }
plot(urine.1)
```

 

which is a plot of each discriminant score against each other
one. You can plot just the first two, like this:

```{r }
plot(urine.1,dimen=2)
```

 
This is easier than using `ggplot`, but (i) less flexible and
(ii) you have to figure out how it works rather than doing things the
standard `ggplot` way. So I went with constructing a data frame
from the predictions, and then
`ggplot`ting that. It's a matter of taste which way is better.
xxxb
 

(f) <a name="part:table">*</a> Obtain a table showing observed and predicted obesity
groups. Comment on the accuracy of the predictions.
 
Solution


Make a table, one way or another:

```{r error=T}
tab=with(urine, table(obesity,class))
tab
```

           

`class` is always the *predicted* group in these. You can
also name things in `table`.
Or, if you prefer (equally good), the `tidyverse` way of
counting all the combinations of true `obesity` and predicted
`class`, which can be done all in one go, or in
two steps by saving the data frame first. I'm saving my results for
later:

```{r }
d %>% count(obesity, class) -> tab
tab
```

 
or if you prefer to make it look more like a table of frequencies:
```{r }
tab %>% spread(class, n, fill=0)
```

 

The thing on the end fills in zero frequencies as such (they would
otherwise be `NA`, which they are not: we know they are zero).
My immediate reaction to this is "it's terrible"! But at least some
of the men have their obesity group correctly predicted: 7 of the
$7+3+2+0=12$ 
men that are actually in group `a` are predicted to be in
`a`; 9 of the 14 actual `b`'s are predicted to be
`b`'s; 5 of the 8 actual `d`'s are predicted to be
`d`'s. These are not so awful. But only 1 of the 11
`c`'s is correctly predicted to be a `c`!

As for what I want to see: I am looking for some kind of statement
about how good you think the predictions are (the word "terrible" is
fine for this) with some kind of support for your statement. For
example, "the predictions are not that good, but at least group B is predicted with some accuracy (9 out of 14)."

I think looking at how well the individual groups were predicted is
the most incisive way of getting at this, because the `c` men
are the hardest to get right and the others are easier, but you could
also think about an overall misclassification rate. This comes most
easily from the "tidy" table:
```{r }
tab %>% count(correct=(obesity==class),wt=n) 
```

 
You can count anything, not just columns that already exist. This one
is a kind of combined mutate-and-count to create the (logical) column
called `correct`. 

It's a shortcut for this:

```{r }
tab %>% mutate(is_correct=(obesity==class)) %>%
count(is_correct, wt=n)
```

 
If I don't put the `wt`, `count` counts the number of
*rows* for which the true and predicted obesity group is the
same. But that's not what I want here: I want the number of
*observations* totalled up, which is what the `wt=`
does. It says "use the things in the given column as weights", which
means to total them up rather than count up the number of rows.

This says that 22 men were classified correctly and 23 were gotten
wrong. We can find the proportions correct and wrong:

```{r }
tab %>% count(correct=(obesity==class),wt=n) %>%
mutate(proportion=nn/sum(nn))
```

 

and we see that 51\% of men had their obesity group predicted
wrongly. This is the overall misclassification rate, which is a simple
summary of how good a job the discriminant analysis did.

I said above that the obesity groups were not equally easy to
predict. A small modification of the above will get the
misclassification rates by (true) obesity group. This is done by
putting an appropriate `group_by` in at the front, before we
do any summarizing:

```{r }
tab %>%
group_by(obesity) %>%
count(correct=(obesity==class),wt=n) %>%
mutate(proportion=nn/sum(nn))
```

 

This gives the proportion wrong and correct for each (true) obesity
group. I'm going to do the one more cosmetic thing to make it easier to
read, a kind of "untidying":

```{r }
tab %>%
group_by(obesity) %>%
count(correct=(obesity==class),wt=n) %>%
mutate(proportion=nn/sum(nn)) %>%
select(-nn) %>%
spread(correct,proportion)
```

 

Looking down the ``TRUE`` column, groups A, B and D were gotten
about 60\% correct (and 40\% wrong), but group C is much worse. The
overall misclassification rate is made bigger by the fact that C is so
hard to predict.

Find out for yourself what happens if I fail to remove the `nn`
column before doing the `spread`.

A slightly more elegant look is obtained this way, by making nicer
values than TRUE and FALSE:

```{r }
tab %>%
group_by(obesity) %>%
mutate(prediction_stat=ifelse(obesity==class,"correct","wrong")) %>%
count(prediction_stat,wt=n) %>%
mutate(proportion=nn/sum(nn)) %>%
select(-nn) %>%
spread(prediction_stat,proportion)
```

 
 

(g) Do your conclusions from (<a href="#part:plot">here</a>) and
(<a href="#part:table">here</a>) appear to be consistent?
 
Solution


On the plot of (<a href="#part:plot">here</a>), we said that there was a
lot of scatter, but that groups `a`, `b` and
`d` tended to be found at the top, right and left
respectively of the plot. That suggests that these three
groups should be somewhat predictable. The `c`'s, on
the other hand, were all over the place on the plot, and
were mostly predicted wrong.

The idea is that the stories you pull from the plot and the
predictions should be more or less consistent. There are
several ways you might say that: another approach is to say
that the observations are all over the place on the plot,
and the predictions are all bad. This is not as insightful
as my comments above, but if that's what the plot told you,
that's what the predictions would seem to be saying as
well. (Or even, the predictions are not so bad compared to
the apparently random pattern on the plot, if that's what
you saw. There are different ways to say something more or
less sensible.)
 


