##  The weather, somewhere


 The data in
[link](http://ritsokiguess.site/datafiles/weather_2014.csv) is of the
weather in a certain location: daily weather records for 2014. The
variables are:



* day of the year (1 through 365)

* day of the month

* number of the month

* season

* low temperature (for the day)

* high temperature

* average temperature

* time of the low temperature

* time of the high temperature

* rainfall (mm)

* average wind speed

* wind gust (highest wind speed)

* time of the wind gust

* wind direction




(a) Read in the data, and create a data frame containing only
the temperature variables, the rainfall and the wind speed variables
(the ones that are actual numbers, not times or text). Display the
first few lines of your data frame.
 
Solution


Read into a temporary data frame, and then process:
```{r message=F}
my_url <- "http://ritsokiguess.site/datafiles/weather_2014.csv"
weather.0 <- read_csv(my_url)
weather.0
```

     

There are lots of columns, of which we only want a few:

```{r }
(weather.0 %>% select(l.temp:ave.temp, rain:gust.wind) -> weather)
```

 
$\blacksquare$

(b) Find five-number summaries for each column by running
`quantile` on all the columns of the data frame (at once, if
you can).
 
Solution

I think this is the easiest way:

```{r}
weather %>% 
  summarize(across(everything(), ~quantile(.)))
```

This loses the actual percents of the percentiles of the five-number summary (because they are "names" of the numerical result, and the tidyverse doesn't like names.) I think you can see which percentile is which, though.

Another way to do it is to make a column of column names, using
`pivot_longer`, and then use `nest` and list-columns to find
the quantiles for each variable:

```{r }
weather %>%
  pivot_longer(everything(), names_to="xname", values_to="x") %>%
  nest_by(xname) %>%
  mutate(q = list(enframe(quantile(data$x)))) %>%
  unnest(q) %>%
  pivot_wider(names_from=name, values_from=value) %>%
  select(-data)
```

 

That was a lot of work, but it depends on how you see it when you're coding it. You should investigate this one line at a time, but the steps are:



* create a "long" data frame with one column of variable names and a second with the values for that variable

* make mini-data-frames `data` containing everything but `xname`: that is, one column `x` with the values for that variable.

* for each mini-data-frame, work out the quantiles of its `x`. The `enframe` saves the labels for what percentiles they are.

* Unnest this to make a long data frame with one row for each quantile for each variable.

* put the variable names in rows and the percentiles in columns.
 
$\blacksquare$ 

(c) Run a principal components analysis (on the correlation matrix).
 
Solution


```{r }
weather.1 <- princomp(weather, cor = T)
```

     
$\blacksquare$ 

(d) Obtain a `summary` of your principal components
analysis. How many components do you think are worth investigating?
 
Solution


```{r }
summary(weather.1)
```

     
The issue is to see where the standard deviations are getting small
(after the second component, or perhaps the third one) and to see
where  the cumulative proportion of variance explained is acceptably
high (again, after the second one, 86\%, or the third, 95\%).
 
$\blacksquare$ 

(e) Make a scree plot. Does this
support your conclusion from the previous part?
 
Solution


`ggscreeplot` from `ggbiplot`:
```{r adskjheuya}
ggscreeplot(weather.1)
```

     

I see elbows at 3 and at 4. Remember you want to be on the mountain
for these, not on the scree, so this suggests 2 or 3 components, which
is exactly what we got from looking at the standard deviations and
cumulative variance explained.

The eigenvalue-greater-than-1 thing 
(that is, the "standard deviation" in the `summary` being greater than 1)
says 2 components, rather than 3.
 
$\blacksquare$ 

(f) Obtain the component loadings. How do the first three
components depend on the original variables? (That is, what kind of
values for the original variables would make the component scores
large or small?)
 
Solution


```{r }
weather.1$loadings
```

     



* 1: This component loads mainly (and positively) on the temperature
variables, so when temperature is high, component 1 is high.
You could also say that it loads negatively on the other
variables, in which case component 1 is high if the temperature
variables are low and the rain and wind variables are high.


* 2: This one loads most heavily, positively, on wind: when wind is high,
component 2 is high. Again, you can make the judgement call that the
other variables also feature in component 2, so that when
*everything* is large, component 2 is large and small with small.

* 3: This one is a bit clearer. The blank loadings are close to 0,
and can be ignored. The main thing in component 3 is rain: when
rainfall is large, component 3 is large. Or, if you like, it is
large (positive) when rainfall is large *and wind is small*.


The interpretation here is kind of muffled, because each component has
bits of everything. One of the advantages of factor analysis that we
see in class later is that
you can do a "rotation" so that each variable (to a greater extent)
is either *in* a factor or *out* of it. Such a varimax rotation
is the default for `factanal`, which I presume I now have to
show you (so this is looking ahead): 

```{r }
weather.2 <- factanal(weather, 3, scores = "r")
weather.2$loadings
```

 

These are a lot less ambiguous: factor 1 is temperature, factor 2 is
rain and wind, and factor 3 is large (positive) if the high
temperature is high or the low temperature is low: that is, if the
high temperature was especially high relative to the low temperature
(or, said differently, if the temperature *range* was high).

These factors are rather pleasantly interpretable.

`ggbiplot` mysteriously doesn't handle factor analyses, so we
have to go back to the base-graphics version, which goes a bit like this:

```{r }
biplot(weather.2$scores, weather.2$loadings)
```

 

Now you see that the factors are aligned with the axes, and it's very
clear what the factors "represent". (You don't see much else, in all
honesty, but you see at least this much.)
 
$\blacksquare$ 

(g) Obtain the principal component scores, for as many components
as you think are reasonable, and display the first
20 of them for each component alongside the other variables in your
data frame.
 
Solution


Something like this. I begin by turning the component scores
(which are a matrix) into a data frame, and selecting the ones I
want (the first three):
```{r echo=F}
ww <- getOption("width")
options(width = ww + 20)
```

      
```{r size="footnotesize"}
as_tibble(weather.1$scores) %>%
  select(1:3) %>%
  bind_cols(weather) %>%
  mutate(day = row_number()) -> d
d 
```

      
I just did the first three scores. I made a column `day` so that I can see which day of the year I am looking at (later).

$\blacksquare$ 

(h) Find a day that scores low on component 1, and explain briefly why
it came out that way (by looking at the measured variables).
 
Solution


We can do this one and the ones following by running
`arrange` appropriately:
```{r }
d %>% arrange(Comp.1)
```

       
Day 40 has the lowest component 1 score. This is one of the
cooler days. Also, there is a
largish amount of rain *and* wind. So low temperature, high
rain and wind.
Some of the other days on my list were cooler than day 4, but they
had less rain and less wind.
  
$\blacksquare$

(i) Find a day that scores high on component 2, and explain
briefly why it came out that way.
 
Solution


```{r }
d %>% arrange(desc(Comp.2))
```

       
Day 37. These are days when the wind speed (average or
gust) is on the high side. 

$\blacksquare$ 

(j) Find a day that scores high on component 3, and explain
briefly why it came out high.
 
Solution


```{r }
d %>% arrange(desc(Comp.3))
```

 
Day 307. Component 3 was mainly rain, so it is not surprising that
the rainfall is the highest on this day.

$\blacksquare$ 

(k) Make a biplot of these data, labelling the days by the day
count (from 1 to 365). You may have to get the day count from the
original data frame that you read in from the file. You can shrink
the day numbers to make them overwrite each other (a bit) less.
 
Solution


`ggbiplot`. I did some digging in the help file to figure
out how to label the points by a variable and how to control the
size of the labels, and I also went digging in the data frame
that I read in from the file to get the count of the day in the
year, which was called `day.count`:
```{r }
ggbiplot(weather.1, labels = weather.0$day.count, labels.size = 2)
```

     

I think the label text is small enough, though you could make it
smaller. I'll be asking you to look at some extreme points in a
moment, so those are the only ones you'll need to be able to
disentangle. 

The variables divide into two groups: the temperature ones, that point
to about 2 o'clock, and the wind and rain ones, that point to about 11
o'clock. These are not straight up or down or across, so they all
feature in both components: component 1 is mostly temperature, but has
a bit of wind/rain in it, while component 2 is mostly wind/rain with a
bit of temperature in it. You might be wondering whether things
couldn't be rotated so that, say, the temperature variables go across
and the rain/wind ones go down, which means you'd have a temperature
component and a rain/wind component. This is what factor analysis
does, and I think I did that earlier (and this is what we found).
 
$\blacksquare$

(l) Looking at your biplot, what do you think was remarkable
about the weather on day 37? Day 211? Confirm your guesses by
looking at the appropriate rows of your data frame (and comparing
with your `summary` from earlier).
 
Solution


Day 37 is at the top left of the plot, at the pointy end of
the arrows for rain, wind gust and average wind. This suggests a
rainy, windy day:
```{r }
weather %>% slice(37)
```

     

Those are high numbers for both rain and wind (the highest for average
wind and above the third quartile otherwise), but the temperatures
are unremarkable.

Day 211 is towards the pointy end of the arrows for temperature, so
this is a hot day:

```{r }
weather %>% slice(211)
```

 

This is actually the hottest day of the entire year: day 211 is
highest on all three temperatures, while the 
wind speeds are right around average (and no rain is not completely
surprising). 

I can do a couple more. Points away from the pointy end of the arrows
are *low* on the variables in question, for example day 265:

```{r }
weather %>% slice(265)
```

 

This is not really low rain, but it is definitely low
*wind*. What about day 47?

```{r }
weather %>% slice(47)
```

 

This is predominantly low on temperature. In fact, it is kind of low
on wind and rain too.^[If you ignore the wind gust, anyway.]
This makes sense, because not only is it at the "wrong" end of the
temperature arrows, it is kind of at the wrong end of the wind/rain
arrows as well.

Having done these by percentile ranks in one of the other questions,
let's see if we can do that here as well:

```{r}
weather %>% mutate(across(everything(), ~percent_rank(.))) %>% 
    slice(c(37, 211, 265, 47))
```

The idea here is that we want to *replace* all the data values by the percent-rank version of themselves, rather than *summarizing* them as we have done before. That's what using an `across` inside a `mutate` will do.^[There are also options to keep the original variables, and to give the new ones new names.]
 

These are:



* Day 37: highly rainy and windy (and below average,
but not remarkably so, on temperature). 

* Day 211: the highest or near-highest temperature, no rain but
unremarkable for wind.

* Day 265: Lowest for wind (and above Q3 for low temperature and
rain). 

* Day 47: Lowest or near-lowest temperature.


The advantage to doing it this way is that you don't need a separate
five-number summary for each variable; the percentile ranks
*give* you a comparison with quartiles (or any other percentile
of interest to you).
In case you are wondering where this is: I was doing a presentation
using these data to some Environmental Science grad students, and I
had them guess where it was. The temperatures for the whole year are
warm-temperate, with a smallish range, and sometimes a lot of
rain. This suggests a maritime climate. I gave the additional clues of
"western Europe" and 
"this place's soccer team plays in blue and white striped shirts". 
The temperatures have about the right range
low-to-high for Britain, but are too warm. Which suggests going south:
perhaps Brittany in France, but actually the west
coast of the Iberian peninsula: Porto, in northern Portugal, with the
weather blowing in off the Atlantic. 

$\blacksquare$

