##  Making a map of Wisconsin


 The file
[link](http://ritsokiguess.site/datafiles/wisconsin.txt) contains
the road distances (in miles) between 12 cities in Wisconsin and
neighbouring states. We are going to try to draw a map of the
area using Leaflet.


(a) Read in the data, displaying
the data that you read in.


Solution


```{r wisconsin2-1 }
my_url <- "http://ritsokiguess.site/datafiles/wisconsin.txt"
wisc <- read_table(my_url)
wisc
```

       

The first time I did this, I had a blank line on the end of the data
file, so I had a blank `location` and missing values for all
the distances for it. I tidied that up before sharing the file with
you, though.

$\blacksquare$


(b) Make a dataframe containing the names of the locations (get rid of the columns containing distances), and add a column of the abbreviations of the states they are in. All of them are in Wisconsin (WI), except for the
last three: Dubuque is in Iowa (IA), St.\ Paul is in Minnesota (MN) and Chicago
is in Illinois (IL).

Solution

There seems to be a bit of base R attached to this, however you do it. I am going to create a dataframe pretending they are all in Wisconsin, and then fix it up afterwards:

```{r wisconsin2-2}
wisc %>% 
  select(!where(is.numeric)) %>% 
  mutate(state = "WI") -> wisc
wisc
```

(I checked that in this question I didn't need the road distances for anything, so I saved it back into the original dataframe. Also, the `select` is unnecessarily fancy: I could have just selected the `location` column, but this one says "don't select any of the columns that are numeric".) 

Now we have to fix up the states of the last three places, which is where the base R seems to come in (but see the Extra):

```{r wisconsin2-3}
wisc$state[10] <- "IA"
wisc$state[11] <- "MN"
wisc$state[12] <- "IL"
wisc
```

The states of the last three locations are now correct.

Extra: I didn't know about the following until literally just now, but there is a `tidyverse` way to do this as well (that may look familiar to those of you that know about SQL). Let's start by pretending again that everything is in Wisconsin:

```{r wisconsin2-4}
wisc %>% 
  mutate(state = "WI") -> wisc2
wisc2
```

and then change the ones that need changing. What you do is to make a little dataframe of the ones that need changing:

```{r wisconsin2-5}
changes <- tribble(
  ~location, ~state,
  "Dubuque", "IA",
  "St.Paul", "MN",
  "Chicago", "IL"
)
changes
```

Note that the columns in here have *exactly* the same names as the ones in the original dataframe where everything was in Wisconsin. 


I did this by copy-pasting the city names whose states needed changing out of the display of my `wisc2`. You might think that a left join is what we need now, and it almost is. Note that I want to match the locations but *not* the states:

```{r wisconsin2-6}
wisc2 %>% left_join(changes, by = "location")
```

and the story here is that if `state.y` has a value, use that, otherwise use  the value in `state.x`. This can even be done: there is a function `coalesce`^[I knew this existed, but I couldn't remember what it was called, which made it hard to search for. My first port of call was `na_if`, which converts values to `NA`, the opposite of what I wanted. But from its See Also I got `na_replace`, and from the See Also of that, I found out what `coalesce` does.]  that will do exactly that:


```{r wisconsin2-7}
wisc2 %>% left_join(changes, by = "location") %>% 
  mutate(state=coalesce(state.y, state.x))
```

The idea behind `coalesce` is that you give it a list of columns, and the first one of those that is *not missing* gives its value to the new column. So I feed it `state.y` *first*, and then `state.x`, and the new `state` contains the right things. (Can you explain what happens if you do it the other way around?)

But, there is a better way. Let's go back to our all-Wisconsin dataframe:

```{r wisconsin2-8}
wisc2
```

and our dataframe of corrections to make:

```{r wisconsin2-9}
changes
```

We can make those changes in one step, thus:

```{r wisconsin2-10}
wisc2 %>% 
  rows_update(changes) -> wisc
wisc
```

This works because the *first* column of `changes`, namely `location`, is the one that is looked up in `wisc2`. (`rows_update` has a `by` in the same way that `left_join` does if you want to change this.) So all three locations in `changes` are looked up in `wisc2`, and any that match have their `state` changed to the one in `changes`.

In database terms, the `location` is known as a "key" column. That means that each city appears *once only* in the column, and so the replacements in `wisc` are only happening once. To a statistician, `location` is an "identifier variable", identifying the individuals in the dataset. Unless you are doing something like repeated measures, each individual will only give you one measurement, but even then, if you have wide format, the identifier variables will still only appear once.

Magic. Now that I have learned about this, I will be using it a lot.

$\blacksquare$

(c) Create a new column in which the abbreviation for the state is glued on to the end of each `location`, separated by a space.

Solution

A couple of ways: a literal gluing, using `paste`:

```{r wisconsin2-11}
wisc %>% 
  mutate(lookup = paste(location, state))
```

or the same idea using `str_c` from `stringr` (part of the tidyverse), only this time you have to supply the space yourself:

```{r wisconsin2-12}
wisc %>% 
  mutate(lookup = str_c(location, " ", state))
```

or a way you might have forgotten, using `unite` (which is the inverse of `separate`):

```{r wisconsin2-13}
wisc %>% 
  unite(lookup, c(location, state), sep = " ") -> wisc
wisc
```

This last one is my favourite, because it gets rid of the two constituent columns `location` and `state` that we don't need any more. The syntax is the name of the new column, a vector of columns to unite together, and (optionally) what to separate the joined-together values with. The default for that is an underscore, but here we want a space because that's what the geocoder (coming up) expects.

$\blacksquare$

(d) Look up the latitudes and longitudes of these twelve places.

Solution

This is geocoding, with the disentangling afterwards that is (I hope) becoming familiar:

```{r wisconsin2-14}
wisc %>% 
  rowwise() %>% 
  mutate(ll = list(geocode_OSM(lookup))) %>% 
  unnest_wider(ll) %>% 
  unnest_wider(coords) -> wisc
wisc
```

Yes, I forgot the `rowwise` as well the first time.

$\blacksquare$

(e) Obtain a Leaflet map of the area containing these
twelve cities. 
 
Solution

The usual:

```{r wisconsin2-15}
leaflet(data = wisc) %>% 
  addTiles() %>% 
  addCircleMarkers(lng = ~x, lat = ~y) -> locs
locs
```

The nice thing about this map is that you can play with it: zoom it (using the plus/minus on the map or your mouse wheel), or move it around by clicking and dragging. To identify the cities: well, the big ones are obvious, and you can zoom in to identify the others. (You have to zoom in quite a long way to find Monroe, and the geocoder seems to have found its airport, which is not actually in the city.)

I like identifying the cities with circles, but there are other possibilities, such as "icon markers" that look like Google map pins:

```{r wisconsin2-16}
leaflet(data = wisc) %>% 
  addTiles() %>% 
  addMarkers(lng = ~x, lat = ~y) -> locs
locs
```

You can even attach text to the markers that appears when you click on them, but that's farther than I would go here.


$\blacksquare$


