##  Making a map of Wisconsin


 The file
[link](http://www.utsc.utoronto.ca/~butler/d29/wisconsin.txt) contains
the road distances (in miles) between 12 cities in Wisconsin and
neighbouring states. We are going to try to reproduce a map of the
area using multidimensional scaling.


(a) Read in the data and create a `dist` object,
bearing in mind that the data in the file are already
distances. Display your `dist` object. Probably, displaying
the data that you read in is a good idea also.
 
Solution


```{r }
my_url <- "http://www.utsc.utoronto.ca/~butler/d29/wisconsin.txt"
wisc <- read_table(my_url)
wisc
```

       

The first time I did this, I had a blank line on the end of the data
file, so I had a blank `location` and missing values for all
the distances for it. I tidied that up before sharing the file with
you, though.

So, the first column is the names of the places, which we should get
rid of before we make a `dist` object using `as.dist`
(since what we read in is already distances). The columns are also the
names  of the places, so we won't lose anything by getting rid of the
`location` column:

```{r }
d <- wisc %>%
  select(-location) %>%
  as.dist()
d
```

 
 

(b) Obtain a vector containing the city names. (This is easy,
and not really necessary, but it was here before when I did things
a different way.)
 
Solution


The `location` column of the data frame you read in
from the file:
```{r }
cities <- wisc$location
cities
```

       
 

(c) Run a (metric) multidimensional scaling on the data, to
obtain a two-dimensional representation of the cities. (You
don't need to look at the results yet.)
 
Solution


```{r }
wisc.1 <- cmdscale(d)
```

       
 

(d) Plot the results of the multidimensional scaling,
labelling the cities with their names. Use your judgement to
decide where to place the city names, and how to  make sure the
whole city names are shown on the map.
 
Solution


Let's see what `wisc.1` has in it, and make a data
frame of the right thing:
```{r }
head(wisc.1)
```

         
Two unnamed columns (the display indicates that it's a `matrix`
rather than a data frame). As we saw in class, if we make a data frame
out of this, the columns will get names `X1` and
`X2`. Those are perfectly good names for coordinates. The city
names on the left of `wisc.1` are actually row names rather
than an actual genuine column. It's probably best *not* to assume
that they will make it through the transition to a data frame, so
we'll explicitly create a column called `city` with the city
names (that we saved before):

```{r }
dd <- data.frame(wisc.1, city = cities)
dd
```

 

There are only 12 rows, so it's fine to display them all.

I'm calling this one `dd` since I have another `d`
elsewhere that I want to keep. I should use better names.

I think the best use of your judgement is to go straight to
`geom_text_repel` from package `ggrepel`:

```{r }
ggplot(dd, aes(x = X1, y = X2, label = city)) +
  geom_point() +
  geom_text_repel() +
  coord_fixed()
```

 

Your map may come out different from mine, but subject to the usual
stuff about rotation and reflection it should be equivalent to
mine. You should include the `coord_fixed` to get the scales
the same (a corresponding distance on the two scales should take the
same space). This one didn't come out quite square because the MDS
says the points should be in a rectangle (stretching further one way
than the other). 
 

(e) Are cities close together on your map also close together
in real life? Give an example or two.
 
Solution


On the map, the trio of cities Madison, Beloit and Fort Atkinson
are closest together. How far apart are they actually? Well, you
can go back to the original file (or display of what I called
`d`) and find them, or you can do this:
```{r }
cities
```

     

Cities 2, 3 and 4, so:

```{r }
wisc %>% slice(2:4) %>% select(c(1, 3:5))
```

 

The column numbers are off by one, since the first column is the names
of the cities, which I decided to display here. It came out right, anyway.

These are all less than 50 miles or less apart. There are some others
this close in the original data: Monroe and Madison are 47 miles
apart, Wausau and Marshfield are 45 miles apart, but these appear
further apart on the map.
Extra: the slice-select thing doesn't work on `d` because that is not
a data frame. It is actually stored internally as a
*one*-dimensional vector that displays nicely, but if you want to
pull things out of it you have to figure out where in the vector they are:

```{r }
print.default(d)
```

 

If you compare that with the usual display of `d`, this way
goes all the way down the first column ending at 130, then all the way
down the second column (which has one fewer entry), ending at 467, and
so on. Thus the three entries we picked out are at $11+1=12$,
$11+2=13$, and $11+10+1=22$:

```{r }
d[12]
d[13]
d[22]
```

 

It's very easy to be off by one in this sort of calculation. There are
12 cities altogether, so *11* distances in the first column, 10
in the second, and so on. This was about my third attempt.

I don't much care which cities you look at. Finding some cities that
are reasonably close on the map and doing some kind of critical
 assessment of their actual distances apart is all I want. 
 


(f) Obtain a Google (or other) map of the area containing these
twelve cities. (The way I expected you to do this when this was a
hand-in problem was to take a screenshot or similar and include that
in your document.)

 
Solution


Since I like to show off, let me show you how you can do this in R,
using the package `ggmap`. (Of course, you can just open the
appropriate map in your browser and copy-paste it, taking a
screenshot if necessary):
```{r }
library(ggmap)
```

   

```{r echo=F}
api_key <- "AIzaSyC7eVXz84-Mj1-zNBW4GTnXNAYdGQJDNXU"
register_google(api_key)
```

 
```{r laskhsal,cache=TRUE,fig.width=6,fig.height=6}
library(ggmap)
wisc.map <- get_map(location = "Milwaukee, WI", zoom = 6)
ggmap(wisc.map)
```

   

I centred this map around Milwaukee (a guess), which is not quite
where the centre should be, since Milwaukee is in southeastern
Wisconsin. The `zoom` option is how zoomed-in the map should be
(a higher number is more zoomed-in). Likewise, 6 was a guess, and it
seems that I need to zoom in a bit more.

The other way of specifying a location, instead of the name or
lat-long of the centre of the map, is to specify the **corners**
of the map in degrees of latitude or longitude. We have to give four
numbers: lower left longitude and latitude, upper right longitude and
latitude. (Degrees west are *negative*, as you see on the
`lon` scale of the above map.) This comes out as west, south,
east and north limits of the map.
Where are the 12 points we want to put on the map? We can get their
latitudes and longitudes, which is called "geocoding", and a
function `geocode` is included in `ggmap`.

First add the state names to the cities, to make sure Google Maps
looks up the right ones. All of them are in Wisconsin, except for the
last three: Dubuque is in Iowa, St.\ Paul is in Minnesota and Chicago
is in Illinois. I am first making a vector with 12 copies of the
abbreviation for Wisconsin, and then overwriting the last three with
the states they are actually in:

```{r }
states <- rep("WI", 12)
states[10] <- "IA"
states[11] <- "MN"
states[12] <- "IL"
cst <- str_c(cities, " ", states)
cst
```

 

And then I look them up. I don't want Google to tell me that I have
done too many lookups, so I look them up first and save the results,
like this:

```{r } 
tibble(cst) %>%
  ggmap::mutate_geocode(cst) -> g
g
```

 
What are the extreme corners of these?

```{r }
range(g$lon)
range(g$lat)
```

 

or even

```{r }
g %>%
  select(lat, lon) %>%
  map_df(range)
```

 

(`range` in R produces the two extreme values, not the
difference between the highest and lowest, which is what you might
think of as a "range".)

We don't
get exactly the corners we ask for, since the map always comes out in the same
proportions (we could ask for a long skinny map, but we'll always get
a rectangular one that fills the page), and also Google Maps converts
the corners into a centre and zoom. I had to tinker with the numbers
below, since on my first attempt the map zoomed in too much.
I also asked for a "roadmap" to maximize the number of places marked
on there.

So:

```{r lhsdljhadkfj,fig.width=6,fig.height=6,cache=T,message=F}
wisc.map.2 <- get_map(location = c(-94, 41.8, -87, 46.8), maptype = "roadmap")
ggmap(wisc.map.2)
```

 

This came out about right. 

Now we need to mark our 12 cities on the map. This is a
`ggplot` map, so the right syntax is as below.

```{r sldkhdlf,echo=-1,message=F, cache=T}
library(ggmap)
ggmap(wisc.map.2) + geom_point(data = g, aes(x = lon, y = lat))
```

 

There's a slight weirdness in that `ggmap` only takes a map,
not a data frame or an `aes`, so those go in
`geom_point` or whatever you're using.

We *just* squeezed all our cities onto the map. The city
southwest of Wausau is Marshfield, the one between Madison and
Milwaukee is Fort Atkinson, and the two east of Dubuque along the
southern border of Wisconsin are Monroe and Beloit. The one *way*
up at the top is Superior.

After that long diversion, we come to:

 


(g) Discuss how the map that came out of the multidimensional
scaling corresponds to the actual (Google) map.

 
Solution


Let's pick a few places from the actual map, and make a table of
where they are on the actual map and the `cmdscale` map:

```

 Place        Real              Cmdscale       
----------------------------------------------
 Superior     northwest         central east   
 St. Paul     central west      southeast      
 Dubuque      central south     central south  
 Chicago      southeast         central west   
 Appleton     central east      central north  


```


This is a bit tricky. Dubuque is the only one in the right place, and
the others that were west have become east and vice versa. So I think
there is a flipping across a line going through Dubuque. That seems to
be the most important thing; if you imagine the other points being
flipped across a line going north-south through Dubuque, they all end
up in about the right place. There might be a little rotation as well,
but I'll call that close enough.

(For you, any comment along the lines of "flipped around this line"
or "rotated about this much" that seems to describe what has happened, is
fine.)
 


(h) Calculate something that demonstrates that a one-dimensional
map of the cities is a much worse representation than the
two-dimensional one that we made before. (I'm planning to get to this
in class, but if we skip it, don't worry about this part.)

 
Solution


Run again with `eig=T` and take a look at `GOF` (uppercase):
```{r }
cmdscale(d, 2, eig = T)$GOF
cmdscale(d, 1, eig = T)$GOF
```

   

The goodness-of-fit of the two-dimensional solution is pretty
good,
`r tufte::margin_note("As it ought to be, since there is a real answer  here: the cities do} exist as locations in two dimensions, if you ignore the curvature of the earth. The goodness of fit isn't  100% because the roads bend a bit.")`
but that of the one-dimensional solution (which arranges all the
cities along a line) is pretty awful in comparison.

How awful? Let's find out. I should have saved it from just above, but
now I have to do it again. For the plot, `ones` is a string of
ones, as many as there are cities.

```{r kashklsahl0}
ones <- rep(1, 12)
v <- cmdscale(d, 1, eig = T)
ddd <- as_tibble(v$points) %>%
  mutate(one = ones, city = cities)
ddd
```

 

(the one-column matrix of points didn't have a name, so it acquired
the name `V1`), and the plot:

```{r }
ggplot(ddd, aes(x = one, y = V1, label = city)) +
  geom_point() + geom_text_repel()
```

 

The cities get mapped onto a line that goes northwest (top) to
southeast (bottom). This is not completely terrible, since there
aren't really any cities in the northeast of the state, but it *is*
pretty awful.
 


