##  Finding wolf spiders on the beach


 <a name="q:wolfspider">*</a> 
A team of Japanese researchers were investigating what would
cause the burrowing wolf spider *Lycosa ishikariana* to be found
on a beach. They hypothesized that it had to do with the size of the
grains of sand on the beach. They went to 28 beaches in Japan,
measured the average sand grain size (in millimetres), and observed
the presence or absence of this particular spider on each beach. The
data are in [link](http://ritsokiguess.site/datafiles/spiders.txt). 



(a) Why would logistic regression be better than regular
regression in this case?
 
Solution


Because the response variable, whether or not the spider is
present, is a categorical yes/no success/failure kind of variable
rather than a quantitative numerical one, and when you have this
kind of response variable, this is when you want to use logistic
regression. 
 
$\blacksquare$

(b) Read in the data and check that you have something
sensible. (Look at the data file first: the columns are aligned but
the headers are not aligned with them.)
 
Solution


The nature of the file means that you need `read_table2`:
```{r wolfspider-1 }
my_url <- "http://ritsokiguess.site/datafiles/spiders.txt"
spider <- read_table2(my_url)
spider
```

     

We have a numerical sand grain size and a categorical variable
`Spiders` that indicates whether the spider was present or
absent. As we were expecting. (The categorical variable is actually
text or `chr`, which will matter in a minute.)
 
$\blacksquare$

(c) Make a boxplot of sand grain size according to whether the
spider is present or absent. Does this suggest that sand grain size
has something to do with presence or absence of the spider?
 
Solution


```{r starcross}
ggplot(spider, aes(x = Spiders, y = Grain.size)) + geom_boxplot()
```

     

The story seems to be that when spiders are present, the sand grain
size tends to be larger. So we would expect to find some kind of
useful relationship in the logistic regression.

Note that we have reversed the cause and effect here: in the boxplot
we are asking "given that the spider is present or absent, how big are the grains of sand?", 
whereas the logistic regression is asking "given the size of the grains of sand, how likely is the spider to be present?". But if one variable has to do with the other, we would
hope to see the link either way around.
 
$\blacksquare$

(d) Fit a logistic regression predicting the presence or
absence of spiders from the grain size, and obtain its
`summary`. Note that you will need to do something with the
response variable first.
 
Solution


The presence or absence of spiders is our response. This is
actually text at the moment:
```{r wolfspider-2 }
spider
```

     

so we need to make a factor version of it first. I'm going to live on
the edge and overwrite everything:

```{r wolfspider-3 }
spider <- spider %>% mutate(Spiders = factor(Spiders))
spider
```

 

`Spiders` is now a factor, correctly. (Sometimes a text
variable gets treated as a factor, sometimes it needs to be an
explicit `factor`. This is one of those times.)
Now we go ahead and fit the model. I'm naming this as
response-with-a-number, so I still have the Capital Letter. Any choice
of name is good, though.

```{r wolfspider-4 }
Spiders.1 <- glm(Spiders ~ Grain.size, family = "binomial", data = spider)
summary(Spiders.1)
```

     
$\blacksquare$

(e) Is there a significant relationship between grain size and
presence or absence of spiders at the $\alpha=0.10$ level? Explain briefly.
 
Solution


The P-value on the `Grain.size` line is *just* less
than 0.10 (it is 0.088), so there is *just* a significant
relationship. 
It isn't a very strong significance, though. This
might be because we don't have that much data: even though we have
28 observations, which, on the face of it, is not a very small
sample size, each observation doesn't tell us much: only whether
the spider was found on that beach or not. Typical sample sizes in
logistic regression are in the hundreds --- the same as for
opinion polls, because you're dealing with the same kind of data.
The weak significance here lends some kind of weak support to the
researchers' hypothesis, but I'm sure they were hoping for
something better.
 
$\blacksquare$

(f) Obtain predicted probabilities of spider presence
for a representative collection of grain sizes.
I only want predicted probabilities, not
any kind of intervals.
 
Solution

```{r}
predictions(Spiders.1, variables = "Grain.size")
```

Another way to do this is to make a data frame of values to predict from directly, using `tribble` or for that matter `datagrid`. For some reason, I chose these four values:

```{r wolfspider-6 }
new <- tribble(
  ~Grain.size,
  0.2,
  0.5,
  0.8,
  1.1
)
new
```

and then

```{r}
predictions(Spiders.1, newdata = new)
```



One of the above is all I need. I prefer the first one, since that way we don't even have to decide what th e representative values are.

Extra:

Note that the probabilities don't go up linearly. (If they did, they
would soon get bigger than 1!). It's actually the *log-odds* that
go up linearly. 

To verify this, you can add a `type` to the `predictions`:

```{r wolfspider-8 }
predictions(Spiders.1, newdata = new, type = "link")
```

This one, as you see shortly, makes more sense with equally-spaced grain sizes. 

The meaning of that slope coefficient in the `summary`, which
is about 5, is that if you increase grain size by 1, you increase the
log-odds by 5. In the table above, we increased the grain size by 0.3
each time, so we would expect to increase the log-odds by
$(0.3)(5)=1.5$, which is (to this accuracy) what happened.

Log-odds are hard to interpret. Odds are a bit easier, and to get
them, we have to `exp` the log-odds:

```{r wolfspider-9 }
predictions(Spiders.1, newdata = new, type = "link") %>% 
  mutate(exp_pred = exp(predicted))
```

 
Thus, with each step of
0.3 in grain size, we *multiply* the odds of finding a spider by
about 

```{r wolfspider-10 }
exp(1.5)
```

 

or about 4.5 times. If you're a gambler, this might give you a feel
for how large the effect of grain size is. Or, of course, you can just
look at the probabilities.
 
$\blacksquare$

(g) Given your previous work in this question, does the trend
you see in your predicted probabilities surprise you? Explain briefly.
 
Solution


My predicted probabilities go up as grain size goes up. This fails
to surprise me for a couple of reasons: first, in my boxplots, I
saw that grain size tended to be larger when spiders were present,
and second, in my logistic regression `summary`, the slope
was positive, so likewise spiders should be more likely as grain
size goes up. Observing just one of these things is enough, though
of course it's nice if you can spot both.
 
$\blacksquare$



