##  The Worcester survey


 The Worcester survey was a long-term study of
all myocardial-infarction^[Heart attack.] victims admitted to hospitals in the
Worcester, Massachusetts area.^[Worcester is pronounced, by locals, *Woo-stuh*.] 
The data have been well studied, and can be found in
the file [link](http://ritsokiguess.site/datafiles/whas100.csv).



(a) Read the data and
 display the first few rows of the data frame. You might get an extra
 column, which you can ignore.
For your information, the variables are:


* patient ID code

* admission date

* date of last followup (this is the date of death if the
patient died)

* length of hospital stay (days)

* followup time (days) (time between admission and last followup)

* followup status: 1=dead, 0=alive

* Age in years (at admission)

* gender (0=male, 1=female)

* body mass index (kg/m$^2$)

 
Solution


```{r worcester-1 }
my_url <- "http://ritsokiguess.site/datafiles/whas100.csv"
whas100 <- read_csv(my_url)
whas100
```

     

I seem to have an extra column called `X1`. This is because I
saved my version of the data using the old `write.csv`, which
comes with row names, and I forgot to
get rid of them. These came back as an extra unnamed variable to which
`read_delim` gave the name `X1`.
 
$\blacksquare$

(b) Create a suitable response variable for a Cox proportional
hazards model for time of survival, using the followup time and
followup status.
 
Solution


`Surv`. The event here is death, so the two parts of the
response variable are followup time `lenfol` and followup
status, 1 being "dead", `fstat`:
```{r worcester-2 }
y <- with(whas100, Surv(lenfol, fstat == 1))
y
```

     

Just using `fstat` alone as the second thing in `Surv`
also works, because anything that gives `TRUE` or 1 when the
event (death) occurs is equally good. (In R, `TRUE` as a number
is 1 and `FALSE` as a number is 0.) 

I listed the values by way of checking. The ones with a `+` are
censored: that is, the patient was still alive the last time the
doctor saw them. Most of the censored values are longer times. Usually
this happens because the patient was still alive at the end of the study.
 

$\blacksquare$

(c) Fit a Cox proportional hazards model predicting survival time
from age, gender and BMI. Obtain the `summary` (but you don't
need to comment on it yet).

 
Solution


This, using the response variable that we just created:
```{r worcester-3 }
whas100.1 <- coxph(y ~ age + gender + bmi, data = whas100)
summary(whas100.1)
```

   
 
$\blacksquare$

(d) Test the overall fit of the model. What does the result mean?
 
Solution


Look at those three P-values at the bottom.  They are all small,
so something in the model is helping to predict survival. As to
what? Well, that's the next part.
 
$\blacksquare$

(e) Can any of your explanatory variables be removed from the
model? Explain briefly.
 
Solution


`gender` has a (very) large P-value, so that can be taken
out of the model. The other two variables have small P-values
(`bmi` only just under 0.05), so they need to stay.
The other way to think about this is `step`, or `drop1`:
```{r worcester-4 }
drop1(whas100.1, test = "Chisq")
```

     

This is here equivalent to^[Not exactly the same as that output, because it  is doing a test that would be the same if you had an infinitely  large sample, but is slightly different with an ordinary finite number of observations.] the output 
from `summary`, but where it
scores is if you have a categorical explanatory variable like
"treatment" with more than two levels: `drop1` will tell you
about keeping or dropping it as a whole.^[Our categorical  variable *gender* has only two levels.]

If you prefer: 

```{r worcester-5 }
step(whas100.1, trace = 0, test = "Chisq")
```

     
`gender` comes out, but the others stay. As usual, put
`trace=1` or `trace=2` to get more output, which will
look like a sequence of `drop1`'s one after the other.
 
$\blacksquare$

(f) Remove your most non-significant explanatory variable from
the model and fit again. Take a look at the results. Are all your
remaining explanatory variables significant? (If all your
explanatory variables were previously significant, you can skip this part.)
 
Solution


So, take out `gender`:
```{r worcester-6 }
whas100.2 <- update(whas100.1, . ~ . - gender)
summary(whas100.2)
```

     

Both explanatory variables are significant: `age` definitely,
`bmi` only just. This is the same model as `step` gave me.
 
$\blacksquare$

(g) Calculate the 1st quartile, median, and 3rd quartiles of
age and BMI. (`quantile`.) Round these off to the
nearest whole number. (Do the rounding off yourself, though R has a
function `round` that does this, which you can investigate if
you want.) As an alternative, you can get these by passing the whole
data frame, or the columns of it you want, into `summary`.
 
Solution


```{r worcester-7 }
quantile(whas100$age)
quantile(whas100$bmi)
```

 

or

```{r worcester-8 }
whas100 %>%
  select(age, bmi) %>%
  summary()
```

 
Or, pure tidyverse: summarize all the columns (after you've done the `select`). 

```{r worcester-9 }
whas100 %>%
  select(age, bmi) %>%
  summarize(across(everything(), ~list(quantile(.)))) %>%
  unnest()
```

 

The reason for the `list()` and the `unnest()` is that
`quantile` returns five numbers rather than just one, so it has
to go into the output data frame as a list-column, which then gets
broken apart with `unnest` so that you can see it. Try it
without the `unnest()` line to see what happens if you don't do
that.
Using whichever of this multitude of ways appeals to you:

60, 71 and 80 for age, 24, 27 and 30 for BMI. 
 
$\blacksquare$

(h) Make a data frame out of all the combinations of
age and BMI values (that you obtained in the previous part) suitable for predicting
with. 
 
Solution


The inevitable `crossing`. This is probably quickest:
```{r worcester-10 }
whas100.new <- crossing(age = c(60, 71, 80), bmi = c(24, 27, 30))
whas100.new
```

     

Or, with some setup beforehand to make the `crossing`
clearer. This is my no-think approach, which is what I recommend;
vectors with plural names containing the values, and inside the
`crossing`, a singular name equals a plural one:

```{r worcester-11 }
ages <- c(60, 71, 80)
bmis <- c(24, 27, 30)
whas100.new <- crossing(age = ages, bmi = bmis)
whas100.new
```

 
 
$\blacksquare$

(i) Obtain predicted survival probabilities for each of the values
in your new data frame. Use your best model. (You don't need to look
at the results, though you can if you want to.)
 
Solution


The magic word is `survfit` (which plays the role of
`predict` here). The best model is `whas100.2`,
with the non-significant `gender` removed:
```{r worcester-12 }
pp2 <- survfit(whas100.2, whas100.new, data = whas100)
```

      

This doesn't need the `data=` at the end (it works perfectly
well without), but the plot (later) seems to need it to be there. I
think the plot needs the information from the original data to be in
the predictions somewhere.

This is kind of long to look at (`summary(pp2)` would be the
thing), so we will need to make a graph of it. I gave it a name, since
I want to use it again later.
 
$\blacksquare$

(j) Make a graph depicting the survival curves from
`survfit` with different colours distinguishing the different
survival curves.
 
Solution


This is actually easy once you work out what to do:

```{r worcester-13 }
ggsurvplot(pp2, conf.int = FALSE)
```

 

Without the `conf.int` thing, you get confidence intervals for
each survival curve, which overlap each other and generally make the
plot look messy.

The "strata" are the different age-BMI combinations that you
predicted for, so it's usually a good idea to list the "new"
prediction data frame, either here or when you assess the effects of
the variables (next part) so that you can see which is which:

```{r worcester-14 }
whas100.new
```

 
 
$\blacksquare$

(k) What is the effect of age on survival? What is the effect
of BMI on survival? Explain briefly. (You will have to disentangle
the meaning of the different coloured lines on the plot to do this.)
 
Solution


Bear in mind that up-and-to-the-right is best for a survival
curve, since that means that people in the upper-right group have
a higher chance of surviving for longer.   

The best survival curve is therefore the olive-green one. According
to the legend, this goes with stratum 3, which is (according to
the listing of `whas100.new`) age 60 (the youngest) and BMI
30 (the highest). So it looks as if the best survival goes with a
lower age (not surprising) and a higher BMI (surprising; see
discussion about BMI below).  

You can also leave one variable constant and see the effects of
changing the other one. Let's pick the oldest age 80: the BMI
values are 24 (stratum 7, blue), 27 (stratum 8, purple), 30
(stratum 9, pink). These survival curves are the bottom one, the
second bottom one, and the fourth bottom one. At this age,
survival chances are not great, but having a higher BMI goes with
a greater chance of surviving longer.  

Or pick a BMI, say 30. These are strata 3 (olive green), 6 (light blue)
and 9 (pink) respectively for ages 60, 71 and 80. These are the
best, 3rd best and 5th best survival curves; that is, as age
increases, the chance of surviving a long time decreases.  

The effect of BMI, though, seems backwards: a higher BMI is
associated with a *higher* chance of survival.   

That's the end of what I wanted you to do, but:

A higher BMI is usually associated with being obese (and therefore
unhealthy), so you'd expect the effect of BMI to be the other way
around. According to Wikipedia
([link](http://en.wikipedia.org/wiki/Body_mass_index)), the BMI values
here are "overweight" or close to it. Maybe being heavier helps
the body recover from a heart attack. 

Let's start with the martingale residual plot:
```{r worcester-15 }
ggcoxdiagnostics(whas100.2) + geom_smooth()
```

     

There is a suspicion of bendiness here, though the left side of the
curve is entirely because of that one positive residual on the
left. In any case, this suggests that nonlinearity (evidently in terms
of BMI, since that's the relationship that currently makes no sense)
would be worth exploring. 

Thus:

```{r worcester-16 }
whas100.3 <- update(whas100.2, . ~ . + I(bmi^2))
summary(whas100.3)
```

     

Ah, that seems to be it. The significant positive coefficient on
`bmi`-squared 
means that the "hazard of dying" increases faster with increasing
`bmi`, so there ought to be an optimal BMI beyond which
survival chances decrease again. 
Have we improved the residuals by adding the squared term?

```{r worcester-17 }
ggcoxdiagnostics(whas100.3) + geom_smooth()
```

 

I call those "inconsequential wiggles" now, so I think we are good.
Let's explore the quadratic relationship on a graph.

I'm going to focus on a close-to-median age of 70, since, in this
model, the effect of BMI is the same for all ages (to make it
different, we would need an interaction term, ANOVA-style). 

First we create a data frame with a bunch of different BMIs in, and
one age 70:

```{r worcester-18 }
bmis <- seq(20, 36, 4)
ages <- 70
whas100.new.2 <- crossing(bmi = bmis, age = ages)
whas100.new.2
```

 

It is rather absurd to have a plural `ages` with only one age
in it, but that's the way it goes, if you're me and trying to avoid thinking.

Predictions, using the model with the squared term in it:

```{r worcester-19 }
pp3 <- survfit(whas100.3, whas100.new.2, data = whas100)
```

 

And then the plot:

```{r worcester-20 }
ggsurvplot(pp3, conf.int = F)
```

 

and the customary reminder of which stratum is which, with its rather
ungainly name:

```{r worcester-21 }
whas100.new.2
```

 

This time, the green survival curve is best, stratum 3, which means that
survival is best at BMI 28, and worse for both higher BMIs and lower
BMIs. You can follow the sequence of colours: red, olive-green, green,
blue, pink, that goes up and then down again. But it's still true
that having a very *low* BMI is worst, which is why our (linear)
model said that having a higher BMI was better.

It would have been better to have you put a squared term in the model,
but the question was already long and complicated enough, and I
didn't want to make your lives more of a nightmare than they are
already becoming!
 
$\blacksquare$



