##  Maximal oxygen uptake in young boys


 A physiologist wanted to understand the relationship between
physical characteristics of pre-adolescent boys and their maximal
oxygen uptake (millilitres of oxygen per kilogram of body weight). The
data are in
[link](http://ritsokiguess.site/datafiles/youngboys.txt) for a
random sample of 10 pre-adolescent boys. The variables are (with
units):



* `uptake`: Oxygen uptake (millitres of oxygen per kilogram
of body weight)

* `age`: boy's age (years)

* `height`: boy's height (cm)

* `weight`: boy's weight (kg)

* `chest`: chest depth (cm).




(a) Read the data into R and confirm that you do indeed have
10 observations.


Solution


```{r }
my_url <- "http://ritsokiguess.site/datafiles/youngboys.txt"
boys <- read_delim(my_url, " ")
boys
```

         

10 boys (rows) indeed.

$\blacksquare$

(b) Fit a regression predicting oxygen uptake from all the
other variables, and display the results.


Solution


Fitting four explanatory variables with only ten observations is likely to be pretty shaky, but we 
press ahead regardless:
```{r }
boys.1 <- lm(uptake ~ age + height + weight + chest, data = boys)
summary(boys.1)
```

         

$\blacksquare$

(c) (A one-mark question.) Would you say, on the evidence so
far, that the regression fits well or badly?  Explain (very)
briefly.


Solution


R-squared of 0.97 (97\%) is very high, so I'd say this
regression fits very well. That's all.
I said "on the evidence so far" to dissuade you from
overthinking this, or thinking that you needed to produce
some more evidence. That, plus the fact that this was only
one mark.

$\blacksquare$

(d) It seems reasonable that an older boy should have a
greater oxygen uptake, all else being equal. Is this supported
by your output?  Explain briefly.


Solution


If an older boy has greater oxygen uptake (the "all else   equal" was a hint), 
the slope of `age` should be
positive. It is not: it is $-0.035$, so it is suggesting
(all else equal) that a greater age goes with a
*smaller* oxygen uptake.
The reason why this happens (which you didn't need, but
you can include it if you like) is that `age` has a
non-small P-value of 0.07, so that the `age` slope
is not significantly different from zero. With all the
other variables, `age` has nothing to *add*
over and above them, and we could therefore remove it.

$\blacksquare$

(e) It seems reasonable that a boy with larger weight
should have larger lungs and thus a \emph{statistically
significantly} larger oxygen uptake. Is that what happens
here? Explain briefly.


Solution


Look at the P-value for `weight`. This is 0.14,
not small, and so a boy with larger weight does not have
a significantly larger oxygen uptake, all else
equal. (The slope for `weight` is not
significantly different from zero either.)
I emphasized "statistically significant" to remind you
that this means to do a test and get a P-value.

$\blacksquare$

(f) Fit a model that contains only the significant
explanatory variables from your first regression. How do
the R-squared values from the two regressions compare?
(The last sentence asks for more or less the same thing as
the next part. Answer it either here or there. Either
place is good.)


Solution


Only `height` is significant, so that's the
only explanatory variable we need to keep. I would
just do the regression straight rather than using
`update` here:
```{r }
boys.2 <- lm(uptake ~ height, data = boys)
summary(boys.2)
```

        

If you want, you can use `update` here, which looks like this:

```{r }
boys.2a <- update(boys.1, . ~ . - age - weight - chest)
summary(boys.2a)
```

 

This doesn't go quite so smoothly here because there are three
variables being removed, and it's a bit of work to type them all. 

$\blacksquare$

(g) How has R-squared changed between your two
regressions? Describe what you see in a few words.


Solution


R-squared has dropped by a bit, from 97\% to 91\%. (Make your own
call: pull out the two R-squared numbers, and say a word or two about
how they compare. I don't much mind what you say: 
"R-squared has decreased (noticeably)", "R-squared has hardly changed". But say
something.)

$\blacksquare$ 

(h) Carry out a test comparing the fit of your
two regression models. What do you conclude, and
therefore what recommendation would you make about the
regression that would be preferred?


Solution


The word "test" again implies something that  produces a P-value with a
null hypothesis that you might reject. In this case, the test that
compares two models differing by more than one $x$ uses
`anova`, testing the null hypothesis that the two regressions
are equally good, against the alternative that the bigger (first) one
is better. Feed `anova` two fitted model objects, smaller first:

```{r }
anova(boys.2, boys.1)
```

 

This P-value of 0.123 is not small, so we do not reject the null
hypothesis. There is not a significant difference in fit between the
two models. Therefore, we should go with the smaller model
`boys.2` because it is simpler. 

That drop in R-squared from 97\% to 91\% was, it turns out, *not*
significant: the three extra variables
could have produced a change in R-squared like that, 
*even if  they were worthless*.^[Recall that adding $x$-variables to a regression will always make R-squared go up, even if they are just random noise.]

If you have learned about "adjusted R-squared", you might recall
that this is supposed to go down *only* if the variables you took
out should not have been taken out. But adjusted R-squared goes down
here as well, from 94\% to 89\% (not quite as much, therefore). What
happens is that adjusted R-squared is rather more relaxed about
keeping variables than the `anova` $F$-test is; if we had used
an $\alpha$ of something like 0.10, the decision between the two
models would have been a lot closer, and this is reflected in the
adjusted R-squared values.

$\blacksquare$

(i) Obtain a table of correlations between all
the variables in the data frame. Do this by feeding
the whole data frame into `cor`. 
We found that a regression predicting oxygen uptake
from just `height` was acceptably good. What
does your table of correlations say about why that
is? (Hint: look for all the correlations that are
*large*.) 


Solution


Correlations first:

```{r }
cor(boys)
```

 
The correlations with `age` are all on the low side, but all
the other correlations are high, not just between `uptake` and the
other variables, but between the explanatory variables as well.

Why is this helpful in understanding what's going on? Well, imagine a
boy with large height (a tall one). The regression `boys.2`
says that this alone is enough to predict that such a boy's oxygen
uptake is likely to be large, since the slope is positive. But the
correlations tell you more: a boy with large height is also (somewhat)
likely to be older (have large age), heavier (large weight) and to have
larger `chest` cavity. So oxygen uptake does depend on those other
variables as well, but once you know `height` you can make a
good guess at their values; you don't need to know them.

Further remarks: `age` has a low correlation with
`uptake`, so its non-significance earlier appears to be
"real": it really does have nothing extra to say, because the other
variables have a stronger link with `uptake` than
`age`. Height, however, seems to be the best way of relating
oxygen uptake to any of the other variables. I think the suppositions
from earlier about relating oxygen uptake to "bigness"^[This  is not, I don't think, a real word, but I mean size emphasizing  how big a boy is generally, rather than how small.] in some sense
are actually sound, but age and weight and `chest` capture
"bigness" worse than height does. Later, when you learn about
Principal Components, you will see that the first principal component,
the one that best captures how the variables vary together, is often
"bigness" in some sense.

Another way to think about these things is via pairwise
scatterplots. The nicest way to produce these is via `ggpairs`
from package `GGally`:

```{r }
boys %>% ggpairs(progress = FALSE)
```

 

A final remark: with five variables, we really ought to have more than
ten observations (something like 50 would be better). But with more
observations and the same correlation structure, the same issues would
come up again, so the question would not be materially changed.

$\blacksquare$

 

