---
title: "Ken's Test"
output: html_notebook
---
---
title: "Ken's Test"
output: html_notebook
---

Here is the built-in data set `mtcars`, with some information about 32 different cars:

```{r}
mtcars
```

We look at a numerical summary of each of the eleven (quantitative) variables:

```{r}
summary(mtcars)
```

For the `mpg` column, the mean is larger than the median, so we suspect that the distribution will be right-skewed. This can be confirmed by looking at a boxplot:

```{r}
boxplot(mtcars$mpg)
```

The longer upper whisker indicates a right skew.

<!--chapter:end:a0-notebook-1.Rmd-->




<!--chapter:end:a1_c33.Rmd-->




<!--chapter:end:a1.Rmd-->




<!--chapter:end:a10_c33.Rmd-->




<!--chapter:end:a10.Rmd-->




<!--chapter:end:a11_c33.Rmd-->




<!--chapter:end:a2_c33.Rmd-->




<!--chapter:end:a2.Rmd-->




<!--chapter:end:a3_c33.Rmd-->




<!--chapter:end:a3.Rmd-->




<!--chapter:end:a4_c33.Rmd-->




<!--chapter:end:a4.Rmd-->




<!--chapter:end:a5_c33.Rmd-->




<!--chapter:end:a5.Rmd-->




<!--chapter:end:a6_2016.Rmd-->




<!--chapter:end:a6_c33.Rmd-->




<!--chapter:end:a6.Rmd-->




<!--chapter:end:a7_c33.Rmd-->


# classoption: answers

Placeholder



<!--chapter:end:a7.Rmd-->




<!--chapter:end:a8_c33.Rmd-->




<!--chapter:end:a8.Rmd-->




<!--chapter:end:a9_c33.Rmd-->




<!--chapter:end:a9.Rmd-->



## packages
## grab the data

<!--chapter:end:airline.Rmd-->




<!--chapter:end:alcohol.Rmd-->




<!--chapter:end:algebra.Rmd-->

```{r}
remotes::install_github("serrat839/mRkov")
```

```{r}
shiny::runGitHub("mRkov_shiny", username = "serrat839")
```


<!--chapter:end:analyze_tweets.Rmd-->




<!--chapter:end:assgt.Rmd-->



## packages
## rainfall

<!--chapter:end:basics.Rmd-->




<!--chapter:end:bayesp.Rmd-->




<!--chapter:end:blood_pressure_paired.Rmd-->




<!--chapter:end:blood_pressure_tidy.Rmd-->




<!--chapter:end:blood_pressure.Rmd-->




<!--chapter:end:boot-irs.Rmd-->



## introduction
## packages
## now, in R, set up these:
## processing
## references

<!--chapter:end:boston-marathon.Rmd-->




<!--chapter:end:braking.Rmd-->




<!--chapter:end:brand_m.Rmd-->


<!--chapter:end:calirain.Rmd-->



## Packages
## Constructing the data

<!--chapter:end:car_speeds.Rmd-->



## Introduction
## Exploratory analysis
## Model 1: a line
## Model 2: a parabola
## Model 3: an asymptote model
## Conclusions

<!--chapter:end:carp.Rmd-->




<!--chapter:end:cholest.Rmd-->




<!--chapter:end:compare_assgt.Rmd-->


<!--chapter:end:county.Rmd-->


<!--chapter:end:crickets.Rmd-->



## packages
## make the data (designed experiment)

<!--chapter:end:crudeoil.Rmd-->




<!--chapter:end:cv.Rmd-->


# A comparison of four shampoos in treating dandruff

Placeholder


## Introduction
## Exploratory analysis
## Analysis of Variance
## Assessment of Assumptions
## Conclusions
## End

<!--chapter:end:dandruff.Rmd-->



## packages
## read in data
## plot caribou and wolf against time with dual y axis

<!--chapter:end:denali.Rmd-->




<!--chapter:end:distance_pref.Rmd-->




<!--chapter:end:dolphins.Rmd-->




<!--chapter:end:drinks.Rmd-->



## packages
## gender bias in grad school admissions at Berkeley
## divorce rate and margarine consumption
## gapminder
## self-reported heights
## gun murders
## missing values
## outlier example
## US election polls
## reported heights
## research funding rates
## falling object (simulated)
## blue and red beads
## contagious diseases

<!--chapter:end:dslabs.Rmd-->


<!--chapter:end:execsal.Rmd-->




<!--chapter:end:exercise.Rmd-->




<!--chapter:end:eyesight.Rmd-->




<!--chapter:end:fbfriends.Rmd-->




<!--chapter:end:flights.Rmd-->



## R Markdown
## Including Plots

<!--chapter:end:fred.Rmd-->




<!--chapter:end:french.Rmd-->


<!--chapter:end:ftccigar.Rmd-->



## packages
## mean and median

<!--chapter:end:funs.Rmd-->



## packages

<!--chapter:end:games-howell.Rmd-->




<!--chapter:end:gpa.Rmd-->



## packages
## the data
## save to read in later
## correlations
## pictures
## multiple regression
## diagnostic graphs

<!--chapter:end:gradrate.Rmd-->




<!--chapter:end:gss_tv.Rmd-->




<!--chapter:end:gss02.Rmd-->




<!--chapter:end:handspan.Rmd-->




<!--chapter:end:handspan2.Rmd-->




<!--chapter:end:heightfoot.Rmd-->



## Packages
## Introduction
## A question
## The length of the `hotpo` sequence
## Further thoughts

<!--chapter:end:hotpo.Rmd-->

---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
mtcars
```

some text

```{r}
ggplot(mtcars, aes(x=cyl, y=mpg)) + geom_point()
```

```{r}
ggplot(mtcars, aes(x=disp, y=mpg)) + geom_point()
```


<!--chapter:end:i.Rmd-->




<!--chapter:end:import1.Rmd-->



## packages
## start again

<!--chapter:end:insect_sprays.Rmd-->


Instructions: Make an R Notebook and in it answer the question or questions below. When you are done, hand in on Quercus the *output* from Previewing (or Knitting) your Notebook, probably an `html` or `pdf` file. An `html` file is easier for the grader to deal with.
Do *not* hand in the Notebook itself. You want to show that you can (i) write code that will answer the questions, (ii) run that code and get some sensible output, (iii) write some words that show you know what is going on and that reflect your conclusions about the data. Your goal is to convince the grader that you *understand* what you are doing: not only doing the right thing, but making it clear that you know *why* it's the right thing.

Do *not* expect to get help on this assignment. The purpose of the assignments is for you to see how much *you* have understood. You will find that you also learn something from grappling with the assignments. The time to get help is after you watch the lectures and work through the problems from PASIAS, via tutorial and the discussion board, that is *before* you start work on the assignment.  The only reasons to contact the instructor while working on an assignment are to report (i) something missing like a data file that cannot possibly be read, (ii) something *beyond your control* that makes it impossible to finish the assignment in time after you have started it.

There is a time limit on this assignment (you will see Quercus counting down the time remaining).


<!--chapter:end:instructions.Rmd-->




<!--chapter:end:isoflavones.Rmd-->




<!--chapter:end:jocko.Rmd-->

---
title: "Testing out jsonlite"
output: html_notebook
---

Array of json objects

```{r}
library(jsonlite)
arry='[{"name":"Erik", "age":43}, {"name":"Anna", "age":32}]'
arry
```

simplify into data frame?

```{r}
fromJSON(arry)
```

Toronto extreme heat days: read from file

```{r}
heat=scan("heat",what = "text")
heat
fromJSON(heat)
```


<!--chapter:end:json.Rmd-->




<!--chapter:end:match_rnw_rmd.Rmd-->




<!--chapter:end:mean_two.Rmd-->




<!--chapter:end:mercury.Rmd-->




<!--chapter:end:mindless.Rmd-->



## packages
## get original data
## boxplot

<!--chapter:end:mm.Rmd-->



## packages
## read data

<!--chapter:end:movies.Rmd-->



## packages

<!--chapter:end:nbapoints.Rmd-->

Normal table
========================================================

```{r,results="asis"}
u=seq(0,3.09,by=0.01)
p=pnorm(u)
m=matrix(p,ncol=10,byrow=TRUE)
options(digits=4)
library(xtable)
newm=xtable(m,digits=4)
print.xtable(newm, type="html")

```


<!--chapter:end:normal-table.Rmd-->


<!--chapter:end:oj.Rmd-->




<!--chapter:end:ozvus.Rmd-->




<!--chapter:end:pasiasify.Rmd-->



## packages
## read data

<!--chapter:end:picture-memory.Rmd-->



## packages

<!--chapter:end:play2.Rmd-->




<!--chapter:end:pop_power.Rmd-->

\usepackage{float}
\newcommand{\bQ}{\begin{questions}}
\newcommand{\eQ}{\end{questions}}

<!--chapter:end:preamble.Rmd-->




<!--chapter:end:problems.Rmd-->



## Introduction
## Data and pre-processing
## Analysis
## Conclusions (see note 9)

<!--chapter:end:proggo.Rmd-->




<!--chapter:end:protein.Rmd-->




<!--chapter:end:pulsemarch.Rmd-->



## Navigate to the course assignment page
## The page for the assignment you want to hand in
## Uploading the file
## Later

<!--chapter:end:quercus1.Rmd-->



## Log in to Quercus
## To grade an assignment
## Showing and hiding grades

<!--chapter:end:quercus2.Rmd-->




<!--chapter:end:quiz.Rmd-->


# classoption: answers

Placeholder



<!--chapter:end:quorko.Rmd-->


<!--chapter:end:r0.Rmd-->




<!--chapter:end:rainfall.Rmd-->




<!--chapter:end:rainfall2.Rmd-->




<!--chapter:end:reg.Rmd-->




<!--chapter:end:reggae.Rmd-->




<!--chapter:end:relig.Rmd-->




<!--chapter:end:reneg.Rmd-->



## Packages
## read in data
## plot

<!--chapter:end:repellent.Rmd-->




<!--chapter:end:rescale.Rmd-->



## packages
## read in file, no header

<!--chapter:end:rl.Rmd-->




<!--chapter:end:roethlisberger.Rmd-->




<!--chapter:end:rsimple.Rmd-->



## The data
## Exploratory analysis
## Regression analysis
### Conclusions

<!--chapter:end:sh.Rmd-->

---
title: "sign power"
output: html_notebook
---

estimating power of sign test vs. t when mean is 50 and SD 15, true mean 40, sample size 10

```{r}
library(tidyverse)
set.seed(457299)
```

```{r}
rerun(1000,rnorm(10,50,15)) %>% 
  map(~tibble(x=.)) %>% 
  map(~count(.,x<40)) %>% 
  map(~summarize(.,the_min=min(n))) %>% 
  map(~mutate(.,is_rejected=(the_min<=1 | the_min==10))) %>% 
  map_lgl(~pull(.,is_rejected)) %>% 
  tibble(was_true=.) %>% 
  count(was_true)
```


<!--chapter:end:signpower.Rmd-->



## Introduction
## Analysis
### Introduction
### Tidying
### Exploratory analysis
### Inference
## Conclusion

<!--chapter:end:skin-cream-report.Rmd-->




<!--chapter:end:sleeping.Rmd-->


<!--chapter:end:soap.Rmd-->




<!--chapter:end:sparrowhawks.Rmd-->




<!--chapter:end:students.Rmd-->

---
title: "Ken's Test"
output:
  pdf_document: default
  html_notebook: default
  word_document: default
---

Here is a data set. I have no idea where it comes from:

```{r}
x=c(10,11,13,17,22,29)
x
```

We summarize the data:

```{r}
summary(x)
```

The mean is a little higher than the median, and the highest value is a fair bit higher than the third quartile, suggesting a distribution that is skewed to the right. To investigate further, we can look at a boxplot:

```{r}
boxplot(x)
```

The long upper whisker further supports the idea of right-skewedness.

<!--chapter:end:test.Rmd-->



## packages
## data
## save the data

<!--chapter:end:touch.Rmd-->




<!--chapter:end:trees.Rmd-->




<!--chapter:end:util_temp.Rmd-->




<!--chapter:end:wood_beams.Rmd-->




<!--chapter:end:woof.Rmd-->




<!--chapter:end:writers.Rmd-->



## xxx title here

<!--chapter:end:xxx.Rmd-->


<!--chapter:end:youngboys.Rmd-->

---
title: "Untitled"
author: "Ken"
date: "05/05/2021"
output: html_document
---

An ordered lettered list:

a) some stuff
with some text under

a) some more stuff
with yet more text under and two blank lines

an odd paragraph

a) yet more

<!--chapter:end:zzz.Rmd-->

