---
title: "Assignment 7"
author: ""
date: ''
documentclass: exam
# classoption: answers
output: 
  pdf_document: 
    includes:
      in_header: preamble.tex
    latex_engine: lualatex
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(smmr)
library(PMCMRplus)
library(MASS)
library(readxl)
library(conflicted)
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
```

Instructions (the same as for Assignment 1): Make an R Notebook and in it answer the two questions below (one Notebook for both questions). When you are done, hand in on Quercus the *output* from Previewing (or Knitting) your Notebook. Do *not* hand in the Notebook itself. You want to show that you can (i) write code that will answer the questions, (ii) run that code and get some sensible output, (iii) write some words that show you know what is going on and that reflect your conclusions about the data. Your goal is to convince the grader that you *understand* what you are doing: not only doing the right thing, but making it clear that you know *why* it's the right thing.

Do *not* expect to get help on this assignment. The purpose of the assignments is for you to see how much *you* have understood. You will find that you also learn something from grappling with the assignments. The time to get help is after you watch the lectures and work through the problems from PASIAS, via tutorial and the discussion board.  The only reason to contact the instructor while working on the assignments is to report something missing like a data file that cannot be read.

You have 5 hours to complete this assignment after you start it. 

My solutions to this assignment, with extra discussion, will be available after everyone has handed in their assignment.



\bQ


```{r utils, child="util_temp.Rmd"}
```


```{r dolphins, child="dolphins.Rmd"}
```



\eQ

<!-- \printendnotes -->


