---
title: "Assignment questions"
author: "Ken"
date: '`r Sys.Date()`'
documentclass: exam
classoption: answers
output: 
  pdf_document: 
    includes:
      in_header: preamble.tex
    latex_engine: lualatex
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(smmr)
library(PMCMRplus)
library(MASS)
library(readxl)
library(conflicted)
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
```



\bQ

\section{Reading in data, graphs and summaries}

some of this. think about open-ended questions.

```{r rainfall, child="rainfall.Rmd"}
```

```{r rainfallx, child="rainfall2.Rmd"}
```

```{r exercise, child="exercise.Rmd"}
```

\section{One-sample inference}  

```{r cholest, child="cholest.Rmd"}
```

```{r trees, child="trees.Rmd"}
```


\section{Two-sample inference}

```{r handspan, child="handspan.Rmd"}
``` 

```{r ozvus, child="ozvus.Rmd"}
```


\section{Power}

```{r mean_two, child="mean_two.Rmd"}
```

```{r pop_power, child="pop_power.Rmd"}
```


\section{Matched pairs}

```{r french, child="french.Rmd"}
```

```{r blood_pressure_paired, child="blood_pressure_paired.Rmd"}
```


\section{Mood's median test}

```{r handspan2, child="handspan2.Rmd"}
``` 

\section{ANOVA}

```{r dandruff, child="dandruff.Rmd"}
```

```{r writers, child="writers.Rmd"}
```


\section{Mood's median test for anova}

```{r reggae, child="reggae.Rmd"}
```

```{r gss_tv, child="gss_tv.Rmd"}
```


\section{Welch anova and games Howell}

```{r relig, child="relig.Rmd"}
```


\section{report writing}

\section{tidying with pivot-longer}

```{r jocko, child="jocko.Rmd"}
```


```{r isoflavones, child="isoflavones.Rmd"}
```

```{r blood_pressure_tidy, child="blood_pressure_tidy.Rmd"}
```


\section{select, slice, filter, sample\_n, sample\_frac, filter, arrange, mutate, group\_by, summarize, pluck, pull, left\_join}

\section{regression}

```{r heightfoot, child="heightfoot.Rmd"}
```

```{r braking, child="braking.Rmd"}
```

\section{multiple regression}

```{r mercury, child="mercury.Rmd"}
```



\section{anova as regression}

```{r pulsemarch, child="pulsemarch.Rmd"}
```




\section{functions}

```{r rescale, child="rescale.Rmd"}
```


\section{map}

eg fitting several different models on the same data

\section{dates and times}

```{r reneg, child="reneg.Rmd"}
```


\eQ

\printendnotes


