---
title: "Endogenous nitrogen excretion in carp"
author: "Ken Butler"
date: "July 1, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

A paper in Fisheries Science reported on variables that
affect ``endogenous nitrogen excretion'' or ENE in carp raised in
Japan. A number of carp were divided into groups based on body weight,
and each group was placed in a different tank. The mean body weight of
the carp placed in each tank was recorded. The carp were then fed a
protein-free diet three times daily for a period of 20 days. At the
end of the experiment, the amount of ENE in each tank was measured, in
milligrams of total fish body weight per day. Ten tanks were used.

The aim in this report is to find a satisfactory model that predicts ENE from body weight.

## Exploratory analysis

We begin by reading in the data:

```{r}
library(tidyverse)
my_url="http://ritsokiguess.site/datafiles/carp.txt"
carp=read_delim(my_url," ")  
carp
```

There are indeed ten tanks. 

With two quantitative variables, a suitable graph is a scatterplot. We are taking ENE as the response, so:

```{r}
ggplot(carp,aes(x=bodyweight,y=ENE))+geom_point()+
  geom_smooth()
```

The trend is downward: as bodyweight increases, ENE
decreases. However, the decrease is rapid at first and then levels
off, so the relationship is nonlinear.

## Model 1: a line

Given our observations above, a straight line should be a poor description of the relationship, but we begin by exploring this:

```{r}
carp.1=lm(ENE~bodyweight,data=carp)
summary(carp.1)
```

R-squared is 47%, which is not that good.

We check the residuals for this model by plotting them against the fitted values:

```{r}
ggplot(carp.1,aes(x=.fitted,y=.resid))+geom_point()
```

This is a clear curve, which means the original relationship must have been a curve. So let's try a second model, a parabola, by adding bodyweight-squared.

## Model 2: a parabola

```{r}
carp.2=lm(ENE~bodyweight+I(bodyweight^2),data=carp)
summary(carp.2)
```

R-squared has gone up from 47\% to 74\%, a substantial
improvement. This suggests to me that the parabola model is a
substantial improvement. This is confirmed by the coefficient for the squared term being significantly nonzero: the parabola is definitely better than the straight line.

Let's make a plot with the data and fitted curve on it. This is most easily done by "fortify"-ing the regression object. What
that means, in the context of `ggplot`, is to add the original
data back onto the regression object, so that we can plot any
combination of original data and values derived from the
regression. That goes like this:

```{r}
carp.2.fort=fortify(carp.2)
carp.2.fort
```

and then we make the plot thus:

```{r}
ggplot(carp.2.fort,aes(x=bodyweight,y=.fitted))+
  geom_line(colour="blue")+
  geom_point(aes(y=ENE))
```

The points in the middle are above the
curve, and the points at the ends are below. (That is, negative
residuals at the ends, and positive ones in the middle.) Another problem is that the parabola curve
fails to capture the \emph{shape} of the relationship; for example, I
see nothing much in the data suggesting that the relationship should go
back up, and even given that, the fitted curve doesn't go especially
near any of the points.

Once again, we need to look at the residuals:

```{r}
ggplot(carp.2,aes(x=.fitted,y=.resid))+geom_point()
```

I think this is \emph{still} a curve (or, it goes down and then
sharply up at the end). Either way, there is still a pattern. 

Thus, our parabola was a
curve all right, but it appears not to be the right *kind* of
curve. I think the original data looks more like a hyperbola (a curve
like $y=1/x$) than a parabola, in that it seems to decrease fast and
then gradually to a limit, and *that* suggests, as in the class
example, that we should try an asymptote model. (This is upside down compared to the windmill one, but it's same idea of tending to a limit.)

## Model 3: an asymptote model

```{r}
carp.3=lm(ENE~I(bodyweight^(-1)),data=carp)
summary(carp.3)
```

That fits *extraordinarily* well, with an R-squared up near
98%. The intercept is the asymptote, which suggests a (lower) limit
of about 5.2 for ENE (in the limit for large bodyweight). We
would have to ask the fisheries scientist whether this kind of thing
is a reasonable biological mechanism. It says that a carp always has
some ENE, no matter how big it gets, but a smaller carp will have a
lot more.

Does the fitted value plot look reasonable now? The "fortify" thing
doesn't quite work here, since `bodyweight` doesn't itself get
added to the fortified data set (it is not part of the model), so it
looks as if we are working with two data frames. I cheated and pulled
the things I wanted out of `carp` without doing a
`data=`:

```{r}
ggplot(carp.3,aes(x=carp$bodyweight,y=.fitted))+
  geom_line(colour="blue")+
  geom_point(aes(y=carp$ENE))
```

I'd say that does a really nice job of fitting the data. But it would
be nice to have a few more tanks with large-bodyweight fish, to
convince us that we have the shape of the trend right.

And, as ever, the residual plot:

```{r}
ggplot(carp.3,aes(x=.fitted,y=.resid))+geom_point()
```

All in all, that looks pretty good (and certainly a vast improvement
over the ones we got before).

## Conclusions

The nonlinear nature of this relationship --- a sharp drop followed by a levelling off --- requires the right kind of nonlinear model, in this case an asymptote model. We found that a straight line and a parabola were not appropriate, both by looking at the residual plots and by looking at R-squared, but the asymptote model has a very high R-squared and a satisfactory residual plot. For these data, the asymptote model, fitted as our model 3, appears to be the best.
