\question Some students in a Statistics class were asked how many
minutes they typically exercised in a week. The data are shown [here](http://ritsokiguess.site/datafiles/STAC32/exercise.txt).

  Some of the
students were male and some were female. Our concern is how the males
and females compare in terms of the amount of exercise they do.

\bP
  \p Take a look at the data file. (Click on the link, or paste the copied link into your web browser.) How is it laid out?
  
  \bS
    Aligned in columns. Or, separated by spaces, but a variable number
    of them. (The latter is a hint that \verb=read_delim= will not
    work, and the former is a hint about what \emph{will} work.)
  \eS
  
  \p Read in and display (some of) the data. (This means to
  display enough of what you read in to convince others that you read
  in the right kind of thing.)
  
  \bS
    "Aligned in columns" is the best way of saying it so that you
    know to use \verb=read_table=. Hence:
    
```{r}
my_url <- "http://ritsokiguess.site/datafiles/STAC32/exercise.txt"
exercise <- read_table(my_url)
exercise
```
    
Just entering the data frame name displays the first ten rows, which
is usually enough to convince anyone that you have the right thing. 

Call the data frame what you like, as long as it describes what's in
the data frame in some way.

Note: Using \verb=read.table= with a dot
    reveals that you have not been paying attention in this class. The instructions in this class (see course outline pages 14-16) say that I expect you to do things as they are done in this class, not as you may have learned elsewhere, so don't expect credit if you use `read.table` here or anywhere in the class; indeed, if you don't say where you got it from (since you certainly didn't get it from me) you are guilty of *plagiarism*, and you definitely don't want to get involved with that. See [this](https://governingcouncil.utoronto.ca/secretariat/policies/code-behaviour-academic-matters-july-1-2019)     section B.I.1(d) and Appendix A, section 2(p).

  \eS
  
  \p Make a suitable plot of this data frame. 
  
  \bS
    Two variables, one quantitative and one categorical, means that a
    boxplot is the thing:
    
```{r}
ggplot(exercise, aes(x=gender, y=minutes)) + geom_boxplot()
```
    
  \eS
  
  \p Does there appear to be any substantial difference in the
  average amount of time that males and females spend exercising?
  Explain briefly. ("average" could be mean or median. Which is it
  here?)
  
  \bS
    A boxplot shows the median. So we learn here that the median time
    spent exercising per week is very slightly higher for
    females. However, there is a lot of variability (the height of the
    boxes), and so the difference between the medians is very
    small. Hence, there is definitely \emph{no} substantial difference
    between males and females. 
    
  Less insightfully, the difference in median between males and females is
    very small (but that doesn't rule out the possibility that the
    spread is very small also).
  \eS
  
  \p How do you know that both distributions, for males as well
  as females, are skewed to the right? Explain (very) briefly.
  
  \bS
    The upper whiskers, at the top of the box, are longer than the
    lower ones (at the bottom).
  \eS
  
  \p For data like this, why does it make practical sense that
  the distributions are skewed to the right?
  
  \bS
    Nobody can exercise less than 0 minutes per week, but there is no
    upper limit: a student in the class can exercise as much as they
    want. This means that there could be unusually high values, but
    not unusually low values. 
    
  Extra: Distributions that have a limit on one side are often skewed to the other side. 
    This one has a limit on the left, so it is skewed to the *right*. This is most likely to be true if 
    there are observations close to the limit, such as the people in this data set that don't exercise at all (and there are some of those).
  
    
  \eS
  
  
  
\eP
