---
title: The relationship between the number of Facebook friends and grey matter in
  the brain
author: "Ken Butler"
date: "October 17, 2016"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Introduction
============

Is there a relationship between the number
of Facebook friends a person has, and the density of grey matter in
the areas of the brain associated with social perception and
associative memory? To find out, a 2012 study measured both of these
variables for a sample of 40 students at City University in London
(England). The grey
matter density is on a $z$-score standardized scale.

Scatterplot
============

We read in the data and make a scatterplot, to see if there is any trend:

```{r}
fb=read.table("facebook.txt",header=T)
library(tidyverse)
ggplot(fb,aes(x=GMdensity,y=FBfriends))+geom_point()+geom_smooth()
```

      There seems to be a weak, upward, apparently linear
      trend. I find that the points are not especially close to the trend.
      
Regression
==========

Next, we fit a regression for predicting the number of Facebook friends from the grey matter density:
      
```{r}
fb.1=lm(FBfriends~GMdensity,data=fb)
summary(fb.1)
```

The regression has a disappointing R-squared (as we would expect from the weakness of the trend on the scatterplot), but the slope is significantly non-zero, indicating that there really is a relationship between number of Facebook friends and grey matter density in the population of which these students are a sample. 

I am surprised that the P-value for the slope is so small, given the weakness of the trend on the scatterplot. Perhaps, with $n=40$ observations, even this weak trend is strong enough to be significant. 

Having found a significant regression, we can obtain a scatterplot of the data with the regression line on it:

```{r}
ggplot(fb,aes(x=GMdensity,y=FBfriends))+geom_point()+
  geom_smooth(method="lm")
```

Insofar as the trend is weak, the regression line appears to do a satisfactory job of capturing it.

Residual analysis
=================

Before accepting this as our final model, we should check by a residual analysis that our model is satisfactory. Plotting the residuals against the fitted values will indicate whether there are any problems with our model:

```{r}
ggplot(fb.1,aes(x=.fitted,y=.resid))+geom_point()
```

I see a completely random scatter of points, and therefore my linear regression model is satisfactory. I see no need to improve it.

