---
title: "Sparrowhawks"
author: "Ken Butler"
date: "October 13, 2015"
output: html_document
---

The data
========
  
We begin by reading in the data:
  
```{r}
library(tidyverse)
my_url="http://ritsokiguess.site/datafiles/sparrowhawk.txt"
sparrowhawks=read_delim(my_url," ")
sparrowhawks
```

There are 13 rows of data altogether, with a value of the number of returning and new adults for each one.

Exploratory analysis
====================

We should look at a scatterplot of the data, to see what, if any, relationship there is. We add a smooth trend, to guide the eye:

```{r}
ggplot(sparrowhawks,aes(x=returning,y=newadults))+geom_point()+geom_smooth()
```

This seems to be something of a downward trend, though it is not completely convincing.  That is, a larger percentage of returning adults is associated with a *smaller* number of new adults. Is the relationship straight? Well, the lowess curve has a strange wiggle in the middle, but the trend overall does not seem obviously curved.

What is the correlation between these two variables?

```{r}
with(sparrowhawks,cor(newadults,returning))
```

The correlation is about -0.75. This seems close to -1 to me, since I didn't think the trend was that strong, but it is no surprise that it is negative, since the trend is definitely downward rather than upward.


Regression analysis
===================

The regression for predicting the number of new adults from the percentage of returning adults is as shown:

```{r}
newadults.1=lm(newadults~returning,data=sparrowhawks)
summary(newadults.1)
```

The intercept is about 32 and the slope is about -0.3. This means that each additional one percentage point in returning adults is associated with a *decrease* of approximately 0.3 new adults. 

The P-value attached to the slope is about 0.003. This means that the slope is *not* zero, and that there is a significant association between the number of returning adults and the number of new adults. (This is despite the apparently weak trend, and the small number, 13, of data points.)

A plot of the data with the regression line superimposed is as shown:

```{r}
ggplot(sparrowhawks,aes(x=returning,y=newadults))+geom_point()+geom_smooth(method="lm")
```

Conclusions
===========

The association is negative, as seen from the scatterplot and
correlation, and also from the negative slope of the regression
line. This suggests that sparrowhawks are long-lived territorial
birds rather than short-lived birds. We would expect to observe
returning birds displaying territorial behaviour that would
discourage new adults from joining the colony.
  
